### 2018-04-11 Draft3.tex
added these manually online

\documentclass[12pt]{amsart}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{lscape}
\usepackage{afterpage}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}\geometry{a4paper} % or letter or a5paper or ... etc
\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.tiff}{png}{.png}{convert #1 \OutputFile}
\AppendGraphicsExtensions{.tiff}

\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
\newcommand{\latcomment}[1]{\textcolor{blue}{LAT: #1}}
\newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}

% \geometry{landscape} % rotated page geometry

% See the ``Article customise'' template for come common customisations

\title{Legacy Data Confounds Modern Studies}
\author{Luke Anderson-Trocm\'e , Simon Gravel}
%\date{} % delete this line to display the current date

%%% BEGIN DOCUMENT
\begin{document}

\maketitle
\tableofcontents
\clearpage
			\section{Introduction}
The last 5 years has seen an increase in number of individuals genotyped through private companies as well as medical research cohorts. Genotype data is often imputed to increase power for association studies in a cost effective way. 
This process is done using a reference database comprised of whole genome sequence data from other individuals in reference populations. 
We often combine these reference databases to increase the sample size and in turn, the accuracy of imputation. Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.  
The errors in one dataset don't disappear when they are combined with another higher quality one. 
Many of the legacy datasets have been known to contain a higher rate of false positives than their more recent counterparts. 
Is it time to retire legacy datasets?

In investigating a mutational signature identified in a Japanese cohort by Harris et. al (2017), we noticed that the individuals responsible for this signal all had lower average quality scores compared to the other Japanese individuals. 
To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
We discovered 000 variants that were associated to low mapping quality in \todo{000} individuals from 000 of the 26 populations included in the 1000 genomes dataset. 
One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.

\subsection{Context}
This project was motivated by an unusual population genetics observation in a recent publication by Harris et. al, that the 1000 Genomes Project Japanese population seem to be partitioned in two clusters of diff mutation rates. 
This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.

Our research center had recently obtained access to a large cohort of several thousand high-quality whole genome sequences from Japanese individuals from the Fumi institute. 
We started off by trying to replicate the results found in Harris et al. 2017, but after several attempts and double checking we had the right quality controls, we were unable to replicate their results with the Fumi dataset. 
Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population. 
Further investigation in the allele frequency spectra of the two Japanese datasets, indicated that the SNPs responsible for the signal were missing in the more recent dataset. 
This result raised several red flags. 
Why would a supposedly genome wide signal be concentrated in private SNPs from a legacy dataset?

Upon investigating the quality control pipelines of the more recent dataset, many of these positions uniquely found in the 1000 Genomes Project had been removed in the standard QC of the sequencing centre. 
Does this legacy dataset meet the current industry standards of quality? 
This study is the result of an investigation in the quality of the 1000 Genomes Project dataset.

After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual. 
This suggested that some individuals with lower quality were driving this apparent population genetics signal. [mention fake pop structure]
Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.

			\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
The date of completion of each sequencing occurrence was not available, however we managed to find the submission dates and sequencing centres for each sample in the phase3 analysis sequence index file.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure. [Average Quality of mapped bases: How was it calculated?]

\subsection{Data Availability}

Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here} 
[link to my compiled metadata file here]

\subsection{Quality Controls}
We reproduced the quality control pipelines used by Harris et. al as they applied the current state of the art quality thresholds to remove questionable sequences especially for the high standards for detecting population level differences. 
Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or basepair complexity compared to the rest of the genome. 
The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker. 
Furthermore, only diallelic autosomal SNPs were considered, with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9.

\subsection{Genome Wide Association}

Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1000 Genomes Project. We used the average quality of mapped bases per individual as the phenotype for the analysis. We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data. [NOT DONE YET]

\subsection{Mutation Spectrum}
We calculated the mutation spectrum for each list of significant SNPs for each population [NOT DONE YET]

We also compared the mutation spectrum ratio between populations using a modified version of the methods used in Harris et al. 2017.

\subsection{Imputation}
Using the Michigan Imputation Server, we imputed the genotype data from 1000 Genomes Project

			\section{Results}
\subsection{Average quality of mapped snps Quality Over Time}
As one would expect, the sequencing quality of individuals increases over time \ref{Figure1}. Sequencing centers improve on their laboratory techniques and sequencing technologies improve from one generation to the next. The magnitude and impact of the differences in quality within and between populations has not yet been explored.

\subsection{Genome Wide Association}
GWAS are commonly used to find genomic regions associated with biologically relevant traits. In this case, we are using this analysis in a more unconventional way to identify regions of the genome that are associated to biologically irrelevant traits like individuals with low average quality of mapped bases.

Using the linear GWAS function offered by PLINK, we were able to identify 000 SNPs that were associated to low quality with  p-values below $10^{-8}$ and 000 snps with p-values below $10^{-6}$ \ref{Figure2}. 
The results from of the GWAS for other populations are available in supplementary data. 
A total of 0000 SNPs were associated to low quality in all populations combined.

\subsection{Overlap of Significant SNPs}
Comparing the results of each independent GWAS, we were able to identify over 0000 variants that were independently associated to low quality in multiple populations. This confirmation using more than one GWAS is a very strong result suggesting that these variants might not be legitimate  \ref{Figure3}.

\subsection{Significant SNPs}
When we remove individuals below a quality of 30, population structure is clarified. 
The signal identified in Harris et. al is enriched in sites associated with low quality. [Need to run harris pipeline without $10^{-6}$ ]
Therefore, when these low quality sites are removed, the mutational signal is also removed. (Or in this case when we remove the individuals with qualities below 30.)

\subsection{Higher quality datasets don't have these SNPs}

Interestingly, the SNPs we identified had already been removed in higher quality datasets. 
These HQ datasets also have filtered more SNPs than the GWAS was able to identify. 
When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWAS. 
This suggests that we lack the power to identify all the questionable variants from these legacy datasets. 
\subsection{Imputation}
~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These  should be removed from reference database

\subsection{Found to be included in other GWAS}
Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWA studies that might have called these erroneous variants as being significantly correlated with some biological trait. 
Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found 6 recent publications that had found at least one of the variants to have reached genome wide significance in their study.

Five of these studies used the 1000 Genomes Project as the reference database for imputation and one used the 1000 Genomes Project cell cultures and sequence data. 
They used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, allele frequency differences using reference populations. 
They also removed rare alleles and alleles with high degrees of missingness. 
Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also to reach genome wide significance for the traits they were testing for.

			\section{Discussion}
\subsection{Why do we care?}
These SNPs matter because they reached genome wide significance for medically relevant traits. 
Could lead to false diagnosis at worst, or spurious correlations at least. 
They were not caught sooner because the batch effect of 1000 genomes had never been properly addressed.

Our GWAS method was able to identify 000 variants that are likely to be false positives.

We compared the number of SNPs identified in the GWAS to the number of sites that differed between our higher quality Japanese dataset and found that 100?\% of the sites were also missing from the Fumi data. 
Because we are running the GWAS on a per-population base, for rarer mutations, we are lacking the power to identify these significant SNPs.  
However, upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified 000 more variants that are beyond the expected frequency spectrum deviation for individuals from the same population. 
This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts. 
The most conservative approach would be to remove all individuals that don't meet the quality threshold. 
In this case, we used a cut off of an average quality of mapped bases over 30. This threshold has been previously used by 000 studies.

\subsection{GWAS from other papers}
Since these variants are present in more than one population from the 1000 genomes project, they are more likely to be associated to biological traits as they would appear to be like any other variant that is shared among multiple populations. 
The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual. 
\subsection{Why were these SNPs even associated in the first place?}
[due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?]

			\section{Conclusion}
Our method identifies spurious mutations by correlating mutations with data quality metrics. 
We propose including our quality control methods to identify possible false positives in sequencing data. 
We have focused on the 1000 Genomes Project dataset as its quality metrics were feely available, however the issues of quality control are not limited to this one consortium. 
This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.

As more and more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the amplification of false positives. 
These results bring forth many questions regarding the reliability of legacy datasets. 
Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.

\afterpage{
\begin{figure}
\includegraphics[width=\hsize]{figures/MapQualOverTime.jpg}

\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project. The average mapping quality per individual was obtained from the raw alignment index data for each sample. The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual. B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
\label{Figure1}
\end{figure}
\clearpage
}

\afterpage{
\begin{landscape}
\begin{figure}[]
\centering
\includegraphics[width=23cm,keepaspectratio, right]{figures/GWAS_Qual_GenomeWide_JPT.jpg}
\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project. This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project. The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
 \label{Figure2}
\end{figure}
\end{landscape}

\clearpage

  }

\afterpage{
\begin{figure}[h]
\centering
\includegraphics[width=\hsize,keepaspectratio, right]{figures/SNPOverlap6.jpg}

\caption{Overlap of SNPs identified independently to be associated with quality. Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants. What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations. }
  \label{Figure3}
\end{figure}
\clearpage
}

\end{document}

### 2018-04-11 Draft3.tex
more words

@@
 \newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
-\newcommand{\latcomment}[1]{\textcolor{blue}{LAT: #1}}
+\newcommand{\lat}[1]{\textcolor{blue}{LAT: #1}}
 \newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}
@@
 			\section{Introduction}
-The last 5 years has seen an increase in number of individuals genotyped through private companies as well as medical research cohorts.
+\subsection{Larger Sample Sizes}			
+The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
+\lat{not a fan of this sentence}
+Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
+Mutations associated to a disease can be identified by comparing individuals afflicted by a disease to healthy individuals. 
+The mutations more common in one group might be associated to the disease.
+Demonstrating that these mutations are biologically relevant to the disease can be difficult.
+Certain factors like age, sex or ancestral origin might be correlated to the trait in question and could potentially be mistakenly associated to a disease. 
+Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
+For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies. 
+Benign mutations at high frequencies in one populations might be exceedingly rare in another population. 
+Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.
+
+\subsection{Imputation}
+ improvements in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases. 
 Genotype data is often imputed to increase power for association studies in a cost effective way. 
@@
 
+\lat{sentence about harris} 
+A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
+The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
+Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
+They argued this by showing that relative difference in mutational types between and even within continental populations.  
 In investigating a mutational signature identified in a Japanese cohort by Harris et.
@@
 To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
-We discovered 000 variants that were associated to low mapping quality in \todo{000} individuals from 000 of the 26 populations included in the 1000 genomes dataset. 
+We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
 One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
@@
 
-			\section{Methods}
-\subsection{Metadata}
-The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
-Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
-Each BAS file has metadata regarding each sequencing event for each sample. 
-If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
-The date of completion of each sequencing occurrence was not available, however we managed to find the submission dates and sequencing centres for each sample in the phase3 analysis sequence index file.
-This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
-The dates of the sequencing are only used to plot Figure. [Average Quality of mapped bases: How was it calculated?]
-
-\subsection{Data Availability}
-
-Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
-Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here} 
-[link to my compiled metadata file here]
-
-\subsection{Quality Controls}
-We reproduced the quality control pipelines used by Harris et.
-al as they applied the current state of the art quality thresholds to remove questionable sequences especially for the high standards for detecting population level differences.
-Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or basepair complexity compared to the rest of the genome. 
-The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
-Furthermore, only diallelic autosomal SNPs were considered, with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9.
-
-\subsection{Genome Wide Association}
-
-Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1000 Genomes Project.
-We used the average quality of mapped bases per individual as the phenotype for the analysis.
-We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data.
-[NOT DONE YET]
-
-\subsection{Mutation Spectrum}
-We calculated the mutation spectrum for each list of significant SNPs for each population [NOT DONE YET]
-
-We also compared the mutation spectrum ratio between populations using a modified version of the methods used in Harris et al. 2017.
-
-\subsection{Imputation}
-Using the Michigan Imputation Server, we imputed the genotype data from 1000 Genomes Project
-
 			\section{Results}
@@
 The magnitude and impact of the differences in quality within and between populations has not yet been explored.
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{MapQualOverTime.jpg}
+
+\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
+The average mapping quality per individual was obtained from the raw alignment index data for each sample.
+The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
+B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
+\label{Figure1}
+\end{figure}
 
@@
 
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
+\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
+This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
+}
+ \label{Figure2}
+\end{figure}
+
 \subsection{Overlap of Significant SNPs}
@@
 This confirmation using more than one GWAS is a very strong result suggesting that these variants might not be legitimate  \ref{Figure3}.
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
+
+\caption{Overlap of SNPs identified independently to be associated with quality.
+Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants.
+What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations.
+}
+  \label{Figure3}
+\end{figure}
 
@@
 
-\afterpage{
-\begin{figure}
-\includegraphics[width=\hsize]{figures/MapQualOverTime.jpg}
+\section{Methods}
+\subsection{Metadata}
+The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
+Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
+Each BAS file has metadata regarding each sequencing event for each sample. 
+If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
+The date of completion of each sequencing occurrence was not available, however we managed to find the submission dates and sequencing centres for each sample in the phase3 analysis sequence index file.
+This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
+The dates of the sequencing are only used to plot Figure. [Average Quality of mapped bases: How was it calculated?]
 
-\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
-The average mapping quality per individual was obtained from the raw alignment index data for each sample.
-The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
-B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
-\label{Figure1}
-\end{figure}
-\clearpage
-}
+\subsection{Data Availability}
 
-\afterpage{
-\begin{landscape}
-\begin{figure}[]
-\centering
-\includegraphics[width=23cm,keepaspectratio, right]{figures/GWAS_Qual_GenomeWide_JPT.jpg}
-\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
-This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
-}
- \label{Figure2}
-\end{figure}
-\end{landscape}
+Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
+Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here} 
+[link to my compiled metadata file here]
 
-\clearpage
+\subsection{Quality Controls}
+We reproduced the quality control pipelines used by Harris et.
+al as they applied the current state of the art quality thresholds to remove questionable sequences especially for the high standards for detecting population level differences.
+Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or basepair complexity compared to the rest of the genome. 
+The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
+Furthermore, only diallelic autosomal SNPs were considered, with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9.
 
-  }
+\subsection{Genome Wide Association}
 
-\afterpage{
-\begin{figure}[h]
-\centering
-\includegraphics[width=\hsize,keepaspectratio, right]{figures/SNPOverlap6.jpg}
+Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1000 Genomes Project.
+We used the average quality of mapped bases per individual as the phenotype for the analysis.
+We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data.
+[NOT DONE YET]
 
-\caption{Overlap of SNPs identified independently to be associated with quality.
-Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants.
-What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations.
-}
-  \label{Figure3}
-\end{figure}
-\clearpage
-}
+\subsection{Mutation Spectrum}
+We calculated the mutation spectrum for each list of significant SNPs for each population [NOT DONE YET]
+
+We also compared the mutation spectrum ratio between populations using a modified version of the methods used in Harris et al. 2017.
+
+\subsection{Imputation}
+Using the Michigan Imputation Server, we imputed the genotype data from 1000 Genomes Project
 

### 2018-04-11 Draft3.tex
more words

@@
 Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
-Mutations associated to a disease can be identified by comparing individuals afflicted by a disease to healthy individuals. 
+Mutations associated to a disease can be identified by comparing healthy individuals to those afflicted by a disease. 
 The mutations more common in one group might be associated to the disease.
@@
 \subsection{Imputation}
- improvements in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases. 
-Genotype data is often imputed to increase power for association studies in a cost effective way. 
-This process is done using a reference database comprised of whole genome sequence data from other individuals in reference populations. 
-We often combine these reference databases to increase the sample size and in turn, the accuracy of imputation. 
-Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.  
+Genome wide associations using large cohorts have lead to developments in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases. 
+Despite drastic reductions in cost of whole genome sequencing, it remains an expensive test for large sample sizes.
+While on average, two humans differ in about 1/10,000 bases, this number is lower in closely related individuals, and higher in individuals from different continental origins.
+Modern chip sequencing will provide the genotype information for over 1 million bases of the genome.
+This data can then be aligned to a reference database of whole genome sequences where the haplotypes that are identical by descent in individuals form the same population can be used to infer the most likely haplotypes of a genotyped individual.
+
+The accuracy of imputation depends on the size of the reference database; this varies significantly from one population to another.
+This means that individuals with ancestry that is less well represented in the reference database will have lower accuracy of imputation.
+To overcome this bias, reference databases are often combined to increase the sample size and in turn, the accuracy of imputation.
+However, combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
 The errors in one dataset don't disappear when they are combined with another higher quality one. 
-Many of the legacy datasets have been known to contain a higher rate of false positives than their more recent counterparts. 
+Many of the legacy datasets produced using vintage sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
 Is it time to retire legacy datasets?
 
-\lat{sentence about harris} 
+\subsection{Mutation Spectra} 
 A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
@@
 Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
-They argued this by showing that relative difference in mutational types between and even within continental populations.  
+They argued this by showing relative differences in mutational types between and even within continental populations.
+
+\lat{rewrite this section}
 In investigating a mutational signature identified in a Japanese cohort by Harris et.
@@
 Our research center had recently obtained access to a large cohort of several thousand high-quality whole genome sequences from Japanese individuals from the Fumi institute.
+\lat{change focus to data not center}
 We started off by trying to replicate the results found in Harris et al.
@@
 This result raised several red flags. 
-Why would a supposedly genome wide signal be concentrated in private SNPs from a legacy dataset?
+Why would a genome wide signal be concentrated in private SNPs from a legacy dataset?
 

### 2018-04-11 Draft3.tex
introduction is functional
now have a menu for the paper

@@
 The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
-\lat{not a fan of this sentence}
 Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
@@
 They argued this by showing relative differences in mutational types between and even within continental populations.
-
-\lat{rewrite this section}
-In investigating a mutational signature identified in a Japanese cohort by Harris et.
-al (2017), we noticed that the individuals responsible for this signal all had lower average quality scores compared to the other Japanese individuals.
-To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
-We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
-One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
+\lat{more here about mutation spectra}
 
@@
 This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.
+In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset.
 
-Our research center had recently obtained access to a large cohort of several thousand high-quality whole genome sequences from Japanese individuals from the Fumi institute.
-\lat{change focus to data not center}
-We started off by trying to replicate the results found in Harris et al.
-2017, but after several attempts and double checking we had the right quality controls, we were unable to replicate their results with the Fumi dataset.
+This study is the result of an investigation in the quality of the 1000 Genomes Project (1kGP) dataset.
+To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent counterpart. 
+We then consider methods to discriminate troublesome variants resulting from dated sequencing technologies. 
+Furthermore, we explore more broadly how low quality variants have been propagated in the other populations included in the 1kGP.
+Finally, we will discuss the impact these variants have on modern analyses.
+
+			\section{Results}
+	\subsection{A heterogeneously distributed mutational signal}			
 Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population.
-Further investigation in the allele frequency spectra of the two Japanese datasets, indicated that the SNPs responsible for the signal were missing in the more recent dataset. 
-This result raised several red flags. 
-Why would a genome wide signal be concentrated in private SNPs from a legacy dataset?
+The SNPs responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
+\todo{plot SFS NAG vs JPT}
+We also noticed that the individuals carrying these mutations all had lower average quality scores.
+This suggested that individuals with lower quality were driving this apparent population genetics signal.
+\todo{plot PC1 vs Qual}
 
-Upon investigating the quality control pipelines of the more recent dataset, many of these positions uniquely found in the 1000 Genomes Project had been removed in the standard QC of the sequencing centre.
-Does this legacy dataset meet the current industry standards of quality? 
-This study is the result of an investigation in the quality of the 1000 Genomes Project dataset.
+To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS) \ref{Figure2}.
+
+We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
+One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
 
 After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual.
-This suggested that some individuals with lower quality were driving this apparent population genetics signal. [mention fake pop structure]
+
 Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
@@
 
-			\section{Results}
 	\subsection{Average quality of mapped snps Quality Over Time}
@@
 When we remove individuals below a quality of 30, population structure is clarified. 
-The signal identified in Harris et. al is enriched in sites associated with low quality. [Need to run harris pipeline without $10^{-6}$ ]
+The signal identified in Harris et. al is enriched in sites associated with low quality
 Therefore, when these low quality sites are removed, the mutational signal is also removed. (Or in this case when we remove the individuals with qualities below 30.)
@@
 This suggests that we lack the power to identify all the questionable variants from these legacy datasets.
+
 	\subsection{Imputation}

### 2018-04-12 Draft3.tex
polished intro

@@
 Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
-Mutations associated to a disease can be identified by comparing healthy individuals to those afflicted by a disease. 
-The mutations more common in one group might be associated to the disease.
-Demonstrating that these mutations are biologically relevant to the disease can be difficult.
-Certain factors like age, sex or ancestral origin might be correlated to the trait in question and could potentially be mistakenly associated to a disease. 
+Mutations associated to a disease can be identified by comparing the genomes of healthy individuals to those afflicted by a disease. 
+The mutations commonly found in patients and rarely in controls might be associated to the disease in question.
+However, demonstrating that these mutations are biologically relevant can be difficult.
 Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
@@
 Despite drastic reductions in cost of whole genome sequencing, it remains an expensive test for large sample sizes.
-While on average, two humans differ in about 1/10,000 bases, this number is lower in closely related individuals, and higher in individuals from different continental origins.
+For this reason, genotype data is often imputed to increase power for association studies in a cost effective way.
+
+Imputation of genotype data is a probabilistic method that infers the bases of a given genome based on its similarity to a set of reference genomes.
+While on average, two humans differ in about 1/10,000 bases, this number is \todo{lower} in closely related individuals, and \todo{higher} in individuals from different continental origins.
 Modern chip sequencing will provide the genotype information for over 1 million bases of the genome.
-This data can then be aligned to a reference database of whole genome sequences where the haplotypes that are identical by descent in individuals form the same population can be used to infer the most likely haplotypes of a genotyped individual.
+The unique combination of genotyped bases can be enough to identify haplotype blocks that are identical by descent in individuals form the reference database.
 
@@
 The errors in one dataset don't disappear when they are combined with another higher quality one. 
-Many of the legacy datasets produced using vintage sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-Is it time to retire legacy datasets?
+Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
+Is it time to retire legacy data?
 
@@
 The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
-Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
+Harris et al have recently put forward that there are patterns of mutations that differ across human populations.
 They argued this by showing relative differences in mutational types between and even within continental populations.
@@
 This project was motivated by an unusual population genetics observation in a recent publication by Harris et.
-al, that the 1000 Genomes Project Japanese population seem to be partitioned in two clusters of diff mutation rates.
+al, that the 1000 Genomes Project (1kGP) Japanese population seem to be partitioned in two clusters of diff mutation rates.
 This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.
-In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset.
+In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset and concluded that this signal can be attributed to sequencing error.
 
-This study is the result of an investigation in the quality of the 1000 Genomes Project (1kGP) dataset.
-To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent counterpart. 
-We then consider methods to discriminate troublesome variants resulting from dated sequencing technologies. 
-Furthermore, we explore more broadly how low quality variants have been propagated in the other populations included in the 1kGP.
+This study is the result of an investigation in the quality of the 1kGP dataset.
+To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent high quality dataset. 
+We then consider methods to discriminate sequencing errors resulting from dated technologies.
+Next, we explore more broadly how low quality variants have proliferated in other populations from the 1kGP.
 Finally, we will discuss the impact these variants have on modern analyses.
@@
 
-To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS) \ref{Figure2}.
+To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure2}.
+GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
+In this case, we are using this analysis in a more unconventional way to identify regions of the genome that are associated to biologically irrelevant traits like SNPs associated to individuals with low quality.
 
-We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
-One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
+Using the linear GWAS function offered by PLINK, we were able to identify \todo{000} SNPs that were associated to low quality with  p-values below $10^{-8}$ and \todo{000} snps with p-values below $10^{-6}$ \ref{Figure2}.
+This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+The results from of the GWAS for other populations are available in supplementary data. 
+A total of \todo{000} SNPs were associated to low quality in all populations combined.
 
-After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual.
-
-Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
-We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
+\todo{does getting rid of these SNPs get rid of the signal?
 
@@
 	\subsection{Genome Wide Association}
-GWAS are commonly used to find genomic regions associated with biologically relevant traits.
-In this case, we are using this analysis in a more unconventional way to identify regions of the genome that are associated to biologically irrelevant traits like individuals with low average quality of mapped bases.
-
-Using the linear GWAS function offered by PLINK, we were able to identify 000 SNPs that were associated to low quality with  p-values below $10^{-8}$ and 000 snps with p-values below $10^{-6}$ \ref{Figure2}.
-The results from of the GWAS for other populations are available in supplementary data. 
-A total of 0000 SNPs were associated to low quality in all populations combined.
 
@@
 \caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
-This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
 The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
@@
 
-Our GWAS method was able to identify 000 variants that are likely to be false positives.
+Our GWAS method was able to identify \todo{000} variants that are likely to be false positives.
 
@@
 Because we are running the GWAS on a per-population base, for rarer mutations, we are lacking the power to identify these significant SNPs.  
-However, upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified 000 more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
+However, upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
 This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts. 
 The most conservative approach would be to remove all individuals that don't meet the quality threshold. 
-In this case, we used a cut off of an average quality of mapped bases over 30. This threshold has been previously used by 000 studies.
+In this case, we used a cut off of an average quality of mapped bases over 30. This threshold has been previously used by \todo{000} studies.
 
@@
 This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
-The dates of the sequencing are only used to plot Figure. [Average Quality of mapped bases: How was it calculated?]
+The dates of the sequencing are only used to plot Figure. \todo:{Average Quality of mapped bases: How was it calculated?}
 
@@
 We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data.
-[NOT DONE YET]
+\todo{NOT DONE YET}
 
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum for each list of significant SNPs for each population [NOT DONE YET]
+We calculated the mutation spectrum for each list of significant SNPs for each population \todo{NOT DONE YET}
 
@@
 \end{document}
+
+We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
+One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
+
+After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual.
+
+Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
+We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.

### 2018-04-12 Draft3.tex
gwas in mutspect section

@@
 \newcommand{\lat}[1]{\textcolor{blue}{LAT: #1}}
-\newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}
+\newcommand{\todo}[1]{\textcolor{blue}{todo: #1}}
 
@@
 To overcome this bias, reference databases are often combined to increase the sample size and in turn, the accuracy of imputation.
-However, combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
+
+Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
 The errors in one dataset don't disappear when they are combined with another higher quality one. 
@@
 A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
-The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
-Harris et al have recently put forward that there are patterns of mutations that differ across human populations.
+The signatures from individuals in the same population will have a tendency to be more similar due to shared ancestry.
+Harris et al have recently posited that there are patterns of mutations that differ across human populations.
 They argued this by showing relative differences in mutational types between and even within continental populations.
+In particular, there is \todo{15 percent} more TTC to TCC mutations in people of European descent.
 \lat{more here about mutation spectra}
@@
 We then consider methods to discriminate sequencing errors resulting from dated technologies.
-Next, we explore more broadly how low quality variants have proliferated in other populations from the 1kGP.
+Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP.
 Finally, we will discuss the impact these variants have on modern analyses.
@@
 Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population.
-The SNPs responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
+Surprisingly, the SNPs responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
 \todo{plot SFS NAG vs JPT}
 We also noticed that the individuals carrying these mutations all had lower average quality scores.
-This suggested that individuals with lower quality were driving this apparent population genetics signal.
+This suggested that individuals with lower quality were driving this signal.
 \todo{plot PC1 vs Qual}
@@
 
-Using the linear GWAS function offered by PLINK, we were able to identify \todo{000} SNPs that were associated to low quality with  p-values below $10^{-8}$ and \todo{000} snps with p-values below $10^{-6}$ \ref{Figure2}.
+Using the linear GWAS function offered by PLINK, we were able to identify \todo{000 $10^{-8}$ and 000 $10^{-6}$} SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure2}.
 This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
-The results from of the GWAS for other populations are available in supplementary data. 
-A total of \todo{000} SNPs were associated to low quality in all populations combined.
 
-\todo{does getting rid of these SNPs get rid of the signal?
+\todo{does getting rid of these SNPs get rid of the signal? What is the lowest frequency SNP that is statistically significant?}
 

### 2018-04-12 Draft3.tex
two columns

@@
-\documentclass[12pt]{amsart}
-\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
+\documentclass[12pt,twocolumn]{article}
+\usepackage[margin=2.2cm]{geometry} % see geometry.pdf on how to lay out the page. There's lots.
 \usepackage{graphicx}
@@
     urlcolor=cyan,
-}\geometry{a4paper} % or letter or a5paper or ... etc
-\usepackage{epstopdf}
-\epstopdfDeclareGraphicsRule{.tiff}{png}{.png}{convert #1 \OutputFile}
-\AppendGraphicsExtensions{.tiff}
+}
+
+%\geometry{a4paper} % or letter or a5paper or ... etc
 
@@
 \newcommand{\lat}[1]{\textcolor{blue}{LAT: #1}}
-\newcommand{\todo}[1]{\textcolor{blue}{todo: #1}}
+\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
 
@@
 
+\onecolumn
 \maketitle
@@
 \clearpage
+\twocolumn
 			\section{Introduction}
@@
 
-\begin{figure}
+\begin{figure*}
 \includegraphics[width=\hsize,keepaspectratio]{MapQualOverTime.jpg}
@@
 \label{Figure1}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
@@
  \label{Figure2}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
@@
   \label{Figure3}
-\end{figure}
+\end{figure*}
 

### 2018-04-13 Draft3.tex
more figures

@@
 			\section{Results}
-	\subsection{A heterogeneously distributed mutational signal}			
-Upon comparing the allele frequencies between the two datasets, we observed an unusually large number of private SNPs despite both cohorts being from the same population.
-Surprisingly, the SNPs responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
+	\subsection{A peculiar mutational signal}			
+One way to contrast two populations is by comparing the frequencies of the same variants in both populations.
+If the two groups are sampled from the same population, we expect there to be little deviation in allele frequencies.
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and a more recent dataset, we observed an unusually large number of private SNPs.
+Surprisingly, the variants responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
 \todo{plot SFS NAG vs JPT}
-We also noticed that the individuals carrying these mutations all had lower average quality scores.
+We also noticed that the individuals carrying these mutations all had lower average quality scores of mapped bases.
 This suggested that individuals with lower quality were driving this signal.
 \todo{plot PC1 vs Qual}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{NAG_JPT_SFS_GenomeWide.jpg}
+\caption{Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
+Out of \todo{000} total SNPs, \todo{000} of them are missing in one population but present in the other by more than 1\%.}
+ \label{SFS}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
+\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
+This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
+}
+ \label{Figure2}
+\end{figure}
 
@@
 
-Using the linear GWAS function offered by PLINK, we were able to identify \todo{000 $10^{-8}$ and 000 $10^{-6}$} SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure2}.
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+Using the linear GWA study function offered by PLINK, we were able to identify \todo{000 $10^{-8}$ and 000 $10^{-6}$} SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure2}.
+This GWA study included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
 
@@
 
-\begin{figure*}
+\begin{figure}
 \includegraphics[width=\hsize,keepaspectratio]{MapQualOverTime.jpg}
@@
 \label{Figure1}
-\end{figure*}
-
-	\subsection{Genome Wide Association}
-
-\begin{figure*}
-\includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
-\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
-}
- \label{Figure2}
-\end{figure*}
+\end{figure}
 
 	\subsection{Overlap of Significant SNPs}
-Comparing the results of each independent GWAS, we were able to identify over 0000 variants that were independently associated to low quality in multiple populations.
+Comparing the results of each independent GWA study, we were able to identify over 0000 variants that were independently associated to low quality in multiple populations.
 This confirmation using more than one GWAS is a very strong result suggesting that these variants might not be legitimate  \ref{Figure3}.
 
-\begin{figure*}
+\begin{figure}
 \includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
@@
 \caption{Overlap of SNPs identified independently to be associated with quality.
-Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants.
+The populations that have the most low-quality individuals also have the most low-quality variants.
 What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations.
@@
   \label{Figure3}
-\end{figure*}
+\end{figure}
 
@@
 Interestingly, the SNPs we identified had already been removed in higher quality datasets. 
-These HQ datasets also have filtered more SNPs than the GWAS was able to identify. 
-When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWAS.
+These HQ datasets also have filtered more SNPs than the GWA study was able to identify. 
+When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWA study.
 This suggests that we lack the power to identify all the questionable variants from these legacy datasets.
@@
 
-	\subsection{Found to be included in other GWAS}
+	\subsection{Found to be included in other GWA studies}
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWA studies that might have called these erroneous variants as being significantly correlated with some biological trait.
@@
 They also removed rare alleles and alleles with high degrees of missingness. 
-Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also to reach genome wide significance for the traits they were testing for.
+Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
 
@@
 We compared the number of SNPs identified in the GWAS to the number of sites that differed between our higher quality Japanese dataset and found that 100?\% of the sites were also missing from the Fumi data.
-Because we are running the GWAS on a per-population base, for rarer mutations, we are lacking the power to identify these significant SNPs.  
+Because we are running the GWA study on a per-population base, for rarer mutations, we are lacking the power to identify these significant SNPs.  
 However, upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
-This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts. 
+This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts. 
 The most conservative approach would be to remove all individuals that don't meet the quality threshold. 
@@
 
-\subsection{GWAS from other papers}
+\subsection{GWA studies from other papers}
 Since these variants are present in more than one population from the 1000 genomes project, they are more likely to be associated to biological traits as they would appear to be like any other variant that is shared among multiple populations.
 The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
+
 \subsection{Why were these SNPs even associated in the first place?}

### 2018-04-13 Draft3.tex
elife formatting :)
minor formatting

@@
-\documentclass[12pt,twocolumn]{article}
-\usepackage[margin=2.2cm]{geometry} % see geometry.pdf on how to lay out the page. There's lots.
-\usepackage{graphicx}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ELIFE ARTICLE TEMPLATE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% PREAMBLE 
+\documentclass[9pt,lineno]{elife}
+
 \usepackage[export]{adjustbox}
@@
 
-%\geometry{a4paper} % or letter or a5paper or ... etc
+\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
+\newcommand{\luke}[1]{\textcolor{blue}{Luke: #1}}
+\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ARTICLE SETUP
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title{Legacy Data Confounds Modern Genomics Studies}
 
-\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
-\newcommand{\lat}[1]{\textcolor{blue}{LAT: #1}}
-\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
+\author[1]{Luke Anderson-Trocm\'e}
+\author[1,2]{Simon Gravel}
 
-% \geometry{landscape} % rotated page geometry
+\affil[1]{Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada}
+\affil[2]{McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada}
 
-% See the ``Article customise'' template for come common customisations
+\corr{luke.anderson-trocme@mail.mcgill.ca}{LAT}
 
-\title{Legacy Data Confounds Modern Studies}
-\author{Luke Anderson-Trocm\'e , Simon Gravel}
-%\date{} % delete this line to display the current date
-
-%%% BEGIN DOCUMENT
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% ARTICLE START
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-\onecolumn
 \maketitle
-\tableofcontents
-\clearpage
-\twocolumn
+\begin{abstract}
+\todo{This is not a real abtract}This study is the result of an investigation in the quality of the 1kGP dataset.
+To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent high quality dataset. 
+We then consider methods to discriminate sequencing errors resulting from dated technologies.
+Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP.
+Finally, we will discuss the impact these variants have on modern analyses.
+\end{abstract}
+
 			\section{Introduction}
-	\subsection{Larger Sample Sizes}			
+	\subsection{Big Data}			
 The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
@@
 In particular, there is \todo{15 percent} more TTC to TCC mutations in people of European descent.
-\lat{more here about mutation spectra}
+\luke{more here about mutation spectra}
 
@@
 
-\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
 The average mapping quality per individual was obtained from the raw alignment index data for each sample.
 The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
-B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
+\textbf{B} Same data as in \textbf{A} except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
 \label{Figure1}
@@
 
+\section{Code Availability}
+\todo{here is where I will put my git hub}
+
+\section{Acknowledgments}
+We would like to thank Kelly Harris for sharing her mutation spectrum pipelines.
+
 \end{document}

### 2018-04-16 Draft3.tex
monday morning

@@
 \begin{abstract}
-\todo{This is not a real abtract}This study is the result of an investigation in the quality of the 1kGP dataset.
-To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent high quality dataset. 
-We then consider methods to discriminate sequencing errors resulting from dated technologies.
-Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP.
-Finally, we will discuss the impact these variants have on modern analyses.
+In investigating a population genetics signal, we noticed some discrepancies between the Japanese samples from the 1000 Genomes Project (1kGP) and a more recent high quality dataset.
+We found that the variants causing the signal were only present in individuals with low average sequencing quality.
+We identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ single nucleotide polymorphisms (SNPs) that were associated to quality in the 1kGP dataset Japanese individuals.
+These variants are present in nearly all of the populations from the 1kGP. 
+We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
+Some of these variants are being imputed onto genotype data and reach genome wide significance in recent publications.
+\todo{update values for all populations}
 \end{abstract}
@@
 
-	\subsection{Mutation Spectra} 
+	\subsection{Mutation Spectrum}
+
 A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
 The signatures from individuals in the same population will have a tendency to be more similar due to shared ancestry.
-Harris et al have recently posited that there are patterns of mutations that differ across human populations.
-They argued this by showing relative differences in mutational types between and even within continental populations.
-In particular, there is \todo{15 percent} more TTC to TCC mutations in people of European descent.
-\luke{more here about mutation spectra}
+In 2015, Harris et al reported an overrepresentation of the TCC to TTC mutation in European populations.
+Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
+In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
+One result was of particular interest to our research group : the heterogeneously distributed signal found in a subset of Japanese individuals. 
+A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.
 
-	\subsection{Context}
-This project was motivated by an unusual population genetics observation in a recent publication by Harris et.
+	\subsection{Motivations}
+This project was motivated by this unusual population genetics observation in a recent publication by Harris et.
 al, that the 1000 Genomes Project (1kGP) Japanese population seem to be partitioned in two clusters of diff mutation rates.
-This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a recent mutagen, or technical bias.
+This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a mutagen, or a technical bias.
 In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset and concluded that this signal can be attributed to sequencing error.
@@
 			\section{Results}
-	\subsection{A peculiar mutational signal}			
-One way to contrast two populations is by comparing the frequencies of the same variants in both populations.
-If the two groups are sampled from the same population, we expect there to be little deviation in allele frequencies.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and a more recent dataset, we observed an unusually large number of private SNPs.
-Surprisingly, the variants responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset. 
-\todo{plot SFS NAG vs JPT}
+	\subsection{A peculiar mutational signature in Japan}			
+If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups.
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and a more recent dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
+Surprisingly, the variants responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset \ref{Figure1}. 
 We also noticed that the individuals carrying these mutations all had lower average quality scores of mapped bases.
 This suggested that individuals with lower quality were driving this signal.
-\todo{plot PC1 vs Qual}
 
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{NAG_JPT_SFS_GenomeWide.jpg}
-\caption{Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
-Out of \todo{000} total SNPs, \todo{000} of them are missing in one population but present in the other by more than 1\%.}
- \label{SFS}
+\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}
+\caption{
+\textbf{A} 
+Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
+Crosses ( + ) are variants that reached genome wide significance in a GWA of sequencing quality. 
+The histogram on the left of the plot is the distribution of significant variants. 
+The lowest frequency variant able to reach genome wide significance is 5\%.
+\textbf{B} 
+Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than  in a in a GWA of sequencing quality. 
+There is a \todo{significant} enrichment in *AC${\rightarrow}$*CC mutations. 
+\textbf{C} 
+Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1000 Genomes Project.
+This GWA study identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
+The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
+ \label{Figure1}
 \end{figure}
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
-\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project.
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
-}
- \label{Figure2}
-\end{figure}
-
-To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure2}.
+To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure1}.
 GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
@@
 
-Using the linear GWA study function offered by PLINK, we were able to identify \todo{000 $10^{-8}$ and 000 $10^{-6}$} SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure2}.
+Using the linear GWA study function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure1}.
 This GWA study included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations.
 
-\todo{does getting rid of these SNPs get rid of the signal? What is the lowest frequency SNP that is statistically significant?}
+Despite removing the variants identified as being low quality, the signal persists. 
+When we remove the variants significantly associated to low quality, the signal identified by Harris et. al (2017) the enrichment in *AC${\rightarrow}$*CC mutations. 
+When we remove individuals with average quality of mapped SNPs below 30, the signal goes away. 
+We suspect that this has to do with the lack of power we have for more rare alleles. 
+The lowest frequency variants that are significantly associated to quality are at 5\%. 
+When we remove all the variants above 1\% in 1kGP but missing from the NAG data, the signal disappears.
+This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
 
-	\subsection{Average quality of mapped snps Quality Over Time}
-As one would expect, the sequencing quality of individuals increases over time \ref{Figure1}.
-Sequencing centers improve on their laboratory techniques and sequencing technologies improve from one generation to the next.
-The magnitude and impact of the differences in quality within and between populations has not yet been explored.
+ \luke{more here about using higher quality data to validate legacy data}.
+Upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
 
+\todo{supplementary figures : genome wide, mutspect without sig snps, and without <30}
+
+	\subsection{Sequencing quality over time}
+We turned our attention to the other populations in the 1kGP, we found that the sequencing done in phase 1 was more variable and overall tended to include lower quality sequencing data \ref{MapQual}.
+The sequencing quality of individuals increases over time \ref{MapQual}. 
+\todo{What sequencers were used in each phase?}
+By 2011 the sequencing quality seems to level off.
 \begin{figure}
@@
 \caption{\textbf{A} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
-The average mapping quality per individual was obtained from the raw alignment index data for each sample.
-The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
-\textbf{B} Same data as in \textbf{A} except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
-\label{Figure1}
+The x-axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
+\textbf{B} Same data as in \textbf{A} except the x-axis is sorted by sequencing date.
+The colors indicate the sequencing centers that produced the data for each individual.}
+\label{MapQual}
 \end{figure}
 
-	\subsection{Overlap of Significant SNPs}
-Comparing the results of each independent GWA study, we were able to identify over 0000 variants that were independently associated to low quality in multiple populations.
-This confirmation using more than one GWAS is a very strong result suggesting that these variants might not be legitimate  \ref{Figure3}.
+	\subsection{Overlap of significant SNPs}
+Comparing the results of each independent GWA study, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations.
+This confirmation using more than one GWAS is a strongly suggests that these variants might not be genuine \ref{Figure3}.
 
@@
 
-	\subsection{Significant SNPs}
-When we remove individuals below a quality of 30, population structure is clarified. 
-The signal identified in Harris et. al is enriched in sites associated with low quality
-Therefore, when these low quality sites are removed, the mutational signal is also removed. (Or in this case when we remove the individuals with qualities below 30.)
-
-	\subsection{Higher quality datasets don't have these SNPs}
-
-Interestingly, the SNPs we identified had already been removed in higher quality datasets. 
-These HQ datasets also have filtered more SNPs than the GWA study was able to identify. 
-When we compare the allele frequencies between the 1000 Genomes project Japanese population to a more contemporary dataset, there are many more sites that are enriched with the same mutational signature as described above that are not identified using a GWA study.
-This suggests that we lack the power to identify all the questionable variants from these legacy datasets.
-
 	\subsection{Imputation}
-~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These  should be removed from reference database
+~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These should be removed from reference database.
 
@@
 Could lead to false diagnosis at worst, or spurious correlations at least. 
-They were not caught sooner because the batch effect of 1000 genomes had never been properly addressed.
+Polygenic risk scores take into account all snps reaching significance, without much manual curation. 
+It's likely that the variants we have identified are being included in these multi-locus risk scores.
 
-Our GWAS method was able to identify \todo{000} variants that are likely to be false positives.
+Despite these variants reaching genome wide significance, the majority of researchers did not pursue these variants for further analysis.
+This is likely because these variants have no peak in LD that is characteristic of a biologically significant variant.
 
-We compared the number of SNPs identified in the GWAS to the number of sites that differed between our higher quality Japanese dataset and found that 100?\% of the sites were also missing from the Fumi data.
-Because we are running the GWA study on a per-population base, for rarer mutations, we are lacking the power to identify these significant SNPs.  
-However, upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
-This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts. 
-The most conservative approach would be to remove all individuals that don't meet the quality threshold. 
-In this case, we used a cut off of an average quality of mapped bases over 30. This threshold has been previously used by \todo{000} studies.
+\subsection{Recommendations}
+The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
+In this case, we used a cut off of an average quality of mapped bases over 30.
+This threshold has been \todo{previously used by studies}.
+It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
 
@@
 
-\subsection{Why were these SNPs even associated in the first place?}
-[due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?]
+\luke{due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?}
 
@@
 If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
-The date of completion of each sequencing occurrence was not available, however we managed to find the submission dates and sequencing centres for each sample in the phase3 analysis sequence index file.
+The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
 This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
-The dates of the sequencing are only used to plot Figure. \todo:{Average Quality of mapped bases: How was it calculated?}
+The dates of the sequencing are only used to plot Figure. \todo{Average Quality of mapped bases: How was it calculated?}
 
@@
 Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
+
 Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here}
-[link to my compiled metadata file here]
+
+\todo{link to my compiled metadata file here}
 
@@
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum for each list of significant SNPs for each population \todo{NOT DONE YET}
+We calculated the mutation spectrum for each list of significant SNPs for each population.
 
@@
 \end{document}
-
-We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
-One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.
-
-After compiling all the metadata from each 1000 genomes sequencing run, we noticed a correlation between PC1 of the mutation spectrum signal and the average quality of mapped bases per individual.
-
-Furthermore, when looking at the days when the individuals were sequenced, we noticed that individuals with this mutational signal were sequenced on the same day in the same sequencing centre.
-We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.

### 2018-04-20 Draft3.tex
updated SFS. Removed HWE sites and fixed bin size issue

@@
 
-\author[1]{Luke Anderson-Trocm\'e}
+\author[1,2]{Luke Anderson-Trocm\'e}
 \author[1,2]{Simon Gravel}
@@
 
-\corr{luke.anderson-trocme@mail.mcgill.ca}{LAT}
+\corr{simon.gravel@mcgill.ca}{SG}
 
@@
 			\section{Introduction}
-	\subsection{Big Data}			
+	\subsection{Reference Cohorts}			
 The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
@@
 
-Imputation of genotype data is a probabilistic method that infers the bases of a given genome based on its similarity to a set of reference genomes.
-While on average, two humans differ in about 1/10,000 bases, this number is \todo{lower} in closely related individuals, and \todo{higher} in individuals from different continental origins.
-Modern chip sequencing will provide the genotype information for over 1 million bases of the genome.
-The unique combination of genotyped bases can be enough to identify haplotype blocks that are identical by descent in individuals form the reference database.
-
-The accuracy of imputation depends on the size of the reference database; this varies significantly from one population to another.
-This means that individuals with ancestry that is less well represented in the reference database will have lower accuracy of imputation.
-To overcome this bias, reference databases are often combined to increase the sample size and in turn, the accuracy of imputation.
+\luke{cut paragraph to discussion}
 
@@
 
+\subsection{Imputation}
+Imputation of genotype data is a probabilistic method that infers the bases of a given genome based on its similarity to a set of reference genomes.
+While on average, two humans differ in about 1/10,000 bases, this number is \todo{lower} in closely related individuals, and \todo{higher} in individuals from different continental origins.
+Modern chip sequencing will provide the genotype information for over 1 million bases of the genome.
+The unique combination of genotyped bases can be enough to identify haplotype blocks that are identical by descent in individuals form the reference database.
+
+The accuracy of imputation depends on the size of the reference database; this varies significantly from one population to another.
+This means that individuals with ancestry that is less well represented in the reference database will have lower accuracy of imputation.
+To overcome this bias, reference databases are often combined to increase the sample size and in turn, the accuracy of imputation.
+
 \subsection{GWA studies from other papers}

### 2018-05-09 Clean_Figure1 (Autosaved).R
minor edits throughout

@@ -1,129 +1,0 @@
-library(ggplot2)
-library(colorspace)
-library(cowplot)
-f1kGP='/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered.freq.frq'
-fnag='/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'
-
-tab5rows <- read.table(fnag, header = TRUE, nrows = 5, row.names=NULL)
-classes <- sapply(tab5rows, class)
-#Allele Frequency from JPT
-JPT<-read.table(f1kGP, fill=T,skip=1,  skipNul = TRUE,colClasses = classes)
-colnames(JPT)<-paste("JPT", colnames(JPT), sep = "_")
-#Allele Frequency from NAG
-NAG<-read.table(fnag, fill=T,skip=1,  skipNul = TRUE,colClasses = classes)
-colnames(NAG)<-paste("NAG", colnames(NAG), sep = "_")
-
-#HWE 10^-6
-NAG_hwe<-read.table('/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered_sig6.freq.hwe')
-JPT_hwe <- read.table('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered_sig6.freq.hwe')
-#ggplot(JPT_hwe, aes(x=V2, y=abs(log10(V8))) )+geom_point(shape='.')+facet_grid(~V1)+coord_flip()+theme_classic()+theme(axis.text.x=element_blank())+xlab('Position (bp)')+ylab('-log10(p) HWE disequilibrium')
-#ggsave('~/Documents/HWE_disequilibrium_JPT.jpg', height=5,width=5)
-hwe=merge(NAG_hwe[,c('V1','V2')], JPT_hwe[,c('V1','V2')], by=c('V1','V2'), all=T)
-
-#merge them together
-join<-merge(JPT[which(JPT$JPT_V6 <= 1),],
- 			NAG[which(NAG$NAG_V6 <= 1),], 
- 			by.x=c("JPT_V1","JPT_V2"), 
- 			by.y=c("NAG_V1","NAG_V2"), all=T)
-join[is.na(join)]<-0
-#REMOVE  HWE SITES
-join= join[! paste(join $JPT_V1, join $JPT_V2) %in% c(paste(hwe$V1,hwe$V2)), ]
-
[…]

### 2018-05-09 Draft3.tex
minor edits throughout
changed a title

--- 2018-04-16 Draft3.tex
@@
 			\section{Introduction}
-	\subsection{Big Data}			
+	\subsection{Reference cohorts}			
 The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 

### 2018-05-11 Draft3.tex
rewrote intro, edits in first section

--- 2018-04-20 Draft3.tex
@@
 			\section{Introduction}
+
 	\subsection{Reference Cohorts}
-The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
+
+The last 5 years have seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
 Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
@@
 
-	\subsection{Imputation}
-Genome wide associations using large cohorts have lead to developments in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases. 
+\sgcomment{This intro paragraph is much too vague -- it reads like general thoughts about genetics at large. We need to be much more concise. I would start with something like:}
+
+The last 5 years have seen a drastic increase in the amount and quality of human genomes sequence data.
+Hapmap \cite{}, the 1000 Genomes Project \cite{}, and the Simons Diversity project \cite{}, for example, have made thousands of genomes publicly available for population and medical genetic analyses.
+Many more genomes are available indirectly through servers providing imputation services \cite{haplotype reference consortium} or variant frequency estimation \cite{GnomAD}.
+
+Many of these large datasets comprise have been assembled over many years.
+The first sequenced genomes in the 1000 Genomes project, for example, were sequenced XX years ago, at a time when \sgcomment{blah blah blah}.
+Because of the extraordinary value of freely available data, this early data from the 1000 Genomes Project is still widely used as a reference panel for ...
+
+Yet this data is far from perfect \sgcomment{cite papers that have pointed out issues}.
+This raises the question of whether and how such legacy data should be included in contemporary analyses.
+Here we point out how large and previously unreported batch effects in the early phases of the 1000 Genomes Project still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
+
+\subsection{Motivations}
+
+A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual.
+\sgcomment{This is a procedural definition.
+You need a conceptual definition!
+Something like:} Mutation rate at a given base pair has long been known to depend on the neighbouring nucleotides \cite{}.
+Finer scale patterns have been reported in \cite{}.
+\sgcomment {then define mutational signature (ideally, more of a biological definition), and move on to the Harris paper}
+
+In 2015, Harris et al reported an overrepresentation of the TCC to TTC mutation in European populations. \sgcomment{quantify}
+Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
+In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
+One result was of particular interest to our research group \sgcomment{unless needed, no need to involve our persons in the study} : the heterogeneously distributed signal found in a subset of Japanese individuals.
+A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.
+
+While trying to follow up on this observation, we were unable to reproduce this particular observation in a different dataset.
+Tracing back the source of the discrepancy between the two datasets, we identified a strong and previously unreported batch effect in the 1000 Genomes Project data, and found that this batch effect led to spurious results in a number of recently published studies.
+
+Here we discuss the source of the batch effect and the ways in which these batch effects contaminated a number of recent studies.
+Fortunately, the main conclusions of \sgcomment{most of} these studies remain supported by other data.
+Yet this begs the question: When should we retire legacy data?
+
+%\subsection{Imputation}
+%Genome wide associations using large cohorts have lead to developments in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases.
 Despite drastic reductions in cost of whole genome sequencing, it remains an expensive test for large sample sizes.
-For this reason, genotype data is often imputed to increase power for association studies in a cost effective way.
+%For this reason, genotype data is often imputed to increase power for association studies in a cost effective way.
+
+%\sgcomment{Too wordy (also, readers of Nature genetics know about imputation!: Despite drastic reduction in sequencing costs, genotyping and imputation remain a more affordable and popular approach for large-scale cohorts.}
 
@@
 Is it time to retire legacy data? 
-
-	\subsection{Mutation Spectrum}
-
-A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
-The signatures from individuals in the same population will have a tendency to be more similar due to shared ancestry.
-In 2015, Harris et al reported an overrepresentation of the TCC to TTC mutation in European populations.
-Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
-In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
-One result was of particular interest to our research group : the heterogeneously distributed signal found in a subset of Japanese individuals. 
-A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.
-
-	\subsection{Motivations}
-This project was motivated by this unusual population genetics observation in a recent publication by Harris et.
-al, that the 1000 Genomes Project (1kGP) Japanese population seem to be partitioned in two clusters of diff mutation rates.
-This heterogeneously distributed mutational signal was unexpected as a signal of this nature could either be due to population structure, a mutagen, or a technical bias.
-In investigating this mutational signature, we were unable to reproduce the results using a larger and higher quality dataset and concluded that this signal can be attributed to sequencing error.
-
-This study is the result of an investigation in the quality of the 1kGP dataset.
-To begin, we will discuss the discrepancy between the Japanese samples from the 1kGP and a more recent high quality dataset. 
-We then consider methods to discriminate sequencing errors resulting from dated technologies.
-Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP.
-Finally, we will discuss the impact these variants have on modern analyses.
+\sgcomment{As a style issue: you tend to write sentences without indicating the logical flow between them. It makes it hard to follow. }
 
@@
 If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and a more recent dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-Surprisingly, the variants responsible for the signal observed by Harris et. al (2017) were missing in the more recent dataset \ref{Figure1}. 
-We also noticed that the individuals carrying these mutations all had lower average quality scores of mapped bases.
+\sgcomment {Here again, you are a bit procedural.
+It's useful to explain to the reader the logical flow of how you got to doing what you are doing: Harris et al reported an excess of XYZ to XWZ mutations in some Japanese individuals.
+While trying to follow up on this observations in a larger cohort of XXX Japanese individuals, [[some info about cohort]], we did not find this particular signature.}
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
+\sgcomment{These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium, ...}
+
+Once mismatch sites were removed from the 1kGP data, the XYZ - W \sgcomment{I would call the effect by the name of the signature rather than harris et al -- more descriptive, less personal.} signal disappears \ref{Figure1} \sgcomment{usually you give a descriptie name to labels, latex will figure out the numbers!}, suggesting a technical artifact rather than a population structure effect.
+Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the XYZ mutational signature in 1kGP, with low-quality individuals showing elevated rates of the signature.
+\ref{Figure1}
+
+ individuals carrying these mutations all had lower average quality scores of mapped bases.
 This suggested that individuals with lower quality were driving this signal.
@@
 
- \luke{more here about using higher quality data to validate legacy data}.
+One way to assess the validity of these unusual variants is to see if they are present in a higher quality dataset.
 Upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
-
-\todo{supplementary figures : genome wide, mutspect without sig snps, and without <30}
 

### 2018-05-11 Clean_Figure2.R
added Mathieu as author and changed sentence about seq tech

@@ -1,0 +1,55 @@
+library(ggplot2)
+library(ggthemes)
+library(cowplot)
+
+Pops<-read.table("~/Dropbox/LukeTemp/NamePop.txt")
+Meta<-read.table("~/Dropbox/LukeTemp/1000GenomesMetaData.txt")
+Meta$SUBMISSION.DATE<-as.Date(Meta$SUBMISSION.DATE, "%Y-%m-%d")
+phase<-read.table('~/genomes/genomes/PHASE1_igsr_samples.tsv', header=T, sep='\t')
+phase$Phase=1
+Meta<-merge(Meta, phase[c('Sample.name','Phase')],by.x='Name',by.y='Sample.name', all=T)
+Meta[which(is.na(Meta$Phase)),]$Phase=3
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
+pop.batch<-unique(
+Meta[,c('Name','SUBMISSION.DATE','average_quality_of_mapped_bases','Pop','BigPop','CENTER_NAME','Phase','INSTRUMENT_MODEL')])
+pop.batch$CENTER_NAME<-gsub('Illumina','ILLUMINA',pop.batch$CENTER_NAME)
+pop.batch$INSTRUMENT_MODEL <-gsub('Illumina','',pop.batch$INSTRUMENT_MODEL)
+pop.batch$INSTRUMENT_MODEL <-gsub('Genome Analyzer','GA',pop.batch$INSTRUMENT_MODEL)
+pop.batch$INSTRUMENT_MODEL <-gsub('HiSeq 2000','HS 2k',pop.batch$INSTRUMENT_MODEL)
+
+pop.batch<-pop.batch[complete.cases(pop.batch),]
+pop.batch<-pop.batch[!duplicated(pop.batch$Name),]
+
+Var<-aggregate(pop.batch$average_quality_of_mapped_bases,list(pop.batch$Pop),var)
+Var<-merge(Var, pop.batch, by.x='Group.1', by.y='Pop')
+plt <- Var[with(Var, order(x, average_quality_of_mapped_bases)),]
+
+Day<-aggregate(plt$SUBMISSION.DATE, list(plt$Group.1), mean)
+Day<-merge(Day, pop.batch, by.x='Group.1', by.y='Pop')
[…]

### 2018-05-11 Draft3.tex
added Mathieu as author and changed sentence about seq tech
A few minor edits and comments

@@
 \author[1,2]{Luke Anderson-Trocm\'e}
+\author[1,2]{Mathieu Bourgey}
 \author[1,2]{Simon Gravel}
@@
 Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the XYZ mutational signature in 1kGP, with low-quality individuals showing elevated rates of the signature.
-\ref{Figure1}
+(\sgcomment{show supplementary Figure where you did the correlations})
 
- individuals carrying these mutations all had lower average quality scores of mapped bases.
-This suggested that individuals with lower quality were driving this signal.
+ Thus sequences with low mappability harbor mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
@@
 
-To further investigate	 the extent to which low quality individuals were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure1}.
+To further investigate	 the extent to which low quality individuals \sgcomment{low quality sequence data!
+"low quality individuals" (= deplorables?) can get you in trouble!}  were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure1}.
+\sgcomment{I would rather have: To identify SNPs that are associated with low quality, we performed }
+
 GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
@@
 The sequencing quality of individuals increases over time \ref{MapQual}. 
-\todo{What sequencers were used in each phase?}
-By 2011 the sequencing quality seems to level off.
+By 2011 the sequencing quality seems to level off, this also coincides with the phasing out of older sequencing technologies. 
 \begin{figure}

### 2018-05-12 Draft3.tex
Updated Figure 3, minor edits introduction

@@
 
-The last 5 years have seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
-Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
-Mutations associated to a disease can be identified by comparing the genomes of healthy individuals to those afflicted by a disease. 
-The mutations commonly found in patients and rarely in controls might be associated to the disease in question.
-However, demonstrating that these mutations are biologically relevant can be difficult.
-Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
-For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies.
-Benign mutations at high frequencies in one populations might be exceedingly rare in another population.
-Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.
+%The last 5 years have seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
+%Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
+%Mutations associated to a disease can be identified by comparing the genomes of healthy individuals to those afflicted by a disease. 
+%The mutations commonly found in patients and rarely in controls might be associated to the disease in question.
+%However, demonstrating that these mutations are biologically relevant can be difficult.
+%Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
+%For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies.
+%Benign mutations at high frequencies in one populations might be exceedingly rare in another population.
+%Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.
 
-\sgcomment{This intro paragraph is much too vague -- it reads like general thoughts about genetics at large. We need to be much more concise. I would start with something like:}
+%\sgcomment{This intro paragraph is much too vague -- it reads like general thoughts about genetics at large. We need to be much more concise. I would start with something like:}
 

### 2018-05-12 Figure3.clean (Autosaved).R

@@ -1,0 +1,66 @@
+#load libraries
+library(ggplot2)
+library(cowplot)
+#pop names, used to read/write files
+pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
+#read colors per pop (used in 1kG Nature)
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
+#read the list of sig snps
+#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
+
+SFS=data.frame()
+for(f in seq(1,nrow(pops))){
+pop=pops[f,]
+print(as.character(pop))
+fname=paste('/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_',pop,'_sig6.assoc.linear',sep='')
+
+freq=read.table(fname, header=F,, col.names=c('CHR','SNP','BP','A1','TEST','NMISS','BETA','STAT','P'))
+freq$Plog10<--log10(freq$P)
+sub=freq[,c('CHR','BP','Plog10')]
+sub$Pop<-pop
+SFS=rbind(SFS,sub)
+}
+
+
+SFS<-unique(SFS)
+NumPop<-as.data.frame(table(SFS$CHR,SFS$BP))
+UnPop<-NumPop[which(NumPop$Freq==1),]
+NumPop<-NumPop[which(NumPop$Freq>1),]
+m=merge(SFS,NumPop, by.y=c('Var1','Var2'), by.x=c('CHR','BP'))
+n=as.data.frame(table(m$Pop))
+plt=merge(m,n,by.x='Pop',by.y='Var1')
+plt <- plt[with(plt, order(-Freq.x, as.numeric(CHR),as.numeric(BP))),]
+plt$BP <- factor(plt$BP, levels = unique(plt$BP))
+plt$nrow<-seq(length=nrow(plt))
+plt$Pcat<-4
+plt[which(plt$Plog10 <8),]$Pcat<-2
+plt[which((plt$Plog10 >8) & (plt$Plog10 <10)),]$Pcat<-3
+
[…]

### 2018-05-12 Figure3.clean.R

@@ -1,0 +1,72 @@
+#load libraries
+library(ggplot2)
+library(cowplot)
+#pop names, used to read/write files
+pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
+#read colors per pop (used in 1kG Nature)
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
+#read the list of sig snps
+bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
+
+SFS=data.frame()
+for(f in seq(1,nrow(pops))){
+pop=pops[f,]
+print(pop)
+fname=paste('/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_',pop,'_sig6.assoc.linear',sep='')
+if (file.exists(fname)){
+freq=read.table(fname, header=F,, col.names=c('CHR','SNP','BP','A1','TEST','NMISS','BETA','STAT','P'))
+freq<-freq[which(as.numeric(freq$P) < 0.000001),]
+freq$Plog10<--log10(freq$P)
+if(nrow(freq)>0){
+f=merge(bad, freq,by=c('CHR', 'BP'))
+if(nrow(f)>0){
+sub=f[,c('CHR','BP','Plog10')]
+sub$Pop<-pop
+SFS=rbind(SFS,sub)
+}
+}
+}
+}
+
+SFS<-unique(SFS)
+NumPop<-as.data.frame(table(SFS$CHR,SFS$BP))
+UnPop<-NumPop[which(NumPop$Freq==1),]
+NumPop<-NumPop[which(NumPop$Freq>1),]
+m=merge(SFS,NumPop, by.y=c('Var1','Var2'), by.x=c('CHR','BP'))
+n=as.data.frame(table(m$Pop))
+plt=merge(m,n,by.x='Pop',by.y='Var1')
+plt <- plt[with(plt, order(-Freq.x, as.numeric(CHR),as.numeric(BP))),]
+plt$BP <- factor(plt$BP, levels = unique(plt$BP))
+plt$nrow<-seq(length=nrow(plt))
[…]

### 2018-05-12 Draft3.tex
implemented some comments. still need to finish a few recommendations

@@
 The first sequenced genomes in the 1000 Genomes project, for example, were sequenced XX years ago, at a time when \sgcomment{blah blah blah}.
-Because of the extraordinary value of freely available data, this early data from the 1000 Genomes Project is still widely used as a reference panel for ...
+Because of the extraordinary value of freely available data, this early data from the 1000 Genomes Project is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
 
@@
 
-A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual.
-\sgcomment{This is a procedural definition.
-You need a conceptual definition!
-Something like:} Mutation rate at a given base pair has long been known to depend on the neighbouring nucleotides \cite{}.
+%A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
+%\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
+Mutation rate at a given base pair has long been known to depend on the neighbouring nucleotides \cite{}. 
 Finer scale patterns have been reported in \cite{}. 
@@
 
-In 2015, Harris et al reported an overrepresentation of the TCC to TTC mutation in European populations. \sgcomment{quantify}
+In 2015, Harris et al reported a \todo{15\%} overrepresentation of the TCC to TTC mutation in European populations.
 Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
 In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
-One result was of particular interest to our research group \sgcomment{unless needed, no need to involve our persons in the study} : the heterogeneously distributed signal found in a subset of Japanese individuals.
+One result was of particular interest the heterogeneously distributed signal found in a subset of Japanese individuals. 
 A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.
@@
 Here we discuss the source of the batch effect and the ways in which these batch effects contaminated a number of recent studies.
-Fortunately, the main conclusions of \sgcomment{most of} these studies remain supported by other data.
+Fortunately, the main conclusions of most of these studies remain supported by other data.
 Yet this begs the question: When should we retire legacy data?
@@
 
-\luke{cut paragraph to discussion}
-
-Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
+Combining data can be an issue especially when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
 The errors in one dataset don't disappear when they are combined with another higher quality one. 
@@
 If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups. 
-\sgcomment {Here again, you are a bit procedural.
-It's useful to explain to the reader the logical flow of how you got to doing what you are doing: Harris et al reported an excess of XYZ to XWZ mutations in some Japanese individuals.
-While trying to follow up on this observations in a larger cohort of XXX Japanese individuals, [[some info about cohort]], we did not find this particular signature.}
+%\sgcomment {Here again, you are a bit procedural. It's useful to explain to the reader the logical flow of how you got to doing what you are doing:} 
+Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in some Japanese individuals. 
+While trying to follow up on this observations in a larger cohort of 886 Japanese individuals, \todo{some info about cohort}, we did not find this particular signature.
 However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-\sgcomment{These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium, ...}
+These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium
 
-Once mismatch sites were removed from the 1kGP data, the XYZ - W \sgcomment{I would call the effect by the name of the signature rather than harris et al -- more descriptive, less personal.} signal disappears \ref{Figure1} \sgcomment{usually you give a descriptie name to labels, latex will figure out the numbers!}, suggesting a technical artifact rather than a population structure effect.
-Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the XYZ mutational signature in 1kGP, with low-quality individuals showing elevated rates of the signature.
+Once mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect.
+Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with low-quality individuals showing elevated rates of the signature.
 (\sgcomment{show supplementary Figure where you did the correlations})
 
- Thus sequences with low mappability harbor mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+ Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
@@
 The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
- \label{Figure1}
+ \label{SFS}
 \end{figure}
 
-To further investigate	 the extent to which low quality individuals \sgcomment{low quality sequence data!
-"low quality individuals" (= deplorables?) can get you in trouble!}  were associated to spurious mutations, we performed a genome-wide association (GWA) study  \ref{Figure1}.
-\sgcomment{I would rather have: To identify SNPs that are associated with low quality, we performed }
-
+To identify SNPs that are associated with low quality, we performed, we performed a genome-wide association (GWA) study  \ref{SFS}. 
 GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
@@
 
-Using the linear GWA study function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{Figure1}.
+Using the linear GWA study function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.
 This GWA study included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.

### 2018-05-14 Draft3.tex
more edits, updated some more comments from SG

@@
 Many of these large datasets comprise have been assembled over many years. 
-The first sequenced genomes in the 1000 Genomes project, for example, were sequenced XX years ago, at a time when \sgcomment{blah blah blah}.
+The first sequenced genomes in the 1000 Genomes project, for example, were sequenced \todo{10} years ago, at a time when the accuracy of sequencing technologies was still being acertained.
+%\sgcomment{blah blah blah}.
 Because of the extraordinary value of freely available data, this early data from the 1000 Genomes Project is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
@@
 Mutation rate at a given base pair has long been known to depend on the neighbouring nucleotides \cite{}. 
-Finer scale patterns have been reported in \cite{}. 
-\sgcomment {then define mutational signature (ideally, more of a biological definition), and move on to the Harris paper}
+Finer scale patterns have been reported in \todo{MORE HERE} \cite{}. 
+Many mutational types are caused by environmental sources or by faulty DNA replication.
+Systematic mutational patterns like thymine dimers caused by ultraviolet light, are measurable by taking the sum of all the different types of mutations in an individual.
+%\sgcomment {then define mutational signature (ideally, more of a biological definition), and move on to the Harris paper}
 
@@
 
-Combining data can be an issue especially when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
+Datasets are often combined to increase power or diversity of a dataset.
+However, combining data can be an issue especially when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
 The errors in one dataset don't disappear when they are combined with another higher quality one.
+This is rapidly becoming a relevant issue as newer technologies have lower error rates, and quality controls become more stringent.
 Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-Is it time to retire legacy data? 
-\sgcomment{As a style issue: you tend to write sentences without indicating the logical flow between them. It makes it hard to follow. }
+Instead of blindly combining old data with new data, is it time to retire legacy data that does not meet modern day sequencing quality thresholds?
+
+%\sgcomment{As a style issue: you tend to write sentences without indicating the logical flow between them. It makes it hard to follow. }
 

### 2018-05-14 Draft3.tex
slight changes, caught typos, removed some duplicated words

@@
 While trying to follow up on this observation, we were unable to reproduce this particular observation in a different dataset.
-Tracing back the source of the discrepancy between the two datasets, we identified a strong and previously unreported batch effect in the 1000 Genomes Project data, and found that this batch effect led to spurious results in a number of recently published studies.
+Tracing back the source of the discrepancy between the two datasets, we identified a strong and previously unreported batch effect in the 1000 Genomes Project data, and found that this led to spurious results in a number of recently published studies.
 
-Here we discuss the source of the batch effect and the ways in which these batch effects contaminated a number of recent studies.
+Here we discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
 Fortunately, the main conclusions of most of these studies remain supported by other data.
@@
 Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-Instead of blindly combining old data with new data, is it time to retire legacy data that does not meet modern day sequencing quality thresholds?
 
@@
 
-To identify SNPs that are associated with low quality, we performed, we performed a genome-wide association (GWA) study  \ref{SFS}. 
+To identify SNPs that are associated with low quality, we performed a genome-wide association (GWA) study  \ref{SFS}. 
 GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
@@
 Despite removing the variants identified as being low quality, the signal persists. 
-When we remove the variants significantly associated to low quality, the signal identified by Harris et. al (2017) the enrichment in *AC${\rightarrow}$*CC mutations. 
+When we remove the variants significantly associated to low quality, the enrichment in *AC${\rightarrow}$*CC mutations remains present, but \todo{less} significant. 
 When we remove individuals with average quality of mapped SNPs below 30, the signal goes away. 
@@
 The populations that have the most low-quality individuals also have the most low-quality variants. 
-What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations.
-}
+The same variants identified as being low quality independently in each population are found in other populations. }
   \label{Figure3}
@@
 
-Five of these studies used the 1000 Genomes Project as the reference database for imputation and one used the 1000 Genomes Project cell cultures and sequence data. 
-They used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, allele frequency differences using reference populations. 
+Five of these studies used the 1kGP as the reference database for imputation and one used the 1kGP cell cultures and sequence data. 
+They used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness. 
@@
 These SNPs matter because they reached genome wide significance for medically relevant traits. 
-Could lead to false diagnosis at worst, or spurious correlations at least. 
-Polygenic risk scores take into account all snps reaching significance, without much manual curation. 
+Including these variants in the GWAS catalogue could lead to false diagnosis at worst, or spurious correlations at least. 
+Polygenic risk scores take into account all SNPs reaching significance, without much manual curation. 
 It's likely that the variants we have identified are being included in these multi-locus risk scores.
@@
 We propose including our quality control methods to identify possible false positives in sequencing data. 
-We have focused on the 1000 Genomes Project dataset as its quality metrics were feely available, however the issues of quality control are not limited to this one consortium. 
+We have focused on the 1000 Genomes Project dataset as its quality metrics were freely available, however the issues of quality control are not limited to this one consortium. 
 This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.
 
-As more and more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the amplification of false positives. 
+As more and more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
 These results bring forth many questions regarding the reliability of legacy datasets. 

### 2018-05-14 legacy.tex
added author, changed manuscript title

--- 2018-05-14 Draft3.tex
@@
 \author[1,2]{Mathieu Bourgey}
+\author[3]{Fumihiko Matsuda}
 \author[1,2]{Simon Gravel}
@@
 \affil[2]{McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada}
-
+\affil[3]{Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
 \corr{simon.gravel@mcgill.ca}{SG}

### 2018-05-15 legacy.tex
updated Motivations
added a few sentences at the start

@@
 
-The last 5 years have seen a drastic increase in the amount and quality of human genomes sequence data. 
+The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
 Hapmap \cite{}, the 1000 Genomes Project \cite{}, and the Simons Diversity project \cite{}, for example, have made thousands of genomes publicly available for population and medical genetic analyses. 
@@
 
-Many of these large datasets comprise have been assembled over many years. 
-The first sequenced genomes in the 1000 Genomes project, for example, were sequenced \todo{10} years ago, at a time when the accuracy of sequencing technologies was still being acertained.
+Many of these large datasets have been assembled over many years. 
+The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when the accuracy of sequencing technologies was still being ascertained.
 %\sgcomment{blah blah blah}.
-Because of the extraordinary value of freely available data, this early data from the 1000 Genomes Project is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
+Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
 
@@
 %\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
-Mutation rate at a given base pair has long been known to depend on the neighbouring nucleotides \cite{}. 
-Finer scale patterns have been reported in \todo{MORE HERE} \cite{}. 
-Many mutational types are caused by environmental sources or by faulty DNA replication.
-Systematic mutational patterns like thymine dimers caused by ultraviolet light, are measurable by taking the sum of all the different types of mutations in an individual.
-%\sgcomment {then define mutational signature (ideally, more of a biological definition), and move on to the Harris paper}
+Mutations can occur in the genome as a result of environmental inflictions as well as genetic factors like DNA replication and repair \cite{}.
+Many mutagens leave signatures that are specific to a genomic context \cite{}.
+This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation \cite{}.
+Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
+However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently \cite{}.
+A large sample size can be used to resolve some of these fine scale mutational signatures \cite{}.
+%Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.
 

### 2018-05-15 legacy.tex
implemented some of alex's comments

@@
 \newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
+\newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Many of these large datasets have been assembled over many years. 
-The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when the accuracy of sequencing technologies was still being ascertained.
-%\sgcomment{blah blah blah}.
+The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when sequencing technologies were still being . %\sgcomment{blah blah blah}. 
 Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
@@
 %\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
-Mutations can occur in the genome as a result of environmental inflictions as well as genetic factors like DNA replication and repair \cite{}.
+Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair \cite{}.
 Many mutagens leave signatures that are specific to a genomic context \cite{}.
@@
 Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
-However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently \cite{}.
-A large sample size can be used to resolve some of these fine scale mutational signatures \cite{}.
+However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently and are more subtle \cite{}.
+A large sample size can be used to increase statistical power to resolve some of these fine scale mutational signatures \cite{}.
 %Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.
@@
 Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
-In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that these changes are present in most human populations. 
+In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that many populations have their own mutational signatures.
 One result was of particular interest the heterogeneously distributed signal found in a subset of Japanese individuals. 
@@
 
-Datasets are often combined to increase power or diversity of a dataset.
-However, combining data can be an issue especially when the quality of the data produced can vary between sequencing technologies and even sequencing centres.
-The errors in one dataset don't disappear when they are combined with another higher quality one.
-This is rapidly becoming a relevant issue as newer technologies have lower error rates, and quality controls become more stringent.
+Many types of DNA sequencing machines have been available on the market over the past decade.
+Different types of sequencing technologies can have different technical limitations in quality control.
+This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
 Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
+The errors in one dataset do not disappear when they are combined with another higher quality one.
+The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
 
@@
 \textbf{B} 
-Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than  in a in a GWA of sequencing quality. 
+Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than in a GWA of sequencing quality. 
 There is a \todo{significant} enrichment in *AC${\rightarrow}$*CC mutations. 
@@
 
+\alex{Having trouble following the first three sentences}
 Despite removing the variants identified as being low quality, the signal persists. 
@@
 
-As more and more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
+As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
 These results bring forth many questions regarding the reliability of legacy datasets. 

### 2018-05-15 Figure3.clean (Autosaved).R

@@ -1,66 +1,0 @@
-#load libraries
-library(ggplot2)
-library(cowplot)
-#pop names, used to read/write files
-pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
-#read colors per pop (used in 1kG Nature)
-colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
-MyColour <- as.character(colors $hex)
-names(MyColour) <- colors $Pop
-
-#read the list of sig snps
-#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
-
-SFS=data.frame()
-for(f in seq(1,nrow(pops))){
-pop=pops[f,]
-print(as.character(pop))
-fname=paste('/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_',pop,'_sig6.assoc.linear',sep='')
-
-freq=read.table(fname, header=F,, col.names=c('CHR','SNP','BP','A1','TEST','NMISS','BETA','STAT','P'))
-freq$Plog10<--log10(freq$P)
-sub=freq[,c('CHR','BP','Plog10')]
-sub$Pop<-pop
-SFS=rbind(SFS,sub)
-}
-
-
-SFS<-unique(SFS)
-NumPop<-as.data.frame(table(SFS$CHR,SFS$BP))
-UnPop<-NumPop[which(NumPop$Freq==1),]
-NumPop<-NumPop[which(NumPop$Freq>1),]
-m=merge(SFS,NumPop, by.y=c('Var1','Var2'), by.x=c('CHR','BP'))
-n=as.data.frame(table(m$Pop))
-plt=merge(m,n,by.x='Pop',by.y='Var1')
-plt <- plt[with(plt, order(-Freq.x, as.numeric(CHR),as.numeric(BP))),]
-plt$BP <- factor(plt$BP, levels = unique(plt$BP))
-plt$nrow<-seq(length=nrow(plt))
-plt$Pcat<-4
-plt[which(plt$Plog10 <8),]$Pcat<-2
-plt[which((plt$Plog10 >8) & (plt$Plog10 <10)),]$Pcat<-3
-
[…]

### 2018-05-15 Figure3.clean.R

@@ -9,27 +9,21 @@ MyColour <- as.character(colors $hex)
 names(MyColour) <- colors $Pop
 
 #read the list of sig snps
-bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
+#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
 
 SFS=data.frame()
 for(f in seq(1,nrow(pops))){
 pop=pops[f,]
-print(pop)
+print(as.character(pop))
 fname=paste('/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_',pop,'_sig6.assoc.linear',sep='')
-if (file.exists(fname)){
+
 freq=read.table(fname, header=F,, col.names=c('CHR','SNP','BP','A1','TEST','NMISS','BETA','STAT','P'))
-freq<-freq[which(as.numeric(freq$P) < 0.000001),]
 freq$Plog10<--log10(freq$P)
-if(nrow(freq)>0){
-f=merge(bad, freq,by=c('CHR', 'BP'))
-if(nrow(f)>0){
-sub=f[,c('CHR','BP','Plog10')]
+sub=freq[,c('CHR','BP','Plog10')]
 sub$Pop<-pop
 SFS=rbind(SFS,sub)
 }
-}
-}
-}
+
 
 SFS<-unique(SFS)
 NumPop<-as.data.frame(table(SFS$CHR,SFS$BP))

### 2018-05-15 HweVsGwas.R

@@ -1,0 +1,19 @@
+##Hardy GWAS qq plot
+library(ggplot2)
+Pops<-read.table('~/genomes/genomes/PopNames/NamePop.txt')
+for(f in seq(1,26,1)){
+pop=Pops[f,]
+print(pop)
+
+fname=paste('~/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_hwe_GWAS_',pop,'_ND.txt',sep='')
+t5<-read.table(fname,header=T,nrows=5)
+classes <- sapply(t5, class)
+t <- read.table(fname, header = TRUE, colClasses = classes, col.names=c('CHR','BP','HWE','GWAS'))
+t<-t[which((t$HWE != 1) & (t$GWAS != 1)),]
+
+t$P.HWE<--log10(t$HWE)
+t$P.GWAS<--log10(t$GWAS)
+
+ggplot(t, aes(x=P.HWE, y=P.GWAS))+geom_bin2d(binwidth=0.15)+theme_classic()+labs(x='Hardy-Weinberg -log10(p)',y='GWAS -log10(p)',title=pop)+scale_fill_gradient(trans='log',breaks=c(1,10,100,1000,10000,100000,1000000,10000000))
+ggsave(paste('~/Documents/GWAS_Qual/HWEvsGWAS_',pop,'.jpg',sep=''))
+}

### 2018-05-15 legacy.tex
some style edits

@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Hapmap \cite{}, the 1000 Genomes Project \cite{}, and the Simons Diversity project \cite{}, for example, have made thousands of genomes publicly available for population and medical genetic analyses. 
+Hapmap [\cite{}], the 1000 Genomes Project [\cite{}], and the Simons Diversity project [\cite{}], for example, have made thousands of genomes publicly available for population and medical genetic analyses.
 Many more genomes are available indirectly through servers providing imputation services \cite{haplotype reference consortium} or variant frequency estimation \cite{GnomAD}.
@@
 Many of these large datasets have been assembled over many years. 
-The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, at a time when sequencing technologies were still being . %\sgcomment{blah blah blah}. 
+The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, before most of the current best technologies and analysis practices were in place. 
 Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
@@
 Yet this data is far from perfect \sgcomment{cite papers that have pointed out issues}.
-This raises the question of whether and how such legacy data should be included in contemporary analyses.
+This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
 Here we point out how large and previously unreported batch effects in the early phases of the 1000 Genomes Project still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
@@
 %\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
-Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair \cite{}.
+Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair [\cite{}].
 Many mutagens leave signatures that are specific to a genomic context \cite{}.
@@
 In 2015, Harris et al reported a \todo{15\%} overrepresentation of the TCC to TTC mutation in European populations.
-Harris and Pritchard published a follow up study comparing how various ethnic groups differ from one another with respect to their mutational spectrum. 
-In this paper, they proposed that not only is there a distinct mutational spectrum difference in Europeans, but that many populations have their own mutational signatures.
-One result was of particular interest the heterogeneously distributed signal found in a subset of Japanese individuals. 
-A mutational signature that is present unevenly across a population is unexpected because it suggested strong population structure; unlikely for a dense population living on an archipelago.
+This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population, and was replicated in a different cohort \cite{}.
+Harris and Pritchard further identified distinct mutational spectra across a range of populations [\cite{}].
+In particular, they identified an axis of variation within 1000 Genomes Japanese individuals.
+Such an axis of variation is intriguing because differences in mutational signatures accumulate over many generations.
+A systematic difference among the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
+We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort.
+}
 
-While trying to follow up on this observation, we were unable to reproduce this particular observation in a different dataset.
-Tracing back the source of the discrepancy between the two datasets, we identified a strong and previously unreported batch effect in the 1000 Genomes Project data, and found that this led to spurious results in a number of recently published studies.
+While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy between the two datasets and identified a strong and previously unreported batch effect in the 1000 Genomes Project data.
+We also found that this batch effect led to spurious results in a number of recently published studies.
 
+\sgcomment{In section ..., we...}
 Here we discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
@@
 %\sgcomment{Too wordy (also, readers of Nature genetics know about imputation!: Despite drastic reduction in sequencing costs, genotyping and imputation remain a more affordable and popular approach for large-scale cohorts.}
-
-Many types of DNA sequencing machines have been available on the market over the past decade.
-Different types of sequencing technologies can have different technical limitations in quality control.
-This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
-Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-The errors in one dataset do not disappear when they are combined with another higher quality one.
-The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
 
@@
 
+Many types of DNA sequencing machines have been available on the market over the past decade.
+Different types of sequencing technologies can have different technical limitations in quality control.
+This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
+Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
+The errors in one dataset do not disappear when they are combined with another higher quality one.
+The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
+
 \section{Methods}

### 2018-05-16 legacy.tex
minor edits

@@
 
-Many of these large datasets have been assembled over many years. 
+Many of these large datasets have been assembled over several years. 
 The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, before most of the current best technologies and analysis practices were in place. 
@@
 %\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:}
-Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair [\cite{}].
-Many mutagens leave signatures that are specific to a genomic context \cite{}.
-This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation \cite{}.
-Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
-However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently and are more subtle \cite{}.
-A large sample size can be used to increase statistical power to resolve some of these fine scale mutational signatures \cite{}.
+
+Mutations in the genome can be caused by exposure to a mutagen, defective DNA repair or errors in DNA replication.
+
+Mutation rates depend on environmental exposure as well as variation in genetic factors like DNA proofreading and repair [\cite{}].
+Some mutagens are more likely to affect a specific genomic context which therefore leaves a mutational signature [\cite{}].
+It has been shown that certain types of cancer have particular mutational profiles that can be used as a diagnostic tool [\cite{}].
+
+This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation [\cite{}].
+Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study [\cite{}].
+However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently and are more subtle [\cite{}].
+A large sample size can be used to increase statistical power to resolve some of these fine scale mutational signatures [\cite{}].
 %Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.

### 2018-05-16 legacy.tex
moderate edits
cleaned up commented out sections of the text
cleaned up the intro, and modified the motivations section to include cancer research instead of thymine dimers

--- 2018-05-15 legacy.tex
@@
 
-%The last 5 years have seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
-%Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
-%Mutations associated to a disease can be identified by comparing the genomes of healthy individuals to those afflicted by a disease. 
-%The mutations commonly found in patients and rarely in controls might be associated to the disease in question.
-%However, demonstrating that these mutations are biologically relevant can be difficult.
-%Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
-%For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies.
-%Benign mutations at high frequencies in one populations might be exceedingly rare in another population.
-%Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.
-
-%\sgcomment{This intro paragraph is much too vague -- it reads like general thoughts about genetics at large. We need to be much more concise. I would start with something like:}
-
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
@@
 
-Many of these large datasets have been assembled over many years. 
-The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced \todo{10} years ago, before most of the current best technologies and analysis practices were in place. 
+Many of these large datasets have been assembled over several years using a variety of sequencing platforms. 
+The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced 10 years ago.
+During this period, sequencing platforms were still being rapidly developed.
+This lead to heterogeneous sample preparations for many of the samples included in the dataset.
 Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
 
-Yet this data is far from perfect \sgcomment{cite papers that have pointed out issues}.
+Although, it is clear that the quality and accuracy of the 1kGP data has increased over the past decade. 
+This data is far from perfect, the publication from the second phase of the 1kGP describe technological and analytical improvements over its previous phases.
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts. 
-Here we point out how large and previously unreported batch effects in the early phases of the 1000 Genomes Project still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP that still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
 
@@
 
-%A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
-%\sgcomment{This is a procedural definition. You need a conceptual definition! Something like:} 
-Mutations can occur in the genome as a result of environmental inflictions as well as variation in genetic factors like DNA proofreading and repair [\cite{}].
-Many mutagens leave signatures that are specific to a genomic context \cite{}.
-This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation \cite{}.
-Thymine dimers caused by UV radiation were detected early on because they were highly specific and easy to study \cite.
-However,  signatures caused by natural variations in the sequence of proteins involved in proofreading and repair are more difficult to study because they tend to occur less frequently and are more subtle \cite{}.
-A large sample size can be used to increase statistical power to resolve some of these fine scale mutational signatures \cite{}.
-%Finer scale patterns have been reported in \todo{MORE HERE} \cite{}.
+Subtle genome wide mutational signatures have been used as a diagnostic tool for cancers.
+Distinct types of mutations that occur more frequently in certain cancers can be used to identify the cause and mechanisms of cancer [\cite].
+Certain mutagens in tobacco smoke have been shown to preferentially bind to certain genomic motifs causing an excess of G to T transversions [\cite{}].
+The excess of certain types of mutations can be measured as a genome wide mutational signature.
+This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation [\cite{}].
+While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity [\cite{}].
+Taking advantage of large cohorts can increase statistical power to resolve some of these fine scale mutational signatures [\cite{}].
 
-In 2015, Harris et al reported a \todo{15\%} overrepresentation of the TCC to TTC mutation in European populations.
-This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population, and was replicated in a different cohort \cite{}.
+Genome wide mutational signatures are also observable on evolutionary scales by comparing the rates of mutations between populations. 
+In 2015, Harris et al reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations.
+This strong population enrichment of a mutational signature was replicated in a different cohort; suggests important genetic or environmental differences in the history of each population \cite{}. 
 Harris and Pritchard further identified distinct mutational spectra across a range of populations [\cite{}]. 
-In particular, they identified an axis of variation within 1000 Genomes Japanese individuals.
+In particular, they identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals. 
 Such an axis of variation is intriguing because differences in mutational signatures accumulate over many generations. 
 A systematic difference among the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow. 
-We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort.
-}
+We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort. }
 
-While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy between the two datasets and identified a strong and previously unreported batch effect in the 1000 Genomes Project data.
+While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy between the two datasets and identified a strong and previously unreported batch effect in the 1kGP data.
 We also found that this batch effect led to spurious results in a number of recently published studies.
 
-\sgcomment{In section ..., we...}
-Here we discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
+This study is the result of an investigation in the quality of the 1kGP dataset. 
+To begin, we will discuss how we discovered these previously unreported batch effects in the 1kGP. 
+Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP. 
+Finally, we will discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
 Fortunately, the main conclusions of most of these studies remain supported by other data.
 Yet this begs the question: When should we retire legacy data?
-
-%\subsection{Imputation}
-%Genome wide associations using large cohorts have lead to developments in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases.
-Despite drastic reductions in cost of whole genome sequencing, it remains an expensive test for large sample sizes.
-%For this reason, genotype data is often imputed to increase power for association studies in a cost effective way.
-
-%\sgcomment{Too wordy (also, readers of Nature genetics know about imputation!: Despite drastic reduction in sequencing costs, genotyping and imputation remain a more affordable and popular approach for large-scale cohorts.}
-
-%\sgcomment{As a style issue: you tend to write sentences without indicating the logical flow between them. It makes it hard to follow. }
 
@@
 \textbf{A} 
-Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent dataset.
+Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent dataset.
 Crosses ( + ) are variants that reached genome wide significance in a GWA of sequencing quality. 
@@
 \textbf{C} 
-Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1000 Genomes Project.
-This GWA study identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
+Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1kGP.
+This GWAS identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
+The same analysis was performed independently for each of the populations in the 1kGP. }
  \label{SFS}
@@
 
-Using the linear GWA study function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.
-This GWA study included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project.
+Using the linear GWAS function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.
+This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
 The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations.
@@
 	\subsection{Overlap of significant SNPs}
-Comparing the results of each independent GWA study, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations.
+Comparing the results of each independent GWAS, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations.
 This confirmation using more than one GWAS is a strongly suggests that these variants might not be genuine \ref{Figure3}.
@@
 \subsection{GWA studies from other papers}
-Since these variants are present in more than one population from the 1000 genomes project, they are more likely to be associated to biological traits as they would appear to be like any other variant that is shared among multiple populations.
+Since these variants are present in more than one population from the 1kGP, they are more likely to be associated to biological traits as they would appear to be like any other variant that is shared among multiple populations.
 The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
@@
 
-Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1000 Genomes Project.
+Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1kGP.
 We used the average quality of mapped bases per individual as the phenotype for the analysis.
@@
 \subsection{Imputation}
-Using the Michigan Imputation Server, we imputed the genotype data from 1000 Genomes Project
+Using the Michigan Imputation Server, we imputed the genotype data from 1kGP
 

### 2018-05-16 legacy.tex
moderate changes
Added bibliography (not functional)

Modified results.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Hapmap [\cite{}], the 1000 Genomes Project [\cite{}], and the Simons Diversity project [\cite{}], for example, have made thousands of genomes publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \cite{haplotype reference consortium} or variant frequency estimation \cite{GnomAD}.
+Hapmap \cite{HapMap2005}, the 1000 Genomes Project \cite{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \cite{Mallick2016}, for example, have made thousands of genomes publicly available for population and medical genetic analyses.
+Many more genomes are available indirectly through servers providing imputation services \cite{ProfJonathanMarchiniProfGoncaloAbecasisProfRichardDurbin2014} or variant frequency estimation \cite{Lek2016}.
 
@@
 
-Subtle genome wide mutational signatures have been used as a diagnostic tool for cancers.
-Distinct types of mutations that occur more frequently in certain cancers can be used to identify the cause and mechanisms of cancer [\cite].
-Certain mutagens in tobacco smoke have been shown to preferentially bind to certain genomic motifs causing an excess of G to T transversions [\cite{}].
+Genome wide mutational signatures have been used as a diagnostic tool for cancers\cite{Pleasance2010,Shiraishi2015a}.
+Distinct types of mutations that occur more frequently in certain cancers can be used to identify the cause and mechanisms of cancer \cite{Pleasance2010,Shiraishi2015a}.
+Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \cite{Pleasance2010}.
 The excess of certain types of mutations can be measured as a genome wide mutational signature.
@@
 While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity [\cite{}].
-Taking advantage of large cohorts can increase statistical power to resolve some of these fine scale mutational signatures [\cite{}].
+Large cohorts have recently been used to increase statistical power to resolve some of these fine scale mutational signatures.
 
 Genome wide mutational signatures are also observable on evolutionary scales by comparing the rates of mutations between populations. 
-In 2015, Harris et al reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations.
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations.
 This strong population enrichment of a mutational signature was replicated in a different cohort; suggests important genetic or environmental differences in the history of each population \cite{}. 
@@
 While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy between the two datasets and identified a strong and previously unreported batch effect in the 1kGP data.
-We also found that this batch effect led to spurious results in a number of recently published studies.
+We also found that this batch effect led to spurious results in a number of recent publications.
 
@@
 To begin, we will discuss how we discovered these previously unreported batch effects in the 1kGP. 
-Next, we explore more broadly how low quality variants remain embedded in other populations of the 1kGP. 
+Next, we explore more broadly how low quality data remain embedded in other populations of the 1kGP. 
 Finally, we will discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
@@
 If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups. 
-%\sgcomment {Here again, you are a bit procedural. It's useful to explain to the reader the logical flow of how you got to doing what you are doing:} 
-Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in some Japanese individuals. 
+Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP. 
 While trying to follow up on this observations in a larger cohort of 886 Japanese individuals, \todo{some info about cohort}, we did not find this particular signature.
@@
 Once mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect. 
-Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with low-quality individuals showing elevated rates of the signature.
-(\sgcomment{show supplementary Figure where you did the correlations})
-
+Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\sgcomment{show supplementary Figure where you did the correlations}
 Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 \textbf{A} 
+Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
+There is a 10 fold enrichment in *AC${\rightarrow}$*CC mutations. 
+\textbf{B} 
 Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent dataset.
@@
 The lowest frequency variant able to reach genome wide significance is 5\%.
-\textbf{B} 
-Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than in a GWA of sequencing quality. 
-There is a \todo{significant} enrichment in *AC${\rightarrow}$*CC mutations. 
 \textbf{C} 
@@
 
-To identify SNPs that are associated with low quality, we performed a genome-wide association (GWA) study  \ref{SFS}. 
-GWA studies are commonly used to find genomic regions associated to biologically relevant traits.
-In this case, we are using this analysis in a more unconventional way to identify regions of the genome that are associated to biologically irrelevant traits like SNPs associated to individuals with low quality.
+To identify SNPs that are associated with low quality, we performed a genome-wide association study (GWAS)  \ref{SFS}. 
+GWAS are commonly used to find genomic regions associated to biologically relevant traits.
+In this case, we are using this analysis to identify regions of the genome that are associated to sequencing quality.
 
@@
 
-\alex{Having trouble following the first three sentences}
-Despite removing the variants identified as being low quality, the signal persists. 
-When we remove the variants significantly associated to low quality, the enrichment in *AC${\rightarrow}$*CC mutations remains present, but \todo{less} significant. 
-When we remove individuals with average quality of mapped SNPs below 30, the signal goes away. 
-We suspect that this has to do with the lack of power we have for more rare alleles. 
-The lowest frequency variants that are significantly associated to quality are at 5\%. 
-When we remove all the variants above 1\% in 1kGP but missing from the NAG data, the signal disappears.
-This suggests that while the GWA approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
+Since the GWAS we performed has a power limit at 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality metrics.
+The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data, as the removal of individuals with average qualities below 30 successfully removes the signal.
 
-One way to assess the validity of these unusual variants is to see if they are present in a higher quality dataset.
-Upon comparing the Japanese 1000 Genomes cohort to a higher quality and larger cohort, we identified \todo{000} more variants that are beyond the expected frequency spectrum deviation for individuals from the same population.
+By taking advantage of the higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the NAG data.
+This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
 
 	\subsection{Sequencing quality over time}
-We turned our attention to the other populations in the 1kGP, we found that the sequencing done in phase 1 was more variable and overall tended to include lower quality sequencing data \ref{MapQual}.
-The sequencing quality of individuals increases over time \ref{MapQual}. 
-By 2011 the sequencing quality seems to level off, this also coincides with the phasing out of older sequencing technologies. 
+The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data \ref{MapQual}.
+This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centers.
+There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
+By 2011, older sequencing technologies are phased out.
+This coincides with the sequencing quality leveling off.
+
 \begin{figure}
@@
 \textbf{B} Same data as in \textbf{A} except the x-axis is sorted by sequencing date.
-The colors indicate the sequencing centers that produced the data for each individual.}
+The colours indicate the sequencing centers that produced the data for each individual.}
 \label{MapQual}
@@
 	\subsection{Overlap of significant SNPs}
-Comparing the results of each independent GWAS, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations.
-This confirmation using more than one GWAS is a strongly suggests that these variants might not be genuine \ref{Figure3}.
+
+We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
+We found that most populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
+\todo{Supplemental Figures : all the GWAS results for each population}
+Comparing the results of each independent GWAS, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations  \ref{OverLap}. 
+This confirmation using more than one GWAS is a strongly suggests that these variants might not be genuine \ref{OverLap}.
 
@@
 The same variants identified as being low quality independently in each population are found in other populations. }
-  \label{Figure3}
+  \label{OverLap}
 \end{figure*}
@@
 	\subsection{Imputation}
-~30\% of the SNPs we identified as being associated with low quality were found to be imputed using the Michigan Imputation Server. These should be removed from reference database.
+The state of the art imputations servers use a combination of many databases including many that are not freely available.
+For this reason, the details of the methods used to impute genomic data are not publicly available either.
+In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted the genotype data from the 1kGP and imputed it using the Michigan Imputation Server.
+We found that \todo{~30\%} of the SNPs we identified as being associated with quality were imputed.
 
 	\subsection{Found to be included in other GWA studies}
-Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWA studies that might have called these erroneous variants as being significantly correlated with some biological trait.
+Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have called these erroneous variants as being significantly correlated with some biological trait.
 Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found 6 recent publications that had found at least one of the variants to have reached genome wide significance in their study.
@@
 
+\subsection{Power limits and Quality Controls}
+If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
+\todo{how many hetd do you need to detect HWE significance?}
+We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.
+However even the GWAS is limited by power as we were able to identify \todo{000} more variants that were missing in a higher quality dataset but present in over 1\% of the 1kGP Japanese data. 
+This suggests that while the GWAS we performed is a step in the right direction, the GWAS is not sufficient because the data is not reliable in the first place.
+
+%\subsection{Technical Source of Error}
+%issues with PCR?
+%repetitive regions, GC bias...
+
 \subsection{Imputation}
@@
 
-\luke{due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?}
+%\luke{due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?}
 
@@
 
+\bibliography{/Users/luke/Dropbox/Work/PHD/BibTeX/Legacy.bib}
+
 \end{document}

### 2018-05-18 legacy.tex
fixed bibliography
some edits and suggestions

@@
 	\subsection{A peculiar mutational signature in Japan}
-If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups. 
-Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP. 
-While trying to follow up on this observations in a larger cohort of 886 Japanese individuals, \todo{some info about cohort}, we did not find this particular signature.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium
 
-Once mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect. 
-Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+Harris et al \cite{} reported a $XX\%$  excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP. 
+While trying to follow up on this interesting observations in a larger cohort of 886 Japanese individuals, \todo{some info about cohort}, we did not find this particular signature.
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) that are private to one or the other cohort.
+This is unexpected given the similarity of the two populations, and suggests a technical artifact rather than a population structure effect. 
+However, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium
+
+Once mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}. 
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
 \sgcomment{show supplementary Figure where you did the correlations}
-Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+Thus sequenced individuals with \sgcomment{low mappability harbour?} mutations that reproduce poorly across cohorts and exhibit a particular mutational signature.
 
@@
 
-To identify SNPs that are associated with low quality, we performed a genome-wide association study (GWAS)  \ref{SFS}. 
-GWAS are commonly used to find genomic regions associated to biologically relevant traits.
-In this case, we are using this analysis to identify regions of the genome that are associated to sequencing quality.
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality \ref{SFS}.
+Using the linear regression as implemented in PLINK, we were able to identify $625$ snips with $p < 10^{-8}$ and $1048$ snps with $ p < 10^{-6}.$ While identifying putative low-quality snps to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
+ This GWAS included 104 Japanese individuals, \todo{000} of which were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
+The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations. 
+\sgcomment{Signal is enriched in GWAS snps, but residual signal remains because...}
 
-Using the linear GWAS function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
-The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations.
-
-Since the GWAS we performed has a power limit at 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality metrics.
+A GWAS threshold of $10^{-6}$ means that variants of less than 5\% minor allele frequency  (MAF) cannot be significantly associated, even if they are only found in low-quality sequences. 
+They therefore remain unidentifiable using association techniques.
+For this reason, the *AC${\rightarrow}$*CC mutational signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality metrics.
 The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data, as the removal of individuals with average qualities below 30 successfully removes the signal.
@@
 By taking advantage of the higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the NAG data.
+\sgcomment{redundant}
 This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
@@
 
-\bibliography{/Users/luke/Dropbox/Work/PHD/BibTeX/Legacy.bib}
+\bibliography{Legacy.bib}
 

### 2018-06-18 legacy.tex
Cleaned up results, added table

--- 2018-05-16 legacy.tex
@@
 
-Genome wide mutational signatures have been used as a diagnostic tool for cancers\cite{Pleasance2010,Shiraishi2015a}.
-Distinct types of mutations that occur more frequently in certain cancers can be used to identify the cause and mechanisms of cancer \cite{Pleasance2010,Shiraishi2015a}.
+Genome wide mutational signatures have been used as a diagnostic tool for cancers \cite{Pleasance2010,Shiraishi2015a}.
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \cite{Pleasance2010}.
-The excess of certain types of mutations can be measured as a genome wide mutational signature.
-This means that the neighbouring nucleotides of a mutation can be used to gain insight into the cause of a mutation [\cite{}].
+This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation [\cite{}].
 While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity [\cite{}].
-Large cohorts have recently been used to increase statistical power to resolve some of these fine scale mutational signatures.
+Large cohorts have recently become available to increase statistical power to resolve some of these fine scale mutational signatures.
 
-Genome wide mutational signatures are also observable on evolutionary scales by comparing the rates of mutations between populations. 
-In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations.
-This strong population enrichment of a mutational signature was replicated in a different cohort; suggests important genetic or environmental differences in the history of each population \cite{}. 
+Fine scale mutational signatures are observable on evolutionary scales by comparing the rates of mutations between populations.
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 [\cite{}].
+This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population [\cite{}]. 
 Harris and Pritchard further identified distinct mutational spectra across a range of populations [\cite{}]. 
@@
 A systematic difference among the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort. }
+We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort.}
 
@@
 Next, we explore more broadly how low quality data remain embedded in other populations of the 1kGP. 
-Finally, we will discuss the source of the batch effect and the ways in which it contaminated a number of recent studies.
+Finally, we will discuss the ways in which it contaminated a number of recent studies.
 Fortunately, the main conclusions of most of these studies remain supported by other data.
@@
 	\subsection{A peculiar mutational signature in Japan}
+
+Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
+While trying to follow up on this observations in a larger Japanese cohort individuals, we did not find this particular signature.
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
 If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups.
-Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP. 
-While trying to follow up on this observations in a larger cohort of 886 Japanese individuals, \todo{some info about cohort}, we did not find this particular signature.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium
+Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
 
-Once mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect. 
-Regressions against different quality metrics provided by the 1kGP revealed that mean mapping quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
-\sgcomment{show supplementary Figure where you did the correlations}
+When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect.
+Regressions against different quality metrics provided by the 1kGP revealed that mean quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
+Individuals with low-quality data have elevated rates of the *AC${\rightarrow}$*CC signature. 
+\sgcomment{supplementary figure of correlations}
 Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 To identify SNPs that are associated with low quality, we performed a genome-wide association study (GWAS)  \ref{SFS}. 
-GWAS are commonly used to find genomic regions associated to biologically relevant traits.
-In this case, we are using this analysis to identify regions of the genome that are associated to sequencing quality.
+We used the mean quality of mapped bases of each individual as the association phenotype.
+Thus, variants that are only found in individuals with low quality data will display significant associations to quality. 
+We were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.  
+This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
+The variants that are associated to the quality of mapped bases have a 10 fold enrichment in *AC${\rightarrow}$*CC mutations.
 
-Using the linear GWAS function offered by PLINK, we were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
-The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations.
+The GWAS has a power limit of 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
+The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data.
+Therefore, the removal of individuals with average qualities below 30 successfully removes the signal.
 
-Since the GWAS we performed has a power limit at 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality metrics.
-The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data, as the removal of individuals with average qualities below 30 successfully removes the signal.
-
-By taking advantage of the higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the NAG data.
+By taking advantage of the larger, higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the \todo{NAG} data.
 This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
 
+%Many of the populations in the 1kGP do not have higher quality counterparts which means that identifying questionably low quality data must be done ad hoc.
+
 	\subsection{Sequencing quality over time}
+The results of the GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
 The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data \ref{MapQual}.
-This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centers.
+This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
-By 2011, older sequencing technologies are phased out.
-This coincides with the sequencing quality leveling off.
+By 2011, older sequencing technologies are phased out, and method development solidify, which coincides with the sequencing quality levelling off.
 
@@
 We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
-We found that most populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
+We found that 25 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
 \todo{Supplemental Figures : all the GWAS results for each population}
-Comparing the results of each independent GWAS, we were able to identify over \todo{0000} variants that were independently associated to low quality in multiple populations  \ref{OverLap}. 
-This confirmation using more than one GWAS is a strongly suggests that these variants might not be genuine \ref{OverLap}.
+Comparing the results of each independent GWAS, we were able to identify over 3625 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
 
@@
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have called these erroneous variants as being significantly correlated with some biological trait.
-Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found 6 recent publications that had found at least one of the variants to have reached genome wide significance in their study.
+Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found 17 recent publications that had found at least one of the variants to have reached genome wide significance in their study.
 
-Five of these studies used the 1kGP as the reference database for imputation and one used the 1kGP cell cultures and sequence data. 
+These studies used the 1kGP as the reference database for imputation. %and one used the 1kGP cell cultures and sequence data. 
 They used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
@@
 Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
+
+\begin{center}
+ \begin{tabular}{||c c c c||} 
+ \hline
+Author & Year & Journal & RSID \\ [0.5ex] 
+ \hline\hline
+Comuzzie AG &	2012	 &PLoS One &	rs9469457\\ 
+ \hline
+Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539\\ 
+ \hline
+Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648\\ 
+ \hline
+Zeng Z	& 2014 &	Caries Res	& rs10495332\\ 
+ \hline
+Park SL & 	2015	 & PLoS One	& rs113626797\\ 
+ \hline
+Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001\\ 
+ \hline
+Spada J & 	2016	 & J Sleep Res &	rs12452225\\ 
+ \hline
+Astle WJ	& 2016 &	Cell	 & rs533229216\\ 
+ \hline
+ Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834\\ 
+ \hline
+Liu TC	& 2017 &	JCI Insight &	rs12481514\\ 
+ \hline
+  McKay JD & 2017 &	Nature Genetics & 	rs13191445\\ 
+ \hline
+Zhou H &	2017	 & JAMA Psychiatry &	rs11179306\\ 
+ \hline
+Tian C	& 2017	 &Nature Communications	 & rs67491322\\ 
+ \hline
+Qian DC	& 2017 &	BMC Genomics &	rs77576211\\ 
+ \hline
+Michailidou K	& 2017	 & Nature &	rs76092474\\ 
+ \hline
+& & & rs80274284, rs201130852,\\ 
+Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,\\ 
+& & & rs201255786, rs201761909\\ 
+ \hline
+ \end{tabular}
+\end{center}
 

### 2018-06-19 Clean_Figure1.R
misc changes from desktop

@@ -13,13 +13,6 @@ colnames(JPT)<-paste("JPT", colnames(JPT), sep = "_")
 NAG<-read.table(fnag, fill=T,skip=1,  skipNul = TRUE,colClasses = classes)
 colnames(NAG)<-paste("NAG", colnames(NAG), sep = "_")
 
-#HWE 10^-6
-NAG_hwe<-read.table('/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered_sig6.freq.hwe')
-JPT_hwe <- read.table('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered_sig6.freq.hwe')
-#ggplot(JPT_hwe, aes(x=V2, y=abs(log10(V8))) )+geom_point(shape='.')+facet_grid(~V1)+coord_flip()+theme_classic()+theme(axis.text.x=element_blank())+xlab('Position (bp)')+ylab('-log10(p) HWE disequilibrium')
-#ggsave('~/Documents/HWE_disequilibrium_JPT.jpg', height=5,width=5)
-hwe=merge(NAG_hwe[,c('V1','V2')], JPT_hwe[,c('V1','V2')], by=c('V1','V2'), all=T)
-
 #merge them together
 join<-merge(JPT[which(JPT$JPT_V6 <= 1),],
  			NAG[which(NAG$NAG_V6 <= 1),], 
@@ -38,7 +31,7 @@ write.table(j,'~/Documents/MutSpect/WrapUpPlots/SFS_NAG_JPT_4bed.plot.txt')
 
 tab5rows <- read.table('/Users/luke/Documents/1FinalGWAS/1kGP_Chr2.4bed_filtered.assoc.linear', header = T, nrows = 5, row.names=NULL)
 classes <- sapply(tab5rows, class)
-GWAS_JPT<-read.table('/Users/luke/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.assoc.linear', header=F,colClasses = classes, col.names=names(tab5rows))
+GWAS_JPT<-read.table('~/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT_sig6.assoc.linear', header=F,colClasses = classes, col.names=names(tab5rows))
 GWAS_JPT$Plog10= -log10(GWAS_JPT$P)
 sig6<-GWAS_JPT[which(GWAS_JPT$Plog10 > 6),]
[…]

### 2018-06-19 legacy.tex

@@
 \begin{abstract}
-In investigating a population genetics signal, we noticed some discrepancies between the Japanese samples from the 1000 Genomes Project (1kGP) and a more recent high quality dataset.
-We found that the variants causing the signal were only present in individuals with low average sequencing quality.
-We identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ single nucleotide polymorphisms (SNPs) that were associated to quality in the 1kGP dataset Japanese individuals.
-These variants are present in nearly all of the populations from the 1kGP. 
+In investigating an unusual population genetics signal we identified a strong and previously unreported batch effect in the 1000 Genomes Project data.
+We also found that this batch effect led to spurious results in a number of recent publications. 
+We identified spurious mutations by correlating mutations with data quality metrics.
 We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
 Some of these variants are being imputed onto genotype data and reach genome wide significance in recent publications.
-\todo{update values for all populations}
 \end{abstract}
@@
 
-			\section{Results}
-	\subsection{A peculiar mutational signature in Japan}
-
-Harris et al reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
-While trying to follow up on this observations in a larger Japanese cohort individuals, we did not find this particular signature.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-If two groups of individuals are sampled from the same population, we expect there to be little deviation in allele frequencies between the two groups.
-Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
-
-When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}, suggesting a technical artifact rather than a population structure effect.
-Regressions against different quality metrics provided by the 1kGP revealed that mean quality per mapped base pair was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
-Individuals with low-quality data have elevated rates of the *AC${\rightarrow}$*CC signature. 
-\sgcomment{supplementary figure of correlations}
-Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
-
 \begin{figure}
@@
 
-To identify SNPs that are associated with low quality, we performed a genome-wide association study (GWAS)  \ref{SFS}. 
-We used the mean quality of mapped bases of each individual as the association phenotype.
-Thus, variants that are only found in individuals with low quality data will display significant associations to quality. 
-We were able to identify $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual \ref{SFS}.  
-This GWAS included 104 individuals, \todo{000} of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
+			\section{Results}
+	\subsection{A peculiar mutational signature in Japan}
+
+Harris and Pritchard reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
+While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
+This is unexpected given the similarity of the two populations, and suggests a technical artifact rather than a population structure effect. 
+Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
+
+When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}.
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
+Individuals with low-quality data showing elevated rates of the signature. \ref{PC1_Correlation}
+Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality \ref{SFS}.
+Using the linear regression as implemented in PLINK, we were able to identify $625$ snips with $p < 10^{-8}$ and $1048$ snps with $ p < 10^{-6}.$ 
+While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
+ This GWAS included 104 Japanese individuals, \todo{000} of which were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
 The variants that are associated to the quality of mapped bases have a 10 fold enrichment in *AC${\rightarrow}$*CC mutations.
 
-The GWAS has a power limit of 5\% minor allele frequency (MAF), many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+The signal is enriched in GWAS snps, but residual signal remains because our GWAS has a power limit of 5\% minor allele frequency (MAF) resulting in many rare alleles found in individuals with low-quality data remaining unidentifiable using association techniques.
 For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
@@
 Therefore, the removal of individuals with average qualities below 30 successfully removes the signal.
-
-By taking advantage of the larger, higher quality dataset, we were also able to get rid of this signal by filtering all the variants above 1\% in the 1kGP Japanese but missing from the \todo{NAG} data.
-This suggests that while the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
+While the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
 
@@
 	\subsection{Overlap of significant SNPs}
-
 We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
@@
 \begin{center}
- \begin{tabular}{||c c c c||} 
+ \begin{tabular}{||c c c c c||} 
  \hline
-Author & Year & Journal & RSID \\ [0.5ex] 
+Author & Year & Journal & RSID & PMID \\ [0.5ex] 
  \hline\hline
-Comuzzie AG &	2012	 &PLoS One &	rs9469457\\ 
+Comuzzie AG &	2012	 &PLoS One &	rs9469457&23251661\\ 
  \hline
-Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539\\ 
+Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539&23166209\\ 
  \hline
-Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648\\ 
+Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\ 
  \hline
-Zeng Z	& 2014 &	Caries Res	& rs10495332\\ 
+Zeng Z	& 2014 &	Caries Res	& rs10495332&24556642\\ 
  \hline
-Park SL & 	2015	 & PLoS One	& rs113626797\\ 
+Park SL & 	2015	 & PLoS One	& rs113626797&26053186\\ 
  \hline
-Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001\\ 
+Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001&26969751\\ 
  \hline
-Spada J & 	2016	 & J Sleep Res &	rs12452225\\ 
+Spada J & 	2016	 & J Sleep Res &	rs12452225&27126917\\ 
  \hline
-Astle WJ	& 2016 &	Cell	 & rs533229216\\ 
+Astle WJ	& 2016 &	Cell	 & rs533229216&27863252\\ 
  \hline
- Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834\\ 
+ Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834&27989323\\ 
  \hline
-Liu TC	& 2017 &	JCI Insight &	rs12481514\\ 
+Liu TC	& 2017 &	JCI Insight &	rs12481514&28352666\\ 
  \hline
-  McKay JD & 2017 &	Nature Genetics & 	rs13191445\\ 
+  McKay JD & 2017 &	Nature Genetics & 	rs13191445&28604730\\ 
  \hline
-Zhou H &	2017	 & JAMA Psychiatry &	rs11179306\\ 
+Zhou H &	2017	 & JAMA Psychiatry &	rs11179306&	29071344\\ 
  \hline
-Tian C	& 2017	 &Nature Communications	 & rs67491322\\ 
+Tian C	& 2017	 &Nature Communications	 & rs67491322&28928442\\
  \hline
-Qian DC	& 2017 &	BMC Genomics &	rs77576211\\ 
+Qian DC	& 2017 &	BMC Genomics &	rs77576211&28927378\\ 
  \hline
-Michailidou K	& 2017	 & Nature &	rs76092474\\ 
+Michailidou K	& 2017	 & Nature &	rs76092474&29059683\\ 
  \hline
-& & & rs80274284, rs201130852,\\ 
-Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,\\ 
-& & & rs201255786, rs201761909\\ 
+& & & rs80274284, rs201130852,&\\
+Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
+& & & rs201255786, rs201761909&\\
  \hline
@@
 This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
-The dates of the sequencing are only used to plot Figure. \todo{Average Quality of mapped bases: How was it calculated?}
+The dates of the sequencing are only used to plot Figure.
 
@@
 We used the average quality of mapped bases per individual as the phenotype for the analysis.
-We also controlled for population structure by including the first 4 principle components of a PCA of each population using genotype data.
-\todo{NOT DONE YET}
+We controlled for population structure by including the first 4 principle components of a PCA of each population with available genotype data, however, considering that population structure did not affect the number of variants associated to quality, and that not all populations had genotype data readily available, we did not include these in the final analysis.
 
@@
 
-\bibliography{/Users/luke/Dropbox/Work/PHD/BibTeX/Legacy.bib}
+\subsection{Supplementary Figures}
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{PC1_Correlation.jpg}
+\caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+}
+ \label{PC1_Correlation}
+\end{figure}
+
+\bibliography{Legacy.bib}
 

### 2018-06-19 legacy.tex
small tweaks, fixed biblio

@@
 \author[1,2]{Mathieu Bourgey}
-\author[3]{Fumihiko Matsuda}
+\author[3]{Fumihiko Matsuda (and coauthors)}
 \author[1,2]{Simon Gravel}
@@
 Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
-Individuals with low-quality data showing elevated rates of the signature. \ref{PC1_Correlation}
+Individuals with low-quality data showing elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}). \sgcomment{fix ref to figs}
 Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 \begin{center}
- \begin{tabular}{||c c c c c||} 
+ \begin{tabular}{l l l r r} 
  \hline
 Author & Year & Journal & RSID & PMID \\ [0.5ex] 
- \hline\hline
-Comuzzie AG &	2012	 &PLoS One &	rs9469457&23251661\\ 
  \hline
+Comuzzie AG \emph{et al}, PLoS One (2012)  &	2012	 &PLoS One &	rs9469457&23251661\\
+
 Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539&23166209\\
- \hline
+
 Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
- \hline
+
 Zeng Z	& 2014 &	Caries Res	& rs10495332&24556642\\
- \hline
+
 Park SL & 	2015	 & PLoS One	& rs113626797&26053186\\
- \hline
+
 Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001&26969751\\
- \hline
+
 Spada J & 	2016	 & J Sleep Res &	rs12452225&27126917\\ 
- \hline
 Astle WJ	& 2016 &	Cell	 & rs533229216&27863252\\
- \hline
+
  Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834&27989323\\
- \hline
+
 Liu TC	& 2017 &	JCI Insight &	rs12481514&28352666\\
- \hline
+
   McKay JD & 2017 &	Nature Genetics & 	rs13191445&28604730\\
- \hline
+
 Zhou H &	2017	 & JAMA Psychiatry &	rs11179306&	29071344\\
- \hline
+
 Tian C	& 2017	 &Nature Communications	 & rs67491322&28928442\\
- \hline
+
 Qian DC	& 2017 &	BMC Genomics &	rs77576211&28927378\\
- \hline
+
 Michailidou K	& 2017	 & Nature &	rs76092474&29059683\\
- \hline
+
 & & & rs80274284, rs201130852,&\\
@@
 
+\bibliographystyle{plain}
 \bibliography{Legacy.bib}

### 2018-07-31 legacy.tex
Finishing up genome wide run

@@
 To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality \ref{SFS}.
-Using the linear regression as implemented in PLINK, we were able to identify $625$ snips with $p < 10^{-8}$ and $1048$ snps with $ p < 10^{-6}.$ 
+\luke{NEW:} 
+Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, our method does the inverse.
+We are looking for loci where the genotypes are predictable based on the data quality scores.
+We were able to identify \todo{000} SNPs with $p < 10^{-8}$ and \todo{000} SNPs with $ p < 10^{-6}$.
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
- This GWAS included 104 Japanese individuals, \todo{000} of which were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1kGP.
-The variants that are associated to the quality of mapped bases have a 10 fold enrichment in *AC${\rightarrow}$*CC mutations.
+ %This GWAS included 104 Japanese individuals, 74 of which were sequenced in using the Genome Analyzer II while the rest were sequenced using the HiSeq 2000 of the 1kGP.
+The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations.
 

### 2018-07-31 Clean_Figure1.R

@@ -90,7 +90,7 @@ scale_fill_distiller(palette = "Spectral",trans = "log", breaks=c(1,10,100,1000,
 scale_x_continuous(expand=c(0.01,0.01))+
 scale_y_continuous(expand=c(0.01,0.01))+
 theme_classic()+theme(axis.line=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())+
-xlab(label="NAG")
+xlab(label="Nagahama")
 
 HIST=ggplot(jj, aes(x= JPT_V6))+geom_histogram(binwidth=0.01, fill='grey')+theme_classic()+scale_x_continuous(limits=c(0,1),expand=c(0.01,0.01))+scale_y_reverse(expand=c(0.01,0.01), breaks=c(0,50,100))+geom_vline(xintercept=min(jj$JPT_V6), linetype=3)+coord_flip()+xlab('')+ylab('')+theme(axis.line=element_blank(),axis.text.x=element_text(angle=90))+xlab(label="1000 Genomes Project")
 B=plot_grid(HIST,SFS,nrow=1,rel_widths=c(1,5), align = 'h', labels=c('','B'))
@@ -141,20 +141,26 @@ nrow=2,rel_heights=c(1,0.1),labels=c('A'))
 AB=plot_grid(A,B, rel_widths=c(1,2.5))
 #ggsave('~/Documents/QualityPaper/NAG_JPT_SFS_GenomeWide_frq.jpg',p3, height=5,width=7)
 
-fname='~/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT_noNA_Dup.assoc.linear'
-tab5rows <- read.table(fname, header = TRUE, nrows = 5)
-classes <- sapply(tab5rows, class)
-assoc<-read.table(fname, header=TRUE, colClasses= classes)
-assoc$Plog10=-log10(assoc$P)
-sig.6<-assoc[which(assoc$Plog10 > 6),]
-NotSig<-assoc[which(assoc$Plog10 < 6),]
+fileNames = list.files(path='/Volumes/gravel/luke_projects/1000Genomes/Regression/', pattern="*.Regression_JPT.csv", full.names = T)
+Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
 
[…]

### 2018-07-31 ComparePvalPerPop (Autosaved).R

@@ -1,32 +1,0 @@
-##All Pops Logistic
-library(ggplot2)
-library(cowplot)
-
-all<-read.table('~/Downloads/CHR22.Regression_ALL.csv', header=T)
-all$Chr<-as.numeric(as.character(all$Chr))
-all$Pos<-as.numeric(as.character(all$Pos))
-all$p<-as.numeric(as.character(all$p))
-
-all<-all[complete.cases(all),]
-
-ggplot(all, aes(x=Pos, y=p, color=Pop))+geom_point()+facet_wrap(~Pop, ncol=2)
-
-JPT<-all[which(all$Pop=='JPT'),]
-plots<-list()
-for(pop in unique(all$Pop)){
-sub<-all[which(all$Pop==pop),]
-m<-merge(sub,JPT, by='Pos')
-plots[[pop]]<-ggplot(m, aes(x=p.x, y=p.y))+geom_point()+labs(x=pop,y='JPT')
-}
-
-
-plot_grid(plots[['ACB']], plots[['ASW']], plots[['BEB']], plots[['CDX']], plots[['CEU']], plots[['CHB']], plots[['CHS']], plots[['CLM']], plots[['ESN']], plots[['FIN']], plots[['GBR']], plots[['GIH']], plots[['GWD']], plots[['IBS']], plots[['ITU']], plots[['KHV']], plots[['LWK']], plots[['MSL']], plots[['MXL']], plots[['PEL']], plots[['PJL']], plots[['PUR']], plots[['STU']], plots[['TSI']], plots[['YRI']])
-
-ggsave('~/Desktop/PvalComparison.jpg',height=10,width=10)
-
-Meta<-all[,c('Pos','p')]
-Meta$invP<- -10^(Meta$p)
-Meta<-aggregate(.~Pos, data=Meta, prod)
-Meta$invP<--log10(Meta$invP)
-ggplot(Meta, aes(x=invP,y=p))+geom_point(shape=1)
-

### 2018-07-31 ComparePvalPerPop.R

@@ -2,7 +2,7 @@
 library(ggplot2)
 library(cowplot)
 
-all<-read.table('~/Downloads/CHR22.Regression_ALL.csv', header=T)
+all<-read.table('~/Documents/Regression/CHR22.Regression_ALL.csv', header=T)
 all$Chr<-as.numeric(as.character(all$Chr))
 all$Pos<-as.numeric(as.character(all$Pos))
 all$p<-as.numeric(as.character(all$p))

### 2018-07-31 Deviations.R

@@ -1,0 +1,63 @@
+##Deviations
+library(dplyr)
+library(data.table)
+library(ggplot2)
+
+Reg<-fread('/Users/luke/Documents/Regression/CHR19-22.Regression.csv')
+Reg<-Reg[which(Reg$Pos != 'Pos'),]
+Reg<-Reg[which(Reg$Chr != '14'),]
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$p))
+Reg$Pop=NULL
+Reg$p=NULL
+
+Meta <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
+
+Meta$plog10 <- - pchisq(Meta$SumDev, Meta$Count, lower.tail=F, log.p=T)/log(10)
+
+ggplot(Meta, aes(x=Pos, y= plog10, color=Chr))+geom_point(shape=1)+facet_grid(~Chr, scales='free_x', space= 'free_x')+theme_classic()+guides(color=F)+geom_hline(yintercept=8, color='blue')
+ggsave('/Users/luke/Documents/Regression/Manhattan_Chr19-22.jpg', height=8, width=11)
+
+ggplot(Meta[which(Meta$Count > 1),], aes(x=Pos, y= MeanDev, color=Chr))+geom_point(shape=1)+facet_grid(Count~Chr, scales='free_x', space= 'free_x')+theme_classic()+guides(color=F)
+ggplot(Meta[which((Meta$Count > 1)&(Meta$MeanDev > 10)),], aes(x=Pos, y= MeanDev, color=Chr))+geom_point(shape=1)+facet_grid(Count~Chr, scales='free_x', space= 'free_x')+theme_classic()+guides(color=F)
+
+Meta[which((Meta$Count > 2)&(Meta$plog10 > 20)),]
+
+
+
+
+gwas<-fread('~/Documents/GWAS_Qual/Meta/INT_19.22.meta')
+#gwas<-fread('/Users/luke/Documents/GWAS_Qual/Meta/NoNormal_19.22.meta')
+names(gwas)<-c('Chr','Pos','SNP','A1','A2','N','P','P(R)','BETA','BETA(R)','Q','I')
+
+gwas$log10P<--log10(gwas$P)
[…]

### 2018-07-31 JustStats.R

@@ -2,21 +2,23 @@
 library(data.table)
 args = commandArgs(trailingOnly=T)
 
-#read input file #/Users/luke/Documents/Regression/CHR22.Format_JPT.csv
-Input <- fread(args[1],sep=',')
-
+#read input file
+#Input <- fread('/Users/luke/Documents/Regression/CHR22.Format_JPT.csv',sep=',')
+Input <- fread(paste("gunzip -c ",args[1]),sep=',', header=T)
+Pop <-args[3]
 Positions<-unique(Input $Pos)
 
 chunks=split(Positions, ceiling(seq_along(Positions)/10000))
 index <- 1
 for(chunk in chunks){
-print(paste('Progress :',index/length(chunks)*100,' %'))
+print(paste(Pop,' Progress :',round(index/length(chunks)*100, digits=2),' %'))
 subChunk<-Input[which(Input $Pos %in% chunk),]	
 stat <- with(subChunk, by(subChunk,Pos, function(x) 
 		glm(value ~ average_quality_of_mapped_bases, data=x, family=binomial)))
-#extract P value
+#extract deviance
 less = unique(subChunk[,c('Chr','Pos','Pop')])
-less$p = -log10(as.numeric(as.character(sapply(stat,function(x) summary(x)$coef[2,4]))))
+less$dev = as.numeric(as.character(sapply(stat,function(x) anova(x, test='Chi')[2,2])))
+
 #write results to file
 if(index == 1){
 	write.table(less,file=args[2], quote=F, row.names=F, append=FALSE)}
@@ -26,4 +28,4 @@ index <- index + 1
 }
 
 ##code dump
[…]

### 2018-07-31 ManhattanPlot (Autosaved).R

@@ -1,0 +1,53 @@
+##Manhattan
+library(dplyr)
+library(magrittr)
+library(data.table)
+
+Tot=data.table()
+for(chrom in seq(1,22)){
+	print(chrom)
+	fileNames = list.files(
+	path='/Volumes/gravel/luke_projects/1000Genomes/Regression/',
+	pattern=paste("CHR",chrom,".Regression",sep=''), full.names = T)
+	
+	Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
+	
+	print(head(Reg))
+	
+	Reg<-Reg[which(Reg$Pos != 'Pos'),]
+	Reg$Chr<-as.numeric(as.character(Reg$Chr))
+	Reg$Pos<-as.numeric(as.character(Reg$Pos))
+	Reg$dev<-as.numeric(as.character(Reg$dev))
+	Reg$Pop=NULL
+	Reg$p=NULL
+
+	Reg <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
+	
+	write.table(paste('/Volumes/gravel/luke_projects/1000Genomes/MeanDev/Chromosome',chrom,'_MeanDev.csv',sep=''))
+	Tot <- rbind(Tot,Reg)
+}
+
+
+
+
+Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
+sig.6<-Reg[which(Reg$plog10 > 6),]
+NotSig<-Reg[which(Reg$plog10 < 6),]
+
+ggplot(NotSig, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
+facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+
+scale_fill_manual (values=getPalette(colourCount))+
+scale_y_continuous(expand=c(0,0))+
+geom_point(alpha=0.3, size=1)+theme_classic()+
+geom_point(data=sig.6, aes(x=Pos, y = plog10,),color='black',shape=3)+
+geom_hline(yintercept = 8, color='red')+
+geom_hline(yintercept = 6, color='blue')+
+labs(y='-log10(p)', x='Chromosome')+
+theme(plot.title = element_text(hjust = 0.5), 
+plot.subtitle = element_text(hjust = 0.5), 
+axis.text.x=element_blank(), 
+axis.ticks.x=element_blank(),
+axis.line.x=element_blank(), 
[…]

### 2018-07-31 Parallel_Regression.py

@@ -1,0 +1,24 @@
+import multiprocessing as mp
+import sys, os
+import time
+
+chrom = sys.argv[1]
+global chrom
+
+def Bash_cmd(pop):
+    print('Working on Pop : '+str(pop))
+    #put your bash command below
+    input='~/Documents/Regression/CHR{0}.Format_{1}.csv.gz'.format(chrom,pop)
+    output='~/Documents/Regression/CHR{0}.Regression_{1}.csv'.format(chrom,pop)
+
+    print(input)
+    os.system('Rscript ~/Documents/QualityPaper/JustStats.R {0} {1} {2}'.format(input, output, pop))
+
+if __name__ == '__main__':
+    with open('/Users/luke/genomes/genomes/PopNames/NamePop.txt') as f:
+        Pops = f.read().splitlines()
+    pool_size=6 #mp.cpu_count()
+    pool=mp.Pool(processes=pool_size)
+    pool_outputs= pool.map(Bash_cmd, Pops)
+    pool.close()
+    pool.join()

### 2018-08-08 Clean_Figure1.R
Updated Figure 1 with regression values

@@ -1,46 +1,72 @@
 library(ggplot2)
 library(colorspace)
 library(cowplot)
-f1kGP='~/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT.frq' #/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered.freq.frq
-fnag='/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'
+library(data.table)
 
-tab5rows <- read.table(fnag, header = TRUE, nrows = 5, row.names=NULL)
-classes <- sapply(tab5rows, class)
-#Allele Frequency from JPT
-JPT<-read.table(f1kGP, fill=T,skip=1,  skipNul = TRUE,colClasses = classes)
-colnames(JPT)<-paste("JPT", colnames(JPT), sep = "_")
-#Allele Frequency from NAG
-NAG<-read.table(fnag, fill=T,skip=1,  skipNul = TRUE,colClasses = classes)
-colnames(NAG)<-paste("NAG", colnames(NAG), sep = "_")
+if(F){
+#f1kGP='~/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered.freq.frq'
+f1kGP='/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_JPT_1.frq'
+fnag='~/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'
+hweFile='/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT_sig6.hwe'
+
+#Allele Frequency files
+JPT<-fread(f1kGP, fill=T, col.names=c('CHROM','POS','N_ALLELES','N_CHR','JPT_AF','JPT_MAF'), colClasses=c('numeric'))
+NAG<-fread(fnag, fill=T,col.names=c('CHROM','POS','N_ALLELES','N_CHR','NAG_AF','NAG_MAF'), colClasses=c('numeric'))
 
 #merge them together
-join<-merge(JPT[which(JPT$JPT_V6 <= 1),],
- 			NAG[which(NAG$NAG_V6 <= 1),], 
- 			by.x=c("JPT_V1","JPT_V2"), 
- 			by.y=c("NAG_V1","NAG_V2"), all=T)
[…]

### 2018-08-08 Clean_Figure3.R

@@ -1,6 +1,7 @@
 #load libraries
 library(ggplot2)
 library(cowplot)
+library(data.table)
 #pop names, used to read/write files
 pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
 #read colors per pop (used in 1kG Nature)
@@ -9,31 +10,38 @@ MyColour <- as.character(colors $hex)
 names(MyColour) <- colors $Pop
 
 #read the list of sig snps
-#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('CHR', 'BP'))
-
+#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('Chr', 'Pos'))
+dir='/Volumes/gravel-1/luke_projects/1000Genomes/Regression/'
 SFS=data.frame()
 for(f in seq(1,nrow(pops))){
 pop=pops[f,]
 print(as.character(pop))
-fname=paste('/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_',pop,'_INT_sig6.assoc.linear',sep='')
-if(file.info(fname)$size != 0){
-freq=read.table(fname, header=F,, col.names=c('CHR','SNP','BP','A1','TEST','NMISS','BETA','STAT','P'))
-freq$Plog10<--log10(freq$P)
-sub=freq[,c('CHR','BP','Plog10')]
-sub$Pop<-pop
-SFS=rbind(SFS,sub)
-}}
+fileNames = list.files(path=dir, pattern=paste(pop), full.names = T)
+Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
+print('Read')
+Reg<-Reg[which(Reg$Pos != 'Pos'),]
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$dev))
+Reg$p=NULL
+print('Clean')
+Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
+sig.6<-Reg[which(Reg$plog10 > 6),]
+rm(Reg)
+write.table(sig.6[,c('Chr','Pos','plog10', 'Pop')], paste('~/Documents/Regression/',pop,'_sig6.csv', sep=','))
[…]

### 2018-08-08 Figure1Rejected.R

@@ -1,0 +1,44 @@
+##Figure 1 Rejected
+
+#f1kGP='~/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_Frequency.txt'#'/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_JPT.frq'
+#'/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'#~/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.frq'
+
+#col.names=c('rsID','CHROM', 'A1','A2','JPT_AF','NCHR','Chr','POS'))
+#Allele Frequency from NAG
+NAG<-fread(fnag, fill=T,col.names=c('CHROM','POS','N_Allele', 'N_Chr','NAG_AF','NAG_MAF'))
+NAG $NAG_AF<-as.numeric(as.character(NAG $NAG_AF))
+NAG $NAG_MAF<-as.numeric(as.character(NAG $NAG_MAF))
+NAG $CHROM<-as.numeric(as.character(NAG $CHROM))
+NAG $POS<-as.numeric(as.character(NAG $POS))
+NAG<-NAG[complete.cases(NAG),]
+
+
+if(F){
+##PYTHON
+JPT=pd.read_table('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_JPT_less.frq',
+                  skiprows=1, names=['CHROM','POS','AF_JPT'])
+JPT['AF_JPT']=JPT['AF_JPT'].str[2:]
+NAG=pd.read_table('/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered_less.frq',
+                  skiprows=1, names=['CHROM','POS','AF_NAG'])
+NAG['AF_NAG']=NAG['AF_NAG'].str[2:]
+join=pd.concat([JPT, NAG])
+join.to_csv('~/Documents/QualityPaper/joinedAF.csv')
+}
+
+
+join<-fread('~/Documents/QualityPaper/joinedAF.csv', colClasses=c('numeric','numeric','numeric','numeric','numeric'))
+join$V1<-NULL
+join$AF_JPT<-as.numeric(as.character(join$AF_JPT))
+join$AF_NAG<-as.numeric(as.character(join$AF_NAG))
[…]

### 2018-08-08 JPTManhattan.R

@@ -1,0 +1,39 @@
+##JPT Manhattan
+
+dir='/Volumes/gravel-1/luke_projects/1000Genomes/Regression/'
+out='/Volumes/gravel-1/luke_projects/1000Genomes/MeanDev/'
+
+
+jptNames = list.files(path=dir,pattern='JPT', full.names = T)
+jReg = do.call(rbind, lapply(jptNames, function(x) read.table(x, header=T, sep=' ')))
+jReg<-jReg[which(jReg$Pos != 'Pos'),]
+jReg$Chr<-as.numeric(as.character(jReg$Chr))
+jReg$Pos<-as.numeric(as.character(jReg$Pos))
+jReg$dev<-as.numeric(as.character(jReg$dev))
+
+jReg$plog10 <- - pchisq(jReg$dev, 1, lower.tail=F, log.p=T)/log(10)
+
+jReg$p <- pchisq(jReg$dev, 1, lower.tail=F)
+
+jsig.6<-jReg[which(jReg$plog10 >= 6),]
+jNotSig<-jReg[which(jReg$plog10 < 6),]
+
+ggplot(jNotSig, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
+facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+
+scale_fill_manual (values=getPalette(colourCount))+
+scale_y_continuous(expand=c(0,0))+
+geom_point(alpha=0.3, size=1)+theme_classic()+
+geom_point(data=jsig.6, aes(x=Pos, y = plog10,),color='black',shape=3)+
+geom_hline(yintercept = 8, color='red')+
+geom_hline(yintercept = 6, color='blue')+
+labs(y='-log10(p)', x='Chromosome')+
+theme(plot.title = element_text(hjust = 0.5), 
+plot.subtitle = element_text(hjust = 0.5), 
+axis.text.x=element_blank(), 
+axis.ticks.x=element_blank(),
+axis.line.x=element_blank(), 
+strip.background = element_blank(),
+strip.text.x = element_text(size = 6))+
+guides(color=F)
+
+ggsave('~/Documents/Regression/JPT_Manhattan.jpg', height=7, width=12)

### 2018-08-08 Parallel_Regression.py

@@ -3,15 +3,14 @@ import sys, os
 import time
 
 chrom = sys.argv[1]
-global chrom
-
+path = sys.argv[2]
 def Bash_cmd(pop):
     print('Working on Pop : '+str(pop))
     #put your bash command below
-    input='~/Documents/Regression/CHR{0}.Format_{1}.csv.gz'.format(chrom,pop)
-    output='~/Documents/Regression/CHR{0}.Regression_{1}.csv'.format(chrom,pop)
+    input='{2}/Format/CHR{0}.Format_{1}.csv.gz'.format(chrom, pop, path)
+    output='{2}/Regression/CHR{0}.Regression_{1}.csv'.format(chrom, pop, path)
 
-    print(input)
+    print('Rscript ~/Documents/QualityPaper/JustStats.R {0} {1} {2}'.format(input, output, pop))
     os.system('Rscript ~/Documents/QualityPaper/JustStats.R {0} {1} {2}'.format(input, output, pop))
 
 if __name__ == '__main__':

### 2018-08-08 legacy.tex
Updated Figure 1 with regression values
Updated Figure 3 (minor edits to F1)

@@
 \author[1,2]{Mathieu Bourgey}
+\author[1,2]{Rick Farouni}
 \author[3]{Fumihiko Matsuda (and coauthors)}
@@
 We are looking for loci where the genotypes are predictable based on the data quality scores.
-We were able to identify \todo{000} SNPs with $p < 10^{-8}$ and \todo{000} SNPs with $ p < 10^{-6}$.
+We were able to identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
@@
 In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted the genotype data from the 1kGP and imputed it using the Michigan Imputation Server.
-We found that \todo{~30\%} of the SNPs we identified as being associated with quality were imputed.
+We found that all of the SNPs we identified as being associated with quality were imputed.
 

### 2018-08-08 Clean_Figure1.R

@@ -122,7 +122,7 @@ scale_y_continuous(expand=c(0.01,0.01))+
 theme_classic()+theme(axis.line=element_blank(), axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank())+
 xlab(label="Nagahama")
 
-HIST=ggplot(jj, aes(x= JPT_MAF))+geom_histogram(binwidth=0.01, fill='grey')+theme_classic()+scale_x_continuous(limits=c(0,1),expand=c(0.01,0.01))+scale_y_reverse(expand=c(0.01,0.01), breaks=c(0,50,100))+geom_vline(xintercept=min(jj$JPT_MAF), linetype=3)+coord_flip()+xlab('')+ylab('')+theme(axis.line=element_blank(),axis.text.x=element_text(angle=90))+xlab(label="1000 Genomes Project")
+HIST=ggplot(jj, aes(x= JPT_MAF))+geom_histogram(binwidth=0.01, fill='grey')+theme_classic()+scale_x_continuous(limits=c(0,1),expand=c(0.01,0.01))+scale_y_reverse(expand=c(0.01,0.01), breaks=c(10,50,100))+geom_vline(xintercept=0.05, linetype=3)+coord_flip()+xlab('')+ylab('')+theme(axis.line=element_blank(),axis.text.x=element_text(angle=90))+xlab(label="1000 Genomes Project")
 B=plot_grid(HIST,SFS,nrow=1,rel_widths=c(1,5), align = 'h', labels=c('','B'))
 
 
@@ -168,11 +168,8 @@ rel_widths=c(1,0.77)),
 mylegend,
 nrow=2,rel_heights=c(1,0.1),labels=c('A'))
 
-AB=plot_grid(A,B, rel_widths=c(1,2.5))
-#ggsave('~/Documents/QualityPaper/NAG_JPT_SFS_GenomeWide_frq.jpg',p3, height=5,width=7)
+AB=plot_grid(A,B, rel_widths=c(1,2.7))
 
-fileNames = list.files(path='/Volumes/gravel/luke_projects/1000Genomes/Regression/', pattern="*.Regression_JPT.csv", full.names = T)
-Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
 Reg<-fread('/Volumes/gravel/luke_projects/1000Genomes/Regression/GenomeWide.Regression_JPT.csv')
 
 Reg<-Reg[which(Reg$Pos != 'Pos'),]
[…]

### 2018-08-08 Clean_Figure3.R

@@ -11,41 +11,31 @@ names(MyColour) <- colors $Pop
 
 #read the list of sig snps
 #bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('Chr', 'Pos'))
-dir='/Volumes/gravel-1/luke_projects/1000Genomes/Regression/'
-SFS=data.frame()
-for(f in seq(1,nrow(pops))){
-pop=pops[f,]
-print(as.character(pop))
-fileNames = list.files(path=dir, pattern=paste(pop), full.names = T)
+dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
+fileNames = list.files(path=dir, pattern='sig.csv', full.names = T)
 Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
-print('Read')
 Reg<-Reg[which(Reg$Pos != 'Pos'),]
 Reg$Chr<-as.numeric(as.character(Reg$Chr))
 Reg$Pos<-as.numeric(as.character(Reg$Pos))
 Reg$dev<-as.numeric(as.character(Reg$dev))
-Reg$p=NULL
-print('Clean')
 Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
-sig.6<-Reg[which(Reg$plog10 > 6),]
-rm(Reg)
-write.table(sig.6[,c('Chr','Pos','plog10', 'Pop')], paste('~/Documents/Regression/',pop,'_sig6.csv', sep=','))
-#SFS=rbind(SFS,sig.6[,c('Chr','Pos','plog10', 'Pop')])
-}
+Reg<-Reg[which(Reg$plog10 >= 6),]
 
-
-SFS<-unique(SFS)
-NumPop<-as.data.frame(table(SFS$Chr,SFS$Pos))
+NumPop<-as.data.frame(table(Reg$Chr,Reg$Pos))
 UnPop<-NumPop[which(NumPop$Freq==1),]
 NumPop<-NumPop[which(NumPop$Freq>1),]
-m=merge(SFS,NumPop, by.y=c('Var1','Var2'), by.x=c('Chr','Pos'))
+NumPop$Var1<-as.numeric(as.character(NumPop$Var1))
+NumPop$Var2<-as.numeric(as.character(NumPop$Var2))
+
[…]

### 2018-08-08 Figure1Rejected.R

@@ -42,3 +42,28 @@ j<-join[which((join$sum > 0.01) & (join$sum < 2)),]
 j<-j[,c('JPT_CHROM', 'JPT_POS','JPT_MAF', 'NAG_MAF', 'sum')]
 write.table(j,'~/Documents/MutSpect/WrapUpPlots/SFS_NAG_JPT_4bed.plot.txt')
 #j<-read.table('~/Documents/MutSpect/WrapUpPlots/SFS_NAG_JPT_4bed.plot.txt')
+
+
+
+##Figure 3
+SFS=data.frame()
+for(f in seq(1,nrow(pops))){
+pop=pops[f,]
+print(as.character(pop))
+fileNames = list.files(path=dir, pattern=paste(pop), full.names = T)
+Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
+print('Read')
+Reg<-Reg[which(Reg$Pos != 'Pos'),]
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$dev))
+Reg$p=NULL
+print('Clean')
+Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
+sig.6<-Reg[which(Reg$plog10 > 6),]
+rm(Reg)
+write.table(sig.6[,c('Chr','Pos','plog10', 'Pop')], paste('~/Documents/Regression/',pop,'_sig6.csv', sep=','))
+#SFS=rbind(SFS,sig.6[,c('Chr','Pos','plog10', 'Pop')])
+}
+
+

### 2018-08-09 legacy.tex
updated text
Methods, results, removed unnecessary text in discussion.

@@
 \textbf{A} 
-Mutation spectrum of the 1048 variants that reached a genome wide significance with a p value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
-There is a 10 fold enrichment in *AC${\rightarrow}$*CC mutations. 
+Mutation spectrum of the 1034 variants that reached a genome wide significance with a p value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
+There majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern.
+There is also an slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as G**${\rightarrow}$GG*.
+(note: the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*)
 \textbf{B} 
-Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent dataset.
-Crosses ( + ) are variants that reached genome wide significance in a GWA of sequencing quality. 
+Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
+Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
 The histogram on the left of the plot is the distribution of significant variants. 
-The lowest frequency variant able to reach genome wide significance is 5\%.
+The dotted line is at 5\% allele frequency. 
 \textbf{C} 
 Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1kGP.
-This GWAS identified $625\ \  p < 10^{-8}$ and $1048\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
 The same analysis was performed independently for each of the populations in the 1kGP. }
@@
 
-Harris and Pritchard reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.
+Harris and Pritchard reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
-Individuals with low-quality data showing elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}). \sgcomment{fix ref to figs}
+Individuals with low-quality data showing elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
 Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 
-The signal is enriched in GWAS snps, but residual signal remains because our GWAS has a power limit of 5\% minor allele frequency (MAF) resulting in many rare alleles found in individuals with low-quality data remaining unidentifiable using association techniques.
+The signal is enriched in GWAS SNPs, but residual signal remains presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
 For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
@@
 We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
-We found that 25 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
+We found that 24 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
 \todo{Supplemental Figures : all the GWAS results for each population}
-Comparing the results of each independent GWAS, we were able to identify over 3625 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
+Comparing the results of each independent GWAS, we were able to identify over 3493 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
 
@@
 \caption{Overlap of SNPs identified independently to be associated with quality. 
-The populations that have the most low-quality individuals also have the most low-quality variants. 
+The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
+The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+The line plot underneath shows the number of populations for which a variant has reached significance.
+The populations that tend to have the most low-quality individuals also tend to have the most variants associated to quality. 
 The same variants identified as being low quality independently in each population are found in other populations. }
@@
 For this reason, the details of the methods used to impute genomic data are not publicly available either.
-In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted the genotype data from the 1kGP and imputed it using the Michigan Imputation Server.
+In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from the 1kGP and imputed it using the Michigan Imputation Server.
 We found that all of the SNPs we identified as being associated with quality were imputed.
+This suggests that the imputation reference panel includes individuals with low sequence quality, as well as the variants that have been shown to be associated to quality.
 
@@
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have called these erroneous variants as being significantly correlated with some biological trait.
-Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found 17 recent publications that had found at least one of the variants to have reached genome wide significance in their study.
+Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found \todo{000} recent publications that had found at least one of the variants to have reached genome wide significance in their study.
 
@@
  \hline
-Comuzzie AG \emph{et al}, PLoS One (2012)  &	2012	 &PLoS One &	rs9469457&23251661\\
 
-Smith JG &	2012	 & Circ Cardiovasc Genet	& rs7729539&23166209\\
-
-Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
-
-Zeng Z	& 2014 &	Caries Res	& rs10495332&24556642\\
-
-Park SL & 	2015	 & PLoS One	& rs113626797&26053186\\
-
-Parmar PG &	2016	 & Circ Cardiovasc Genet	& rs12370001&26969751\\
-
-Spada J & 	2016	 & J Sleep Res &	rs12452225&27126917\\ 
-Astle WJ	& 2016 &	Cell	 & rs533229216&27863252\\
-
- Ahola-Olli AV &	2017 &	Am J Hum Genet &	rs76842834&27989323\\
-
-Liu TC	& 2017 &	JCI Insight &	rs12481514&28352666\\
-
-  McKay JD & 2017 &	Nature Genetics & 	rs13191445&28604730\\
-
-Zhou H &	2017	 & JAMA Psychiatry &	rs11179306&	29071344\\
-
-Tian C	& 2017	 &Nature Communications	 & rs67491322&28928442\\
-
-Qian DC	& 2017 &	BMC Genomics &	rs77576211&28927378\\
-
-Michailidou K	& 2017	 & Nature &	rs76092474&29059683\\
-
-& & & rs80274284, rs201130852,&\\
-Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
-& & & rs201255786, rs201761909&\\
  \hline
@@
 
-%\subsection{Technical Source of Error}
+\subsection{Technical Source of Error}
+%B tail trimming  (Minoche et al. Genome Biology 2011)
 %issues with PCR?
@@
 
-\subsection{Imputation}
-Imputation of genotype data is a probabilistic method that infers the bases of a given genome based on its similarity to a set of reference genomes.
-While on average, two humans differ in about 1/10,000 bases, this number is \todo{lower} in closely related individuals, and \todo{higher} in individuals from different continental origins.
-Modern chip sequencing will provide the genotype information for over 1 million bases of the genome.
-The unique combination of genotyped bases can be enough to identify haplotype blocks that are identical by descent in individuals form the reference database.
-
-The accuracy of imputation depends on the size of the reference database; this varies significantly from one population to another.
-This means that individuals with ancestry that is less well represented in the reference database will have lower accuracy of imputation.
-To overcome this bias, reference databases are often combined to increase the sample size and in turn, the accuracy of imputation.
-
-\subsection{GWA studies from other papers}
-Since these variants are present in more than one population from the 1kGP, they are more likely to be associated to biological traits as they would appear to be like any other variant that is shared among multiple populations.
-The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
-
-%\luke{due to the temporal nature of the batch effects, and because entire populations were sequenced in one centre on one day, the false positives are more likely to cluster with population structure or case/control?}
+\subsection{GWAS from other papers}
+%The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
 
@@
 
-\subsection{Genome Wide Association}
-
-Using PLINK v1.90b4.4 we ran a linear genome wide association study independently for each population of the 1kGP.
-We used the average quality of mapped bases per individual as the phenotype for the analysis.
-We controlled for population structure by including the first 4 principle components of a PCA of each population with available genotype data, however, considering that population structure did not affect the number of variants associated to quality, and that not all populations had genotype data readily available, we did not include these in the final analysis.
+\subsection{Genome Wide Association Study}
+Using an in-house R script, we performed a logistic regression using the glm package.
+\cite{[R and GLM]} We performed this statistical test independently for each population of the 1kGP.
+The deviations from the null model for each test was used to compute the chi-squared p value for each SNP for each population.
+Since each deviation follows a chi-squared distribution, the sum of the deviations also follow a chi-squared distribution.
+However, since not all SNPs are present in all populations, when computing the p values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.
 
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum for each list of significant SNPs for each population.
-
-We also compared the mutation spectrum ratio between populations using a modified version of the methods used in Harris et al. 2017.
+We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. 
+We also calculated the spectrum of triplets of a list the same number of randomly selected SNPs as a control.
 
 \subsection{Imputation}
-Using the Michigan Imputation Server, we imputed the genotype data from 1kGP
+Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2. 
+The VCF file returned from the server was then used to search for the number of significant SNPs imputed.
 

### 2018-08-09 ManhattanPlot (Autosaved).R

@@ -30,32 +30,29 @@ for(chrom in seq(1,22)){
 fileNames = list.files(path= out,pattern='*_MeanDev.csv', full.names = T)
 	
 Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
-
 Reg$plog10 <- - pchisq(Reg$SumDev, Reg$Count, lower.tail=F, log.p=T)/log(10)
-
 Reg$p <- pchisq(Reg$SumDev, Reg$Count, lower.tail=F)
 
-procedures <- c( "TSBH")
-adjusted <- mt.rawp2adjp(Reg$p, procedures)
-adj<-as.data.frame(adjusted$adj[order(adjusted$index), ])
-Reg$adjusted<-adj$TSBH_0.05
-Reg$log10a<--log10(adj$TSBH_0.05)
+##RICK Adjustment
+#procedures <- c( "TSBH")
+#adjusted <- mt.rawp2adjp(Reg$p, procedures)
+#adj<-as.data.frame(adjusted$adj[order(adjusted$index), ])
+#Reg$adjusted<-adj$TSBH_0.05
+#Reg$log10a<--log10(adj$TSBH_0.05)
+#ggplot(Reg, aes(x=p, y=adjusted))+geom_point(shape=1)+theme_classic()
+
+sig.6<-Reg[which((Reg$plog10 >= 6)&(Reg$Count > 1)),]
+sig.6[which(sig.6$plog10 > 20),]$log10a <- 20
+NotSig<-Reg[which((Reg$plog10 < 6)&(Reg$Count > 1)),]
 
 
-ggplot(Reg, aes(x=p, y=adjusted))+geom_point(shape=1)+theme_classic()
-
-sig.6<-Reg[which((Reg$log10a >= 6)&(Reg$Count > 1)),]
-sig.6[which(sig.6$log10a > 20),]$log10a<-20
-NotSig<-Reg[which((Reg$log10a < 6)&(Reg$Count > 1)),]
-
-
-ggplot(NotSig, aes(x=Pos, y = log10a, color=as.factor(Chr)))+
+ggplot(NotSig, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
 facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+
 scale_fill_manual (values=getPalette(colourCount))+
 scale_y_continuous(expand=c(0,0))+
 geom_point(alpha=0.3, size=1)+theme_classic()+
[…]

### 2018-08-09 legacy.tex
added figure 4 : Manhattan Plot

@@
 
+\subsection{Combined Test}
+To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
+Each population will have it's own deviation from the null for each position in the genome.
+We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
+This resulted in an increase in the number of SNPs reaching genome wide significance to 3664. \ref{Manhattan}
+
+\begin{figure*}
+\includegraphics[width=\hsize,keepaspectratio]{ManhattanPlot.jpg}
+
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
+There are 2007 and 3664 variants that reach p values greater than $ p < 10^{-8}$ and $ p < 10^{-6}$ respectively.
+Crosses ( + ) represent SNPs that reach a genome wide significance greater than $ p < 10^{-6}$, the circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+  \label{Manhattan}
+\end{figure*}
+
 	\subsection{Imputation}

### 2018-08-09 ManhattanPlot (Autosaved).R

@@ -1,102 +1,0 @@
-##Manhattan
-library(dplyr)
-library(magrittr)
-library(data.table)
-library(ggplot2)
-library(multtest)
-
-dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
-out='/Volumes/gravel/luke_projects/1000Genomes/MeanDev/'
-if(F){
-for(chrom in seq(1,22)){
-	print(chrom)
-	fileNames = list.files(path=dir,pattern=paste("CHR",chrom,".Regression",sep=''), full.names = T)
-	if(length(fileNames) != 0){
-	Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T, sep=' ')))
-	
-	print(head(Reg))
-	
-	Reg<-Reg[which(Reg$Pos != 'Pos'),]
-	Reg$Chr<-as.numeric(as.character(Reg$Chr))
-	Reg$Pos<-as.numeric(as.character(Reg$Pos))
-	Reg$dev<-as.numeric(as.character(Reg$dev))
-	
-	Reg <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
-	
-	write.table(Reg,paste(out,chrom,'_MeanDev.csv',sep=''))
-	}
-}
-}
-fileNames = list.files(path= out,pattern='*_MeanDev.csv', full.names = T)
-	
-Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
-Reg$plog10 <- - pchisq(Reg$SumDev, Reg$Count, lower.tail=F, log.p=T)/log(10)
-Reg$p <- pchisq(Reg$SumDev, Reg$Count, lower.tail=F)
-
-##RICK Adjustment
-#procedures <- c( "TSBH")
-#adjusted <- mt.rawp2adjp(Reg$p, procedures)
-#adj<-as.data.frame(adjusted$adj[order(adjusted$index), ])
-#Reg$adjusted<-adj$TSBH_0.05
-#Reg$log10a<--log10(adj$TSBH_0.05)
-#ggplot(Reg, aes(x=p, y=adjusted))+geom_point(shape=1)+theme_classic()
-
-sig.6<-Reg[which((Reg$plog10 >= 6)&(Reg$Count > 1)),]
-sig.6[which(sig.6$plog10 > 20),]$log10a <- 20
[…]

### 2018-08-09 ManhattanPlot.R

@@ -30,31 +30,22 @@ for(chrom in seq(1,22)){
 fileNames = list.files(path= out,pattern='*_MeanDev.csv', full.names = T)
 	
 Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
-
 Reg$plog10 <- - pchisq(Reg$SumDev, Reg$Count, lower.tail=F, log.p=T)/log(10)
-
 Reg$p <- pchisq(Reg$SumDev, Reg$Count, lower.tail=F)
 
-procedures <- c( "TSBH")
-adjusted <- mt.rawp2adjp(Reg$p, procedures)
-adj<-as.data.frame(adjusted$adj[order(adjusted$index), ])
-Reg$adjusted<-adj$TSBH_0.05
-Reg$log10a<--log10(adj$TSBH_0.05)
+sig.6<-Reg[which((Reg$plog10 >= 6)&(Reg$plog10 < 20)&(Reg$Count > 1)),]
+sig.20<-Reg[which((Reg$plog10 >= 20)&(Reg$Count > 1)),]
+sig.20$plog10=20
+NotSig<-Reg[which((Reg$plog10 < 6)&(Reg$Count > 1)),]
 
 
-ggplot(Reg, aes(x=p, y=adjusted))+geom_point(shape=1)+theme_classic()
-
-sig.6<-Reg[which((Reg$log10a >= 6)&(Reg$Count > 1)),]
-sig.6[which(sig.6$log10a > 20),]$log10a<-20
-NotSig<-Reg[which((Reg$log10a < 6)&(Reg$Count > 1)),]
-
-
-ggplot(NotSig, aes(x=Pos, y = log10a, color=as.factor(Chr)))+
+ggplot(NotSig, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
 facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+
 scale_fill_manual (values=getPalette(colourCount))+
 scale_y_continuous(expand=c(0,0))+
 geom_point(alpha=0.3, size=1)+theme_classic()+
 geom_point(data=sig.6, aes(x=Pos, y = plog10,),color='black',shape=3)+
+geom_point(data=sig.20, aes(x=Pos, y = plog10,),color='black',shape=1)+
 labs(y='-log10(p)', x='Chromosome')+
 theme(plot.title = element_text(hjust = 0.5), 
[…]

### 2018-08-10 legacy.tex
wrote some text

@@
 \todo{Supplemental Figures : all the GWAS results for each population}
-Comparing the results of each independent GWAS, we were able to identify over 3493 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
+Comparing the results of each independent GWAS, we were able to identify over 3517 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
 
@@
 
-	\subsection{Found to be included in other GWA studies}
+	\subsection{Found to be included in other GWAS }
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have called these erroneous variants as being significantly correlated with some biological trait.
-Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found \todo{000} recent publications that had found at least one of the variants to have reached genome wide significance in their study.
+Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found two recent publications that had found at least one of the variants to have reached genome wide significance in their study.
 
-These studies used the 1kGP as the reference database for imputation. %and one used the 1kGP cell cultures and sequence data. 
-They used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+One of these studies genotyped individuals and imputed the data using the HapMapII as a reference  database for imputation. 
+The other study used the 1kGP sequence data directly.
+Both papers used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness. 
@@
  \hline
+Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
 
+Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
+& & & rs201255786, rs201761909&\\
  \hline
@@
 \subsection{Why do we care?}
-These SNPs matter because they reached genome wide significance for medically relevant traits. 
+These SNPs reached genome wide significance for medically relevant traits. 
 Including these variants in the GWAS catalogue could lead to false diagnosis at worst, or spurious correlations at least. 

### 2018-08-10 legacy.tex
added text, added supplementary figures

@@
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 [\cite{}].
+
 This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population [\cite{}]. 
@@
 We found that 24 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
-\todo{Supplemental Figures : all the GWAS results for each population}
 Comparing the results of each independent GWAS, we were able to identify over 3517 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
@@
 			\section{Discussion}
-\subsection{Why do we care?}
-These SNPs reached genome wide significance for medically relevant traits. 
-Including these variants in the GWAS catalogue could lead to false diagnosis at worst, or spurious correlations at least. 
-Polygenic risk scores take into account all SNPs reaching significance, without much manual curation. 
-It's likely that the variants we have identified are being included in these multi-locus risk scores.
+%\subsection{Why do we care?}
+The variants identified in this study are likely to be technical artifacts from legacy technology.
+If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
+However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
+We have shown that these variants have been significantly associated using traditional GWAS methods.
+However this is not the only method that would be affected by these false positives. 
+Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
+The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.
 
-Despite these variants reaching genome wide significance, the majority of researchers did not pursue these variants for further analysis.
-This is likely because these variants have no peak in LD that is characteristic of a biologically significant variant.
+%\subsection{Recommendations}
+The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
+In this case, we used a cut off of an average quality of mapped bases over 30. It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
 
-\subsection{Recommendations}
-The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
-In this case, we used a cut off of an average quality of mapped bases over 30.
-This threshold has been \todo{previously used by studies}.
-It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
-
-\subsection{Power limits and Quality Controls}
+%\subsection{Power limits and Quality Controls}
 If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
@@
 We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.
-However even the GWAS is limited by power as we were able to identify \todo{000} more variants that were missing in a higher quality dataset but present in over 1\% of the 1kGP Japanese data. 
-This suggests that while the GWAS we performed is a step in the right direction, the GWAS is not sufficient because the data is not reliable in the first place.
 
-\subsection{Technical Source of Error}
+%\subsection{Technical Source of Error}
 %B tail trimming  (Minoche et al. Genome Biology 2011)
@@
 
-\subsection{GWAS from other papers}
+%\subsection{GWAS from other papers}
 %The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
@@
 \section{Code Availability}
-\todo{here is where I will put my git hub}
+https://github.com/LukeAndersonTrocme/QualityPaper
 
@@
 
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{RegressionPlot_mostSig2.jpg}
+\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to quality.}
+\label{MostSig}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{RegressionPlot.jpg}
+\caption{Logistic regression of rs6057648 for the populations with significant association to quality.}
+\label{TwinsSNP}
+\end{figure}
+
 \bibliographystyle{plain}

### 2018-08-10 ManhattanPlot.R

@@ -5,6 +5,8 @@ library(data.table)
 library(ggplot2)
 library(multtest)
 
+rsID<-fread('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_Chr.Pos.rsID.txt')
+
 dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
 out='/Volumes/gravel/luke_projects/1000Genomes/MeanDev/'
 if(F){
@@ -32,12 +34,14 @@ fileNames = list.files(path= out,pattern='*_MeanDev.csv', full.names = T)
 Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T)))
 Reg$plog10 <- - pchisq(Reg$SumDev, Reg$Count, lower.tail=F, log.p=T)/log(10)
 Reg$p <- pchisq(Reg$SumDev, Reg$Count, lower.tail=F)
+Reg<-merge(Reg, rsID, by.x=c('Chr','Pos'), by.y=c('V1','V2'), all.x=T)
 
 sig.6<-Reg[which((Reg$plog10 >= 6)&(Reg$plog10 < 20)&(Reg$Count > 1)),]
 sig.20<-Reg[which((Reg$plog10 >= 20)&(Reg$Count > 1)),]
 sig.20$plog10=20
 NotSig<-Reg[which((Reg$plog10 < 6)&(Reg$Count > 1)),]
 
+write.table(Reg[which((Reg$plog10 >= 6)&(Reg$Count > 1)),], '~/Documents/QualityPaper/Significant6SNPs.txt', quote=F, row.names=F)
 
 ggplot(NotSig, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
 facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+

### 2018-08-10 PlotRegression.R

@@ -1,0 +1,34 @@
+library(ggplot2)
+
+#pop names, used to read/write files
+pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
+#read colors per pop (used in 1kG Nature)
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
+snp<-read.table('~/Documents/QualityPaper/31384880.txt',sep=',')
+dev<-read.table('~/Documents/Regression/31384880.dev.txt')
+dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
+dev<-dev[which(dev$p>6),]
+sig<-merge(snp, dev, by.x=c('V1','V2','V5'), by.y=c('V1','V2','V3'))
+sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$V4.y, digits=2))
+
+ggplot(sig, aes(x=V6, y=V4.x))+geom_point(shape=1)+facet_wrap(~Name,ncol=1)+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))
+
+ggsave('~/Documents/QualityPaper/RegressionPlot.jpg',height=5,width=5)
+
+sig6<-read.table('~/Documents/QualityPaper/Significant6SNPs.txt', header=T)
+
+snp<-read.table('~/Documents/QualityPaper/33264017.txt',sep=',', fill=T)
+dev<-read.table('~/Documents/QualityPaper/33264017.dev.txt')
+dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
+dev<-dev[which(dev$p>6),]
+sig<-merge(snp, dev, by.x=c('V2','V5'), by.y=c('V2','V3'))
+sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$p, digits=2))
+sig<-sig[order(sig$p),]
+sig$Name<-factor(sig$Name, levels=unique(sig$Name))
+
[…]

### 2018-08-16 legacy.tex
minor edits

@@
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \cite{Pleasance2010}.
-This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation [\cite{}].
+This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
 While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity [\cite{}].
@@
 Fine scale mutational signatures are observable on evolutionary scales by comparing the rates of mutations between populations.
-In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 [\cite{}].
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \cite{Harris2015a}.
+A more in depth analysis of these subtle mutational signatures was recently released in preprint by Aikens, Johnson and Voight.
 
-This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population [\cite{}]. 
-Harris and Pritchard further identified distinct mutational spectra across a range of populations [\cite{}]. 
+This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population \cite{Harris2015a}. 
+Harris and Pritchard further identified distinct mutational spectra across a range of populations \cite{Harris2017a}. 
 In particular, they identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals. 

### 2018-08-17 legacy.tex
updated text for Aikens et al

@@
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \cite{Harris2015a}.
-A more in depth analysis of these subtle mutational signatures was recently released in preprint by Aikens, Johnson and Voight.
+A more in depth analysis of these subtle mutational signatures was recently released in preprint by Aikens, Johnson and Voight. \cite{}
+In their study they expanded the investigations of mutational signals in the 1kGP by investigations to broader sequence contexts.
 
@@
 In particular, they identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+In addition, multiple 7-mer mutational signatures were also identified by Aikens et. al in the same populations. \cite{}
 Such an axis of variation is intriguing because differences in mutational signatures accumulate over many generations.
@@
 
-Harris and Pritchard reported an excess of *AC${\rightarrow}$*CC mutations in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
+Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 
-%\subsection{GWAS from other papers}
-%The only way to distinguish some of these more covert false positives is to use statistical tests associating the quality metrics of each position relative to each individual.
+\subsection{Twins Research Human Genetics}
+Genotype + Imputation + GWAS
+
+\subsection{PLos One}
+Using 1kGP data directly, estimating count of EBV
 

### 2018-08-17 legacy.tex
added sections for the two papers

@@
 This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
-While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity [\cite{}].
+While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
 Large cohorts have recently become available to increase statistical power to resolve some of these fine scale mutational signatures.
@@
 A more in depth analysis of these subtle mutational signatures was recently released in preprint by Aikens, Johnson and Voight. \cite{}
-In their study they expanded the investigations of mutational signals in the 1kGP by investigations to broader sequence contexts.
+In their study they expanded the investigations of mutational signals in the 1kGP by investigations to broader sequence contexts and found differences in mutation profiles across populations. \cite{}
 
@@
 The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations. 
+A slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations is also present.
+By taking the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*, these three enrichments can be summarized as G**${\rightarrow}$GG*.
 
@@
 One of these studies genotyped individuals and imputed the data using the HapMapII as a reference  database for imputation. 
-The other study used the 1kGP sequence data directly.
+The other study used the 1kGP sequence data and cell lines directly.
 Both papers used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
@@
 \section{Discussion}
-%\subsection{Why do we care?}
+
+\subsection{Twins Research Human Genetics}
+This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
+While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
+The data used was imputed genotype data using Hapmap II as the reference.
+One of the top 25 SNPs in the GWAS on ADHD was identified as being significantly associated to data quality.
+
+While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified. 
+Furthermore, the approach they used to impute genotype data for GWAS is very common.
+A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results.
+
+\subsection{PLos One}
+This paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the coverage of mapped reads to EBV.
+They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
+They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
+However they reported the SNPs reading significance below $ p < 10^{-6}$. 
+Four of the SNPs in the top hits of the Asian populations we're identified by our regression against data quality.
+
+The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphobastoid cell lines (LCLs).
+This is a commonly used transformation that helps preserve cell lines for long periods of time.
+It's unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that perhaps different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
+Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics.
+
+\subsection{Technical Artifact}
 The variants identified in this study are likely to be technical artifacts from legacy technology.
@@
 
-%\subsection{Recommendations}
+\subsection{Recommendations}
 The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
-In this case, we used a cut off of an average quality of mapped bases over 30. It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
+In this case, we used a cut off of an average quality of mapped bases over 30. 
+It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
 
 %\subsection{Power limits and Quality Controls}
-If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
-\todo{how many hetd do you need to detect HWE significance?}
-We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.
+%If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
+%how many hetd do you need to detect HWE significance?
+%We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.
 
@@
 %repetitive regions, GC bias...
-
-\subsection{Twins Research Human Genetics}
-Genotype + Imputation + GWAS
-
-\subsection{PLos One}
-Using 1kGP data directly, estimating count of EBV
 

### 2018-08-20 legacy.tex
updated text.
minor updates

@@
 \author[1,2]{Rick Farouni}
-\author[3]{Fumihiko Matsuda (and coauthors)}
+\author[3]{Fumihiko Matsuda}
+\author[3]{Yoichiro Kamatani}
 \author[1,2]{Simon Gravel}
@@
 Hapmap \cite{HapMap2005}, the 1000 Genomes Project \cite{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \cite{Mallick2016}, for example, have made thousands of genomes publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \cite{ProfJonathanMarchiniProfGoncaloAbecasisProfRichardDurbin2014} or variant frequency estimation \cite{Lek2016}.
+Many more genomes are available indirectly through servers providing imputation services \cite{ProfJonathanMarchiniProfGoncaloAbecasisProfRichardDurbin2014} or summary statistics for variant frequency estimation \cite{Lek2016}.
 
@@
 During this period, sequencing platforms were still being rapidly developed.
-This lead to heterogeneous sample preparations for many of the samples included in the dataset.
+This lead to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
 Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
@@
 
-This strong population enrichment of a mutational signature suggests important genetic or environmental differences in the history of each population \cite{Harris2015a}. 
+Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \cite{Harris2015a}. 
 Harris and Pritchard further identified distinct mutational spectra across a range of populations \cite{Harris2017a}. 
@@
 
-When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears \ref{SFS}.
+When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
 Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
-Individuals with low-quality data showing elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
-Thus sequences with low mappability harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+Individuals with low-quality data show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
+Thus sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality \ref{SFS}.
-\luke{NEW:} 
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
 Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, our method does the inverse.
@@
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
- %This GWAS included 104 Japanese individuals, 74 of which were sequenced in using the Genome Analyzer II while the rest were sequenced using the HiSeq 2000 of the 1kGP.
 The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations. 
@@
 
-%Many of the populations in the 1kGP do not have higher quality counterparts which means that identifying questionably low quality data must be done ad hoc.
-
 	\subsection{Sequencing quality over time}
-The results of the GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
-The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data \ref{MapQual}.
+The results of the logistic regression GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
 This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
@@
 By 2011, older sequencing technologies are phased out, and method development solidify, which coincides with the sequencing quality levelling off.
+It's clear that there are many covariates that could be used to identify spurious sites like sequencing date, sequencing technology or even the sequencing centre. 
+However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
 
@@
 \caption{\textbf{A} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
-The x-axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual.
-\textbf{B} Same data as in \textbf{A} except the x-axis is sorted by sequencing date.
-The colours indicate the sequencing centers that produced the data for each individual.}
+The earliest sequencing data is used for individuals sequenced more than once.
+The x-axis is ranked by the mean sequencing date per population.
+\textbf{B} The x-axis is sorted by the sequencing date per individual.
+The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project.
+The bottom plot indicates the sequencing technologies used over time.}
 \label{MapQual}
@@
 	\subsection{Overlap of significant SNPs}
-We applied the same GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
+We applied the same logistic regression GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
 We found that 24 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
-Comparing the results of each independent GWAS, we were able to identify over 3517 variants that were independently associated to low quality in at least two populations  \ref{OverLap}.
+Comparing the results of each independent GWAS, we were able to identify over 3517 variants that were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
 
@@
 We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
-This resulted in an increase in the number of SNPs reaching genome wide significance to 3664. \ref{Manhattan}
-
-\begin{figure*}
-\includegraphics[width=\hsize,keepaspectratio]{ManhattanPlot.jpg}
-
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
-There are 2007 and 3664 variants that reach p values greater than $ p < 10^{-8}$ and $ p < 10^{-6}$ respectively.
-Crosses ( + ) represent SNPs that reach a genome wide significance greater than $ p < 10^{-6}$, the circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
-  \label{Manhattan}
-\end{figure*}
+This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).
 
@@
 
-\begin{center}
- \begin{tabular}{l l l r r} 
- \hline
-Author & Year & Journal & RSID & PMID \\ [0.5ex] 
- \hline
-Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
-
-Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
-& & & rs201255786, rs201761909&\\
- \hline
- \end{tabular}
-\end{center}
-
 \section{Discussion}
@@
 The data used was imputed genotype data using Hapmap II as the reference.
-One of the top 25 SNPs in the GWAS on ADHD was identified as being significantly associated to data quality.
+One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality.\cite{}
 
@@
 However they reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations we're identified by our regression against data quality.
+Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality. \cite{}
 
-The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphobastoid cell lines (LCLs).
+The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphobastoid cell lines.
 This is a commonly used transformation that helps preserve cell lines for long periods of time.
@@
 It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
-
+Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
+(See supplementary material for lists of SNPs)
 %\subsection{Power limits and Quality Controls}
@@
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. 
-We also calculated the spectrum of triplets of a list the same number of randomly selected SNPs as a control.
+We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. \cite{Harris2017a}
+We modified the methods as necessary for our purposes, scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
 
@@
 Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2.
-The VCF file returned from the server was then used to search for the number of significant SNPs imputed.
+We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip} genotype data.
+The VCF file returned from the server was then downloaded and used to search for the number of significant SNPs successfully imputed.
 
@@
 \section{Acknowledgments}
-We would like to thank Kelly Harris for sharing her mutation spectrum pipelines.
+We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
+
+\bibliographystyle{plain}
+\bibliography{Legacy.bib}
 
 \subsection{Supplementary Figures}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{ManhattanPlot.jpg}
+
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
+There are 2007 and 3664 variants that reach p values greater than $ p < 10^{-8}$ and $ p < 10^{-6}$ respectively.
+Crosses ( + ) represent SNPs that reach a genome wide significance greater than $ p < 10^{-6}$, the circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+  \label{Manhattan}
+\end{figure}
+
 \begin{figure}
@@
 
-\bibliographystyle{plain}
-\bibliography{Legacy.bib}
+\end{document}
 
-\end{document}
+%\begin{center}
+ %\begin{tabular}{l l l r r} 
+ %\hline
+%Author & Year & Journal & RSID & PMID \\ [0.5ex] 
+ %\hline
+%Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
+
+%Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
+%& & & rs201255786, rs201761909&\\
+%\hline
+%\end{tabular}
+%\end{center}

### 2018-08-21 legacy.tex
moved misc files to misc folder and figures to figures folder
comments throughout

@@
 \begin{abstract}
-In investigating an unusual population genetics signal we identified a strong and previously unreported batch effect in the 1000 Genomes Project data.
-We also found that this batch effect led to spurious results in a number of recent publications. 
-We identified spurious mutations by correlating mutations with data quality metrics.
-We then turned our attention to the rest of the 1000 Genomes populations and saw that there were similar batch effects in many of these populations.
-Some of these variants are being imputed onto genotype data and reach genome wide significance in recent publications.
+Recent reports have identified differences in the mutational spectra across human populations.
+While some of these reports have been replicated in other cohorts, most have been reported in the 1000 Genomes project data.
+While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1000 Genomes Project data and to the apparent population stratification.
+Because the 1000 Genomes data is used extensively, we find that the spurious calls also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
+Lower-quality data from the early phases of the 1000 Genomes project thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
 \end{abstract}
@@
 
-Many of these large datasets have been assembled over several years using a variety of sequencing platforms. 
-The first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced 10 years ago.
-During this period, sequencing platforms were still being rapidly developed.
-This lead to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
-Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and a wide range of applications.
+Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+However, the first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced 10 years ago.
+Since then, sequencing platforms have rapidly improved.
+The second phase of the 1kGP describe multiple technological and analytical improvements over its earlier phases \cite{}.
+This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
 
-Although, it is clear that the quality and accuracy of the 1kGP data has increased over the past decade. 
-This data is far from perfect, the publication from the second phase of the 1kGP describe technological and analytical improvements over its previous phases.
+Even though it is intuitively clear that such batch effects have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data makes such issues particularly insidious.
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 
-Genome wide mutational signatures have been used as a diagnostic tool for cancers \cite{Pleasance2010,Shiraishi2015a}.
+Different mutagenic processes preferentially affect different DNA motifs.
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \cite{Pleasance2010}.
-This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
-While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
-Large cohorts have recently become available to increase statistical power to resolve some of these fine scale mutational signatures.
+Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of `signatures' of different mutational processes\cite{Pleasance2010}.
+Such genome wide mutational signatures have been used as a diagnostic tool for cancers \cite{Pleasance2010,Shiraishi2015a}.
 
-Fine scale mutational signatures are observable on evolutionary scales by comparing the rates of mutations between populations.
+%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
+%Large cohorts have recently become available to increase statistical power to resolve some of these finer scale mutational signatures.
+
+Differences in mutational signatures across human populations have recently been reported.
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \cite{Harris2015a}.
-A more in depth analysis of these subtle mutational signatures was recently released in preprint by Aikens, Johnson and Voight. \cite{}
-In their study they expanded the investigations of mutational signals in the 1kGP by investigations to broader sequence contexts and found differences in mutation profiles across populations. \cite{}
 
 Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \cite{Harris2015a}. 
-Harris and Pritchard further identified distinct mutational spectra across a range of populations \cite{Harris2017a}. 
-In particular, they identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
-In addition, multiple 7-mer mutational signatures were also identified by Aikens et. al in the same populations. \cite{}
-Such an axis of variation is intriguing because differences in mutational signatures accumulate over many generations.
+Harris and Pritchard further identified distinct mutational spectra across a range of populations \cite{Harris2017a}, which were further studied in \cite{Voight}
+
+ In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+This heterogeneity is intriguing because differences in mutational signatures accumulate over many generations.
 A systematic difference among the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
@@
 
-While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy between the two datasets and identified a strong and previously unreported batch effect in the 1kGP data.
-We also found that this batch effect led to spurious results in a number of recent publications.
+While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
+We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.
 
-This study is the result of an investigation in the quality of the 1kGP dataset. 
-To begin, we will discuss how we discovered these previously unreported batch effects in the 1kGP. 
-Next, we explore more broadly how low quality data remain embedded in other populations of the 1kGP. 
-Finally, we will discuss the ways in which it contaminated a number of recent studies.
-Fortunately, the main conclusions of most of these studies remain supported by other data.
-Yet this begs the question: When should we retire legacy data?
+The results section is organized as follows:
+First, we describe the observed discrepancies within and between studies.
+We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.
+Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
+Even though these SNPs are likely technical artifacts, we two recent studies report such SNPs as genome-wide significant hits.
+
+Fortunately, the main conclusions of most of the affected studies remain supported by other data. Yet this begs the question: When should we retire legacy reference data?
 
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Figure1.jpg}
 \caption{
@@
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of private single nucleotide polymorphisms (SNPs), only found in one of the two groups.
-This is unexpected given the similarity of the two populations, and suggests a technical artifact rather than a population structure effect. 
+However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
+This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
 Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
@@
 When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
-Individuals with low-quality data show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
+Individuals with low mapping quality show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
 Thus sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
-Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, our method does the inverse.
-We are looking for loci where the genotypes are predictable based on the data quality scores.
-We were able to identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
-While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., the traditional genome-wide significance threshold might not be a good guide for filtering).
-The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations. 
-A slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations is also present.
-By taking the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*, these three enrichments can be summarized as G**${\rightarrow}$GG*.
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
+Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, we are looking for loci where the genotypes are predictable based on the data quality scores.
+This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
+While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
+The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.
 
-The signal is enriched in GWAS SNPs, but residual signal remains presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality.
-The *AC${\rightarrow}$*CC mutational signal is limited to individuals with low quality sequencing data.
-Therefore, the removal of individuals with average qualities below 30 successfully removes the signal.
-While the GWAS approach can identify some of the low hanging bad apples, there are likely more of these false positives nested inside these legacy cohorts.
+Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+\sgcomment{do we excluse singletons?
+We should quantify how much reduction in signal we see.}
+%For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
+The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal \sgcomment{How about the others?}.
+For population genetic analyses, the removal of low-mapping quality individuals appears preferable to filtering.
 
@@
 It's clear that there are many covariates that could be used to identify spurious sites like sequencing date, sequencing technology or even the sequencing centre. 
-However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
+However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites. \sgcomment{supplementary figures with the general regressions?}
 
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{MapQualOverTime.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
 
-\caption{\textbf{A} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
 The earliest sequencing data is used for individuals sequenced more than once.
@@
 \subsection{Overlap of significant SNPs}
-We applied the same logistic regression GWAS approach on the other 1kGP populations as it appears to capture a large portion of problematic sites. 
-We found that 24 of the 26 populations in the 1kGP had SNPs significantly associated to the average quality of mapped bases per individual.
-Comparing the results of each independent GWAS, we were able to identify over 3517 variants that were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
+The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
 
 \begin{figure*}
-\includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
 
@@
 \subsection{Combined Test}
-To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
-Each population will have it's own deviation from the null for each position in the genome.
-We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
-This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).
+%To increase statistical power, we performed a meta-analysis, computing a combined p-value for each position by taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
+%To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
+%Each population will have it's own deviation from the null for each position in the genome.
+%We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
+%This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).
 
 \subsection{Imputation}
-The state of the art imputations servers use a combination of many databases including many that are not freely available.
-For this reason, the details of the methods used to impute genomic data are not publicly available either.
-In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from the 1kGP and imputed it using the Michigan Imputation Server.
-We found that all of the SNPs we identified as being associated with quality were imputed.
-This suggests that the imputation reference panel includes individuals with low sequence quality, as well as the variants that have been shown to be associated to quality.
+The state of the art imputation servers use a combination of many databases including many that are not freely available.
+From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
+In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from low-quality 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality were imputed.
+\sgcomment{All of them?
+How is it possible if we only submitted Japanese?
+Also, I thought that we had only submitted a couple of chromosomes.}
+This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
+\sgcomment{It may be useful to impute some of the NAG data.}
 
 \subsection{Found to be included in other GWAS }
-Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have called these erroneous variants as being significantly correlated with some biological trait.
-Using the NHGRI-EBI Catalog of published genome-wide association studies we queried the rsIDs of the SNPs we identified as being low quality and found two recent publications that had found at least one of the variants to have reached genome wide significance in their study.
+Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
+The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found at least one of the variants to have reached genome wide significance.
 
-One of these studies genotyped individuals and imputed the data using the HapMapII as a reference  database for imputation. 
+One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
+\sgcomment{If they used HapMap, how did they get affected by the 1000 Genomes results?}
 The other study used the 1kGP sequence data and cell lines directly.
@@
 \subsection{Twins Research Human Genetics}
+\sgcomment{HAd we ever discusssed that these GWAS hits were not genome-wide significant?}
 This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
@@
 Furthermore, the approach they used to impute genotype data for GWAS is very common.
-A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results.
+A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
 
@@
 \subsection{Technical Artifact}
+\sgcomment{needs some editing}
 The variants identified in this study are likely to be technical artifacts from legacy technology.
@@
 \subsection{Recommendations}
+\sgcomment{needs some editing}
 The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
@@
 \section{Conclusion}
-Our method identifies spurious mutations by correlating mutations with data quality metrics. 
-We propose including our quality control methods to identify possible false positives in sequencing data. 
-We have focused on the 1000 Genomes Project dataset as its quality metrics were freely available, however the issues of quality control are not limited to this one consortium. 
-This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.
 
-As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
-These results bring forth many questions regarding the reliability of legacy datasets. 
-Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.
+On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before.
+The reverse GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.
 
-Many types of DNA sequencing machines have been available on the market over the past decade.
-Different types of sequencing technologies can have different technical limitations in quality control.
-This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
-Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-The errors in one dataset do not disappear when they are combined with another higher quality one.
-The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
+More generally, to improve the quality of genomic reference datasets, we can proceed by addition of new and better data and by better curation of existing data.
+Given rapid technological progress, the focus of genomic research is naturally on the data generation side.
+However, cleaning up is also important to avoid generating spurious results.
+The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing.
+
+Somewhat embarrassingly, we have not identified the precise mechanism through which the technical artifact was introduced: we have ruled out neither cell-line, sequencer, nor algorithmic artifact.
+Given that modern platforms seem to have resolved this technical issue but that recent studies continue being affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while the sequencing forensics investigation can take place.
+
+%but evidence points to the fact that the artifact disappeared in the later stages of the 1kGP, and does not corrupt more recent data.
+
+%As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
+%These results bring forth many questions regarding the reliability of legacy datasets. 
+%Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.
+
+%Our method identifies spurious mutations by correlating mutations with data quality metrics. 
+%We propose including our quality control methods to identify possible false positives in sequencing data. 
+%We have focused on the 1000 Genomes Project dataset as its quality metrics were freely available, however the issues of quality control are not limited to this one consortium. 
+%This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.
+
+%Many types of DNA sequencing machines have been available on the market over the past decade.
+%Different types of sequencing technologies can have different technical limitations in quality control.
+%This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
+%Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
+%The errors in one dataset do not disappear when they are combined with another higher quality one.
+%The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
 
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{ManhattanPlot.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot.jpg}
 
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{PC1_Correlation.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
 \caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{RegressionPlot_mostSig2.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
 \caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to quality.}
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{RegressionPlot.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
 \caption{Logistic regression of rs6057648 for the populations with significant association to quality.}

### 2018-08-22 legacy.tex
updated manhattan plot

@@
 
-\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
 The earliest sequencing data is used for individuals sequenced more than once.
@@
 In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from low-quality 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality were imputed.
-\sgcomment{All of them?
-How is it possible if we only submitted Japanese?
-Also, I thought that we had only submitted a couple of chromosomes.}
+\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
+\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
@@
 Furthermore, the approach they used to impute genotype data for GWAS is very common.
-A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
+%A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
 

### 2018-08-22 Clean_Figure1.R

@@ -186,7 +186,7 @@ Reg$p<-pchisq(Reg$dev, 1, lower.tail=F)
 colnames(Reg)[c(1,2)]<-c('CHROM','POS')
 
 RegJoin<-left_join(Reg, join, by=c('CHROM','POS'))
-RegJoin<-RegJoin[which(RegJoin$JPT_MAF > 1/208),]
+#RegJoin<-RegJoin[which(RegJoin$JPT_MAF > 1/208),]
 
 ggplot(RegJoin, aes(sample=RegJoin$dev))+stat_qq(distribution=stats::qchisq, dparams=list(df=1))+geom_qq_line(distribution=stats::qchisq, dparams=list(df=1))
 
@@ -194,6 +194,18 @@ ggsave('~/Documents/QualityPaper/Figures/QQplot_JPT.jpg', height=5, width=7)
 
 ggplot(Reg, aes(sample=Reg$dev))+stat_qq(distribution=stats::qchisq, dparams=list(df=1))+geom_qq_line(distribution=stats::qchisq, dparams=list(df=1))
 ggsave('~/Documents/QualityPaper/Figures/QQplot_JPT_withSingles.jpg', height=5, width=7)
+
+single<-RegJoin[which(RegJoin$JPT_MAF == 0.00480769),]
+ggplot(single, aes(sample= single$dev))+stat_qq(distribution=stats::qchisq, dparams=list(df=1))+geom_qq_line(distribution=stats::qchisq, dparams=list(df=1))
+ggsave('~/Documents/QualityPaper/Figures/QQplot_JPT_justSingles.jpg', height=5, width=7)
+
+double<-RegJoin[which(RegJoin$JPT_MAF == 2*0.00480769),]
+ggplot(double, aes(sample= double$dev))+stat_qq(distribution=stats::qchisq, dparams=list(df=1))+geom_qq_line(distribution=stats::qchisq, dparams=list(df=1))
+ggsave('~/Documents/QualityPaper/Figures/QQplot_JPT_justDouble.jpg', height=5, width=7)
+
+triple<-RegJoin[which(RegJoin$JPT_MAF == 3*0.00480769),]
+ggplot(triple, aes(sample= triple$dev))+stat_qq(distribution=stats::qchisq, dparams=list(df=1))+geom_qq_line(distribution=stats::qchisq, dparams=list(df=1))
[…]

### 2018-08-22 PlotRegression (Autosaved).R

@@ -1,38 +1,0 @@
-library(ggplot2)
-
-#pop names, used to read/write files
-pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
-#read colors per pop (used in 1kG Nature)
-colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
-MyColour <- as.character(colors $hex)
-names(MyColour) <- colors $Pop
-
-snp<-read.table('~/Documents/QualityPaper/31384880.txt',sep=',')
-dev<-read.table('~/Documents/Regression/31384880.dev.txt')
-dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
-dev<-dev[which(dev$p>6),]
-sig<-merge(snp, dev, by.x=c('V1','V2','V5'), by.y=c('V1','V2','V3'))
-sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$V4.y, digits=2))
-
-ggplot(sig, aes(x=V6, y=V4.x))+geom_point(shape=1)+facet_wrap(~Name,ncol=1)+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))
-
-ggsave('~/Documents/QualityPaper/RegressionPlot.jpg',height=5,width=5)
-
-sig6<-read.table('~/Documents/QualityPaper/Significant6SNPs.txt', header=T)
-
-snp<-read.table('~/Documents/QualityPaper/33264017.txt',sep=',', fill=T)
-dev<-read.table('~/Documents/QualityPaper/33264017.dev.txt')
-dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
-dev<-dev[which(dev$p>6),]
-sig<-merge(snp, dev, by.x=c('V2','V5'), by.y=c('V2','V3'))
-sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$p, digits=2))
-sig<-sig[order(sig$p),]
-sig$Name<-factor(sig$Name, levels=unique(sig$Name))
-
[…]

### 2018-08-22 PlotRegression.R

@@ -32,3 +32,7 @@ sig$Name<-factor(sig$Name, levels=unique(sig$Name))
 ggplot(sig, aes(x=V6, y=V4.x, color=V5))+geom_point(shape=1)+facet_wrap(~Name,nrow=1)+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))+geom_vline(xintercept=30, color='grey60',linetype=3)+coord_flip()+scale_color_manual(breaks= sig$V5,values = MyColour)+guides(color=F)
 
 ggsave('~/Documents/QualityPaper/RegressionPlot_mostSig2.jpg',height=6,width=12)
+
+
+weird<-fread('~/Documents/QualityPaper/Misc/LowAFsig_Format.txt')
+ggplot(weird, aes(x=V6, y=V4))+facet_grid(.~V2)+geom_point()+coord_flip()+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme(axis.text.x=element_blank())+labs(x='Genotypes', y='Quality')

### 2018-08-27 legacy.tex
updated figures

@@
 Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
+Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, in essence we would not expect any homozygous alternates for low frequency variants.
 
@@
 The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
 \sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
@@
 
-\subsection{Combined Test}
-%To increase statistical power, we performed a meta-analysis, computing a combined p-value for each position by taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
-%To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
-%Each population will have it's own deviation from the null for each position in the genome.
-%We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
-%This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).
-
 \subsection{Imputation}
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
-In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from low-quality 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality were imputed.
-\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
-\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
+In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed.
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
+%\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
+%\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
 \sgcomment{It may be useful to impute some of the NAG data.}
@@
 
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
+\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
+\label{Singles}
+\end{figure}
+
 \end{document}

### 2018-08-27 Clean_Figure1.R

@@ -42,10 +42,10 @@ join=fread('~/Documents/QualityPaper/Misc/join.txt')
 
 ##JPT Manhattan
 if(f){
-dir='/Volumes/gravel-1/luke_projects/1000Genomes/Regression/'
-out='/Volumes/gravel-1/luke_projects/1000Genomes/MeanDev/'
+dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
+out='/Volumes/gravel/luke_projects/1000Genomes/MeanDev/'
 
-jptNames = list.files(path=dir,pattern='JPT', full.names = T)
+jptNames = list.files(path=dir,pattern='_JPT_dev10.csv', full.names = T)
 jReg = do.call(rbind, lapply(jptNames, function(x) fread(x, header=T, sep=' ')))
 jReg<-jReg[which(jReg$Pos != 'Pos'),]
 jReg$Chr<-as.numeric(as.character(jReg$Chr))

### 2018-08-27 Clean_Figure3.R

@@ -27,38 +27,50 @@ NumPop<-NumPop[which(NumPop$Freq>1),]
 NumPop$Var1<-as.numeric(as.character(NumPop$Var1))
 NumPop$Var2<-as.numeric(as.character(NumPop$Var2))
 
+
 m=merge(Reg,NumPop, by.y=c('Var1','Var2'), by.x=c('Chr','Pos'))
 n=as.data.frame(table(m$Pop))
 plt=merge(m,n,by.x='Pop',by.y='Var1')
+plt$title = paste(plt$Pop, ' : ',plt$Freq.y)
 plt <- plt[with(plt, order(-Freq.x, as.numeric(Chr),as.numeric(Pos))),]
 plt$Pos <- factor(plt$Pos, levels = unique(plt$Pos))
 plt$nrow<-seq(length=nrow(plt))
-plt$Pcat<-4
-plt[which(plt$plog10 <8),]$Pcat<-2
-plt[which((plt$plog10 >8) & (plt$plog10 <10)),]$Pcat<-3
+plt$Pcat<-10
+plt[which(plt$plog10 <8),]$Pcat<-1
+plt[which((plt$plog10 >8) & (plt$plog10 <10)),]$Pcat<-5
+
 
 xaxis<-plt[!duplicated(plt[,c('Freq.x')]), ]
 xaxis<-xaxis[which(xaxis$Freq.x<11),]
 xaxis<-rbind(xaxis,tail(plt,1))
 
-p1=ggplot(plt, aes(x=Pos,y= reorder(Pop,-Freq.y), color=Pop, size=Pcat))+geom_point(shape=3)+theme_classic()+scale_size(breaks=c(2,3,4), range=c(1,3),labels=c('6-8','8-10','10+'), guide = guide_legend(direction = "horizontal"), name='-log10(p)')+scale_color_manual(breaks= plt$Pop,values = MyColour)+ylab('Populations')+ggtitle('Overlap of SNPs found to be Significant in Quality-GWAS')+theme(legend.position = c(0.85,0.95),legend.box.background = element_rect(colour = "black"), plot.title = element_text(hjust = 0.5), axis.title.x=element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank(), plot.margin = unit(c(1,1,0,1), "cm"))+guides(color=F)
[…]

### 2018-08-27 ManhattanPlot.R

@@ -94,5 +94,21 @@ NoSingles$adjustedP.01 <- adj$TSBH_0.01
 adj.sig.05<-NoSingles[which((NoSingles$adjusted.05 >= -log10(0.05))&(NoSingles$Count > 1)),]
 adj.sig.01<-NoSingles[which((NoSingles$adjusted.01 >= -log10(0.01))&(NoSingles$Count > 1)),]
 
-write.table(adj.sig.05, '~/Documents/QualityPaper/Significant0.5SNPs_adjusted.txt', quote=F, row.names=F)
-write.table(adj.sig.01, '~/Documents/QualityPaper/Significant0.1SNPs_adjusted.txt', quote=F, row.names=F)
+write.table(adj.sig.05, 
+	'~/Documents/QualityPaper/Significant0.5SNPs_adjusted.txt', 
+	quote=F, row.names=F)
+write.table(adj.sig.01, 
+	'~/Documents/QualityPaper/Significant0.1SNPs_adjusted.txt', 
+	quote=F, row.names=F)
+write.table(adj.sig.05$rsID.y, 
+	'~/Documents/QualityPaper/Significant0.5SNPs_adjusted_rsID.txt', 
+	quote=F, row.names=F, col.names=F)
+write.table(adj.sig.01$rsID.y, 
+	'~/Documents/QualityPaper/Significant0.1SNPs_adjusted_rsID.txt', 
+	quote=F, row.names=F, col.names=F)
+write.table(adj.sig.05[,c('Chr','Pos')],
+	'~/Documents/QualityPaper/Significant0.5SNPs_adjusted_POS.txt', 
+	quote=F, row.names=F, col.names=F,  sep='\t')
+write.table(adj.sig.01[,c('Chr','Pos')], 
+	'~/Documents/QualityPaper/Significant0.1SNPs_adjusted_POS.txt', 
+	quote=F, row.names=F, col.names=F, sep='\t')

### 2018-08-27 PlotRegression.R

@@ -36,3 +36,15 @@ ggsave('~/Documents/QualityPaper/RegressionPlot_mostSig2.jpg',height=6,width=12)
 
 weird<-fread('~/Documents/QualityPaper/Misc/LowAFsig_Format.txt')
 ggplot(weird, aes(x=V6, y=V4))+facet_grid(.~V2)+geom_point()+coord_flip()+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme(axis.text.x=element_blank())+labs(x='Genotypes', y='Quality')
+
+
+snp<-read.table('/Users/luke/Documents/QualityPaper/sig/57694260.format.txt',sep=',')
+dev<-read.table('/Users/luke/Documents/QualityPaper/sig/57694260.deviance.txt')
+dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
+dev<-dev[which(dev$p>6),]
+sig<-merge(snp, dev, by.x=c('V1','V2','V5'), by.y=c('V1','V2','V3'))
+sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$V4.y, digits=2))
+
+ggplot(snp, aes(x=V6, y=V4, color=V5))+geom_point(shape=1)+facet_wrap(~V5,nrow=1)+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))+geom_vline(xintercept=30, color='grey60',linetype=3)+coord_flip()+scale_color_manual(breaks= snp$V5,values = MyColour)+guides(color=F)
+
+ggsave('~/Documents/QualityPaper/RegressionPlot.jpg',height=5,width=5)

### 2018-08-29 legacy.tex
updated plots

@@
 %For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
-The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal \sgcomment{How about the others?}.
+The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal observed \sgcomment{How about the others?}.
 For population genetic analyses, the removal of low-mapping quality individuals appears preferable to filtering.
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}
 
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
-There are 2007 and 3664 variants that reach p values greater than $ p < 10^{-8}$ and $ p < 10^{-6}$ respectively.
-Crosses ( + ) represent SNPs that reach a genome wide significance greater than $ p < 10^{-6}$, the circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini & Hochberg FDR adjustment. 
+The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+  \label{Manhattan}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
+
+\caption{Empirical distribution plots of the sum of deviances grouped by N, the number of populations included.
+The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom.
+Note $\sum_i^N  \chi^2_1= \chi^2_N$.
+The grey dotted line indicated the significance level of $10^{-5}$ for the respective distribution which happens to be very close to the adjusted p value threshold at 0.01 FDR.
+}
   \label{Manhattan}

### 2018-08-29 PlotRegression.R

@@ -1,50 +1,0 @@
-library(ggplot2)
-
-#pop names, used to read/write files
-pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
-#read colors per pop (used in 1kG Nature)
-colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
-MyColour <- as.character(colors $hex)
-names(MyColour) <- colors $Pop
-
-snp<-read.table('~/Documents/QualityPaper/31384880.txt',sep=',')
-dev<-read.table('~/Documents/Regression/31384880.dev.txt')
-dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
-dev<-dev[which(dev$p>6),]
-sig<-merge(snp, dev, by.x=c('V1','V2','V5'), by.y=c('V1','V2','V3'))
-sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$V4.y, digits=2))
-
-ggplot(sig, aes(x=V6, y=V4.x))+geom_point(shape=1)+facet_wrap(~Name,ncol=1)+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))
-
-ggsave('~/Documents/QualityPaper/RegressionPlot.jpg',height=5,width=5)
-
-sig6<-read.table('~/Documents/QualityPaper/Significant6SNPs.txt', header=T)
-
-snp<-read.table('~/Documents/QualityPaper/33264017.txt',sep=',', fill=T)
-dev<-read.table('~/Documents/QualityPaper/33264017.dev.txt')
-dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
-dev<-dev[which(dev$p>6),]
-sig<-merge(snp, dev, by.x=c('V2','V5'), by.y=c('V2','V3'))
-sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$p, digits=2))
-sig<-sig[order(sig$p),]
-sig$Name<-factor(sig$Name, levels=unique(sig$Name))
-
[…]

### 2018-08-29 PlottingScripts/Clean_Figure2.R

@@ -38,7 +38,7 @@ scale_colour_manual(values= MyColour, breaks=unique(plt$Group.1))+
 scale_x_discrete(expand=c(0.1,0))+scale_shape_manual(values=c(1,3))+
 guides(color=guide_legend(title="Population",
 override.aes = list(shape = 15, size=3),ncol=1),shape=F)+
-labs(title="Mapping Quality of each Population over Time",x="Populations ranked by Sequencing Date", y="Average quality of mapped bases")+
+labs(title="Data quality over time",x="Populations ranked by sequencing date", y="Average quality of mapped bases")+
 theme(plot.title = element_text(hjust= 0.5), 
 axis.text.x=element_blank(), axis.ticks.x=element_blank())+geom_hline(yintercept=30, linetype=2, color='grey')
 
@@ -48,8 +48,7 @@ guides(color=guide_legend(title="Sequencing\n    Center",
 override.aes = list(shape = 15, size=5)))+
 labs(y="Average quality of mapped bases")+geom_hline(yintercept=30, linetype=2, color='grey')+theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(), axis.line.x=element_blank())
 
-c=ggplot(plt, aes(x= SUBMISSION.DATE, y= INSTRUMENT_MODEL,shape= INSTRUMENT_MODEL))+geom_point()+theme(axis.title.y=element_blank(),axis.line.y=element_blank())+labs(x="Sequencing Date")+ guides(shape=guide_legend(title="",label=F,override.aes = list(alpha = 0))) +
-scale_x_date(date_labels = "%Y",date_breaks = "1 year")+theme(axis.text.y=element_text(size=10,hjust=0))
+c=ggplot(plt, aes(x= SUBMISSION.DATE, y= INSTRUMENT_MODEL,shape= INSTRUMENT_MODEL))+geom_point()+theme(axis.line.y=element_blank())+labs(x="Individuals ranked by sequencing date", y='Sequencer')+ guides(shape=guide_legend(title="",label=F,override.aes = list(alpha = 0))) +scale_x_date(date_labels = "%Y",date_breaks = "1 year")+theme(axis.text.y=element_text(size=10,hjust=0))
 
 d=plot_grid(a,b,c, nrow=3, labels=c('A','B',''),rel_heights=c(4,4,1),align='v')
[…]

### 2018-08-29 PlottingScripts/ManhattanPlot.R

@@ -91,6 +91,35 @@ adj <- as.data.frame(adjusted$adj[order(adjusted$index), ])
 NoSingles$adjusted.01 <- -log10(adj$TSBH_0.01)
 NoSingles$adjustedP.01 <- adj$TSBH_0.01
 
+#make min for dev
+#geom_vline(xintercept=
+#min(adj.sig.01[which(adj.sig.01$Count==df),]$SumDev),
+#color='grey60',linetype=2)+
+#labs(x=paste('df =',df), y='density')+
+
+MakePlot<-function(file){
+	ggplot(file, aes(x=SumDev))+
+				stat_function(aes(x=seq(0,10,
+				length=nrow(file))),fun=dchisq, 
+				args=list(df=df), color='blue')+
+				geom_density()+xlim(c(0,80))+
+				geom_vline(xintercept=
+				qchisq(0.000001, df=df, lower.tail=F),
+				color='grey60',linetype=2)+
+				labs(x='deviance', y='density',subtitle=paste('df =',df))
+}
+
+LP<-list()
+for(df in seq(2,26)){
+	file <- NoSingles[which(NoSingles$Count==df),]
+	LP[[df]] <- MakePlot(file)	
+	}
+
+pf <- grid.arrange(LP[[2]], LP[[3]],LP[[4]],LP[[5]],LP[[6]],LP[[7]],LP[[8]],LP[[9]],LP[[10]], LP[[11]],LP[[12]],LP[[13]],LP[[14]],LP[[15]],LP[[16]],LP[[17]],LP[[18]],LP[[19]],LP[[20]], LP[[21]],LP[[22]],LP[[23]], LP[[24]],LP[[25]],LP[[26]])
+
+ggsave('~/Documents/QualityPaper/Figures/AllDeviances.jpg',pf, height=16,width=22)
+ggsave('~/Documents/QualityPaper/Figures/AllDeviances.tiff',pf, height=16,width=22)
+	
 adj.sig.05<-NoSingles[which((NoSingles$adjusted.05 >= -log10(0.05))&(NoSingles$Count > 1)),]
 adj.sig.01<-NoSingles[which((NoSingles$adjusted.01 >= -log10(0.01))&(NoSingles$Count > 1)),]
 

### 2018-08-30 legacy.tex
minor changes

@@
 Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-\sgcomment{do we excluse singletons?
-We should quantify how much reduction in signal we see.}
+%\sgcomment{do we excluse singletons? We should quantify how much reduction in signal we see.}
 %For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
-The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal observed \sgcomment{How about the others?}.
-For population genetic analyses, the removal of low-mapping quality individuals appears preferable to filtering.
+The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
+For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.
 
@@
 It's clear that there are many covariates that could be used to identify spurious sites like sequencing date, sequencing technology or even the sequencing centre. 
-However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites. \sgcomment{supplementary figures with the general regressions?}
-
+However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
 \begin{figure}
@@
 The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
 \sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

### 2018-08-30 Misc/HapMap+GnomAD.R

@@ -1,0 +1,22 @@
+library(data.table)
+library(ggplot2)
+library(stringr)
+
+GP<-fread('/Users/luke/Documents/QualityPaper/sig/GenomeWide_sig0.01_inHapMap.frq', fill=T)
+HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_Significant0.1SNPs.frq', sep='\t')
+
+HM$AF<-str_split_fixed(HM$V4, ";", 3)[,2]
+HM$AF<-as.numeric(as.character(str_split_fixed(HM$AF, "=", 2)[,2]))
+
+plot<-merge(GP, HM, by.x=c('CHROM','POS'), by.y=c('V1','V2'))
+
+ggplot(plot, aes(x=V6, y=AF))+geom_point()+geom_point(data=plot[which(plot$V3=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
+
+GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq', colClasses=c('numeric','numeric','character','numeric','numeric','numeric','numeric','numeric','numeric','numeric','numeric'))
+
+ggplot(GN, aes(x=reorder(POS,CHROM),y=AF_raw))+geom_point(shape=1)+theme_classic()+theme(axis.text.x=element_blank())
+
+GNm<-melt(GN[,-c('AF_raw')], id=c('CHROM','POS','ID'))
+GNm$value<-as.numeric(as.character(GNm$value))
+GNm<-GNm[which(GNm$value!=0),]
+ggplot(GNm, aes(x=reorder(POS,CHROM),y=value, color=variable))+geom_point(alpha=0.4)+theme_classic()+theme(axis.text.x=element_blank())+labs(y='Frequency',x='Position')

### 2018-09-04 legacy.tex
small style edits

--- 2018-08-27 legacy.tex
@@
 Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
-Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, in essence we would not expect any homozygous alternates for low frequency variants.
 
@@
 Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-\sgcomment{do we excluse singletons?
-We should quantify how much reduction in signal we see.}
 %For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
@@
 
-\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project.
 The earliest sequencing data is used for individuals sequenced more than once.
@@
 The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
 \sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
@@
 
+\subsection{Combined Test}
+%To increase statistical power, we performed a meta-analysis, computing a combined p-value for each position by taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
+%To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
+%Each population will have it's own deviation from the null for each position in the genome.
+%We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
+%This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).
+
 \subsection{Imputation}
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
-In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed.
+In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from low-quality 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality were imputed.
+\sgcomment{All of them?
+How is it possible if we only submitted Japanese?
+Also, I thought that we had only submitted a couple of chromosomes.}
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
-%\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
-%\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
 \sgcomment{It may be useful to impute some of the NAG data.}
@@
 \subsection{Twins Research Human Genetics}
-\sgcomment{HAd we ever discusssed that these GWAS hits were not genome-wide significant?}
+\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
 This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
@@
 Furthermore, the approach they used to impute genotype data for GWAS is very common.
-%A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
+A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
 
@@
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
-\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
-\label{Singles}
-\end{figure}
-
 \end{document}

### 2018-09-04 legacy.tex

--- 2018-08-30 legacy.tex
@@
 Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-%\sgcomment{do we excluse singletons? We should quantify how much reduction in signal we see.}
+
 %For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 

### 2018-09-04 legacy.tex
minor changes to remove errors

@@
 Using an in-house R script, we performed a logistic regression using the glm package.
-\cite{[R and GLM]} We performed this statistical test independently for each population of the 1kGP.
+\cite{R,GLM} We performed this statistical test independently for each population of the 1kGP.
 The deviations from the null model for each test was used to compute the chi-squared p value for each SNP for each population.
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
 \bibliography{Legacy.bib}
 
-\subsection{Supplementary Figures}
+\section{Supplementary Figures}
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
-There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini & Hochberg FDR adjustment. 
+There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}

### 2018-09-05 legacy.tex
minor edits
Updated comments from meeting

@@
 The histogram on the left of the plot is the distribution of significant variants. 
-The dotted line is at 5\% allele frequency. 
 \textbf{C} 
@@
 
-\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
-The earliest sequencing data is used for individuals sequenced more than once.
+\caption{\textbf{A}The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
+Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
 The x-axis is ranked by the mean sequencing date per population.
@@
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
-The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found at least one of the variants to have reached genome wide significance.
+The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
 
@@
 The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.
+luke{Multiple explanations for the signal, technical artifact, cell line artifact.
+The presence of these SNPs in the HapMap suggests a cell line effect, however with only 1\% of sites identified in 1kGP present in HapMap it's not a smoking gun.}
 
@@
 }
-  \label{Manhattan}
+  \label{Deviations}
 \end{figure}

### 2018-09-05 Misc/HapMap+GnomAD.R
updated HapMap + gnomAD SFS

@@ -2,21 +2,27 @@ library(data.table)
 library(ggplot2)
 library(stringr)
 
-GP<-fread('/Users/luke/Documents/QualityPaper/sig/GenomeWide_sig0.01_inHapMap.frq', fill=T)
-HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_Significant0.1SNPs.frq', sep='\t')
+GP<-fread('/Users/luke/Documents/QualityPaper/sig/GenomeWide_sig0.01.frq', skip=1,fill=T,col.names=c('CHROM','POS','count','AC','AF', 'AN'))
+GP$AC1<-round(GP$AC*(1-GP$AF), digits=0)
 
-HM$AF<-str_split_fixed(HM$V4, ";", 3)[,2]
-HM$AF<-as.numeric(as.character(str_split_fixed(HM$AF, "=", 2)[,2]))
+HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_Significant0.1SNPs_format.frq', sep='\t',fill=T, col.names=c('CHROM','POS','rsID','AC','AF', 'AN'))
+HM<-HM[which(complete.cases(HM)==T),]
 
-plot<-merge(GP, HM, by.x=c('CHROM','POS'), by.y=c('V1','V2'))
 
-ggplot(plot, aes(x=V6, y=AF))+geom_point()+geom_point(data=plot[which(plot$V3=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
+plot<-merge(GP, HM, by=c('CHROM','POS'))
 
-GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq', colClasses=c('numeric','numeric','character','numeric','numeric','numeric','numeric','numeric','numeric','numeric','numeric'))
+ggplot(plot, aes(x=1-AF.x, y=AF.y))+geom_point()+geom_point(data=plot[which(plot$rsID=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
 
+GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
[…]

### 2018-09-10 Misc/HapMap+GnomAD.R
half way through Ricks comments

@@ -1,7 +1,7 @@
 library(data.table)
 library(ggplot2)
 library(stringr)
-
+library(cowplot)
 GP<-fread('/Users/luke/Documents/QualityPaper/sig/GenomeWide_sig0.01.frq', skip=1,fill=T,col.names=c('CHROM','POS','count','AC','AF', 'AN'))
 GP$AC1<-round(GP$AC*(1-GP$AF), digits=0)
 
@@ -11,7 +11,7 @@ HM<-HM[which(complete.cases(HM)==T),]
 
 plot<-merge(GP, HM, by=c('CHROM','POS'))
 
-ggplot(plot, aes(x=1-AF.x, y=AF.y))+geom_point()+geom_point(data=plot[which(plot$rsID=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
+A=ggplot(plot, aes(x=1-AF.x, y=AF.y))+geom_point(shape=1)+geom_point(data=plot[which(plot$rsID=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
 
 GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
 GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
@@ -25,4 +25,7 @@ ggplot(GNm, aes(x=reorder(POS,CHROM),y=value, color=variable))+geom_point(alpha=
 
 plt<-merge(GP, GN, by=c('CHROM','POS'))
 
-ggplot(plt, aes(y=AC1, x=AC_raw))+geom_point(shape=1)+labs(x='Allele Frequency in GnomAD', y='Allele Frequency in 1kGP', title='Significant SNPs found in GnomAD and 1kGP')+theme_classic()+xlim(c(0,1000))+ylim(c(0,1000))
+B=ggplot(plt, aes(y=1-AF, x=AF_raw))+geom_point(shape=1)+labs(x='Allele Frequency in GnomAD', y='Allele Frequency in 1kGP', title='Significant SNPs found in GnomAD and 1kGP')+theme_classic()
+C=plot_grid(A,B)
+
+ggsave('/Users/luke/Documents/QualityPaper/Figures/Hap_GnomAD.jpg',height=7,width=11)

### 2018-09-10 PlottingScripts/PC1_Correlation.R

@@ -15,6 +15,7 @@ Meta$SUBMISSION.DATE<-as.Date(Meta$SUBMISSION.DATE, "%Y-%m-%d")
 NamePop<-read.table("~/bin/smaller_mut_spectrum_pipeline/1000genomes_phase3_sample_IDs_BigPop_NAG.txt")
 NamePop<-NamePop[,c(1,2,3)]
 names(NamePop)<-c('Name', 'BigPop', 'Pop')
+
 ##Name of OutLiers
 CutOffIDs<-read.table("/Users/luke/Documents/MutSpect/OutputFiles/PCA_cutoff/CutOffID.txt")
 CutOffIDs$Out<-'OutLiers'
@@ -31,59 +32,42 @@ colnames(count)[1]<- "Mut_type"
 tot.count <- sapply(seq(2,ncol(count),2), function(i) {
   rowMeans(count[,c(i, i+1)], na.rm=T)})
   
-colnames(tot.count)<-NamePop$Pop #adds the POP name to the column
+colnames(tot.count)<-NamePop$BigPop #adds the POP name to the column
 
 tot.count <- prop.table(as.matrix(tot.count),2)
 tot.count <- tot.count[ , apply(tot.count, 2, var) != 0]
 
-pop.count<-tot.count[, grep('JPT', colnames(tot.count))]
-colnames(pop.count)<-NamePop[which(NamePop$Pop == 'JPT'),]$Name
+pop.count<-tot.count[, grep('EAS', colnames(tot.count))]
+
+EAS <- c('JPT','CHS','CHB','CDX','KHV')
+colnames(pop.count)<-NamePop[which(NamePop$BigPop == 'EAS'),]$Name
+pop.count<-pop.count[, -grep('NAG', colnames(pop.count))]
+
+
 pop.pca = prcomp(t(pop.count), scale=T)
 pop.pca.scores = as.data.frame(pop.pca$x)[,1:5]
 pop.pca.scores$Name<-rownames(pop.pca.scores)
 
 plot<-merge(pop.pca.scores, Meta, by='Name')
-									
-									
-r1=lm(plot$average_quality_of_mapped_bases ~ plot$PC1, plot)
-t1=paste("Adj R2 = ",signif(summary(r1)$adj.r.squared, 3),
-                     " P =",signif(summary(r1)$coef[2,4], 3))
+plot$Out<-gsub('OutLiers', 'Outliers',plot$Out)
 
[…]

### 2018-09-10 legacy.tex
Finished implementing Ricks comments

@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Hapmap \cite{HapMap2005}, the 1000 Genomes Project \cite{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \cite{Mallick2016}, for example, have made thousands of genomes publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \cite{ProfJonathanMarchiniProfGoncaloAbecasisProfRichardDurbin2014} or summary statistics for variant frequency estimation \cite{Lek2016}.
+The International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
 
-Because of the extraordinary value of freely available data, this early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-However, the first sequenced genomes in the 1000 Genomes project (1kGP), for example, were sequenced 10 years ago.
+Because of the extraordinary value of freely available data, this early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+However, the first sequenced genomes in the 1kGP, for example, were sequenced 10 years ago \citep{VanDijk2014}.
 Since then, sequencing platforms have rapidly improved.
-The second phase of the 1kGP describe multiple technological and analytical improvements over its earlier phases \cite{}.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}.
 This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
@@
 
-Different mutagenic processes preferentially affect different DNA motifs.
-Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \cite{Pleasance2010}.
-Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of `signatures' of different mutational processes\cite{Pleasance2010}.
-Such genome wide mutational signatures have been used as a diagnostic tool for cancers \cite{Pleasance2010,Shiraishi2015a}.
+Some mutagenic processes will affect different DNA motifs. 
+Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
+Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
+Such genome wide mutational signatures have been used as a diagnostic tool for cancers \citep{Alexandrov2013,Shiraishi2015a}.
 
-%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
+%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
 %Large cohorts have recently become available to increase statistical power to resolve some of these finer scale mutational signatures.
@@
 Differences in mutational signatures across human populations have recently been reported.
-In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \cite{Harris2015a}.
-
-Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \cite{Harris2015a}. 
-Harris and Pritchard further identified distinct mutational spectra across a range of populations \cite{Harris2017a}, which were further studied in \cite{Voight}
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
+Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
+Harris and Pritchard further identified distinct mutational spectra across a range of populations \citep{Harris2017a}, which were further studied in \citep{Aikens2018}
 
@@
 
-The results section is organized as follows:
+The results section is organized as follows.
 First, we describe the observed discrepancies within and between studies.
@@
 Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
-Even though these SNPs are likely technical artifacts, we two recent studies report such SNPs as genome-wide significant hits.
-
-Fortunately, the main conclusions of most of the affected studies remain supported by other data. Yet this begs the question: When should we retire legacy reference data?
+Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
+Fortunately, the main conclusions of the affected studies remain supported by other data. 
+Yet this begs the question: When should we retire legacy reference data?
 
@@
 \textbf{A} 
-Mutation spectrum of the 1034 variants that reached a genome wide significance with a p value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
-There majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern.
-There is also an slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
+Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
+The majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern.
+There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
 These three enrichments can be summarized as G**${\rightarrow}$GG*.
@@
 
-Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
+Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\citep{Harris2015a}
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
-Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, in essence we would not expect any homozygous alternates for low frequency variants.
+Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants.
 
@@
 By 2011, older sequencing technologies are phased out, and method development solidify, which coincides with the sequencing quality levelling off.
-It's clear that there are many covariates that could be used to identify spurious sites like sequencing date, sequencing technology or even the sequencing centre. 
+It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
 However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
@@
 The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{meta-analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
 \sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
-In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed.
+In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
+We found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. 
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. 
@@
 
-\subsection{Twins Research Human Genetics}
-\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
+%\subsection{Twins Research Human Genetics}
+%\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
 This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
@@
 The data used was imputed genotype data using Hapmap II as the reference.
-One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality.\cite{}
+One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
 
@@
 
-\subsection{PLos One}
-This paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the coverage of mapped reads to EBV.
+%\subsection{PLos One}
+This paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
 They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
@@
 However they reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality. \cite{}
+Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality \citep{Mandage2017}.
 
@@
 This is a commonly used transformation that helps preserve cell lines for long periods of time.
-It's unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that perhaps different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
+It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that perhaps different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
 Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics.
@@
 \sgcomment{needs some editing}
-The variants identified in this study are likely to be technical artifacts from legacy technology.
+The variants identified in this study are likely to be technical artifacts from legacy technology.\citep{Minoche2011} \luke{<-citation} 
 If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
@@
 \sgcomment{needs some editing}
-The most conservative approach would be to remove all individuals that don't meet the quality threshold as well as all the variants associated to low quality.
-In this case, we used a cut off of an average quality of mapped bases over 30. 
-It is the minimum requirements for \todo{GATK variant calling} for them to have a minimum quality of 30.
+The most conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
+In this case, we used a cut-off of an average quality of mapped bases over 30.
 Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
@@
 
-%Many types of DNA sequencing machines have been available on the market over the past decade.
-%Different types of sequencing technologies can have different technical limitations in quality control.
-%This makes combining data an issue especially when multiple sequencing technologies are involved in the data production.
-%Many of the legacy datasets produced using dated sequencing technologies have been known to contain a higher rate of false positives than their more recent counterparts.
-%The errors in one dataset do not disappear when they are combined with another higher quality one.
-%The technical biases caused by legacy sequencing technology is becoming increasingly relevant as newer technologies produce data with lower error rates, and more stringent quality controls.
-
 \section{Methods}
@@
 al as they applied the current state of the art quality thresholds to remove questionable sequences especially for the high standards for detecting population level differences.
-Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or basepair complexity compared to the rest of the genome. 
+Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
 The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
@@
 \subsection{Genome Wide Association Study}
-Using an in-house R script, we performed a logistic regression using the glm package.
-\cite{R,GLM} We performed this statistical test independently for each population of the 1kGP.
-The deviations from the null model for each test was used to compute the chi-squared p value for each SNP for each population.
-Since each deviation follows a chi-squared distribution, the sum of the deviations also follow a chi-squared distribution.
-However, since not all SNPs are present in all populations, when computing the p values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.
+Using a custom-made R script, we performed a logistic regression using the glm package \citep{RDevelopmentCoreTeam2016}.
+We ran the regression independently on data for each genomic locus and population independently for each of the 26 populations of the 1kGP.
+For each of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero.
+$H_{0}: \beta_{1}=0$.  
+To test the null-hypothesis we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal important of the quality variable. 
+The deviance test statistics is approximately chi-square distributed with one degrees of freedom.  
+The deviations from the null model for each test was used to compute the chi-squared \textit{p}-value for each SNP for each population.
+To test whether the quality variable is associated with genotype across all populations, we used the sum of the deviances across the populations.
+However, since not all SNPs are present in all populations, when computing the \textit{p}-values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.
 
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. \cite{Harris2017a}
+We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}
 We modified the methods as necessary for our purposes, scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
@@
 Note $\sum_i^N  \chi^2_1= \chi^2_N$.
-The grey dotted line indicated the significance level of $10^{-5}$ for the respective distribution which happens to be very close to the adjusted p value threshold at 0.01 FDR.
+The grey dotted line indicates the significance level of $10^{-5}$ for the respective distribution which happens to be very close to the adjusted \textit{p}-value threshold at 0.01 FDR.
 }
-  \label{Deviations}
+\label{Deviances}
 \end{figure}

### 2018-09-11 legacy.tex
updated discussion, removed 'reverse GWAS'

@@
 
-%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
-%Large cohorts have recently become available to increase statistical power to resolve some of these finer scale mutational signatures.
-
+While certain mutagens have very distinct mutational signatures, others are more subtle and require much larger datasets to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
 Differences in mutational signatures across human populations have recently been reported.
@@
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
 Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, we are looking for loci where the genotypes are predictable based on the data quality scores.
@@
 Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-
-%For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
 The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
@@
 \subsection{Overlap of significant SNPs}
-The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{meta-analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
+The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
 \sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
@@
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. 
-%\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
-%\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
-\sgcomment{It may be useful to impute some of the NAG data.}
+%\sgcomment{It may be useful to impute some of the NAG data.}
 
@@
 One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
-\sgcomment{If they used HapMap, how did they get affected by the 1000 Genomes results?}
 The other study used the 1kGP sequence data and cell lines directly.
@@
 
-%\subsection{Twins Research Human Genetics}
-%\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
-This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
+One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
 While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
@@
 One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
+While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
+Finding some of these SNPs in imputed HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
+We found that roughly 1\% of the SNPs associated to quality were found in HapMap,  rs6057648 is among those found in HapMap.
 
-While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified. 
-Furthermore, the approach they used to impute genotype data for GWAS is very common.
-%A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}
-
-%\subsection{PLos One}
-This paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
+Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
 They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
@@
 \subsection{Technical Artifact}
-\sgcomment{needs some editing}
-The variants identified in this study are likely to be technical artifacts from legacy technology.\citep{Minoche2011} \luke{<-citation} 
+\luke{edited}
+The variants identified in this study are likely to be technical artifacts from legacy technology.
+Different sequencing technologies will have different error profiles. 
+A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}.
+The quality of sequencing data decreases along the length of the read.
+These regions of low quality at the ends of reads are know as B-tails.
+The same report found that the GAII also had much higher rates of reads with B-tails.
+However, many variant calling algorithms will discard reads with high error rates \citep{Minoche2011}.
+There are indeed multiple explanations for the signal. 
+The presence of these SNPs in the HapMap suggests a cell line effect, however this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
+Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
+
 If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
@@
 The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.
-luke{Multiple explanations for the signal, technical artifact, cell line artifact.
-The presence of these SNPs in the HapMap suggests a cell line effect, however with only 1\% of sites identified in 1kGP present in HapMap it's not a smoking gun.}
 
 \subsection{Recommendations}
-\sgcomment{needs some editing}
-The most conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
+\luke{edited}
+While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
+A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
 In this case, we used a cut-off of an average quality of mapped bases over 30.
@@
 (See supplementary material for lists of SNPs)
-%\subsection{Power limits and Quality Controls}
-%If these variants are truly spurious, why were they not identified by the Hardy-Weinberg test of heterozygosity?
-%how many hetd do you need to detect HWE significance?
-%We were able to identify many variants with frequencies below \todo{the power threshold for HWE}.
-
-%\subsection{Technical Source of Error}
-%B tail trimming  (Minoche et al. Genome Biology 2011)
-%issues with PCR?
-%repetitive regions, GC bias...
 
@@
 On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before. 
-The reverse GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.
+The GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.
 
 More generally, to improve the quality of genomic reference datasets, we can proceed by addition of new and better data and by better curation of existing data. 
-Given rapid technological progress, the focus of genomic research is naturally on the data generation side.
-However, cleaning up is also important to avoid generating spurious results.
+Given rapid technological progress, the focus of genomic research is naturally on the data generation side. However, cleaning up is also important to avoid generating spurious results. 
 The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing.
@@
 Given that modern platforms seem to have resolved this technical issue but that recent studies continue being affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while the sequencing forensics investigation can take place.
-
-%but evidence points to the fact that the artifact disappeared in the later stages of the 1kGP, and does not corrupt more recent data.
-
-%As more large scale genotyping efforts are being imputed on the same legacy datasets, we must scrutinize the quality of the reference databases to avoid the propagation of false positives. 
-%These results bring forth many questions regarding the reliability of legacy datasets. 
-%Moreover, since there are so many broad applications of imputation, it frames the question for reference data turnover.
-
-%Our method identifies spurious mutations by correlating mutations with data quality metrics. 
-%We propose including our quality control methods to identify possible false positives in sequencing data. 
-%We have focused on the 1000 Genomes Project dataset as its quality metrics were freely available, however the issues of quality control are not limited to this one consortium. 
-%This study only used one dataset and one quality metric, but using this same approach can be used to identify more false positives in many more datasets.
 
@@
 
-%\bibliographystyle{plain}
 \bibliography{Legacy.bib}

### 2018-09-12 legacy.tex
suggestions

@@
 \newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
+
+%\newcommand{\qq}{m}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Different sequencing technologies will have different error profiles. 
-A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}.
-The quality of sequencing data decreases along the length of the read.
-These regions of low quality at the ends of reads are know as B-tails.
-The same report found that the GAII also had much higher rates of reads with B-tails.
-However, many variant calling algorithms will discard reads with high error rates \citep{Minoche2011}.
-There are indeed multiple explanations for the signal. 
-The presence of these SNPs in the HapMap suggests a cell line effect, however this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
+A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}, with e.g.
+different patterns of quality decrease along reads (known as B-tails).
+Differences in read quality in turn require for different calling pipelines.
+
+We were not able to pinpoint the precise technical source of the discrepancy.
+Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced with an entirely different platform.
+\sgcomment{Check that is sanger, if so, say so}.
+We find that \sgcomment{$99\%$--check} of the suspicious snps are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
+
+The presence of some of these SNPs in HapMap suggests a cell line effect, however this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
 Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
@@
 \subsection{Genome Wide Association Study}
-Using a custom-made R script, we performed a logistic regression using the glm package \citep{RDevelopmentCoreTeam2016}.
-We ran the regression independently on data for each genomic locus and population independently for each of the 26 populations of the 1kGP.
+We performed logistic regression using the R glm package \citep{RDevelopmentCoreTeam2016}.
+We ran the regression independently for each genomic locus and each of the 26 populations of the 1kGP.
 For each of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero.
-$H_{0}: \beta_{1}=0$.  
-To test the null-hypothesis we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal important of the quality variable. 
-The deviance test statistics is approximately chi-square distributed with one degrees of freedom.  
+\sgcomment{define $\beta_1$ or not mention?
+$H_{0}: \beta_{1}=0$  }.  
+To test the null hypothesis, we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of the average quality of mapped bases. 
+The deviance test statistic is approximately chi-square distributed with one degrees of freedom.  
 The deviations from the null model for each test was used to compute the chi-squared \textit{p}-value for each SNP for each population.

### 2018-09-13 legacy.tex
modified model specs

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Legacy Data Confounds Modern Genomics Studies}
+%\title{Legacy Data Confounds Modern Genomics Studies}
+\title{Batch Effects in the 1000 Genomes Project Data Confounds Modern Genomics Studies}
 
@@
 
-\subsection{Reference Cohorts}
+\subsection{Batch Effects in Aging Reference Cohort Data}
 
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-The International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
 Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
@@
 Because of the extraordinary value of freely available data, this early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-However, the first sequenced genomes in the 1kGP, for example, were sequenced 10 years ago \citep{VanDijk2014}.
-Since then, sequencing platforms have rapidly improved.
+However, the first sequenced genomes in the 1kGP, for example, were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
 The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}. 
@@
 
-\subsection{Motivations}
+\subsection{Mutational Signatures}
 
@@
 
+ \subsection{Irreproducibility of a Japanese Mutational Signature}
+
  In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
@@
 We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.
-
-The results section is organized as follows.
-First, we describe the observed discrepancies within and between studies.
-We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.
-Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
-Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
-Fortunately, the main conclusions of the affected studies remain supported by other data. 
-Yet this begs the question: When should we retire legacy reference data?
 
@@
 			\section{Results}
+The results section is organized as follows.
+First, we describe the observed discrepancies within and between studies.
+We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.
+Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
+Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
+Fortunately, the main conclusions of the affected studies remain supported by other data. 
+Yet this begs the question: When should we retire legacy reference data?
+
 	\subsection{A peculiar mutational signature in Japan}
@@
 \subsection{Overlap of significant SNPs}
+%Define the approach before calling a genome wide simple logistic regression analysis
 The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
 Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
-\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
+%Wanted a test statistic to represent the association across all populations, to do that we summed across the deviations for each population. 
+%
+%\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
 
@@
 
+\subsection{Model Specification}
+
+${logit}E(Y_{si}) = \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{i} $
+
+for $i = 1,\hdots, n_p$
+
+where ${Q}$ is the average quality of mapped bases, ${i}$ is the individual, ${s}$ is SNP, ${p}$ is the population.
+
 \subsection{Mutation Spectrum}
@@
 
+\clearpage
 \section{Supplementary Figures}
+
 \renewcommand{\thefigure}{S\arabic{figure}}

### 2018-09-13 legacy.tex
updated methods section with Rick

@@
 This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
+Yet this begs the question: When should we retire legacy reference data?
 
@@
 Harris and Pritchard further identified distinct mutational spectra across a range of populations \citep{Harris2017a}, which were further studied in \citep{Aikens2018}
-
- \subsection{Irreproducibility of a Japanese Mutational Signature}
 
@@
 The results section is organized as follows.
-First, we describe the observed discrepancies within and between studies.
-We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.
-Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
-Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
-Fortunately, the main conclusions of the affected studies remain supported by other data. 
-Yet this begs the question: When should we retire legacy reference data?
+First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs from the whose validity is highly debatable.
+Finally, we investigate how these SNPs have impacted modern genomics analyses. 
+We report that major imputation servers impute the dubious SNPs. 
+We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.
 
@@
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome-wide association study (GWAS) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
-Whereas the majority of GWAS use linear regression predicting a phenotype based on genotype, we are looking for loci where the genotypes are predictable based on the data quality scores.
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome wide simple logistic regression analysis (GWLR) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
+Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
+In this case, we are looking for loci where the genotypes are predictable based on the data quality scores.
 This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
@@
 
-Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+Thus this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
 The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
@@
 
-	\subsection{Sequencing quality over time}
-The results of the logistic regression GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+	\subsection{Identifying Suspicious SNPs in the 1kGP}
+The results of the GWLR performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
 The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
@@
 However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
+
 \begin{figure}
@@
 
-\subsection{Overlap of significant SNPs}
-%Define the approach before calling a genome wide simple logistic regression analysis
 The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
-%Wanted a test statistic to represent the association across all populations, to do that we summed across the deviations for each population. 
-%
-%\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
+We wanted a test statistic to represent the association across all populations, to do that we summed across the deviations from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
 
@@
 
-\subsection{Imputation}
+\subsection{Suspicious SNPs impact modern genomics analyses}
 The state of the art imputation servers use a combination of many databases including many that are not freely available.
@@
 This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
-%\sgcomment{It may be useful to impute some of the NAG data.}
 
-\subsection{Found to be included in other GWAS }
 Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
@@
 
-\subsection{Genome Wide Association Study}
-We performed logistic regression using the R glm package \citep{RDevelopmentCoreTeam2016}.
-We ran the regression independently for each genomic locus and each of the 26 populations of the 1kGP.
-For each of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero.
-\sgcomment{define $\beta_1$ or not mention?
-$H_{0}: \beta_{1}=0$  }.  
+\subsection{Genome wide simple logistic regression analysis}
+We fitted a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
+The simple logistic model can be expressed as such.
+
+\begin{align*}
+{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
+\\
+\textit{for}\quad i &= 1,\hdots, n_{p'}
+\end{align*}
+where $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$.
+In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$.
+We run a total of $K (= \sum_{s=1}^S P_s)$ regressions where $ P_s$, ranging from 2 to 26, is the number of populations with a segregating variant for genomic locus $s$.
+
+For each combination of of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero.
+$H_{0}: \beta_{1}=0$.
 To test the null hypothesis, we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of the average quality of mapped bases. 
@@
 However, since not all SNPs are present in all populations, when computing the \textit{p}-values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.
-
-\subsection{Model Specification}
-
-${logit}E(Y_{si}) = \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{i} $
-
-for $i = 1,\hdots, n_p$
-
-where ${Q}$ is the average quality of mapped bases, ${i}$ is the individual, ${s}$ is SNP, ${p}$ is the population.
 

### 2018-09-13 legacy.tex
finalized stats methods with Rick, and added reference

@@
 Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
-We wanted a test statistic to represent the association across all populations, to do that we summed across the deviations from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
+We wanted a test statistic to represent the association across all populations, to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
 
@@
 \subsection{Genome wide simple logistic regression analysis}
-We fitted a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
-The simple logistic model can be expressed as such.
-
+We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
+The model we fitted to each subset of data can be expressed as such.
 \begin{align*}
@@
 \end{align*}
-where $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$.
+Here $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
 In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$.
-We run a total of $K (= \sum_{s=1}^S P_s)$ regressions where $ P_s$, ranging from 2 to 26, is the number of populations with a segregating variant for genomic locus $s$.
+We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the \todo{total number of genomic loci} , and $P_s$ is the number of populations with a segregating variant for locus $s$.
+In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.
 
-For each combination of of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero.
-$H_{0}: \beta_{1}=0$.
-To test the null hypothesis, we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of the average quality of mapped bases. 
+For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{1}^{(sp)}=0$. That is, the average quality of mapped bases $Q_{p}$ variable is not associated with genotype $Y_{sp}$.
+In other words, the slope coefficient of the simple logistic regression is zero.
+To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q_{p}$ in the model. 
 The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
-The deviations from the null model for each test was used to compute the chi-squared \textit{p}-value for each SNP for each population.
-To test whether the quality variable is associated with genotype across all populations, we used the sum of the deviances across the populations.
-However, since not all SNPs are present in all populations, when computing the \textit{p}-values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.
+
+To obtain a population-wide measure of association for each locus across its respective $P_s$ populations, we summed the $P_s$ deviances for each $s$.
+The resulting test statistic $T_s$ is chi-squared distributed with $P_s$ degrees of freedom (i.e.
+$T_s \sim \chi^2_{P_s}$), see Figure \ref{Deviances}.
+The observed summed deviances were used to compute the \textit{p}-value for each locus.
+
+Given the large number of \textit{p}-values of which there are $S$ of them, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values
+\citep{Benjamini2006}.
+By using a very conservative nominal Type-I error rate $\alpha = 0.01$, a total of 3990 loci were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 
@@
 
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
 There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 

### 2018-09-14 legacy.tex
minor grammatical corrections and changes

@@
 %\title{Legacy Data Confounds Modern Genomics Studies}
-\title{Batch Effects in the 1000 Genomes Project Data Confounds Modern Genomics Studies}
+\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}
 
 \author[1,2]{Luke Anderson-Trocm\'e}
+\author[1,2]{Rick Farouni}
 \author[1,2]{Mathieu Bourgey}
-\author[1,2]{Rick Farouni}
 \author[3]{Fumihiko Matsuda}
@@
 This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
-Yet this begs the question: When should we retire legacy reference data?
+This begs the question: when should we retire legacy reference data?
 
-Even though it is intuitively clear that such batch effects have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data makes such issues particularly insidious.
+Even though it is intuitively clear that such batch effects have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Here we point out how large and previously unreported batch effects in the early phases of the 1kGP that still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
 
@@
 Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
-Such genome wide mutational signatures have been used as a diagnostic tool for cancers \citep{Alexandrov2013,Shiraishi2015a}.
+Such genome wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
 
@@
 Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
-Harris and Pritchard further identified distinct mutational spectra across a range of populations \citep{Harris2017a}, which were further studied in \citep{Aikens2018}
+Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.
 
@@
 This heterogeneity is intriguing because differences in mutational signatures accumulate over many generations.
-A systematic difference among the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
+A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
 We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort.}
 
-While we were unable to reproduce the heterogeneity among the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
+While we were unable to reproduce the heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
 We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.
@@
 			\section{Results}
-The results section is organized as follows.
+%The results section is organized as follows.
 First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs from the whose validity is highly debatable.
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
 Finally, we investigate how these SNPs have impacted modern genomics analyses. 
@@
 
-Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\citep{Harris2015a}
+Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a}.
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 Individuals with low mapping quality show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
-Thus sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
@@
 
-Thus this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+Thus, this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
 The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
@@
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
-By 2011, older sequencing technologies are phased out, and method development solidify, which coincides with the sequencing quality levelling off.
+By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
 It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
+\includegraphics[width=0.95\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
 
@@
 
-The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
+The GWAS approach similarly identified mapping-quality-associated SNPs in 24 of the 26 populations in the 1kGP.
 Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
-We wanted a test statistic to represent the association across all populations, to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
+We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
 
@@
 They also removed rare alleles and alleles with high degrees of missingness. 
-Despite using the state of the art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
+Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
 
 \section{Discussion}
-
 One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
-While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
+While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
 The data used was imputed genotype data using Hapmap II as the reference.
@@
 While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
-Finding some of these SNPs in imputed HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
-We found that roughly 1\% of the SNPs associated to quality were found in HapMap,  rs6057648 is among those found in HapMap.
+Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
+We found that roughly 1\% of the SNPs associated to quality were found in HapMap; rs6057648 is among those found in HapMap.
 
 Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
-They did PCR confirmation of EBV copy number using 7 samples from some of the populations, however all of these samples have relatively high quality data.
+They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high average quality of mapped bases.
 They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
-However they reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) we're identified by our regression against data quality \citep{Mandage2017}.
+They also reported the SNPs reading significance below $ p < 10^{-6}$. 
+Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against data quality \citep{Mandage2017}.
 
-The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphobastoid cell lines.
+The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphoblastoid cell lines.
 This is a commonly used transformation that helps preserve cell lines for long periods of time.
-It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that perhaps different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
+It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
 Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics.
@@
 \subsection{Technical Artifact}
-\luke{edited}
 The variants identified in this study are likely to be technical artifacts from legacy technology.
 Different sequencing technologies will have different error profiles. 
-A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}, with e.g.
-different patterns of quality decrease along reads (known as B-tails).
-Differences in read quality in turn require for different calling pipelines.
+A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}, with, for instance, different patterns of quality decrease along reads (known as B-tails).
+Differences in read quality and error profiles in turn require different calling pipelines.
 
@@
 
-The presence of some of these SNPs in HapMap suggests a cell line effect, however this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
+The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
 Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
 
-If these sequencing platforms are no longer being used then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
+If these sequencing platforms are no longer being used, then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
 However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
 We have shown that these variants have been significantly associated using traditional GWAS methods.
-However this is not the only method that would be affected by these false positives. 
+However, this is not the only method that would be affected by these false positives. 
 Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
@@
 \subsection{Recommendations}
-\luke{edited}
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
@@
 On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before. 
-The GWAS approach is rather straightforward, and should probably be a standard in a variety of -omics studies.
+The genome wide simple logistic regression analysis approach is rather straightforward, and could be a standard in a variety of -omics studies.
 
 More generally, to improve the quality of genomic reference datasets, we can proceed by addition of new and better data and by better curation of existing data.
-Given rapid technological progress, the focus of genomic research is naturally on the data generation side. However, cleaning up is also important to avoid generating spurious results. 
+Given rapid technological progress, the focus of genomic research is naturally on the data generation side. However, cleaning up data is also important to avoid generating spurious results. 
 The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing.
 
-Somewhat embarrassingly, we have not identified the precise mechanism through which the technical artifact was introduced: we have ruled out neither cell-line, sequencer, nor algorithmic artifact. 
-Given that modern platforms seem to have resolved this technical issue but that recent studies continue being affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while the sequencing forensics investigation can take place.
+We have not identified the precise mechanism through which the technical artifact was introduced: we have ruled out neither cell-line, sequencer, nor algorithmic artifact. 
+Given that modern platforms seem to have resolved this technical issue but that recent studies continue to be affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while the sequencing forensics investigation can take place.
 
@@
 \end{align*}
-Here $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
+Here $Y_{spi}$ is a Bernoulli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
 In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$.

### 2018-09-19 legacy.tex
changed 'GWAS' and 'Q'

@@
 
+The results section is organized as follows.
+First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
+Finally, we investigate how these SNPs have impacted modern genomics analyses. 
+We report that major imputation servers impute the dubious SNPs. 
+We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.
+
 \begin{figure}
@@
 Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
-The majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern.
+The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern.
 There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
@@
 \textbf{C} 
-Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1kGP.
-This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
+Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1kGP.
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
 The same analysis was performed independently for each of the populations in the 1kGP. }
@@
 			\section{Results}
-%The results section is organized as follows.
-First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
-Finally, we investigate how these SNPs have impacted modern genomics analyses. 
-We report that major imputation servers impute the dubious SNPs. 
-We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.
 
@@
 When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
-Individuals with low mapping quality show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that $Q$ per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
+Individuals with low $Q$ show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
 Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome wide simple logistic regression analysis (GWLR) in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that correlate strongly with low $Q$ (Figure \ref{SFS}).
 Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
-In this case, we are looking for loci where the genotypes are predictable based on the data quality scores.
+In this case, we are using simple logistic regression looking for loci where the genotypes are predictable based on the data quality scores.
 This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
-These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.
+The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
 
-Thus, this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
-The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
-For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.
+Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques.
+The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
+For population genetic analyses, the removal of individuals with low $Q$ appears preferable to filtering positions.
 
 	\subsection{Identifying Suspicious SNPs in the 1kGP}
-The results of the GWLR performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
-The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
+The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
 This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
@@
 It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
-However, the average quality of mapped bases per individual appeared to be the best metric to identify spurious sites.
+However, the $Q$ per individual appeared to be the best metric to identify spurious sites.
 
@@
 
-\caption{\textbf{A}The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A}The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
 Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
@@
 
-The GWAS approach similarly identified mapping-quality-associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
+The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
 We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
@@
 
-\caption{Overlap of SNPs identified independently to be associated with quality. 
+\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most low-quality individuals also tend to have the most variants associated to quality. 
+The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
 The same variants identified as being low quality independently in each population are found in other populations. }
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
-In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
-We found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. 
-This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed.
+In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
+We found that all of the SNPs we identified as being associated with $Q$ for chromosomes 1 and 2 were imputed. 
+This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants have not been removed.
 
-Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
+Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
 The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
@@
 The data used was imputed genotype data using Hapmap II as the reference.
-One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
+One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ \citep{Ebejer2013}.
 While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
 Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
-We found that roughly 1\% of the SNPs associated to quality were found in HapMap; rs6057648 is among those found in HapMap.
+We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.
 
 Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
-They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high average quality of mapped bases.
+They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high $Q$.
 They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
 They also reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against data quality \citep{Mandage2017}.
+Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$ \citep{Mandage2017}.
 
@@
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
-A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
-In this case, we used a cut-off of an average quality of mapped bases over 30.
+A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
+In this case, we used a cut-off of $Q$ over 30.
 Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
@@
 The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
-Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
+Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file. 
 Each BAS file has metadata regarding each sequencing event for each sample. 
-If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
+If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
 The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
-This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
+This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the $Q$ scores and only kept the earliest sequencing date per sample.
 The dates of the sequencing are only used to plot Figure.
@@
 
-\subsection{Genome wide simple logistic regression analysis}
+\subsection{Genome wide association study}
 We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
-\caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\caption{Regressions against the average quality per mapped bases $Q$ per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
 }
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
-\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to quality.}
+\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
 \label{MostSig}
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
-\caption{Logistic regression of rs6057648 for the populations with significant association to quality.}
+\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
 \label{TwinsSNP}

### 2018-09-19 legacy.tex
updated total number of SNPs

@@
 Furthermore, only diallelic autosomal SNPs were considered, with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9.
+In total, 59,272,030 SNPs passed our quality controls.
 
@@
 In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$.
-We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the \todo{total number of genomic loci} , and $P_s$ is the number of populations with a segregating variant for locus $s$.
+We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the total number of genomic loci, and $P_s$ is the number of populations with a segregating variant for locus $s$.
 In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.

### 2018-09-20 legacy.tex
updated text

@@
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that correlate strongly with low $Q$ (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
 Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
@@
 
+The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
+We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.
+
 \begin{figure}
@@
 \end{figure}
-
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
 
@@
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
+Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait.
 The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
@@
 
-On a technical front, we were surprised that strong correlations between SNPs and technical covariates in the 1kGP project had not been identified before. 
+On a technical front, we were surprised that strong association between SNPs and technical covariates in the 1kGP project had not been identified before. 
 The genome wide simple logistic regression analysis approach is rather straightforward, and could be a standard in a variety of -omics studies.
@@
 The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
-Furthermore, only diallelic autosomal SNPs were considered, with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9.
-In total, 59,272,030 SNPs passed our quality controls.
+Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 we're considered.
+In total, 57,838,849 diallelic autosomal SNPs passed our quality controls for the mutation spectrum analysis and reverse GWAS. 
+We also ran the same reverse GWAS separately on 2,952,684 insertions and deletions that passed the same quality controls.
 
@@
 We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the total number of genomic loci, and $P_s$ is the number of populations with a segregating variant for locus $s$. 
-In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.
+In the data, $P_s$ ranges from 1 to 26 since not all SNPs were present in all populations with singletons and doubletons filtered out.
 
@@
 
-Given the large number of \textit{p}-values of which there are $S$ of them, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values
-\citep{Benjamini2006}.
-By using a very conservative nominal Type-I error rate $\alpha = 0.01$, a total of 3990 loci were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 3990 SNPs and 1134 indels were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 
@@
 \begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}
+
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
+There are 1134 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+  \label{indels_Manhattan}
+\end{figure}
+
+\begin{figure}
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
-\caption{Regressions against the average quality per mapped bases $Q$ per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\caption{Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
 }

### 2018-09-20 legacy.tex
edits throughout

--- 2018-09-19 legacy.tex
@@
 While certain mutagens have very distinct mutational signatures, others are more subtle and require much larger datasets to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
+\sgcomment{Missing logical link}
 Differences in mutational signatures across human populations have recently been reported.
-In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations; this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
 Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
@@
 
-While we were unable to reproduce the heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
-We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.
-
-The results section is organized as follows.
-First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
-Finally, we investigate how these SNPs have impacted modern genomics analyses. 
-We report that major imputation servers impute the dubious SNPs. 
-We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.
+While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
+We also found that this batch effect led to spurious imputation results which found their way in recent publications.
 
@@
 Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
-The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern.
+The majority of the variants with significant associations to quality have the *AC${\rightarrow}$*CC mutational pattern.
 There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
@@
 \textbf{C} 
-Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1kGP.
-This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
+Genome wide association of the average quality of mapped bases for the 104 Japanese individuals included in the 1kGP.
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average quality of SNPs mapped for an individual
 The same analysis was performed independently for each of the populations in the 1kGP. }
@@
 \section{Results}
+We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of suspicious SNPs.
+Finally, we investigate how these SNPs have impacted modern genomics analyses.
 
@@
 
-Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a}.
+Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a} \sgcomment{That would be 2017, no?}.
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
-Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
+These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
 Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants.
+\sgcomment{this sentence needs work}
 
 When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-Regressions against different individual-level quality metrics provided by the 1kGP revealed that $Q$ per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
-Individuals with low $Q$ show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that mean quality per mapped base pair per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
+Individuals with low mapping quality show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
 Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that correlate strongly with low $Q$ (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a genome wide simple logistic regression analysis (GWLR) \sgcomment{I don't like this name} in the JPT for SNPs that correlate strongly with low quality (Figure \ref{SFS}).
 Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
-In this case, we are using simple logistic regression looking for loci where the genotypes are predictable based on the data quality scores.
+In this case, we are looking for loci where the genotypes are predictable based on the data quality scores.
 This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
-These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
+The variants that are associated to the quality of mapped bases have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in low-mapping quality individuals.
 
-Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques.
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
-For population genetic analyses, the removal of individuals with low $Q$ appears preferable to filtering positions.
+Thus, this mutational signal is heavily enriched in GWLR suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques.
+The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
+For population genetic analyses, the removal of individuals with low quality sequence data appears preferable to filtering positions.
 
 	\subsection{Identifying Suspicious SNPs in the 1kGP}
-The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
-The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
+The results of the GWLR performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+The average quality of sequencing data of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
 This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
-By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
-It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
-However, the $Q$ per individual appeared to be the best metric to identify spurious sites.
+By 2011, older sequencing technologies were phased out and methods became more consistent, resulting in higher and more uniform quality.
+Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases per individual because it best correlated with mismatched sites across Japanese cohorts.
 
@@
 
-\caption{\textbf{A}The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A}The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project.
 Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
@@
 
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.
+The GWAS approach similarly identified mapping-quality-associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}).
+We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs (Figure \ref{Manhattan}).
 
@@
 
-\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
+\caption{Overlap of SNPs identified independently to be associated with quality. 
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
+The populations that tend to have the most low-quality individuals also tend to have the most variants associated to quality. 
 The same variants identified as being low quality independently in each population are found in other populations. }
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
-In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
-We found that all of the SNPs we identified as being associated with $Q$ for chromosomes 1 and 2 were imputed. 
-This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants have not been removed.
+In order to investigate the proportion of variants that we have identified as being associated to quality that are being imputed, we submitted genotype data from chomosomes 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+We found that all the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed back into these individuals. 
+This suggests that the imputation reference panel likely includes individuals with low sequence quality, and will be imputed in individuals who most closely match the low-quality individual.
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
+Once we identified SNPs that were clearly associated with low quality, we searched the literature for any GWAS that might have reported these dubious variants as being significantly correlated with some biological trait.
 The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
@@
 The data used was imputed genotype data using Hapmap II as the reference.
-One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ \citep{Ebejer2013}.
+One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality \citep{Ebejer2013}.
 While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
 Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
-We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.
+We found that roughly 1\% of the SNPs associated to quality were found in HapMap; rs6057648 is among those found in HapMap.
 
 Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
-They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high $Q$.
+They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high average quality of mapped bases.
 They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
 They also reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$ \citep{Mandage2017}.
+Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against data quality \citep{Mandage2017}.
 
@@
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
-A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-In this case, we used a cut-off of $Q$ over 30.
+A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low quality.
+In this case, we used a cut-off of an average quality of mapped bases over 30.
 Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
@@
 The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
-Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file. 
+Average quality of mapped bases per sample was obtained from the BAS files associated with each alignment file. 
 Each BAS file has metadata regarding each sequencing event for each sample. 
-If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
+If a sample was sequenced more than once, we took the average of the each quality score from each sequencing instance. 
 The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
-This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the $Q$ scores and only kept the earliest sequencing date per sample.
+This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the quality scores and only kept the earliest sequencing date per sample.
 The dates of the sequencing are only used to plot Figure.
@@
 
-\subsection{Genome wide association study}
+\subsection{Genome wide simple logistic regression analysis}
 We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
-\caption{Regressions against the average quality per mapped bases $Q$ per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\caption{Regressions against mean quality per mapped base pair per individual is an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
 }
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
-\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
+\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to quality.}
 \label{MostSig}
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
-\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
+\caption{Logistic regression of rs6057648 for the populations with significant association to quality.}
 \label{TwinsSNP}

### 2018-09-20 legacy.tex

--- 2018-09-20 legacy.tex
@@
 
-While we were unable to reproduce the heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
-We also found that this batch effect led to spurious results in a additional recent publications as a consequence of biased imputation.
+While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
+We also found that this batch effect led to spurious imputation results which found their way in recent publications.
 
 The results section is organized as follows.
-First, we explore the mutational signal found in the 1kGP Japanese population and methods by which we can identify this technical artifact.
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs whose validity is highly debatable.
+We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of suspicious SNPs.
 Finally, we investigate how these SNPs have impacted modern genomics analyses.
-We report that major imputation servers impute the dubious SNPs. 
-We also identify two publications that report such SNPs as genome-wide significant or nearly significant hits.
 
@@
 This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
-Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
+These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
 Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants.
+\sgcomment{this sentence needs work}
 
@@
 To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
-Traditionally, genome wide association studies use linear regression by predicting a phenotype based on genotype. 
-In this case, we are using simple logistic regression looking for loci where the genotypes are predictable based on the data quality scores.
+Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
+Here we use logistic regression of the genotypes on a quantitative measure of quality (i.e., the genotype is the dependent variable).
 This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
@@
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
-By 2011, older sequencing technologies were phased out, and method development solidify, which coincides with the sequencing quality levelling off.
-It is clear that there are many covariates that could be used to identify spurious sites such as sequencing date, sequencing technology or even the sequencing centre. 
-However, the $Q$ per individual appeared to be the best metric to identify spurious sites.
 
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
-Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-We wanted a test statistic to represent the association across all populations; to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.
+By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
+Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.
 
@@
 \end{figure}
+
+The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
+Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.
 
@@
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
-In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server.
-We found that all of the SNPs we identified as being associated with $Q$ for chromosomes 1 and 2 were imputed. 
-This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants have not been removed.
+
+\sgcomment{In order to -> To, in general} In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+We found that all of the \sgomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
+This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
 

### 2018-09-20 legacy.tex
more comments

@@
 	\subsection{Identifying Suspicious SNPs in the 1kGP}
-The results of the \textit{reverse} GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
 The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
-This variability could be as a result of fluctuating sequence platform protocols or variation between sequencing centres.
+This variability could be as a result of \sgcomment{could result from?} fluctuating sequence platform protocols or variation between sequencing centres.
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
@@
 
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP.
+The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
 Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan}.
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan} \sgcomment{figure missing for me}.
 
@@
 \sgcomment{In order to -> To, in general} In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that all of the \sgomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
+We found that all of the \sgcomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
 This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
@@
 The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
+\sgcomment{This is misleading!
+Neither has reached $10^-8$}
 
@@
 The other study used the 1kGP sequence data and cell lines directly.
-Both papers used strict quality thresholds, including population genetic statistical tests such as the Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+Both papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds. 
 They also removed rare alleles and alleles with high degrees of missingness. 

### 2018-10-02 legacy.tex
made recommended changes
added text about hap map and gnomAD sites.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\title{Legacy Data Confounds Modern Genomics Studies}
-\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}
+\title{Legacy Data Confounds Genomics Studies}
+%\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}
 
@@
 
-While certain mutagens have very distinct mutational signatures, others are more subtle and require much larger datasets to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}.
-\sgcomment{Missing logical link}
-Differences in mutational signatures across human populations have recently been reported.
+In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures, which can be revealed in large sequencing panels. 
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
@@
 
-Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a} \sgcomment{That would be 2017, no?}.
+Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
 While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect.
+
 These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
-Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants.
-\sgcomment{this sentence needs work}
+Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.
 
@@
 
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
 Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
@@
 The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
-This variability could be as a result of \sgcomment{could result from?} fluctuating sequence platform protocols or variation between sequencing centres.
+This variability could result from fluctuating sequence platform protocols or variation between sequencing centres.
 There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
@@
 Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan} \sgcomment{figure missing for me}.
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels) (Figure \ref{Manhattan} and Figure \ref{indels_Manhattan}).
 
@@
 
-\sgcomment{In order to -> To, in general} In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that all of the \sgcomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
+In general, to investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+We found that all the SNPs we identified as being associated with $Q$ were imputed back in the samples. 
 This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
@@
 Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait.
-The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$.
-\sgcomment{This is misleading!
-Neither has reached $10^-8$}
+The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had described variants that \todo{nearly reached genome wide significance} of $ p < 10^{-8}$.
 
@@
 We were not able to pinpoint the precise technical source of the discrepancy. 
-Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced with an entirely different platform.
-\sgcomment{Check that is sanger, if so, say so}.
-We find that \sgcomment{$99\%$--check} of the suspicious snps are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
+Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}. 
+We find that $99\%$ of the suspicious SNPs are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
 
-The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
+The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 42 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
+We also note that 164 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
 Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
@@
 \label{Singles}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
+\caption{Site frequency spectrum of suspicious SNPs found in HapMap and GnomAD.
+The blue dot in the HapMap plot is the joint frequency of a variant from a published GWAS.
+There are 42 suspicious SNPs present in the HapMap database and 162 suspicious SNPs present in the GnomAD database.}
+\label{HapMap_GnomAD}
 \end{figure}

### 2018-10-02 PlottingScripts/ManhattanPlot.R

@@ -4,7 +4,7 @@ library(magrittr)
 library(data.table)
 library(ggplot2)
 library(multtest)
-
+library(cowplot)
 #dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
 #out='/Volumes/gravel/luke_projects/1000Genomes/MeanDev/'
 dir='~/Documents/Regression/'
@@ -143,7 +143,15 @@ for(df in seq(1,26)){
 	LP[[df]] <- MakePlot(file)	
 	}
 
-pf <- grid.arrange(LP[[1]],LP[[2]], LP[[3]],LP[[4]],LP[[5]],LP[[6]],LP[[7]],LP[[8]],LP[[9]],LP[[10]], LP[[11]],LP[[12]],LP[[13]],LP[[14]],LP[[15]],LP[[16]],LP[[17]],LP[[18]],LP[[19]],LP[[20]], LP[[21]],LP[[22]],LP[[23]], LP[[24]],LP[[25]],LP[[26]])
+pf <- plot_grid(
+		plot_grid(LP[[1]],LP[[2]], LP[[3]],LP[[4]],LP[[5]],LP[[6]],nrow=1),
+		plot_grid(LP[[7]],LP[[8]],LP[[9]],LP[[10]],LP[[11]],nrow=1),
+		plot_grid(LP[[12]],LP[[13]],LP[[14]],LP[[15]],LP[[16]],nrow=1),
+		plot_grid(LP[[17]],LP[[18]],LP[[19]],LP[[20]],LP[[21]],nrow=1), 
+		plot_grid(LP[[22]],LP[[23]],LP[[24]],LP[[25]],LP[[26]],nrow=1),ncol=1)
+
+ggsave('~/Documents/QualityPaper/Figures/AllDeviances_1-26.jpg',pf, height=16,width=22)
+ggsave('~/Documents/QualityPaper/Figures/AllDeviances_1-26.tiff',pf, height=16,width=22)
 
 ggsave('~/Documents/QualityPaper/Figures/AllDeviances.jpg',pf, height=16,width=22)
 ggsave('~/Documents/QualityPaper/Figures/AllDeviances.tiff',pf, height=16,width=22)

### 2018-10-03 legacy.tex
edits throughout

@@
 Because of the extraordinary value of freely available data, this early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-However, the first sequenced genomes in the 1kGP, for example, were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
+However, the first sequenced genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
 The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}. 
 This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
-This begs the question: when should we retire legacy reference data?
 
@@
 
-In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures, which can be revealed in large sequencing panels. 
+In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures which can be revealed in large sequencing panels. 
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
@@
  In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
-This heterogeneity is intriguing because differences in mutational signatures accumulate over many generations.
+This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-We therefore decided to follow up on this observation, by using a newly sequenced dataset \sgcomment{description of cohort.}
+We therefore decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals.
 
-While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a batch effect in the 1kGP data that we were not previously aware of.
-We also found that this batch effect led to spurious imputation results which found their way in recent publications.
+While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
+In addition to creating biases in mutational signatures, this artefactt leads to spurious imputation results which found their way in recent publications.
 
@@
 We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of suspicious SNPs.
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs that show evidence for technical bias.
 Finally, we investigate how these SNPs have impacted modern genomics analyses.
@@
 Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
-However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
-This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect.
+While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
+Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
 
-These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
+These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests.
 Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.
@@
 When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-Regressions against different individual-level quality metrics provided by the 1kGP revealed that $Q$ per individual was an excellent correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
+Regressions against different individual-level quality metrics provided by the 1kGP revealed that average quality of mapped bases  $Q$ per individual was a strong correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
 Individuals with low $Q$ show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
@@
 The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
-For population genetic analyses, the removal of individuals with low $Q$ appears preferable to filtering positions.
+For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs.
+For other analyses, removing suspicious SNPs may be adequate.
+\sgcomment{Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.}
 
 	\subsection{Identifying Suspicious SNPs in the 1kGP}
-The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
+The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether other populations in the 1kGP exhibited similar artefacts.
 The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
-This variability could result from fluctuating sequence platform protocols or variation between sequencing centres.
-There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.
+This variability could result from evolving sequence platform and protocols or variation between sequencing centres.
 
@@
 To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels) (Figure \ref{Manhattan} and Figure \ref{indels_Manhattan}).
+\sgcomment{We should discuss significance threshold and FDR!}
 
@@
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait.
-The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had described variants that \todo{nearly reached genome wide significance} of $ p < 10^{-8}$.
-
+Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
+The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had reported these variants as close to the genome-wide significant threshold.
 One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
@@
 \section{Discussion}
-One paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
-While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
-The data used was imputed genotype data using Hapmap II as the reference.
-One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ \citep{Ebejer2013}.
-While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
-Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
-We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.
+One of these studies, which identified SNPs associated to ADHD symptoms, used HapMap as a reference panel, rather than 1000 Genomes \citep{Ebejer2013}.
+The variant identified is among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
+%While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
+%The data used was imputed genotype data using Hapmap II as the reference.
+%One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ .
+%While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
+%Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
+%We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.
 
-Another paper used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV.
-They did PCR confirmation of EBV copy number using 7 samples from some of the populations, although as it turns out, all of these samples have relatively high $Q$.
-They performed a GWAS on the EBV copy number and identified only one SNP with significance below $ p < 10^{-8}$. 
-They also reported the SNPs reading significance below $ p < 10^{-6}$. 
-Four of the SNPs in the top hits of the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$ \citep{Mandage2017}.
+The second study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
+Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as \sgcomment{number} SNPs with $ p < 10^{-6}$. 
+Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
+The study further confirmed EBV copy number estimation using PCR on 7 samples, but all of them had high $Q$.
 

### 2018-10-03 legacy.tex
more edits throughout

@@
 It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
-Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics.
+Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics. \sgcomment{this could be cleaned up}
 
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
-\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
+\caption{SNPs associated with $Q$ in a single population. The x-axis is sorted by genomic position.}
 \label{Singles}
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
-\caption{Site frequency spectrum of suspicious SNPs found in HapMap and GnomAD.
-The blue dot in the HapMap plot is the joint frequency of a variant from a published GWAS.
-There are 42 suspicious SNPs present in the HapMap database and 162 suspicious SNPs present in the GnomAD database.}
+\caption{Frequencies of $Q$-associated SNPs shared between 1000G and either HapMap or GnomAD.
+The blue dot in the HapMap plot is the is the variant reported in \citep{Mandage2017}, as discussed in the text.
+There are 42 such SNPs in HapMap and 162 in the GnomAD.}
 \label{HapMap_GnomAD}

### 2018-10-05 legacy.tex
misc

@@
 
+\subsection{Frequencies of SNPs in other datasets}
+Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
+We then compared the frequencies of these 42 SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
+Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.
+
+Similarly, using the allele frequencies from GnomAD, we were able to extract 164 sites that are statistically significantly associated to the average quality of mapped bases.
+However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).
+
 \section{Code Availability}

### 2018-10-23 legacy.tex
updated text, added figures for indels and repeats

@@
 \title{Legacy Data Confounds Genomics Studies}
-%\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}
 
@@
 
+The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
+Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
+To account for the large number of tests we performed a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
+\sgcomment{We should discuss significance threshold and FDR!}
+\luke{is this more appropriate for the methods section?}
+
+\begin{table}[h]
+\begin{tabular}{l  l  l}
+                      & \textbf{Repeat}  & \textbf{Non-Repeat}       \\ \cline{2-3} 
+\textbf{SNP}  & 2207 & 3839 \\ \cline{2-3} 
+\textbf{INDEL} & 750  & 1175 \\ \cline{2-3}
+\end{tabular}
+\caption{Number of statistically significant variants per category}
+\label{sigTable}
+\end{table}
+
 \begin{figure}
@@
 
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
-Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}).
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels) (Figure \ref{Manhattan} and Figure \ref{indels_Manhattan}).
-\sgcomment{We should discuss significance threshold and FDR!}
-
-\begin{figure*}
+\begin{figure}
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
@@
   \label{OverLap}
-\end{figure*}
+\end{figure}
 
@@
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 3990 SNPs and 1134 indels were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 7971 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}
 
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 3839 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
-  \label{Manhattan}
+ \label{NRS_Manhattan}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}
 
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 1134 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 1175 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
-  \label{indels_Manhattan}
+ \label{NRI_Manhattan}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}
+
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
+There are 2207 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+\label{RS_Manhattan}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_Indels.jpg}
+
+\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
+There are 750 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+ \label{RI_Manhattan}
 \end{figure}
@@
 \begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
+\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
+\label{DiabetesSNP}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
@@
 \end{document}
-
-%\begin{center}
- %\begin{tabular}{l l l r r} 
- %\hline
-%Author & Year & Journal & RSID & PMID \\ [0.5ex] 
- %\hline
-%Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\
-
-%Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
-%& & & rs201255786, rs201761909&\\
-%\hline
-%\end{tabular}
-%\end{center}

### 2018-10-23 PlottingScripts/PlotRegression.R

@@ -65,3 +65,19 @@ snp<-read.table('~/Documents/Regression/Chr1_149214891.txt',sep=',')
 ggplot(snp, aes(x=V6, y=V4, color=V5))+geom_point(shape=1)+facet_wrap(~V5,ncol=1, strip.position="right")+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Quality', y='Genotype')+scale_y_continuous(breaks=c(0,1))+geom_vline(xintercept=30, color='grey60',linetype=3)+scale_color_manual(breaks= snp$V5,values = MyColour)+guides(color=F)
 
 ggsave('~/Documents/QualityPaper/Figures/RegressionPlot_Indels.jpg',height=10,width=5)
+
+
+###DIABETES
+snp<-read.table('~/Documents/QualityPaper/sig/Chr8_19816934_Diabetes.txt',sep=',')
+dev<-read.table('~/Documents/QualityPaper/sig/Chr8_19816934_Diabetes_dev.txt')
+dev$p<-- pchisq(dev$V4, 1, lower.tail=F, log.p=T)/log(10)
+dev<-dev[which(dev$p>6),]
+sig<-merge(snp, dev, by.x=c('V2','V5'), by.y=c('V2','V3'))
+sig$Name<-paste(sig$V5, '\n -log10(p) :',round(sig$p, digits=2))
+
+ggplot(sig, aes(x=V6, y=V4.x, color=V5))+geom_point(shape=1)+facet_wrap(~Name,ncol=2, strip.position='right')+geom_smooth(method = "glm",     method.args = list(family = "binomial"), se = FALSE,size=0.5)+theme_classic()+labs(x='Average quality of mapped bases', y='Presence of Alternate Allele')+scale_y_continuous(breaks=c(0,1))+theme(strip.text.y = element_text(angle = 0))+geom_vline(xintercept=30, color='grey60',linetype=3)+scale_color_manual(breaks= sig$V5,values = MyColour)+guides(color=F)
+
[…]

### 2018-10-23 PlottingScripts/TotalRegression.R

@@ -5,6 +5,7 @@ library(data.table)
 library(ggplot2)
 library(multtest)
 library(cowplot)
+library(reshape2)
 
 ##read Regression
 Regression<-fread('~/Documents/QualityPaper/Misc/Reg_ALL_SITES_5col.txt', fill=T)
@@ -31,11 +32,7 @@ Regression$drop<-NULL
 
 ##read Repeats
 Repeats <- fread('~/Documents/QualityPaper/Misc/Reg_strictMask_nestedRepeats.POS',fill=T,col.names=c('Chr','Pos'))
-Repeats$Type='Repeat'
-
-##read Non-Repeats
-NoRepeats <- fread('~/Documents/QualityPaper/Misc/Reg_strictMask_NOnestedRepeats.POS',fill=T,col.names=c('Chr','Pos'))
-NoRepeats$Type='NOT_Repeat'
+Repeats$Repeat='yes'
 
 ##read Indels
 Indels <- fread('/Users/luke/Documents/QualityPaper/Misc/Reg_indels.POS',header=T)
@@ -47,21 +44,23 @@ colnames(rsID)<-c('Chr','Pos','rsID')
 
 Regression <- left_join(Regression, Repeats, by=c('Chr','Pos'))
 
-Regression <- left_join(Regression, NoRepeats, by=c('Chr','Pos'))
-
 Regression <- left_join(Regression, Indels, by=c('Chr','Pos'))
 
+Regression <- left_join(Regression, rsID, by=c('Chr','Pos'))
+
+write.table(Regression, file='/Users/luke/Documents/QualityPaper/Misc/TotalRegression.csv',row.names=F)
+
 Repeat_Indels <- Regression[which(Regression$Indel=='yes' 
-								& Regression$Type.x=='Repeat'),]
+								& Regression$Repeat=='yes'),]
 
 NORepeat_Indels <- Regression[which(Regression$Indel=='yes' 
-								& Regression$Type.y=='NOT_Repeat'),]
+								& is.na(Regression$Repeat),]
 								
 Repeat_SNPs <- Regression[which(is.na(Regression$Indel) 
-								& Regression$Type.x=='Repeat'),]
+								& Regression$Repeat=='yes'),]
 								
 NORepeat_SNPs <- Regression[which(is.na(Regression$Indel) 
[…]

### 2018-10-23 legacy.tex
small edits

@@
 \begin{table}[h]
-\begin{tabular}{l  l  l}
-                      & \textbf{Repeat}  & \textbf{Non-Repeat}       \\ \cline{2-3} 
-\textbf{SNP}  & 2207 & 3839 \\ \cline{2-3} 
-\textbf{INDEL} & 750  & 1175 \\ \cline{2-3}
+\begin{tabular}{l  r r}
+                      & {Repeat}  & {Non-Repeat}       \\ \hline
+{SNP}  & 2207 & 3839 \\  
+{INDEL} & 750  & 1175 \\ \hline
 \end{tabular}
@@
 It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
-Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics. \sgcomment{this could be cleaned up}
+Given that the bias appears to have been resolved in later stages of the project using updated sequencing platforms, we did not pursue this line of investigation.
 

### 2018-10-24 legacy.tex
slight updates, prep for merge

--- 2018-10-23 legacy.tex
@@
 The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
-Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
+Over \todo{3517} variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
 To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
 To account for the large number of tests we performed a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
-\sgcomment{We should discuss significance threshold and FDR!}
-\luke{is this more appropriate for the methods section?}
 
@@
 
-\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
+\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.\luke{update figure with all variants} 
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had reported these variants as close to the genome-wide significant threshold.
-One of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
+The NHGRI-EBI Catalog of published genome-wide association studies identified \todo{three} recent publications that had reported these variants as close to the genome-wide significant threshold.
+Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
 The other study used the 1kGP sequence data and cell lines directly.
-Both papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds. 
+All three papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
@@
 \section{Discussion}
-One of these studies, which identified SNPs associated to ADHD symptoms, used HapMap as a reference panel, rather than 1000 Genomes \citep{Ebejer2013}.
-The variant identified is among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
+Two of these studies, which identified SNPs associated to ADHD symptoms and metabolic traits respectively, used HapMap as a reference panel, rather than 1000 Genomes \citep{Ebejer2013}.
+\todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
 %While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
@@
 Further investigations in the source of the issue could be done using the raw sequence data from the 1kGP along with the per sample per base quality metrics. \sgcomment{this could be cleaned up}
+
+%%%meeting stuff
+%Association with EBV strain raises the question of whether it's a cell line artifact, but the fact that it also correlates with sequencing tech makes it less likely.
 
@@
 \end{document}
+
+%not necessarily all fake SNPs, there;s this bias, in most cases we expect them not to be real, there's a small chance that they are real
+%could be false posittive, or have a strong calling bias
+%majority of them don't reproduce, but a small portion of them do reproduce and should be looked at with care
+
+%if the SNP was imputed with HapMap maybe it is real, but if this SNP is imputed using the other database, it might throw things off.

### 2018-10-24 legacy.tex
update tex

@@
 
-%%%meeting stuff
-%Association with EBV strain raises the question of whether it's a cell line artifact, but the fact that it also correlates with sequencing tech makes it less likely.
-
 \subsection{Technical Artifact}
@@
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 we're considered.
-In total, 57,838,849 diallelic autosomal SNPs passed our quality controls for the mutation spectrum analysis and reverse GWAS. 
-We also ran the same reverse GWAS separately on 2,952,684 insertions and deletions that passed the same quality controls.
+In total, 57,838,849 diallelic autosomal SNPs passed our quality controls for the mutation spectrum analysis.
+
+For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis.
+The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in \todo{number} of SNPs and  \todo{2,952,684} insertions and deletions that passed the same quality controls.
+We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.
+
+\begin{table}[h]
+\begin{tabular}{l  r r}
+                      & {Repeat}  & {Non-Repeat}       \\ \hline
+{SNP}  & \todo{number} & \todo{number}  \\  
+{INDEL} & \todo{number}   & \todo{number}  \\ \hline
+\end{tabular}
+\caption{Number of variants included in the analysis per category}
+\label{totTable}
+\end{table}
 
@@
 Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
-We then compared the frequencies of these 42 SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
+We then compared the frequencies of these \todo{42} SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
 Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.
 
-Similarly, using the allele frequencies from GnomAD, we were able to extract 164 sites that are statistically significantly associated to the average quality of mapped bases.
+Similarly, using the allele frequencies from GnomAD, we were able to extract \todo{164} sites that are statistically significantly associated to the average quality of mapped bases.
 However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).
@@
 
-\caption{Empirical distribution plots of the sum of deviances grouped by N, the number of populations included.
+\caption{Empirical distribution plots of the sum of deviances for Non-Repeat SNPs grouped by N, the number of populations included.
 The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom.

### 2018-10-25 legacy.tex
updated figures

@@
 For other analyses, removing suspicious SNPs may be adequate. 
-\sgcomment{Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.}
+Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.
 
@@
 \section{Discussion}
-Two of these studies, which identified SNPs associated to ADHD symptoms and metabolic traits respectively, used HapMap as a reference panel, rather than 1000 Genomes \citep{Ebejer2013}.
-\todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
+Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
+The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data.
+It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
+However, considering that there are only \todo{number} suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
+One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}.
+Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real, however we cannot discount the chance that they are real.
+It is clear that the majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
+For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.
+
+%\todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
 %While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
@@
 For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis.
-The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in \todo{number} of SNPs and  \todo{2,952,684} insertions and deletions that passed the same quality controls.
+The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in a total of 80,971,616 variants ($S$) included in the test (see Table \ref{totTable}).
 We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & \todo{number} & \todo{number}  \\  
-{INDEL} & \todo{number}   & \todo{number}  \\ \hline
+{SNP}  & 22,651,762 & 51,643,590  \\  
+{INDEL} &  2,176,825  & 4,499,439 \\ \hline
 \end{tabular}
-\caption{Number of variants included in the analysis per category}
+\caption{Number of variants included in the analysis by category}
 \label{totTable}
@@
 \caption{Frequencies of $Q$-associated SNPs shared between 1000G and either HapMap or GnomAD.
-The blue dot in the HapMap plot is the is the variant reported in \citep{Mandage2017}, as discussed in the text.
-There are 42 such SNPs in HapMap and 162 in the GnomAD.}
+The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
+There are 70 such SNPs in HapMap and 186 in the GnomAD.}
 \label{HapMap_GnomAD}
@@
 \end{document}
-
-%not necessarily all fake SNPs, there;s this bias, in most cases we expect them not to be real, there's a small chance that they are real
-%could be false posittive, or have a strong calling bias
-%majority of them don't reproduce, but a small portion of them do reproduce and should be looked at with care
-
-%if the SNP was imputed with HapMap maybe it is real, but if this SNP is imputed using the other database, it might throw things off.

### 2018-10-25 Misc/HapMap+GnomAD.R

@@ -1,31 +1,0 @@
-library(data.table)
-library(ggplot2)
-library(stringr)
-library(cowplot)
-GP<-fread('/Users/luke/Documents/QualityPaper/sig/GenomeWide_sig0.01.frq', skip=1,fill=T,col.names=c('CHROM','POS','count','AC','AF', 'AN'))
-GP$AC1<-round(GP$AC*(1-GP$AF), digits=0)
-
-HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_Significant0.1SNPs_format.frq', sep='\t',fill=T, col.names=c('CHROM','POS','rsID','AC','AF', 'AN'))
-HM<-HM[which(complete.cases(HM)==T),]
-
-
-plot<-merge(GP, HM, by=c('CHROM','POS'))
-
-A=ggplot(plot, aes(x=1-AF.x, y=AF.y))+geom_point(shape=1)+geom_point(data=plot[which(plot$rsID=='rs6057648'),], color='blue')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()
-
-GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
-GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
-GN$AC_raw<-as.numeric(as.character(GN$AC_raw))
-ggplot(GN, aes(x=reorder(POS,CHROM),y=AF_raw))+geom_point(shape=1)+theme_classic()+theme(axis.text.x=element_blank())
-
-GNm<-melt(GN[,-c('AF_raw')], id=c('CHROM','POS','ID'))
-GNm$value<-as.numeric(as.character(GNm$value))
-GNm<-GNm[which(GNm$value!=0),]
-ggplot(GNm, aes(x=reorder(POS,CHROM),y=value, color=variable))+geom_point(alpha=0.4)+theme_classic()+theme(axis.text.x=element_blank())+labs(y='Frequency',x='Position')
-
-plt<-merge(GP, GN, by=c('CHROM','POS'))
-
-B=ggplot(plt, aes(y=1-AF, x=AF_raw))+geom_point(shape=1)+labs(x='Allele Frequency in GnomAD', y='Allele Frequency in 1kGP', title='Significant SNPs found in GnomAD and 1kGP')+theme_classic()
-C=plot_grid(A,B)
-
[…]

### 2018-10-25 PlottingScripts/Clean_Figure3.R

@@ -11,9 +11,11 @@ names(MyColour) <- colors $Pop
 
 #read the list of sig snps
 #bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('Chr', 'Pos'))
-dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
-fileNames = list.files(path=dir, pattern='dev10.csv', full.names = T)
-Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
+#dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
+#fileNames = list.files(path=dir, pattern='dev10.csv', full.names = T)
+#Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
+
+Reg<- fread('/Users/luke/Documents/QualityPaper/Misc/AllRegressionsOver10.txt')
 Reg<-Reg[which(Reg$Pos != 'Pos'),]
 Reg$Chr<-as.numeric(as.character(Reg$Chr))
 Reg$Pos<-as.numeric(as.character(Reg$Pos))

### 2018-10-25 PlottingScripts/HapMap+GnomAD.R

@@ -1,0 +1,44 @@
+library(data.table)
+library(ggplot2)
+library(stringr)
+library(cowplot)
+
+rsID<-fread('~/Documents/QualityPaper/sig/Total_Sig_rsID.txt', header=F, col.names=c('rsID'))
+pos<-fread('~/Documents/QualityPaper/sig/Total_Sig_POS.txt',col.names=c('CHROM','POS'))
+tot<-cbind(rsID,pos)
+tot $CHROM<-as.numeric(as.character(tot $CHROM))
+tot $POS<-as.numeric(as.character(tot $POS))
+
+GP<-fread('~/Documents/QualityPaper/sig/Total_Sig_FREQ.frq',fill=TRUE,col.names=c('CHROM','POS','count','AC','AF', 'AN'))
+GP$AC1<-round(GP$AC*(1-GP$AF), digits=0)
+GP$CHROM<-as.numeric(as.character(GP$CHROM))
+GP$POS<-as.numeric(as.character(GP$POS))
+
+GP<-merge(GP,tot, by=c('CHROM','POS'))
+
+
+HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_sig_rsID.frq', sep='\t',fill=T, col.names=c('CHROM','POS','rsID','AF', 'AN'))
+HM<-HM[which(complete.cases(HM)==T),]
+
+
+plt1<-merge(GP, HM, by=c('rsID'))
+plt1 $AF.x<-as.numeric(as.character(plt $AF.x))
+plt1 $AF.y<-as.numeric(as.character(plt $AF.y))
+
+A=ggplot(plt1, aes(x=1-AF.x, y=AF.y))+geom_point(shape=1)+geom_point(data=plt1[which(plt1$rsID=='rs6057648'),], color='blue')+geom_point(data=plt1[which(plt1$rsID=='rs301'),], color='green')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()+xlim(c(0,1))+ylim(c(0,1))
+
+GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
+GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
+GN$AC_raw<-as.numeric(as.character(GN$AC_raw))
[…]

### 2018-10-29 legacy.tex
edits throughout

@@
 
-\subsection{Batch Effects in Aging Reference Cohort Data}
+\sgcomment{Check time concordance throughout}		
+%\subsection{Batch Effects in Aging Reference Cohort Data}
 
@@
 
-Because of the extraordinary value of freely available data, this early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-However, the first sequenced genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}. 
-This led to heterogeneous sample preparations and data quality for many of the samples included in the dataset.
+The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
+Yet, because of the extraordinary value of freely available data, early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
 
-Even though it is intuitively clear that such batch effects have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
+Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
-Such genome wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
+Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
 
@@
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
-In addition to creating biases in mutational signatures, this artefactt leads to spurious imputation results which found their way in recent publications.
+In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications.
 
@@
 
-Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
+Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
 While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
+These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests \sgcomment{did we have a specific result for this?
+As in "$94\%$ of sites passed HW"...}.
+%Even though some of the mismatched variants are at low frequency in the 1kGP, and thus can evade Hardy-Weinberg detection, but many are at high enough frequency that HW detection
+%Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.
 
-These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests.
-Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.
+When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
+To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Supplementary Figure \ref{PC1_Correlation}).
+The average quality of mapped bases  $Q$ per individual stood out as a strong correlate (\sgcomment {p value?}): Individuals with low $Q$ show elevated rates of the signature .
+Thus, sequences with low mapping quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
-When mismatch sites were removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-Regressions against different individual-level quality metrics provided by the 1kGP revealed that average quality of mapped bases  $Q$ per individual was a strong correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP:
-Individuals with low $Q$ show elevated rates of the signature (see Supplementary Figure \ref{PC1_Correlation}).
-Thus, sequences with low quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
-
-To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
-Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
-Here we use logistic regression of the genotypes on a quantitative measure of quality (i.e., the genotype is the dependent variable).
-This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
+To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
+Traditionally, genome wide association studies use genotypes as the independent variable. 
+Here we perform a "reverse GWAS", in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
+We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
@@
 
-Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques.
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged. 
+Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques\sgcomment{You have to give some details or point to the results!}.
+The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged \sgcomment{You have to give some details or point to the results!}.
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
@@
 
-	\subsection{Identifying Suspicious SNPs in the 1kGP}
-The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether other populations in the 1kGP exhibited similar artefacts.
-The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
+	%\subsection{Identifying Suspicious SNPs in the 1kGP}
+The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
 This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
+By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
+%Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.
 
-By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
-Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.
-
-The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
-Over \todo{3517} variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
-To account for the large number of tests we performed a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
+We therefore performed the reverse GWAS approach in all populations, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected \sgcomment{quantify!}.
+Over \todo{3517} variants were independently associated to low $Q$ in at least two populations  \sgcomment{At what threshold?}(Figure \ref{OverLap}). 
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}):
+To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 
@@
 \end{tabular}
-\caption{Number of statistically significant variants per category}
+\caption{Number of statistically significant variants per category \sgcomment{Repeat/Non-repeat is super ambiguous need to clarify!}}
 \label{sigTable}
@@
 
-\caption{\textbf{A}The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
+\caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
 Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
@@
 
-In general, to investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that all the SNPs we identified as being associated with $Q$ were imputed back in the samples. 
-This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
+To investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+We found that all the SNPs associated with $Q$ were imputed back in the samples. 
+This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
 
@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
+\sgcomment{This is \emph{very} misleading, and wrong.
+We should discuss.
+}
+
+\sgcomment{Where are all your results comparing with Hapmap, etc? These should be in results.}
 
@@
 Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
-The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data.
+The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data \sgcomment{comma splice}.
 It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
 However, considering that there are only \todo{number} suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}.
-Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real, however we cannot discount the chance that they are real.
-It is clear that the majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
+One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}).
+Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real, however we cannot discount the chance that they are real \sgcomment{check style}.
+\sgcomment{It is clear that -- see Strunk-White} the majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
 For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.
+\sgcomment{This paragraph reads as wishy washy. }
 

### 2018-10-29 legacy.tex
forgot to push these changes

--- 2018-10-25 legacy.tex
@@
 The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
-Over \todo{3517} variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
+Over 3826 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
 To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
@@
 
-\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.\luke{update figure with all variants} 
+\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 In general, to investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that all the SNPs we identified as being associated with $Q$ were imputed back in the samples. 
+We found that \todo{all} the SNPs we identified as being associated with $Q$ were imputed back in the samples. 
 This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified \todo{three} recent publications that had reported these variants as close to the genome-wide significant threshold.
+Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
+The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
 Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
-The other study used the 1kGP sequence data and cell lines directly.
+The third study used the 1kGP sequence data and cell lines directly.
 All three papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
@@
 It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
-However, considering that there are only \todo{number} suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
+However, considering that there are only 70 suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
 One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}.
@@
 Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
-We then compared the frequencies of these \todo{42} SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
+We then compared the frequencies of these 70 SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
 Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.
 
-Similarly, using the allele frequencies from GnomAD, we were able to extract \todo{164} sites that are statistically significantly associated to the average quality of mapped bases.
+Similarly, using the allele frequencies from GnomAD, we were able to extract 186 sites that are statistically significantly associated to the average quality of mapped bases.
 However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).

### 2018-10-30 legacy.tex
edits before meeting

--- 2018-10-29 legacy.tex
@@
 
-We therefore performed the reverse GWAS approach in all populations, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected \sgcomment{quantify!}.
-Over \todo{3517} variants were independently associated to low $Q$ in at least two populations  \sgcomment{At what threshold?}(Figure \ref{OverLap}). 
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}):
+We therefore performed the reverse GWAS approach in all populations, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+% \sgcomment{quantify!}
+Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}). 
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
+We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable})
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 
+\luke{doing a final update on these values. Repeat entries, rerunning final adjustment with unique positions.}
 \begin{table}[h]
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 2207 & 3839 \\  
-{INDEL} & 750  & 1175 \\ \hline
+{SNP}  & 2207 & 6919 \\  
+{INDEL} & 750  & 2017 \\ \hline
 \end{tabular}
-\caption{Number of statistically significant variants per category \sgcomment{Repeat/Non-repeat is super ambiguous need to clarify!}}
+\caption{Number of statistically significant variants per category.
+Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately.
+SNPs and INDELs were also analyzed separately.}
 \label{sigTable}
@@
 
-\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.\luke{update figure with all variants} 
+\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
 The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+Phase 1 populations are marked by a star ( * ).
 The line plot underneath shows the number of populations for which a variant has reached significance.
@@
 To investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that all the SNPs associated with $Q$ were imputed back in the samples. 
+We found that 11\% of the SNPs associated with $Q$ were imputed back in the samples. 
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified \todo{three} recent publications that had reported these variants as close to the genome-wide significant threshold.
+Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
+The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
 Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
-The other study used the 1kGP sequence data and cell lines directly.
+The third study used the 1kGP sequence data and cell lines directly.
 All three papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
-Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for biological traits.
-\sgcomment{This is \emph{very} misleading, and wrong.
-We should discuss.
-}
+Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
+\sgcomment{This is \emph{very} misleading, and wrong. We should discuss. }
+\luke{Agreed, that sentence slipped by, there was another sentence like that in the discussion that I had fixed. Although the 3rd paper's SNP has $p < 10^{-11}$ }
 
-\sgcomment{Where are all your results comparing with Hapmap, etc? These should be in results.}
+We searched other databases to see if the suspicious variants had been sequenced in different cohorts.
+70 of these variants were present in HapMap and 186 in GnomAD (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+These variants include the two variants found in the studies that imputed genotypes using HapMap.
+With our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
 
-\section{Discussion}
-Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
-The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data \sgcomment{comma splice}.
-It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
-However, considering that there are only \todo{number} suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}).
-Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real, however we cannot discount the chance that they are real \sgcomment{check style}.
-\sgcomment{It is clear that -- see Strunk-White} the majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
-For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.
-\sgcomment{This paragraph reads as wishy washy. }
-
-%\todo{The variants identified} are among the \sgcomment{roughly $1\%$ this is a bit sloppy -- clarify} of suspicious SNPS that are concordant between hapmap and 1000 Genomes data, consistent with it being a false positive association.
 %While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
@@
 
+\section{Discussion}
+Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
+The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data \sgcomment{comma splice}.
+It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
+
+However, considering that there are only 70 suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
+One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}).
+Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real; however we cannot discount the chance that they are real.
+%\sgcomment{check style}
+%\sgcomment{It is clear that -- see Strunk-White} 
+The majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
+For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.
+\sgcomment{This paragraph reads as wishy washy. }
+\luke{Agreed, I'm not sure what we're trying to say here.}
+
 The second study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
-Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as \sgcomment{number} SNPs with $ p < 10^{-6}$. 
+Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
 Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
@@
 
-The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 42 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
-We also note that 164 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
+The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 70 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
+We also note that 186 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
 Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
@@
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 7971 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 10756 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 

### 2018-10-30 PlottingScripts/Clean_Figure3.R

@@ -9,6 +9,8 @@ colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.ch
 MyColour <- as.character(colors $hex)
 names(MyColour) <- colors $Pop
 
+phase1=c('YRI','LWK','CHB','CHD','JPT','CEU','LWK')
+
 #read the list of sig snps
 #bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('Chr', 'Pos'))
 #dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
@@ -32,8 +34,9 @@ NumPop$Var2<-as.numeric(as.character(NumPop$Var2))
 
 m=merge(Reg,NumPop, by.y=c('Var1','Var2'), by.x=c('Chr','Pos'))
 n=as.data.frame(table(m$Pop))
-plt=merge(m,n,by.x='Pop',by.y='Var1')
+plt=unique(merge(m,n,by.x='Pop',by.y='Var1'))
 plt$title = paste(plt$Pop, ' : ',plt$Freq.y)
+plt[which(plt$Pop %in% phase1),]$title = paste('*',plt[which(plt$Pop %in% phase1),]$title)
 plt <- plt[with(plt, order(-Freq.x, as.numeric(Chr),as.numeric(Pos))),]
 plt$Pos <- factor(plt$Pos, levels = unique(plt$Pos))
 plt$nrow<-seq(length=nrow(plt))

### 2018-10-30 legacy.tex
some suggestions

@@
 Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
-The HapMap project and 1000 Genomes project shared some cell lines, however used different sequencing platforms to produce the data \sgcomment{comma splice}.
+The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
 It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
-\caption{Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\caption{\sgcomment{not regressions}Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
 }
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
-\caption{Frequencies of $Q$-associated SNPs shared between 1000G and either HapMap or GnomAD.
-The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
+\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
+\sgcomment{Change title, check targeted  sequencing region.
+Specify proportion of SNPs not included} The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
 There are 70 such SNPs in HapMap and 186 in the GnomAD.}

### 2018-11-07 QE/thesis_1/main.tex
QE draft outline

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Masters/Doctoral Thesis 
% LaTeX Template
% Version 2.5 (27/8/17)
%
% This template was downloaded from:
% http://www.LaTeXTemplates.com
%
% Version 2.x major modifications by:
% Vel (vel@latextemplates.com)
%
% This template is based on a template by:
% Steve Gunn (http://users.ecs.soton.ac.uk/srg/softwaretools/document/templates/)
% Sunil Patel (http://www.sunilpatel.co.uk/thesis-template/)
%
% Template license:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[
11pt, % The default document font size, options: 10pt, 11pt, 12pt
oneside, % Two side (alternating margins) for binding by default, uncomment to switch to one side
english, % ngerman for German
doublespacing, % Single line spacing, alternatives: onehalfspacing or doublespacing
%draft, % Uncomment to enable draft mode (no pictures, no links, overfull hboxes indicated)
%nolistspacing, % If the document is onehalfspacing or doublespacing, uncomment this to set spacing in lists to single
%liststotoc, % Uncomment to add the list of figures/tables/etc to the table of contents
%toctotoc, % Uncomment to add the main table of contents to the table of contents
%parskip, % Uncomment to add space between paragraphs
%nohyperref, % Uncomment to not load the hyperref package
headsepline, % Uncomment to get a line under the header
%chapterinoneline, % Uncomment to place the chapter title next to the number on one line
%consistentlayout, % Uncomment to change the layout of the declaration, abstract and acknowledgements pages to match the default layout
]{MastersDoctoralThesis} % The class file specifying the document structure

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters

\usepackage{mathpazo} % Use the Palatino font by default

\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)

\addbibresource{example.bib} % The filename of the bibliography

\usepackage[autostyle=true]{csquotes} % Required to generate language-dependent quotes in the bibliography

%----------------------------------------------------------------------------------------
%	MARGIN SETTINGS
%----------------------------------------------------------------------------------------

\geometry{
	paper=a4paper, % Change to letterpaper for US letter
	inner=2.5cm, % Inner margin
	outer=3.8cm, % Outer margin
	bindingoffset=.5cm, % Binding offset
	top=1.5cm, % Top margin
	bottom=1.5cm, % Bottom margin
	%showframe, % Uncomment to show how the type block is set on the page
}

%----------------------------------------------------------------------------------------
%	THESIS INFORMATION
%----------------------------------------------------------------------------------------

\thesistitle{Mutational Patterns in Founder Populations} % Your thesis title, this is used in the title and abstract, print it elsewhere with \ttitle
\supervisor{Dr. Simon \textsc{Gravel}} % Your supervisor's name, this is used in the title page, print it elsewhere with \supname
\examiner{} % Your examiner's name, this is not currently used anywhere in the template, print it elsewhere with \examname
\degree{Doctor of Philosophy} % Your degree name, this is used in the title page and abstract, print it elsewhere with \degreename
\author{Luke \textsc{Anderson-Trocm\'e}} % Your name, this is used in the title page and abstract, print it elsewhere with \authorname
\addresses{} % Your address, this is not currently used anywhere in the template, print it elsewhere with \addressname

\subject{Statistical Population Genetics} % Your subject area, this is not currently used anywhere in the template, print it elsewhere with \subjectname
\keywords{} % Keywords for your thesis, this is not currently used anywhere in the template, print it elsewhere with \keywordnames
\university{\href{http://www.mcgill.ca}{McGill University}} % Your university's name and URL, this is used in the title page and abstract, print it elsewhere with \univname
\department{\href{https://www.mcgill.ca/humangenetics/}{Department of Human Genetics}} % Your department's name and URL, this is used in the title page and abstract, print it elsewhere with \deptname
\group{\href{http://simongravel.lab.mcgill.ca/Home.html}{Gravel Lab}} % Your research group's name and URL, this is used in the title page, print it elsewhere with \groupname
\faculty{\href{https://www.mcgill.ca/medicine/}{Faculty of Medecine}} % Your faculty's name and URL, this is used in the title page and abstract, print it elsewhere with \facname

\AtBeginDocument{
\hypersetup{pdftitle=\ttitle} % Set the PDF's title to your title
\hypersetup{pdfauthor=\authorname} % Set the PDF's author to your name
\hypersetup{pdfkeywords=\keywordnames} % Set the PDF's keywords to your keywords
}

\begin{document}

\frontmatter % Use roman page numbering style (i, ii, iii, iv...) for the pre-content pages

\pagestyle{plain} % Default to the plain heading style until the thesis style is called for the body content

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\begin{titlepage}
\begin{center}

\vspace*{.06\textheight}
{\scshape\LARGE \univname\par}\vspace{1.5cm} % University name
\textsc{\Large Doctoral Thesis Project Report}\\[0.5cm] % Thesis type

\HRule \\[0.4cm] % Horizontal line
{\huge \bfseries \ttitle\par}\vspace{0.4cm} % Thesis title
\HRule \\[1.5cm] % Horizontal line

\begin{minipage}[t]{0.4\textwidth}
\begin{flushleft} \large
\emph{Author:}\\
{\authorname} % Author name - remove the \href bracket to remove the link
\end{flushleft}
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}
\begin{flushright} \large
\emph{Supervisor:} \\
{\supname} % Supervisor name - remove the \href bracket to remove the link  
\end{flushright}
\end{minipage}\\[3cm]

\vfill

%\large \textit{A thesis submitted in fulfillment of the requirements\\ for the degree of \degreename}\\[0.3cm] % University requirement text
%\textit{in the}\\[0.4cm]
\groupname\\\deptname\\[2cm] % Research group name and department name

\vfill

{\large \today}\\[4cm] % Date
%\includegraphics{Logo} % University/department logo - uncomment to place it

\vfill
\end{center}
\end{titlepage}

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

\mainmatter % Begin numeric (1,2,3...) page numbering

\pagestyle{thesis} % Return the page headers back to the "thesis" style

\section{Introduction}

\subsection{founder populations}

Founder populations in theoretical population genetics
historical changes in population size or structure.

Founder populations in human history

Impact of studying founder populations. Why is it important?

\subsection{mutational signatures}

\chapter{A curious mutational signature in Japan}

\section{summary of results from paper}

\section{link to manuscript}

\chapter{Characterization of the Quebec population}

\section{Brief History of Quebec}

\section{Genizon Data and Balsac}
\subsection{approach and methods}

\section{Linking population structure to geography (results)}

\section{Future Directions}

\chapter{Estimating mutation rates}

\section{Review of current methods and limitations}

\subsection{Approach and methods for new calculation}

\subsection{Apply state of the art methods to new data}

\subsection{can we do better?}
Estimating mutation rate spatially across the genome.
Using IBD and Genealogy to better estimate the number of generations.

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - APPENDICES
%----------------------------------------------------------------------------------------

\appendix % Cue to tell LaTeX that the following "chapters" are Appendices

% Include the appendices of the thesis as separate files from the Appendices folder
% Uncomment the lines as you write the Appendices

%\include{Appendices/AppendixA}

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\printbibliography[heading=bibintoc]

%----------------------------------------------------------------------------------------

\end{document}

### 2018-11-07 PlottingScripts/Han_1kGP_SFS.R

@@ -1,0 +1,44 @@
+##plot SFS sig Han
+library(data.table)
+library(ggplot2)
+library(reshape2)
+
+Qual<-fread('~/Dropbox/LukeTemp/SubMeta.txt')
+Qual$V1<-NULL
+
+Han<- fread('~/genomes/genomes/Han90/83_Han_Sig.frq')
+
+KGP<- fread('~/Documents/QualityPaper/SigVCF/1kGP_90Han_shared.frq')
+
+plt<-unique(merge(Han, KGP, by=c('V1','V2'), all=T))
+
+ggplot(plt, aes(x=V6.x, y=V6.y))+geom_bin2d(binwidth=c(1/83,1/83))+scale_fill_distiller(palette = "Spectral")+labs(x='high depth',y='1kGP')+theme_classic()
+
+gt1<-fread('/Users/luke/Documents/QualityPaper/SigVCF/1kGP_all.genotypes.txt')
+gt1.m<-melt(gt1, id.vars = c('#CHROM','POS'))
+colnames(gt1.m)=c('CHROM','POS','Sample','kGT')
+
+
+gtHan<-fread('~/genomes/genomes/Han90/83_Han_all.genotypes.txt')
+gtHan$'#CHROM' <- as.integer(as.character(gsub('chr','',gtHan$'#CHROM')))
+gtHan.m<-melt(gtHan, id.vars = c('#CHROM','POS'))
+colnames(gtHan.m)=c('CHROM','POS','Sample','HanGT')
+
+gt<-merge(gtHan.m, gt1.m, by=c('CHROM','POS','Sample'))
+gt <-merge(gt, Qual, by.x='Sample',by.y='Name')
+gt$diff <- (1-as.numeric(gt$HanGT == gt$kGT))
+
+agg.gt<-aggregate(. ~ Sample, data = gt, sum)
+agg.gt$diff <- agg.gt$diff/149*100
+
+agg.pos<-aggregate(gt[,c('POS','average_quality_of_mapped_bases')], by=list(gt$HanGT,gt$POS),FUN= mean)
+
+ggplot(agg.gt, aes(x=HanGT, y= kGT, color= average_quality_of_mapped_bases))+geom_point()+ scale_color_continuous(name = "Quality")+theme_classic()+labs(x='# of Mutations in Resequence',y='# of Mutations in 1kGP')
+
[…]

### 2018-11-16 QE/thesis_1/main.tex
QE write up

@@
 
-\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
+\usepackage[backend=bibtex,style=numeric,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
 
-\addbibresource{example.bib} % The filename of the bibliography
+\addbibresource{Legacy.bib} % The filename of the bibliography
 
@@
 	inner=2.5cm, % Inner margin
-	outer=3.8cm, % Outer margin
+	outer=2.5cm, % Outer margin
 	bindingoffset=.5cm, % Binding offset
@@
 {\scshape\LARGE \univname\par}\vspace{1.5cm} % University name
-\textsc{\Large Doctoral Thesis Project Report}\\[0.5cm] % Thesis type
+\textsc{\Large Qualifying Exam}\\[0.5cm] % Thesis type
 
@@
 
-\subsection{founder populations}
+\subsection{Serial Founder Effects}
+The OOA is a series of founder effects.
+This can drastically change the allele frequencies within the population, which can lead to rapid divergence.
+Moreover, because of the limited population size, inbreeding will have impacts on heterozygosity.
+This can impact the rates of genetic diseases.
+
+Mutational patterns in founder populations have recently become possible to measure, and useful for quantifying changes in mutation rates, as well as demographic histories of populations.
 
@@
 
+Studying recent founder events can lead to better understanding of the mechanisms and dynamics of founder events.
+
 Founder populations in human history
@@
 Impact of studying founder populations. Why is it important?
+These populations often harbour a unique set of disease risks.
+They may also have a distinct mutational signature.
+
+\subsection{Thesis proposal}
+Aim : Investigating mutational patterns in founder populations. 
+Characterizing the genetic diversity of founder populations.
+
+Hypothesis : 
+Founder effects will influence the mutational signatures of a population leading to unique patterns in populations that can be used to infer demographic histories.
+
+Combining genetic, genealogical and geographic data can explain the population structure observed today, recreating historical demographics.
+
+This project will be broken down into three chapters :
+
+1. Unique signature in Japan
+Identify the source of the mutational pattern in Japanese population.
+
+2. Characterize the genetic diversity of the Quebec population (a case study)
+Explain population structure. Using Fst and UMAP/GEO.
+
+3. Explore mutation rate in Quebec population
+apply existing methods of estimating human mutation rate.
+can we do better? combine data to improve estimation.
 
 \subsection{mutational signatures}
+Some mutagenic processes will affect different DNA motifs. 
+Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
+Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
+Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
+
+In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures which can be revealed in large sequencing panels. 
+In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
+Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
+Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.
+
+ In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
+A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
+We therefore decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals.
 
 \chapter{A curious mutational signature in Japan}
+The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
 
+The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
+Yet, because of the extraordinary value of freely available data, early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
+
+Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
+This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
 \section{summary of results from paper}
+Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
+While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
+Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
+These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests.
 
+To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
+Traditionally, genome wide association studies use genotypes as the independent variable. 
+Here we perform a "reverse GWAS", in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
+We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
 \section{link to manuscript}
@@
 \section{Brief History of Quebec}
+The Quebec population was founded over four centuries ago from \~8,500 settlers from France.
+This small number of settlers are the direct ancestors of the great majority of Quebec's population of 7 million.
+In population genetics, a founder effect is characterized by a small number of founders from a larger source population giving rise to a population expansion.
+This tends to increased homozygosity and rapidly shift allele frequencies compared to the source population.
 
-\section{Genizon Data and Balsac}
+The north east of Quebec (e.g.
+Saguenay-Lac-Saint-Jean and Charlevoix) displays higher rates of some rare genetic diseases and increased homozygosity which is suggestive of a severe founder effect.
+(Gagnon, 2000)
+Western Quebec, on the other hand, has similar prevalence of diseases compared to Europe. (Gagnon, 2000)
+This discrepancy of disease rates across the Saint-Lawrence valley indicates that the severity of the bottleneck varies across geographic regions. (Gagnon, 2000)
+
+Indeed, the intensity of the founder effect in Quebec is not consistent across the population as there is evidence of geographical stratification which has been attributed to a number of factors like varying concentration of founders, native ancestry and geographic isolation.
+(Gagnon, 2000).
+The founders of Quebec were not all from the same regions in France, providing some initial diversity despite a small initial population size.
+Moreover, Native American admixture with founders has been estimated to be around 1\% which would have also been a source of genetic diversity among these settlements. (Moreau, 2013)
+Some settlements were more isolated than others resulting in less gene flow between sub-populations.
+The north-east of Quebec is characterized historically as being more isolated with higher emigration rates compared to western Quebec.
+
+\subsection{Quebec : A case study of founder populations}
+
+The Quebec founder effect is an interesting case study for population genetics for multiple reasons.
+Firstly, the source population of this bottleneck is still present, which may not be the case for older founder populations. 
+This provides us with the possibility of comparing allele frequencies between Quebec and its ancestral population.
+Secondly, many populations experienced founder effects as humans expanded out of Africa and populated the continents.
+Studying examples of founder effects in human populations are a great way to improve our understanding of human demographic histories.
+Finally, one unique aspect of this population is the deep genealogical data that spans four centuries and includes nearly 3 million entries.
+Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events.
+
+There have been several studies in the past few decades that have examined the genetic diversity and population structure of Quebec.
+Using genotype data, linkage disequilibrium and homozygosity analyses of sub-populations within Quebec showed significant structure and relatedness that persists in the current population (Roy-Gagnon, 2011).
+Quebec is not a homogeneous population, but rather made up of a combination of distinct sub-populations.
+This was confirmed with Fst statistics of these sub-populations, indicating substantial genetic structure.
+Of particular interest, they reported that there was greater differentiation between sub-populations of Quebec than between European populations.
+While this study included 140 genotyped individuals - a small sample by today's standards - they postulate that this difference is accounted for by genetic drift.
+Indeed, the heterogeneity of the founder effect in the Quebec population could be explained by a series of separate founder events influenced by variation in the relative isolation, and migration between settlements.
+
+\section{Genomic and genealogical data}
+
+The second chapter of my thesis will examine the genetic structure of the Quebec population using more recent and larger genomic datasets in combination with the deep-genealogical and geographical data.
+
+While the deep-genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes.
+
+Genizon data has 000 samples, genotyped using a variety of chips. These data require some cleaning and processing before analysis. This cohort has yet to be published.
+
+We will also include data from the Cartagene project which has 000 samples from...
+
+With these data, we hope to reproduce the results from previous work and answer the following question :
+(2) Are the sub-populations of Quebec the result of genetic drift from founder events and isolation, or the preservation of genetic structure from France?
+
 \subsection{approach and methods}
+Using Fst statistics we can infer the relative differences between sub-populations of Quebec, as done in (Roy-Gagnon, 2011).
+Dominic Nelson has written software that can produce forward simulations of the Quebec population using the genealogical data from BALSAC.
+These simulations can be used to produce estimated differences between sub-populations resulting purely from a founder effect.
+
+Comparing the Fst statistics between the real data and simulations will determine if the structure observed in Quebec is the result of preserved genetic structure from France or the result of drift from founder events.
 
 \section{Linking population structure to geography (results)}
+Preliminary analyses of the Genizon dataset are in agreement with previous work.
+In collaboration with Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data.
+This method takes the first *10* principle components of the data as input and reduces this high dimensional data preserving both local and global structure.
+This projection shows significant population structure within the Quebec population. 
+These data cluster into 5 roughly defined sub-populations, in agreement with what (Roy-Gagnon, 2011) observed \ref{Cluster}.
+
+We colored these plots based on the three dimensional (X,Y,Z) coordinates of the UMAP projection space converted to three dimensional color space (R,G,B).
+Thus, points that are close together in UMAP space will have similar color.
+
+We were able to link 000 of the Genizon individuals to the BALSAC database which includes marriage location of their ancestors.
+Combining the ancestral locations of these individuals with the three dimensional UMAP coloring scheme, geographic clustering of the sub-populations becomes apparent \ref{UMAP}.
+
+These preliminary results are very promising as they appear to capture the geographic and genetic clustering of the Quebec population.
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Genizon_BalSac_mapsInMaps4.jpg}
+\label{UMAP}
+\end{figure}
 
 \section{Future Directions}
+We know that there are regional differences in disease rates that are linked to population substructure.
+The genetic data used in UMAP analysis also identify similar clusters.
+Identifying the variants that are responsible for the clustering could also provide us with a shortlist of variants that could also be linked to the diseases that are associated with these sub-populations.
+This method has yet to be formalized, but in brief, performing a logistic regression predicting presence of genotypes using the axis of variation that separates one group from the others could be used to identify these variants.
+
+Expanding the current analyses to include data from French individuals could be used to confirm the results from the Fst analysis as well as provide insight into which regions in France are source populations to regions in Quebec.
 
@@
 \section{Review of current methods and limitations}
+Three main categories of methods :
 
+1. Phylogenetic rate : Chimp vs human estimates 2.5 x 10-8 but limitations about the number of generations that separate the two species. Based on fossil evidence.
+
+2. Direct approach : One generation using trio sequencing. 
+Tend to give lower estimates compared to phylogenetic approach. (1-1.25 X10-8)
+
+Pedigree approaches are simple in and direct approaches for per generation estimates but are met with several limitations.
+Sex bias : fathers germline mutations increase with age and could be different among men.
+Difficulty distinguishing difference between mosaicism and germline/somatic mutations.
+
+Using Trio : sequencing tech limitations in error rates.
+Mosaicism in parent or postzygotic mutations in child.
+Father's age : spermatogenesis cell division.
+Different types of mutations occur at different rates (and changes with age)
+
+Converting per generation to per year is also difficult. 
+Current approach is to divide generation rate by age of reproduction. (assumes same reproductive age for all pops over all time)
+Estimates may only be accurate to the population studied.
+
+Mutational signatures specific to populations suggest recent changes in mutation rates.
+
+One possible solution is to limit estimates to CpG sites as they appear to have more consistent mutation rates across primates.
+*(but apparently no diff in paternal age for CpG transitions vs other mutations)
+
+3. Population genetics approach : 
+Estimates around 1.6-1.7 x 10-8
+
+Taking advantage of founder populations with known pedigrees can get better estimates of mutation rates over multiuple generations.
+Identify mutations in regions of autozygosity and count the number of generations that separate these segments. 
+Reduces errors from somatic mutations. 
 \subsection{Approach and methods for new calculation}
@@
 \appendix % Cue to tell LaTeX that the following "chapters" are Appendices
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/5Clusters.png}
+\label{Cluster}
+\end{figure}
 

### 2018-11-16 QE/thesis_1/main.tex
more text for QE

@@
 \subsection{Serial Founder Effects}
-The OOA is a series of founder effects.
-This can drastically change the allele frequencies within the population, which can lead to rapid divergence.
+Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago, and resulted in a significant loss of diversity in the populations the colonized new habitats (Henn, 2012).
+This loss of diversity is the result of a population bottleneck from a small group leaving a source population. 
+Estimates of the parameters of this bottleneck using sequence data and modelling population dynamics range from 2,500 to as low as 450 individuals (Eller, 2004; Fagundes, 2007).
+However, as humans expanded into new habitats, each new colony was founded by small groups leading to successive bottlenecks and reduction in diversity.
+This trend is still apparent in modern humans, with distance from Africa being correlated with loss of genetic diversity (Ramachandran, 2005).
+Bottlenecks can drastically change the allele frequencies within the population, which can lead to rapid divergence (Rosenberg, 2002; Charlesworth, 2003).
 Moreover, because of the limited population size, inbreeding will have impacts on heterozygosity.
-This can impact the rates of genetic diseases.
+Accounting for population divergence and population specific allele frequencies are necessary for disease phenotype association tests (Gravel, 2010).
 
@@
 
-Founder populations in theoretical population genetics
-historical changes in population size or structure.
+Studying recent founder events can lead to better understanding of the mechanisms and dynamics of historical founder events.
 
-Studying recent founder events can lead to better understanding of the mechanisms and dynamics of founder events.
+Inferring historical population size based on local density of heterozygosity (Li \& Durbin, 2011)
 
-Founder populations in human history
+Admixture with Neanderthals?
 
-Impact of studying founder populations. Why is it important?
-These populations often harbour a unique set of disease risks.
-They may also have a distinct mutational signature.
+%Impact of studying founder populations. Why is it important?
+%These populations often harbour a unique set of disease risks.
+%They may also have a distinct mutational signature.
 
@@
 Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
+
 \section{summary of results from paper}
@@
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
-\section{link to manuscript}
+
+link to manuscript
 
@@
 \section{Review of current methods and limitations}
+
+Mutation rates are important for understanding the timing of evolutionary processes.
+It is used as an evolutionary clock if you assume a constant accumulation rate of neutral mutations.
+The basic equation of mutation rate is $N / LM$ where N is the number of mutations, L is the length of the region and M is the number of generations.
+Estimating mutation rates in humans has proven to be difficult for a number of reasons.
+[list of reasons here]
+
 Three main categories of methods :
 
-1. Phylogenetic rate : Chimp vs human estimates 2.5 x 10-8 but limitations about the number of generations that separate the two species. Based on fossil evidence.
+\subsection{Phylogenetic method :} Chimp vs human estimates $2.5\times10^{-8}$ but limitations about the number of generations that separate the two species. Based on fossil evidence.
 
-2. Direct approach : One generation using trio sequencing. 
-Tend to give lower estimates compared to phylogenetic approach. (1-1.25 X10-8)
+\subsection{Trio method :} One generation using trio sequencing. 
+Tend to give lower estimates compared to phylogenetic approach. ($1-1.25\times10^{-8}$)
 
@@
 
-3. Population genetics approach : 
-Estimates around 1.6-1.7 x 10-8
+\subsection{Population genetics method : }
+Estimates around $1.6-1.7\times10^{-8}$
 

### 2018-11-16 PlottingScripts/Han_1kGP_SFS.R

@@ -3,14 +3,41 @@ library(data.table)
 library(ggplot2)
 library(reshape2)
 
+Reg<- fread('/Users/luke/Documents/QualityPaper/Misc/AllRegressionsOver10.txt')
+Reg<-Reg[which(Reg$Pos != 'Pos'),]
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$dev))
+Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
+Reg<-Reg[which(Reg$plog10 >= 6),]
+RegCH<-Reg[which(Reg$Pop == 'CHS' | Reg$Pop == 'CHB'),]
+
+write.table(RegCH[,c('Chr','Pos')],'/Users/luke/Documents/QualityPaper/sig/CHINESE_pos.txt', quote=F, col.names=F, row.names=F)
+
+awkPos<-fread('/Users/luke/Documents/QualityPaper/sig/Total_Sig_awk_POS.txt')
+
+Overlap<- unique(RegCH[which(RegCH$Pos %in% awkPos$V2),c('Chr','Pos')])
+
 Qual<-fread('~/Dropbox/LukeTemp/SubMeta.txt')
 Qual$V1<-NULL
 
 Han<- fread('~/genomes/genomes/Han90/83_Han_Sig.frq')
+Han$V1 <- as.numeric(Han$V1)
+Han$V2 <- as.numeric(Han$V2)
 
-KGP<- fread('~/Documents/QualityPaper/SigVCF/1kGP_90Han_shared.frq')
 
-plt<-unique(merge(Han, KGP, by=c('V1','V2'), all=T))
+KGP1 <- read.table('/Users/luke/Documents/QualityPaper/SigVCF/1kGP_GenomeWide_83Han.frq',
+					sep="\t",fill=TRUE,col.names=paste("V",1:8,sep=''), skip=1)
+KGP <- read.table('~/Documents/QualityPaper/SigVCF/1kGP_90Han_shared.frq',
+					sep="\t",fill=TRUE,col.names=paste("V",1:8,sep=''), skip=1)
+					
+KGP2 <- unique(merge(KGP, Overlap, by.x=c('V1','V2'), by.y=c('Chr','Pos')))
[…]

### 2018-11-19 QE/thesis_1/main.tex
more words

@@
 
-Inferring historical population size based on local density of heterozygosity (Li \& Durbin, 2011)
-
-Admixture with Neanderthals?
-
-%Impact of studying founder populations. Why is it important?
-%These populations often harbour a unique set of disease risks.
-%They may also have a distinct mutational signature.
-
 \subsection{Thesis proposal}
@@
 \section{Future Directions}
-We know that there are regional differences in disease rates that are linked to population substructure.
-The genetic data used in UMAP analysis also identify similar clusters.
-Identifying the variants that are responsible for the clustering could also provide us with a shortlist of variants that could also be linked to the diseases that are associated with these sub-populations.
-This method has yet to be formalized, but in brief, performing a logistic regression predicting presence of genotypes using the axis of variation that separates one group from the others could be used to identify these variants.
+%We know that there are regional differences in disease rates that are linked to population substructure.
+%The genetic data used in UMAP analysis also identify similar clusters.
+%Identifying the variants that are responsible for the clustering could also provide us with a shortlist of variants that could also be linked to the diseases that are associated with these sub-populations.
+%This method has yet to be formalized, but in brief, performing a logistic regression predicting presence of genotypes using the axis of variation that separates one group from the others could be used to identify these variants.
+While Fst gives a good estimate of genome wide past coalescent rates, the analysis of long IBD segments can provide us with more detailed insight of more recent demographic history (Ralph, 2013; Charlesworth 2003).
+Because Fst takes in all polymorphisms as input for the statistic, it does not distinguish between recent and more ancient common ancestry.
+Analysis of IBD segments shared among individuals from Quebec can be useful for reconstructing the recent structure of relatedness.
+The length of the shared segments can be used to infer the number of generations leading back to a common ancestor.
+Combining this with geographic data can infer patterns of shared ancestry across the Saint-Lawrence valley.
+Ralph (2013) identified a general trend of decreased of shared ancestry with geographic distance as well as more regional signals of IBD in European populations.
+Europe has a rich history of complex migratory patterns and demographic histories that is difficult to trace back with certainty. 
+Quebec on the other hand has a much more recent history that may be easier to trace with IBD analysis.
 
+While studying Quebec populations on their own can provide detailed historical inference of demographics, including samples from other European populations can provide more context.
 Expanding the current analyses to include data from French individuals could be used to confirm the results from the Fst analysis as well as provide insight into which regions in France are source populations to regions in Quebec.
+French genome data is obtainable through Dr. Christian Dina, one of our collaborators in the University of Nantes.
 

### 2018-11-21 QE/thesis_1/main.tex
Mutation rate section

@@
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/Genizon_BalSac_mapsInMaps4.jpg}
-\label{UMAP}
-\end{figure}
-
 \section{Future Directions}
@@
 
+Mutation rate estimations are necessary for understanding the timing of evolutionary processes and events.
+Neutral mutations are expected to arise at a constant rate in the genome, and can therefore be used as an evolutionary clock that can be used to estimate divergence times (Moorjani, 2016). 
+The basic equation of mutation rate is $N / LM$ where N is the number of mutations, L is the length of the region and M is the number of generations. (Narasimah, 2016)
+Estimating mutation rates in humans has proven to be difficult for a number of reasons.
+Mutation rates vary across the genome as some regions are more conserved than others. 
+Mutation rates also vary over time.
+Shorter branches within the Ape family compared to other primates suggests a "hominoid rate slowdown" whereby mutation rates decreased in apes. (Moorjani, 2016; Scally, 2012)
+More recent studies have found that mutation rates vary among human populations. (Harris, 2015, Harris Pritchard2017)
+Converting the mutation rate per generation to a yearly rate is also met with difficulties when estimating average human generation time.
+Generation times are known to vary among current populations, and likely varied over longer time spans (Henn, Moorjani, 2016)
+
 \section{Review of current methods and limitations}
 
-Mutation rates are important for understanding the timing of evolutionary processes.
-It is used as an evolutionary clock if you assume a constant accumulation rate of neutral mutations.
-The basic equation of mutation rate is $N / LM$ where N is the number of mutations, L is the length of the region and M is the number of generations.
-Estimating mutation rates in humans has proven to be difficult for a number of reasons.
-[list of reasons here]
+The first human mutation rates were estimated by Haldane (1935) using the rates of children born with rare diseases from unaffected parents.
+However, once sequence data became available, estimates were based on phylogenetic data and fossil records.
+By using fossil record dating as an estimate for coalescence time $t$ and the sequence divergence $d$ between two species as the number of mutation that have occurred over that time span, mutation rates $\mu$ can be estimated as $\mu = 2d/t$ (Scally, 2012).
+Using this method with the time of divergence of Chimps and Humans, the mutation rates are estimated to be around $1\times10^{-9}$ up to $2.5\times10^{-8}$ (Takahata, 1997; Scally, 2012; Moorjani, 2016, Narasimah, 2016).
+The limitations of this approach are numerous, ranging from accuracy of fossil record dating, to estimating the average generation time for chimps and humans over millions of years.
 
-Three main categories of methods :
+Instead of using distant ancestry and fossil records, another more direct approach to estimating mutation rates is to sequence the genomes of parents and offspring. 
+This approach tends to give lower estimates compared to phylogenetic approach ($1-1.25\times10^{-8}$) (Narasimah, 2016).
+This rate is less than half of the rates estimated from phylogenies (Shendure, 2015). 
+One issue with trio sequencing for mutation rates is that false-positive rates of sequencing technologies are difficult to account for considering how low the true mutation rate is.
+Another issue is that the samples used for sequencing are not zygotes.
+This leaves room for germline mosaicisms to artificially increase the number of mutations observed in a generation.
 
-\subsection{Phylogenetic method :} Chimp vs human estimates $2.5\times10^{-8}$ but limitations about the number of generations that separate the two species. Based on fossil evidence.
-
-\subsection{Trio method :} One generation using trio sequencing. 
-Tend to give lower estimates compared to phylogenetic approach. ($1-1.25\times10^{-8}$)
-
-Pedigree approaches are simple in and direct approaches for per generation estimates but are met with several limitations.
-Sex bias : fathers germline mutations increase with age and could be different among men.
-Difficulty distinguishing difference between mosaicism and germline/somatic mutations.
-
-Using Trio : sequencing tech limitations in error rates.
-Mosaicism in parent or postzygotic mutations in child.
-Father's age : spermatogenesis cell division.
-Different types of mutations occur at different rates (and changes with age)
-
-Converting per generation to per year is also difficult. 
-Current approach is to divide generation rate by age of reproduction. (assumes same reproductive age for all pops over all time)
-Estimates may only be accurate to the population studied.
-
-Mutational signatures specific to populations suggest recent changes in mutation rates.
-
-One possible solution is to limit estimates to CpG sites as they appear to have more consistent mutation rates across primates.
-*(but apparently no diff in paternal age for CpG transitions vs other mutations)
-
-\subsection{Population genetics method : }
-Estimates around $1.6-1.7\times10^{-8}$
-
-Taking advantage of founder populations with known pedigrees can get better estimates of mutation rates over multiuple generations.
-Identify mutations in regions of autozygosity and count the number of generations that separate these segments. 
-Reduces errors from somatic mutations. 
-\subsection{Approach and methods for new calculation}
+Most recently, methods for estimating mutation rates using population genetics are increasingly being used.
+These approaches expand on the theory behind trio sequencing by including more distantly related individuals.
+One approach involves autozygous segments observed in a founder population with a known pedigree (Campbell, 2012).
+These segments shared by multiple individuals are compared and mutations in these segments are then used in combination with the number of generations separating two individuals (Campbell, 2012).
+Palamara et al. (2015) describe an even more generalized method by identifying IBD segments and inferring number of generations using coalescent theory.
+This method then regresses the number of mutations observed per base against the time of most recent common ancestor.
+The slope of the regression is the mutation rate per generation.
+Applying this method to 498 individuals from 250 families, they measured a mutation rate of ($2.08\times10^{-8}$).
+This method is able to infer the genotyping false-positive rate as the intercept of the regression line.
+The major limitation to this method comes from the coalescence estimations.
 
 \subsection{Apply state of the art methods to new data}
+For the third chapter of my thesis I plan on exploring mutation rate estimations using the combination of sequence data and phylogenetic data from the Quebec population.
+One preliminary objective might be to replicate the population genetics methods described above by Palamara et al. and Campbell et al..
+Knowing that mutation rates can vary across populations, applying these methods to a new dataset may in itself be worthy of publication.
 
-\subsection{can we do better?}
-Estimating mutation rate spatially across the genome.
-Using IBD and Genealogy to better estimate the number of generations.
+Considering the deep phylogenetic data available for the Quebec population, we may be able to improve on the Palamara method by acquiring more accurate estimates of generation time separating two IBD segments.
+Indeed, by linking individuals to the phylogenetic tree of the Quebec population, we could estimate the most recent common ancestor for a given segment of the genome.
 
@@
 \appendix % Cue to tell LaTeX that the following "chapters" are Appendices
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Genizon_BalSac_mapsInMaps4.jpg}
+\label{UMAP}
+\end{figure}
 

### 2018-11-22 QE/thesis_1/main.tex
Filled in the Intro

@@
 \mainmatter % Begin numeric (1,2,3...) page numbering
-
 \pagestyle{thesis} % Return the page headers back to the "thesis" style
@@
 \section{Introduction}
+Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago (Henn, 2012). 
+As humans expanded into new habitats, each new colony was founded by a smaller group departing a larger, more diverse source population.
+The impacts of the OOA bottleneck are still measurable in modern human populations, with distance from Africa being strongly correlated with decreasing genetic diversity (Ramachandran, 2005).
 
-\subsection{Serial Founder Effects}
-Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago, and resulted in a significant loss of diversity in the populations the colonized new habitats (Henn, 2012).
-This loss of diversity is the result of a population bottleneck from a small group leaving a source population. 
-Estimates of the parameters of this bottleneck using sequence data and modelling population dynamics range from 2,500 to as low as 450 individuals (Eller, 2004; Fagundes, 2007).
-However, as humans expanded into new habitats, each new colony was founded by small groups leading to successive bottlenecks and reduction in diversity.
-This trend is still apparent in modern humans, with distance from Africa being correlated with loss of genetic diversity (Ramachandran, 2005).
-Bottlenecks can drastically change the allele frequencies within the population, which can lead to rapid divergence (Rosenberg, 2002; Charlesworth, 2003).
-Moreover, because of the limited population size, inbreeding will have impacts on heterozygosity.
-Accounting for population divergence and population specific allele frequencies are necessary for disease phenotype association tests (Gravel, 2010).
+Bottlenecks can drastically change the allele frequencies within a population and lead to rapid divergences (Rosenberg, 2002; Charlesworth, 2003).
+Rare alleles in the source population can increase significantly in frequency or even become fixed in the founder population.
+Isolation over time can lead to increased diversity between small populations, while the diversity within may remain relatively low. 
+For this reason, migration between populations can increase the genetic diversity of both populations.
 
-Mutational patterns in founder populations have recently become possible to measure, and useful for quantifying changes in mutation rates, as well as demographic histories of populations.
+Recent events of admixture are tractable with genomic analyses like identity by descent (IBD) mapping and coalescence estimations (Baharian, 2016).
+However, recombination over time will fragment haplotypes eventually making it difficult to determine their ancestry.
+Despite this, a population will retain a unique set of alleles at frequencies waxing and waning over time.
+These allele frequencies can be used to estimate demographic histories of multiple populations (Gravel, 2010).
 
-Studying recent founder events can lead to better understanding of the mechanisms and dynamics of historical founder events.
+Many disease or phenotype association tests are susceptible to spurious results caused by population structure.
+In order to properly account for this, a better understanding of human demographic history is needed (Gravel, 2010).
+Moreover, studying recent founder populations can lead to better understanding of the mechanisms and dynamics of historical founder events.
 
+%The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data.
+%As a result of decreasing costs per sample, research centres are now able to perform genomic studies involving larger and larger cohorts.
+%Need, A. C. & Goldstein, D. B. Next generation disparities in human genomics: concerns and remedies. Trends Genet. 25, 489?494 (2009).
+%These large scale genomics studies have increased statistical power to a point where we can resolve fine scale mutational signatures observable on evolutionary scales by comparing the frequencies of mutations between populations.
+%These mutational patterns are useful for inferring population demographic histories, as well as variation in the human mutation rate.
+
+%Estimates of the parameters of this OOA bottleneck using sequence data and modelling population dynamics range from 2,500 to as low as 450 individuals founding Europe and Asia (Eller, 2004; Fagundes, 2007).
 \subsection{Thesis proposal}
-Aim : Investigating mutational patterns in founder populations. 
-Characterizing the genetic diversity of founder populations.
 
-Hypothesis : 
-Founder effects will influence the mutational signatures of a population leading to unique patterns in populations that can be used to infer demographic histories.
+The aim of this PhD project is to study the genomics of founder populations to infer mutational signatures, demographic history, population structure and mutation rates.
+\textbf{Our hypotheses are }: \textbf{1}. An unusual mutational patterns observed in Japan is not reproducible and is the result of a technical artifact.
+\textbf{2}. Quebec has more structure than would be expected from a series of founder events which can be accounted for by ancestral population structure predating the founder events.
+\textbf{3}. Leveraging the deep phylogenetic records from Quebec in combination with sequence data can improve the estimates of human mutation rates.
 
-Combining genetic, genealogical and geographic data can explain the population structure observed today, recreating historical demographics.
+%Combining genetic, genealogical and geographic data can explain the population structure observed today, recreating historical demographics.
+%This project will be broken down into three chapters :
+%1. Unique signature in Japan
+%Identify the source of the mutational pattern in Japanese population. 
+%2. Characterize the genetic diversity of the Quebec population (a case study)
+%Explain population structure. Using Fst and UMAP/GEO.
+%3. Explore mutation rate in Quebec population
+%apply existing methods of estimating human mutation rate.
+%can we do better? combine data to improve estimation.
 
-This project will be broken down into three chapters :
+\chapter{A curious mutational signature in Japan}
 
-1. Unique signature in Japan
-Identify the source of the mutational pattern in Japanese population.
+The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
 
-2. Characterize the genetic diversity of the Quebec population (a case study)
-Explain population structure. Using Fst and UMAP/GEO.
+The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. 
+Since then, sequencing platforms have rapidly improved. 
+Yet, because of the extraordinary value of freely available data, early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
 
-3. Explore mutation rate in Quebec population
-apply existing methods of estimating human mutation rate.
-can we do better? combine data to improve estimation.
+Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
+This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
 
-\subsection{mutational signatures}
-Some mutagenic processes will affect different DNA motifs. 
-Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
-Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
-Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
-
-In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures which can be revealed in large sequencing panels. 
+\subsection{Mutational Signatures}
+There has been recent interest in population variation in germline mutational signatures which can be revealed in large sequencing panels. 
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
@@
 
-\chapter{A curious mutational signature in Japan}
-The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
-
-The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
-Yet, because of the extraordinary value of freely available data, early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
-
-Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
-This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
-
-\section{summary of results from paper}
+\section{Results}
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
@@
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
+The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
 
-link to manuscript
+The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
+This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
+By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
+We performed the reverse GWAS approach in all populations and identified 3826 variants that were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
 
@@
 \section{Brief History of Quebec}
-The Quebec population was founded over four centuries ago from \~8,500 settlers from France.
+The Quebec population was founded over four centuries ago from around 8500 settlers from France.
 This small number of settlers are the direct ancestors of the great majority of Quebec's population of 7 million.
-In population genetics, a founder effect is characterized by a small number of founders from a larger source population giving rise to a population expansion.
+In population genetics, a founder effect is characterized by a small number of founders from a larger source population giving rise to a population expansion.[***NEEDS WORK***]
 This tends to increased homozygosity and rapidly shift allele frequencies compared to the source population.
@@
 These segments shared by multiple individuals are compared and mutations in these segments are then used in combination with the number of generations separating two individuals (Campbell, 2012).
-Palamara et al. (2015) describe an even more generalized method by identifying IBD segments and inferring number of generations using coalescent theory.
+Palamara et al. (2015) describe an even more generalized method by identifying IBD segments and inferring number of generations using coalescent theory (Baharian 2016).
 This method then regresses the number of mutations observed per base against the time of most recent common ancestor.
@@
 Indeed, by linking individuals to the phylogenetic tree of the Quebec population, we could estimate the most recent common ancestor for a given segment of the genome.
+By grouping IBD segments according to genomic regions (e.g.
+coding, CpG, etc.) and running the analysis separately would provide us with a more detailed view of the genome wide mutation rate in humans.
 

### 2018-11-22 QE/main.tex
Solid draft

--- 2018-11-22 QE/thesis_1/main.tex
@@
 headsepline, % Uncomment to get a line under the header
-%chapterinoneline, % Uncomment to place the chapter title next to the number on one line
+chapterinoneline, % Uncomment to place the chapter title next to the number on one line
 %consistentlayout, % Uncomment to change the layout of the declaration, abstract and acknowledgements pages to match the default layout
@@
 
-\usepackage[backend=bibtex,style=numeric,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
+\usepackage[backend=bibtex, style=numeric, natbib=true, sorting=none]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
 
-\addbibresource{Legacy.bib} % The filename of the bibliography
+\addbibresource{QE_bib2.bib} % The filename of the bibliography
 
 \usepackage[autostyle=true]{csquotes} % Required to generate language-dependent quotes in the bibliography
+
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd{\chapter}{\if@openright\cleardoublepage\else\clearpage\fi}{}{}{}
+\makeatother
 
@@
 	paper=a4paper, % Change to letterpaper for US letter
-	inner=2.5cm, % Inner margin
-	outer=2.5cm, % Outer margin
-	bindingoffset=.5cm, % Binding offset
+	inner=2cm, % Inner margin
+	outer=2cm, % Outer margin
+	bindingoffset=0cm, % Binding offset
 	top=1.5cm, % Top margin
@@
 
+\includegraphics[width=40mm]{./Figures/McGill.png} % University/department logo
+
 \vspace*{.06\textheight}
@@
 
-\vfill
-
-%\large \textit{A thesis submitted in fulfillment of the requirements\\ for the degree of \degreename}\\[0.3cm] % University requirement text
-%\textit{in the}\\[0.4cm]
-\groupname\\\deptname\\[2cm] % Research group name and department name
-
-\vfill
+\deptname\\\facname\\[2cm] % Research group name and department name
 
 {\large \today}\\[4cm] % Date
-%\includegraphics{Logo} % University/department logo - uncomment to place it
 
@@
 
-\mainmatter % Begin numeric (1,2,3...) page numbering
+\mainmatter % Begin numeric (1,2,3...) page numberingc
 \pagestyle{thesis} % Return the page headers back to the "thesis" style
@@
 \section{Introduction}
-Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago (Henn, 2012). 
+The human genome is composed of a mosaic of ancestral genetic material passed on for millennia.
+This pattern of interconnected relatedness is beginning to unveil humanities rich and complex history.
+Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago \citep{Henn2012}. 
 As humans expanded into new habitats, each new colony was founded by a smaller group departing a larger, more diverse source population.
-The impacts of the OOA bottleneck are still measurable in modern human populations, with distance from Africa being strongly correlated with decreasing genetic diversity (Ramachandran, 2005).
+The impacts of the OOA bottleneck are still measurable, as distance from Africa is strongly correlated with decreasing genetic diversity \citep{Ramachandran2005}.
 
-Bottlenecks can drastically change the allele frequencies within a population and lead to rapid divergences (Rosenberg, 2002; Charlesworth, 2003).
+Bottlenecks can drastically change the allele frequencies within a population and lead to rapid divergences \citep{Rosenberg2002, Charlesworth2003}.
 Rare alleles in the source population can increase significantly in frequency or even become fixed in the founder population.
 Isolation over time can lead to increased diversity between small populations, while the diversity within may remain relatively low. 
-For this reason, migration between populations can increase the genetic diversity of both populations.
+For this reason, migration between populations can increase the genetic diversity of both populations \citep{Charlesworth2003}.
 
-Recent events of admixture are tractable with genomic analyses like identity by descent (IBD) mapping and coalescence estimations (Baharian, 2016).
+Recent events of admixture are tractable with genomic analyses like identity by descent (IBD) mapping and coalescence estimations \citep{Baharian2016}.
 However, recombination over time will fragment haplotypes eventually making it difficult to determine their ancestry.
 Despite this, a population will retain a unique set of alleles at frequencies waxing and waning over time.
-These allele frequencies can be used to estimate demographic histories of multiple populations (Gravel, 2010).
+These allele frequencies can then be used to estimate demographic histories of multiple populations \citep{Gravel2011}.
 
 Many disease or phenotype association tests are susceptible to spurious results caused by population structure.
-In order to properly account for this, a better understanding of human demographic history is needed (Gravel, 2010).
+In order to properly account for this, a better understanding of human demographic history is needed \citep{Gravel2011}.
 Moreover, studying recent founder populations can lead to better understanding of the mechanisms and dynamics of historical founder events.
 
-%The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data.
-%As a result of decreasing costs per sample, research centres are now able to perform genomic studies involving larger and larger cohorts.
-%Need, A. C. & Goldstein, D. B. Next generation disparities in human genomics: concerns and remedies. Trends Genet. 25, 489?494 (2009).
-%These large scale genomics studies have increased statistical power to a point where we can resolve fine scale mutational signatures observable on evolutionary scales by comparing the frequencies of mutations between populations.
-%These mutational patterns are useful for inferring population demographic histories, as well as variation in the human mutation rate.
+\subsection{Thesis proposal}
+The aim of this PhD project is to study the genomics of founder populations to infer population specific mutational signatures, demographic history and mutation rates.
+\textbf{Our hypotheses are}: \textbf{1}. An unusual non reproducible mutational pattern observed in Japan is the result of a technical artifact.
+\textbf{2}. The heterogeneity of Quebec's population structure can be accounted for by ancestral population structure predating the colonization of Canada.
+\textbf{3}. Leveraging phylogenetic records in combination with sequence data can improve the estimates of mutation rates.
+\vfill
+\chapter{Mutational signatures in Japan}
 
-%Estimates of the parameters of this OOA bottleneck using sequence data and modelling population dynamics range from 2,500 to as low as 450 individuals founding Europe and Asia (Eller, 2004; Fagundes, 2007).
-\subsection{Thesis proposal}
-
-The aim of this PhD project is to study the genomics of founder populations to infer mutational signatures, demographic history, population structure and mutation rates.
-\textbf{Our hypotheses are }: \textbf{1}. An unusual mutational patterns observed in Japan is not reproducible and is the result of a technical artifact.
-\textbf{2}. Quebec has more structure than would be expected from a series of founder events which can be accounted for by ancestral population structure predating the founder events.
-\textbf{3}. Leveraging the deep phylogenetic records from Quebec in combination with sequence data can improve the estimates of human mutation rates.
-
-%Combining genetic, genealogical and geographic data can explain the population structure observed today, recreating historical demographics.
-%This project will be broken down into three chapters :
-%1. Unique signature in Japan
-%Identify the source of the mutational pattern in Japanese population. 
-%2. Characterize the genetic diversity of the Quebec population (a case study)
-%Explain population structure. Using Fst and UMAP/GEO.
-%3. Explore mutation rate in Quebec population
-%apply existing methods of estimating human mutation rate.
-%can we do better? combine data to improve estimation.
-
-\chapter{A curious mutational signature in Japan}
-
+\subsection{Batch Effects in Aging Reference Cohort Data}	
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
@@
 \section{Results}
+\subsection{A peculiar mutational signature in Japan}
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
 While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
-When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
 These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests.
-
-To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
+\subsection{Reverse GWAS}
+To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the Japanese from the 1kGP (JPT) for variants that associate strongly with low average quality of mapped bases ($Q$) (Figure \ref{SFS}).
 Traditionally, genome wide association studies use genotypes as the independent variable. 
@@
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
-The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
-These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
 
+\subsection{Identifying Suspicious SNPs in the 1kGP}
 The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
@@
 
-\chapter{Characterization of the Quebec population}
+\subsection{Suspicious variants impact modern genomics analyses}
+We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
+The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
+Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
+The third study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
+Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
+
+The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
+It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
+
+\subsection{Recommendations}
+A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
+In this case, we used a cut-off of $Q$ over 30.
+Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations.
+\vfill
+\chapter{Population structure of Quebec}
 
 \section{Brief History of Quebec}
-The Quebec population was founded over four centuries ago from around 8500 settlers from France.
+The Quebec population was founded over four centuries ago from around 8500 settlers from France \citep{Bherer2011,Laberge2005}.
 This small number of settlers are the direct ancestors of the great majority of Quebec's population of 7 million.
-In population genetics, a founder effect is characterized by a small number of founders from a larger source population giving rise to a population expansion.[***NEEDS WORK***]
-This tends to increased homozygosity and rapidly shift allele frequencies compared to the source population.
+In addition, ancestral contribution is uneven, whereby two thirds of the modern gene pool is attributed to the first 2600 founders \citep{Gagnon2001, Laberge2005}.
+While the rates of consanguineous unions in Quebec were not higher than in Europe, small population sizes and isolation over time led to increased distant consanguinity \citep{Laberge2005}.
 
 The north east of Quebec (e.g.
-Saguenay-Lac-Saint-Jean and Charlevoix) displays higher rates of some rare genetic diseases and increased homozygosity which is suggestive of a severe founder effect.
-(Gagnon, 2000)
-Western Quebec, on the other hand, has similar prevalence of diseases compared to Europe. (Gagnon, 2000)
-This discrepancy of disease rates across the Saint-Lawrence valley indicates that the severity of the bottleneck varies across geographic regions. (Gagnon, 2000)
+Saguenay-Lac-Saint-Jean and Charlevoix) displays higher rates of some rare genetic diseases and increased homozygosity which is suggestive of a severe founder effect \citep{Gagnon2001}.
+Western Quebec, on the other hand, has similar prevalence of diseases compared to Europe \citep{Gagnon2001}.
+This discrepancy of disease rates across the Saint-Lawrence valley indicates that the severity of the bottleneck varies across geographic regions \citep{Gagnon2001}.
 
-Indeed, the intensity of the founder effect in Quebec is not consistent across the population as there is evidence of geographical stratification which has been attributed to a number of factors like varying concentration of founders, native ancestry and geographic isolation.
-(Gagnon, 2000).
-The founders of Quebec were not all from the same regions in France, providing some initial diversity despite a small initial population size.
-Moreover, Native American admixture with founders has been estimated to be around 1\% which would have also been a source of genetic diversity among these settlements. (Moreau, 2013)
+Indeed, the intensity of the founder effect in Quebec is not consistent across the population as there is evidence of geographical stratification which has been attributed to a number of factors like varying concentration of founders, native ancestry and geographic isolation \citep{Gagnon2001,Laberge2005,Bherer2011}.
+The founders of Quebec were not all from the same regions in France, providing some initial diversity despite small initial population sizes \citep{Laberge2005}.
+Moreover, Native American admixture with founders has been estimated to be around 1\% which would have also been a source of genetic diversity among these settlements \citep{Moreau2013}.
 Some settlements were more isolated than others resulting in less gene flow between sub-populations.
@@
 
-\subsection{Quebec : A case study of founder populations}
+\subsection{A case study of founder populations}
 
 The Quebec founder effect is an interesting case study for population genetics for multiple reasons.
-Firstly, the source population of this bottleneck is still present, which may not be the case for older founder populations. 
-This provides us with the possibility of comparing allele frequencies between Quebec and its ancestral population.
+Firstly, the source population of this bottleneck is still present, which may not be the case for more ancient founder populations. 
+This provides us with the possibility of comparing allele frequencies between Quebec and its ancestral populations of French Atlantic seaports and Paris.
 Secondly, many populations experienced founder effects as humans expanded out of Africa and populated the continents.
 Studying examples of founder effects in human populations are a great way to improve our understanding of human demographic histories.
-Finally, one unique aspect of this population is the deep genealogical data that spans four centuries and includes nearly 3 million entries.
-Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events.
+Finally, one unique aspect of this population is the deep genealogical data that spans four centuries and includes over 3 million entries.
+Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events \citep{Larmuseau2013}.
 
 There have been several studies in the past few decades that have examined the genetic diversity and population structure of Quebec.
-Using genotype data, linkage disequilibrium and homozygosity analyses of sub-populations within Quebec showed significant structure and relatedness that persists in the current population (Roy-Gagnon, 2011).
+Using genotype data, linkage disequilibrium and homozygosity analyses of sub-populations within Quebec showed significant structure and relatedness that persists in the current population \citep{Bherer2010,Gagnon2011,Roy-Gagnon2011}.
 Quebec is not a homogeneous population, but rather made up of a combination of distinct sub-populations.
-This was confirmed with Fst statistics of these sub-populations, indicating substantial genetic structure.
-Of particular interest, they reported that there was greater differentiation between sub-populations of Quebec than between European populations.
+This was confirmed with Fst statistics of these groups, indicating substantial genetic structure.
+Of particular interest, Roy-Gagnon (2011) reported that there was greater differentiation between sub-populations of Quebec than between European populations.
 While this study included 140 genotyped individuals - a small sample by today's standards - they postulate that this difference is accounted for by genetic drift.
-Indeed, the heterogeneity of the founder effect in the Quebec population could be explained by a series of separate founder events influenced by variation in the relative isolation, and migration between settlements.
+Alternatively, the heterogeneity of the founder effect in the Quebec population could be explained by a series of separate founder events influenced by variation in the relative isolation, and migration between settlements \citep{Gagnon2001,Laberge2005}.
 
-\section{Genomic and genealogical data}
+\section{Materials and Methods}
 
-The second chapter of my thesis will examine the genetic structure of the Quebec population using more recent and larger genomic datasets in combination with the deep-genealogical and geographical data.
+While the deep-genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes \citep{Gagnon2001,Roy-Gagnon2011}.
+The aim of the second chapter of this thesis will be to examine the genetic structure of the Quebec population using more recent and larger genomic datasets in combination with the deep-genealogical and geographical data.
+Recently developed methodologies will also be of use to further characterize this genomic landscape.
 
-While the deep-genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes.
+We plan on using two main cohorts for this chapter. 
+The Genizon cohort has 9961 samples from Quebec, genotyped on a variety of chips.
+While this data has yet to be published, it has required a fair amount of cleaning and processing to account for the different genotyping chips used.
+One key advantage of using this data is that we already have consent to link 516 individuals to genealogical records.
+We will also include data from the Cartagene project which has over 12,000 sequenced Quebecers.
 
-Genizon data has 000 samples, genotyped using a variety of chips. These data require some cleaning and processing before analysis. This cohort has yet to be published.
+The BALSAC database is a genealogical tree of the French-Canadian Quebec population.
+It was compiled from birth, marriage and death certificates from the Catholic Church.
+Because of the major role the Church played in French-Canadian society, these records are remarkably detailed and comprehensive.
+As mentioned above, 516 individuals from the Genizon cohort have been genotyped and are linked to this genealogy.
 
-We will also include data from the Cartagene project which has 000 samples from...
-
-With these data, we hope to reproduce the results from previous work and answer the following question :
-(2) Are the sub-populations of Quebec the result of genetic drift from founder events and isolation, or the preservation of genetic structure from France?
-
-\subsection{approach and methods}
-Using Fst statistics we can infer the relative differences between sub-populations of Quebec, as done in (Roy-Gagnon, 2011).
-Dominic Nelson has written software that can produce forward simulations of the Quebec population using the genealogical data from BALSAC.
+\subsection{Founder effect or Ancient population structure?}
+Using Fst statistics we can infer the relative differences between sub-populations of Quebec (see \citep{Roy-Gagnon2011}).
+Dominic Nelson has written software that can produce single locus forward simulations of the Quebec population using the genealogical data from BALSAC.
 These simulations can be used to produce estimated differences between sub-populations resulting purely from a founder effect.
-
 Comparing the Fst statistics between the real data and simulations will determine if the structure observed in Quebec is the result of preserved genetic structure from France or the result of drift from founder events.
 
-\section{Linking population structure to geography (results)}
+\section{Results : the geography of genetics}
+
 Preliminary analyses of the Genizon dataset are in agreement with previous work.
-In collaboration with Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data.
-This method takes the first *10* principle components of the data as input and reduces this high dimensional data preserving both local and global structure.
+In collaboration with Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data \citep{mcinnes2018umap,diaz2018}.
+This method takes the first 20 principle components of the data as input and reduces this high dimensional data preserving both local and global structure (see \citep{diaz2018} and \citep{mcinnes2018umap}).
 This projection shows significant population structure within the Quebec population.
-These data cluster into 5 roughly defined sub-populations, in agreement with what (Roy-Gagnon, 2011) observed \ref{Cluster}.
 
-We colored these plots based on the three dimensional (X,Y,Z) coordinates of the UMAP projection space converted to three dimensional color space (R,G,B).
-Thus, points that are close together in UMAP space will have similar color.
-
-We were able to link 000 of the Genizon individuals to the BALSAC database which includes marriage location of their ancestors.
-Combining the ancestral locations of these individuals with the three dimensional UMAP coloring scheme, geographic clustering of the sub-populations becomes apparent \ref{UMAP}.
-
+We colored these plots based on the three dimensional coordinates of the UMAP projection space (X,Y,Z) converted to three dimensional color space (R,G,B).
+Therefore, Individuals with similar color are closer to each other in the projection and presumably share more recent ancestry.
+To link the genetic data to geographic coordinates, we used the marriage locations of the ancestors of those linked to the BALSAC database.
+Combining the ancestral locations of these individuals with the three dimensional UMAP coloring scheme, geographic clustering of the sub-populations becomes apparent.
+Five genetic clusters are visible in the UMAP projection, these clusters also appear to be geographically linked.
+This is in agreement with what Roy-Gagnon observed \citep{Roy-Gagnon2011}.
 These preliminary results are very promising as they appear to capture the geographic and genetic clustering of the Quebec population.
 
-\section{Future Directions}
-%We know that there are regional differences in disease rates that are linked to population substructure.
-%The genetic data used in UMAP analysis also identify similar clusters.
-%Identifying the variants that are responsible for the clustering could also provide us with a shortlist of variants that could also be linked to the diseases that are associated with these sub-populations.
-%This method has yet to be formalized, but in brief, performing a logistic regression predicting presence of genotypes using the axis of variation that separates one group from the others could be used to identify these variants.
-While Fst gives a good estimate of genome wide past coalescent rates, the analysis of long IBD segments can provide us with more detailed insight of more recent demographic history (Ralph, 2013; Charlesworth 2003).
-Because Fst takes in all polymorphisms as input for the statistic, it does not distinguish between recent and more ancient common ancestry.
+\section{Future directions}
+While Fst gives a good estimate of genome wide past coalescent rates, we plan on analyzing IBD segments as they can provide us with more detailed insight of demographic history \citep{Charlesworth2003, Ralph2013, Baharian2016}.
+Because Fst takes in all polymorphisms as input for the statistic, it does not distinguish between recent and more ancient common ancestry \citep{Ralph2013}.
 Analysis of IBD segments shared among individuals from Quebec can be useful for reconstructing the recent structure of relatedness.
-The length of the shared segments can be used to infer the number of generations leading back to a common ancestor.
-Combining this with geographic data can infer patterns of shared ancestry across the Saint-Lawrence valley.
-Ralph (2013) identified a general trend of decreased of shared ancestry with geographic distance as well as more regional signals of IBD in European populations.
+The length of the shared segments can be used to estimate the number of generations leading back to a common ancestor \citep{Palamara2015,Campbell2012,Baharian2016}.
+Combining this with geographical data can infer patterns of shared ancestry across the Saint-Lawrence valley.
+Ralph (2013) identified a general trend of decreasing recent common ancestry with geographic distance as well as distinct regional signals of IBD in different European populations \citep{Ralph2013}.
 Europe has a rich history of complex migratory patterns and demographic histories that is difficult to trace back with certainty. 
@@
 
-While studying Quebec populations on their own can provide detailed historical inference of demographics, including samples from other European populations can provide more context.
-Expanding the current analyses to include data from French individuals could be used to confirm the results from the Fst analysis as well as provide insight into which regions in France are source populations to regions in Quebec.
-French genome data is obtainable through Dr. Christian Dina, one of our collaborators in the University of Nantes.
+While studying the French-Canadian populations on their own can provide a detailed historical inference of demographics, including samples from other European populations can provide more context.
+We plan on expanding the current analyses to include data from French individuals which could be used to confirm the results from the Fst analysis as well as provide insight into which regions in France are source populations to regions in Quebec.
+Dr. Christian Dina in the University of Nantes in France has expressed interest in sharing French genomics data for this project.
+\clearpage
 
-\chapter{Estimating mutation rates}
+\chapter{Estimating human mutation rates}
 
 Mutation rate estimations are necessary for understanding the timing of evolutionary processes and events.
-Neutral mutations are expected to arise at a constant rate in the genome, and can therefore be used as an evolutionary clock that can be used to estimate divergence times (Moorjani, 2016). 
-The basic equation of mutation rate is $N / LM$ where N is the number of mutations, L is the length of the region and M is the number of generations. (Narasimah, 2016)
+Neutral mutations are expected to arise at a constant rate in the genome, and can therefore be used as an evolutionary clock that can estimate divergence times \citep{Moorjani2016}. 
+The basic equation of mutation rate is $\mu = N / LM$ where N is the number of mutations, L is the length of the region and M is the number of generations \citep{Narasimhan2016}.
 Estimating mutation rates in humans has proven to be difficult for a number of reasons.
 Mutation rates vary across the genome as some regions are more conserved than others. 
-Mutation rates also vary over time.
-Shorter branches within the Ape family compared to other primates suggests a "hominoid rate slowdown" whereby mutation rates decreased in apes. (Moorjani, 2016; Scally, 2012)
-More recent studies have found that mutation rates vary among human populations. (Harris, 2015, Harris Pritchard2017)
-Converting the mutation rate per generation to a yearly rate is also met with difficulties when estimating average human generation time.
-Generation times are known to vary among current populations, and likely varied over longer time spans (Henn, Moorjani, 2016)
+Mutation rates are also known to vary over time as shorter branches within the Ape family compared to other primates suggest a "hominoid rate slowdown" whereby mutation rates decreased in apes \citep{Moorjani2016, Scally2012}.
+In addition, more recent studies have found that mutation rates can even vary among human populations. \citep{Harris2015a, Harris2017a}.
+%Converting the mutation rate per generation to a yearly rate is also met with difficulties when estimating average human generation time.
+%Generation times are known to vary among current populations, and likely varied over longer time spans \citep{Henn2012, Moorjani2016}
 
@@
 
-The first human mutation rates were estimated by Haldane (1935) using the rates of children born with rare diseases from unaffected parents.
+The first human mutation rates were estimated by Haldane (1935) using the rates of children born with rare diseases from unaffected parents \citep{Haldane1935}.
 However, once sequence data became available, estimates were based on phylogenetic data and fossil records.
-By using fossil record dating as an estimate for coalescence time $t$ and the sequence divergence $d$ between two species as the number of mutation that have occurred over that time span, mutation rates $\mu$ can be estimated as $\mu = 2d/t$ (Scally, 2012).
-Using this method with the time of divergence of Chimps and Humans, the mutation rates are estimated to be around $1\times10^{-9}$ up to $2.5\times10^{-8}$ (Takahata, 1997; Scally, 2012; Moorjani, 2016, Narasimah, 2016).
+By using fossil record dating as an estimate for coalescence time $t$ and the sequence divergence $d$ between two species as the number of mutation that have occurred over that time span, mutation rates $\mu$ can be estimated as $\mu = 2d/t$ \citep{Scally2012}.
+Using this method with the time of divergence of Chimps and Humans of 6 to 10 million years, the mutation rates are estimated to be around $1\times10^{-9}$ up to $2.5\times10^{-8}$ \citep{Takahata1997, Scally2012, Moorjani2016, Narasimhan2016}.
 The limitations of this approach are numerous, ranging from accuracy of fossil record dating, to estimating the average generation time for chimps and humans over millions of years.
@@
 Instead of using distant ancestry and fossil records, another more direct approach to estimating mutation rates is to sequence the genomes of parents and offspring. 
-This approach tends to give lower estimates compared to phylogenetic approach ($1-1.25\times10^{-8}$) (Narasimah, 2016).
-This rate is less than half of the rates estimated from phylogenies (Shendure, 2015). 
+This approach tends to give estimates that are less than half of the rate compared to the phylogenetic approach ($1-1.25\times10^{-8}$) \citep{Shendure2015,Narasimhan2016}.
 One issue with trio sequencing for mutation rates is that false-positive rates of sequencing technologies are difficult to account for considering how low the true mutation rate is.
@@
 
-Most recently, methods for estimating mutation rates using population genetics are increasingly being used.
+Methods for estimating mutation rates using population genetics are increasingly being used.
 These approaches expand on the theory behind trio sequencing by including more distantly related individuals.
-One approach involves autozygous segments observed in a founder population with a known pedigree (Campbell, 2012).
-These segments shared by multiple individuals are compared and mutations in these segments are then used in combination with the number of generations separating two individuals (Campbell, 2012).
-Palamara et al. (2015) describe an even more generalized method by identifying IBD segments and inferring number of generations using coalescent theory (Baharian 2016).
+One approach involves autozygous segments observed in a founder population with a known pedigree \citep{Campbell2012}.
+These segments shared by multiple individuals are compared and mutations in these segments are then used in combination with the number of generations separating two individuals \citep{Campbell2012}.
+Using autozygous regions from 70 trio sequenced Hutterite families, they obtained an estimated mutation rate of $1.2\times10^{-8}$.
+Palamara et al.
+(2015) describe an even more generalized method by identifying IBD segments from unrelated individuals and inferring the number of generations using coalescent theory \citep{Palamara2015,Baharian2016}.
 This method then regresses the number of mutations observed per base against the time of most recent common ancestor.
-The slope of the regression is the mutation rate per generation.
+The slope of the regression is the mutation rate per generation and the intercept is the inferred genotyping false-positive rate.
 Applying this method to 498 individuals from 250 families, they measured a mutation rate of ($2.08\times10^{-8}$).
-This method is able to infer the genotyping false-positive rate as the intercept of the regression line.
-The major limitation to this method comes from the coalescence estimations.
+%The major limitation to this method comes from the coalescence estimations.
 
-\subsection{Apply state of the art methods to new data}
-For the third chapter of my thesis I plan on exploring mutation rate estimations using the combination of sequence data and phylogenetic data from the Quebec population.
-One preliminary objective might be to replicate the population genetics methods described above by Palamara et al. and Campbell et al..
+\subsection{Applying new methods to new data}
+The aim of the third chapter of this thesis will be to explore mutation rate estimations using a combination of sequence data and genealogical data from the French-Canadian population.
+One preliminary objective might be to replicate the population genetics methods described above by Palamara et al. and Campbell et al. \citep{Palamara2015,Campbell2012}.
 Knowing that mutation rates can vary across populations, applying these methods to a new dataset may in itself be worthy of publication.
 
-Considering the deep phylogenetic data available for the Quebec population, we may be able to improve on the Palamara method by acquiring more accurate estimates of generation time separating two IBD segments.
-Indeed, by linking individuals to the phylogenetic tree of the Quebec population, we could estimate the most recent common ancestor for a given segment of the genome.
-By grouping IBD segments according to genomic regions (e.g.
+Considering the deep genealogical records available for the French-Canadian population, we may be able to improve on the Palamara method by acquiring more accurate estimates of generation time separating two IBD segments.
+Indeed, by linking individuals to the phylogenetic tree of the Quebec population, we may be able to better estimate the most recent common ancestor for a given segment of the genome.
+Furthermore, by grouping IBD segments according to genomic regions (e.g.
 coding, CpG, etc.) and running the analysis separately would provide us with a more detailed view of the genome wide mutation rate in humans.
@@
 %----------------------------------------------------------------------------------------
-
+\clearpage
 \appendix % Cue to tell LaTeX that the following "chapters" are Appendices
@@
 \begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Figure1.jpg}
+\caption{
+\textbf{A} 
+Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
+The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern.
+There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations.
+These three enrichments can be summarized as G**${\rightarrow}$GG*.
+(note: the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*)
+\textbf{B} 
+Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
+Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
+The histogram on the left of the plot is the distribution of significant variants. 
+\textbf{C} 
+Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1kGP.
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
+The same analysis was performed independently for each of the populations in the 1kGP. }
+ \label{SFS}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.85\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
+
+\caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
+Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
+The x-axis is ranked by the mean sequencing date per population.
+\textbf{B} The x-axis is sorted by the sequencing date per individual.
+The colors indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project.
+The bottom plot indicates the sequencing technologies used over time.}
+\label{MapQual}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
+
+\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.
+The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
+The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+Phase 1 populations are marked by a star ( * ).
+The line plot underneath shows the number of populations for which a variant has reached significance.
+The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
+The same variants identified as being low quality independently in each population are found in other populations. }
+  \label{OverLap}
+\end{figure}
+
+\begin{figure}
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/Genizon_BalSac_mapsInMaps4.jpg}
+
+\caption{A map of Quebec colored by a 3D UMAP projection. 
+The coloring of each individual is based on the three dimensional (X,Y,Z) coordinates of the UMAP projection space converted to three dimensional color space (R,G,B).
+Therefore, Individuals with similar color are closer to each other in the projection.
+Five genetic clusters are visible in the UMAP projection, these clusters also appear to be geographically linked. 
+A flattened 2D view of of the 3D projection used for coloring is presented in the bottom right.}
 \label{UMAP}
@@
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/5Clusters.png}
-\label{Cluster}
-\end{figure}
+\clearpage
+%\begin{figure}
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/5Clusters.png}
+%\label{Cluster}
+%\end{figure}
 

### 2018-11-23 QE/main.tex
minor edits

@@
 
-\section{Brief History of Quebec}
+\section{Brief history of Quebec}
 The Quebec population was founded over four centuries ago from around 8500 settlers from France \citep{Bherer2011,Laberge2005}.
 This small number of settlers are the direct ancestors of the great majority of Quebec's population of 7 million.
-In addition, ancestral contribution is uneven, whereby two thirds of the modern gene pool is attributed to the first 2600 founders \citep{Gagnon2001, Laberge2005}.
+Moreover, ancestral contributions of these settlers across the Saint-Lawrence valley is uneven, whereby two thirds of the modern gene pool is attributed to the first 2600 founders \citep{Gagnon2001, Laberge2005}.
 While the rates of consanguineous unions in Quebec were not higher than in Europe, small population sizes and isolation over time led to increased distant consanguinity \citep{Laberge2005}.
 
-The north east of Quebec (e.g.
-Saguenay-Lac-Saint-Jean and Charlevoix) displays higher rates of some rare genetic diseases and increased homozygosity which is suggestive of a severe founder effect \citep{Gagnon2001}.
-Western Quebec, on the other hand, has similar prevalence of diseases compared to Europe \citep{Gagnon2001}.
-This discrepancy of disease rates across the Saint-Lawrence valley indicates that the severity of the bottleneck varies across geographic regions \citep{Gagnon2001}.
-
-Indeed, the intensity of the founder effect in Quebec is not consistent across the population as there is evidence of geographical stratification which has been attributed to a number of factors like varying concentration of founders, native ancestry and geographic isolation \citep{Gagnon2001,Laberge2005,Bherer2011}.
-The founders of Quebec were not all from the same regions in France, providing some initial diversity despite small initial population sizes \citep{Laberge2005}.
-Moreover, Native American admixture with founders has been estimated to be around 1\% which would have also been a source of genetic diversity among these settlements \citep{Moreau2013}.
-Some settlements were more isolated than others resulting in less gene flow between sub-populations.
-The north-east of Quebec is characterized historically as being more isolated with higher emigration rates compared to western Quebec.
+The North-East of Quebec is characterized historically as being more isolated with higher emigration rates compared to Western Quebec.
+Saguenay-Lac-Saint-Jean and Charlevoix also display higher rates of some rare genetic diseases as well increased homozygosity compared to Western Quebec \citep{Gagnon2001}.
+Variation of disease prevalence across the Saint-Lawrence valley indicate that population structure fluctuates across geographic regions \citep{Gagnon2001}.
+Indeed, this pattern of genetic diversity is attributed to a number of factors like concentration of founders, native ancestry and geographic isolation \citep{Gagnon2001,Laberge2005,Bherer2011}.
+The founders of Quebec are not all from the same regions in France.
+In the seventeenth-century, while France was evenly populated, there were not many marriages across great distances \citep{Laberge2005}.
+French settlers came from regions around Paris and Atlantic port towns.
+Each group of settlers carried a distinct gene pool and founded a region of 'Nouvelle-France'.
+The population structure from each of these regions in France provides initial diversity despite small population sizes \citep{Laberge2005}.
+Moreover, Native American ancestry has been estimated to be around 1\% among these settlements \citep{Moreau2013}.
 
@@
 
-The Quebec founder effect is an interesting case study for population genetics for multiple reasons.
-Firstly, the source population of this bottleneck is still present, which may not be the case for more ancient founder populations. 
-This provides us with the possibility of comparing allele frequencies between Quebec and its ancestral populations of French Atlantic seaports and Paris.
-Secondly, many populations experienced founder effects as humans expanded out of Africa and populated the continents.
+Many populations experienced founder effects as humans expanded out of Africa and populated the continents.
 Studying examples of founder effects in human populations are a great way to improve our understanding of human demographic histories.
-Finally, one unique aspect of this population is the deep genealogical data that spans four centuries and includes over 3 million entries.
-Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events \citep{Larmuseau2013}.
+The Quebec population is an good case study for population genetics for multiple reasons.
+The source population of this bottleneck is still present, which may not be the case for more ancient migration events. 
+Comparing allele frequencies between France and Quebec can determine more accurately the nature of the colonization.
+The French-Canadian population also has deep genealogical records that span four centuries and includes over 3 million entries.
+Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events \citep{Larmuseau2013,Bherer2010,Gagnon2011,Roy-Gagnon2011}.
 
-There have been several studies in the past few decades that have examined the genetic diversity and population structure of Quebec.
-Using genotype data, linkage disequilibrium and homozygosity analyses of sub-populations within Quebec showed significant structure and relatedness that persists in the current population \citep{Bherer2010,Gagnon2011,Roy-Gagnon2011}.
+There have been several studies in the past few decades that have examined the population structure of Quebec.
+By analyzing the genealogical data they described distinct sub-populations and inferred the relative contribution of founders \citep{Gagnon2001, Bherer2010}.
+Genotype data used to analyze linkage disequilibrium and homozygosity has also revealed significant structure measurable today \citep{ Roy-Gagnon2011}.
 Quebec is not a homogeneous population, but rather made up of a combination of distinct sub-populations.
-This was confirmed with Fst statistics of these groups, indicating substantial genetic structure.
-Of particular interest, Roy-Gagnon (2011) reported that there was greater differentiation between sub-populations of Quebec than between European populations.
+This was confirmed with Fst statistics of these groups, indicating substantial genetic structure \citep{Roy-Gagnon2011}.
+Roy-Gagnon (2011) also reported that there was greater differentiation between sub-populations of Quebec than between European populations.
 While this study included 140 genotyped individuals - a small sample by today's standards - they postulate that this difference is accounted for by genetic drift.

### 2018-11-23 QE/main.tex
edits

@@
 The human genome is composed of a mosaic of ancestral genetic material passed on for millennia.
-This pattern of interconnected relatedness is beginning to unveil humanities rich and complex history.
+This pattern of interconnected relatedness is beginning to unveil humanity's rich and complex history.
 Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago \citep{Henn2012}. 
@@
 The French-Canadian population also has deep genealogical records that span four centuries and includes over 3 million entries.
-Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events \citep{Larmuseau2013,Bherer2010,Gagnon2011,Roy-Gagnon2011}.
+Quebec is not the only recent founder population, but it is among the largest and in combination with its rich genealogical records, it makes it an excellent cohort to study the impact of founder events \citep{Larmuseau2013,Bherer2011,Gagnon2001,Roy-Gagnon2011}.
 
 There have been several studies in the past few decades that have examined the population structure of Quebec.
-By analyzing the genealogical data they described distinct sub-populations and inferred the relative contribution of founders \citep{Gagnon2001, Bherer2010}.
+By analyzing the genealogical data they described distinct sub-populations and inferred the relative contribution of founders \citep{Gagnon2001, Bherer2011}.
 Genotype data used to analyze linkage disequilibrium and homozygosity has also revealed significant structure measurable today \citep{ Roy-Gagnon2011}.
@@
 \caption{A map of Quebec colored by a 3D UMAP projection. 
-The coloring of each individual is based on the three dimensional (X,Y,Z) coordinates of the UMAP projection space converted to three dimensional color space (R,G,B).
+The coloring of each individual is based on the three dimensional coordinates of the UMAP projection space (X,Y,Z) converted to three dimensional color space (R,G,B).
 Therefore, Individuals with similar color are closer to each other in the projection.
 Five genetic clusters are visible in the UMAP projection, these clusters also appear to be geographically linked. 
-A flattened 2D view of of the 3D projection used for coloring is presented in the bottom right.}
+A flattened 2D view of of the 3D projection used for coloring is in the bottom right.}
 \label{UMAP}

### 2018-11-25 QE/main.tex
finished

@@
 
+\newgeometry{
+	paper=a4paper, % Change to letterpaper for US letter
+	inner=2cm, % Inner margin
+	outer=2cm, % Outer margin
+	bindingoffset=0cm, % Binding offset
+	top=3cm, % Top margin
+	bottom=1.5cm, % Bottom margin
+	%showframe, % Uncomment to show how the type block is set on the page
+}
+
 \section{Introduction}
-The human genome is composed of a mosaic of ancestral genetic material passed on for millennia.
+The human genome is composed of a mosaic of ancestral genetic material that has been passed on for millennia.
 This pattern of interconnected relatedness is beginning to unveil humanity's rich and complex history.
-Human expansion out of Africa (OOA) is estimated to have occurred between 45,000 and 60,000 years ago \citep{Henn2012}. 
+For example, sequence data estimates that Human expansion out of Africa (OOA) occurred between 45,000 and 60,000 years ago \citep{Henn2012}. 
 As humans expanded into new habitats, each new colony was founded by a smaller group departing a larger, more diverse source population.
@@
 In order to properly account for this, a better understanding of human demographic history is needed \citep{Gravel2011}.
-Moreover, studying recent founder populations can lead to better understanding of the mechanisms and dynamics of historical founder events.
+Moreover, studying recent founder populations can lead to a better understanding of the mechanisms and dynamics of historical founder events.
 
@@
 The aim of this PhD project is to study the genomics of founder populations to infer population specific mutational signatures, demographic history and mutation rates.
-\textbf{Our hypotheses are}: \textbf{1}. An unusual non reproducible mutational pattern observed in Japan is the result of a technical artifact.
-\textbf{2}. The heterogeneity of Quebec's population structure can be accounted for by ancestral population structure predating the colonization of Canada.
+\textbf{Our hypotheses are}:
+
+\textbf{1}. Using higher quality data can clarify the nature of the mutational pattern observed in Japan. Alternatively, the mutational pattern may be a technical artifact unique to the original data.
+
+\textbf{2}. The heterogeneity of Quebec's population structure can be accounted for both by a series of founder events and ancestral population structure predating the colonization of Canada.
+
 \textbf{3}. Leveraging phylogenetic records in combination with sequence data can improve the estimates of mutation rates.
 \vfill
+\restoregeometry 
 \chapter{Mutational signatures in Japan}
@@
 In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
-Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
+Strong population enrichments of a mutational signature suggest important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
 Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.
 
-In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Project (1kGP) Japanese individuals.
 This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
@@
 \subsection{Reverse GWAS}
-To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the Japanese from the 1kGP (JPT) for variants that associate strongly with low average quality of mapped bases ($Q$) (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study on the Japanese from the 1kGP (JPT) for variants that associate strongly with low average quality of mapped bases ($Q$) (Figure \ref{SFS}).
 Traditionally, genome wide association studies use genotypes as the independent variable. 
@@
 
-The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
-It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
+The HapMap project and 1kGP shared some cell lines but used different sequencing platforms to produce the data.
+It is possible that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
 
-\subsection{Recommendations}
-A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-In this case, we used a cut-off of $Q$ over 30.
-Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations.
+\section{Conclusion}
+The genome wide simple logistic regression analysis approach is rather straightforward, and could be a standard in a variety of -omics studies.
+Given rapid technological progress, the focus of genomic research is naturally on the data generation side. 
+However, cleaning up data is also important to avoid generating spurious results. 
+The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing. 
 \vfill
+
 \chapter{Population structure of Quebec}
@@
 \section{Brief history of Quebec}
-The Quebec population was founded over four centuries ago from around 8500 settlers from France \citep{Bherer2011,Laberge2005}.
-This small number of settlers are the direct ancestors of the great majority of Quebec's population of 7 million.
+The French-Canadian population of Quebec was founded over four centuries ago from around 8500 settlers from France \citep{Bherer2011,Laberge2005}.
+This small number of settlers are the direct ancestors of the majority of Quebec's population of 7 million.
 Moreover, ancestral contributions of these settlers across the Saint-Lawrence valley is uneven, whereby two thirds of the modern gene pool is attributed to the first 2600 founders \citep{Gagnon2001, Laberge2005}.
@@
 
-The North-East of Quebec is characterized historically as being more isolated with higher emigration rates compared to Western Quebec.
-Saguenay-Lac-Saint-Jean and Charlevoix also display higher rates of some rare genetic diseases as well increased homozygosity compared to Western Quebec \citep{Gagnon2001}.
-Variation of disease prevalence across the Saint-Lawrence valley indicate that population structure fluctuates across geographic regions \citep{Gagnon2001}.
-Indeed, this pattern of genetic diversity is attributed to a number of factors like concentration of founders, native ancestry and geographic isolation \citep{Gagnon2001,Laberge2005,Bherer2011}.
+The North-East of Quebec is characterized historically as being more isolated with lower immigration rates compared to Western Quebec.
+Saguenay-Lac-Saint-Jean and Charlevoix also display higher rates of some rare genetic diseases as well as increased homozygosity compared to Western Quebec \citep{Gagnon2001}.
+Variation of disease prevalence across the Saint-Lawrence valley indicate that population structure varies across geographic regions \citep{Gagnon2001}.
+Indeed, this pattern of genetic diversity is attributed to a number of factors like variation in native ancestry, geographic isolation and concentration of founders.\citep{Gagnon2001,Laberge2005,Bherer2011}.
+
 The founders of Quebec are not all from the same regions in France.
-In the seventeenth-century, while France was evenly populated, there were not many marriages across great distances \citep{Laberge2005}.
-French settlers came from regions around Paris and Atlantic port towns.
-Each group of settlers carried a distinct gene pool and founded a region of 'Nouvelle-France'.
-The population structure from each of these regions in France provides initial diversity despite small population sizes \citep{Laberge2005}.
-Moreover, Native American ancestry has been estimated to be around 1\% among these settlements \citep{Moreau2013}.
+In the seventeenth-century, while France consisted of a network of small densely populated towns, there were not many marriages across great distances \citep{Laberge2005}.
+Thus, the French settlers from regions around Paris and Atlantic port towns provide initial population structure \citep{Laberge2005}.
+Moreover, Native American ancestry has been estimated to be around 1\% among French-Canadians. \citep{Moreau2013}.
 
@@
 
-Many populations experienced founder effects as humans expanded out of Africa and populated the continents.
-Studying examples of founder effects in human populations are a great way to improve our understanding of human demographic histories.
+Many populations experienced founder effects as humans expanded out of Africa and populated the other continents.
+Studying examples of founder effects in human populations is a great way to improve our understanding of human demographic histories.
 The Quebec population is an good case study for population genetics for multiple reasons.
@@
 
-While the deep-genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes \citep{Gagnon2001,Roy-Gagnon2011}.
-The aim of the second chapter of this thesis will be to examine the genetic structure of the Quebec population using more recent and larger genomic datasets in combination with the deep-genealogical and geographical data.
+While the deep genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes \citep{Gagnon2001,Roy-Gagnon2011}.
+The aim of the second chapter of this thesis will be to examine the genetic structure of the Quebec population using more recent and larger genomic datasets in combination with the deep genealogical and geographical data.
 Recently developed methodologies will also be of use to further characterize this genomic landscape.
@@
 The Genizon cohort has 9961 samples from Quebec, genotyped on a variety of chips.
-While this data has yet to be published, it has required a fair amount of cleaning and processing to account for the different genotyping chips used.
-One key advantage of using this data is that we already have consent to link 516 individuals to genealogical records.
-We will also include data from the Cartagene project which has over 12,000 sequenced Quebecers.
+While this data has yet to be published, our first contribution has been cleaning and processing these data to account for the different genotyping chips used.
+One key advantage of using this cohort is that we already linked 516 individuals to the BALSAC genealogical records.
+We will also include data from the Cartagene project which will soon have over 12,000 genotyped Quebecers.
 
-The BALSAC database is a genealogical tree of the French-Canadian Quebec population.
+The BALSAC database is a genealogical tree of predominantly French-Canadian Quebecers.
 It was compiled from birth, marriage and death certificates from the Catholic Church.
 Because of the major role the Church played in French-Canadian society, these records are remarkably detailed and comprehensive.
-As mentioned above, 516 individuals from the Genizon cohort have been genotyped and are linked to this genealogy.
 
@@
 Using Fst statistics we can infer the relative differences between sub-populations of Quebec (see \citep{Roy-Gagnon2011}).
-Dominic Nelson has written software that can produce single locus forward simulations of the Quebec population using the genealogical data from BALSAC.
+Dominic Nelson, a graduate student from the Gravel lab, has written software that can produce single locus forward simulations of the Quebec population using the genealogical data from BALSAC.
 These simulations can be used to produce estimated differences between sub-populations resulting purely from a founder effect.
-Comparing the Fst statistics between the real data and simulations will determine if the structure observed in Quebec is the result of preserved genetic structure from France or the result of drift from founder events.
+Comparing the Fst statistics between the real data and simulations will allow us to partition the variation as the result of preserved genetic structure from France or of drift from founder events.
 
-\section{Results : the geography of genetics}
+\section{Preliminary Results}
+
+\subsection{The geography of genetics}
 
 Preliminary analyses of the Genizon dataset are in agreement with previous work.
-In collaboration with Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data \citep{mcinnes2018umap,diaz2018}.
-This method takes the first 20 principle components of the data as input and reduces this high dimensional data preserving both local and global structure (see \citep{diaz2018} and \citep{mcinnes2018umap}).
-This projection shows significant population structure within the Quebec population.
+In collaboration with another graduate student from the Gravel lab, Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data \citep{mcinnes2018umap,diaz2018}.
+This method takes the first $n$ principal components of genetic data as input ($n = 20$ in this case) and reduces this high dimensional data to a lower dimension that preserves the high dimensional topology.
+UMAP has been shown to outperform dimension reduction techniques like tSNE both computationally and at preserving both local and global population structure \citep{diaz2018, mcinnes2018umap}.
+This projection shows significant population structure in Quebec (Figure \ref{UMAP}).
 
-We colored these plots based on the three dimensional coordinates of the UMAP projection space (X,Y,Z) converted to three dimensional color space (R,G,B).
+We colored this plot based on the three dimensional coordinates of the UMAP projection space (X,Y,Z) converted to three dimensional color space (R,G,B) (Figure \ref{UMAP}).
 Therefore, Individuals with similar color are closer to each other in the projection and presumably share more recent ancestry.
 To link the genetic data to geographic coordinates, we used the marriage locations of the ancestors of those linked to the BALSAC database.
-Combining the ancestral locations of these individuals with the three dimensional UMAP coloring scheme, geographic clustering of the sub-populations becomes apparent.
+The ancestral locations of these individuals in combination with the three dimensional UMAP coloring scheme show geographic clustering of the sub-populations becomes apparent.
 Five genetic clusters are visible in the UMAP projection, these clusters also appear to be geographically linked.
@@
 
-\section{Future directions}
+\section{Future Goals}
 While Fst gives a good estimate of genome wide past coalescent rates, we plan on analyzing IBD segments as they can provide us with more detailed insight of demographic history \citep{Charlesworth2003, Ralph2013, Baharian2016}.
@@
 We plan on expanding the current analyses to include data from French individuals which could be used to confirm the results from the Fst analysis as well as provide insight into which regions in France are source populations to regions in Quebec.
+If our hypothesis is correct, we expect French data projected onto the preserved embeddings from Figure \ref{UMAP} to cluster with the distinct sub-groups of Quebec.
 Dr. Christian Dina in the University of Nantes in France has expressed interest in sharing French genomics data for this project.
@@
 One preliminary objective might be to replicate the population genetics methods described above by Palamara et al. and Campbell et al. \citep{Palamara2015,Campbell2012}.
-Knowing that mutation rates can vary across populations, applying these methods to a new dataset may in itself be worthy of publication.
+There remains much controversy over mutation rate estimations and applying each of these methods to the same dataset would be a useful comparison.
 
-Considering the deep genealogical records available for the French-Canadian population, we may be able to improve on the Palamara method by acquiring more accurate estimates of generation time separating two IBD segments.
+Considering the deep genealogical records available for the French-Canadian population, we may be able to improve on the Palamara method by acquiring more precise estimates of generation time separating two IBD segments.
 Indeed, by linking individuals to the phylogenetic tree of the Quebec population, we may be able to better estimate the most recent common ancestor for a given segment of the genome.
+In addition, we expect French-Canadians to share long IBD segments as a result of the relatively recent bottleneck, which increases the power for this method.
 Furthermore, by grouping IBD segments according to genomic regions (e.g.
-coding, CpG, etc.) and running the analysis separately would provide us with a more detailed view of the genome wide mutation rate in humans.
+coding, CpG islands, etc.) and running the analysis separately would provide us with a more detailed view of the genome wide mutation rate in humans.
 

### 2018-11-26 QE/main.tex
now I'm done

@@
 
-\addbibresource{QE_bib2.bib} % The filename of the bibliography
+\addbibresource{QE_bib.bib} % The filename of the bibliography
 
@@
 \section{Brief history of Quebec}
-The French-Canadian population of Quebec was founded over four centuries ago from around 8500 settlers from France \citep{Bherer2011,Laberge2005}.
+The French-Canadian population of Quebec was founded over four centuries ago from \textasciitilde 8500 settlers from France \citep{Bherer2011,Laberge2005}.
 This small number of settlers are the direct ancestors of the majority of Quebec's population of 7 million.
@@
 Thus, the French settlers from regions around Paris and Atlantic port towns provide initial population structure \citep{Laberge2005}.
-Moreover, Native American ancestry has been estimated to be around 1\% among French-Canadians. \citep{Moreau2013}.
+Another source of genetic diversity in the French-Canadian population is an estimated 1\% Native American ancestry \citep{Moreau2013}.
 
@@
 
+\subsection{Previous studies of Quebec population structure}
 There have been several studies in the past few decades that have examined the population structure of Quebec.
@@
 \section{Materials and Methods}
-
+\subsection{Genetic and genealogical data}
 While the deep genealogical data has been studied for over a decade, linking the genealogy to genetic data has only been done with relatively small sample sizes \citep{Gagnon2001,Roy-Gagnon2011}.
@@
 
-\subsection{Founder effect or Ancient population structure?}
+\subsection{Founder effect or ancient population structure?}
 Using Fst statistics we can infer the relative differences between sub-populations of Quebec (see \citep{Roy-Gagnon2011}).
@@
 
-\section{Preliminary Results}
+\section{Preliminary results}
 
@@
 
-\section{Future Goals}
+\section{Goals}
 While Fst gives a good estimate of genome wide past coalescent rates, we plan on analyzing IBD segments as they can provide us with more detailed insight of demographic history \citep{Charlesworth2003, Ralph2013, Baharian2016}.
@@
 
+%\subsection{Phylogenetic methods}
 The first human mutation rates were estimated by Haldane (1935) using the rates of children born with rare diseases from unaffected parents \citep{Haldane1935}.
@@
 
+%\subsection{Direct methods}
 Instead of using distant ancestry and fossil records, another more direct approach to estimating mutation rates is to sequence the genomes of parents and offspring. 
@@
 
+%\subsection{Population genetics methods}
 Methods for estimating mutation rates using population genetics are increasingly being used.
@@
 
-\subsection{Applying new methods to new data}
+\section{Goals}
+%\subsection{Applying new methods to new data}
 The aim of the third chapter of this thesis will be to explore mutation rate estimations using a combination of sequence data and genealogical data from the French-Canadian population.

### 2018-12-18 legacy.tex
added comment about SNPs vs Variants
finished
Added 90Han Text, still need to update values and plots with the final run.

@@
 We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of SNPs that show evidence for technical bias.
-Finally, we investigate how these SNPs have impacted modern genomics analyses.
+Next, we expand our analysis to the other populations in the 1kGP and identify lists of variants that show evidence for technical bias.
+Finally, we investigate how these variants have impacted modern genomics analyses.
 
@@
 To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Supplementary Figure \ref{PC1_Correlation}).
-The average quality of mapped bases  $Q$ per individual stood out as a strong correlate (\sgcomment {p value?}): Individuals with low $Q$ show elevated rates of the signature .
+The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature.
+%(\sgcomment {p value?})
 Thus, sequences with low mapping quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 
-Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques\sgcomment{You have to give some details or point to the results!}.
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged \sgcomment{You have to give some details or point to the results!}.
+Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
+The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
@@
 
-\luke{doing a final update on these values. Repeat entries, rerunning final adjustment with unique positions.}
+\luke{Need a final update on these values.}
 \begin{table}[h]
@@
 
-\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$.
+\caption{Overlap of variants identified independently to be associated with average quality of mapped bases $Q$.
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 
-\subsection{Suspicious SNPs impact modern genomics analyses}
+\subsection{Suspicious variants impact modern genomics analyses}
 The state of the art imputation servers use a combination of many databases including many that are not freely available.
@@
 
-To investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
-We found that 11\% of the SNPs associated with $Q$ were imputed back in the samples. 
+To investigate the proportion of suspicious variants that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+We found that all of the variants associated with $Q$ were imputed back in the samples. 
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
 
-Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these SNPS to be associated with phenotypes.
+Once we identified variants that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
-\sgcomment{This is \emph{very} misleading, and wrong. We should discuss. }
-\luke{Agreed, that sentence slipped by, there was another sentence like that in the discussion that I had fixed. Although the 3rd paper's SNP has $p < 10^{-11}$ }
 
@@
 
-%While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below a significance threshold of $ p < 10^{-5}$.
-%The data used was imputed genotype data using Hapmap II as the reference.
-%One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to $Q$ .
-%While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified.
-%Finding some of these SNPs in HapMap data was unexpected, however, not impossible considering that HapMap and 1kGP shared many samples.
-%We found that roughly 1\% of the SNPs associated to $Q$ were found in HapMap; rs6057648 is among those found in HapMap.
+\section{Discussion}
 
-\section{Discussion}
-Two of these studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
-The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
-It is possible that that a cell line artifact may have produced the same suspicious SNPs in the two sequencing projects.
-
-However, considering that there are only 70 suspicious SNPS that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-One of the SNPs was imputed solely using the CEU population of HapMap, and this population shows no significant association to quality for that SNP (Figure \ref{DiabetesSNP}).
-Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real; however we cannot discount the chance that they are real.
-%\sgcomment{check style}
-%\sgcomment{It is clear that -- see Strunk-White} 
-The majority of these variants do not reproduce in other datasets, but a small portion of them do reproduce and should be looked at with care. 
-For example, if the SNP was imputed with HapMap maybe it is real, but if this same SNP is imputed using the 1kGP, it might be imputed improperly.
-\sgcomment{This paragraph reads as wishy washy. }
-\luke{Agreed, I'm not sure what we're trying to say here.}
-
-The second study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
+One study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
 Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
 Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
-The study further confirmed EBV copy number estimation using PCR on 7 samples, but all of them had high $Q$.
 
@@
 It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
-Given that the bias appears to have been resolved in later stages of the project using updated sequencing platforms, we did not pursue this line of investigation.
+
+Two studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
+The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
+It is possible that that a cell line artifact may have produced the same suspicious variants in the two sequencing projects.
+
+However, considering that there are only 70 suspicious variants that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
+Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real; however we cannot discount the chance that they are real.
+In 2017 Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
+We find that \todo{000} out of the \todo{000} the suspicious variants from the single population test, and \todo{000} out of \todo{000} the variants from the combined test are present in the resequenced data.
+While this is more than 1\% of the sites, we suspect that they are indeed false associations to quality.
+To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
+We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
+These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
 
@@
 Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}. 
-We find that $99\%$ of the suspicious SNPs are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
+We find that $99\%$ of the suspicious variants are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
 
-The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 70 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
+The presence of some of these variants in HapMap suggests a cell line effect; although, this result is not very convincing with only 70 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
 We also note that 186 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
@@
 Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
-(See supplementary material for lists of SNPs)
+(See supplementary material for lists of variants)
 
@@
 
-On a technical front, we were surprised that strong association between SNPs and technical covariates in the 1kGP project had not been identified before. 
+On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
 The genome wide simple logistic regression analysis approach is rather straightforward, and could be a standard in a variety of -omics studies.
@@
 More generally, to improve the quality of genomic reference datasets, we can proceed by addition of new and better data and by better curation of existing data.
-Given rapid technological progress, the focus of genomic research is naturally on the data generation side. However, cleaning up data is also important to avoid generating spurious results. 
+Given rapid technological progress, the focus of genomic research is naturally on the data generation side. 
+However, cleaning up data is also important to avoid generating spurious results. 
 The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing.
@@
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 we're considered.
-In total, 57,838,849 diallelic autosomal SNPs passed our quality controls for the mutation spectrum analysis.
+In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
 
@@
 We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the total number of genomic loci, and $P_s$ is the number of populations with a segregating variant for locus $s$. 
-In the data, $P_s$ ranges from 1 to 26 since not all SNPs were present in all populations with singletons and doubletons filtered out.
+In the data, $P_s$ ranges from 1 to 26 since not all variants were present in all populations with singletons and doubletons filtered out.
 
@@
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum of triplets for the list of significant SNPs for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}
+We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}
 We modified the methods as necessary for our purposes, scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
@@
 We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip} genotype data.
-The VCF file returned from the server was then downloaded and used to search for the number of significant SNPs successfully imputed.
+The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.
 
-\subsection{Frequencies of SNPs in other datasets}
+\subsection{Frequencies of variants in other datasets}
 Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
-We then compared the frequencies of these 70 SNPs in the 1kGP Japanese samples to those of the HapMap Japanese samples.
+We then compared the frequencies of these 70 variants in the 1kGP Japanese samples to those of the HapMap Japanese samples.
 Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.
@@
 There are 3839 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{NRS_Manhattan}
@@
 There are 1175 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{NRI_Manhattan}
@@
 There are 2207 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
 \label{RS_Manhattan}
@@
 There are 750 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{RI_Manhattan}
@@
 
-\caption{Empirical distribution plots of the sum of deviances for Non-Repeat SNPs grouped by N, the number of populations included.
+\caption{Empirical distribution plots of the sum of deviances for Non-Repeat variants grouped by N, the number of populations included.
 The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom.
@@
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
-\caption{\sgcomment{not regressions}Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.
+\caption{Plots of data metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
+The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.
 }
 \label{PC1_Correlation}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/MutationSpectrum_cutOff.png}
+\caption{\textbf{A} 
+The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality. 
+\textbf{B} 
+Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
+\label{MutSpect}
 \end{figure}
@@
 \begin{figure}[h]
+\includegraphics[width=8cm,keepaspectratio]{./Figures/ResequencePvals.jpg}
+\caption{Comparison of p-values from the CHS and CHB populations using the original 1kGP data and the 90 Han Chinese resequenced data from \citep{Lan2017}.
+Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.
+There is no correlation in the p-values between the two datasets indicating that the suspicious variants found in the more recent dataset are false discoveries.}
+\label{90Han}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
@@
 Specify proportion of SNPs not included} The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
-There are 70 such SNPs in HapMap and 186 in the GnomAD.}
+There are 70 such variants in HapMap and 186 in the GnomAD.}
 \label{HapMap_GnomAD}

### 2018-12-18 Misc/JustStats.R

@@ -1,11 +1,17 @@
 ##Just the Stats
 library(data.table)
 args = commandArgs(trailingOnly=T)
+#ARGUMENTS :
+#1 : input file in format (Name,Chr,Pos,GTvalue,Pop,average_quality_of_mapped_bases)
+#2 : output file name
+#3 : Pop Name
 
 #read input file
-#Input <- fread('/Users/luke/Documents/Regression/CHR22.Format_JPT.csv',sep=',')
+#Input <- fread(paste("gunzip -c",'~/Documents/Regression/CHR22_NOT.Format_JPT.csv'),sep=',')
 Input <- fread(paste("gunzip -c ",args[1]),sep=',', header=T)
 Pop <-args[3]
+PC <- fread(paste('~/Documents/PCAperPop/',Pop,'.eigenvec',sep=''))
+Input<-merge(Input, PC[,c(2,3,4)], by.x='Name',by.y='V2')
 Positions<-unique(Input $Pos)
 
 chunks=split(Positions, ceiling(seq_along(Positions)/10000))
@@ -14,10 +20,11 @@ for(chunk in chunks){
 print(paste(Pop,' Progress :',round(index/length(chunks)*100, digits=2),' %'))
 subChunk<-Input[which(Input $Pos %in% chunk),]	
 stat <- with(subChunk, by(subChunk,Pos, function(x) 
-		glm(value ~ average_quality_of_mapped_bases, data=x, family=binomial)))
+		glm(value ~ V3 + V4 + average_quality_of_mapped_bases, data=x, family=binomial)))
 #extract deviance
 less = unique(subChunk[,c('Chr','Pos','Pop')])
-less$dev = as.numeric(as.character(sapply(stat,function(x) anova(x, test='Chi')[2,2])))
+#the deviance of the quality is the last row
+less$dev = as.numeric(as.character(sapply(stat,function(x) anova(x, test='Chi')[4,2])))
 
 #write results to file
 if(index == 1){
@@ -29,3 +36,5 @@ index <- index + 1
 
 ##code dump
[…]

### 2018-12-18 Misc/PCvsQual.R

@@ -1,0 +1,59 @@
+library(data.table)
+library(ggplot2)
+library(cowplot)
+
+Pops <- fread('~/genomes/genomes/PopNames/NamePop.txt', header=F, nrows=26)
+
+Qual <- fread('~/genomes/genomes/misc/1000GenomesQual_INT.pheno')
+
+# GET EQUATION AND R-SQUARED AS STRING
+# SOURCE: http://goo.gl/K4yh
+
+lm_eqn <- function(df){
+	names(df)=c('x','y')
+    m <- lm(y ~ x, df);
+    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2*","~~italic(p)~"="~pv,
+         list(a = format(coef(m)[1], digits = 2), 
+              b = format(coef(m)[2], digits = 2), 
+             r2 = format(summary(m)$r.squared, digits = 3),
+             pv = format(summary(m)$coefficients[2,4], digits = 3)))
+    as.character(as.expression(eq));                 
+}
+
+p1 <- p + geom_text(x = 25, y = 300, label = lm_eqn(df), parse = TRUE)
+
+
+for(i in seq(1,26)){
+	pop = as.character(Pops$V1[i])
+	PC = fread(paste('~/Documents/PCAperPop/',pop,'.eigenvec',sep=''))
+
+	plt = merge(Qual, PC, by=c('V1','V2'))
+	p1 = ggplot(plt, aes(x=V3.y,y=V3.x))+geom_point()+
+	geom_smooth(method='lm',formula=y~x, se=F)+
+	theme_classic()+theme(plot.title = element_text(hjust = 0.5))+
+	labs(y='Normalized Quality',x='PC1',title=pop)+
+	annotate("text", x = -0.25, y = 2.5, 
+		label = lm_eqn(plt[,c('V3.y','V3.x')]), size = 2, parse=TRUE)
+	
+	p2 = ggplot(plt, aes(x=V4,y=V3.x))+geom_point()+
+	geom_smooth(method='lm',formula=y~x, se=F)+
+	theme_classic()+
+	labs(y='Normalized Quality',x='PC2')+
+	annotate("text", x = -0.25, y = 2.5, 
+		label = lm_eqn(plt[,c('V4','V3.x')]), size = 2, parse=TRUE)
+		
+	p3 = plot_grid(p1,p2,ncol = 1)
+	
[…]

### 2018-12-18 Misc/Parallel_Regression2.py

@@ -5,12 +5,12 @@ def Bash_cmd(i):
     #print statement to keep track of where you are
     print('Working on Chrom : '+str(i))
     #put your bash command below
-    subprocess.call('for pop in CDX CLM IBS PEL PJL KHV ACB GWD ESN BEB MSL STU ITU CEU YRI CHB JPT LWK ASW; \
+    subprocess.call('for pop in GBR FIN CHS PUR CDX CLM IBS PEL PJL KHV ACB GWD ESN BEB MSL STU ITU CEU YRI CHB JPT LWK ASW; \
     do echo ${{pop}} {0}; \
-    (ls /Users/luke/Documents/Regression/CHR{0}_NOT.Regression_${{pop}}.csv && echo "File Exists") \
+    (ls /Users/luke/Documents/FinalRegression/CHR{0}_NOT.Regression_${{pop}}.csv && echo "File Exists") \
     || Rscript /Users/luke/Documents/QualityPaper/Misc/JustStats.R \
     /Users/luke/Documents/Regression/CHR{0}_NOT.Format_${{pop}}.csv \
-    /Users/luke/Documents/Regression/CHR{0}_NOT.Regression_${{pop}}.csv \
+    /Users/luke/Documents/FinalRegression/CHR{0}_NOT.Regression_${{pop}}.csv \
     ${{pop}};done'.format(i), shell=True)
 
 if __name__ == '__main__':

### 2018-12-18 PlottingScripts/Resequenced_Dev.R

@@ -1,0 +1,49 @@
+library(data.table)
+library(ggplot2)
+library(cowplot)
+
+CHB.kgp<-fread('/Users/luke/Documents/Regression/GenomeWideRegression_CHB.csv')
+CHS.kgp<-fread('/Users/luke/Documents/Regression/GenomeWideRegression_CHS.csv')
+
+CHB.kgp$Pos <- as.numeric(as.character(CHB.kgp$Pos))
+CHS.kgp$Pos <- as.numeric(as.character(CHS.kgp $Pos))
+CHB.kgp$Chr <- as.numeric(as.character(CHB.kgp$Chr))
+CHS.kgp$Chr <- as.numeric(as.character(CHS.kgp $Chr))
+CHB.kgp$dev <- as.numeric(as.character(CHB.kgp$dev))
+CHS.kgp$dev <- as.numeric(as.character(CHS.kgp $dev))
+
+CHB.kgp<-CHB.kgp[complete.cases(CHB.kgp),]
+CHS.kgp<-CHS.kgp[complete.cases(CHS.kgp),]
+
+
+CHB.han<-fread('~/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_CHB_Regression.txt')
+CHS.han<-fread('~/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_CHS_Regression.txt')
+
+CHB.han$Chr <- as.numeric(as.character(gsub('chr','',CHB.han$Chr)))
+CHS.han$Chr <- as.numeric(as.character(gsub('chr','',CHS.han$Chr)))
+CHB.han$Pos <- as.numeric(as.character(CHB.han$Pos))
+CHS.han$Pos <- as.numeric(as.character(CHS.han$Pos))
+CHB.han$dev <- as.numeric(as.character(CHB.han$dev))
+CHS.han$dev <- as.numeric(as.character(CHS.han$dev))
+
+CHB.han<-CHB.han[complete.cases(CHB.han),]
+CHS.han<-CHS.han[complete.cases(CHS.han),]
+
+CHB<-merge(CHB.han, CHB.kgp, by=c('Chr', 'Pos', 'Pop'), all =T)
+CHS<-merge(CHS.han, CHS.kgp, by=c('Chr', 'Pos', 'Pop'), all= T)
+CHB[is.na(CHB)] <- 0
+CHS[is.na(CHS)] <- 0
+
[…]

### 2018-12-19 legacy.tex
suggestions

@@
 Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}). 
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
+To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable})
@@
 
-\caption{Overlap of variants identified independently to be associated with average quality of mapped bases $Q$.
+\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
 The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
+\sgcomment{The previous paragraph should be clarified: two of the studies are likely ok.
+the third is likely problematic, and why.
+For "conventional" GWAS, we do not see evidence that these contaminated hits.
+but they may still affect, e.g., risk scores.}
 
@@
 With our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
+\sgcomment{This could be fleshed out a bit.
+Currently reader has no idea whether the proportion of variants that are present in other cohorts is more or less than expected.
+Also, I would probably put this paragraph before the GWAS section.}
 
@@
 It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
+\sgcomment{Do you still stand by this?}
 
@@
 However, considering that there are only 70 suspicious variants that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real; however we cannot discount the chance that they are real.
+Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real \sgcomment{previous phrase needs rewriting.
+I don't love the "real" vs "not real".
+Or at least you should clarify that we detect problematic mutations.
+THe next question is whether these are real and badly called, or completely spurious I think this paragraph suggest that the majority are spurious}; however we cannot discount the chance that they are real.
 In 2017 Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
@@
 These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
+\sgcomment{I would have the discussion of cell line artifacts as a separate paragraph}
 
@@
 We were not able to pinpoint the precise technical source of the discrepancy. 
-Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}. 
-We find that $99\%$ of the suspicious variants are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.
+Sequencer-specific errors would be expected to be absent from Hapmap \sgcomment{why?
+Do you mean that they would be different between 1kgp and hapmap?}, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}.
+We find that $99\%$ of the suspicious variants are indeed absent from HapMap \sgcomment{what is this $99\%$?
+concordance?
+I thought that the result was a bit different}, which is consistent with the expected false positive rate in our test.
 
 The presence of some of these variants in HapMap suggests a cell line effect; although, this result is not very convincing with only 70 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
+\sgcomment{I don't think you want to start sentence with "although,".
+}
 We also note that 186 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
-Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.
+Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal. \sgcomment{This paragraph needs a bit of work}
 
-If these sequencing platforms are no longer being used, then it it likely that these artifacts are no longer being actively introduced in recent sequence data.
+If these \sgcomment{Which?} sequencing platforms are no longer being used, then it it likely that these \sgcomment{which?} artifacts are no longer being actively introduced in recent sequence data.
+\sgcomment{It's good to start paragraph in more self-contained manner}
 However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
-We have shown that these variants have been significantly associated using traditional GWAS methods.
-However, this is not the only method that would be affected by these false positives. 
+\sgcomment{Even though we have found only one GWAS study that reported associations with the problematic variants, the imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores and biases in fine mapping analyses.
+}
+%However, this is not the only method that would be affected by these false positives.
 Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
-The inclusion of these variants could also affect population genetics analyses that compare and contrast populations based on allele frequencies.
+These variants do have a sizeable effect on population on analyses that compare and contrast populations based on mutational profiles.
+Even though we expect that present-day technologies will be less sensitive to such artefacts, if is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
 
 \subsection{Recommendations}
-While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet the current quality standards.
+While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
 A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-In this case, we used a cut-off of $Q$ over 30.
+We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signal.
 Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
-(See supplementary material for lists of variants)
+(See supplementary material for lists of variants) \sgcomment{This is not really a recommendation.
+We should be a bit more specific}
 
@@
 On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
-The genome wide simple logistic regression analysis approach is rather straightforward, and could be a standard in a variety of -omics studies.
+The genome wide simple logistic regression analysis approach is straightforward, and should probably be a standard in a variety of -omics studies.
 
@@
 
-We have not identified the precise mechanism through which the technical artifact was introduced: we have ruled out neither cell-line, sequencer, nor algorithmic artifact. 
-Given that modern platforms seem to have resolved this technical issue but that recent studies continue to be affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while the sequencing forensics investigation can take place.
+We have not identified the precise mechanism through which the technical artifact was introduced.
+Given that modern platforms seem to have resolved this technical issue, but that recent studies continue to be affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while a sequencing forensics investigation can take place.
 

### 2018-12-19 legacy.tex
made some changes, more needed

@@
 \subsection{Suspicious variants impact modern genomics analyses}
+\subsubsection{Imputation}
 The state of the art imputation servers use a combination of many databases including many that are not freely available.
@@
 We found that all of the variants associated with $Q$ were imputed back in the samples. 
-This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual
+This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
 
-Once we identified variants that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
+\subsubsection{Cell line or technical artifact}
+
+To determine whether this artifact is the result of a cell line artifact or a technical artifact, we investigated other cohorts that used the same cell lines.
+The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
+If these variants are a cell line artifact then the same suspicious variants should be present in the two sequencing projects.
+We found that only 70 variants were present in HapMap and 186 in GnomAD (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
+
+% \sgcomment{This could be fleshed out a bit. Currently reader has no idea whether the proportion of variants that are present in other cohorts is more or less than expected.}
+
+%However, considering that there are only 70 suspicious variants that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
+%Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real \sgcomment{previous phrase needs rewriting.
+I don't love the "real" vs "not real".
+Or at least you should clarify that we detect problematic mutations.
+The next question is whether these are real and badly called, or completely spurious I think this paragraph suggest that the majority are spurious}; however we cannot discount the chance that they are real.
+In 2017, Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
+We find that \todo{000} out of the \todo{000} the suspicious variants from the single population test, and \todo{000} out of \todo{000} the variants from the combined test are present in the resequenced data.
+While this is more than 1\% of the sites, we suspect that they are indeed false associations to $Q$.
+To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
+We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
+These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
+\luke{More here with the SFS and concordance between 90Han and 1kGP}
+
+\subsubsection{GWAS}
+We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
-Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation.
-The third study used the 1kGP sequence data and cell lines directly.
+Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
+A third study used the 1kGP sequence data and cell lines directly.
 All three papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
-\sgcomment{The previous paragraph should be clarified: two of the studies are likely ok.
-the third is likely problematic, and why.
-For "conventional" GWAS, we do not see evidence that these contaminated hits.
-but they may still affect, e.g., risk scores.}
+The two studies that imputed using HapMap are likely not impacted by this as these variants are likely falsely associated to $Q$.
 
-We searched other databases to see if the suspicious variants had been sequenced in different cohorts.
-70 of these variants were present in HapMap and 186 in GnomAD (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
-These variants include the two variants found in the studies that imputed genotypes using HapMap.
-With our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
-\sgcomment{This could be fleshed out a bit.
-Currently reader has no idea whether the proportion of variants that are present in other cohorts is more or less than expected.
-Also, I would probably put this paragraph before the GWAS section.}
-
-\section{Discussion}
-
-One study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
+The third study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
 Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
@@
 
-Two studies used HapMap as a reference panel, rather than 1000 Genomes \citep{Kraja2011, Ebejer2013}. 
-The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
-It is possible that that a cell line artifact may have produced the same suspicious variants in the two sequencing projects.
+For "conventional" GWAS, we do not see evidence that these contaminated hits. but they may still affect, e.g., risk scores.
 
-However, considering that there are only 70 suspicious variants that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real \sgcomment{previous phrase needs rewriting.
-I don't love the "real" vs "not real".
-Or at least you should clarify that we detect problematic mutations.
-THe next question is whether these are real and badly called, or completely spurious I think this paragraph suggest that the majority are spurious}; however we cannot discount the chance that they are real.
-In 2017 Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-We find that \todo{000} out of the \todo{000} the suspicious variants from the single population test, and \todo{000} out of \todo{000} the variants from the combined test are present in the resequenced data.
-While this is more than 1\% of the sites, we suspect that they are indeed false associations to quality.
-To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
-We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
-These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
-\sgcomment{I would have the discussion of cell line artifacts as a separate paragraph}
+\section{Discussion}
 

### 2018-12-20 legacy.tex
minor edits

@@
 \subsubsection{Imputation}
-The state of the art imputation servers use a combination of many databases including many that are not freely available.
+The state of the art imputation servers use a combination of many databases including some that are not freely available.
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
 
-To investigate the proportion of suspicious variants that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
+To investigate the proportion of suspicious variants that are being imputed, we submitted the first two chromosomes of  the 1kGP genotype data to the Michigan Imputation Server.
 We found that all of the variants associated with $Q$ were imputed back in the samples.
@@
 If these variants are a cell line artifact then the same suspicious variants should be present in the two sequencing projects.
-We found that only 70 variants were present in HapMap and 186 in GnomAD (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+We found that only 70 variants were present in HapMap and 186 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
 However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.

### 2018-12-20 legacy.tex
new figures

@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NonRepeatSNPs.jpg}
 
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NonRepeatIndels.jpg}
 
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_RepeatSNPs.jpg}
 
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_Indels.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_RepeatIndels.jpg}
 

### 2018-12-20 Misc/ProcessRegression.R

@@ -1,0 +1,33 @@
+##Manhattan
+library(dplyr)
+library(magrittr)
+library(data.table)
+library(ggplot2)
+library(multtest)
+library(cowplot)
+
+dir='~/Documents/GenomeWideRegression/'
+out='~/Documents/GenomeWideRegression/'
+
+Reg = fread('~/Documents/GenomeWideRegression/ALLPOP_GenomeWide.Regression.csv', header=F, sep=' ', col.names=c('Chr','Pos','Pop','dev'))
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$dev))
+
+
+for(i in 1:22){
+print(i)	
+Sub<- Reg[which(Reg$Chr == i),]
+#chisquared random variable
+#taking the sum of the deviances (mean is to take into account the DF)
+Sub <- Sub %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
+#write results to file
+if(i == 1){
+	write.table(Sub,file='~/Documents/GenomeWideRegression/GenomeWide.MeanDev_ALLPOPS.csv', quote=F, row.names=F, append=FALSE)}
+if(i > 1){
+    write.table(Sub,file='~/Documents/GenomeWideRegression/GenomeWide.MeanDev_ALLPOPS.csv', quote=F, row.names=F, append=TRUE)}
+}
+
+
+	
+

### 2018-12-20 PlottingScripts/TotalRegression.R

@@ -7,8 +7,30 @@ library(multtest)
 library(cowplot)
 library(reshape2)
 
+dir='~/Documents/GenomeWideRegression/'
+out='~/Documents/GenomeWideRegression/'
+
+Reg = fread('~/Documents/GenomeWideRegression/ALLPOP_GenomeWide.Regression.csv', header=F, sep=' ', col.names=c('Chr','Pos','Pop','dev'))
+Reg$Chr<-as.numeric(as.character(Reg$Chr))
+Reg$Pos<-as.numeric(as.character(Reg$Pos))
+Reg$dev<-as.numeric(as.character(Reg$dev))
+
+
+for(i in 1:22){
+print(i)	
+Sub<- Reg[which(Reg$Chr == i),]
+#chisquared random variable
+#taking the sum of the deviances (mean is to take into account the DF)
+Sub <- Sub %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
+#write results to file
+if(i == 1){
+	write.table(Sub,file='~/Documents/GenomeWideRegression/GenomeWide.MeanDev_ALLPOPS.csv', quote=F, row.names=F, append=FALSE)}
+if(i > 1){
+    write.table(Sub,file='~/Documents/GenomeWideRegression/GenomeWide.MeanDev_ALLPOPS.csv', quote=F, row.names=F, append=TRUE)}
+}
+
 ##read Regression
-Regression<-fread('~/Documents/QualityPaper/Misc/Reg_ALL_SITES_5col.txt', fill=T)
+Regression<-fread('~/Documents/GenomeWideRegression/GenomeWide.MeanDev_ALLPOPS.csv', fill=T)
 Regression <-Regression[which(Regression $Pos != 'Pos'),]
 Regression $Chr<-as.numeric(as.character(Regression $Chr))
 Regression $Pos<-as.numeric(as.character(Regression $Pos))
@@ -52,7 +74,7 @@ write.table(Regression, file='/Users/luke/Documents/QualityPaper/Misc/TotalRegre
 
 Regression<-fread('/Users/luke/Documents/QualityPaper/Misc/TotalRegression.csv')
 
-Repeat_Indels <- unique(Regression[which(Regression$Indel=='yes' 
+`Repeat_Indels <- unique(Regression[which(Regression$Indel=='yes' 
[…]

### 2019-02-12 legacy.tex
updated table and figures

--- 2018-12-20 legacy.tex
@@
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and indirect use through prominent imputation servers.
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses \luke{and spurious GWAS associations as a result of imputation using the 1kGP as a reference.}
 
@@
 The results section is organized as follows.
-We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1000 genomes project. 
+We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1kGP. 
 Next, we expand our analysis to the other populations in the 1kGP and identify lists of variants that show evidence for technical bias.
@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests \sgcomment{did we have a specific result for this?
-As in "$94\%$ of sites passed HW"...}.
+These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests. 
+%\sgcomment{did we have a specific result for this? As in "$94\%$ of sites passed HW"...}. 
 %Even though some of the mismatched variants are at low frequency in the 1kGP, and thus can evade Hardy-Weinberg detection, but many are at high enough frequency that HW detection   
@@
 For other analyses, removing suspicious SNPs may be adequate. 
-Lists of suspicious SNPs and low-quality individuals is provided as supplementary material.
+Lists of suspicious SNPs and low-quality individuals is provided as supplementary material \luke{link to file here}.
 
@@
 
-We therefore performed the reverse GWAS approach in all populations, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-% \sgcomment{quantify!}
+We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
 Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
-To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+
+\luke{Added text here}To build a test statistic with more power to represent the association across all populations simultaneously, we performed a logistic regression predicting genotype with the top five principal components as covariates to account for population structure and $Q$ as the last covariate.
+This method identifies a total of 1,409,401 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+%we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable})
@@
 
-\luke{Need a final update on these values.}
 \begin{table}[h]
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 2207 & 6919 \\  
-{INDEL} & 750  & 2017 \\ \hline
+{SNP}  & 284,551 & 1,059,354 \\  
+{INDEL} & 15,680 & 49,816 \\ \hline
 \end{tabular}
@@
 In 2017, Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-We find that \todo{000} out of the \todo{000} the suspicious variants from the single population test, and \todo{000} out of \todo{000} the variants from the combined test are present in the resequenced data.
+We find that 9 out of the 297 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
 While this is more than 1\% of the sites, we suspect that they are indeed false associations to $Q$.
@@
 These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
-\luke{More here with the SFS and concordance between 90Han and 1kGP}
 
@@
 We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified three recent publications that had reported these variants as close to or above the genome-wide significant threshold.
-Two of these studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
-A third study used the 1kGP sequence data and cell lines directly.
-All three papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+The NHGRI-EBI Catalog of published genome-wide association studies identified eight recent publications that had reported these variants as close to or above the genome-wide significant threshold.
+Four of these studies used the 1kGP in their reference panel for imputation.
+One study used genotype data directly using the Omni2-5 and Omni5 chips.
+Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
+And one study used the 1kGP sequence data and cell lines directly.
+All eight papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
-Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also nearly reached genome wide significance for biological traits.
+Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for some biological traits.
 The two studies that imputed using HapMap are likely not impacted by this as these variants are likely falsely associated to $Q$.
 
-The third study used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
+The study that used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
 Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
@@
 This is a commonly used transformation that helps preserve cell lines for long periods of time.
-It is unclear what exactly might be causing the association to EBV copy number and the data quality, however, one could speculate that different protocols were used in the cell culturing of certain samples from the 1kGP which might affect the replication of the virus in the cells, or somehow affect the overall quality of the sequencing data produced.
-\sgcomment{Do you still stand by this?}
 
 For "conventional" GWAS, we do not see evidence that these contaminated hits. but they may still affect, e.g., risk scores.
+
+\begin{table}[h]
+\begin{tabular}{l l l r r r}
+  {Pubmed ID}  & {Journal} & {rsID} & {GWAS -$\log_{10} p$} & {Adjusted $Q$ -$\log_{10} p$} & {odds ratio}     \\ \hline
+ 28654678	& PLoS One	& rs201761909	& 5.70	& 100.67 & 0.63\\
+ 	& 		& rs201130852	& 5.05	& 94.28 & 0.65\\
+ 	& 		& rs200655768	& 6.53	& 85.84 & 0.66\\
+ 	& 		& rs201255786	& 5.70	& 80.65 & 0.65\\	
+ 	& 		& rs184202621	& 5.52	& 71.13 & 0.68\\
+	& 		& rs80274284	& 6	& 69.02 &  0.70\\
+ 	& 		& rs200699422	& 5.30	& 7.82 & 0.87\\ 
+23527680 &	Twin Res Hum Genet	& rs6057648	& 5.40	& 25.50 & 0.82\\  
+21386085 &	Diabetes &	rs301 &*10.52	 & 8.22 & 0.90\\
+28928442	& Nature Commun	& rs201471471	& 6.52	& 7.14 & 0.67\\
+29255261	& Nature Genetics	& rs11039216	& *14.30	& 2.59 & 1.08\\
+25918132	& Toxicol Sci	& rs76780579	& 6	& 2.58	7 & 1.17 \\
+27863252	& Cell	& rs3794738	& *13.16	& 2.39 & 0.91\\
+28698626	& Science Reports	& rs11015915	& 5.05	& 2.05 & 1.08\\
+
+ \hline
+\end{tabular}
+\caption{A list of eight recent publications that reported suspicious variants as close to or above the genome-wide significant threshold. The variants reaching genome wide significance have a star.}
+\label{gwasTable}
+\end{table}
 
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Non_Repeat_SNPs.jpg}
 
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 3839 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 1,059,354 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Non_Repeat_Indels.jpg}
 
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 1175 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 49,816 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
@@
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 2207 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 284,551 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
@@
 \caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 750 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
+There are 15,680 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
@@
 \begin{figure}[h]
+\includegraphics[width=8cm,keepaspectratio]{./Figures/Han83.jpg}
+\caption{Site frequency spectrum plot of the significant SNPs identified in the single population tests of CHS and CHB populations.
+Only 9 of the 297 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
+Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
+\label{90HanSFS}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[width=8cm,keepaspectratio]{./Figures/ResequencePvals.jpg}

### 2019-02-12 Misc/Pub_Sig (Autosaved).R

@@ -1,0 +1,52 @@
+library(dplyr)
+library(data.table)
+library(ggplot2)
+library(glm2)
+
+path = '/Users/luke/genomes/genomes/hg19/Genotypes/'
+
+GT = fread('~/Desktop/PublishedHits.txt')
+ColNames = read.table(paste(path,"ColNames.txt", sep=''), header=F)
+names(GT)= as.character(unlist(ColNames))
+samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
+
+logP = apply(GT[,-(1:3)], 1, function(x) 
+				anova(
+					glm2(x ~
+							samples$PC1 + 
+							samples$PC2 +
+							samples$PC3 +
+							samples$PC4 +
+							samples$PC5 + 
+							samples$average_quality_of_mapped_bases,
+						family=binomial),
+					test="Chi")[7,2])
+
+model = apply(GT[,-(1:3)], 1, function(x)
+					glm2(x ~
+							samples$PC1 + 
+							samples$PC2 +
+							samples$PC3 +
+							samples$PC4 +
+							samples$PC5 + 
+							samples$average_quality_of_mapped_bases,
+						family=binomial))
+						
+exp.coef = apply(GT[,-(1:3)], 1, function(x)
+					exp(coefficients(glm2(x ~
+							samples$PC1 + 
+							samples$PC2 +
+							samples$PC3 +
+							samples$PC4 +
+							samples$PC5 + 
+							samples$average_quality_of_mapped_bases,
+						family=binomial)))[[7]])						
+
+dat = cbind(GT[,c(1,2)], exp.coef)
+rsid = fread('~/Desktop/SigPos.txt')
+rsID <- merge(dat, rsid, by.x=c("CHROM",'POS'), by.y=c('rsID','CHROM'))
+
+mGT = melt(GT,id=c('CHROM','POS','ID'))
+mGT <- merge(samples, mGT, by.y='variable',by.x='Name')
+cGT <- mGT %>% group_by(Pop) %>% summarize(var = sum(value))
+keep <- cGT[which(cGT$var > 0),]$Pop

### 2019-02-12 PlottingScripts/MakeQQ.R

@@ -1,0 +1,59 @@
+#Make QQ
+library(data.table)
+library(ggplot2)
+
+Repeat_Indels<-fread('/Users/luke/Documents/QualityPaper/Misc/Repeat_Indels.csv')
+NORepeat_Indels <-fread('/Users/luke/Documents/QualityPaper/Misc/NORepeat_Indels.csv')
+Repeat_SNPs <-fread('/Users/luke/Documents/QualityPaper/Misc/Repeat_SNPs.csv')
+NORepeat_SNPs <-fread('/Users/luke/Documents/QualityPaper/Misc/NORepeat_SNPs.csv')
+
+makeQQ <- function(pvalues, name){
+set.seed(1234)
+pvalue <- runif(length(pvalues), min=0, max=1)
+inflation <- function(ps) {
+  chisq <- qchisq(1 - pvalue, 1)
+  lambda <- median(chisq) / qchisq(0.5, 1)
+  lambda
+}
+inflation(pvalue)
+
+gg_qqplot(pvalues, name) +
+  theme_bw(base_size = 24) +
+  annotate(
+    geom = "text",
+    x = -Inf,
+    y = Inf,
+    hjust = -0.15,
+    vjust = 1 + 0.15 * 3,
+    label = sprintf("λ = %.2f", inflation(ps)),
+    size = 8
+  ) + ggtitle(name) +
+  theme(
+    axis.ticks = element_line(size = 0.5),
+    panel.grid = element_blank(),
+    plot.title = element_text(hjust = 0.5)
+  )
+ }
+ ##modified from : https://slowkow.com/notes/ggplot2-qqplot/
+ gg_qqplot <- function(ps,name) {
+  n  <- length(ps)
+  df <- data.frame(
+    observed = -log10(sort(ps)),
+    expected = -log10(ppoints(n)))
+  log10Pe <- expression(paste("Expected -log"[10], plain(P)))
+  log10Po <- expression(paste("Observed -log"[10], plain(P)))
+  ggplot(df) +
+    geom_point(aes(expected, observed), shape = 1, size = 3) +
+    geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
+    xlab(log10Pe) +
+    ylab(log10Po)
+}
+
+makeQQ(Repeat_Indels$p, 'Repeat Indels')
[…]

### 2019-02-12 legacy.tex
microArray + minor edits

@@
 This method identifies a total of 1,409,401 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+The total number of sites tested is 37,597,679 meaning that 3.75\% of the sites tested are statistically significantly associated to $Q$.
 %we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756
@@
 Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately.
-SNPs and INDELs were also analyzed separately.}
+SNPs and INDELs were also analyzed separately.
+The total number of sites tested is 37,597,679 meaning that 3.75\% of the sites tested are statistically significantly associated to $Q$.}
 \label{sigTable}
@@
  	& 		& rs184202621	& 5.52	& 71.13 & 0.68\\
-	& 		& rs80274284	& 6	& 69.02 &  0.70\\
+	& 		& rs80274284	& 6.00	& 69.02 &  0.70\\
  	& 		& rs200699422	& 5.30	& 7.82 & 0.87\\ 
@@
 29255261	& Nature Genetics	& rs11039216	& *14.30	& 2.59 & 1.08\\
-25918132	& Toxicol Sci	& rs76780579	& 6	& 2.58	7 & 1.17 \\
+25918132	& Toxicol Sci	& rs76780579	& 6.00	& 2.58	& 1.17 \\
 27863252	& Cell	& rs3794738	& *13.16	& 2.39 & 0.91\\
@@
 \end{tabular}
-\caption{A list of eight recent publications that reported suspicious variants as close to or above the genome-wide significant threshold. The variants reaching genome wide significance have a star.}
+\caption{A list of eight recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
+The variants reaching genome wide significance have a star.
+Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of -$\log_{10} p > 2$ are deemed significant.
+}
 \label{gwasTable}
 \end{table}
+
+\subsubsection{Microarrays}
+We found that 6456 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
+This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
+Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
 
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
-\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
-\label{DiabetesSNP}
-\end{figure}
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
+%\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
+%\label{DiabetesSNP}
+%\end{figure}
 

### 2019-02-13 legacy.tex
Added new plot, updated scripts

@@
 
-	%\subsection{Identifying Suspicious SNPs in the 1kGP}
+\subsection{Identifying suspicious variants in the 1kGP}
 The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
@@
 
-We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
+\luke{removed paragraph about meta-analysis}
+%We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+%Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
 
-\luke{Added text here}To build a test statistic with more power to represent the association across all populations simultaneously, we performed a logistic regression predicting genotype with the top five principal components as covariates to account for population structure and $Q$ as the last covariate.
-This method identifies a total of 1,409,401 statistically significant variants using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+\luke{Added text here}To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype with the top five principal components as covariates to account for population structure and $Q$ as the last covariate.
+This method identifies a total of 1,409,401 statistically significant variants associated to $Q$ using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 The total number of sites tested is 37,597,679 meaning that 3.75\% of the sites tested are statistically significantly associated to $Q$.
@@
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
+\luke{removed figure 3}
+%\begin{figure}
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
 
-\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
-The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
-Phase 1 populations are marked by a star ( * ).
-The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
-The same variants identified as being low quality independently in each population are found in other populations. }
-  \label{OverLap}
-\end{figure}
+%\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
+%The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
+%The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+%Phase 1 populations are marked by a star ( * ).
+%The line plot underneath shows the number of populations for which a variant has reached significance.
+%The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
+%The same variants identified as being low quality independently in each population are found in other populations. }
+%  \label{OverLap}
+%\end{figure}
 
@@
 The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
-If these variants are a cell line artifact then the same suspicious variants should be present in the two sequencing projects.
-We found that only 70 variants were present in HapMap and 186 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+Technological artifacts in one technology aren't expected to be present in another sequencing platform.
+Moreover, cell line artifacts would likely be present in data from two separate platforms.
+
+Using the single population association test we produced a list of variants associated to $Q$ for the samples shared between the two sequencing projects.
+From this list, we found that only 70 variants were present in HapMap and 186 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
 However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
@@
 
+\subsubsection{Microarrays}
+We found that 6,456 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
+This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
+Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
+
 \subsubsection{GWAS}
 We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified eight recent publications that had reported these variants as close to or above the genome-wide significant threshold.
-Four of these studies used the 1kGP in their reference panel for imputation.
-One study used genotype data directly using the Omni2-5 and Omni5 chips.
+The NHGRI-EBI Catalog of published genome-wide association studies identified eight recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
+
+Four of these studies used the 1kGP in their reference panel for imputation \citep{astle2016allelic, lopez2017genome, tian2017genome, luciano2018association}.
+One study used genotype data directly using the Omni2-5 and Omni5 chips \citep{yucesoy2015genome}.
 Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
-And one study used the 1kGP sequence data and cell lines directly.
+And one study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
+
 All eight papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
@@
 
-The study that used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
-Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
-Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
+\luke{removed a paragraph here}
+%The study that used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
+%Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
+%Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
 
-The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphoblastoid cell lines.
-This is a commonly used transformation that helps preserve cell lines for long periods of time.
-
-For "conventional" GWAS, we do not see evidence that these contaminated hits. but they may still affect, e.g., risk scores.
+%The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphoblastoid cell lines.
+%This is a commonly used transformation that helps preserve cell lines for long periods of time.
 
@@
 \caption{A list of eight recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
-The variants reaching genome wide significance have a star.
+The variants reaching genome wide significance have a star ( * ).
 Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of -$\log_{10} p > 2$ are deemed significant.
@@
 \end{table}
-
-\subsubsection{Microarrays}
-We found that 6456 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
-This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
-Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
 
@@
 Even though we expect that present-day technologies will be less sensitive to such artefacts, if is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
+
+For \textit{conventional} GWAS, we do not see evidence that these contaminated hits impact the conclusions made in each of these publications.
+However, considering many polygenic risk score estimates use the association scores from multiple sources, the inclusion of these variants could possibly impact downstream analyses.
 
@@
 \subsection{Genome wide association study}
-We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}.
+We ran a simple logistic regression independently for each genomic locus $s$ using the R glm2 package\citep{RDevelopmentCoreTeam2016}.
 The model we fitted to each subset of data can be expressed as such.
 \begin{align*}
-{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
+{logit}{E(Y_{s'i})} &= \beta_{0}{PC1}_{i} + \beta_{1}{PC2}_{i} + \beta_{2}{PC3}_{i} + \beta_{3}{PC4}_{i} + \beta_{4}{PC5}_{i} + \beta_{5}Q_{i}
 \\
-\textit{for}\quad i &= 1,\hdots, n_{p'}
+\textit{for}\quad i &= 1,\hdots, n
 \end{align*}
-Here $Y_{spi}$ is a Bernoulli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
-In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$.
-We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the total number of genomic loci, and $P_s$ is the number of populations with a segregating variant for locus $s$. 
-In the data, $P_s$ ranges from 1 to 26 since not all variants were present in all populations with singletons and doubletons filtered out.
+Here $Y_{si}$ is a Bernoulli random variable denoting the genotype of individual $i$ at genomic locus $s$. 
+In the model, the regression coefficient $\beta_{0}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{1}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
+We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
 
-For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{1}^{(sp)}=0$. That is, the average quality of mapped bases $Q_{p}$ variable is not associated with genotype $Y_{sp}$.
+For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{1}^{(s)}=0$. That is, the average quality of mapped bases $Q_{p}$ variable is not associated with genotype $Y_{s}$.
 In other words, the slope coefficient of the simple logistic regression is zero.
-To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q_{p}$ in the model. 
+To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q$ in the model. 
 The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
 
-To obtain a population-wide measure of association for each locus across its respective $P_s$ populations, we summed the $P_s$ deviances for each $s$.
-The resulting test statistic $T_s$ is chi-squared distributed with $P_s$ degrees of freedom (i.e.
-$T_s \sim \chi^2_{P_s}$), see Figure \ref{Deviances}.
-The observed summed deviances were used to compute the \textit{p}-value for each locus.
-
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 10756 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 1,409,401 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 

### 2019-02-13 Misc/Pub_Sig.R

@@ -3,12 +3,19 @@ library(data.table)
 library(ggplot2)
 library(glm2)
 
+#read colors per pop (used in 1kG Nature)
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
 path = '/Users/luke/genomes/genomes/hg19/Genotypes/'
 
 GT = fread('~/Desktop/PublishedHits.txt')
 ColNames = read.table(paste(path,"ColNames.txt", sep=''), header=F)
 names(GT)= as.character(unlist(ColNames))
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
+SigPub = fread('~/Documents/QualityPaper/sig/SignificantPublications.txt')
+
 
 logP = apply(GT[,-(1:3)], 1, function(x) 
 				anova(
@@ -23,15 +30,15 @@ logP = apply(GT[,-(1:3)], 1, function(x)
 					test="Chi")[7,2])
 
 model = apply(GT[,-(1:3)], 1, function(x)
-					glm2(x ~
+					predict(glm2(x ~
 							samples$PC1 + 
 							samples$PC2 +
 							samples$PC3 +
 							samples$PC4 +
 							samples$PC5 + 
 							samples$average_quality_of_mapped_bases,
-						family=binomial))
-						
+						family=binomial)))
+
 exp.coef = apply(GT[,-(1:3)], 1, function(x)
 					exp(coefficients(glm2(x ~
 							samples$PC1 + 
@@ -40,9 +47,31 @@ exp.coef = apply(GT[,-(1:3)], 1, function(x)
 							samples$PC4 +
 							samples$PC5 + 
 							samples$average_quality_of_mapped_bases,
-						family=binomial)))[[7]])						
+						family=binomial)))[[7]])
+
+plt = cbind(samples, model)
+plt = melt(plt, id=c('Name','Pop','average_quality_of_mapped_bases','PC1','PC2','PC3','PC4','PC5'))
+q = list()
+for(p in seq(1,14)){
+	sub = plt[which(plt$variable == paste('V',p,sep='')),]
[…]

### 2019-02-13 PlottingScripts/Clean_RegressionPlots.R

@@ -1,0 +1,162 @@
+## Deal with Regression output
+
+library(dplyr)
+library(magrittr)
+library(data.table)
+library(ggplot2)
+library(multtest)
+library(cowplot)
+library(reshape2)
+
+Regression <-fread('/Users/luke/Documents/PC5/GenomeWide.5PC.Regression.clean.txt', colClasses='numeric',col.names=c('CHROM','POS','ID','dev.p'))
+
+Regression $ID<-as.numeric(as.character(Regression $ID))
+
+#read in list of global singletons
+singles<-fread(
+	'/Users/luke/genomes/genomes/1kGP_NotFiltered/GenomeWide_doubles.frq',
+	col.names=c('CHROM','POS'))
+singles$drop<-'Yes' #dummy variable for merging
+
+#join the singles to the Reg to get info of what sites to remove
+Regression <-left_join(Regression, singles, by=c('CHROM','POS'))
+rm(singles)
+
+#this removes the singletons
+Regression <-Regression[which(is.na(Regression$drop)),]
+Regression$drop<-NULL
+
+##read Repeats
+Repeats <- fread('~/Documents/QualityPaper/Misc/Reg_strictMask_nestedRepeats.POS',fill=T,col.names=c('CHROM','POS'))
+Repeats$Repeat='yes'
+
+##read Indels
+Indels <- fread('/Users/luke/Documents/QualityPaper/Misc/Reg_indels.POS', skip=1 ,col.names=c('CHROM','POS'))
+Indels$Indel = 'yes'
+
+#read in rsIDs (used for GWAS catalogue search)
+rsID<-fread('/Users/luke/genomes/genomes/hg19/phase3/rsID.txt')
+colnames(rsID)<-c('CHROM','POS','rsID')
+
+Regression <- left_join(Regression, Repeats, by=c('CHROM','POS'))
+
+Regression <- left_join(Regression, Indels, by=c('CHROM','POS'))
+
+Regression <- left_join(Regression, rsID, by=c('CHROM','POS'))
+
+#free up memory
+rm(Repeats)
+rm(Indels)
+rm(rsID)
+
+write.table(Regression, file='/Users/luke/Documents/QualityPaper/Misc/TotalRegressionPC5.csv',row.names=F)
+
[…]

### 2019-02-13 PlottingScripts/TotalRegression.R

@@ -85,7 +85,7 @@ Regression <- left_join(Regression, rsID, by=c('CHROM','POS'))
 
 write.table(Regression, file='/Users/luke/Documents/QualityPaper/Misc/TotalRegressionPC5.csv',row.names=F)
 
-Regression<-fread('/Users/luke/Documents/QualityPaper/Misc/TotalRegression.csv')
+Regression<-fread('/Users/luke/Documents/QualityPaper/Misc/TotalRegressionPC5.csv')
 
 Repeat_Indels <- unique(Regression[which(Regression$Indel=='yes' 
 								& Regression$Repeat=='yes'),])
@@ -117,6 +117,13 @@ Repeat_Indels <-TSBH(Repeat_Indels)
 NORepeat_Indels <-TSBH(NORepeat_Indels)
 Repeat_SNPs <-TSBH(Repeat_SNPs)
 NORepeat_SNPs <-TSBH(NORepeat_SNPs)
+
+
+SNPs <- rbind(Repeat_SNPs,NORepeat_SNPs)
+###OMNI CHIP
+omni <- fread('~/Documents/QualityPaper/sig/Omni2-5.txt',header=F)
+omni <- merge(omni, SNPs, by.x='V2', by.y='rsID')
+omni <- omni[which(omni$V2 != '.'),]
 
 write.table(Repeat_Indels, 
 file='/Users/luke/Documents/QualityPaper/Misc/Repeat_Indels.csv',row.names=F)

### 2019-02-14 legacy.tex
Updated text

@@
 \usepackage{afterpage}
+\usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage{hyperref}
@@
 
+\begin{figure}
+\includegraphics[width=0.95\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
+
+\caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
+Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
+The x-axis is ranked by the mean sequencing date per population.
+\textbf{B} The x-axis is sorted by the sequencing date per individual.
+The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project.
+The bottom plot indicates the sequencing technologies used over time.}
+\label{MapQual}
+\end{figure}
+
 To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
@@
 
-\luke{Added text here}To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype with the top five principal components as covariates to account for population structure and $Q$ as the last covariate.
-This method identifies a total of 1,409,401 statistically significant variants associated to $Q$ using a FDR cutoff of $\alpha = 0.01$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
-The total number of sites tested is 37,597,679 meaning that 3.75\% of the sites tested are statistically significantly associated to $Q$.
-%we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 10756
-We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable})
+\luke{Added text here}To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$. 
+To account for population structure, we also included the top five global principal components.
+This method identifies a total of 22,186 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
@@
 \begin{table}[h]
+\centering
 \begin{tabular}{l  r r}
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 284,551 & 1,059,354 \\  
-{INDEL} & 15,680 & 49,816 \\ \hline
+{SNP}  & 4,405 & 15,018 \\  
+{INDEL} & 642 & 2,121 \\ \hline
 \end{tabular}
@@
 SNPs and INDELs were also analyzed separately.
-The total number of sites tested is 37,597,679 meaning that 3.75\% of the sites tested are statistically significantly associated to $Q$.}
+The total number of 22,186 are statistically significantly associated to $Q$.}
 \label{sigTable}
 \end{table}
-
-\begin{figure}
-\includegraphics[width=0.95\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
-
-\caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project.
-Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
-The x-axis is ranked by the mean sequencing date per population.
-\textbf{B} The x-axis is sorted by the sequencing date per individual.
-The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project.
-The bottom plot indicates the sequencing technologies used over time.}
-\label{MapQual}
-\end{figure}
 
@@
 
+\subsection{Cell line or technical artifact}
+
+To determine whether this signal is the result of a cell line or technical artifact, we investigated other cohorts that used the same cell lines.
+Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
+The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{}.
+If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
+From the list of 22,186 variants identified as being associated to $Q$, only 22 variants were present in HapMap and 143 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
+
+In 2017, Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
+We find that 9 out of the 297 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
+While this is more than 1\% of the sites, we suspect that they are indeed false associations to $Q$.
+To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
+We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
+These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
+
 \subsection{Suspicious variants impact modern genomics analyses}
+
 \subsubsection{Imputation}
@@
 
-\subsubsection{Cell line or technical artifact}
-
-To determine whether this artifact is the result of a cell line artifact or a technical artifact, we investigated other cohorts that used the same cell lines.
-The HapMap project and 1000 Genomes project shared some cell lines but used different sequencing platforms to produce the data.
-Technological artifacts in one technology aren't expected to be present in another sequencing platform.
-Moreover, cell line artifacts would likely be present in data from two separate platforms.
-
-Using the single population association test we produced a list of variants associated to $Q$ for the samples shared between the two sequencing projects.
-From this list, we found that only 70 variants were present in HapMap and 186 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
-However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
-
-% \sgcomment{This could be fleshed out a bit. Currently reader has no idea whether the proportion of variants that are present in other cohorts is more or less than expected.}
-
-%However, considering that there are only 70 suspicious variants that are concordant between hapmap and 1000 Genomes data, this is consistent with it being a false positive association.
-%Given the false discovery rate adjustment with a type-I error rate $\alpha = 0.01$, we expect most the variants we identified not to be real \sgcomment{previous phrase needs rewriting.
-I don't love the "real" vs "not real".
-Or at least you should clarify that we detect problematic mutations.
-The next question is whether these are real and badly called, or completely spurious I think this paragraph suggest that the majority are spurious}; however we cannot discount the chance that they are real.
-In 2017, Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-We find that 9 out of the 297 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
-While this is more than 1\% of the sites, we suspect that they are indeed false associations to $Q$.
-To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
-We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
-These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
-
 \subsubsection{Microarrays}
-We found that 6,456 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
+We found that 476 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
 This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
@@
 And one study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
+The two studies that imputed using HapMap are likely not impacted by this as these variants are likely falsely associated to $Q$.
 
@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for some biological traits.
-The two studies that imputed using HapMap are likely not impacted by this as these variants are likely falsely associated to $Q$.
-
-\luke{removed a paragraph here}
-%The study that used the raw BAM files from the 1kGP to identify EBV copy number by comparing the number of reads that mapped to EBV \citep{Mandage2017}.
-%Based on GWAS on the EBV copy number, the study reports only one SNP with significance below $ p < 10^{-8}$, as well as 104 SNPs in the range of $p < 10^{-7}$ to $ p < 10^{-6}$. 
-%Four of the SNPs in the top hits for the Asian populations (rs200655768, rs184202621, rs201255786, rs201761909) were identified by our regression against $Q$.
-
-%The EBV strain was used to transform B-cells from blood samples from the 1kGP into lymphoblastoid cell lines.
-%This is a commonly used transformation that helps preserve cell lines for long periods of time.
 
@@
 \begin{tabular}{l l l r r r}
-  {Pubmed ID}  & {Journal} & {rsID} & {GWAS -$\log_{10} p$} & {Adjusted $Q$ -$\log_{10} p$} & {odds ratio}     \\ \hline
+  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)} & \multicolumn{1}{p{1cm}}{\centering Odds \\ ratio}     \\ \hline
  28654678	& PLoS One	& rs201761909	& 5.70	& 100.67 & 0.63\\
@@
  	& 		& rs200699422	& 5.30	& 7.82 & 0.87\\ 
-23527680 &	Twin Res Hum Genet	& rs6057648	& 5.40	& 25.50 & 0.82\\  
-21386085 &	Diabetes &	rs301 &*10.52	 & 8.22 & 0.90\\
-28928442	& Nature Commun	& rs201471471	& 6.52	& 7.14 & 0.67\\
-29255261	& Nature Genetics	& rs11039216	& *14.30	& 2.59 & 1.08\\
-25918132	& Toxicol Sci	& rs76780579	& 6.00	& 2.58	& 1.17 \\
-27863252	& Cell	& rs3794738	& *13.16	& 2.39 & 0.91\\
-28698626	& Science Reports	& rs11015915	& 5.05	& 2.05 & 1.08\\
+	&		&			&		&		&\\
+23527680 &	\multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics}	& rs6057648	& 5.40	& 25.50 & 0.82\\  
+21386085 &	\multicolumn{1}{p{3cm}}{\raggedright Diabetes\\} &	rs301 &*10.52	 & 8.22 & 0.90\\
+28928442	& \multicolumn{1}{p{3cm}}{\raggedright Nature Communications}	& rs201471471	& 6.52	& 7.14 & 0.67\\
+29255261	& \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics\\}	& rs11039216	& *14.30	& 2.59 & 1.08\\
+25918132	& \multicolumn{1}{p{3cm}}{\raggedright Toxicological\\Sciences}	& rs76780579	& 6.00	& 2.58	& 1.17 \\
+27863252	& \multicolumn{1}{p{3cm}}{\raggedright Cell\\}	& rs3794738	& *13.16	& 2.39 & 0.91\\
+28698626	& Scientific Reports	& rs11015915	& 5.05	& 2.05 & 1.08\\
 
@@
 
+\begin{figure}
+\includegraphics[width=15cm,keepaspectratio]{./Figures/PublishedSNPs.jpg}
+\caption{For each of the suspicious variants identified in the GWAS catalogue a plot of the average quality of mapped bases $Q$ compared to the predicted value from logistic regression.
+The strong correlation of $Q$ and the predicted value indicates that $Q$ is strongly associated to genotype for these variants.
+A negative correlation indicates that individuals with low $Q$ have higher likelihood of having a variant, whereas a positive correlation indicates that these individuals have a lower likelihood of having a variant at this position.}
+\label{PubSig}
+\end{figure}
 \section{Discussion}
@@
 
+Sequencer-specific errors from the 1kGP would likely be absent from HapMap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}.  
+Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
+This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
+
  We were not able to pinpoint the precise technical source of the discrepancy. 
-Sequencer-specific errors would be expected to be absent from Hapmap \sgcomment{why?
-Do you mean that they would be different between 1kgp and hapmap?}, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}.
-We find that $99\%$ of the suspicious variants are indeed absent from HapMap \sgcomment{what is this $99\%$?
-concordance?
-I thought that the result was a bit different}, which is consistent with the expected false positive rate in our test.
+Further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal. 
+For instance, our test was limited by the genome wide resolution of the quality per sample.
+Data quality at a SNP-wise resolution would be a better metric to use, but unfortunately was not available.
 
-The presence of some of these variants in HapMap suggests a cell line effect; although, this result is not very convincing with only 70 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
-\sgcomment{I don't think you want to start sentence with "although,".
-}
-We also note that 186 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
-Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal. \sgcomment{This paragraph needs a bit of work}
-
-If these \sgcomment{Which?} sequencing platforms are no longer being used, then it it likely that these \sgcomment{which?} artifacts are no longer being actively introduced in recent sequence data.
-\sgcomment{It's good to start paragraph in more self-contained manner}
+If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer being actively introduced in recent sequence data.
 However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
-\sgcomment{Even though we have found only one GWAS study that reported associations with the problematic variants, the imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores and biases in fine mapping analyses.
-}
-%However, this is not the only method that would be affected by these false positives.
-Polygenic risk scores, meta analyses and other methods that combine the scores of multiple variants might also include some of these variants in their analyses.
-These variants do have a sizeable effect on population on analyses that compare and contrast populations based on mutational profiles.
+Even though we have found only one GWAS study that reported associations with the problematic variants, the imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores and biases in fine mapping analyses.
+These variants do have a sizeable effect on population analyses that compare populations based on mutational profiles.
 Even though we expect that present-day technologies will be less sensitive to such artefacts, if is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
@@
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
-A conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
+In order to avoid spurious results, the most conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
 We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signal.
-Another approach is to remove all the sites that reach a significance below $ p < 10^{-6}$ in at least two populations or using the sites reaching a significance below $ p < 10^{-6}$ in the combined test.
-(See supplementary material for lists of variants) \sgcomment{This is not really a recommendation.
-We should be a bit more specific}
+Another approach is to remove all the sites that reach a significance below $ p < 10^{-2}$ in the reverse GWAS (See supplementary material for lists of variants statistically associated to $Q$).
+We also recommend the resequencing of samples below an acceptable $Q$ threshold to render them usable.
 
@@
 For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis.
-The only filtration used was the application of the  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} which resulted in a total of 80,971,616 variants ($S$) included in the test (see Table \ref{totTable}).
+The only filtration used was the application of an allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of 37,674,915 variants ($S$) included in the test (see Table \ref{totTable}).
 We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 22,651,762 & 51,643,590  \\  
-{INDEL} &  2,176,825  & 4,499,439 \\ \hline
+{SNP}  & 8,393,290 & 26,240,754   \\  
+{INDEL} &  749,724  & 2,291,150 \\ \hline
 \end{tabular}
@@
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 1,409,401 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 22,186 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Non_Repeat_SNPs.jpg}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}
+        \label{fig:a}
+    \end{subfigure} %
 
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 1,059,354 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=90mm]{./Figures/NORepeat_SNPs_QQ.jpg}
+        \label{fig:b}    
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
+There are 15,018 SNPs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.}
   \label{NRS_Manhattan}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Non_Repeat_Indels.jpg}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}
+        \label{fig:a}
+    \end{subfigure} %
 
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 49,816 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=90mm]{./Figures/NORepeat_Indels_QQ.jpg}
+        \label{fig:b}    
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
+There are 2,121 INDELs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.}
   \label{NRI_Manhattan}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}
+        \label{fig:a}
+    \end{subfigure} %
 
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 284,551 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=90mm]{./Figures/Repeat_SNPs_QQ.jpg}
+        \label{fig:b}    
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
+There are 4,405 SNPs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.}
   \label{RS_Manhattan}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_Indels.jpg}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_Repeat_Indels.jpg}
+        \label{fig:a}
+    \end{subfigure} %
 
-\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
-There are 15,680 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=90mm]{./Figures/Repeat_Indels_QQ.jpg}
+        \label{fig:b}    
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
+There are 642 INDELs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.}
   \label{RI_Manhattan}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
 
-\caption{Empirical distribution plots of the sum of deviances for Non-Repeat variants grouped by N, the number of populations included.
+%\caption{Empirical distribution plots of the sum of deviances for Non-Repeat variants grouped by N, the number of populations included.
 The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom.
@@
 }
-\label{Deviances}
-\end{figure}
+%\label{Deviances}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
-\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
-\label{MostSig}
-\end{figure}
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
+%\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
+%\label{MostSig}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
-\caption{SNPs associated with $Q$ in a single population. The x-axis is sorted by genomic position.}
-\label{Singles}
-\end{figure}
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
+%\caption{SNPs associated with $Q$ in a single population. The x-axis is sorted by genomic position.}
+%\label{Singles}
+%\end{figure}
 

### 2019-02-18 legacy.tex
Updated text.

@@
 
-\luke{removed paragraph about meta-analysis}
-%We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-%Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
+We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
 
@@
 
-\luke{removed figure 3}
-%\begin{figure}
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
 
-%\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-%The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
-%The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
-%Phase 1 populations are marked by a star ( * ).
-%The line plot underneath shows the number of populations for which a variant has reached significance.
-%The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
-%The same variants identified as being low quality independently in each population are found in other populations. }
-%  \label{OverLap}
-%\end{figure}
+\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
+The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
+The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+Phase 1 populations are marked by a star ( * ).
+The line plot underneath shows the number of populations for which a variant has reached significance.
+The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
+The same variants identified as being low quality independently in each population are found in other populations. }
+  \label{OverLap}
+\end{figure}
 
@@
 Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
-The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{}.
+The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
 If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
 From the list of 22,186 variants identified as being associated to $Q$, only 22 variants were present in HapMap and 143 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
-However, with our FDR type-I error rate of $\alpha = 0.01$, these variants could be explained by false positive associations.
+Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
+This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
 
@@
 
-\begin{figure}
-\includegraphics[width=15cm,keepaspectratio]{./Figures/PublishedSNPs.jpg}
-\caption{For each of the suspicious variants identified in the GWAS catalogue a plot of the average quality of mapped bases $Q$ compared to the predicted value from logistic regression.
+%\begin{figure}
+%\includegraphics[width=15cm,keepaspectratio]{./Figures/PublishedSNPs.jpg}
+%\caption{For each of the suspicious variants identified in the GWAS catalogue a plot of the average quality of mapped bases $Q$ compared to the predicted value from logistic regression.
 The strong correlation of $Q$ and the predicted value indicates that $Q$ is strongly associated to genotype for these variants.
 A negative correlation indicates that individuals with low $Q$ have higher likelihood of having a variant, whereas a positive correlation indicates that these individuals have a lower likelihood of having a variant at this position.}
-\label{PubSig}
-\end{figure}
+%\label{PubSig}
+%\end{figure}
 \section{Discussion}
@@
 
-Sequencer-specific errors from the 1kGP would likely be absent from HapMap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}.  
-Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
-This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
-
  We were not able to pinpoint the precise technical source of the discrepancy. 
@@
 If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer being actively introduced in recent sequence data.
-However, because the 1kGP data is widely used as a reference database, these variants are being imputed onto new genotype data.
-Even though we have found only one GWAS study that reported associations with the problematic variants, the imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores and biases in fine mapping analyses.
+However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes.
+Our model only reveals association to quality; some variants may be false positives in cases where only individuals with low $Q$ scores have the variant and other variants may be false negatives, in cases where individuals with low $Q$ appear to have lower calling rates for particular.
+
+The imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses as well as impact population specific allele frequency estimates.
 These variants do have a sizeable effect on population analyses that compare populations based on mutational profiles. 
@@
 \begin{align*}
-{logit}{E(Y_{s'i})} &= \beta_{0}{PC1}_{i} + \beta_{1}{PC2}_{i} + \beta_{2}{PC3}_{i} + \beta_{3}{PC4}_{i} + \beta_{4}{PC5}_{i} + \beta_{5}Q_{i}
+{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{PC1}_{i} + \beta_{2}{PC2}_{i} + \beta_{3}{PC3}_{i} + \beta_{4}{PC4}_{i} + \beta_{5}{PC5}_{i} + \beta_{6}Q_{i}
 \\
@@
 Here $Y_{si}$ is a Bernoulli random variable denoting the genotype of individual $i$ at genomic locus $s$. 
-In the model, the regression coefficient $\beta_{0}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{1}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
+In the model, the regression coefficient $\beta_{1}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
 We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
 
-For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{1}^{(s)}=0$. That is, the average quality of mapped bases $Q_{p}$ variable is not associated with genotype $Y_{s}$.
+For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{6}^{(s)}=0$. That is, the average quality of mapped bases $Q$ variable is not associated with genotype $Y_{s}$.
 In other words, the slope coefficient of the simple logistic regression is zero.
@@
 
-\subsection{Frequencies of variants in other datasets}
-Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
+%\subsection{Frequencies of variants in other datasets}
+%Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
 We then compared the frequencies of these 70 variants in the 1kGP Japanese samples to those of the HapMap Japanese samples.
@@
 
-Similarly, using the allele frequencies from GnomAD, we were able to extract 186 sites that are statistically significantly associated to the average quality of mapped bases.
+%Similarly, using the allele frequencies from GnomAD, we were able to extract 186 sites that are statistically significantly associated to the average quality of mapped bases.
 However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
-\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
-\label{TwinsSNP}
-\end{figure}
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
+%\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
+%\label{TwinsSNP}
+%\end{figure}
 
@@
 \begin{figure}[h]
-\includegraphics[width=8cm,keepaspectratio]{./Figures/Han83.jpg}
-\caption{Site frequency spectrum plot of the significant SNPs identified in the single population tests of CHS and CHB populations.
+\centering
+\includegraphics[width=10cm,keepaspectratio]{./Figures/Han83.jpg}
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
+This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
 Only 9 of the 297 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
 Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
+\label{90HanSFS}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=10cm,keepaspectratio]{./Figures/Han_1kGP_SFS.jpg}
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
+This plot shows significant SNPs identified in the full model including all populations.
+Only 1899 of the variants are present in the 83 individuals included in both the original 1kGP and the high depth resequenced individuals.}
 \label{90HanSFS}
@@
 
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
-\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
+%\begin{figure}[h]
+%\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
+%\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
 \sgcomment{Change title, check targeted  sequencing region.
@@
 There are 70 such variants in HapMap and 186 in the GnomAD.}
-\label{HapMap_GnomAD}
-\end{figure}
+%\label{HapMap_GnomAD}
+%\end{figure}
 

### 2019-02-18 Misc/VCF_to_R_format.py

@@ -11,9 +11,8 @@ def Bash_cmd(i):
 
     os.system('java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar \
     extractFields {0}/ALL.chr{1}.{2} CHROM POS AF "GEN[*].GT" \
-    | sed "s/0|[1-9]/1/g ; s/[1-9]|0/1/g ; s/[1-9]|[1-9]/1/g ; /|/d"\
-    | tail -n +2 | awk "\$3 > 0.000599" | gzip > /Users/luke/genomes/genomes/hg19/Genotypes/CHR{1}.Genotypes.txt.gz'.format(path,i,name))
-    #NOTE : the [1-9] portion of the SED allows for multiallelic Snp
+    | sed "s/0|0/0/g ; s/0|[1-9]/1/g ; s/[1-9]|0/1/g ; s/[1-9]|[1-9]/1/g ; /|/d"\
+    | tail -n +2 | awk "\$3 > 0.000599" | gzip > /Users/luke/genomes/genomes/hg19/Genotypes/DoubleCheck/CHR{1}.Genotypes.txt.gz'.format(path,i,name))
 
 if __name__ == '__main__':
     #you can replace ListOfChrom with whatever you are iterating through
@@ -26,5 +25,3 @@ if __name__ == '__main__':
     #some magic, not sure what this does
     pool.close()
     pool.join()
-
-#s/\//|/g ; s/0|0/0/g ; s/0|1/1/g ; s/1|0/1/g ; s/1|1/1/g ; /|/d

### 2019-02-18 PlottingScripts/Han_1kGP_SFS_clean (Autosaved).R

@@ -29,7 +29,7 @@ agg <- combo %>% group_by(CHROM,POS) %>% summarize(hanAF = sum(Han.gt), kgpAF = 
 
 
 ggplot(agg, aes(x=hanAF, y=kgpAF))+geom_bin2d()+scale_fill_distiller(palette = "Spectral")+labs(x='Resequenced Han', y='1000 Genomes Project')+ggtitle('Joint Frequency Plot of Suspicious variants')
-ggsave('~/Documents/QualityPaper/Figures/Han_1kGP_SFS.jpg',height=4, width=12)
+ggsave('~/Documents/QualityPaper/Figures/Han_1kGP_SFS.jpg',height=10, width=10)
 
 #concordance per sample
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')

### 2019-02-18 legacy.tex
Added text about resequenced data

@@
 If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-From the list of 22,186 variants identified as being associated to $Q$, only 22 variants were present in HapMap and 143 in GnomAD - a compilation of multiple databases (Figure \ref{HapMap_GnomAD}) \citep{HapMap2005,Lek2016}.
+From the list of 22,186 variants identified as being associated to $Q$, only 108 variants were present in HapMap \citep{HapMap2005}.
+%and 143 in GnomAD - a compilation of multiple databases %Lek2016
 Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
@@
 We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
-These result suggest that the variants we have identified using this reverse-GWAS approach are the result of a technical artifact and not a cell line artifact.
+
+We find that of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
+These variants are likely unaffected by $Q$ in these particular individuals.
+We find that 772 variants are at higher frequency in the resequenced data.
+These variants may be associated to $Q$ due to false negatives, whereby individuals with lower $Q$ tend to have lower calling rates for these variants.
+Only 194 variants are at higher frequencies in the original 1kGP data.
+These variants may be associated to $Q$ due to false positives, whereby individuals with lower $Q$ have higher calling rates for these variants (See Figure \ref{90HanSFS_full}).
 
@@
 \subsubsection{Imputation}
+
 The state of the art imputation servers use a combination of many databases including some that are not freely available.
@@
 \subsubsection{Microarrays}
+
 We found that 476 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
@@
 \subsubsection{GWAS}
+
 We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
@@
 Only 1899 of the variants are present in the 83 individuals included in both the original 1kGP and the high depth resequenced individuals.}
-\label{90HanSFS}
+\label{90HanSFS_full}
 \end{figure}

### 2019-02-19 legacy.tex
cleaned up misc hashed out text
small edits

@@
 
-\sgcomment{Check time concordance throughout}		
-%\subsection{Batch Effects in Aging Reference Cohort Data}
+\subsection{Batch Effects in Aging Reference Cohort Data}
 
@@
 
-Even though such batch effects clearly have the potential to confound analyses, the usefulness and ubiquity of the 1kGP data make such issues particularly insidious.   
+Even though such batch effects clearly have the potential to confound analyses, the unrestricted availability of the 1kGP data lead to ubiquitous use making batch effects particularly insidious.   
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 
-Some mutagenic processes will affect different DNA motifs. 
+Different mutagenic processes may preferentially affect different DNA motifs. 
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-These mismatches were maintained after filtering for low-quality regions of the human genome and sites failing Hardy-Weinberg equilibrium tests. 
-%\sgcomment{did we have a specific result for this? As in "$94\%$ of sites passed HW"...}. 
-%Even though some of the mismatched variants are at low frequency in the 1kGP, and thus can evade Hardy-Weinberg detection, but many are at high enough frequency that HW detection   
-%Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.
+These mismatches were present despite filtering for low-quality regions of the human genome and Hardy-Weinberg disequilibrium.
 
@@
 %(\sgcomment {p value?})
-Thus, sequences with low mapping quality harbour mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+Thus, sequences called from low-$Q$ data contain mutations that reproduce poorly across studies and exhibit a particular mutational signature.
 
@@
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses, removing suspicious SNPs may be adequate. 
-Lists of suspicious SNPs and low-quality individuals is provided as supplementary material \luke{link to file here}.
+For other analyses where quality of imputation matters, identifying suspicious SNPs may be preferable. 
+Lists of suspicious SNPs and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
 
@@
 By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
-%Among the many covariates that could be used to identify spurious sites, such as sequencing date, sequencing technology, or sequencing centre, we used average quality of mapped bases ($Q$) per individual because it best correlated with mismatched sites across Japanese cohorts.
 
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ (Figure \ref{OverLap}).
+Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
@@
 
-We find that of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
+Of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
 These variants are likely unaffected by $Q$ in these particular individuals.
@@
 
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}
-
-%\caption{Empirical distribution plots of the sum of deviances for Non-Repeat variants grouped by N, the number of populations included.
-The blue curve is the theoretical null reference chi squared distribution with $N$ degrees of freedom.
-Note $\sum_i^N  \chi^2_1= \chi^2_N$.
-The grey dotted line indicates the significance level of $10^{-5}$ for the respective distribution which happens to be very close to the adjusted \textit{p}-value threshold at 0.01 FDR.
-}
-%\label{Deviances}
-%\end{figure}
-
 \begin{figure}[h]
@@
 \end{figure}
-
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
-%\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
-%\label{MostSig}
-%\end{figure}
-
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
-%\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
-%\label{DiabetesSNP}
-%\end{figure}
-
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
-%\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
-%\label{TwinsSNP}
-%\end{figure}
-
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
-%\caption{SNPs associated with $Q$ in a single population. The x-axis is sorted by genomic position.}
-%\label{Singles}
-%\end{figure}
 

### 2019-02-19 legacy.tex
new figures for Han Reseq + minor text

@@
 Recent reports have identified differences in the mutational spectra across human populations.
-While some of these reports have been replicated in other cohorts, most have been reported in the 1000 Genomes project data.
-While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1000 Genomes Project data and to the apparent population stratification.
-Because the 1000 Genomes data is used extensively, we find that the spurious calls also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
-Lower-quality data from the early phases of the 1000 Genomes project thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
+While some of these reports have been replicated in other cohorts, most have been reported in the 1000 Genomes Project (1kGP) data.
+While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
+Because the 1kGP data is used extensively, we find that the spurious calls also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
+Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
 \end{abstract}
@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
 Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
@@
 The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved. 
-Yet, because of the extraordinary value of freely available data, early data from the 1000 Genomes project (1kGP) is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
 The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
@@
 Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
+
+%This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
+%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
+%Large cohorts have recently become available to increase statistical power to resolve some of these fine scale mutational signatures.
 

### 2019-02-19 Misc/MakeRegressionPlot.R

@@ -22,15 +22,24 @@ mGT = melt(GT[,-(1:3)])
 mGT <- merge(samples, mGT, by.y='variable',by.x='Name')
 cGT <- mGT %>% group_by(Pop) %>% summarize(var = sum(value))
 keep <- cGT[which(cGT$var > 0),]$Pop
+if(length(keep) < 8){
+	cols = 1
+}
+if(length(keep) > 8){
+	cols = 2
+}
+if(length(keep) > 16){
+	cols = 3
+}
 
 ggplot(mGT[which(mGT$Pop %in% keep),], aes(x= average_quality_of_mapped_bases, y=value)) +
- 	facet_wrap(.~Pop, ncol=1) + 
+ 	facet_wrap(.~Pop, ncol=cols) + 
  	geom_smooth(method = "glm", 
      			method.args = list(family = "binomial"), 
      			se = FALSE) +
-     geom_point() +
+     geom_point(shape=1) +
      theme_classic() + 
      labs(y='genotypes', title = paste(Chr," : ",Pos)) +
-     theme(plot.title = element_text(hjust = 0.5))
+     theme(plot.title = element_text(hjust = 0.5), axis.text.y=element_blank())
 
 ggsave(paste('~/Documents/QualityPaper/Misc/Rplot_LogisticReg/',Chr,'_',Pos,'.jpg',sep=''), height = 8, width=6)

### 2019-02-19 PlottingScripts/AllSig_RunModel (Autosaved).R

@@ -1,0 +1,44 @@
+library(dplyr)
+library(data.table)
+library(ggplot2)
+library(glm2)
+
+#read colors per pop (used in 1kG Nature)
+colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
+MyColour <- as.character(colors $hex)
+names(MyColour) <- colors $Pop
+
+Sig = fread('/Users/luke/Documents/QualityPaper/Misc/SignificantVariants.csv',col.names=c("CHROM", "POS", "AF", "Rawlog10P", "Repeat", "Indel", "rsID", "log10P_0.01", "P_0.01", "p"))
+
+#read Genotypes
+GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_1kGP_clean_GT.txt')
+ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
+names(GT) = as.character(unlist(ColNames))
+samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
+samples$Pop = factor(samples$Pop, levels=c('FIN','GBR','CEU','IBS','TSI','CHS','CDX','CHB','JPT','KHV','GIH','STU','PJL','ITU','BEB','PEL','MXL','CLM','PUR','ASW','ACB','GWD','YRI','LWK','ESN','MSL'))
+
+model = apply(GT[,-(1:3)], 1, function(x)
+						glm2(x ~
+							samples$PC1 + 
+							samples$PC2 +
+							samples$PC3 +
+							samples$PC4 +
+							samples$PC5 + 
+							samples$average_quality_of_mapped_bases,))
+#get coef of Qual per site
+coef = lapply(model, function(x) coef(x)[[7]])				
+
+dev = lapply(model, function(x) anova(x, test='Chi')[[7,2]])
+
+coef = cbind(unlist(coef), GT[,c(1:3)])
+coef = cbind(coef, unlist(dev))
+coef$ID = as.numeric(as.character(coef$ID))
+coef$pval = -log10(pchisq(coef$V2, 1, lower.tail=F))
+names(coef) = c('Beta','CHROM','POS','AF','deviance','-log10(p)')
[…]

### 2019-02-19 PlottingScripts/Han_1kGP_SFS_clean (Autosaved).R

@@ -1,58 +1,0 @@
-##plot SFS sig Han
-library(data.table)
-library(ggplot2)
-library(reshape2)
-
-#read Genotypes
-kgp1 = fread('~/Documents/QualityPaper/sig/SigInHan_GenomeWide_gt.txt')
-ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
-names(kgp1) = as.character(unlist(ColNames))
-#melt to long format
-kgp = melt(kgp1, id=c('CHROM','POS','ID'))
-
-#read Genotypes
-han1=fread('~/Documents/QualityPaper/sig/Han90_SigInHan_GenomeWide_gt.txt')
-HanNames=fread('~/genomes/genomes/Han90/SignificantSNPs_Han90_COLNAMES.txt',header=F)
-names(han1) = as.character(unlist(HanNames))
-#melt to long format
-han = melt(han1, id=c('#CHROM','POS','ID'))
-#fix chromosome names
-han$CHROM = as.numeric(as.character(substring(han$'#CHROM',4)))
-han$'#CHROM' <- NULL
-
-#merge 1kGP and han resequenced data.
-combo <- merge(han, kgp, by=c('CHROM','POS','variable'), all=F)
-#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
-combo <- combo %>% group_by(CHROM,POS,variable) %>% summarize(Han.gt = max(value.x), Kgp.gt = max(value.y))
-#to get allele frequencies we group by POS and summarize GT
-agg <- combo %>% group_by(CHROM,POS) %>% summarize(hanAF = sum(Han.gt), kgpAF = sum(Kgp.gt)) 
-
-
-ggplot(agg, aes(x=hanAF, y=kgpAF))+geom_bin2d()+scale_fill_distiller(palette = "Spectral")+labs(x='Resequenced Han', y='1000 Genomes Project')+ggtitle('Joint Frequency Plot of Suspicious variants')
-ggsave('~/Documents/QualityPaper/Figures/Han_1kGP_SFS.jpg',height=10, width=10)
-
-#concordance per sample
-samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
-#get quality per sample
-combo = merge(combo, samples[,c('Name','average_quality_of_mapped_bases')], by.x='variable',by.y='Name')
[…]

### 2019-02-19 PlottingScripts/Han_1kGP_SFS_clean.R

@@ -4,13 +4,6 @@ library(ggplot2)
 library(reshape2)
 
 #read Genotypes
-kgp1 = fread('~/Documents/QualityPaper/sig/SigInHan_GenomeWide_gt.txt')
-ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
-names(kgp1) = as.character(unlist(ColNames))
-#melt to long format
-kgp = melt(kgp1, id=c('CHROM','POS','ID'))
-
-#read Genotypes
 han1=fread('~/Documents/QualityPaper/sig/Han90_SigInHan_GenomeWide_gt.txt')
 HanNames=fread('~/genomes/genomes/Han90/SignificantSNPs_Han90_COLNAMES.txt',header=F)
 names(han1) = as.character(unlist(HanNames))
@@ -20,16 +13,37 @@ han = melt(han1, id=c('#CHROM','POS','ID'))
 han$CHROM = as.numeric(as.character(substring(han$'#CHROM',4)))
 han$'#CHROM' <- NULL
 
+#read Genotypes
+kgp1 = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_1kGP_clean_GT.txt')
+ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
+names(kgp1) = as.character(unlist(ColNames))
+#melt to long format
+kgp = melt(kgp1, id=c('CHROM','POS','ID'))
+#keep only the 83 resequenced individuals
+kgp = kgp[which(kgp$variable %in% han$variable),]
+
 #merge 1kGP and han resequenced data.
-combo <- merge(han, kgp, by=c('CHROM','POS','variable'), all=F)
+combo <- merge(han, kgp, by=c('CHROM','POS','variable'), all=T)
+#missing sites replaced with 0
+combo[is.na(combo)] <- 0 
 #to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
 combo <- combo %>% group_by(CHROM,POS,variable) %>% summarize(Han.gt = max(value.x), Kgp.gt = max(value.y))
+
 #to get allele frequencies we group by POS and summarize GT
 agg <- combo %>% group_by(CHROM,POS) %>% summarize(hanAF = sum(Han.gt), kgpAF = sum(Kgp.gt)) 
+#remove sites missing in both groups
[…]

### 2019-02-21 legacy.tex
minor edits

@@
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
-Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
+Thus, the impact of mutational processes can be inferred by considering the DNA context of a polymorphism \citep{Alexandrov2013,Shiraishi2015a}. 
+%Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
 Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.

### 2019-02-26 legacy.tex
more stuff

@@
 
-\luke{Added text here}To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$. 
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$. 
 To account for population structure, we also included the top five global principal components.
-This method identifies a total of 22,186 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 14,322 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
+To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.001$ \citep{Benjamini2006}.
 
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 4,405 & 15,018 \\  
-{INDEL} & 642 & 2,121 \\ \hline
+{SNP}  & 2,819 & 9,846 \\  
+{INDEL} & 378 & 1,279 \\ \hline
 \end{tabular}
@@
 
-To determine whether this signal is the result of a cell line or technical artifact, we investigated other cohorts that used the same cell lines.
+To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
 Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
@@
 If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-From the list of 22,186 variants identified as being associated to $Q$, only 108 variants were present in HapMap \citep{HapMap2005}.
-%and 143 in GnomAD - a compilation of multiple databases %Lek2016
+From the list of 14,322 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
 Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
@@
 
-In 2017, Lan et al. resequenced 90 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-We find that 9 out of the 297 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
-While this is more than 1\% of the sites, we suspect that they are indeed false associations to $Q$.
+In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
+After downsampling the 1kGP data to match the resequenced individuals, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
+While this is more than 0.1\% of the sites, we suspect that they are indeed false associations to $Q$.
 To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
@@
 
-Of the 1899 variants from the full model that are present in both Resequenced and original 1kGP data, 933 of the variants are present in both samples at matching allele frequencies.
-These variants are likely unaffected by $Q$ in these particular individuals.
-We find that 772 variants are at higher frequency in the resequenced data.
-These variants may be associated to $Q$ due to false negatives, whereby individuals with lower $Q$ tend to have lower calling rates for these variants.
-Only 194 variants are at higher frequencies in the original 1kGP data.
-These variants may be associated to $Q$ due to false positives, whereby individuals with lower $Q$ have higher calling rates for these variants (See Figure \ref{90HanSFS_full}).
+We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
+Of the 14,322 suspicious variants identified globally, only 8,697 of them are found in the 83 individuals in the original 1kGP data.
+From this subset, 8,085 of these variants are missing entirely from the resequenced data.
+We noticed a unusually large number of INDELs that are fixed in the resequenced data that are at lower frequencies in the 1kGP. (See Figure \ref{90HanSFS_full})
 
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{./Figures/Han83.jpg}
+\includegraphics[width=10cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
 This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
-Only 9 of the 297 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
+Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
 Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
@@
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot shows significant SNPs identified in the full model including all populations.
-Only 1899 of the variants are present in the 83 individuals included in both the original 1kGP and the high depth resequenced individuals.}
+This plot shows  8697 significant SNPs and INDELs identified in the \textit{5PC} model including all populations.
+Only 612 of the variants are present in the high depth resequenced individuals.}
 \label{90HanSFS_full}
@@
 
-%\begin{figure}[h]
-%\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
-%\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap.jpg}
+\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
 \sgcomment{Change title, check targeted  sequencing region.
 Specify proportion of SNPs not included} The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
-There are 70 such variants in HapMap and 186 in the GnomAD.}
-%\label{HapMap_GnomAD}
-%\end{figure}
+There are 17 such variants in HapMap and 186 in the GnomAD.}
+\label{HapMap_GnomAD}
+\end{figure}
 

### 2019-02-26 Misc/DoStatsLocal.R

@@ -24,15 +24,10 @@ for(chunk in chunks){
 	out = apply(gt, 1, function(x) 
 				anova(
 					glm2(x ~
-							samples$Pop + 
-							samples$PC1 + 
-							samples$PC2 +
-							samples$PC3 +
-							samples$PC4 +
-							samples$PC5 + 
+							samples$Pop +
 							samples$average_quality_of_mapped_bases,
 						family=binomial),
-					test="Chi")[8,2])
+					test="Chi")[3,2])
 					
 	positions$dev.p = -log10(pchisq(out, 1, lower.tail=F))
 

### 2019-02-26 PlottingScripts/Clean_RegressionPlots.R

@@ -71,10 +71,10 @@ TSBH <-function(df){
 procedures <- c( "TSBH") 
 #0.01
 df$p <- 10^-(df$dev.p)
-adjusted <- mt.rawp2adjp(df$p, procedures, alpha = 0.01)
+adjusted <- mt.rawp2adjp(df$p, procedures, alpha = 0.001)
 adj <- as.data.frame(adjusted$adj[order(adjusted$index), ])
-df$log10P_0.01 <- -log10(adj$TSBH_0.01)
-df$P_0.01 <- adj$TSBH_0.01
+df$log10P_0.001 <- -log10(adj$TSBH_0.001)
+df$P_0.001 <- adj$TSBH_0.001
 
 return(df)
 }
@@ -94,9 +94,9 @@ write.table(NORepeat_SNPs,file='/Users/luke/Documents/QualityPaper/Misc/NORepeat
 
 #write significant positions
 writeSig <- function(df, Name){
-write.table(df[which(df$log10P_0.01>-log10(0.01)),]$rsID, file=paste('~/Documents/QualityPaper/sig/',Name,'_rsID.txt',sep=''), quote=F, col.names=F, row.names=F)
+write.table(df[which(df$log10P_0.001>-log10(0.001)),]$rsID, file=paste('~/Documents/QualityPaper/sig/',Name,'_rsID.txt',sep=''), quote=F, col.names=F, row.names=F)
 
-write.table(df[which(df$log10P_0.01>-log10(0.01)),c('CHROM','POS')], file=paste('~/Documents/QualityPaper/sig/',Name,'_POS.txt',sep=''), quote=F, col.names=F, row.names=F)
+write.table(df[which(df$log10P_0.001>-log10(0.001)),c('CHROM','POS')], file=paste('~/Documents/QualityPaper/sig/',Name,'_POS.txt',sep=''), quote=F, col.names=F, row.names=F)
 }
 
 writeSig(Repeat_Indels, 'Repeat_Indels')
@@ -104,21 +104,26 @@ writeSig(NORepeat_Indels, 'NORepeat_Indels')
 writeSig(Repeat_SNPs, 'Repeat_SNPs')
 writeSig(NORepeat_SNPs, 'NORepeat_SNPs')
 
+#combine them back together and write output file
+Regression <- bind_rows(list(Repeat_Indels, NORepeat_Indels, Repeat_SNPs, NORepeat_SNPs))
[…]

### 2019-02-26 PlottingScripts/CompareModels.R

@@ -6,13 +6,13 @@ library(cowplot)
 
 #read Genotypes
 GT = fread('gzcat ~/Desktop/CHR22.Genotypes.txt.gz',nrows=10000)
-GT = fread('~/Desktop/SigVar_1kGP_clean_GT.txt')
-SigPos = fread('~/Desktop/Total_Sig_POSfreq.txt', col.names=c('CHROM','POS','rsID','ID'))
+GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.genotypes.txt')
+SigPos = fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt', col.names=c('CHROM','POS','rsID','ID'))
 
-ColNames = read.table('~/Desktop/ColNames.txt', header=F)
+ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
 names(GT) = as.character(unlist(ColNames))
-samples = fread('~/Desktop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt') 
-#/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt
+samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt') 
+
 
 Pop = apply(GT[,-(1:3)], 1, function(x)
 						glm2(x ~
@@ -27,50 +27,34 @@ PC = apply(GT[,-(1:3)], 1, function(x)
 							samples$PC4 +
 							samples$PC5 + 
 							samples$average_quality_of_mapped_bases))
-							
-PopPC = apply(GT[,-(1:3)], 1, function(x)
-						glm2(x ~
-							samples$Pop +
-							samples$PC1 + 
-							samples$PC2 +
-							samples$PC3 +
-							samples$PC4 +
-							samples$PC5 + 
-							samples$average_quality_of_mapped_bases))														
+																					
 #get coef of Qual per site
 coefPop = lapply(Pop, function(x) coef(x)[[27]])
 coefPC = lapply(PC, function(x) coef(x)[[7]])
-coefPopPC = lapply(PopPC, function(x) coef(x)[[32]])
 
 coef = data.frame('Pop'=unlist(coefPop),
[…]

### 2019-02-26 PlottingScripts/Han_1kGP_SFS_clean.R

@@ -3,8 +3,13 @@ library(data.table)
 library(ggplot2)
 library(reshape2)
 
+#bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariants_chr_CHROM_POS.txt /Users/luke/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_sorted.vcf.gz  -Oz -o /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001.vcf.gz
+
+#java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar extractFields /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001.vcf.gz CHROM POS AF "GEN[*].GT" | sed "s/0\/0/0/g ; s/0\/[1-9]/1/g ; s/[1-9]\/0/1/g ; s/[1-9]\/[1-9]/1/g ; /\//d" | tail -n +2 > /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001_genotype.txt
+
+
 #read Genotypes
-han1=fread('~/Documents/QualityPaper/sig/Han90_SigInHan_GenomeWide_gt.txt')
+han1=fread('/Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001_genotype.txt')
 HanNames=fread('~/genomes/genomes/Han90/SignificantSNPs_Han90_COLNAMES.txt',header=F)
 names(han1) = as.character(unlist(HanNames))
 #melt to long format
@@ -13,8 +18,14 @@ han = melt(han1, id=c('#CHROM','POS','ID'))
 han$CHROM = as.numeric(as.character(substring(han$'#CHROM',4)))
 han$'#CHROM' <- NULL
 
+#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
+han <- han %>% group_by(CHROM,POS,variable) %>% summarize(han.gt = max(value))
+#to get allele frequencies we group by POS and summarize GT
+hanAF <- han %>% group_by(CHROM,POS) %>% summarize(hanAF = sum(han.gt))
+
+
 #read Genotypes
-kgp1 = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_1kGP_clean_GT.txt')
+kgp1 = fread('~/Documents/QualityPaper/sig/SigVar_0.001.genotypes.txt')
[…]

### 2019-02-26 PlottingScripts/HapMap+GnomAD (Autosaved).R

@@ -1,0 +1,40 @@
+library(data.table)
+library(ggplot2)
+library(stringr)
+library(cowplot)
+
+#GP<-fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
+GP<-fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.frq',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
+GP$CHROM<-as.numeric(as.character(GP$CHROM))
+GP$POS<-as.numeric(as.character(GP$POS))
+
+
+
+HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_sig_rsID.frq', sep='\t',fill=T, col.names=c('CHROM','POS','rsID','AF', 'AN'))
+HM<-HM[which(complete.cases(HM)==T),]
+
+
+plt1<-merge(GP, HM, by=c('CHROM','POS','rsID'))
+
+ggplot(plt1, aes(x=AF.y, y=AF.x))+geom_point(shape=1)+geom_point(data=plt1[which(plt1$rsID=='rs6057648'),], color='blue')+geom_point(data=plt1[which(plt1$rsID=='rs301'),], color='green')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()+xlim(c(0,1))+ylim(c(0,1))
+
+ggsave('/Users/luke/Documents/QualityPaper/Figures/Hap.jpg',height=7,width=7)
+
+
+GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
+GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
+GN$AC_raw<-as.numeric(as.character(GN$AC_raw))
+ggplot(GN, aes(x=reorder(POS,CHROM),y=AF_raw))+geom_point(shape=1)+theme_classic()+theme(axis.text.x=element_blank())
+
+GNm<-melt(GN[,-c('AF_raw')], id=c('CHROM','POS','ID'))
+GNm$value<-as.numeric(as.character(GNm$value))
+GNm<-GNm[which(GNm$value!=0),]
+
+plt<-merge(GP, GN, by=c('CHROM','POS'))
+
[…]

### 2019-02-26 rick/1KGP.Rmd
update

---
title: "HGP analysis"
author: 
- name: Rick Farouni
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

# Preparations

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
options(scipen=999)
```

```{r}
library(Matrix)
library(irlba)
library(threejs)
library(tidyverse)
library(lfa)
```

# Download data

Download a fast parser program (after compilation you'll have a program called a.out)

```{bash}
wget https://raw.githubusercontent.com/bwlewis/1000_genomes_examples/master/parse.c
cc -O2 parse.c
```

Download 1000 genomes phenotype data file

```{bash}
wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped
```

Download 1000 genomes genotype data file for chr 20

```{bash}
wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz

```

Read just the header of the chromosome file to obtain the sample identifiers

```{r}
ids <- readLines(pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
                      sed -n /^#CHROM/p | 
                      tr '\t' '\n' | 
                      tail -n +10"))
head(ids)
```

Write ids to file

```{r}
writeLines(ids, "ids.txt")
```

Download bas files for all the samples

```{bash message=FALSE}
for id in `cat ids.txt`
  do
  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.chrom20.*.bam.bas
done
wait
```

# Read bas files

```{r}
read_basfiles <- function(bas_filepath) {

  bas_file <- read_tsv(bas_filepath,
                  col_types = cols_only('sample' = 'c', 
                                        'platform' = 'c',        
                                       'average_quality_of_mapped_bases' = col_guess())) 
  return( bas_file)
}
```

```{r}
bas_files <- list.files(path = ".",
                        pattern="*bam.bas")

bas_dt <- 
  map_dfr(bas_files,read_basfiles ) %>%
  rename(avg_qual="average_quality_of_mapped_bases")
bas_dt
```

```{r}
bas_dt <-
  bas_dt %>%
  group_by(sample) %>%
  mutate(id=1:n()) %>%
  select(-platform)%>%
  spread(id,avg_qual)
bas_dt
```
```{r}
n_not_na <- function(x) {
 sum(!is.na(x))
}
```

```{r}
 bas_dt <- 
  bas_dt %>%
  ungroup()%>% 
    mutate( avg_qual_sd =pmap_dbl(select(., -sample), lift_vd(sd,na.rm =TRUE )),
           avg_qual_mean = rowMeans(select(., -sample),na.rm =TRUE),
           avg_qual_n= pmap_dbl(select(., -sample), lift_vd(n_not_na)),
           avg_qual_min= pmap_dbl(select(., -sample), lift_vd(min, na.rm =TRUE)),
           avg_qual_max= pmap_dbl(select(., -sample), lift_vd(max, na.rm =TRUE))) %>%
  select (sample, avg_qual_n, avg_qual_min,avg_qual_max,avg_qual_mean,avg_qual_sd, everything() ) 
```

```{r}
ped <- 
  read_tsv(url("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped"))%>% 
  rename(sample="Individual ID")
ped 
```

```{r}
bas_dt <- inner_join( ped  %>% 
                       select(sample, Population),
                       bas_dt,
                     by="sample")
bas_dt
```

```{r}
write_tsv(bas_dt, "average_quality_chrom21.txt")
```

## Plot

```{r fig.width=20}
ggplot(bas_dt,
       aes(x=avg_qual_min,
           y=avg_qual_max, 
           colour=avg_qual_n )) +
  geom_point(size=2)

```

```{r fig.width=20}
ggplot(bas_dt,
       aes(x=avg_qual_mean,
           y=avg_qual_sd, 
           colour=avg_qual_n )) +
  geom_point(size=2)

```

# Load Gentype data

Load VCF file into a sparse matrix

```{r}
p <- pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
          sed /^#/d  |
          cut  -f '10-' |
          ./a.out | 
          cut -f '1-2'")
x <- read.table(p,
                colClasses=c("integer","integer"), 
                fill=TRUE,
                row.names=NULL)
```

```{r}
# Convert to a sparse matrix of people (rows) x variant (columns)
chr20 <- sparseMatrix(i=x[,2], j=x[,1], x=1.0)

# Inspect the dimensions of this matrix
print(dim(chr20))
```

The matrix has 2,504 rows (people) and 1,812,841 columns (variants)

# PCA

compute the first three principal component vectors

```{r}
cm <- colMeans(chr20)
p <- irlba(chr20,
           nv=3,
           nu=3,
           tol=0.1,
           center=cm)
```

 add to dataset
```{r}
colnames(p$u) <- paste0("pc",1:3)
bas_dt <- bind_cols(bas_dt, 
                    as_tibble(p$u))
```

```{r}
ggplot(bas_dt,
       aes(pc1, pc3, color=Population)) +
  geom_point() +
  theme_bw() 
```

# Logistic factor analysis

```{r}
chr20_mtx <- t(as.matrix(chr20[,1:50000]))
```

compute the first three factors including the intercept

```{r}
LF <- lfa(chr20_mtx, 4)
dim(LF)
```

```{r}

bas_dt <- bind_cols(bas_dt,
                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
```

```{r}
ggplot(bas_dt,
       aes(lf1, lf2, color=Population)) +
  geom_point() +
  theme_bw()
```

computes the individual-specific allele frequencies

```{r}
subset <- af(chr20_mtx, LF)

subset[1:5,1]
```

```{r}
sessionInfo()
```

### 2019-02-27 rick/1kGP_lukeMod.Rmd
Added a modified version of the R script from Rick

---
title: "Individual Specific Allele Frequency"
author: "Rick + Luke"
date: "2/27/2019"
output: html_document
---

# Preparations

```{r}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
options(scipen=999)
```
```{bash}
wget https://raw.githubusercontent.com/bwlewis/1000_genomes_examples/master/parse.c
cc -O2 parse.c
```

```{r}
library(Matrix)
library(irlba)
library(threejs)
library(tidyverse)
library(lfa)

library(glm2)
```

Read just the header of the chromosome file to obtain the sample identifiers

```{r}
GenoPath='/Users/luke/genomes/genomes/hg19/phase3/'

ids <- readLines(pipe(paste("gzcat ",GenoPath,"ALL.chr22.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
                      sed -n /^#CHROM/p | 
                      tr '\t' '\n' | 
                      tail -n +10",sep='')))
head(ids)
```

Write ids to file

```{r}
Dir = '/Users/luke/Documents/QualityPaper/rick/'
setwd(Dir)
writeLines(ids, "ids.txt")
```

# Read bas files

```{r}
read_basfiles <- function(bas_filepath) {

  bas_file <- read_tsv(bas_filepath,
                  col_types = cols_only('sample' = 'c',        
                                       'average_quality_of_mapped_bases' = col_guess())) 
  return( bas_file)
}
```

```{r}
bas_files <- list.files(path = "/Users/luke/genomes/bas_file/1000GenomesBAS/",
                        pattern="*bam.bas", full.names = TRUE)

bas_dt <- 
  map_dfr(bas_files,read_basfiles ) %>%
  rename(avg_qual="average_quality_of_mapped_bases") %>%
  group_by(sample) %>% summarize(avg_qual = mean(avg_qual))

#keep samples in 1kGP VCF files
bas_dt <- bas_dt[which((bas_dt$sample %in% ids)==T),]

bas_dt
```
Get Sample Info, 
```{r}
ped <- 
  read_tsv(url("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped"))%>% 
  rename(sample="Individual ID")

bas_dt <- inner_join( ped  %>% 
                       select(sample, Population),
                       bas_dt,
                     by="sample")
bas_dt
```

# Load Gentype data

Load VCF file into a sparse matrix

```{r}
p <- pipe(paste("gzcat ",GenoPath,"ALL.chr22.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
          sed /^#/d  |
          cut  -f '10-' |
          ./a.out | 
          cut -f '1-2'",sep=''))
x <- read.table(p,
                colClasses=c("integer","integer"), 
                fill=TRUE,
                row.names=NULL)
```

```{r}
# Convert to a sparse matrix of people (rows) x variant (columns)
chr22 <- sparseMatrix(i=x[,2], j=x[,1], x=1.0)

# Inspect the dimensions of this matrix
print(dim(chr22))
```

# PCA

compute the first three principal component vectors

```{r}
cm <- colMeans(chr22)
p <- irlba(chr22,
           nv=3,
           nu=3,
           tol=0.1,
           center=cm)
```

 add to dataset
```{r}
colnames(p$u) <- paste0("pc",1:3)
bas_dt <- bind_cols(bas_dt, 
                    as_tibble(p$u))
```

```{r}
ggplot(bas_dt,
       aes(pc1, pc3, color=Population)) +
  geom_point() +
  theme_bw() 
```

# Logistic factor analysis

```{r}
chr22_mtx <- t(as.matrix(chr22[,1:50000]))
```

compute the first three factors including the intercept

```{r}
LF <- lfa(chr22_mtx, 4)
dim(LF)
```

```{r}

bas_dt <- bind_cols(bas_dt,
                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
```

computes the individual-specific allele frequencies

```{r}
subset <- af(chr22_mtx, LF)

subset[1:5,1]
```

```{r}
ggplot(bas_dt,
       aes(lf1, lf2, color=Population)) +
  geom_point() +
  theme_bw()
```

Logistic Regression
```{r}
LF = mapply(function(x, y){
                           glm(x 
                                ~ bas_dt$avg_qual 
                                + offset(y))}, 
                           as.data.frame(t(chr22_mtx)),
                           as.data.frame(t(subset)))

```

```{r}
Pop = apply(head(chr22_mtx,1000), 1, function(x)
                    glm2(x 
                          ~ bas_dt$Population
                          + bas_dt$avg_qual))
```

### 2019-02-27 PlottingScripts/HapMap+GnomAD.R

@@ -3,7 +3,8 @@ library(ggplot2)
 library(stringr)
 library(cowplot)
 
-GP<-fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
+#GP<-fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
+GP<-fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.frq',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
 GP$CHROM<-as.numeric(as.character(GP$CHROM))
 GP$POS<-as.numeric(as.character(GP$POS))
 
@@ -15,7 +16,10 @@ HM<-HM[which(complete.cases(HM)==T),]
 
 plt1<-merge(GP, HM, by=c('CHROM','POS','rsID'))
 
-A=ggplot(plt1, aes(x=AF.x, y=AF.y))+geom_point(shape=1)+geom_point(data=plt1[which(plt1$rsID=='rs6057648'),], color='blue')+geom_point(data=plt1[which(plt1$rsID=='rs301'),], color='green')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()+xlim(c(0,1))+ylim(c(0,1))
+ggplot(plt1, aes(x=AF.y, y=AF.x))+geom_point(shape=1)+geom_point(data=plt1[which(plt1$rsID=='rs6057648'),], color='blue')+geom_point(data=plt1[which(plt1$rsID=='rs301'),], color='green')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()+xlim(c(0,1))+ylim(c(0,1))
+
+ggsave('/Users/luke/Documents/QualityPaper/Figures/Hap.jpg',height=7,width=7)
+
 
 GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
 GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
@@ -31,6 +35,6 @@ plt<-merge(GP, GN, by=c('CHROM','POS'))
[…]

### 2019-02-27 rick/1KGP.Rmd
update

@@
 library(lfa)
+library(furrr)
+options(future.globals.maxSize= 10000000000)
+plan(multicore)
 ```
@@
 
+```{r}
+rm(x)
+gc()
+```
+
 The matrix has 2,504 rows (people) and 1,812,841 columns (variants)
+
+### work with a subset
+
+```{r}
+chr20_sub <- chr20[,1:100000]
+```
 
@@
 ```{r}
-cm <- colMeans(chr20)
-p <- irlba(chr20,
+cm <- colMeans(chr20_sub)
+p <- irlba(chr20_sub,
            nv=3,
@@
 ggplot(bas_dt,
-       aes(pc1, pc3, color=Population)) +
+       aes(pc1, pc2, color=Population)) +
   geom_point() +
@@
 ```{r}
-chr20_mtx <- t(as.matrix(chr20[,1:50000]))
+chr20_mtx <- t(as.matrix(chr20_sub))
 ```
@@
 dim(LF)
+```
+
+Determine genome-wide goodness-of-fit for a particular value of d by computing SNP-by-SNP goodness-of-fit when compared to population structure.
+
+```{r}
+#gof_4 <- model.gof(chr20_mtx, LF, 3)
+#LF <- lfa(chr20_mtx, 10)
+#gof_10 <- model.gof(chr20_mtx, LF, 3)
+#hist(gof_4)
 ```
@@
 ```{r}
-subset <- af(chr20_mtx, LF)
+chr20sub_ial <- af(chr20_mtx, LF)
 
-subset[1:5,1]
+```
+
+```{r}
+
+to_datalist <- function(i, data_genotype, data_iaf, qual){
+
+    data <- tibble(genotype=data_genotype[i,],
+               iaf=data_iaf[i,],
+               qual=qual)
+
+     return(data)       
+}
+
+fit_model_1 <- function(data){
+
+  fit <- glm(genotype ~ qual + offset(iaf), family ="binomial", data=data)
+  return(fit)
+
+}
+
+plot_data <- function(data){
+
+      p <- ggplot(data,
+       aes(qual, genotype, color=iaf)) +
+      geom_point() +
+      theme_bw()
+  return(p)
+
+}
+```
+
+```{r}
+
+loci <- c(1,4,5)
+names(loci) <- loci
+
+data_list <- map(loci, 
+                 to_datalist,
+                 data_genotype=chr20_mtx,
+                 data_iaf=chr20sub_ial,
+                 qual=bas_dt$avg_qual_mean)
+
+```
+
+```{r}
+fits <- future_map(data_list, fit_model_1)
+
+```
+
+```{r}
+plots <-  future_map(data_list, plot_data) 
 ```

### 2019-02-27 rick/1kGP_lukeMod.Rmd

@@
 ```{r}
-Pop = apply(head(chr22_mtx,1000), 1, function(x)
+t= list()
+for(f in seq(1,10000)){
+  t1 = chr22_mtx[f,]
+  t2 = subset[f,]
+  t[[f]] = glm(t1 ~ bas_dt$avg_qual + offset(t2))
+}
+
+```
+
+```{r}
+Pop = apply(head(chr22_mtx,10000), 1, function(x)
                     glm2(x 

### 2019-02-28 rick/1kGP_lukeMod.Rmd
modified ricks code a bit

@@
 
-```{r}
-knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
-options(scipen=999)
-```
+#```{r}
+#knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
+#options(scipen=999)
+#```
 ```{bash}
@@
 library(lfa)
-
+library(furrr)plots
 library(glm2)
@@
 
-Logistic Regression
+computes the individual-specific allele frequencies
+
 ```{r}
-LF = mapply(function(x, y){
-                           glm(x 
-                                ~ bas_dt$avg_qual 
-                                + offset(y))}, 
-                           as.data.frame(t(chr22_mtx)),
-                           as.data.frame(t(subset)))
+chr22sub_ial <- af(chr22_mtx, LF)
 
+```
+
+```{r}
+
+to_datalist <- function(i, data_genotype, data_iaf, qual){
+
+    data <- tibble(genotype=data_genotype[i,],
+               iaf=data_iaf[i,],
+               qual=qual)
+
+     return(data)       
+}
+
+fit_model_1 <- function(data){
+
+  fit <- glm2(genotype ~ qual + offset(iaf), family ="binomial", data=data)
+  return(fit)
+
+}
+
+plot_data <- function(data){
+
+      p <- ggplot(data,
+       aes(qual, genotype, color=iaf)) +
+      geom_point() +
+      theme_bw()
+  return(p)
+
+}
+```
+
+```{r}
+
+loci <- seq(1,50000)
+names(loci) <- loci
+
+data_list <- map(loci, 
+                 to_datalist,
+                 data_genotype=chr22_mtx,
+                 data_iaf=chr22sub_ial,
+                 qual=bas_dt$avg_qual)
+
+```
+
+```{r}
+options(future.globals.maxSize= 6000000000)
+
+fits <- future_map(data_list, fit_model_1)
+
+```
+
+```{r}
+coefLF = lapply(fits, function(x) coef(x)[[2]])
+
+```
+
+```{r}
+plots <-  future_map(data_list, plot_data) 
 ```

### 2019-03-01 rick/1KGP.Rmd
update

@@
 library(furrr)
+library(broom)
 options(future.globals.maxSize= 10000000000)
@@
 
+```{r}
+bas_dt<- read_tsv( "average_quality_chrom21.txt")
+```
+
 ## Plot
@@
 # Convert to a sparse matrix of people (rows) x variant (columns)
-chr20 <- sparseMatrix(i=x[,2], j=x[,1], x=1.0)
+chr20 <- sparseMatrix(i=x[,2], j=x[,1], x=1L)
 
@@
 ```{r}
-chr20_mtx <- t(as.matrix(chr20_sub))
+chr20_mtx <- t(as.matrix(chr20[,1:100000]))
 ```
@@
 ggplot(bas_dt,
-       aes(lf1, lf2, color=Population)) +
+       aes(lf2, lf3, color=Population)) +
   geom_point() +
@@
 
-to_datalist <- function(i, data_genotype, data_iaf, qual){
+to_datalist <- function(i, data_genotype, data_iaf){
 
     data <- tibble(genotype=data_genotype[i,],
-               iaf=data_iaf[i,],
-               qual=qual)
+               iaf=data_iaf[i,])
 
@@
 
-fit_model_1 <- function(data){
+fit_model_1 <- function(data, qual){
 
@@
 
-plot_data <- function(data){
+plot_data <- function(data,qual){
 
@@
                  data_genotype=chr20_mtx,
-                 data_iaf=chr20sub_ial,
-                 qual=bas_dt$avg_qual_mean)
+                 data_iaf=chr20sub_ial)
 
@@
 ```{r}
-fits <- future_map(data_list, fit_model_1)
-
+data_list <- enframe(data_list,
+                     name = "locus", 
+                     value = "dataset")
 ```
@@
 ```{r}
-plots <-  future_map(data_list, plot_data) 
+avg_qual_mean <- bas_dt$avg_qual_mean
+data_list <- 
+  data_list %>%
+  mutate(
+    fit = future_map(dataset,~ glm(genotype ~ avg_qual_mean + offset(iaf), family ="binomial", data=.x)),
+    tidied = future_map(fit, tidy),
+    glanced = future_map(fit, glance)
+)
+```
+
+get all p-values
+```{r}
+p_values_dt <-
+  data_list  %>% 
+  select(tidied) %>% 
+  map_depth( 2, "p.value") %>% 
+  map_depth( 2, c(2)) %>% # c(1) for the intercept
+  unlist(use.names = FALSE) %>% 
+  enframe(name = "locus" ,
+          value= "p.value")%>%
+  mutate(locus= data_list$locus)
+
+p_values_dt
+```
+get a p-value for target specific loci
+
+```{r}
+target_loci <- c("4", "5")
+```
+
+```{r}
+p_values_subset <-
+  data_list  %>% 
+  filter(locus %in% target_loci) %>% 
+  select(tidied) %>% 
+  map_depth( 2, "p.value") %>% 
+   map_depth( 2, c(2)) %>%
+  unlist(use.names = FALSE) %>% 
+  enframe(name = "locus" ,
+          value= "p.value")%>%
+  mutate(locus= target_loci)
+p_values_subset 
 ```

### 2019-03-01 rick/1KGP.Rmd
update

@@
 
-loci <- c(1,4,5)
+loci <- c(1:1000)
 names(loci) <- loci
 
-data_list <- map(loci, 
+data_list <- future_map(loci, 
                  to_datalist,
@@
 ```{r}
-data_list <- enframe(data_list,
+data_list <- 
+  enframe(data_list,
           name = "locus", 
@@
 avg_qual_mean <- bas_dt$avg_qual_mean
+
 data_list <- 
@@
   mutate(
-    fit = future_map(dataset,~ glm(genotype ~ avg_qual_mean + offset(iaf), family ="binomial", data=.x)),
+    fit = future_map(dataset,
+                     ~ glm(genotype ~ avg_qual_mean + offset(iaf),
+                           family ="binomial",
+                           data=.x)),
     tidied = future_map(fit, tidy),
@@
 get all p-values
+
 ```{r}
@@
 ```
+
+```{r}
+estimates_dt <-
+  data_list  %>% 
+  select(tidied) %>% 
+  map_depth( 2, "estimate")%>% 
+  map_depth( 2, c(2)) %>% # c(1) for the intercept
+  unlist(use.names = FALSE) %>% 
+  enframe(name = "locus" ,
+          value= "beta_hat")%>%
+  mutate(locus= data_list$locus)
+
+estimates_dt
+```
+```{r}
+ggplot(p_values_dt)+
+  geom_histogram((aes(x=p.value)), bins=100) +
+  theme_bw()
+```
+
 get a p-value for target specific loci

### 2019-03-02 rick/1KGP.Rmd
update

@@
 
-```{r}
-bas_dt<- read_tsv( "average_quality_chrom21.txt")
-```
-
-## Plot
+ Plot
 
@@
 
-# Load Gentype data
-
-Load VCF file into a sparse matrix
+# Logistic factor analysis
 
 ```{r}
+bas_dt <- read_tsv( "average_quality_chrom21.txt")
+avg_qual_mean <- bas_dt$avg_qual_mean
+```
+
+```{r}
+
+to_datalist <- function(i, data_genotype, data_iaf){
+
+    data <- tibble(genotype=data_genotype[i,],
+               iaf=data_iaf[i,])
+
+     return(data)       
+}
+
+get_data <- function(loci){
+
   p <- pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
@@
                 row.names=NULL)
+
+  data <- sparseMatrix(i=x[,2], j=x[,1], x=1L)
+
+  data <- t(as.matrix(data[,loci]))
+
+  return(data)
+
+}
+
+get_af_data <- function(data, d=4){
+
+  # compute the first d factors including the intercept
+  LF <- lfa(data, d)
+  # compute the individual-specific allele frequencies
+  data_iaf <- af(data, LF)
+  loci <- c(1:dim(data)[1])
+  names(loci) <- loci
+
+  data_list <- future_map(loci, 
+                 to_datalist,
+                 data_genotype=data,
+                 data_iaf=data_iaf)
+
+  data_list <- 
+  enframe(data_list,
+          name = "locus", 
+          value = "dataset")
+
+  return(data_list)
+
+}
+
+get_predicted_mean <- function(model_fit) {
+ predicted <- 
+   augment(model_fit, type.predict = "response") %>%
+   select("genotype","offset.iaf.",".fitted") %>%
+   rename(iaf="offset.iaf.", mu=".fitted" )
+  return(predicted)
+}
+
+get_stats <- function(model_fit){
+  stats_dt <- 
+    tidy(model_fit)  %>%
+    filter(term=="avg_qual_mean") %>%
+    select(estimate, p.value) %>%
+    mutate(deviance= anova(model_fit)["Deviance"][[1]][2])
+  return(stats_dt)
+}
+
+fit_model <- function(dataset){
+
+   fit <- glm(genotype ~ avg_qual_mean + offset(iaf),
+                           family ="binomial",
+                           data=dataset)
+
+   stats <- get_stats(fit)
+   predicted <- get_predicted_mean(fit)
+
+   return(list(stats, predicted))
+}
+
+get_plot <- function(data){
+
+ p <- ggplot(bind_cols(data, 
+                   bas_dt %>% 
+                     select(Population,
+                            avg_qual_mean))) +
+    geom_point(aes(avg_qual_mean, 
+                   genotype),
+               size=.5,
+               alpha= 0.4) +
+    geom_line(aes(avg_qual_mean,
+                  iaf, 
+                  colour="af"),
+              linetype="dashed") +
+    geom_line(aes(avg_qual_mean,
+                  mu,
+                  colour="mu")) +
+      theme_bw() +
+    facet_wrap(~Population)
+
+  return(p)
+}
 ```
@@
 ```{r}
-# Convert to a sparse matrix of people (rows) x variant (columns)
-chr20 <- sparseMatrix(i=x[,2], j=x[,1], x=1L)
+run_workflow <- function(loci){
 
-# Inspect the dimensions of this matrix
-print(dim(chr20))
+  data <- get_data(loci)
+  data_list <- get_af_data(data)
+
+  rm(data)
+  gc()
+
+  data_list <- 
+    data_list %>%
+  #slice(1:10) %>%
+  transmute(
+    locus=locus,
+    fit1 = future_map(dataset, fit_model))  %>%
+  transmute(
+    locus=locus,
+    stats = future_map(fit1, c(1)),
+    predicted =future_map(fit1, c(2))) %>%
+    unnest(stats) %>%
+    mutate(plotted= future_map(predicted, get_plot)) %>%
+    arrange(-deviance)
+
+  return(data_list)
+
+}
+
 ```
@@
 ```{r}
-rm(x)
-gc()
-```
-
-The matrix has 2,504 rows (people) and 1,812,841 columns (variants)
-
-### work with a subset
-
-```{r}
-chr20_sub <- chr20[,1:100000]
-```
-
-# PCA
-
-compute the first three principal component vectors
-
-```{r}
-cm <- colMeans(chr20_sub)
-p <- irlba(chr20_sub,
-           nv=3,
-           nu=3,
-           tol=0.1,
-           center=cm)
-```
-
- add to dataset
-```{r}
-colnames(p$u) <- paste0("pc",1:3)
-bas_dt <- bind_cols(bas_dt, 
-                    as_tibble(p$u))
+data_list <- run_workflow (loci=c(1:100))
+data_list
 ```
@@
 ```{r}
-ggplot(bas_dt,
-       aes(pc1, pc2, color=Population)) +
-  geom_point() +
-  theme_bw() 
+n <-5
+target_loci <- 
+  data_list %>% 
+  slice(1:n) %>% 
+  pull(locus)
+target_loci 
 ```
 
-# Logistic factor analysis
+```{r}
+data_list_subset %>%
+  filter(locus %in% target_loci )%>%
+  pull(plotted)
+```
 
 ```{r}
-chr20_mtx <- t(as.matrix(chr20[,1:100000]))
+sessionInfo()
 ```
 
-compute the first three factors including the intercept
+## Draft
 
-```{r}
-LF <- lfa(chr20_mtx, 4)
-dim(LF)
-```
+The matrix has 2,504 rows (people) and 1,812,841 columns (variants)
 
@@
 
-bas_dt <- bind_cols(bas_dt,
-                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
+data_list
 ```
-
-```{r}
-ggplot(bas_dt,
-       aes(lf2, lf3, color=Population)) +
-  geom_point() +
-  theme_bw()
-```
-
-computes the individual-specific allele frequencies
-
-```{r}
-chr20sub_ial <- af(chr20_mtx, LF)
-
-```
-
-```{r}
-
-to_datalist <- function(i, data_genotype, data_iaf){
-
-    data <- tibble(genotype=data_genotype[i,],
-               iaf=data_iaf[i,])
-
-     return(data)       
-}
-
-fit_model_1 <- function(data, qual){
-
-  fit <- glm(genotype ~ qual + offset(iaf), family ="binomial", data=data)
-  return(fit)
-
-}
-
-plot_data <- function(data,qual){
-
-      p <- ggplot(data,
-       aes(qual, genotype, color=iaf)) +
-      geom_point() +
-      theme_bw()
-  return(p)
-
-}
-```
-
-```{r}
-
-loci <- c(1:1000)
-names(loci) <- loci
-
-data_list <- future_map(loci, 
-                 to_datalist,
-                 data_genotype=chr20_mtx,
-                 data_iaf=chr20sub_ial)
-
-```
-
-```{r}
-data_list <- 
-  enframe(data_list,
-          name = "locus", 
-          value = "dataset")
-```
-
-```{r}
-avg_qual_mean <- bas_dt$avg_qual_mean
-
-data_list <- 
-  data_list %>%
-  mutate(
-    fit = future_map(dataset,
-                     ~ glm(genotype ~ avg_qual_mean + offset(iaf),
-                           family ="binomial",
-                           data=.x)),
-    tidied = future_map(fit, tidy),
-    glanced = future_map(fit, glance)
-)
-```
-
-get all p-values
 
@@
 ```{r}
-target_loci <- c("4", "5")
+
 ```
@@
 ```{r}
-sessionInfo()
+
+bas_dt <- bind_cols(bas_dt,
+                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
 ```
+
+```{r}
+ggplot(bas_dt,
+       aes(lf2, lf3, color=Population)) +
+  geom_point() +
+  theme_bw()
+```
+
+### work with a subset
+
+```{r}
+chr20_sub <- chr20[,1:50000]
+```
+
+# PCA
+
+compute the first three principal component vectors
+
+```{r}
+cm <- colMeans(chr20_sub)
+p <- irlba(chr20_sub,
+           nv=3,
+           nu=3,
+           tol=0.1,
+           center=cm)
+```
+
+ add to dataset
+```{r}
+colnames(p$u) <- paste0("pc",1:3)
+bas_dt <- bind_cols(bas_dt, 
+                    as_tibble(p$u))
+```
+
+```{r}
+ggplot(bas_dt,
+       aes(pc1, pc2, color=Population)) +
+  geom_point() +
+  theme_bw() 
+```

### 2019-03-02 rick/1KGP.Rmd
update

@@
 
-get_data <- function(loci){
+get_data <- function(){
 
@@
   data <- sparseMatrix(i=x[,2], j=x[,1], x=1L)
-
-  data <- t(as.matrix(data[,loci]))
 
@@
 
-  data <- get_data(loci)
+  data <- get_data()
+  data <- t(as.matrix(data[,loci]))
   data_list <- get_af_data(data)
@@
     data_list %>%
-  #slice(1:10) %>%
+  slice(1:10) %>%
   transmute(
@@
     unnest(stats) %>%
-    mutate(plotted= future_map(predicted, get_plot)) %>%
     arrange(-deviance)
@@
 ```{r}
-data_list <- run_workflow (loci=c(1:100))
+data_list <- run_workflow (loci=c(1:50000))
 data_list
@@
 ```{r}
+data_list_subset <- 
+  data_list %>%
+   filter(locus %in% target_loci ) %>%
+    mutate(plotted= future_map(predicted, get_plot)) 
+```
+
+```{r}
 data_list_subset %>%
-  filter(locus %in% target_loci )%>%
   pull(plotted)

### 2019-03-02 rick/LFA_Analysis.Rmd
Added LFA Analysis

--- 2019-03-02 rick/1KGP.Rmd
@@
 ---
-title: "HGP analysis"
+title: "LFA analysis"
 author: 
-- name: Rick Farouni
+- name: Rick Farouni + Luke
   affiliation:
@@
 options(future.globals.maxSize= 10000000000)
-plan(multicore)
-```
-
-# Download data
-
-Download a fast parser program (after compilation you'll have a program called a.out)
-
-```{bash}
-wget https://raw.githubusercontent.com/bwlewis/1000_genomes_examples/master/parse.c
-cc -O2 parse.c
-```
-
-Download 1000 genomes phenotype data file
-
-```{bash}
-wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped
-```
-
-Download 1000 genomes genotype data file for chr 20
-
-```{bash}
-wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz
-
-```
-
-Read just the header of the chromosome file to obtain the sample identifiers
-
-```{r}
-ids <- readLines(pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
-                      sed -n /^#CHROM/p | 
-                      tr '\t' '\n' | 
-                      tail -n +10"))
-head(ids)
-```
-
-Write ids to file
-
-```{r}
-writeLines(ids, "ids.txt")
-```
-
-Download bas files for all the samples
-
-```{bash message=FALSE}
-for id in `cat ids.txt`
-  do
-  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.chrom20.*.bam.bas
-done
-wait
-```
-
-# Read bas files
-
-```{r}
-read_basfiles <- function(bas_filepath) {
-
-  bas_file <- read_tsv(bas_filepath,
-                  col_types = cols_only('sample' = 'c', 
-                                        'platform' = 'c',        
-                                       'average_quality_of_mapped_bases' = col_guess())) 
-  return( bas_file)
-}
-```
-
-```{r}
-bas_files <- list.files(path = ".",
-                        pattern="*bam.bas")
-
-bas_dt <- 
-  map_dfr(bas_files,read_basfiles ) %>%
-  rename(avg_qual="average_quality_of_mapped_bases")
-bas_dt
-```
-
-```{r}
-bas_dt <-
-  bas_dt %>%
-  group_by(sample) %>%
-  mutate(id=1:n()) %>%
-  select(-platform)%>%
-  spread(id,avg_qual)
-bas_dt
-```
-```{r}
-n_not_na <- function(x) {
- sum(!is.na(x))
-}
-```
-
-```{r}
- bas_dt <- 
-  bas_dt %>%
-  ungroup()%>% 
-    mutate( avg_qual_sd =pmap_dbl(select(., -sample), lift_vd(sd,na.rm =TRUE )),
-           avg_qual_mean = rowMeans(select(., -sample),na.rm =TRUE),
-           avg_qual_n= pmap_dbl(select(., -sample), lift_vd(n_not_na)),
-           avg_qual_min= pmap_dbl(select(., -sample), lift_vd(min, na.rm =TRUE)),
-           avg_qual_max= pmap_dbl(select(., -sample), lift_vd(max, na.rm =TRUE))) %>%
-  select (sample, avg_qual_n, avg_qual_min,avg_qual_max,avg_qual_mean,avg_qual_sd, everything() ) 
-```
-
-```{r}
-ped <- 
-  read_tsv(url("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped"))%>% 
-  rename(sample="Individual ID")
-ped 
-```
-
-```{r}
-bas_dt <- inner_join( ped  %>% 
-                       select(sample, Population),
-                       bas_dt,
-                     by="sample")
-bas_dt
-```
-
-```{r}
-write_tsv(bas_dt, "average_quality_chrom21.txt")
-```
-
- Plot
-
-```{r fig.width=20}
-ggplot(bas_dt,
-       aes(x=avg_qual_min,
-           y=avg_qual_max, 
-           colour=avg_qual_n )) +
-  geom_point(size=2)
-
-```
-
-```{r fig.width=20}
-ggplot(bas_dt,
-       aes(x=avg_qual_mean,
-           y=avg_qual_sd, 
-           colour=avg_qual_n )) +
-  geom_point(size=2)
-
+future::plan(multicore)
 ```
@@
 ```{r}
-bas_dt <- read_tsv( "average_quality_chrom21.txt")
-avg_qual_mean <- bas_dt$avg_qual_mean
-```
-
-```{r}
+GenoPath='/Users/luke/genomes/genomes/Genotype/'
 
@@
 
-  p <- pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
+  p <- pipe(paste("cat ",GenoPath,"1KG_DownSample_noDup_header.chip.omni.vcf |
           sed /^#/d  |
@@
           ./a.out | 
-          cut -f '1-2'")
+          cut -f '1-2'",sep=''))
   x <- read.table(p,
@@
 
-get_predicted_mean <- function(model_fit) {
- predicted <- 
-   augment(model_fit, type.predict = "response") %>%
-   select("genotype","offset.iaf.",".fitted") %>%
-   rename(iaf="offset.iaf.", mu=".fitted" )
-  return(predicted)
-}
-
-get_stats <- function(model_fit){
-  stats_dt <- 
-    tidy(model_fit)  %>%
-    filter(term=="avg_qual_mean") %>%
-    select(estimate, p.value) %>%
-    mutate(deviance= anova(model_fit)["Deviance"][[1]][2])
-  return(stats_dt)
-}
-
-fit_model <- function(dataset){
-
-   fit <- glm(genotype ~ avg_qual_mean + offset(iaf),
-                           family ="binomial",
-                           data=dataset)
-
-   stats <- get_stats(fit)
-   predicted <- get_predicted_mean(fit)
-
-   return(list(stats, predicted))
-}
-
-get_plot <- function(data){
-
- p <- ggplot(bind_cols(data, 
-                   bas_dt %>% 
-                     select(Population,
-                            avg_qual_mean))) +
-    geom_point(aes(avg_qual_mean, 
-                   genotype),
-               size=.5,
-               alpha= 0.4) +
-    geom_line(aes(avg_qual_mean,
-                  iaf, 
-                  colour="af"),
-              linetype="dashed") +
-    geom_line(aes(avg_qual_mean,
-                  mu,
-                  colour="mu")) +
-      theme_bw() +
-    facet_wrap(~Population)
-
-  return(p)
-}
-```
-
-```{r}
-run_workflow <- function(loci){
-
-  data <- get_data()
-  data <- t(as.matrix(data[,loci]))
-  data_list <- get_af_data(data)
-
-  rm(data)
-  gc()
-
-  data_list <- 
-    data_list %>%
-  slice(1:10) %>%
-  transmute(
-    locus=locus,
-    fit1 = future_map(dataset, fit_model))  %>%
-  transmute(
-    locus=locus,
-    stats = future_map(fit1, c(1)),
-    predicted =future_map(fit1, c(2))) %>%
-    unnest(stats) %>%
-    arrange(-deviance)
-
-  return(data_list)
-
-}
-
-```
-
-```{r}
-data_list <- run_workflow (loci=c(1:50000))
-data_list
-```
-
-```{r}
-n <-5
-target_loci <- 
-  data_list %>% 
-  slice(1:n) %>% 
-  pull(locus)
-target_loci 
-```
-
-```{r}
-data_list_subset <- 
-  data_list %>%
-   filter(locus %in% target_loci ) %>%
-    mutate(plotted= future_map(predicted, get_plot)) 
-```
-
-```{r}
-data_list_subset %>%
-  pull(plotted)
-```
-
-```{r}
-sessionInfo()
-```
-
-## Draft
-
-The matrix has 2,504 rows (people) and 1,812,841 columns (variants)
-
-Determine genome-wide goodness-of-fit for a particular value of d by computing SNP-by-SNP goodness-of-fit when compared to population structure.
-
-```{r}
-#gof_4 <- model.gof(chr20_mtx, LF, 3)
-#LF <- lfa(chr20_mtx, 10)
-#gof_10 <- model.gof(chr20_mtx, LF, 3)
-#hist(gof_4)
 ```
@@
 
-data_list
-```
+data <- get_data()
+data <- t(as.matrix(data))
+data_list <- get_af_data(data)
 
-```{r}
-p_values_dt <-
-  data_list  %>% 
-  select(tidied) %>% 
-  map_depth( 2, "p.value") %>% 
-  map_depth( 2, c(2)) %>% # c(1) for the intercept
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "p.value")%>%
-  mutate(locus= data_list$locus)
+rm(data)
 
-p_values_dt
-```
-
-```{r}
-estimates_dt <-
-  data_list  %>% 
-  select(tidied) %>% 
-  map_depth( 2, "estimate")%>% 
-  map_depth( 2, c(2)) %>% # c(1) for the intercept
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "beta_hat")%>%
-  mutate(locus= data_list$locus)
-
-estimates_dt
-```
-```{r}
-ggplot(p_values_dt)+
-  geom_histogram((aes(x=p.value)), bins=100) +
-  theme_bw()
-```
-
-get a p-value for target specific loci
-
-```{r}
+save(data_list, '/Users/luke/Documents/QualityPaper/rick/LFA_GenotypeData.RData')
 
 ```
-
-```{r}
-p_values_subset <-
-  data_list  %>% 
-  filter(locus %in% target_loci) %>% 
-  select(tidied) %>% 
-  map_depth( 2, "p.value") %>% 
-   map_depth( 2, c(2)) %>%
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "p.value")%>%
-  mutate(locus= target_loci)
-p_values_subset 
-```
-
-```{r}
-
-bas_dt <- bind_cols(bas_dt,
-                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
-```
-
-```{r}
-ggplot(bas_dt,
-       aes(lf2, lf3, color=Population)) +
-  geom_point() +
-  theme_bw()
-```
-
-### work with a subset
-
-```{r}
-chr20_sub <- chr20[,1:50000]
-```
-
-# PCA
-
-compute the first three principal component vectors
-
-```{r}
-cm <- colMeans(chr20_sub)
-p <- irlba(chr20_sub,
-           nv=3,
-           nu=3,
-           tol=0.1,
-           center=cm)
-```
-
- add to dataset
-```{r}
-colnames(p$u) <- paste0("pc",1:3)
-bas_dt <- bind_cols(bas_dt, 
-                    as_tibble(p$u))
-```
-
-```{r}
-ggplot(bas_dt,
-       aes(pc1, pc2, color=Population)) +
-  geom_point() +
-  theme_bw() 
-```

### 2019-03-02 rick/LFA_Analysis.Rmd
slight changes

@@
 ---
-
-# Preparations
-
-```{r setup}
-knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
-options(scipen=999)
-```
 
@@
 data <- t(as.matrix(data))
+
+```
+
+```{r}
 data_list <- get_af_data(data)
@@
 save(data_list, '/Users/luke/Documents/QualityPaper/rick/LFA_GenotypeData.RData')
-
 ```

### 2019-03-05 legacy.tex
updated stuff

@@
 
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$. 
-To account for population structure, we also included the top five global principal components.
-This method identifies a total of 14,322 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
+This method identifies a total of 11,416 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 2,819 & 9,846 \\  
-{INDEL} & 378 & 1,279 \\ \hline
+{SNP}  & 2,007 & 8,009 \\  
+{INDEL} & 330 & 1,070 \\ \hline
 \end{tabular}
@@
 SNPs and INDELs were also analyzed separately.
-The total number of 22,186 are statistically significantly associated to $Q$.}
+The total number of 11,416 are statistically significantly associated to $Q$.}
 \label{sigTable}
@@
 
+\subsection{Model selection}
+We considered multiple approaches to account for population structure in order to best capture the variants associated to $Q$.
+Each of these methods is somewhat similar to the genotype-conditional association test described by Song et al. \citep{song2015testing}.
+One approach included the top five global principal components (PCs)as covariates to account for population structure without using a categorical variable.
+Another combined both the population categorical variable as well as all five global PCs.
+
 \subsection{Cell line or technical artifact}
@@
 If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-From the list of 14,322 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
+From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
 Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
@@
 We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 14,322 suspicious variants identified globally, only 8,697 of them are found in the 83 individuals in the original 1kGP data.
+Of the 11,416 suspicious variants identified globally, only 8,697 of them are found in the 83 individuals in the original 1kGP data.
 From this subset, 8,085 of these variants are missing entirely from the resequenced data.
@@
 
-Four of these studies used the 1kGP in their reference panel for imputation \citep{astle2016allelic, lopez2017genome, tian2017genome, luciano2018association}.
-One study used genotype data directly using the Omni2-5 and Omni5 chips \citep{yucesoy2015genome}.
+One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome}.
 Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
 And one study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-The two studies that imputed using HapMap are likely not impacted by this as these variants are likely falsely associated to $Q$.
 
-All eight papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+All four papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for some biological traits.
-
+\todo{UPDATE LOG ODDDS}
 \begin{table}[h]
@@
   {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)} & \multicolumn{1}{p{1cm}}{\centering Odds \\ ratio}     \\ \hline
- 28654678	& PLoS One	& rs201761909	& 5.70	& 100.67 & 0.63\\
- 	& 		& rs201130852	& 5.05	& 94.28 & 0.65\\
- 	& 		& rs200655768	& 6.53	& 85.84 & 0.66\\
- 	& 		& rs201255786	& 5.70	& 80.65 & 0.65\\	
- 	& 		& rs184202621	& 5.52	& 71.13 & 0.68\\
-	& 		& rs80274284	& 6.00	& 69.02 &  0.70\\
- 	& 		& rs200699422	& 5.30	& 7.82 & 0.87\\ 
+ 28654678	& PLoS One	& rs201130852	& 5.05	& 73.18 & 0.65\\
+ 	& 		& rs201761909	& 5.70	& 72.90 & 0.63\\
+ 	& 		& rs200655768	& 6.53	& 69.28 & 0.66\\
+ 	& 		& rs201255786	& 5.70	& 61.20 & 0.65\\
+ 	& 		& rs80274284	& 6.00	& 57.76 & 0.70\\		
+ 	& 		& rs184202621	& 5.52	& 53.72 & 0.68\\
+	& 		& rs200699422	& 5.30	& 6.33 & 0.87\\	
 	&		&			&		&		&\\
-23527680 &	\multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics}	& rs6057648	& 5.40	& 25.50 & 0.82\\  
-21386085 &	\multicolumn{1}{p{3cm}}{\raggedright Diabetes\\} &	rs301 &*10.52	 & 8.22 & 0.90\\
-28928442	& \multicolumn{1}{p{3cm}}{\raggedright Nature Communications}	& rs201471471	& 6.52	& 7.14 & 0.67\\
-29255261	& \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics\\}	& rs11039216	& *14.30	& 2.59 & 1.08\\
-25918132	& \multicolumn{1}{p{3cm}}{\raggedright Toxicological\\Sciences}	& rs76780579	& 6.00	& 2.58	& 1.17 \\
-27863252	& \multicolumn{1}{p{3cm}}{\raggedright Cell\\}	& rs3794738	& *13.16	& 2.39 & 0.91\\
-28698626	& Scientific Reports	& rs11015915	& 5.05	& 2.05 & 1.08\\
+23527680 &	\multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics}	& rs6057648	& 5.40	& 14.17 & 0.82\\  
+21386085 &	\multicolumn{1}{p{3cm}}{\raggedright Diabetes\\} &	rs301 &*10.52	 & 9.21 & 0.90\\
+28928442	& \multicolumn{1}{p{3cm}}{\raggedright Nature Communications}	& rs201471471	& 6.52	& 2.5 & 0.67\\
 
@@
 \end{tabular}
-\caption{A list of eight recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
+\caption{A list of four recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
 The variants reaching genome wide significance have a star ( * ).
@@
 
-For the reverse GWAS we used less stringent quality controls to broaden the scope of our analysis.
-The only filtration used was the application of an allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of 37,674,915 variants ($S$) included in the test (see Table \ref{totTable}).
+For the reverse GWAS, the only filtration used was the application of an allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) as well as a strict removal of sites with any missing data resulting in a total of 37,700,882 variants ($S$) included in the test (see Table \ref{totTable}).
 We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 8,393,290 & 26,240,754   \\  
-{INDEL} &  749,724  & 2,291,150 \\ \hline
+{SNP}  & 8,393,467 & 26,241,186   \\  
+{INDEL} &  756,889  & 2,309,280 \\ \hline
 \end{tabular}
@@
 \begin{align*}
-{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{PC1}_{i} + \beta_{2}{PC2}_{i} + \beta_{3}{PC3}_{i} + \beta_{4}{PC4}_{i} + \beta_{5}{PC5}_{i} + \beta_{6}Q_{i}
+%{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{PC1}_{i} + \beta_{2}{PC2}_{i} + \beta_{3}{PC3}_{i} + \beta_{4}{PC4}_{i} + \beta_{5}{PC5}_{i} + \beta_{6}Q_{i}
+{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{Pop}_{i}  + \beta_{2}Q_{i}
 \\
@@
 Here $Y_{si}$ is a Bernoulli random variable denoting the genotype of individual $i$ at genomic locus $s$. 
-In the model, the regression coefficient $\beta_{1}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
+%In the model, the regression coefficient $\beta_{1}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
+In the model, the regression coefficient $\beta_{1}{Pop}_{i}$ represents the coefficient of the categorical variable of Population, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
 We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
 
-For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{6}^{(s)}=0$. That is, the average quality of mapped bases $Q$ variable is not associated with genotype $Y_{s}$.
+For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{2}^{(s)}=0$. That is, the average quality of mapped bases $Q$ variable is not associated with genotype $Y_{s}$.
 In other words, the slope coefficient of the simple logistic regression is zero.
@@
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 22,186 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 11,416 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 

### 2019-03-05 Misc/DownSampleVCF.sh

@@ -14,7 +14,7 @@ $GenoPath/ALL.chip.POS
 for f in `seq 1 22`
   do echo $f
   #download VCF if needed
-  if [ ! -f $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz]; then
+  if [ ! -f $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz ]; then
     echo "Downloading VCF CHR ${f}"
     wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz
   fi

### 2019-03-05 Misc/MeanDev.R

@@ -1,31 +1,0 @@
-##Manhattan
-library(dplyr)
-library(magrittr)
-library(data.table)
-library(ggplot2)
-library(multtest)
-library(cowplot)
-
-dir='~/Documents/Regression/'
-out='~/Documents/Regression/'
-
-for(chrom in seq(1,5)){
-	print(chrom)
-	#read in the regressions for each pop
-	fileNames = list.files(path=dir,pattern=paste("CHR",chrom,"_NOT.Regression",sep=''), full.names = T)
-	if(length(fileNames) != 0){
-	Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T, sep=' ')))
-	
-	print(head(Reg))
-	#had duplicate headers, so had to remove them and reset columns to numeric
-	Reg<-Reg[which(Reg$Pos != 'Pos'),]
-	Reg$Chr<-as.numeric(as.character(Reg$Chr))
-	Reg$Pos<-as.numeric(as.character(Reg$Pos))
-	Reg$dev<-as.numeric(as.character(Reg$dev))
-	#chisquared random variable
-	#taking the sum of the deviances (mean is to take into account the DF)
-	Reg <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
-	
-	write.table(Reg,paste(out,chrom,'NOT_MeanDev.csv',sep=''))
-	}
-	}

### 2019-03-05 PlottingScripts/CompareModels.R

@@ -7,17 +7,14 @@ library(lfa)
 library(tidyverse)
 
 #read Genotypes
-#GT = fread('gzcat ~/genomes/genomes/hg19/Genotypes/CHR22.Genotypes.txt.gz',nrows=50000)
-GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.genotypes.txt')
-
-
-SigPos = fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt', col.names=c('CHROM','POS','rsID','ID'))
+GT = fread('gzcat ~/genomes/genomes/hg19/Genotypes/CHR22.Genotypes.txt.gz',nrows=50000)
+#GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.genotypes.txt')
+#SigPos = fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt', col.names=c('CHROM','POS','rsID','ID'))
 
 ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
 names(GT) = as.character(unlist(ColNames))
-GT = merge(GT, SigPos, by=c('CHROM','POS','ID'))
-GT$rsID = NULL
-
+#GT = merge(GT, SigPos, by=c('CHROM','POS','ID'))
+#GT$rsID = NULL
 
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt') 
 
@@ -54,12 +51,12 @@ fit_model_1 <- function(data){
 
 fits <- lapply(head(data_list,10000), fit_model_1)
 
-Pop = apply(GT[c(1:10000),-(1:3)], 1, function(x)
+Pop = apply(GT[,-(1:3)], 1, function(x)
 						glm2(x ~
 							samples$Pop +
 							samples$average_quality_of_mapped_bases))
 
-PC = apply(GT[c(1:10000),-(1:3)], 1, function(x)
+PC = apply(GT[,-(1:3)], 1, function(x)
 						glm2(x ~
 							samples$PC1 + 
 							samples$PC2 +
@@ -67,7 +64,17 @@ PC = apply(GT[c(1:10000),-(1:3)], 1, function(x)
 							samples$PC4 +
 							samples$PC5 + 
 							samples$average_quality_of_mapped_bases))
-																					
+							
+PopPC = apply(GT[,-(1:3)], 1, function(x)
+						glm2(x ~
[…]

### 2019-03-05 PlottingScripts/Han_1kGP_SFS_clean.R

@@ -3,7 +3,12 @@ library(data.table)
 library(ggplot2)
 library(reshape2)
 
-#bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariants_chr_CHROM_POS.txt /Users/luke/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_sorted.vcf.gz  -Oz -o /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001.vcf.gz
+#grep '^#' $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz     > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf
+
+#for f in `seq 1 22`; do echo $f; bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariantsCHROM_POS.txt $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz | grep -v '#' >> ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf; done
+
+
+#bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariants_chr_CHROM_POS.txt /Users/luke/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_sorted.vcf.gz  -Oz -o /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.01.vcf.gz
 
 #java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar extractFields /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001.vcf.gz CHROM POS AF "GEN[*].GT" | sed "s/0\/0/0/g ; s/0\/[1-9]/1/g ; s/[1-9]\/0/1/g ; s/[1-9]\/[1-9]/1/g ; /\//d" | tail -n +2 > /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.001_genotype.txt
 

### 2019-03-05 legacy.tex
updated Han Resequenced plots

@@
 We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 11,416 suspicious variants identified globally, only 8,697 of them are found in the 83 individuals in the original 1kGP data.
-From this subset, 8,085 of these variants are missing entirely from the resequenced data.
+Of the 11,416 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data.
+From this subset, only 613 of these variants are present in the resequenced data, the rest are missing entirely.
 We noticed a unusually large number of INDELs that are fixed in the resequenced data that are at lower frequencies in the 1kGP. (See Figure \ref{90HanSFS_full})
@@
 
-We found that 476 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
+We found that 219 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
 This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.

### 2019-03-05 PlottingScripts/Han_1kGP_SFS_clean.R

@@ -2,19 +2,21 @@
 library(data.table)
 library(ggplot2)
 library(reshape2)
+library(cowplot)
+library(dplyr)
 
-#grep '^#' $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz     > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf
+#zgrep '^#' $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz     > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf
 
 #for f in `seq 1 22`; do echo $f; bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariantsCHROM_POS.txt $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz | grep -v '#' >> ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf; done
-
+#java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar extractFields ~/Documents/QualityPaper/sig/SigVar_0.01_Pop.vcf CHROM POS AF "GEN[*].GT" | sed "s/0|0/0/g ; s/0|[1-9]/1/g ; s/[1-9]|0/1/g ; s/[1-9]|[1-9]/1/g ; /|/d" | tail -n +2 > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01_genotype.txt
 
 #bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariants_chr_CHROM_POS.txt /Users/luke/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_sorted.vcf.gz  -Oz -o /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.01.vcf.gz
 
[…]

### 2019-03-05 rick/1KGP.Rmd
update

@@
 author: 
-- name: Rick Farouni
+- name: Rick Farouni, 
   affiliation:
@@
 library(Matrix)
-library(irlba)
-library(threejs)
+#library(irlba)
+#library(threejs)
 library(tidyverse)
@@
 
-# Download data
-
-Download a fast parser program (after compilation you'll have a program called a.out)
-
-```{bash}
-wget https://raw.githubusercontent.com/bwlewis/1000_genomes_examples/master/parse.c
-cc -O2 parse.c
-```
-
-Download 1000 genomes phenotype data file
-
-```{bash}
-wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped
-```
-
-Download 1000 genomes genotype data file for chr 20
-
-```{bash}
-wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz
-
-```
-
-Read just the header of the chromosome file to obtain the sample identifiers
-
-```{r}
-ids <- readLines(pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
-                      sed -n /^#CHROM/p | 
-                      tr '\t' '\n' | 
-                      tail -n +10"))
-head(ids)
-```
-
-Write ids to file
-
-```{r}
-writeLines(ids, "ids.txt")
-```
+# Download metadata
 
@@
   do
-  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.chrom20.*.bam.bas
+  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.mapped.*.bam.bas -P ./data/bas
 done
@@
 
-# Read bas files
+## Read bas files
 
-```{r}
+```{r read_bas}
 read_basfiles <- function(bas_filepath) {
@@
 ```
+
+### Preprocess table
 
@@
 ```{r}
-write_tsv(bas_dt, "average_quality_chrom21.txt")
+bas_filepath <- "./data/metadata/average_quality_dt.txt"
+write_tsv(bas_dt, bas_filepath)
 ```
@@
 ```{r}
-bas_dt <- read_tsv( "average_quality_chrom21.txt")
+bas_dt <- read_tsv(bas_filepath)
 avg_qual_mean <- bas_dt$avg_qual_mean
@@
 
-get_data <- function(){
-
-  p <- pipe("zcat ALL.chr20.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
-          sed /^#/d  |
-          cut  -f '10-' |
-          ./a.out | 
-          cut -f '1-2'")
-  x <- read.table(p,
-                colClasses=c("integer","integer"), 
-                fill=TRUE,
-                row.names=NULL)
-
-  data <- sparseMatrix(i=x[,2], j=x[,1], x=1L)
-
-  return(data)
-
-}
-
 get_af_data <- function(data, d=4){
@@
   # compute the first d factors including the intercept
+
+  message("running lfa")
+
   LF <- lfa(data, d)
   # compute the individual-specific allele frequencies
+
+  message("running  af")
+
   data_iaf <- af(data, LF)
@@
   names(loci) <- loci
+
+  message("to datalist")
 
@@
 
-  data <- get_data()
-  data <- t(as.matrix(data[,loci]))
+  data <- read.bed("./data/plink/ALL_Pops")
+
+  message("plink data loaded")
+
   data_list <- get_af_data(data)
@@
 
+  message("fitting model")
+
   data_list <- 
     data_list %>%
-  slice(1:10) %>%
+  #slice(1:10) %>%
   transmute(
@@
 ```{r}
-data_list <- run_workflow (loci=c(1:50000))
+data_list <- run_workflow ()
 data_list
@@
 
+```{r save_data}
+saveRDS(data_list, "./data/preprocessed/data_fit.rds")
+```
+
 ```{r}
-n <-5
+n <-20
 target_loci <- 
@@
     mutate(plotted= future_map(predicted, get_plot)) 
+data_list_subset
 ```
@@
 ```
-
-## Draft
-
-The matrix has 2,504 rows (people) and 1,812,841 columns (variants)
-
-Determine genome-wide goodness-of-fit for a particular value of d by computing SNP-by-SNP goodness-of-fit when compared to population structure.
-
-```{r}
-#gof_4 <- model.gof(chr20_mtx, LF, 3)
-#LF <- lfa(chr20_mtx, 10)
-#gof_10 <- model.gof(chr20_mtx, LF, 3)
-#hist(gof_4)
-```
-
-```{r}
-
-data_list
-```
-
-```{r}
-p_values_dt <-
-  data_list  %>% 
-  select(tidied) %>% 
-  map_depth( 2, "p.value") %>% 
-  map_depth( 2, c(2)) %>% # c(1) for the intercept
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "p.value")%>%
-  mutate(locus= data_list$locus)
-
-p_values_dt
-```
-
-```{r}
-estimates_dt <-
-  data_list  %>% 
-  select(tidied) %>% 
-  map_depth( 2, "estimate")%>% 
-  map_depth( 2, c(2)) %>% # c(1) for the intercept
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "beta_hat")%>%
-  mutate(locus= data_list$locus)
-
-estimates_dt
-```
-```{r}
-ggplot(p_values_dt)+
-  geom_histogram((aes(x=p.value)), bins=100) +
-  theme_bw()
-```
-
-get a p-value for target specific loci
-
-```{r}
-
-```
-
-```{r}
-p_values_subset <-
-  data_list  %>% 
-  filter(locus %in% target_loci) %>% 
-  select(tidied) %>% 
-  map_depth( 2, "p.value") %>% 
-   map_depth( 2, c(2)) %>%
-  unlist(use.names = FALSE) %>% 
-  enframe(name = "locus" ,
-          value= "p.value")%>%
-  mutate(locus= target_loci)
-p_values_subset 
-```
-
-```{r}
-
-bas_dt <- bind_cols(bas_dt,
-                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
-```
-
-```{r}
-ggplot(bas_dt,
-       aes(lf2, lf3, color=Population)) +
-  geom_point() +
-  theme_bw()
-```
-
-### work with a subset
-
-```{r}
-chr20_sub <- chr20[,1:50000]
-```
-
-# PCA
-
-compute the first three principal component vectors
-
-```{r}
-cm <- colMeans(chr20_sub)
-p <- irlba(chr20_sub,
-           nv=3,
-           nu=3,
-           tol=0.1,
-           center=cm)
-```
-
- add to dataset
-```{r}
-colnames(p$u) <- paste0("pc",1:3)
-bas_dt <- bind_cols(bas_dt, 
-                    as_tibble(p$u))
-```
-
-```{r}
-ggplot(bas_dt,
-       aes(pc1, pc2, color=Population)) +
-  geom_point() +
-  theme_bw() 
-```

### 2019-03-05 rick/1KGP_draft.Rmd
update

--- 2019-03-02 rick/1KGP.Rmd
@@
     data_list %>%
-  slice(1:10) %>%
+  #slice(1:10) %>%
   transmute(

### 2019-03-05 legacy.tex
minor changes, prepping plots

@@
 Each of these methods is somewhat similar to the genotype-conditional association test described by Song et al. \citep{song2015testing}.
-One approach included the top five global principal components (PCs)as covariates to account for population structure without using a categorical variable.
-Another combined both the population categorical variable as well as all five global PCs.
+These methods model each genotype as a logistic function using principal components (PCs), Population membership or both as a covariate. 
+We found that these three models to be consistent except a small number of variants with higher associations with the PC model.
+We chose the more conservative approach using Population membership as the sole covariate.
 
@@
 In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
-After downsampling the 1kGP data to match the resequenced individuals, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
+After restricting the analysis to individuals in both datasets, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
 While this is more than 0.1\% of the sites, we suspect that they are indeed false associations to $Q$.
-To confirm this, we also ran the reverse-GWAS approach on these resequenced individuals, using the $Q$ scores from the 1kGP sequencing \ref{90Han}.
-We found no correlation in p-values indicating that the SNPs in the resequenced dataset had no association to $Q$.
 
 We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 11,416 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data.
+Of the 11,416 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
 From this subset, only 613 of these variants are present in the resequenced data, the rest are missing entirely.
-We noticed a unusually large number of INDELs that are fixed in the resequenced data that are at lower frequencies in the 1kGP. (See Figure \ref{90HanSFS_full})
+Figure \ref{90HanSFS_full} shows a weak correlation to allele frequency for the variants present in the resequenced data.
+Whereas the vast majority of the SNPs are absent in resequenced data, a high proportion of INDELs are fixed in the resequenced data.
+A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are real but genotyped incorrectly.
 
@@
 \begin{figure}[h]
-\includegraphics[width=8cm,keepaspectratio]{./Figures/ResequencePvals.jpg}
-\caption{Comparison of p-values from the CHS and CHB populations using the original 1kGP data and the 90 Han Chinese resequenced data from \citep{Lan2017}.
-Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.
-There is no correlation in the p-values between the two datasets indicating that the suspicious variants found in the more recent dataset are false discoveries.}
-\label{90Han}
-\end{figure}
-
-\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap.jpg}

### 2019-03-05 PlottingScripts/AllSig_RunModel.R

@@ -2,6 +2,7 @@ library(dplyr)
 library(data.table)
 library(ggplot2)
 library(glm2)
+library(cowplot)
 
 #read colors per pop (used in 1kG Nature)
 colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
@@ -11,30 +12,63 @@ names(MyColour) <- colors $Pop
 path = '/Users/luke/genomes/genomes/hg19/Genotypes/'
 
 #read Genotypes
-GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_1kGP_clean_GT.txt')
+GT = fread('~/Documents/QualityPaper/sig/SigVar_Pop_0.01_genotype.txt')
 ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
 names(GT) = as.character(unlist(ColNames))
+
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
 samples$Pop = factor(samples$Pop, levels=c('FIN','GBR','CEU','IBS','TSI','CHS','CDX','CHB','JPT','KHV','GIH','STU','PJL','ITU','BEB','PEL','MXL','CLM','PUR','ASW','ACB','GWD','YRI','LWK','ESN','MSL'))
 
 model = apply(GT[,-(1:3)], 1, function(x)
 						glm2(x ~
-							samples$PC1 + 
-							samples$PC2 +
-							samples$PC3 +
-							samples$PC4 +
-							samples$PC5 + 
-							samples$average_quality_of_mapped_bases,))
+							samples$Pop +
+							samples$average_quality_of_mapped_bases))
 #get coef of Qual per site
-coef = lapply(model, function(x) coef(x)[[7]])				
+coef = lapply(model, function(x) coef(x)[[27]])				
 
-dev = lapply(model, function(x) anova(x, test='Chi')[[7,2]])
+dev = lapply(model, function(x) anova(x, test='Chi')[[3,2]])
 
-coef = cbind(unlist(coef), GT[,c(1:3)])
-coef = cbind(coef, unlist(dev))
-coef$ID = as.numeric(as.character(coef$ID))
-coef$pval = -log10(pchisq(coef$V2, 1, lower.tail=F))
[…]

### 2019-03-05 PlottingScripts/CompareModels.R

@@ -3,53 +3,14 @@ library(data.table)
 library(ggplot2)
 library(glm2)
 library(cowplot)
-library(lfa)
-library(tidyverse)
 
 #read Genotypes
 GT = fread('gzcat ~/genomes/genomes/hg19/Genotypes/CHR22.Genotypes.txt.gz',nrows=50000)
-#GT = fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.genotypes.txt')
-#SigPos = fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt', col.names=c('CHROM','POS','rsID','ID'))
 
 ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
 names(GT) = as.character(unlist(ColNames))
-#GT = merge(GT, SigPos, by=c('CHROM','POS','ID'))
-#GT$rsID = NULL
 
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt') 
-
-# Logistic factor analysis
-GT_mtx <- as.matrix(GT[,-(1:3)])
-#compute the first three factors including the intercept
-LF <- lfa(GT_mtx, 4)
-subset <- af(GT_mtx, LF)
-subset <- cbind(GT[,c(1:3)], subset)
-
-to_datalist <- function(i, data_genotype, data_iaf){
-  
-    data <- tibble(genotype=data_genotype[i,],
-               iaf= subset[i,])
-               
-    return(data)       
-}
-
-loci <- seq(1,nrow(GT_mtx))
-names(loci) <- loci
-
-data_list <- map(loci, 
-                 to_datalist,
-                 data_genotype = GT_mtx,
-                 data_iaf = subset)
-
-fit_model_1 <- function(data){
-	
-	fit <- glm2(genotype ~ samples$average_quality_of_mapped_bases + offset(iaf), 
-				family ="binomial", data=data)
-	
-	return(fit)
-}
-
-fits <- lapply(head(data_list,10000), fit_model_1)
 
 Pop = apply(GT[,-(1:3)], 1, function(x)
 						glm2(x ~
@@ -74,46 +35,26 @@ PopPC = apply(GT[,-(1:3)], 1, function(x)
 							samples$PC4 +
 							samples$PC5 + 
[…]

### 2019-03-05 legacy.tex
Updated text as well as OverUnder30 Plot

@@
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses \luke{and spurious GWAS associations as a result of imputation using the 1kGP as a reference.}
+Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
 
@@
 Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
-Thus, the impact of mutational processes can be inferred by considering the DNA context of a polymorphism \citep{Alexandrov2013,Shiraishi2015a}. 
-%Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
+Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
 Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
-
 %This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{./Figures/Han_1kGP_SFS.jpg}
+\includegraphics[width=15cm,keepaspectratio]{./Figures/Han_1kGP_SFS.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot shows  8697 significant SNPs and INDELs identified in the \textit{5PC} model including all populations.
-Only 612 of the variants are present in the high depth resequenced individuals.}
+This plot shows  7,607 significant SNPs and INDELs identified in the reverse GWAS model including all populations.
+Only 613 of the variants are present in the high depth resequenced individuals.}
+\label{90HanSFS_full}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of suspicious variants for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.}
 \label{90HanSFS_full}

### 2019-03-05 PlottingScripts/AllSig_RunModel.R

@@ -12,9 +12,11 @@ names(MyColour) <- colors $Pop
 path = '/Users/luke/genomes/genomes/hg19/Genotypes/'
 
 #read Genotypes
-GT = fread('~/Documents/QualityPaper/sig/SigVar_Pop_0.01_genotype.txt')
-ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
+GT = fread('~/Documents/QualityPaper/sig/SigVar_Pop_0.01_rsID.genotype.txt')
+ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames_rsID.txt', header=F)
 names(GT) = as.character(unlist(ColNames))
+names(GT)[4] = 'rsID'
+GT$ID <- as.numeric(as.character(GT$ID))
 
 samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
 samples$Pop = factor(samples$Pop, levels=c('FIN','GBR','CEU','IBS','TSI','CHS','CDX','CHB','JPT','KHV','GIH','STU','PJL','ITU','BEB','PEL','MXL','CLM','PUR','ASW','ACB','GWD','YRI','LWK','ESN','MSL'))
@@ -35,6 +37,7 @@ plt$pval = -log10(pchisq(plt$V2, 1, lower.tail=F))
 
 #to deal with duplicate entries, merge based on Chr Pos and AF
 SigPos = fread('/Users/luke/Documents/QualityPaper/sig/SignificantVariants.txt')
+
 plt = merge(plt, SigPos, by = c('CHROM','POS','ID'))
 
 names(plt) = c('Beta','CHROM','POS','AF','deviance','-log10(p)')
@@ -45,30 +48,34 @@ ggplot(plt, aes(x=ID, y=V1, color=log10P_0.01))+geom_point()+theme_bw()+labs(x='
 
 ##Compare over and under 30
 
-mGT = melt(unique(GT), id = c('CHROM','POS','ID'))
+mGT = melt(unique(GT), id = c('CHROM','POS','ID','rsID'))
 mGT = merge(samples[,c('Name','Pop','average_quality_of_mapped_bases')], mGT, by.x='Name',by.y='variable')
-mGT = merge(mGT, SigPos, by=c('CHROM','POS','ID'))
+mGT = merge(mGT, SigPos, by=c('CHROM','POS','ID','rsID'))
 
[…]

### 2019-03-06 rick/1KGP.Rmd
update

@@
 ---
-title: "HGP analysis"
+title: "HGP analysis: fit model"
 author: 
@@
 ```{r}
-library(Matrix)
-#library(irlba)
-#library(threejs)
 library(tidyverse)
@@
 library(broom)
+library(tictoc)
 options(future.globals.maxSize= 10000000000)
@@
 
-# Download metadata
-
-Download bas files for all the samples
-
-```{bash message=FALSE}
-for id in `cat ids.txt`
-  do
-  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.mapped.*.bam.bas -P ./data/bas
-done
-wait
-```
-
-## Read bas files
-
-```{r read_bas}
-read_basfiles <- function(bas_filepath) {
-
-  bas_file <- read_tsv(bas_filepath,
-                  col_types = cols_only('sample' = 'c', 
-                                        'platform' = 'c',        
-                                       'average_quality_of_mapped_bases' = col_guess())) 
-  return( bas_file)
-}
-```
-
-### Preprocess table
-
 ```{r}
-bas_files <- list.files(path = ".",
-                        pattern="*bam.bas")
-
-bas_dt <- 
-  map_dfr(bas_files,read_basfiles ) %>%
-  rename(avg_qual="average_quality_of_mapped_bases")
-bas_dt
+gc()
 ```
@@
 ```{r}
-bas_dt <-
-  bas_dt %>%
-  group_by(sample) %>%
-  mutate(id=1:n()) %>%
-  select(-platform)%>%
-  spread(id,avg_qual)
-bas_dt
-```
-```{r}
-n_not_na <- function(x) {
- sum(!is.na(x))
-}
+gcinfo(TRUE) 
 ```
 
-```{r}
- bas_dt <- 
-  bas_dt %>%
-  ungroup()%>% 
-    mutate( avg_qual_sd =pmap_dbl(select(., -sample), lift_vd(sd,na.rm =TRUE )),
-           avg_qual_mean = rowMeans(select(., -sample),na.rm =TRUE),
-           avg_qual_n= pmap_dbl(select(., -sample), lift_vd(n_not_na)),
-           avg_qual_min= pmap_dbl(select(., -sample), lift_vd(min, na.rm =TRUE)),
-           avg_qual_max= pmap_dbl(select(., -sample), lift_vd(max, na.rm =TRUE))) %>%
-  select (sample, avg_qual_n, avg_qual_min,avg_qual_max,avg_qual_mean,avg_qual_sd, everything() ) 
-```
-
-```{r}
-ped <- 
-  read_tsv(url("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped"))%>% 
-  rename(sample="Individual ID")
-ped 
-```
-
-```{r}
-bas_dt <- inner_join( ped  %>% 
-                       select(sample, Population),
-                       bas_dt,
-                     by="sample")
-bas_dt
-```
-
-```{r}
-bas_filepath <- "./data/metadata/average_quality_dt.txt"
-write_tsv(bas_dt, bas_filepath)
-```
-
-Plot
-
-```{r fig.width=20}
-ggplot(bas_dt,
-       aes(x=avg_qual_min,
-           y=avg_qual_max, 
-           colour=avg_qual_n )) +
-  geom_point(size=2)
-
-```
-
-```{r fig.width=20}
-ggplot(bas_dt,
-       aes(x=avg_qual_mean,
-           y=avg_qual_sd, 
-           colour=avg_qual_n )) +
-  geom_point(size=2)
-
-```
-
-# Logistic factor analysis
-
-```{r}
-bas_dt <- read_tsv(bas_filepath)
-avg_qual_mean <- bas_dt$avg_qual_mean
-```
+#  Fit model
 
@@
 
-get_af_data <- function(data, d=4){
+get_af_data <- function(){
+
+  data <- read.bed("./data/plink/ALL_Pops")
+  message("plink data loaded")
+  str(data)
+  LF <- readRDS("./data/preprocessed/lf.rds")
 
@@
 
-  message("running lfa")
+  message("running  af")
+  tic()
+  data_iaf <- af(data, LF)
+  toc()
 
-  LF <- lfa(data, d)
-  # compute the individual-specific allele frequencies
-
-  message("running  af")
-
-  data_iaf <- af(data, LF)
   loci <- c(1:dim(data)[1])
@@
 
+  tic()
   data_list <- future_map(loci, 
@@
                  data_iaf=data_iaf)
+
+  toc()
 
@@
 ```{r}
+bas_filepath <- "./data/metadata/average_quality_dt.txt"
+bas_dt <- read_tsv(bas_filepath)
+avg_qual_mean <- bas_dt$avg_qual_mean
+```
+
+```{r}
 run_workflow <- function(loci){
 
-  data <- read.bed("./data/plink/ALL_Pops")
-
-  message("plink data loaded")
-
-  data_list <- get_af_data(data)
-
-  rm(data)
-  gc()
+  data_list <- get_af_data()
 
   message("fitting model")
-
+  tic()
   data_list <- 
@@
     arrange(-deviance)
+
+  toc()
 

### 2019-03-06 legacy.tex
minor updates

@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
+Over 3,826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.001$ \citep{Benjamini2006}.
+To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 
@@
 \subsection{Cell line or technical artifact}
-
+\luke{Do we keep HapMap? It may be worth keeping some of it to explain why 2/4 papers impute using HapMap and find the same variants.}
 To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
@@
 After restricting the analysis to individuals in both datasets, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
-While this is more than 0.1\% of the sites, we suspect that they are indeed false associations to $Q$.
+While this is more than 0.1\% of the sites, we suspect that some of these variants may be associated to $Q$ as a result of incorrect genotyping.
 
@@
 
-One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome}.
+One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
 Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
-And one study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
 

### 2019-03-07 rick/1KGP.Rmd
update

@@
 library(tictoc)
-options(future.globals.maxSize= 10000000000)
+options(future.globals.maxSize= 1000000000000)
 plan(multicore)
@@
   toc()
+
+  saveRDS(data_iaf, "./data/preprocessed/iaf.rds")
 
@@
 
+  saveRDS(data_list, "./data/preprocessed/data_list.rds")
+
   message("fitting model")
@@
 ```{r save_data}
-saveRDS(data_list, "./data/preprocessed/data_fit.rds")
+saveRDS(data_list, "./data/preprocessed/data_fits.rds")
 ```

### 2019-03-07 legacy.tex
update

@@
 The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. 
-%(\sgcomment {p value?})
 Thus, sequences called from low-$Q$ data contain mutations that reproduce poorly across studies and exhibit a particular mutational signature.
@@
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses where quality of imputation matters, identifying suspicious SNPs may be preferable. 
-Lists of suspicious SNPs and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
+For other analyses where quality of imputation matters, identifying suspicious variants may be preferable. 
+Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
 

### 2019-03-08 legacy.tex
update

@@
 We found that these three models to be consistent except a small number of variants with higher associations with the PC model.
-We chose the more conservative approach using Population membership as the sole covariate.
+We chose the more conservative approach using Population membership as the sole covariate (See Figure \ref{CompareModel}).
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/dev_Pop_PC_PopPC.jpg}
+\caption{Comparison of three logistic regression models for testing association to $Q$.
+These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or both (PopPC) as a covariate.
+In these plots we are comparing the deviance from the null model in 50,000 variants from chromosome 22.}  
+\label{CompareModel}
+\end{figure}
 
 \subsection{Cell line or technical artifact}
-\luke{Do we keep HapMap? It may be worth keeping some of it to explain why 2/4 papers impute using HapMap and find the same variants.}
+%\luke{Do we keep HapMap? It may be worth keeping some of it to explain why 2/4 papers impute using HapMap and find the same variants.}
 To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.

### 2019-03-08 Misc/Parallel_VCFtoBED.py

@@ -1,0 +1,27 @@
+import multiprocessing as mp
+import sys, os
+import time
+
+
+def Bash_cmd(i):
+    GenoName='phase3_shapeit2_mvncall_integrated_v5a.20130502'
+    #print statement to keep track of where you are
+    print('Working on Chrom : '+str(i))
+    #put your bash command below
+    path='/Users/luke/genomes/genomes/hg19/'
+    os.system('/Users/luke/bin/plink_mac/plink \
+    --vcf {0}/phase3//ALL.chr{1}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz \
+    --make-bed \
+    --out {0}/plink/chr{1}'.format(path, i))
+
+if __name__ == '__main__':
+    #you can replace ListOfChrom with whatever you are iterating through
+    ListOfChrom=list(range(1,23))
+    #pool_size is number of CPU
+    pool_size=7 #mp.cpu_count()
+    pool=mp.Pool(processes=pool_size)
+    #this is where you are calling the parallelization
+    pool_outputs= pool.map(Bash_cmd, ListOfChrom)
+    #some magic, not sure what this does
+    pool.close()
+    pool.join()

### 2019-03-08 PlottingScripts/CompareModels.R

@@ -69,37 +69,40 @@ plt = data.frame('coefPopPC' = unlist(coefPopPC),
 				  'devPC' = unlist(devPC))
 				  
 c1<-ggplot(plt, aes(coefPop, coefPC))+
-				geom_point(shape=1)+
+				geom_point(shape=1, alpha=0.6)+
 				geom_abline()+
 				labs(x='Pop',y='PC')
 c2<-ggplot(plt, aes(coefPopPC, coefPC))+
-				geom_point(shape=1)+
+				geom_point(shape=1, alpha=0.6)+
 				geom_abline()+
 				labs(x='PopPC',y='PC')
 c3<-ggplot(plt, aes(coefPopPC, coefPop))+
-				geom_point(shape=1)+
+				geom_point(shape=1, alpha=0.6)+
 				geom_abline()+
 				labs(x='PopPC',y='Pop')				  
 
 d1<-ggplot(plt, aes(devPop, devPC))+
-			geom_point(shape=1)+
-			geom_abline()+
+			geom_abline(alpha=0.3, linetype=3)+
+			geom_point(shape=1, alpha=0.6)+
 			labs(x='Pop',y='PC')
 d2<-ggplot(plt, aes(devPopPC, devPC))+
-			geom_point(shape=1)+
-			geom_abline()+
+			geom_abline(alpha=0.3, linetype=3)+
+			geom_point(shape=1, alpha=0.6)+
 			labs(x='PopPC',y='PC')
-d3<-ggplot(plt, aes(devPopPC, devPop))+
-			geom_point(shape=1)+
-			geom_abline()+
+d3<-ggplot(plt, aes(devPop, devPopPC))+
+			geom_abline(alpha=0.3, linetype=3)+
+			geom_point(shape=1, alpha=0.6)+
 			labs(x='PopPC',y='Pop')
 
 plot_grid(c1,c2,c3,d1,d2,d3, nrow=2)
-ggsave('~/Documents/QualityPaper/Figures/CoefDEV_Pop_PC_PopPC.jpg',height=9,width=12)
+ggsave('~/Documents/QualityPaper/Figures/CoefDEV_Pop_PC_PopPC.jpg',height=6,width=12)
 
 
-plot_grid(c1,c2,c3,d1,nrow=2)
-ggsave('~/Documents/QualityPaper/Figures/Coef_Pop_PC_PopPC.jpg',height=10,width=10)
[…]

### 2019-03-11 rick/final/1_make_quality_dt.Rmd
update

--- 2019-03-05 rick/1KGP.Rmd
@@
 ```{r}
-library(Matrix)
-#library(irlba)
-#library(threejs)
 library(tidyverse)
-library(lfa)
-library(furrr)
-library(broom)
-options(future.globals.maxSize= 10000000000)
-plan(multicore)
-```
-
-# Download metadata
-
-Download bas files for all the samples
-
-```{bash message=FALSE}
-for id in `cat ids.txt`
-  do
-  wget ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/${id}/alignment/${id}.mapped.*.bam.bas -P ./data/bas
-done
-wait
 ```
@@
 ```{r}
-bas_files <- list.files(path = ".",
-                        pattern="*bam.bas")
+bas_files <- list.files(path = "./data/bas",
+                        pattern="*bam.bas",
+                        full.names=TRUE)
+```
+
+```{r}
 
 bas_dt <- 
-  map_dfr(bas_files,read_basfiles ) %>%
+  map_dfr(bas_files, read_basfiles ) %>%
   rename(avg_qual="average_quality_of_mapped_bases")
@@
 ped <- 
-  read_tsv(url("ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped"))%>% 
+  read_tsv("./data/metadata/20130606_g1k.ped")%>% 
   rename(sample="Individual ID")
@@
 ```
-
-# Logistic factor analysis
-
-```{r}
-bas_dt <- read_tsv(bas_filepath)
-avg_qual_mean <- bas_dt$avg_qual_mean
-```
-
-```{r}
-
-to_datalist <- function(i, data_genotype, data_iaf){
-
-    data <- tibble(genotype=data_genotype[i,],
-               iaf=data_iaf[i,])
-
-     return(data)       
-}
-
-get_af_data <- function(data, d=4){
-
-  # compute the first d factors including the intercept
-
-  message("running lfa")
-
-  LF <- lfa(data, d)
-  # compute the individual-specific allele frequencies
-
-  message("running  af")
-
-  data_iaf <- af(data, LF)
-  loci <- c(1:dim(data)[1])
-  names(loci) <- loci
-
-  message("to datalist")
-
-  data_list <- future_map(loci, 
-                 to_datalist,
-                 data_genotype=data,
-                 data_iaf=data_iaf)
-
-  data_list <- 
-  enframe(data_list,
-          name = "locus", 
-          value = "dataset")
-
-  return(data_list)
-
-}
-
-get_predicted_mean <- function(model_fit) {
- predicted <- 
-   augment(model_fit, type.predict = "response") %>%
-   select("genotype","offset.iaf.",".fitted") %>%
-   rename(iaf="offset.iaf.", mu=".fitted" )
-  return(predicted)
-}
-
-get_stats <- function(model_fit){
-  stats_dt <- 
-    tidy(model_fit)  %>%
-    filter(term=="avg_qual_mean") %>%
-    select(estimate, p.value) %>%
-    mutate(deviance= anova(model_fit)["Deviance"][[1]][2])
-  return(stats_dt)
-}
-
-fit_model <- function(dataset){
-
-   fit <- glm(genotype ~ avg_qual_mean + offset(iaf),
-                           family ="binomial",
-                           data=dataset)
-
-   stats <- get_stats(fit)
-   predicted <- get_predicted_mean(fit)
-
-   return(list(stats, predicted))
-}
-
-get_plot <- function(data){
-
- p <- ggplot(bind_cols(data, 
-                   bas_dt %>% 
-                     select(Population,
-                            avg_qual_mean))) +
-    geom_point(aes(avg_qual_mean, 
-                   genotype),
-               size=.5,
-               alpha= 0.4) +
-    geom_line(aes(avg_qual_mean,
-                  iaf, 
-                  colour="af"),
-              linetype="dashed") +
-    geom_line(aes(avg_qual_mean,
-                  mu,
-                  colour="mu")) +
-      theme_bw() +
-    facet_wrap(~Population)
-
-  return(p)
-}
-```
-
-```{r}
-run_workflow <- function(loci){
-
-  data <- read.bed("./data/plink/ALL_Pops")
-
-  message("plink data loaded")
-
-  data_list <- get_af_data(data)
-
-  rm(data)
-  gc()
-
-  message("fitting model")
-
-  data_list <- 
-    data_list %>%
-  #slice(1:10) %>%
-  transmute(
-    locus=locus,
-    fit1 = future_map(dataset, fit_model))  %>%
-  transmute(
-    locus=locus,
-    stats = future_map(fit1, c(1)),
-    predicted =future_map(fit1, c(2))) %>%
-    unnest(stats) %>%
-    arrange(-deviance)
-
-  return(data_list)
-
-}
-
-```
-
-```{r}
-data_list <- run_workflow ()
-data_list
-```
-
-```{r save_data}
-saveRDS(data_list, "./data/preprocessed/data_fit.rds")
-```
-
-```{r}
-n <-20
-target_loci <- 
-  data_list %>% 
-  slice(1:n) %>% 
-  pull(locus)
-target_loci 
-```
-
-```{r}
-data_list_subset <- 
-  data_list %>%
-   filter(locus %in% target_loci ) %>%
-    mutate(plotted= future_map(predicted, get_plot)) 
-data_list_subset
-```
-
-```{r}
-data_list_subset %>%
-  pull(plotted)
-```
-
-```{r}
-sessionInfo()
-```

### 2019-03-11 rick/final/5_make_plots.Rmd
update

---
title: "HGP analysis: plot"
author: 
- name: Rick Farouni, 
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

# Preparations

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
options(scipen=999)
```

```{r}
library(tidyverse)
library(lfa)
library(glm2)
library(furrr)
library(broom)
library(tictoc)
options(future.globals.maxSize= 1000000000000)
plan(multicore)
```

```{r}
gc()
```

```{r}
bas_filepath <- "./data/metadata/average_quality_dt.txt"
bas_dt <- read_tsv(bas_filepath)
avg_qual_mean <- bas_dt$avg_qual_mean
```
```{r}
head(avg_qual_mean)
```

```{r}
S <-30
m <- 76871
subset <- vector("list", length = S)
for(i in 1:S){
  subset[[i]] <-readRDS(file.path("./data/preprocessed/subsets", paste0("fits_subset_", i, ".rds")))
}
```

```{r}
gc()
```

```{r}

get_plot <- function(data) {
  p <- ggplot(bind_cols(data,
                        bas_dt %>%
                          select(Population,
                                 avg_qual_mean))) +
    geom_point(aes(avg_qual_mean,
                   genotype),
               size = .5,
               alpha = 0.4) +
    geom_point(aes(avg_qual_mean,
                  iaf,
                  colour = "af")) +
    geom_line(aes(avg_qual_mean,
              p_0,
              colour = "p_0"),
          linetype = "dashed") +
    geom_line(aes(avg_qual_mean,
                  p_1,
                  colour = "p_1")) +
    theme_bw() +
    facet_wrap( ~ Population)

  return(p)
}

data_list_subset <- 
  data_list_sub %>%
    mutate(plotted= future_map(data, get_plot)) 
data_list_subset
```

```{r, fig.width=20}
data_list_subset %>%
  arrange(dev_logp )%>%
  slice(1:5)%>%
  pull(plotted)
```

```{r}
sessionInfo()
```

### 2019-03-11 Misc/AncestralContext.py
starting to make R Notebook for all figures

@@ -1,0 +1,72 @@
+import pandas as pd
+import argparse
+
+def main(args):
+    #read input file (VCF file cut first 4 cols)
+    vcfPos=pd.read_table(args.input,
+                         names=["Chrom","Pos", "Ref", "Alt"])
+
+    vcfPos.drop_duplicates(subset='Pos', keep = False)
+    #set default Derived Allele to 1
+    vcfPos['DerivedAllele'] = 1
+
+    #read chimp human alignment
+    chimp=pd.read_table('/Users/luke/bin/smaller_mut_spectrum_pipeline/'
+    +'hg19_chimp_align/human_chimp_diffs_chr'
+    + args.chrom + '.txt', sep=" ")
+    chimp= chimp[chimp['SNP/Indel'] == 'SNP']
+
+    #read oneLine fasta ref file
+    ref=open('/Users/luke/bin/smaller_mut_spectrum_pipeline/'
+    +'hg19_reference/chr' + args.chrom + '_oneline.txt')
+    refseq=ref.read()
+
+    #get context (i.e. one base up- and downstream)
+    vcfPos['Context'] = vcfPos.apply(\
+    lambda row : refseq[row.Pos - 2 : row.Pos + 1], axis=1)
+    vcfPos=vcfPos.drop_duplicates(subset='Pos')
+
+    #only keep SNPs
+    DNA=['A','C','G','T']
+    vcfPos= vcfPos[vcfPos['Ref'].isin(DNA)]
+    vcfPos= vcfPos[vcfPos['Alt'].isin(DNA)]
+    vcfPos= vcfPos[~vcfPos['Context'].isin(['N'])]
+
+    #get sites shared by both files
+    iPos = set(chimp['Pos']) & set(vcfPos['Pos'])
+    iChimp = chimp[chimp['Pos'].isin(list(iPos))]
+    iVCF = vcfPos[vcfPos['Pos'].isin(list(iPos))]
+    print('Chrom : '+args.chrom)
+    print(len(iVCF), len(iChimp))
+    assert len(iVCF) == len(iChimp)
+    #find sites where Chimp is same as Human Alt
+    ChimpDiff = iVCF[iVCF['Alt'] == list(iChimp['Chimp'])]['Pos']
+    #ChimpDiff = vcfPos['Pos'][vcfPos['Pos'].isin(chimp['Pos'])][vcfPos['Alt'] == list(chimp['Chimp'])]['Pos']
+    #print(ChimpDiff.head())
+
[…]

### 2019-03-14 rick/final/3_fit_model_compareModels.Rmd
update

---
title: "HGP analysis: fit model"
author: 
- name: Rick Farouni, 
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

  # Preparations

```{r setup}
args <- commandArgs(trailingOnly=T)

#path <- args[1]
path <- '/Users/luke/Documents/QualityPaper/rick/final/'
setwd(path)

#plinkPath <- args[2] 
plinkPath <- '/Users/luke/genomes/genomes/hg19/plink/'

#chr <- args[3]
chr <- '22_10000'

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
options(scipen=999)
```

```{r}
library(tidyverse)
library(lfa)
library(furrr)
library(glm2)
library(tictoc)
library(data.table)
library(cowplot)
options(future.globals.maxSize= 1000000000000)
plan(multicore)
```
load plink format genotype data
```{r}
data_genotype <- read.bed(paste0(plinkPath,'chr',chr))

message("plink data loaded")
```
get rsIDs and add them as rownames for genotype data
```{r}
data_bim <- fread(paste0(plinkPath,'chr',chr,'.bim'), 
                  col.names = c('Chr','rsID','X',
                                'Pos','Ref','Alt'))

rownames(data_genotype) <- data_bim$rsID
```
run lfa if needed
```{r lfa}
tic()
#LF <- lfa(data_genotype, 5)
toc()
message("lfa run complete")
```
save lfa if needed
```{r save}
tic()
#saveRDS(LF, "./data/preprocessed/lf5.rds")
toc()
```
load logistic factor analysis results
```{r read lf}
LF <- readRDS( "./data/preprocessed/lf5.rds")
```

```{r}
#this runs the logistic regression
lreg_dich <- function(y, X){

  b <- glm2(cbind(y, 2-y) ~ -1 + X, family="binomial")$coef

  return(b)

}

#modified from gcat
fit_glm_dich <- function(..., LF, trait){
  snp <- c(...)
  snp <- as.numeric((snp==1) | (snp==2))
  ind <- !is.na(snp) #logical index vector
  snp_no_na <- snp[ind]
  LF_no_na <- LF[ind,]
  p0 <- rep(NA, length(snp))
  p1 <- rep(NA, length(snp))
  trait_no_na <- trait[ind]
  b0 <- lreg_dich(snp_no_na, LF_no_na) #coefficients from logreg 
  b1 <- lreg_dich(snp_no_na, cbind(LF_no_na, trait_no_na))
  est0 <- .Call("mv", LF_no_na, b0)
  est1 <- .Call("mv", cbind(LF_no_na, trait_no_na), b1)
  p0[ind] <- exp(est0)/(1+exp(est0))
  p1[ind] <- exp(est1)/(1+exp(est1))

  dataset <-
    tibble(snp=snp,
           p0 = p0,
           p1 = p1)

  stat <-
    dataset %>%
    summarize( #cross entropy (log likelihood)
      dev0 = -2 * sum(snp * log(p0) +
                        (1 - snp) * log(1 - p0),
                      na.rm = TRUE),
      dev1 = -2 * sum(snp * log(p1) +
                        (1 - snp) * log(1 - p1),
                      na.rm = TRUE),
      devdiff = dev0 - dev1,
      dev_logp = -log10(pchisq(devdiff,
                               1,
                               lower.tail = FALSE))
    ) %>% 
    mutate(num_nas = sum(!ind),
	   num_zeros= sum(snp_no_na ==0),
           b=b1[length(b1)])

  return(list(dataset=dataset, stat=stat))

}
```

define functions
```{r}

#this runs the logistic regression
lreg <- function(y, X){
  X <- rbind(X, X)
  y1 <- as.numeric((y==1) | (y==2))
  y2 <- as.numeric(y==2)
  y <- c(y1,y2)
  b <- glm2(cbind(y, 2-y) ~ -1 + X, family="binomial")$coef

  return(b)

}
#modified from gcat
fit_glm <- function(..., LF, trait){
  snp <- c(...)
  ind <- !is.na(snp) #logical index vector
  snp_no_na <- snp[ind]
  LF_no_na <- LF[ind,]
  p0 <- rep(NA, length(snp))
  p1 <- rep(NA, length(snp))
  trait_no_na <- trait[ind]
  b0 <- lreg(snp_no_na, LF_no_na) #coefficients from logreg 
  b1 <- lreg(snp_no_na, cbind(LF_no_na, trait_no_na))
  est0 <- .Call("mv", LF_no_na, b0)
  est1 <- .Call("mv", cbind(LF_no_na, trait_no_na), b1)
  p0[ind] <- exp(est0)/(1+exp(est0))
  p1[ind] <- exp(est1)/(1+exp(est1))

  dataset <-
    tibble(snp=snp,
           p0 = p0,
           p1 = p1)

  stat <-
    dataset %>%
    summarize(
      dev0 = -2 * sum(snp * log(p0) +
                        (2 - snp) * log(1 - p0),
                      na.rm = TRUE),
      dev1 = -2 * sum(snp * log(p1) +
                        (2 - snp) * log(1 - p1),
                      na.rm = TRUE),
      devdiff = dev0 - dev1,
      dev_logp = -log10(pchisq(devdiff,
                               1,
                               lower.tail = FALSE))
    ) %>% 
    mutate(num_nas = sum(!ind),
	   num_zeros= sum(snp_no_na ==0),
           b=b1[length(b1)])

  return(list(dataset=dataset, stat=stat))

}
```

```{r}
gc()
```

```{r}
fam_file <-fread(paste0(plinkPath,'chr',chr,'.fam'), 
                  col.names=c('sample','a','b','c','d','e'))

bas_filepath <- "./data/metadata/average_quality_dt.txt"
bas_dt <- read_tsv(bas_filepath)
bas_dt <- merge(fam_file, 
                bas_dt,
                by = 'sample') 
avg_qual_mean <- bas_dt$avg_qual_mean
```

```{r}
head(avg_qual_mean)
```
#LFA
```{r, warning=FALSE}
tic("fitting model")
labels <- rownames(data_genotype)
data_fit <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm,
	LF=LF, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)

data_fit_dich <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm_dich,
	LF= LF, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)
   toc()    
```

```{r}

saveRDS(data_fit, 
	file.path("./data/preprocessed/subsets",
	paste0("fits_chrom_", chr, ".rds")))

write.table(data_fit[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("fits_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)

saveRDS(data_fit_dich, 
	file.path("./data/preprocessed/subsets",
	paste0("dich_fits_chrom_", chr, ".rds")))

write.table(data_fit_dich[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("dich_fits_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)
```
#PCA
```{r, warning=FALSE}
#load PCA
library(data.table)
samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
PC <- as.matrix(samples[,c('PC1','PC2','PC3','PC4','PC5')])
PC <- cbind(PC, 1)

tic("fitting model")
labels <- rownames(data_genotype)
data_fit_PCA <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm,
	LF= PC, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)

data_fit_PCA_dich <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm_dich,
	LF= PC, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)
   toc()    
```

```{r}

saveRDS(data_fit_PCA, 
	file.path("./data/preprocessed/subsets",
	paste0("fits_chrom_", chr, ".rds")))

write.table(data_fit_PCA[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("fitsPCA_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)

saveRDS(data_fit_PCA_dich, 
	file.path("./data/preprocessed/subsets",
	paste0("dich_fits_chrom_", chr, ".rds")))

write.table(data_fit_PCA_dich[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("dich_fitsPCA_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)
```
#Population
```{r, warning=FALSE}

testModel <- model.matrix(data = bas_dt, object = ~Population)

tic("fitting model")
labels <- rownames(data_genotype)
data_fit_Pop <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm,
	LF= testModel, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)
   toc()

data_fit_Pop_dich <- 
	data_genotype %>%
	as_tibble() %>%
	transmute(
	fit = future_pmap(., fit_glm_dich,
	LF= testModel, 
	trait=avg_qual_mean),
		rsID=labels)  %>%
	unnest(fit) %>%
	mutate(l= rep(c("data","stat"), n()/2))%>%
	spread(l, fit)%>%
	unnest(stat)
   toc()   
```
```{r}

saveRDS(data_fit_Pop, 
	file.path("./data/preprocessed/subsets",
	paste0("fits_chrom_", chr, ".rds")))

write.table(data_fit[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("fitsPOP_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)

saveRDS(data_fit_Pop_dich, 
	file.path("./data/preprocessed/subsets",
	paste0("dich_fits_chrom_", chr, ".rds")))

write.table(data_fit_dich[-2], 
              file.path("./data/preprocessed/subsets", 
                        paste0("dich_fitsPOP_chrom_", chr, ".txt")),
              quote = FALSE,
              row.names = FALSE)
```

```{r}
gc()
```

```{r}
sessionInfo()
```

```{r,fig.height = 10, fig.width = 10}

plt <- merge(data_fit[-2], data_fit_PCA[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
plt <- merge(plt, data_fit_Pop[-2], by=c('rsID'), suffixes = c('','Pop'))

p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

plot_grid(pp1,pp2,pp3,ncol=3)
ggsave('./CompareModels.jpg', height=10, width = 10)

```

```{r,fig.height = 10, fig.width = 10}

plt <- merge(data_fit_dich[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
plt <- merge(plt, data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('','Pop'))

p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

plot_grid(pp1,pp2,pp3,ncol=3)
ggsave('./CompareModels_dich.jpg', height=10, width = 10)

```

```{r,fig.height = 10, fig.width = 10}

LFAp <- merge(data_fit[-2], data_fit_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
PCAp <- merge(data_fit_PCA[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
Popp <- merge(data_fit_Pop[-2], data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))

p1<-ggplot(LFAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model LFA')
p2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
p2.2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(LFAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(PCAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model PCA')
p2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
p2.2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(PCAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

p1<-ggplot(Popp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model Pop')
p2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
p2.2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
p3<-ggplot(Popp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

plot_grid(pp1,pp2,pp3,ncol=3)
ggsave('./CompareModels_VSdich.jpg', height=10, width = 10)

```

```{r,fig.height = 10, fig.width = 6}
#plot diff
plt <- merge(data_fit, data_fit_Pop_dich, by=c('rsID'), suffixes = c('LFA_Full','Pop_Dich'))

p1<-ggplot(plt, aes(dev0LFA_Full, dev0Pop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA Full', y='Pop Dich')+ggtitle('Null Model')+coord_flip()
p2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance')+coord_flip()
p2.2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))+coord_flip()
p3<-ggplot(plt, aes(bLFA_Full, bPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Beta')+coord_flip()
pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)

pp1
ggsave('./CompareModels_FullLFA_Popdich.jpg', height=10, width = 6)
```

### 2019-03-15 PlottingScripts/ProcessPlotData.Rmd
update

---
title: "Legacy Data Confounds Modern Genomics Studies"
author: 
- name: Luke Anderson-Trocmé, 
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

load libraries
```{r, message = FALSE}
library(ggplot2)
library(RColorBrewer)
library(plotly)
library(colorspace)
library(cowplot)
library(data.table)
library(tidyverse)
library(purrr)

#Directories
DataPath <- '~/Documents/QualityPaper/ProcessedData/'
figurePath <- '~/Documents/QualityPaper/Figures/'
```

#Joint Frequency Spectrum
```{r}
if(!file.exists(paste0(DataPath,
                       'jointAF.txt'))) {
  #Load Allele Frequency files
  AF.1kGP <- paste0(DataPath,
                    '1kGP_GenomeWide_JPT_1.frq')
  JPT <- fread(
    AF.1kGP,
    fill = TRUE,
    col.names = c('Chr', 'Pos', 'N_ALLELES',
                  'N_CHR', 'JPT_AF', 'JPT_MAF')
  )
  AF.NAG <- paste0(DataPath,
                   'NAG_GenomeWide.4bed_filtered.freq.frq')
  NAG <- fread(
    AF.NAG,
    fill = TRUE,
    col.names = c('Chr', 'Pos', 'N_ALLELES',
                  'N_CHR', 'NAG_AF', 'NAG_MAF')
  )

  #merge them together to get joint frequency spectrum
  join <- merge(JPT[which(JPT$JPT_AF <= 1), ],
                NAG[which(NAG$NAG_AF <= 1), ],
                by = c('Chr', 'Pos'), all = TRUE)
  join[is.na(join)] <- 0 #missing positions are frequency of 0

  #REMOVE  HWE SITES
  hwe_1kGP <- paste0(DataPath,
                     '1kGP_GenomeWide.4bed_filtered_sig6.freq.hwe')
  JPT_hwe <- fread(hwe_1kGP)
  hwe_NAG <- paste0(DataPath,
                    'NAG_GenomeWide.4bed_filtered_sig6.freq.hwe')
  NAG_hwe <- fread(hwe_NAG)

  hwe <- merge(NAG_hwe[, c('V1', 'V2')],
               JPT_hwe[, c('V1', 'V2')],
               by = c('V1', 'V2'),
               all = TRUE)
  #remove sites from AF that don't meet HWE
  join <- join[!paste(join$CHROM, join$POS)
               %in% c(paste(hwe$V1, hwe$V2)),]

  join <- join[, c('Chr', 'Pos', 'JPT_MAF',
                   'JPT_AF', 'NAG_MAF', 'NAG_AF')]
  #make numeric
  join <- sapply(join, as.numeric)
  join <- join[complete.cases(join), ]

  rm(NAG) #free up memory
  rm(JPT)
  rm(hwe)

  join

  #save to file
  write.table(join,
              paste0(DataPath,
                     'jointAF.txt'))
}
```

## Mutation Spectrum enrichment
```{r}
#get positions of significant SNPs
sig.pos <- SigPos[,c('Chr','Pos')]
write.table(sig.pos, 
            file=paste0(DataPath,
                        'sig.pos.txt'), 
            col.names = FALSE, 
            row.names = FALSE, 
            quote = FALSE, 
            sep = '\t')

#get positions of random SNPs as control
set.seed(123)
random.pos <- NotSig[sample(nrow(NotSig), 
                            nrow(sig.6)),
                     c('Chr','Pos')]
write.table(random.pos, 
            file = paste0(DataPath,
                        'random.pos.txt'), 
            col.names = FALSE, 
            row.names = FALSE, 
            quote = FALSE, 
            sep = '\t')
rm(NotSig) #save memory
```
extract context for snps of interest
```{bash}
#Using GenomeWide.AncestralContext.txt
#To get mutation context, use $path/AncestralContext.py
#compare first two columns in both files, extract matches
path='/Users/luke/Documents/QualityPaper/Misc/'
if [ ! -f $path/random.pos.context.txt ]; then
  echo "random pos context"
  awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
  $path/random.pos.txt \
  $path/GenomeWide.AncestralContext.txt \
  > $path/random.pos.context.txt
fi

if [ ! -f $path/sig.pos.context.txt ]; then
  echo "sig pos context"
  awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
  $path/sig.pos.txt \
  $path/GenomeWide.AncestralContext.txt \
  > $path/sig.pos.context.txt
fi 
```
Load mutation context and plot enrichment
```{r}
contextHeader = c('Chr','Pos','Ref','Alt','Flip','Context')

#load SNP context
SNP <- read.table(paste0(DataPath,
                         'sig.pos.context.txt'), 
                sep = '\t', 
                header  = FALSE,
                col.names = contextHeader)
rSNP <- read.table(paste0(DataPath,
                          'random.pos.context.txt'), 
                sep = '\t',
                header = FALSE,
                col.names = contextHeader)

ReverseComp <- function(x)
        chartr("ATGC","TACG",
        sapply(
          lapply(
          strsplit(x, NULL), 
          rev), 
          paste, 
          collapse = ""))

processContext <- function(data){

#get reverse complement of context, ref and alt
data$Rev<-ReverseComp(as.character(data$Context))
data$RevRef<-ReverseComp(as.character(data$Ref))
data$RevAlt<-ReverseComp(as.character(data$Alt))

#only get one half (to fold over)
AC<-data[which(data$Ref 
               %in% 
                 c('A','C')),]
TG<-data[which(data$Ref 
               %in% 
                 c('T','G')),]
#fold over
TG$Context <- TG$Rev
TG$Ref <- TG$RevRef
TG$Alt <- TG$RevAlt
#bind them together
data <- rbind(AC,TG)
#combine it all together to get a context
data$Mut <- paste0(data$Ref,'->', 
                substr(data$Context,1,1),
                data$Alt,
                substr(data$Context,3,3))
#get the counts of each bin
binCount <- as.data.frame(table(data$Mut))
#process for plotting
binCount$Start <- substr(binCount$Var1,4,4)
binCount$End <- substr(binCount$Var1,6,6)
binCount$Mut <- paste(substr(binCount$Var1,1,3),
                      substr(binCount$Var1,5,5))
return(binCount)
}

SNP <- processContext(SNP)
rSNP <- processContext(rSNP)

SNP
#save to file
saveRDS(SNP,
            paste0(DataPath,
                   'SNP_Context.rds'))

#save to file
saveRDS(rSNP,
            paste0(DataPath,
                   'rSNP_Context.rds'))
```

```{r}
BasPath = '/Users/luke/genomes/bas_file/1000GenomesBAS/'

#load bas_dt
#see 1_make_quality_dt.Rmd for making this file
load('/Users/luke/Documents/QualityPaper/rick/bas_dt')

#get the index files for extra meta data
#wget http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index
index <- read.table(paste0(BasPath, 
                           '20130502.phase3.analysis.sequence.index'),
               comment.char = "?", 
               fill = TRUE,
               sep = "\t", 
               header = TRUE,
               stringsAsFactors = FALSE)
#keep columns of interest
index <- unique(index[,c('SAMPLE_NAME',
                         'CENTER_NAME',
                         'SUBMISSION_DATE',
                         'INSTRUMENT_MODEL')])

#Phase 1 samples, list downloaded from http://www.internationalgenome.org/data-portal/sample
phase <- read.table(paste0(DataPath,
                           'Phase1_IDs.txt'))
phase$Phase = 1

#merge tables
bas_dt <- merge(bas_dt, 
                index, 
                by.x = "sample", 
                by.y = "SAMPLE_NAME",
                all.x = TRUE)
bas_dt <- merge(bas_dt, 
                phase, 
                by.x = "sample", 
                by.y = "V1", 
                all.x = TRUE)
bas_dt[which(is.na(bas_dt$Phase)),]$Phase <- 3
#clean up the text
bas_dt$SUBMISSION_DATE <- gsub(" 00:00:00",
                               "",
                               bas_dt$SUBMISSION_DATE)
bas_dt$SUBMISSION_DATE <- as.Date(bas_dt$SUBMISSION_DATE, 
                                  "%Y-%m-%d")
bas_dt$CENTER_NAME<-gsub('Illumina',
                         'ILLUMINA',
                         bas_dt$CENTER_NAME)
bas_dt$INSTRUMENT_MODEL <-gsub('Illumina',
                               '',
                               bas_dt$INSTRUMENT_MODEL)
bas_dt$INSTRUMENT_MODEL <-gsub('Genome Analyzer',
                               'GA',
                               bas_dt$INSTRUMENT_MODEL)
bas_dt$INSTRUMENT_MODEL <-gsub('HiSeq 2000',
                               'HS 2k',
                               bas_dt$INSTRUMENT_MODEL)
#remove missing data
bas_dt<-bas_dt[complete.cases(bas_dt),]

#Get the average sequencing date per population
#Used for ranking the populations
Day<-aggregate(bas_dt$SUBMISSION_DATE, 
               list(bas_dt$Population), 
               mean)
#add it to data
metaData<-merge(bas_dt, 
                Day, 
                by.x = 'Population', 
                by.y = 'Group.1')
#reorder data by average date then quality
metaData <- metaData[with(metaData, 
                          order(x,
                                Population, 
                                avg_qual)),]
#set factor with order in mind (for plotting)
metaData$sample <- factor(metaData$sample, 
                          levels = unique(metaData$sample))

metaData
#save to file
saveRDS(metaData,
            paste0(DataPath,
                   'metaData.rds'))
```

```{r}
phase1 <- c('YRI','LWK','CHB','CHD','JPT','CEU','LWK')
#read single population regressions
Reg <- fread(paste0(DataPath,
                    'AllRegressionsOver10.txt'), 
             col.names = c('Chr','Pos',
                           'Population','deviance'), 
             colClasses = c('numeric', 'numeric', 
                            'character', 'numeric'))
#Get P-Values and significant sites
Reg$plog10 <- -log10(pchisq(Reg$deviance, 1, 
                            lower.tail = FALSE))
Reg<-Reg[which(Reg$plog10 >= 6),]
#Count number of populations where a variant is significant
NumPop <- as.data.frame(table(Reg$Chr,Reg$Pos))
NumPop <- NumPop[which(NumPop$Freq>1),]
names(NumPop) <- c('Chr','Pos', 'SigHits')
NumPop$Chr<-as.numeric(as.character(NumPop$Chr)) 
NumPop$Pos<-as.numeric(as.character(NumPop$Pos))

m <- merge(Reg,
           NumPop, 
           by = c('Chr','Pos'))
#count number of significant sites per population
n <- as.data.frame(table(m$Pop))
#put it all together for plotting
overLap <- unique(merge(m,n,
                        by.x = 'Population',
                        by.y = 'Var1'))
#include number of significant sites in titles
overLap$title = paste0(overLap$Pop,
                       ' : ', 
                       overLap$Freq)
#add asterix for Phase 1 pops
Phase1Title <- paste('*',
                     overLap[which(overLap$Pop 
                                   %in% phase1),]$title)
overLap[which(overLap$Pop
              %in% phase1),]$title <- Phase1Title
#set order by SigHits, Chromosome and Position
overLap <- overLap[with(overLap, 
                        order(-SigHits,
                              Chr,
                              Pos)),]
overLap$Pos <- factor(overLap$Pos, 
                      levels = unique(overLap$Pos))
#this is for keeping track of the number of variants
overLap$nrow <- seq(length = nrow(overLap))
#this is for discrete point size
overLap$Pcat <- 4
overLap[which(overLap$plog10 < 8),]$Pcat <- 2
overLap[which((overLap$plog10 > 8) & 
              (overLap$plog10 < 10)),]$Pcat <- 3
#this is for the labels of the xaxis
xaxis <- overLap[!duplicated(overLap[,c('SigHits')]),]
xaxis <- xaxis[which(xaxis$SigHits < 11),]
xaxis <- rbind(xaxis,tail(overLap, 1))

overLap

#save to file
saveRDS(overLap,
            paste0(DataPath,
                   'overLap.rds'))
```
```{r}
#load 1kgp 
kgp1= paste0(DataPath,
             '1kGP_genomeWide_SINGLEPOP_83.genotypes.txt')
kgp1.n=fread(paste0(DataPath,
                    '1kGP_genomeWide_SINGLEPOP_83.header'))
gt1<-unique(fread(kgp1,
                  col.names = colnames(kgp1.n)))
gt1.m<-melt(gt1, 
            id.vars = c('#CHROM','POS','ID'))
colnames(gt1.m) = c('CHROM','POS',
                    'ID','sample','kGT')

han1=paste0(DataPath,
            '90_Han_Chinese_SigPos.genotypes.txt')
han1.n=fread(paste0(DataPath,
                    '90_Han_Chinese_SigPos.header'))
gtHan <-unique(fread(han1,
                     col.names = colnames(han1.n)))
gtHan$'#CHROM' <- as.integer(as.character(gsub('chr','',gtHan$'#CHROM')))
gtHan.m<-melt(gtHan, 
              id.vars = c('#CHROM','POS','ID'))
colnames(gtHan.m)=c('CHROM','POS',
                    'ID','sample','HanGT')

#merge
HanKgpAF<-merge(gtHan.m, 
          gt1.m, 
          by = c('CHROM','POS',
                 'ID','sample'), 
          all.y=T)

HanKgpAF[is.na(HanKgpAF)] <- 0

#to get allele frequencies we group by POS and summarize GT
HanKgpAF <- HanKgpAF %>% 
  group_by(CHROM,POS) %>% 
  summarize(hanAF = sum(HanGT), 
            kgpAF = sum(kGT))

#remove sites missing in this subset of individuals
HanKgpAF <- HanKgpAF[which(HanKgpAF$kgpAF > 0),]

HanKgpAF

#save to file
saveRDS(HanKgpAF,
            paste0(DataPath,
                   'HanKgpAF.rds'))
```

```{r}
#read Genotypes
han1=fread(paste0(DataPath,
                  'SignificantSNPs_Han90_0.01_genotype.txt'))
HanNames=fread(paste0(DataPath,
                      'SignificantSNPs_Han90_COLNAMES.txt'),
               header=F)
names(han1) = as.character(unlist(HanNames))
#melt to long format
han = melt(han1, id=c('#CHROM','POS','ID'))
#fix chromosome names
han$CHROM = as.numeric(as.character(substring(han$'#CHROM',4)))
han$'#CHROM' <- NULL

#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
han <- han %>% 
  group_by(CHROM,POS,variable) %>% 
  summarize(han.gt = max(value))
#to get allele frequencies we group by POS and summarize GT
hanAF <- han %>% 
  group_by(CHROM,POS) %>% 
  summarize(hanAF = sum(han.gt))

#read Genotypes
kgp1 = fread(paste0(DataPath,
                    'SigVar_Pop_0.01_genotype.txt'))
ColNames = read.table(paste0(DataPath,
                             'ColNames.txt'),
                      header=F)
names(kgp1) = as.character(unlist(ColNames))
#melt to long format
kgp = melt(kgp1, id=c('CHROM','POS','ID'))
#keep only the 83 resequenced individuals
kgp = kgp[which(kgp$variable %in% han$variable),]

#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
kgp <- kgp %>% 
  group_by(CHROM,POS,variable) %>% 
  summarize(Kgp.gt = max(value))
#to get allele frequencies we group by POS and summarize GT
kgpAF <- kgp %>% 
  group_by(CHROM,POS) %>% 
  summarize(kgpAF = sum(Kgp.gt))
#keep only positions that are present in the 83 individuals
kgpAF <- kgpAF[which(kgpAF$kgpAF > 0),]

#merge 1kGP and han resequenced data.
combo <- merge(hanAF, kgpAF, by=c('CHROM','POS'), all.y=T)
#missing sites replaced with 0
combo[is.na(combo)] <- 0

##read Indels
Indels <- fread(paste0(DataPath,
                    'Reg_indels.POS'), 
                skip=1 ,col.names=c('CHROM','POS'))
Indels$Indel = 'yes'

INDEL = merge(Indels, combo, by=c('CHROM','POS'))
SNP = combo[which((combo$POS %in% INDEL$POS)==F),]

#save to file
saveRDS(INDEL,
            paste0(DataPath,
                   'INDEL_HanKgp.rds'))
#save to file
saveRDS(SNP,
            paste0(DataPath,
                   'SNP_HanKgp.rds'))
```

### 2019-03-15 rick/final/3_fit_model_compareModels.Rmd
update

@@
   trait_no_na <- trait[ind]
+
   b0 <- lreg(snp_no_na, LF_no_na) #coefficients from logreg 
@@
 
-plt <- merge(data_fit[-2], data_fit_PCA[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
-plt <- merge(plt, data_fit_Pop[-2], by=c('rsID'), suffixes = c('','Pop'))
+plt <- merge(data_fit[-2], 
+             data_fit_PCA[-2], 
+             by=c('rsID'), 
+             suffixes = c('LFA','PCA'))
+plt <- merge(plt, 
+             data_fit_Pop[-2], 
+             by=c('rsID'), 
+             suffixes = c('','Pop'))
 
-p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
+beta1<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
+beta2<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
+beta3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
+betas <- plot_grid(beta1,beta2,beta3,nrow=1)
+
+af1<-ggplot(plt, aes(1-num_zeros/2504, bLFA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='LFA')+ggtitle('Beta')+xlim(c(0,0.3))
+af2<-ggplot(plt, aes(1-num_zeros/2504, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='PCA')+ggtitle('Beta')+xlim(c(0,0.3))
+af3<-ggplot(plt, aes(1-num_zeros/2504, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='Pop')+ggtitle('Beta')+xlim(c(0,0.3))
+af <- plot_grid(af1,af2,af3,nrow=1)
+af
+ggsave('./Beta_AF.jpg', height=4, width=8)
+
+plt <- plt[which(plt$num_zeros < 2501),]
+
+beta1<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
+beta2<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
+beta3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
+betas2 <- plot_grid(beta1,beta2,beta3,nrow=1)
+
+af1<-ggplot(plt, aes(1-num_zeros/2504, bLFA))+geom_point(shape=1)+labs(x='AF', y='LFA')+ggtitle('Beta')+xlim(c(0,0.3))
+af2<-ggplot(plt, aes(1-num_zeros/2504, bPCA))+geom_point(shape=1)+labs(x='AF', y='PCA')+ggtitle('Beta')+xlim(c(0,0.3))
+af3<-ggplot(plt, aes(1-num_zeros/2504, b))+geom_point(shape=1)+labs(x='AF', y='Pop')+ggtitle('Beta')+xlim(c(0,0.3))
+af2 <- plot_grid(af1,af2,af3,nrow=1)
+af2
+ggsave('./Beta_AF2.jpg', height=4, width=8)
+
+makePlot <- function(x,y,xlab,ylab,title) {
+p <- ggplot(plt, aes(x, y)) + 
+  geom_abline(color='grey') +
+  geom_point(shape=1) +
+  labs(x=xlab, y=ylab) +
+  ggtitle(title)
+
+return(p)
+}
+
+p1 <- makePlot(plt$dev0LFA, 
+               plt$dev0PCA, 
+               'Null Deviance LFA', 
+               'Null Deviance PCA', 
+               'Null Model')
+p2 <- makePlot(plt$devdiffLFA, 
+               plt$devdiffPCA, 
+               'Deviance LFA', 
+               'Deviance PCA', 
+               'Deviance')
+
 p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
+
+p3<-ggplot(plt, aes(bLFA, bPCA, color=1-num_zeros/2504))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
 pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
@@
 p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
+p3<-ggplot(plt, aes(bLFA, b, color=1-num_zeros/2504))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
 pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
@@
 #plot diff
-plt <- merge(data_fit, data_fit_Pop_dich, by=c('rsID'), suffixes = c('LFA_Full','Pop_Dich'))
+plt <- merge(data_fit[-2], data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('LFA_Full','Pop_Dich'))
 

### 2019-03-18 PlottingScripts/ProcessPlotData.Rmd
update

@@
 
-#Joint Frequency Spectrum
+##GWAS
+```{r}
+#Load single population regression results
+Reg <- fread(
+  paste0(DataPath,
+         '1KGP.GenomeWide.Regression_JPT.csv'),
+  col.names = c('Chr', 'Pos',
+                'Population', 'deviance'),
+  colClasses = c('numeric', 'numeric',
+                 'character', 'numeric'))
+
+#Compute P values from deviance from null model
+Reg$plog10 <- -log10(pchisq(Reg$deviance,
+                            df = 1,
+                            lower.tail = FALSE))
+
+#split data into sig / not sig for plotting
+SigPos <- Reg[which(Reg$plog10 > 6), ]
+NotSig <- Reg[which(Reg$plog10 < 6), ]
+rm(Reg) #save memory
+
+#save to file
+saveRDS(SigPos,
+        paste0(DataPath,
+               'SigPos.rds'))
+#save to file
+saveRDS(NotSig,
+        paste0(DataPath,
+               'NotSig.rds'))
+```
+##Joint Frequency Spectrum
 ```{r}
 if(!file.exists(paste0(DataPath,
-                       'jointAF.txt'))) {
+               'jointAF.rds'))){
+
 #Load Allele Frequency files
@@
   col.names = c('Chr', 'Pos', 'N_ALLELES',
-                  'N_CHR', 'JPT_AF', 'JPT_MAF')
-  )
+                'N_CHR', 'JPT_AF', 'JPT_MAF'))
 AF.NAG <- paste0(DataPath,
@@
   col.names = c('Chr', 'Pos', 'N_ALLELES',
-                  'N_CHR', 'NAG_AF', 'NAG_MAF')
-  )
+                'N_CHR', 'NAG_AF', 'NAG_MAF'))
 
 #merge them together to get joint frequency spectrum
-  join <- merge(JPT[which(JPT$JPT_AF <= 1), ],
-                NAG[which(NAG$NAG_AF <= 1), ],
+join <- merge(JPT[which(JPT$JPT_AF <= 1),],
+              NAG[which(NAG$NAG_AF <= 1),],
               by = c('Chr', 'Pos'), all = TRUE)
@@
 #remove sites from AF that don't meet HWE
-  join <- join[!paste(join$CHROM, join$POS)
-               %in% c(paste(hwe$V1, hwe$V2)),]
+join <- join[!paste(join$Chr, join$Pos)
+             %in% c(paste(hwe$V1, hwe$V2)), ]
 
-  join <- join[, c('Chr', 'Pos', 'JPT_MAF',
-                   'JPT_AF', 'NAG_MAF', 'NAG_AF')]
+join <- join[, c('Chr', 'Pos', 'JPT_MAF', 'NAG_MAF')]
 #make numeric
 join <- sapply(join, as.numeric)
-  join <- join[complete.cases(join), ]
+join <- join[complete.cases(join),]
 
@@
 #save to file
-  write.table(join,
+saveRDS(join,
         paste0(DataPath,
-                     'jointAF.txt'))
+               'jointAF.rds'))
 }
@@
 random.pos <- NotSig[sample(nrow(NotSig), 
-                            nrow(sig.6)),
+                            nrow(sig.pos)),
                      c('Chr','Pos')]
@@
 
+##Data Quality Over Time
 ```{r}
@@
 ```
-
+##Overlap of Significant SNPs from Single Population test
 ```{r}
@@
 ```
+## Compare Han Resequenced to 1kGP
+##Single Population Test
 ```{r}
@@
 
+##Full Model
 ```{r}
@@
                    'SNP_HanKgp.rds'))
+
+SNP
+INDEL
 ```
+##Compare Over and Under Qual 30
+```{r}
+#read Genotypes
+GT = fread(paste0(DataPath,
+                   'SigVar_Pop_0.01_rsID.genotype.txt'))
+ColNames = read.table(paste0(DataPath,
+                   'ColNames_rsID.txt'), 
+                   header=F)
+
+names(GT) = as.character(unlist(ColNames))
+names(GT)[4] = 'rsID'
+GT$ID <- as.numeric(as.character(GT$ID))
+
+Sig.rsID <- fread(paste0(DataPath,
+                         'SignificantVariants.txt'), 
+                  col.names = c('Chr','Pos','rsID', 
+                                'ID', 'dev', 'log10_p'))
+##Compare over and under 30
+
+mGT = melt(unique(GT), 
+           id = c('Chr','Pos',
+                  'ID','rsID'))
+mGT = merge(unique(bas_dt[,
+                          c('sample',
+                            'Population',
+                            'avg_qual')]), 
+            mGT, 
+            by.x='sample',
+            by.y='variable')
+
+mGT = merge(mGT, 
+            Sig.rsID, 
+            by=c('Chr','Pos',
+                 'ID','rsID'))
+
+over30 = mGT[which(mGT$avg_qual > 30),]
+
+under30 = mGT[which(mGT$avg_qual < 30),]
+
+o30AF <- 
+		over30 %>% 
+		group_by(Chr,Pos,
+		         ID,rsID) %>% 
+		summarize(oAF = sum(value))
+
+u30AF <- 
+		under30 %>% 
+		group_by(Chr,Pos,
+		         ID,rsID) %>% 
+		summarize(uAF = sum(value))
+
+AF30 <- merge(u30AF, 
+              o30AF, 
+              by=c('Chr','Pos',
+                   'ID','rsID'))
+
+#save to file
+saveRDS(AF30,
+            paste0(DataPath,
+                   'AF30.rds'))
+```

### 2019-03-22 legacy.tex
working
Processed and Plotted New Results

@@
 \begin{tabular}{l l l r r r}
-  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)} & \multicolumn{1}{p{1cm}}{\centering Odds \\ ratio}     \\ \hline
- 28654678	& PLoS One	& rs201130852	& 5.05	& 73.18 & 0.65\\
- 	& 		& rs201761909	& 5.70	& 72.90 & 0.63\\
- 	& 		& rs200655768	& 6.53	& 69.28 & 0.66\\
- 	& 		& rs201255786	& 5.70	& 61.20 & 0.65\\
- 	& 		& rs80274284	& 6.00	& 57.76 & 0.70\\		
- 	& 		& rs184202621	& 5.52	& 53.72 & 0.68\\
-	& 		& rs200699422	& 5.30	& 6.33 & 0.87\\	
-	&		&			&		&		&\\
-23527680 &	\multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics}	& rs6057648	& 5.40	& 14.17 & 0.82\\  
-21386085 &	\multicolumn{1}{p{3cm}}{\raggedright Diabetes\\} &	rs301 &*10.52	 & 9.21 & 0.90\\
-28928442	& \multicolumn{1}{p{3cm}}{\raggedright Nature Communications}	& rs201471471	& 6.52	& 2.5 & 0.67\\
+  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
+28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
+	& 	&  	rs201130852 & 5.05 & 72.28\\
+	& 	&  	rs201255786 & 5.7 & 68.97\\
+	& 	&  	rs200655768 & 6.52 & 66.67\\
+	& 	&  	rs184202621 & 5.52 & 60.45\\
+	& 	& 	rs80274284 & 6 & 56.15\\
+	& 	& 	rs200699422 & 5.3 & 7.43\\
+23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
+28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
+28270201 & \multicolumn{1}{p{3cm}}{\raggedright Genome Medicine} & rs453755 & 7.52 & 5.29\\
+23023329 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs103294 & *15.3 & 4.32\\
+28334899 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs103294 & *29.3 & 4.32\\
+28240269 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs103294 & *72.7 & 4.32\\
+27863252 & \multicolumn{1}{p{3cm}}{\raggedright Cell\\} & rs3794738 & *13.15 & 3.73\\
+29534301 & \multicolumn{1}{p{3cm}}{\raggedright Hepatology\\} & rs9273062 & *9.7 & 3.36\\
+21386085 & \multicolumn{1}{p{3cm}}{\raggedright Diabetes\\}  & rs301 & *10.52 & 3.02\\
+26830138 & \multicolumn{1}{p{3cm}}{\raggedright Molecular Psychiatry} & rs77894924 & 6.7 & 2.77\\
+29617998 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs4963156 & *22.4 & 2.52\\
+28698626 & \multicolumn{1}{p{3cm}}{\raggedright Scientific Reports} & rs11015915 & 5.05 & 2.45\\
+26974007 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs3124998 & *8.05 & 2.33\\
+26634245 & \multicolumn{1}{p{3cm}}{\raggedright BMC Genetics} & rs451000 & 6 & 2.28\\
+ 	& 	& 	rs443874 & 5.3 & 2.26\\
+	& 	&	rs400942 & 6 & 2.2\\
+26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\
+
+25918132 & \multicolumn{1}{p{3cm}}{\raggedright Toxicological Sciences} & rs76780579 & 6 & 2.09\\
 
@@
 \end{tabular}
-\caption{A list of four recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
+\caption{A list of recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
 The variants reaching genome wide significance have a star ( * ).
@@
 \end{table}
+
+%& \multicolumn{1}{p{1cm}}{\centering Odds \\ ratio}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatSNPs.jpg}
         \label{fig:a}
@@
     	\center    
-        \includegraphics[width=90mm]{./Figures/NORepeat_SNPs_QQ.jpg}
+        \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatSNPs.jpg}
         \label{fig:b}    
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatIndels.jpg}
         \label{fig:a}
@@
     	\center    
-        \includegraphics[width=90mm]{./Figures/NORepeat_Indels_QQ.jpg}
+        \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatIndels.jpg}
         \label{fig:b}    
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatSNPs.jpg}
         \label{fig:a}
@@
     	\center    
-        \includegraphics[width=90mm]{./Figures/Repeat_SNPs_QQ.jpg}
+        \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatSNPs.jpg}
         \label{fig:b}    
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_Repeat_Indels.jpg}
+        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatIndels.jpg}
         \label{fig:a}
@@
     	\center    
-        \includegraphics[width=90mm]{./Figures/Repeat_Indels_QQ.jpg}
+        \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatIndels.jpg}
         \label{fig:b}    

### 2019-03-24 PlottingScripts/ProcessPlotData.Rmd
update

@@
 library(purrr)
+library(lfa)
 
@@
 ```
+##Extract Significant Positions from 1kGP Data
+```{bash}
+DataPath='/Users/luke/Documents/QualityPaper/ProcessedData'
+RawDataPath='/Users/luke/genomes/genomes/hg19/phase3'
+VCF='.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz'
+
+#GetHeader
+zgrep '^#' $RawDataPath/ALL.chr22$VCF > $DataPath/Significant_Positions.vcf
+
+#AddGenotypes
+for f in `seq 1 22`;
+do echo $f;
+bcftools view -R $DataPath/Significant_Positions.txt \
+$RawDataPath/ALL.chr${f}$VCF \
+| grep -v "^#" \
+>> $DataPath/Significant_Positions.vcf ; \
+done
+
+#bcftools doesn't take rsID
+#this deals with indels and other mulitallelic variants
+vcftools \
+--vcf $DataPath/Significant_Positions.vcf \
+--snps $DataPath/Significant_rsID.txt \
+--recode --recode-INFO-all \
+--out $DataPath/Significant_Positions
+
+/Users/luke/bin/plink_mac/plink \
+--vcf $DataPath/Significant_Positions.recode.vcf \
+--recode \
+--out $DataPath/Significant_Positions
+
+```
+```{r}
+data_bim <- fread(paste0(DataPath,
+                         'Significant_Positions.bim'), 
+                  col.names = c('Chr','rsID','X',
+                                'Pos','Ref','Alt')) %>%
+  select(rsID)
+
+data_fam <- fread(paste0(DataPath,
+                         'Significant_Positions.fam'), 
+                  col.names = c('familyID','sample',
+                                'father','mother',
+                                'sex','pheno')) %>%
+  select(sample)
+
+data_genotype <- read.bed(paste0(DataPath,'Significant_Positions'))
+rownames(data_genotype) <- data_bim$rsID
+colnames(data_genotype) <- data_fam$sample
+data_genotype[1:5,1:5]
+
+```
+
 ##Compare Over and Under Qual 30

### 2019-03-24 PlottingScripts/ProcessPlotData.Rmd
update

@@
 for f in `seq 1 22`;
-do echo $f;
-bcftools view -R $DataPath/Significant_Positions.txt \
+do bcftools view \
+-R $DataPath/Significant_Positions.txt \
 $RawDataPath/ALL.chr${f}$VCF \
@@
 --vcf $DataPath/Significant_Positions.recode.vcf \
---recode \
+--make-bed \
 --out $DataPath/Significant_Positions
@@
 ```
+##Compare Over and Under Qual 30
 ```{r}
@@
 
-```
-
-##Compare Over and Under Qual 30
-```{r}
-#read Genotypes
-GT = fread(paste0(DataPath,
-                   'SigVar_Pop_0.01_rsID.genotype.txt'))
-ColNames = read.table(paste0(DataPath,
-                   'ColNames_rsID.txt'), 
-                   header=F)
-
-names(GT) = as.character(unlist(ColNames))
-names(GT)[4] = 'rsID'
-GT$ID <- as.numeric(as.character(GT$ID))
-
-Sig.rsID <- fread(paste0(DataPath,
-                         'SignificantVariants.txt'), 
-                  col.names = c('Chr','Pos','rsID', 
-                                'ID', 'dev', 'log10_p'))
 ##Compare over and under 30
 
-mGT = melt(unique(GT), 
-           id = c('Chr','Pos',
-                  'ID','rsID'))
+mGT = melt(unique(data_genotype), 
+           varnames = c('rsID','sample'), 
+           value.name = 'Genotype')
+
 mGT = merge(unique(bas_dt[,
@@
             mGT, 
-            by.x='sample',
-            by.y='variable')
+            by = 'sample')
 
-mGT = merge(mGT, 
-            Sig.rsID, 
-            by=c('Chr','Pos',
-                 'ID','rsID'))
+over30 = unique(mGT[which(mGT$avg_qual > 30),])
 
-over30 = mGT[which(mGT$avg_qual > 30),]
-
-under30 = mGT[which(mGT$avg_qual < 30),]
+under30 = unique(mGT[which(mGT$avg_qual < 30),])
 
@@
 		over30 %>% 
-		group_by(Chr,Pos,
-		         ID,rsID) %>% 
-		summarize(oAF = sum(value))
+		group_by(rsID) %>% 
+		summarize(oAF = sum(Genotype))
 
@@
 		under30 %>% 
-		group_by(Chr,Pos,
-		         ID,rsID) %>% 
-		summarize(uAF = sum(value))
+		group_by(rsID) %>% 
+		summarize(uAF = sum(Genotype))
 
@@
               o30AF, 
-              by=c('Chr','Pos',
-                   'ID','rsID'))
+              by=c('rsID'))
 

### 2019-03-25 PlottingScripts/ProcessPlotData.Rmd
update

@@
 ```
-
-##Full Model
-```{r}
-#read Genotypes
-han1=fread(paste0(DataPath,
-                  'SignificantSNPs_Han90_0.01_genotype.txt'))
-HanNames=fread(paste0(DataPath,
-                      'SignificantSNPs_Han90_COLNAMES.txt'),
-               header=F)
-names(han1) = as.character(unlist(HanNames))
-#melt to long format
-han = melt(han1, id=c('#CHROM','POS','ID'))
-#fix chromosome names
-han$CHROM = as.numeric(as.character(substring(han$'#CHROM',4)))
-han$'#CHROM' <- NULL
-
-#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
-han <- han %>% 
-  group_by(CHROM,POS,variable) %>% 
-  summarize(han.gt = max(value))
-#to get allele frequencies we group by POS and summarize GT
-hanAF <- han %>% 
-  group_by(CHROM,POS) %>% 
-  summarize(hanAF = sum(han.gt))
-
-#read Genotypes
-kgp1 = fread(paste0(DataPath,
-                    'SigVar_Pop_0.01_genotype.txt'))
-ColNames = read.table(paste0(DataPath,
-                             'ColNames.txt'),
-                      header=F)
-names(kgp1) = as.character(unlist(ColNames))
-#melt to long format
-kgp = melt(kgp1, id=c('CHROM','POS','ID'))
-#keep only the 83 resequenced individuals
-kgp = kgp[which(kgp$variable %in% han$variable),]
-
-#to deal with INDELs we only count one instance of duplicate positions (i.e. pick max GT)
-kgp <- kgp %>% 
-  group_by(CHROM,POS,variable) %>% 
-  summarize(Kgp.gt = max(value))
-#to get allele frequencies we group by POS and summarize GT
-kgpAF <- kgp %>% 
-  group_by(CHROM,POS) %>% 
-  summarize(kgpAF = sum(Kgp.gt))
-#keep only positions that are present in the 83 individuals
-kgpAF <- kgpAF[which(kgpAF$kgpAF > 0),]
-
-#merge 1kGP and han resequenced data.
-combo <- merge(hanAF, kgpAF, by=c('CHROM','POS'), all.y=T)
-#missing sites replaced with 0
-combo[is.na(combo)] <- 0
-
-##read Indels
-Indels <- fread(paste0(DataPath,
-                    'Reg_indels.POS'), 
-                skip=1 ,col.names=c('CHROM','POS'))
-Indels$Indel = 'yes'
-
-INDEL = merge(Indels, combo, by=c('CHROM','POS'))
-SNP = combo[which((combo$POS %in% INDEL$POS)==F),]
-
-#save to file
-saveRDS(INDEL,
-            paste0(DataPath,
-                   'INDEL_HanKgp.rds'))
-#save to file
-saveRDS(SNP,
-            paste0(DataPath,
-                   'SNP_HanKgp.rds'))
-
-SNP
-INDEL
-```
 ##Extract Significant Positions from 1kGP Data
@@
 ```
+
+```{bash}
+DataPath='/Users/luke/Documents/QualityPaper/ProcessedData'
+RawDataPath='/Users/luke/genomes/genomes/Han90/90_Han_Chinese_SNP_gatkvqsr_20170424_chr'
+VCF='_validator.vcf.gz'
+
+#Han vcf has `chr` before chromosome number
+sed -e 's/^/chr/' \
+$DataPath/Significant_Positions.txt \
+> $DataPath/Significant_Positions_chr.txt
+
+#GetHeader
+zgrep '^#' $RawDataPath\22$VCF > $DataPath/Significant_Positions_Han.vcf
+
+#AddGenotypes
+for f in `seq 1 22`;
+do echo $f;
+bcftools view \
+-R $DataPath/Significant_Positions_chr.txt \
+$RawDataPath${f}$VCF \
+| grep -v "^#" \
+>> $DataPath/Significant_Positions_Han.vcf ; \
+done
+
+#bcftools doesn't take rsID
+#this deals with indels and other mulitallelic variants
+vcftools \
+--vcf $DataPath/Significant_Positions_Han.vcf \
+--snps $DataPath/Significant_rsID.txt \
+--recode --recode-INFO-all \
+--out $DataPath/Significant_Positions_Han
+
+/Users/luke/bin/plink_mac/plink \
+--vcf $DataPath/Significant_Positions_Han.recode.vcf \
+--make-bed \
+--out $DataPath/Significant_Positions_Han
+```
+
+##Full Model
+```{r}
+sharedID <- fread(paste0(DataPath,
+                          '83samples.txt'), 
+                  header = FALSE,
+                  col.names = 'sample')
+
+kgp_bim <- fread(paste0(DataPath,
+                         'Significant_Positions.bim'), 
+                  col.names = c('Chr','rsID','X',
+                                'Pos','Ref','Alt')) %>%
+  select(rsID)
+
+kgp_fam <- fread(paste0(DataPath,
+                         'Significant_Positions.fam'), 
+                  col.names = c('familyID','sample',
+                                'father','mother',
+                                'sex','pheno')) %>%
+  select(sample)
+
+kgp_genotype <- read.bed(paste0(DataPath,'Significant_Positions'))
+rownames(kgp_genotype) <- kgp_bim$rsID
+colnames(kgp_genotype) <- kgp_fam$sample
+#Only keep the 83 shared samples
+keep <- colnames(kgp_genotype)[(colnames(kgp_genotype) 
+                                 %in% sharedID$sample)]
+kgp_genotype <- kgp_genotype[,keep]
+kgp_genotype <- kgp_genotype[!duplicated(rownames(kgp_genotype)),]
+kgp_genotype[1:5,1:5]
+
+kgp <- melt(kgp_genotype, 
+           varnames = c('rsID','sample'), 
+           value.name = 'Genotype')
+kgp <- 
+		kgp %>% 
+		group_by(rsID) %>% 
+		summarize(kgp.AF = sum(Genotype))
+
+kgp <- kgp[which(kgp$kgp.AF > 0),]
+#read Han Genotypes
+han_bim <- fread(paste0(DataPath,
+                         'Significant_Positions_Han.bim'), 
+                  col.names = c('Chr','rsID','X',
+                                'Pos','Ref','Alt')) %>%
+  select(rsID)
+
+han_fam <- fread(paste0(DataPath,
+                         'Significant_Positions_Han.fam'), 
+                  col.names = c('familyID','sample',
+                                'father','mother',
+                                'sex','pheno')) %>%
+  select(sample)
+
+han_genotype <- read.bed(paste0(DataPath,'Significant_Positions_Han'))
+rownames(han_genotype) <- han_bim$rsID
+colnames(han_genotype) <- han_fam$sample
+#Only keep the 83 shared samples
+keep <- colnames(han_genotype)[(colnames(han_genotype) 
+                                 %in% sharedID$sample)]
+han_genotype <- han_genotype[,keep]
+han_genotype <- han_genotype[!duplicated(rownames(han_genotype)),]
+han_genotype[1:5,1:5]
+
+han = melt(han_genotype, 
+           varnames = c('rsID','sample'), 
+           value.name = 'Genotype')
+
+han <- 
+		han %>% 
+		group_by(rsID) %>% 
+		summarize(han.AF = sum(Genotype))
+
+#merge 1kGP and han resequenced data.
+combo <- merge(han, kgp, by='rsID', all.y=T)
+#missing sites replaced with 0
+combo[is.na(combo)] <- 0
+
+#save to file
+saveRDS(combo,
+            paste0(DataPath,
+                   'HanKgp_AF.rds'))
+
+combo
+```
+
 ##Compare Over and Under Qual 30

### 2019-03-25 PlottingScripts/ProcessPlotData.Rmd
update

@@
 author: 
-- name: Luke Anderson-Trocmé, 
+- name: Luke Anderson-Trocmé, Rick Farouni
   affiliation:
@@
 ```
+## Compare Models
+```{r}
+
+```
+
 ## Compare Han Resequenced to 1kGP
@@
 for f in `seq 1 22`;
-do echo $f;
-bcftools view \
+do bcftools view \
 -R $DataPath/Significant_Positions_chr.txt \
@@
 
-kgp <- melt(kgp_genotype, 
+kgp1 <- melt(kgp_genotype, 
            varnames = c('rsID','sample'), 
@@
 kgp <- 
-		kgp %>% 
+		kgp1 %>% 
 		group_by(rsID) %>% 
@@
 
-han = melt(han_genotype, 
+han1 = melt(han_genotype, 
            varnames = c('rsID','sample'), 
@@
 han <- 
-		han %>% 
+		han1 %>% 
 		group_by(rsID) %>% 

### 2019-03-25 legacy.tex
update

@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
-This method identifies a total of 11,416 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 24,390 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 2,007 & 8,009 \\  
-{INDEL} & 330 & 1,070 \\ \hline
+{SNP}  & 5,186 & 17,124 \\  
+{INDEL} & 509 & 1,571 \\ \hline
 \end{tabular}
@@
 SNPs and INDELs were also analyzed separately.
-The total number of 11,416 are statistically significantly associated to $Q$.}
+The total number of 24,390 are statistically significantly associated to $Q$.}
 \label{sigTable}
@@
 If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
+From the list of 24,390 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
 Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
@@
 We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 11,416 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
+Of the 24,390 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
 From this subset, only 613 of these variants are present in the resequenced data, the rest are missing entirely.
@@
 Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 11,416 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
 
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
+\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/Han_1kGP_SFS.jpg}
+\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_FullModel.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot shows  7,607 significant SNPs and INDELs identified in the reverse GWAS model including all populations.
-Only 613 of the variants are present in the high depth resequenced individuals.}
+This plot shows  11,946 significant SNPs and INDELs identified in the reverse GWAS model including all populations.
+Only 1,586 of the variants are present in the high depth resequenced individuals.}
 \label{90HanSFS_full}

### 2019-03-26 legacy.tex
style edits

--- 2019-03-08 legacy.tex
@@
 While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the spurious calls also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
+Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
 Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
@@
 The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
-Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
 The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
 
-Even though such batch effects clearly have the potential to confound analyses, the unrestricted availability of the 1kGP data lead to ubiquitous use making batch effects particularly insidious.   
+Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+%Even though such batch effects clearly have the potential to confound analyses, the unrestricted availability of the 1kGP 
+%data lead to ubiquitous use making batch effects particularly insidious.   
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
-Such genome-wide mutational signatures have been used as diagnostic tools for cancers \citep{Alexandrov2013,Shiraishi2015a}.
+Such genome-wide mutational signatures have been used as diagnostic tools for cancers (e.g., \cite{Alexandrov2013,Shiraishi2015a}).
 %This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
@@
 
- In particular, both studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+ In particular, the latter two studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
 This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
@@
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
-In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications.
+In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in a number of recent publications.
 
@@
 While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
-When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
@@
 The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. 
-Thus, sequences called from low-$Q$ data contain mutations that reproduce poorly across studies and exhibit a particular mutational signature.
+Thus, sequences called from low-$Q$ data contain variants that reproduce poorly across studies and exhibit a particular mutational signature.
 
@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a "reverse GWAS", in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
+Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
+The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
@@
 For other analyses where quality of imputation matters, identifying suspicious variants may be preferable.
-Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
+
+Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here} \sgcomment{Add this after all variants identified.}.
 
@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
-This method identifies a total of 11,416 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 11,416 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}): 
+To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 
@@
 \end{figure}
+To ensure that population structure was properly taken into account, we considered alternative models with principal components (PCs), Population membership or both as a covariate.
+These models were broadly consistent (See Figure \ref{CompareModel}).
 
-\subsection{Model selection}
-We considered multiple approaches to account for population structure in order to best capture the variants associated to $Q$.
-Each of these methods is somewhat similar to the genotype-conditional association test described by Song et al. \citep{song2015testing}.
-These methods model each genotype as a logistic function using principal components (PCs), Population membership or both as a covariate. 
-We found that these three models to be consistent except a small number of variants with higher associations with the PC model.
-We chose the more conservative approach using Population membership as the sole covariate (See Figure \ref{CompareModel}).
+%\subsection{Model selection}
+%We considered multiple approaches to account for population structure in order to best capture the variants associated to $Q$.
+%Each of these methods is somewhat similar to the genotype-conditional association test described by Song et al. \citep{song2015testing}.
+%These methods model each genotype as a logistic function using principal components (PCs), Population membership or both as a covariate. 
+%We found that these three models to be consistent except a small number of variants with higher associations with the PC model.
+%We chose the more conservative approach using Population membership as the sole covariate (See Figure \ref{CompareModel}).
 
@@
 \subsection{Cell line or technical artifact}
-%\luke{Do we keep HapMap? It may be worth keeping some of it to explain why 2/4 papers impute using HapMap and find the same variants.}
-To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
-Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
-The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
-If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
-Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
-This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
+%
+%To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
+%Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
+%The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
+%If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
+%From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
+%Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
+%This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
 
 In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
-After restricting the analysis to individuals in both datasets, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
-While this is more than 0.1\% of the sites, we suspect that some of these variants may be associated to $Q$ as a result of incorrect genotyping.
+After restricting the analysis to individuals present in both datasets, we find 296 suspicious variants polymorphic in 1kGP. Of those, 9 are present in the resequenced data \ref{90HanSFS}.
+While this provides strong validation that the variants are not cell-line specific mutations, it is more than the 0.1\% of the sites, some variants associated to $Q$ may be present in the population but have biased genotypes.
 

### 2019-03-26 legacy.tex

@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
-This method identifies a total of 11,416 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}): 
+
+This method identifies a total of 24,390 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.

### 2019-03-26 rick/final/3_fit_model_compareModels.Rmd

@@
 #chr <- args[3]
-chr <- '22_10000'
+chr <- '22'
 
@@
 
-```{r}
-#this runs the logistic regression
-lreg_dich <- function(y, X){
-
-  b <- glm2(cbind(y, 2-y) ~ -1 + X, family="binomial")$coef
-
-  return(b)
-
-}
-
-#modified from gcat
-fit_glm_dich <- function(..., LF, trait){
-  snp <- c(...)
-  snp <- as.numeric((snp==1) | (snp==2))
-  ind <- !is.na(snp) #logical index vector
-  snp_no_na <- snp[ind]
-  LF_no_na <- LF[ind,]
-  p0 <- rep(NA, length(snp))
-  p1 <- rep(NA, length(snp))
-  trait_no_na <- trait[ind]
-  b0 <- lreg_dich(snp_no_na, LF_no_na) #coefficients from logreg 
-  b1 <- lreg_dich(snp_no_na, cbind(LF_no_na, trait_no_na))
-  est0 <- .Call("mv", LF_no_na, b0)
-  est1 <- .Call("mv", cbind(LF_no_na, trait_no_na), b1)
-  p0[ind] <- exp(est0)/(1+exp(est0))
-  p1[ind] <- exp(est1)/(1+exp(est1))
-
-  dataset <-
-    tibble(snp=snp,
-           p0 = p0,
-           p1 = p1)
-
-  stat <-
-    dataset %>%
-    summarize( #cross entropy (log likelihood)
-      dev0 = -2 * sum(snp * log(p0) +
-                        (1 - snp) * log(1 - p0),
-                      na.rm = TRUE),
-      dev1 = -2 * sum(snp * log(p1) +
-                        (1 - snp) * log(1 - p1),
-                      na.rm = TRUE),
-      devdiff = dev0 - dev1,
-      dev_logp = -log10(pchisq(devdiff,
-                               1,
-                               lower.tail = FALSE))
-    ) %>% 
-    mutate(num_nas = sum(!ind),
-	   num_zeros= sum(snp_no_na ==0),
-           b=b1[length(b1)])
-
-  return(list(dataset=dataset, stat=stat))
-
-}
-```
-
 define functions
 ```{r}
-
 #this runs the logistic regression
@@
 
-data_fit_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= LF, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()    
 ```
@@
 	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, ".rds")))
+	paste0("fits_chrom_", chr, "_LFA.rds")))
 
@@
               file.path("./data/preprocessed/subsets", 
-                        paste0("fits_chrom_", chr, ".txt")),
+                        paste0("fits_chrom_", chr, "_LFA.txt")),
               quote = FALSE,
               row.names = FALSE)
-
-saveRDS(data_fit_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fits_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
+data_fit[,2]<-NULL
 ```
@@
 	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, ".rds")))
+	paste0("fits_chrom_", chr, "_PCA.rds")))
 
@@
               file.path("./data/preprocessed/subsets", 
-                        paste0("fitsPCA_chrom_", chr, ".txt")),
+                        paste0("fits_chrom_", chr, "_PCA.txt")),
               quote = FALSE,
               row.names = FALSE)
-
-saveRDS(data_fit_PCA_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_PCA_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fitsPCA_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
+data_fit_PCA[,2]<-NULL
 ```
@@
 
-data_fit_Pop_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= testModel, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()   
 ```
@@
 	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, ".rds")))
+	paste0("fits_chrom_", chr, "_Pop.rds")))
 
-write.table(data_fit[-2], 
+write.table(data_fit_Pop[-2], 
               file.path("./data/preprocessed/subsets", 
-                        paste0("fitsPOP_chrom_", chr, ".txt")),
+                        paste0("fits_chrom_", chr, "_Pop.txt")),
               quote = FALSE,
@@
 
-saveRDS(data_fit_Pop_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fitsPOP_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
+data_fit_Pop[,2]<-NULL
 ```

### 2019-03-26 rick/final/Luke_Cluster/3_fit_model.Rmd

--- 2019-03-14 rick/final/3_fit_model_compareModels.Rmd
@@
 
-#path <- args[1]
-path <- '/Users/luke/Documents/QualityPaper/rick/final/'
+path <- args[1]
+#path <- '/Users/luke/Documents/QualityPaper/rick/final/'
 setwd(path)
 
-#plinkPath <- args[2] 
-plinkPath <- '/Users/luke/genomes/genomes/hg19/plink/'
+plinkPath <- args[2] 
+#plinkPath <- '/Users/luke/genomes/genomes/hg19/plink/'
 
-#chr <- args[3]
-chr <- '22_10000'
+chr <- args[3]
+#chr <- '22_10000'
 
@@
 library(data.table)
-library(cowplot)
 options(future.globals.maxSize= 1000000000000)
@@
 ```
-
-```{r}
-#this runs the logistic regression
-lreg_dich <- function(y, X){
-
-  b <- glm2(cbind(y, 2-y) ~ -1 + X, family="binomial")$coef
-
-  return(b)
-
-}
-
-#modified from gcat
-fit_glm_dich <- function(..., LF, trait){
-  snp <- c(...)
-  snp <- as.numeric((snp==1) | (snp==2))
-  ind <- !is.na(snp) #logical index vector
-  snp_no_na <- snp[ind]
-  LF_no_na <- LF[ind,]
-  p0 <- rep(NA, length(snp))
-  p1 <- rep(NA, length(snp))
-  trait_no_na <- trait[ind]
-  b0 <- lreg_dich(snp_no_na, LF_no_na) #coefficients from logreg 
-  b1 <- lreg_dich(snp_no_na, cbind(LF_no_na, trait_no_na))
-  est0 <- .Call("mv", LF_no_na, b0)
-  est1 <- .Call("mv", cbind(LF_no_na, trait_no_na), b1)
-  p0[ind] <- exp(est0)/(1+exp(est0))
-  p1[ind] <- exp(est1)/(1+exp(est1))
-
-  dataset <-
-    tibble(snp=snp,
-           p0 = p0,
-           p1 = p1)
-
-  stat <-
-    dataset %>%
-    summarize( #cross entropy (log likelihood)
-      dev0 = -2 * sum(snp * log(p0) +
-                        (1 - snp) * log(1 - p0),
-                      na.rm = TRUE),
-      dev1 = -2 * sum(snp * log(p1) +
-                        (1 - snp) * log(1 - p1),
-                      na.rm = TRUE),
-      devdiff = dev0 - dev1,
-      dev_logp = -log10(pchisq(devdiff,
-                               1,
-                               lower.tail = FALSE))
-    ) %>% 
-    mutate(num_nas = sum(!ind),
-	   num_zeros= sum(snp_no_na ==0),
-           b=b1[length(b1)])
-
-  return(list(dataset=dataset, stat=stat))
-
-}
-```
-
 define functions
 ```{r}
-
 #this runs the logistic regression
@@
 ```
-#LFA
-```{r, warning=FALSE}
+
+```{r}
 tic("fitting model")
@@
 	unnest(stat)
+   toc()
 
-data_fit_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= LF, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()    
 ```
@@
               row.names = FALSE)
-
-saveRDS(data_fit_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fits_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
-```
-#PCA
-```{r, warning=FALSE}
-#load PCA
-library(data.table)
-samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
-PC <- as.matrix(samples[,c('PC1','PC2','PC3','PC4','PC5')])
-PC <- cbind(PC, 1)
-
-tic("fitting model")
-labels <- rownames(data_genotype)
-data_fit_PCA <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm,
-	LF= PC, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-
-data_fit_PCA_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= PC, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()    
-```
-
-```{r}
-
-saveRDS(data_fit_PCA, 
-	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_PCA[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("fitsPCA_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
-
-saveRDS(data_fit_PCA_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_PCA_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fitsPCA_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
-```
-#Population
-```{r, warning=FALSE}
-
-testModel <- model.matrix(data = bas_dt, object = ~Population)
-
-tic("fitting model")
-labels <- rownames(data_genotype)
-data_fit_Pop <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm,
-	LF= testModel, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()
-
-data_fit_Pop_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= testModel, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()   
-```
-```{r}
-
-saveRDS(data_fit_Pop, 
-	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, ".rds")))
-
-write.table(data_fit[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("fitsPOP_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
-
-saveRDS(data_fit_Pop_dich, 
-	file.path("./data/preprocessed/subsets",
-	paste0("dich_fits_chrom_", chr, ".rds")))
-
-write.table(data_fit_dich[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("dich_fitsPOP_chrom_", chr, ".txt")),
-              quote = FALSE,
-              row.names = FALSE)
 ```
@@
 ```
-
-```{r,fig.height = 10, fig.width = 10}
-
-plt <- merge(data_fit[-2], data_fit_PCA[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
-plt <- merge(plt, data_fit_Pop[-2], by=c('rsID'), suffixes = c('','Pop'))
-
-p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
-p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 10}
-
-plt <- merge(data_fit_dich[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
-plt <- merge(plt, data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('','Pop'))
-
-p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
-p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels_dich.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 10}
-
-LFAp <- merge(data_fit[-2], data_fit_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-PCAp <- merge(data_fit_PCA[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-Popp <- merge(data_fit_Pop[-2], data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-
-p1<-ggplot(LFAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model LFA')
-p2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(LFAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(PCAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model PCA')
-p2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(PCAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(Popp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model Pop')
-p2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(Popp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels_VSdich.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 6}
-#plot diff
-plt <- merge(data_fit, data_fit_Pop_dich, by=c('rsID'), suffixes = c('LFA_Full','Pop_Dich'))
-
-p1<-ggplot(plt, aes(dev0LFA_Full, dev0Pop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA Full', y='Pop Dich')+ggtitle('Null Model')+coord_flip()
-p2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance')+coord_flip()
-p2.2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))+coord_flip()
-p3<-ggplot(plt, aes(bLFA_Full, bPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Beta')+coord_flip()
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-pp1
-ggsave('./CompareModels_FullLFA_Popdich.jpg', height=10, width = 6)
-```

### 2019-03-26 PlottingScripts/ProcessPlotData.Rmd
update

@@
               (overLap$plog10 < 10)),]$Pcat <- 3
-#this is for the labels of the xaxis
-xaxis <- overLap[!duplicated(overLap[,c('SigHits')]),]
-xaxis <- xaxis[which(xaxis$SigHits < 11),]
-xaxis <- rbind(xaxis,tail(overLap, 1))
 

### 2019-03-26 rick/final/3_fit_model_compareModels.Rmd
update

@@
 #plinkPath <- args[2] 
-plinkPath <- '/Users/luke/genomes/genomes/hg19/plink/'
+plinkPath <- '/Users/luke/Documents/QualityPaper/ProcessedData' #'/Users/luke/genomes/genomes/hg19/plink/'
 
 #chr <- args[3]
-chr <- '22'
+chr <- 'PublishedVariants'
 
@@
 library(cowplot)
-options(future.globals.maxSize= 1000000000000)
+options(future.globals.maxSize= 25000000000)
 plan(multicore)
@@
 
+data_bim <- 
+  data_bim %>%  
+  rowid_to_column() %>% 
+  mutate(rsID= if_else(!duplicated(rsID),
+                       rsID,
+                       paste(rsID,
+                             rowid,
+                             sep="_" ))) %>% 
+  select(-rowid)
+
 rownames(data_genotype) <- data_bim$rsID
+
+#data_genotype <- head(data_genotype, 50000)
 ```
@@
 
-```
-
-```{r}
-
 saveRDS(data_fit, 
@@
               row.names = FALSE)
-data_fit[,2]<-NULL
+rm(data_fit)
 ```
@@
 
-data_fit_PCA_dich <- 
-	data_genotype %>%
-	as_tibble() %>%
-	transmute(
-	fit = future_pmap(., fit_glm_dich,
-	LF= PC, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
-	mutate(l= rep(c("data","stat"), n()/2))%>%
-	spread(l, fit)%>%
-	unnest(stat)
-   toc()    
-```
-
-```{r}
-
 saveRDS(data_fit_PCA, 
@@
               row.names = FALSE)
-data_fit_PCA[,2]<-NULL
+rm(data_fit_PCA)
 ```
@@
 
-```
-```{r}
-
 saveRDS(data_fit_Pop, 
@@
 
-data_fit_Pop[,2]<-NULL
+rm(data_fit_Pop)
 ```
@@
 ```
-
-```{r,fig.height = 10, fig.width = 10}
-
-plt <- merge(data_fit[-2], 
-             data_fit_PCA[-2], 
-             by=c('rsID'), 
-             suffixes = c('LFA','PCA'))
-plt <- merge(plt, 
-             data_fit_Pop[-2], 
-             by=c('rsID'), 
-             suffixes = c('','Pop'))
-
-beta1<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-beta2<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-beta3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-betas <- plot_grid(beta1,beta2,beta3,nrow=1)
-
-af1<-ggplot(plt, aes(1-num_zeros/2504, bLFA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='LFA')+ggtitle('Beta')+xlim(c(0,0.3))
-af2<-ggplot(plt, aes(1-num_zeros/2504, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='PCA')+ggtitle('Beta')+xlim(c(0,0.3))
-af3<-ggplot(plt, aes(1-num_zeros/2504, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='AF', y='Pop')+ggtitle('Beta')+xlim(c(0,0.3))
-af <- plot_grid(af1,af2,af3,nrow=1)
-af
-ggsave('./Beta_AF.jpg', height=4, width=8)
-
-plt <- plt[which(plt$num_zeros < 2501),]
-
-beta1<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-beta2<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-beta3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-betas2 <- plot_grid(beta1,beta2,beta3,nrow=1)
-
-af1<-ggplot(plt, aes(1-num_zeros/2504, bLFA))+geom_point(shape=1)+labs(x='AF', y='LFA')+ggtitle('Beta')+xlim(c(0,0.3))
-af2<-ggplot(plt, aes(1-num_zeros/2504, bPCA))+geom_point(shape=1)+labs(x='AF', y='PCA')+ggtitle('Beta')+xlim(c(0,0.3))
-af3<-ggplot(plt, aes(1-num_zeros/2504, b))+geom_point(shape=1)+labs(x='AF', y='Pop')+ggtitle('Beta')+xlim(c(0,0.3))
-af2 <- plot_grid(af1,af2,af3,nrow=1)
-af2
-ggsave('./Beta_AF2.jpg', height=4, width=8)
-
-makePlot <- function(x,y,xlab,ylab,title) {
-p <- ggplot(plt, aes(x, y)) + 
-  geom_abline(color='grey') +
-  geom_point(shape=1) +
-  labs(x=xlab, y=ylab) +
-  ggtitle(title)
-
-return(p)
-}
-
-p1 <- makePlot(plt$dev0LFA, 
-               plt$dev0PCA, 
-               'Null Deviance LFA', 
-               'Null Deviance PCA', 
-               'Null Model')
-p2 <- makePlot(plt$devdiffLFA, 
-               plt$devdiffPCA, 
-               'Deviance LFA', 
-               'Deviance PCA', 
-               'Deviance')
-
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-
-p3<-ggplot(plt, aes(bLFA, bPCA, color=1-num_zeros/2504))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, b, color=1-num_zeros/2504))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
-p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 10}
-
-plt <- merge(data_fit_dich[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('LFA','PCA'))
-plt <- merge(plt, data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('','Pop'))
-
-p1<-ggplot(plt, aes(dev0LFA, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='PCA')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0LFA, dev0))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiffLFA, devdiff))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(plt, aes(bLFA, b))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA', y='Pop')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(plt, aes(dev0, dev0PCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Null Model')
-p2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance')
-p2.2<-ggplot(plt, aes(devdiff, devdiffPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Deviance Zoom')+xlim(c(0,50))+ylim(c(0,50))
-p3<-ggplot(plt, aes(b, bPCA))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Pop', y='PCA')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels_dich.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 10}
-
-LFAp <- merge(data_fit[-2], data_fit_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-PCAp <- merge(data_fit_PCA[-2], data_fit_PCA_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-Popp <- merge(data_fit_Pop[-2], data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('Poly','Dich'))
-
-p1<-ggplot(LFAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model LFA')
-p2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(LFAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(LFAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(PCAp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model PCA')
-p2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(PCAp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(PCAp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp2 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-p1<-ggplot(Popp, aes(dev0Poly, dev0Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Null Model Pop')
-p2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance')
-p2.2<-ggplot(Popp, aes(devdiffPoly, devdiffDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))
-p3<-ggplot(Popp, aes(bPoly, bDich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='Poly', y='Dich')+ggtitle('Beta')
-pp3 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-plot_grid(pp1,pp2,pp3,ncol=3)
-ggsave('./CompareModels_VSdich.jpg', height=10, width = 10)
-
-```
-
-```{r,fig.height = 10, fig.width = 6}
-#plot diff
-plt <- merge(data_fit[-2], data_fit_Pop_dich[-2], by=c('rsID'), suffixes = c('LFA_Full','Pop_Dich'))
-
-p1<-ggplot(plt, aes(dev0LFA_Full, dev0Pop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA Full', y='Pop Dich')+ggtitle('Null Model')+coord_flip()
-p2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance')+coord_flip()
-p2.2<-ggplot(plt, aes(devdiffLFA_Full, devdiffPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Deviance Zoom')+xlim(c(0,30))+ylim(c(0,30))+coord_flip()
-p3<-ggplot(plt, aes(bLFA_Full, bPop_Dich))+geom_abline(color='grey')+geom_point(shape=1)+labs(x='LFA_Full', y='Pop_Dich')+ggtitle('Beta')+coord_flip()
-pp1 <- plot_grid(p1,p2,p2.2,p3,ncol=1)
-
-pp1
-ggsave('./CompareModels_FullLFA_Popdich.jpg', height=10, width = 6)
-```

### 2019-03-26 rick/final/3_fit_model_compareModels.Rmd
update

@@
 #chr <- args[3]
-chr <- 'PublishedVariants'
+plinkName <- 'PublishedVariants'
 
@@
 ```{r}
-data_genotype <- read.bed(paste0(plinkPath,'chr',chr))
+data_genotype <- read.bed(paste0(plinkPath,plinkName))
 

### 2019-03-26 rick/final/3_fit_model_compareModels.Rmd
update

@@
 #plinkPath <- args[2] 
-plinkPath <- '/Users/luke/Documents/QualityPaper/ProcessedData' #'/Users/luke/genomes/genomes/hg19/plink/'
-
+plinkPath <- '/Users/luke/Documents/QualityPaper/ProcessedData/' #'/Users/luke/genomes/genomes/hg19/plink/'
+DataPath <- '~/Documents/QualityPaper/ProcessedData/'
 #chr <- args[3]
-plinkName <- 'PublishedVariants'
+plinkName <- 'Significant_Positions'
 
@@
 ```{r}
-data_bim <- fread(paste0(plinkPath,'chr',chr,'.bim'), 
+data_bim <- fread(paste0(plinkPath,plinkName,'.bim'), 
                   col.names = c('Chr','rsID','X',
@@
 ```{r}
-fam_file <-fread(paste0(plinkPath,'chr',chr,'.fam'), 
+fam_file <-fread(paste0(plinkPath,plinkName,'.fam'), 
                   col.names=c('sample','a','b','c','d','e'))
@@
 saveRDS(data_fit, 
-	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, "_LFA.rds")))
+	file.path(DataPath,
+	paste0("fits_",plinkName, "_LFA.rds")))
 
 write.table(data_fit[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("fits_chrom_", chr, "_LFA.txt")),
+              file.path(DataPath, 
+                        paste0("fits_",plinkName,"_LFA.txt")),
               quote = FALSE,
@@
 saveRDS(data_fit_PCA, 
-	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, "_PCA.rds")))
+	file.path(DataPath,
+	paste0("fits_",plinkName, "_PCA.rds")))
 
 write.table(data_fit_PCA[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("fits_chrom_", chr, "_PCA.txt")),
+              file.path(DataPath, 
+                        paste0("fits_",plinkName, "_PCA.txt")),
               quote = FALSE,
@@
 
+#split data to run separately
+chunks <- split(as.data.frame(data_genotype), (as.numeric(1:nrow(data_genotype)) - 1) %/% 5000) 
+i<-1
+out = list()
+for(chunk in chunks){
+	print(i)
+
 tic("fitting model")
-labels <- rownames(data_genotype)
-data_fit_Pop <- 
-	data_genotype %>%
+labels <- rownames(chunk)
+out <- 
+	chunk %>%
 	as_tibble() %>%
@@
 	          fit = future_pmap(., fit_glm,
-	LF= testModel, 
-	trait=avg_qual_mean),
-		rsID=labels)  %>%
-	unnest(fit) %>%
+	          LF = testModel, 
+	          trait = avg_qual_mean),
+		        rsID = labels)  %>%
+	unnest(fit)%>%
 	mutate(l= rep(c("data","stat"), n()/2))%>%
@@
 
-saveRDS(data_fit_Pop, 
-	file.path("./data/preprocessed/subsets",
-	paste0("fits_chrom_", chr, "_Pop.rds")))
+   if(i==1){
+   write.table(unlist(out), 
+              file.path(DataPath, 
+                        paste0("fits_",plinkName, "_Pop.txt")),
+              quote = FALSE,
+              row.names = FALSE)}
+      if(i!=1){
+   write.table(unlist(out), 
+               append = TRUE,
+               file.path(DataPath, 
+                        paste0("fits_",plinkName, "_Pop.txt")),
+              quote = FALSE,
+              row.names = FALSE,
+              col.names = FALSE)}
 
-write.table(data_fit_Pop[-2], 
-              file.path("./data/preprocessed/subsets", 
-                        paste0("fits_chrom_", chr, "_Pop.txt")),
-              quote = FALSE,
-              row.names = FALSE)
+i <- i+1
+}
 
-rm(data_fit_Pop)
 ```

### 2019-03-27 PlottingScripts/ProcessPlotData.Rmd
update

@@
 ```
-## Compare Models
-```{r}
-
-```
 
@@
 --snps $DataPath/Significant_rsID.txt \
+--keep $DataPath/83samples.txt \
 --recode --recode-INFO-all \
@@
 
+#Get Allele Frequencies
+/Users/luke/bin/plink_mac/plink \
+--bfile $DataPath/Significant_Positions \
+--freq \
+--out $DataPath/Significant_Positions
 ```
 
+##Extract Significant Positions from Han Resequence Data
 ```{bash}
@@
 --out $DataPath/Significant_Positions_Han
-```
 
-##Full Model
-```{r}
-sharedID <- fread(paste0(DataPath,
-                          '83samples.txt'), 
-                  header = FALSE,
-                  col.names = 'sample')
-
-kgp_bim <- fread(paste0(DataPath,
-                         'Significant_Positions.bim'), 
-                  col.names = c('Chr','rsID','X',
-                                'Pos','Ref','Alt')) %>%
-  select(rsID)
-
-kgp_fam <- fread(paste0(DataPath,
-                         'Significant_Positions.fam'), 
-                  col.names = c('familyID','sample',
-                                'father','mother',
-                                'sex','pheno')) %>%
-  select(sample)
-
-kgp_genotype <- read.bed(paste0(DataPath,'Significant_Positions'))
-rownames(kgp_genotype) <- kgp_bim$rsID
-colnames(kgp_genotype) <- kgp_fam$sample
-#Only keep the 83 shared samples
-keep <- colnames(kgp_genotype)[(colnames(kgp_genotype) 
-                                 %in% sharedID$sample)]
-kgp_genotype <- kgp_genotype[,keep]
-kgp_genotype <- kgp_genotype[!duplicated(rownames(kgp_genotype)),]
-kgp_genotype[1:5,1:5]
-
-kgp1 <- melt(kgp_genotype, 
-           varnames = c('rsID','sample'), 
-           value.name = 'Genotype')
-kgp <- 
-		kgp1 %>% 
-		group_by(rsID) %>% 
-		summarize(kgp.AF = sum(Genotype))
-
-kgp <- kgp[which(kgp$kgp.AF > 0),]
-#read Han Genotypes
-han_bim <- fread(paste0(DataPath,
-                         'Significant_Positions_Han.bim'), 
-                  col.names = c('Chr','rsID','X',
-                                'Pos','Ref','Alt')) %>%
-  select(rsID)
-
-han_fam <- fread(paste0(DataPath,
-                         'Significant_Positions_Han.fam'), 
-                  col.names = c('familyID','sample',
-                                'father','mother',
-                                'sex','pheno')) %>%
-  select(sample)
-
-han_genotype <- read.bed(paste0(DataPath,'Significant_Positions_Han'))
-rownames(han_genotype) <- han_bim$rsID
-colnames(han_genotype) <- han_fam$sample
-#Only keep the 83 shared samples
-keep <- colnames(han_genotype)[(colnames(han_genotype) 
-                                 %in% sharedID$sample)]
-han_genotype <- han_genotype[,keep]
-han_genotype <- han_genotype[!duplicated(rownames(han_genotype)),]
-han_genotype[1:5,1:5]
-
-han1 = melt(han_genotype, 
-           varnames = c('rsID','sample'), 
-           value.name = 'Genotype')
-
-han <- 
-		han1 %>% 
-		group_by(rsID) %>% 
-		summarize(han.AF = sum(Genotype))
-
-#merge 1kGP and han resequenced data.
-combo <- merge(han, kgp, by='rsID', all.y=T)
-#missing sites replaced with 0
-combo[is.na(combo)] <- 0
-
-#save to file
-saveRDS(combo,
-            paste0(DataPath,
-                   'HanKgp_AF.rds'))
-
-combo
+#Get Allele Frequencies
+/Users/luke/bin/plink_mac/plink \
+--bfile $DataPath/Significant_Positions_Han \
+--freq \
+--out $DataPath/Significant_Positions_Han
 ```
@@
 ```
+
+##OMNI CHIP
+```{r}
+omni <- fread(paste0(DataPath,
+                     'HumanOmni2-5-8-v1-2-A-b138-rsIDs.txt'),
+              header = T)
+omni <- merge(omni, 
+              plt, 
+              by.x='RsID', 
+              by.y='rsID')
+omni <- omni[which(omni$RsID != '.'),]
+
+#save to file
+saveRDS(omni,
+            paste0(DataPath,
+                   'omni.rds'))
+```

### 2019-03-27 legacy.tex
update

@@
 
-Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here} \sgcomment{Add this after all variants identified.}.
-
 \subsection{Identifying suspicious variants in the 1kGP}
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
+Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/dev_Pop_PC_PopPC.jpg}
-\caption{Comparison of three logistic regression models for testing association to $Q$.
-These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or both (PopPC) as a covariate.
-In these plots we are comparing the deviance from the null model in 50,000 variants from chromosome 22.}  
-\label{CompareModel}
-\end{figure}
-
 \subsection{Cell line or technical artifact}
@@
 We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 24,390 suspicious variants identified globally, only 7,607 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
-From this subset, only 613 of these variants are present in the resequenced data, the rest are missing entirely.
-Figure \ref{90HanSFS_full} shows a weak correlation to allele frequency for the variants present in the resequenced data.
-Whereas the vast majority of the SNPs are absent in resequenced data, a high proportion of INDELs are fixed in the resequenced data.
+Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
+From this subset, only 1,637 of these variants are present in the resequenced data, the rest are missing entirely.
+Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in the resequenced data.
 A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are real but genotyped incorrectly.
@@
 
-We found that 219 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
+We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
 This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
@@
                       & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 8,393,467 & 26,241,186   \\  
-{INDEL} &  756,889  & 2,309,280 \\ \hline
+{SNP}  & 6,312,620 & 19,846,786   \\  
+{INDEL} &  586,342  & 1,770,315 \\ \hline
 \end{tabular}
@@
 \section{Supplementary Figures}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/fits_Significant_Positions_CompareModels.jpg}
+\caption{Comparison of three logistic regression models for testing association to $Q$.
+These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
+In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model.}  
+\label{CompareModel}
+\end{figure}
 
@@
 \begin{figure}[h]
+\centering
+\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
+This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
+Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
+Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
+\label{90HanSFS}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_FullModel.jpg}
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequenced data. 
+This plot shows  12,086 significant variants identified in the reverse GWAS model including all populations.
+ Only 1,637 of the variants are present in the high depth resequenced individuals.}  
+\label{90HanSFS_full}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of suspicious variants for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.}
+\label{90HanSFS_full}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\hsize,keepaspectratio]{./Figures/Omni_AF.jpg}
+\caption{We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip. 
+This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
+Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.}
+\label{HapMap_GnomAD}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
-Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
-Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
-\label{90HanSFS}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot shows  11,946 significant SNPs and INDELs identified in the reverse GWAS model including all populations.
-Only 1,586 of the variants are present in the high depth resequenced individuals.}
-\label{90HanSFS_full}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of suspicious variants for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.}
-\label{90HanSFS_full}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap.jpg}
-\caption{Frequencies of $Q$-associated variants polymorphic in both 1000G and either HapMap or GnomAD.
-\sgcomment{Change title, check targeted  sequencing region.
-Specify proportion of SNPs not included} The blue and green dots in the HapMap plot are the variants reported in \citep{Mandage2017} and \citep{Kraja2011} respectively.
-There are 17 such variants in HapMap and 186 in the GnomAD.}
-\label{HapMap_GnomAD}
-\end{figure}
-
 \end{document}

### 2019-03-27 legacy.tex
update

@@
 Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for some biological traits.
-\todo{UPDATE LOG ODDDS}
+
 \begin{table}[h]
@@
 
-%& \multicolumn{1}{p{1cm}}{\centering Odds \\ ratio}
-
-%\begin{figure}
-%\includegraphics[width=15cm,keepaspectratio]{./Figures/PublishedSNPs.jpg}
-%\caption{For each of the suspicious variants identified in the GWAS catalogue a plot of the average quality of mapped bases $Q$ compared to the predicted value from logistic regression.
-The strong correlation of $Q$ and the predicted value indicates that $Q$ is strongly associated to genotype for these variants.
-A negative correlation indicates that individuals with low $Q$ have higher likelihood of having a variant, whereas a positive correlation indicates that these individuals have a lower likelihood of having a variant at this position.}
-%\label{PubSig}
-%\end{figure}
 \section{Discussion}
@@
 
-For the reverse GWAS, the only filtration used was the application of an allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) as well as a strict removal of sites with any missing data resulting in a total of 37,700,882 variants ($S$) included in the test (see Table \ref{totTable}).
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons)  resulting in a total of 28,516,063 variants ($S$) included in the test (see Table \ref{totTable}).
 We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.

### 2019-03-29 legacy.tex
update

@@
 
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure \citep{song2015testing}.
 
@@
 
-\begin{table}[h]
+\begin{table}
 \centering
@@
 The variants reaching genome wide significance have a star ( * ).
-Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of -$\log_{10} p > 2$ are deemed significant.
+Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of $ p > -\log_{10}(\alpha)$ are deemed statistically significant.
 }

### 2019-04-01 legacy.tex
update

@@
 
+\subsection{Testing the association of quality to genotype}
+When conducting a statistical analysis of population genetics data, 
+an essential consideration is  to adjust or account for population structure (i.e.
+the heterogeneity in biallelic frequencies observed in individuals).
+Examples of models that widely used to account population structure include the Balding-Nichols model (Balding  Nichols, 1995) \luke{add citation}, and the Pritchard- Stephens-Donnelly model (Pritchard et al., 2000) \luke{add citation}.
+These and several other simlilar models used in GWAS studies can be understood in terms of the following matrix factorization.
+\begin{align}
+\mathbf{L }= \mathbf{AH}
+\end{align} 
+where the $i$'th column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
+When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
+\begin{align} \label{lfa}
+y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
+\end{align} 
+for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
+
+To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Aassociation Test  (GCAT) proposed by ....
+\luke{add citation} The GCAT is a reverse regression approach that assumes the following model.
+\begin{align}\label{gcat}
+ y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
+\end{align} 
+for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$.
+The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \luke{add citation} and plugged in the model.
+We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs.
+The matrix \luke{Describe matrix}
+
+Under the null hypothesis the slope coefficient is zero $\beta_s=0$ and Model \ref{gcat} reduces to Model \ref{lfa}.
+$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$.
+To test the null hypothesis, the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model.
+The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
+
+We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
+Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+
 \subsection{Genome wide association study}

### 2019-04-01 legacy.tex
extra space removed

--- 2019-03-27 legacy.tex
@@
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
-The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome , highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
+The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome, highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 we're considered.

### 2019-04-01 legacy.tex
update

@@
 Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
-%Even though such batch effects clearly have the potential to confound analyses, the unrestricted availability of the 1kGP 
-%data lead to ubiquitous use making batch effects particularly insidious.   
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 Such genome-wide mutational signatures have been used as diagnostic tools for cancers (e.g., \cite{Alexandrov2013,Shiraishi2015a}).
-%This means that the genomic context of a mutation can be used to gain insight into the cause of this mutation \cite{Pleasance2010}.
-%While certain mutagens have very distinct mutational signatures, others are more subtle and difficult to identify because they tend to occur less frequently and with less genomic specificity \cite{Pleasance2010}.
-%Large cohorts have recently become available to increase statistical power to resolve some of these fine scale mutational signatures.
 
@@
 
-%\subsection{Model selection}
-%We considered multiple approaches to account for population structure in order to best capture the variants associated to $Q$.
-%Each of these methods is somewhat similar to the genotype-conditional association test described by Song et al. \citep{song2015testing}.
-%These methods model each genotype as a logistic function using principal components (PCs), Population membership or both as a covariate. 
-%We found that these three models to be consistent except a small number of variants with higher associations with the PC model.
-%We chose the more conservative approach using Population membership as the sole covariate (See Figure \ref{CompareModel}).
-
 \subsection{Cell line or technical artifact}
-%
-%To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
-%Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
-%The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
-%If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
-%From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
-%Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
-%This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.
 
@@
 We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified eight recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
+The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
 
-One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
+Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
+One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence data \citep{yucesoy2015genome, ellinghaus2016analysis}
 
-All four papers used strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+All papers used a variety of strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
@@
   {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
-28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
-	& 	&  	rs201130852 & 5.05 & 72.28\\
-	& 	&  	rs201255786 & 5.7 & 68.97\\
-	& 	&  	rs200655768 & 6.52 & 66.67\\
-	& 	&  	rs184202621 & 5.52 & 60.45\\
-	& 	& 	rs80274284 & 6 & 56.15\\
-	& 	& 	rs200699422 & 5.3 & 7.43\\
-23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
-28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
+28654678 & PLoS One & *rs201761909 & 5.7 & 78.11\\
+	& 	&  	*rs201130852 & 5.05 & 72.28\\
+	& 	&  	*rs201255786 & 5.7 & 68.97\\
+	& 	&  	*rs200655768 & 6.52 & 66.67\\
+	& 	&  	*rs184202621 & 5.52 & 60.45\\
+	& 	& 	*rs80274284 & 6 & 56.15\\
+	& 	& 	*rs200699422 & 5.3 & 7.43\\
+23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & *rs6057648 & 5.4 & 20.5\\
+28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & *rs201471471 & 6.52 & 7.87\\
 28270201 & \multicolumn{1}{p{3cm}}{\raggedright Genome Medicine} & rs453755 & 7.52 & 5.29\\
@@
 	& 	&	rs400942 & 6 & 2.2\\
-26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\
+26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & *rs60136336 & 5.7 & 2.25\\
 
@@
 the heterogeneity in biallelic frequencies observed in individuals).
-Examples of models that widely used to account population structure include the Balding-Nichols model (Balding  Nichols, 1995) \luke{add citation}, and the Pritchard- Stephens-Donnelly model (Pritchard et al., 2000) \luke{add citation}.
+Examples of models that widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
 These and several other simlilar models used in GWAS studies can be understood in terms of the following matrix factorization.
@@
 
-To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Aassociation Test  (GCAT) proposed by ....
-\luke{add citation} The GCAT is a reverse regression approach that assumes the following model.
+To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Aassociation Test  (GCAT) proposed by \citep{song2015testing} The GCAT is a reverse regression approach that assumes the following model.
 \begin{align}\label{gcat}
@@
 for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$.
-The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \luke{add citation} and plugged in the model.
+The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model.
 We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs.

### 2019-04-02 legacy.tex
clarifications in conclusion and methods

@@
 \newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
-
+\newcommand{\logit}{\operatorname{logit}}
 %\newcommand{\qq}{m}
@@
 
-The state of the art imputation servers use a combination of many databases including some that are not freely available.
+State of the art imputation servers use a combination of many databases including some that are not freely available.
 From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
@@
 \subsection{Technical Artifact}
-The variants identified in this study are likely to be technical artifacts from legacy technology.
+The variants identified in this study are likely to be technical artifacts from legacy technologies.
 Different sequencing technologies will have different error profiles. 
-A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011}, with, for instance, different patterns of quality decrease along reads (known as B-tails).
+A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011} with, for instance, different patterns of quality decrease along reads.
+%(known as B-tails).
 Differences in read quality and error profiles in turn require different calling pipelines.
 
-We were not able to pinpoint the precise technical source of the discrepancy. 
-Further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal. 
-For instance, our test was limited by the genome wide resolution of the quality per sample.
-Data quality at a SNP-wise resolution would be a better metric to use, but unfortunately was not available.
+To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kgp.
+Given the progress in sequencing and calling that occurred since the early phases of the 1kgp (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
 
-If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer being actively introduced in recent sequence data.
+%For instance, our test was limited by the genome wide resolution of the quality per sample.
+%Data quality at a SNP-wise resolution would be a better metric to use, but unfortunately was not available.
+
+%If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer 
 However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes.
-Our model only reveals association to quality; some variants may be false positives in cases where only individuals with low $Q$ scores have the variant and other variants may be false negatives, in cases where individuals with low $Q$ appear to have lower calling rates for particular.
+The significant association of a variant with a quality metric is not in itself an indication that the variant is spurious.
+It may simply be subject to a weak but significant calling bias --- not an extraordinary surprise in a low-coverage dataset with a complex, joint sample variant calling pipeline.
+It may nevertheless be prudent to discard such variants from analyses that may be sensitive to calling biases, such as mutation profiles and allele frequency estimates.
+The imputation of such variants can also lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses.
 
-The imputation of incorrect variants can lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses as well as impact population specific allele frequency estimates.
-These variants do have a sizeable effect on population analyses that compare populations based on mutational profiles. 
-Even though we expect that present-day technologies will be less sensitive to such artefacts, if is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
+%Even though we expect that present-day technologies will be less sensitive to such artefacts, it is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
 
-For \textit{conventional} GWAS, we do not see evidence that these contaminated hits impact the conclusions made in each of these publications.
-However, considering many polygenic risk score estimates use the association scores from multiple sources, the inclusion of these variants could possibly impact downstream analyses.
+%For conventional GWAS, we do not see evidence that these contaminated hits impact the conclusions made in each of these publications.
+%However, considering many polygenic risk score estimates use the association scores from multiple sources, the inclusion of these variants could possibly impact downstream analyses.
 
@@
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
-In order to avoid spurious results, the most conservative approach would be to remove all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signal.
-Another approach is to remove all the sites that reach a significance below $ p < 10^{-2}$ in the reverse GWAS (See supplementary material for lists of variants statistically associated to $Q$).
-We also recommend the resequencing of samples below an acceptable $Q$ threshold to render them usable.
+In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
+We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{}).
+Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p < 10^{-2}$ in the reverse GWAS.
+(The supplementary material contains a list of variants statistically associated to $Q$).
 
@@
 On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
-The genome wide simple logistic regression analysis approach is straightforward, and should probably be a standard in a variety of -omics studies.
+The genome-wide simple logistic regression analysis approach is straightforward, and should probably be a standard in a variety of -omics studies.
+The logistic factor analysis is more computationally demanding and produces robust results.
+Both approaches produce comparable results.
 
@@
 However, cleaning up data is also important to avoid generating spurious results. 
-The present findings suggest that a substantial fraction of the final release of the 1kGP project is overdue for removal or re-sequencing.
+The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for removal or re-sequencing.
 
-We have not identified the precise mechanism through which the technical artifact was introduced.
+%We have not identified the precise mechanism through which the technical artifact was introduced.
 Given that modern platforms seem to have resolved this technical issue, but that recent studies continue to be affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while a sequencing forensics investigation can take place.
@@
 \subsection{Testing the association of quality to genotype}
-When conducting a statistical analysis of population genetics data, 
-an essential consideration is  to adjust or account for population structure (i.e.
-the heterogeneity in biallelic frequencies observed in individuals).
-Examples of models that widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
-These and several other simlilar models used in GWAS studies can be understood in terms of the following matrix factorization.
-\begin{align}
-\mathbf{L }= \mathbf{AH}
-\end{align} 
-where the $i$'th column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
-When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
+When conducting a statistical analysis of population genetics data, an essential consideration is  to adjust or account for population structure.
+In a typical GWAS, we are interested in modelling the phenotype as a function of the genotype.
+Here we have the opposite situation, where the quantitative variable ($Q$) is used as an explanatory variable.
+So we consider models where the genotype $y$ is a function of an expected frequency $\pi_{si},$ based on population structure, and $Q$.
+The null model is
 \begin{align} \label{lfa}
@@
 \end{align} 
-for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
+The expected frequency for a SNP $s$ and individual $i$ can be estimated using principal component analysis, categorical population labels, or logistic factor analysis \citep{song2015testing}.
+The alternative model then takes in $Q$ as a covariate:
 
-To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Aassociation Test  (GCAT) proposed by \citep{song2015testing} The GCAT is a reverse regression approach that assumes the following model.
 \begin{align}\label{gcat}
- y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
+ y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg)
 \end{align}
-for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$.
+
+%Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
+These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization.
+%\begin{align}
+%\mathbf{L }= \mathbf{AH}
+%\end{align} 
+%where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
+When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
+%\begin{align} \label{lfa}
+%y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
+%\end{align} 
+%for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
+%
+%To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing} The GCAT is a regression approach that assumes the following model.
+%\begin{align}\label{gcat}
+% y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
+%\end{align} 
+%for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$.
 The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model.
@@
 
-Under the null hypothesis the slope coefficient is zero $\beta_s=0$ and Model \ref{gcat} reduces to Model \ref{lfa}.
+Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}.
 $\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$.
-To test the null hypothesis, the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model.
+To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model.
 The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
@@
 We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
-Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
+Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
 By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}

### 2019-04-03 legacy.tex
edits and suggestions

@@
 Recent reports have identified differences in the mutational spectra across human populations.
-While some of these reports have been replicated in other cohorts, most have been reported in the 1000 Genomes Project (1kGP) data.
+While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
 While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
@@
 The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012}, leading to heterogeneous sample preparations and data quality.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012} \sgcomment{also cite phase 3}, leading to heterogeneous sample preparations and data quality.
 
@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure \citep{song2015testing}.
+\sgcomment{We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates.
+}
 
@@
 After restricting the analysis to individuals present in both datasets, we find 296 suspicious variants polymorphic in 1kGP. Of those, 9 are present in the resequenced data \ref{90HanSFS}.
-While this provides strong validation that the variants are not cell-line specific mutations, it is more than the 0.1\% of the sites, some variants associated to $Q$ may be present in the population but have biased genotypes.
+While this provides strong validation that the variants are not cell-line specific mutations, it is more than the 0.1\% of the sites: some variants that are associated to $Q$ may be present in the population but have biased genotypes.
 
-We did a similar analysis using the variants identified in the \textit{5PC} model present in the 83 resequenced individuals.
-Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data. (See Figure \ref{90HanSFS_full})
+We did a similar analysis using the variants identified in the \textit{5PC} \sgcomment{I don't think that this has been described?
+I also don't understand why the 5pc approach has so many more variants than the logistic factor analysis above} model present in the 83 resequenced individuals.
+Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data (See Figure \ref{90HanSFS_full}).
 From this subset, only 1,637 of these variants are present in the resequenced data, the rest are missing entirely.
 Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in the resequenced data.
-A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are real but genotyped incorrectly.
+A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are present in the cell lines real but exhibit genotyping bias.
 
@@
 \subsubsection{Microarrays}
-
+\sgcomment{It seems like this paragraph is out of place and would fit better in the validation section, above? }
 We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
@@
 \subsubsection{GWAS}
+\sgcomment{"these dubious", below, refers to all dubious positions rather than only the ones on the chips, right? Even more reason to move the previous paragraph. }
 
-We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biologic al trait, even though there is no particular reason for these variants to be associated with phenotypes.
+We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
@@
 Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence data \citep{yucesoy2015genome, ellinghaus2016analysis}
+One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence data \citep{yucesoy2015genome, ellinghaus2016analysis}.
 
-All papers used a variety of strict quality thresholds, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+All papers used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
-Despite using state-of-the-art quality controls, these erroneous variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for some biological traits.
+Despite using state-of-the-art quality controls, these erroneous \sgcomment{we don't know that all or even most are erroneous} variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.
+
+These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
+To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.
+\sgcomment{you know, we could simply compare the relative effect size to predicted frequency $\pi$}
 

### 2019-04-03 legacy.tex
update : greyed out text and minor text changes

--- 2019-04-02 legacy.tex
@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
+Here we perform a ``reverse GWAS'' or Genotype-Conditional Association Test, in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
@@
   {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
-28654678 & PLoS One & *rs201761909 & 5.7 & 78.11\\
-	& 	&  	*rs201130852 & 5.05 & 72.28\\
-	& 	&  	*rs201255786 & 5.7 & 68.97\\
-	& 	&  	*rs200655768 & 6.52 & 66.67\\
-	& 	&  	*rs184202621 & 5.52 & 60.45\\
-	& 	& 	*rs80274284 & 6 & 56.15\\
-	& 	& 	*rs200699422 & 5.3 & 7.43\\
-23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & *rs6057648 & 5.4 & 20.5\\
-28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & *rs201471471 & 6.52 & 7.87\\
-28270201 & \multicolumn{1}{p{3cm}}{\raggedright Genome Medicine} & rs453755 & 7.52 & 5.29\\
-23023329 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs103294 & *15.3 & 4.32\\
-28334899 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs103294 & *29.3 & 4.32\\
-28240269 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs103294 & *72.7 & 4.32\\
-27863252 & \multicolumn{1}{p{3cm}}{\raggedright Cell\\} & rs3794738 & *13.15 & 3.73\\
-29534301 & \multicolumn{1}{p{3cm}}{\raggedright Hepatology\\} & rs9273062 & *9.7 & 3.36\\
-21386085 & \multicolumn{1}{p{3cm}}{\raggedright Diabetes\\}  & rs301 & *10.52 & 3.02\\
-26830138 & \multicolumn{1}{p{3cm}}{\raggedright Molecular Psychiatry} & rs77894924 & 6.7 & 2.77\\
-29617998 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs4963156 & *22.4 & 2.52\\
-28698626 & \multicolumn{1}{p{3cm}}{\raggedright Scientific Reports} & rs11015915 & 5.05 & 2.45\\
-26974007 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs3124998 & *8.05 & 2.33\\
-26634245 & \multicolumn{1}{p{3cm}}{\raggedright BMC Genetics} & rs451000 & 6 & 2.28\\
- 	& 	& 	rs443874 & 5.3 & 2.26\\
-	& 	&	rs400942 & 6 & 2.2\\
-26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & *rs60136336 & 5.7 & 2.25\\
-
-25918132 & \multicolumn{1}{p{3cm}}{\raggedright Toxicological Sciences} & rs76780579 & 6 & 2.09\\
+28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
+	& 	&  	rs201130852 & 5.05 & 72.28\\
+	& 	&  	rs201255786 & 5.7 & 68.97\\
+	& 	&  	rs200655768 & 6.52 & 66.67\\
+	& 	&  	rs184202621 & 5.52 & 60.45\\
+	& 	& 	rs80274284 & 6 & 56.15\\
+	& 	& 	rs200699422 & 5.3 & 7.43\\
+23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
+28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
+26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
+\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
+\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}}\\
+\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
+\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
+\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
+\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
+\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
+\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
+\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
+\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
+\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
+\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
+ 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
+	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
+\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\
 
@@
 Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of $ p > -\log_{10}(\alpha)$ are deemed statistically significant.
-}
+The black text colour indicates that this variant may be spurious, grey text colour indicates that these variants are likely correct but may be impacted by a calling bias.}
 \label{gwasTable}
@@
 In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{}).
-Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p < 10^{-2}$ in the reverse GWAS.
+We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
+Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS.
 (The supplementary material contains a list of variants statistically associated to $Q$).
@@
 If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
-The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
-This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the $Q$ scores and only kept the earliest sequencing date per sample.
-The dates of the sequencing are only used to plot Figure.
+The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files \ref{ProcessPlotData.Rmd}.
 
@@
 Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here}
-
-\todo{link to my compiled metadata file here}
 
@@
 
-%Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
-These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization.
-%\begin{align}
-%\mathbf{L }= \mathbf{AH}
-%\end{align} 
-%where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
-When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
-%\begin{align} \label{lfa}
-%y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
-%\end{align} 
-%for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
-%
-%To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing} The GCAT is a regression approach that assumes the following model.
-%\begin{align}\label{gcat}
-% y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
-%\end{align} 
-%for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$.
-The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model.
-We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs.
-The matrix \luke{Describe matrix}
-
 Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. 
@@
 Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
+See supplementary file {here}. \luke{make this file}
 
-\subsection{Genome wide association study}
-We ran a simple logistic regression independently for each genomic locus $s$ using the R glm2 package\citep{RDevelopmentCoreTeam2016}.
-The model we fitted to each subset of data can be expressed as such.
-\begin{align*}
-%{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{PC1}_{i} + \beta_{2}{PC2}_{i} + \beta_{3}{PC3}_{i} + \beta_{4}{PC4}_{i} + \beta_{5}{PC5}_{i} + \beta_{6}Q_{i}
-{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{Pop}_{i}  + \beta_{2}Q_{i}
-\\
-\textit{for}\quad i &= 1,\hdots, n
-\end{align*}
-Here $Y_{si}$ is a Bernoulli random variable denoting the genotype of individual $i$ at genomic locus $s$. 
-%In the model, the regression coefficient $\beta_{1}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
-In the model, the regression coefficient $\beta_{1}{Pop}_{i}$ represents the coefficient of the categorical variable of Population, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$.
-We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
+\subsection{Individual-specific allele frequency}
+Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
+These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization. 
+\begin{align}
+\mathbf{L }= \mathbf{AH}
+\end{align} 
+where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i^\text{th}$ individual and the $s^\text{th}$ row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
+When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
+\begin{align} \label{lfa}
+y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
+\end{align} 
+for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
 
-For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{2}^{(s)}=0$. That is, the average quality of mapped bases $Q$ variable is not associated with genotype $Y_{s}$.
-In other words, the slope coefficient of the simple logistic regression is zero.
-To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q$ in the model. 
-The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
-
-Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}
+To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing}.
+The GCAT is a regression approach that assumes the following model.
+\begin{align}\label{gcat}
+ y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
+\end{align} 
+for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
+The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model. 
+We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs. 
+The matrix includes the sequence data from the 2,504 individuals from the 1kGP down-sampled to the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip}
 

### 2019-04-04 legacy.tex
complete.

@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
 Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
@@
 The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012} \sgcomment{also cite phase 3}, leading to heterogeneous sample preparations and data quality.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012,phase3}, leading to heterogeneous sample preparations and data quality.
 
@@
 
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure \citep{song2015testing}.
-\sgcomment{We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates.
-}
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
+We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates (Supplementary Figure \ref{CompareModel}).
 
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \luke{link to file here}.
+Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \ref{rsIDList}.
 
@@
 
-We did a similar analysis using the variants identified in the \textit{5PC} \sgcomment{I don't think that this has been described?
-I also don't understand why the 5pc approach has so many more variants than the logistic factor analysis above} model present in the 83 resequenced individuals.
+We did a similar analysis using the variants identified in the GCAT model present in the 83 resequenced individuals.
 Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data (See Figure \ref{90HanSFS_full}).
@@
 
+We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip. 
+This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
+Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
+
 \subsection{Suspicious variants impact modern genomics analyses}
-
-\subsubsection{Imputation}
 
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-
-\subsubsection{Microarrays}
-\sgcomment{It seems like this paragraph is out of place and would fit better in the validation section, above? }
-We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
-This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position.
-Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
-
-\subsubsection{GWAS}
-\sgcomment{"these dubious", below, refers to all dubious positions rather than only the ones on the chips, right? Even more reason to move the previous paragraph. }
 
@@
 They also removed rare alleles and alleles with high degrees of missingness.
-Despite using state-of-the-art quality controls, these erroneous \sgcomment{we don't know that all or even most are erroneous} variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.
+Despite using state-of-the-art quality controls, these variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.
 
@@
 To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.
-\sgcomment{you know, we could simply compare the relative effect size to predicted frequency $\pi$}
 
@@
 \textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
-\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}}\\
+\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
 \textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
@@
 
-\subsection{Technical Artifact}
 The variants identified in this study are likely to be technical artifacts from legacy technologies.
@@
 A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011} with, for instance, different patterns of quality decrease along reads.
-%(known as B-tails).
 Differences in read quality and error profiles in turn require different calling pipelines.
@@
 
-%For instance, our test was limited by the genome wide resolution of the quality per sample.
-%Data quality at a SNP-wise resolution would be a better metric to use, but unfortunately was not available.
-
-%If the Illumina Genome Analyzer sequencing platforms are no longer being used, then it it likely that these sequencer specific artifacts are no longer 
 However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes.
@@
 
-%Even though we expect that present-day technologies will be less sensitive to such artefacts, it is worth pointing out that cancer mutational signature analysis can also be sensitive to technological artefacts that are difficult to detect using classical quality metrics.
-
-%For conventional GWAS, we do not see evidence that these contaminated hits impact the conclusions made in each of these publications.
-%However, considering many polygenic risk score estimates use the association scores from multiple sources, the inclusion of these variants could possibly impact downstream analyses.
-
-\subsection{Recommendations}
 While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
@@
 
-\subsection{Data Availability}
+\subsection{Code and data availability}
+For more details regarding data processing or figure plotting, our pipelines and R Notebooks are publicly available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
 
@@
 By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
-See supplementary file {here}. \luke{make this file}
+See supplementary file  \ref{rsIDList}.
 
@@
 The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.
-
-%\subsection{Frequencies of variants in other datasets}
-%Using the allele frequency files for the Japanese HapMap samples, we extracted the positions we identified as being significantly associated to the average quality of mapped bases.
-We then compared the frequencies of these 70 variants in the 1kGP Japanese samples to those of the HapMap Japanese samples.
-Figure \ref{HapMap_GnomAD} compares the allele frequencies for suspicious sites that are present in HapMap.
-
-%Similarly, using the allele frequencies from GnomAD, we were able to extract 186 sites that are statistically significantly associated to the average quality of mapped bases.
-However, in this case we compared global allele frequencies in both the 1kGP and GnomAD (Figure \ref{HapMap_GnomAD}).
-
-\section{Code Availability}
-https://github.com/LukeAndersonTrocme/QualityPaper
 

### 2019-04-05 legacy.tex
MBE formatting

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ELIFE ARTICLE TEMPLATE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% PREAMBLE 
-\documentclass[9pt,lineno]{elife}
-
-\usepackage[export]{adjustbox}
-\usepackage{lscape}
-\usepackage{afterpage}
-\usepackage{graphicx}
+% PREAMBLE 
+%\documentclass[9pt,lineno]{elife}
+\documentclass[custompaper]{MBE}%
+%\DeclareOption{a4paper}
+\usepackage{url}
+\usepackage{xcolor}
 \usepackage{subcaption}
@@
 \newcommand{\luke}[1]{\textcolor{blue}{Luke: #1}}
-\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
-\newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
 \newcommand{\logit}{\operatorname{logit}}
-%\newcommand{\qq}{m}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Legacy Data Confounds Genomics Studies}
+%\jshort{mst}
 
-\author[1,2]{Luke Anderson-Trocm\'e}
-\author[1,2]{Rick Farouni}
-\author[1,2]{Mathieu Bourgey}
-\author[3]{Fumihiko Matsuda}
-\author[3]{Yoichiro Kamatani}
-\author[1,2]{Simon Gravel}
+%\volname{}
 
-\affil[1]{Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada}
-\affil[2]{McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada}
-\affil[3]{Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
-\corr{simon.gravel@mcgill.ca}{SG}
+%\jvolume{0}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE START
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\jvol{}
+
+%\jissue{0}
+
+%\pubyear{2019}
+
+\mstype{Article}
+
+%\artid{012}
+
+%\access{Advance Access publication March 3, 2013}
 
@@
 
-\maketitle
-\begin{abstract}
-Recent reports have identified differences in the mutational spectra across human populations.
+\keyword{1000 Genomes Project, Batch Effect, Population Structure, GWAS, GCAT}
+
+\title{\ Legacy Data Confounds Genomics Studies}
+
+\author[Anderson-Trocm\'e et al.]{\ Luke \surname{Anderson-Trocm\'e},$^{\ast,1,2}$ Rick Farouni,$^{1,2}$ Mathieu Bourgey,$^{1,2}$ Fumihiko Matsuda,$^{3}$, Yoichiro Kamatani$^{3}$, and Simon Gravel$^{1,2}$,}
+
+\address{$^{1}$Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada\\
+$^{2}$McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada\\ 
+$^{3}$Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
+
+\coresp{simon.gravel@mcgill.ca}
+
+\abstract{Recent reports have identified differences in the mutational spectra across human populations.
 While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
@@
 Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
-\end{abstract}
+}
+
+\maketitle
 
@@
 
+\newpage{}
 \section{Methods}
@@
 
+\bibliographystyle{natbib}
 \bibliography{Legacy.bib}

### 2019-04-16 legacy.tex
fixed citation, reverted back to eLife formatting

--- 2019-04-04 legacy.tex
@@
 Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.
+Many more genomes are available indirectly through servers providing imputation services \citep{mccarthy2016reference} or summary statistics for variant frequency estimation \citep{Lek2016}.
 
@@
 
-\bibliography{Legacy.bib}
+\bibliography{Legacy}
 

### 2019-04-29 legacy.tex
edits up to GWAS

@@
 Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, and a community effort may be required to remove or upgrade such legacy sequencing data from reference databases.
+Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, it may be time to remove or upgrade such legacy sequencing data from reference databases.
 \end{abstract}
@@
 The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012,phase3}, leading to heterogeneous sample preparations and data quality.
+The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012,phase3}, leading to heterogeneous sample preparations and data quality over the course of the project.
 
-Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used as a reference panel for imputation, allele frequency estimations and to answer a wide range of medical and evolutionary questions.
+Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 These mismatches were present despite filtering for low-quality regions of the human genome and Hardy-Weinberg disequilibrium.
+\sgcomment{Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts?
+}
 
@@
 To ensure that population structure was properly taken into account, we considered alternative models with principal components (PCs), Population membership or both as a covariate. 
-These models were broadly consistent (See Figure \ref{CompareModel}).
+These models were broadly consistent (See Figure \ref{CompareModel}). \sgcomment{Is this not redundant with the paragraph before last?}
 
@@
 
-In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
-After restricting the analysis to individuals present in both datasets, we find 296 suspicious variants polymorphic in 1kGP. Of those, 9 are present in the resequenced data \ref{90HanSFS}.
-While this provides strong validation that the variants are not cell-line specific mutations, it is more than the 0.1\% of the sites: some variants that are associated to $Q$ may be present in the population but have biased genotypes.
+In 2017, Lan et al.
+resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
+Here we consider whether variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
 
-We did a similar analysis using the variants identified in the GCAT model present in the 83 resequenced individuals.
-Of the 24,390 suspicious variants identified globally, only 12,086 of them are found in the 83 individuals in the original 1kGP data (See Figure \ref{90HanSFS_full}).
-From this subset, only 1,637 of these variants are present in the resequenced data, the rest are missing entirely.
-Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in the resequenced data.
-A number of variants shared across cohorts suggests that roughly 10\% of the variants identified are present in the cell lines real but exhibit genotyping bias.
+Among the 296 variants associated with $Q$ within the 1kGP CHB and CHS, and polymorphic the 83 resequenced individuals, only 9 are present in the resequenced data (Figure \ref{90HanSFS}).
+This is more than our nominal false positive rate of 1\% of the sites.
+Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
+However, the vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
 
-We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip. 
-This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
-Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
+We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
+Of the 24,390 suspicious variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Supplementary Figure \ref{90HanSFS_full}).
+From this subset, only 1,637 are present in the resequenced data.
+Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in both datasets.
+Thus roughly 10\% of the variants identified are associated with $Q$ are likely present in the cell lines but exhibit biased genotypes.
+
+Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip \sgcomment{Figure}.
+These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
 

### 2019-04-29 legacy.tex
update

@@
 \author[1,2]{Mathieu Bourgey}
+\author[3]{Yoichiro Kamatani}
+\author[3]{Koichiro Higasa}
+\author[4,5]{Jeong-Sun Seo}
+\author[4]{Changhoon Kim}
 \author[3]{Fumihiko Matsuda}
-\author[3]{Yoichiro Kamatani}
 \author[1,2]{Simon Gravel}
@@
 \affil[3]{Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
+\affil[4]{Bioinformatics Institute, Macrogen Inc., Seoul, 08511, Republic of Korea}
+\affil[5]{Precision Medicine Center, Seoul National University Bundang Hospital, Seongnam, 13605, Republic of Korea}
 \corr{simon.gravel@mcgill.ca}{SG}
@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-These mismatches were present despite filtering for low-quality regions of the human genome and Hardy-Weinberg disequilibrium.
-\sgcomment{Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts?
-}
+Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
 
@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
-We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population labels as covariates (Supplementary Figure \ref{CompareModel}).
+We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population membership as covariates. 
+These models were broadly consistent (See Figure \ref{CompareModel}).
 
@@
 \end{figure}
-To ensure that population structure was properly taken into account, we considered alternative models with principal components (PCs), Population membership or both as a covariate. 
-These models were broadly consistent (See Figure \ref{CompareModel}). \sgcomment{Is this not redundant with the paragraph before last?}
 
@@
 
-In 2017, Lan et al.
-resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
+In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
 Here we consider whether variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
@@
 Among the 296 variants associated with $Q$ within the 1kGP CHB and CHS, and polymorphic the 83 resequenced individuals, only 9 are present in the resequenced data (Figure \ref{90HanSFS}). 
-This is more than our nominal false positive rate of 1\% of the sites.
-Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
+This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes. 
 However, the vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
@@
 
-Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip \sgcomment{Figure}.
+Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Supplementary Figure \ref{Omni}). 
 These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
@@
  Only 1,637 of the variants are present in the high depth resequenced individuals.}  
-\label{90HanSFS_full}
+\label{90HanSFS_30}
 \end{figure}
@@
 Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.}
-\label{HapMap_GnomAD}
+\label{Omni}
 \end{figure}

### 2019-04-29 legacy.tex
more edits

@@
 
-To investigate the proportion of suspicious variants that are being imputed, we submitted the first two chromosomes of  the 1kGP genotype data to the Michigan Imputation Server.
+To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted \sgcomment{genotype data for} the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
 We found that all of the variants associated with $Q$ were imputed back in the samples.
@@
 Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence data \citep{yucesoy2015genome, ellinghaus2016analysis}.
+One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence\sgcomment{genotyping chip?} data \citep{yucesoy2015genome, ellinghaus2016analysis}.
 
-All papers used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
@@
 To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.
+\sgcomment{Some discussion is missing here!
+Probably discuss the PLOS one in more detail}
 
@@
 
-To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kgp.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kgp (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
+To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
+Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
 
@@
 The significant association of a variant with a quality metric is not in itself an indication that the variant is spurious.
-It may simply be subject to a weak but significant calling bias --- not an extraordinary surprise in a low-coverage dataset with a complex, joint sample variant calling pipeline.
-It may nevertheless be prudent to discard such variants from analyses that may be sensitive to calling biases, such as mutation profiles and allele frequency estimates.
-The imputation of such variants can also lead to milder problems such as the addition of noise in polygenic risk scores, biases in fine mapping analyses.
+As we mentioned above, it may simply be subject to a weak but significant calling bias.
+It may nevertheless be prudent to carefully examine GWAS associations for such variants, e.g.
+by repeating the analysis without the 1kGP as part of the imputation panel.
 
-While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
-In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
-Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS.
+For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the associates SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary material).
+We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
+Given the value of freely accessible data, resequencing individuals with low Q would  also likely be worth it for the community.
+
+%While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
+%In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
+%We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
+%Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS.
 (The supplementary material contains a list of variants statistically associated to $Q$).
@@
 On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
-The genome-wide simple logistic regression analysis approach is straightforward, and should probably be a standard in a variety of -omics studies.
-The logistic factor analysis is more computationally demanding and produces robust results.
+The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies.
+The logistic factor analysis is more computationally demanding and produces more robust results.
 Both approaches produce comparable results.
@@
 Given rapid technological progress, the focus of genomic research is naturally on the data generation side. 
-However, cleaning up data is also important to avoid generating spurious results. 
+However, cleaning up existing databases is also important to avoid generating spurious results. 
 The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for removal or re-sequencing.

### 2019-04-29 legacy.tex
updated results about two-fold threshold

@@
 
-This method identifies a total of 24,390 variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
@@
 
-To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted \sgcomment{genotype data for} the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
+To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
 We found that all of the variants associated with $Q$ were imputed back in the samples.
@@
 Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used chip sequence\sgcomment{genotyping chip?} data \citep{yucesoy2015genome, ellinghaus2016analysis}.
+One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used genotyping chip data \citep{yucesoy2015genome, ellinghaus2016analysis}.
 
@@
 To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.
-\sgcomment{Some discussion is missing here!
-Probably discuss the PLOS one in more detail}
+We would expect the variants above this threshold to be enriched in real variants with genotyping biases.
+We also note that the majority of the papers whose GWAS variants are associated to $Q$ are above this threshold (See Figure \ref{Pub_30}).
+However, one study that has 7 variants below this threshold is a study that specifically looked at the copy number variation of EBV.
+The results from this study could be confounded by the data quality of the 1kGP as these 7 variants are only significantly associated to EBV copy number in populations we identified as having strong batch effects.
 
@@
 \centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/Published_OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of suspicious variants identified in publications, for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Each of the rsIDs of the variants are labelled for clarity.}
+\label{Pub_30}
+\end{figure}
+
+\begin{figure}[h]
+\centering
 \includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}

### 2019-04-29 legacy.tex
more edits

@@
 These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations. 
-To distinguish between weak but significant biases and dubious SNPS, we also considered the effect size and distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two.
-We would expect the variants above this threshold to be enriched in real variants with genotyping biases.
-We also note that the majority of the papers whose GWAS variants are associated to $Q$ are above this threshold (See Figure \ref{Pub_30}).
-However, one study that has 7 variants below this threshold is a study that specifically looked at the copy number variation of EBV.
-The results from this study could be confounded by the data quality of the 1kGP as these 7 variants are only significantly associated to EBV copy number in populations we identified as having strong batch effects.
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{...}.
+\sgcomment{Give the distribution of large and weak}.
+Most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}).
+However, three studies reported on SNPs with strong bias.
+One such study reports associations with seven suspicious variants that we find to be highly biased.
+This study considered copy number of \sgcomment{Epstein-Barr Virus} sequence in the 1kGP as a phenotype.
+Thus the phenotype in that study is likely confounded by the same technical artefacts.
 

### 2019-04-29 legacy.tex
more edits

@@
 \textbf{A} 
-Mutation spectrum of the 1034 variants that reached a genome wide significance with a \textit{p}-value less than $p < 10^{-6}$  in a GWAS of sequencing quality. 
+Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
 The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern.
@@
 These three enrichments can be summarized as G**${\rightarrow}$GG*.
-(note: the reverse complement of *AC${\rightarrow}$*CC is GT*${\rightarrow}$GG*)
 \textbf{B}
-Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
+Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset \sgcomment{Citation? That will be discussed in more detail elsewhere?}.
 Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
@@
 Each BAS file has metadata regarding each sequencing event for each sample. 
-If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
+If a sample was sequenced more than once, we took the average of each $Q$ score from each sequencing instance. 
 The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files \ref{ProcessPlotData.Rmd}.
@@
 \subsection{Quality Controls}
-We reproduced the quality control pipelines used by Harris et.
-al as they applied the current state of the art quality thresholds to remove questionable sequences especially for the high standards for detecting population level differences.
+For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al.
+as they applied the current state of the art quality thresholds to remove questionable sequences for detecting population level differences.
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
 The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome, highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
-Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 we're considered.
+Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
 In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
 
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of 28,516,063 variants ($S$) included in the test (see Table \ref{totTable}).
-We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately.
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test (see Table \ref{totTable}).
+We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
 
@@
 \subsection{Testing the association of quality to genotype}
-When conducting a statistical analysis of population genetics data, an essential consideration is  to adjust or account for population structure.
+When conducting a statistical analysis of population genetics data, we must account for population structure.
 In a typical GWAS, we are interested in modelling the phenotype as a function of the genotype.
@@
 \begin{align} \label{lfa}
-y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big)
+y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big).
 \end{align} 
@@
 \begin{align}\label{gcat}
- y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg)
+ y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg).
 \end{align}
 
-Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. 
+Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. \sgcomment{Not sure why, but the equation numbers don't line up}
 $\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
 To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
-The deviance test statistic is approximately chi-square distributed with one degrees of freedom.
+The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.
 
@@
 By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
-See supplementary file  \ref{rsIDList}.
+See supplementary file  \ref{rsIDList} for a list of variants.
 
@@
 for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
-The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model. 
+The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model.
 We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs.
-The matrix includes the sequence data from the 2,504 individuals from the 1kGP down-sampled to the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip}
+\sgcomment{This is not consistent with what is said in the text.
+Also, population labels?
+what is M?}
+To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip} \sgcomment{OMNI 1M?}
 
 \subsection{Mutation Spectrum}
-We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}
-We modified the methods as necessary for our purposes, scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
+\sgcomment{This paragraph should perhaps be in the QC section above?}
+We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al.
+2017.
+\citep{Harris2017a}.
+Scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
 
@@
 Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2.
-We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip} genotype data.
+We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni 1M chip} genotype data.
 The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.

### 2019-04-29 legacy.tex
more edits in captions

@@
 Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset \sgcomment{Citation? That will be discussed in more detail elsewhere?}.
-Crosses ( + ) are variants that reached genome wide significance in a GWAS of sequencing quality. 
+Crosses $(+)$ are variants that associate with $Q$ in the JPT. 
 The histogram on the left of the plot is the distribution of significant variants. 
@@
 This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
-The same analysis was performed independently for each of the populations in the 1kGP. }
+The same analysis was performed independently for each of the populations in the 1kGP.  \sgcomment{Label axes!}}
  \label{SFS}
@@
 The x-axis is ranked by the mean sequencing date per population.
+\sgcomment{Not sure that I understand last 2 sentences.
+What it looks like to me: Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates).}.
 \textbf{B} The x-axis is sorted by the sequencing date per individual.
@@
 SNPs and INDELs were also analyzed separately.
-The total number of 24,390 are statistically significantly associated to $Q$.}
+\sgcomment{All four variant categories were subjected to a distinct FDR estimation.?}
+The total number of 24,390 are statistically significantly associated to $Q$. \sgcomment{Have sums in table?}}
 \label{sigTable}
@@
 \caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
+The size of the crosses ( + ) \sgcomment{I don't see crosses} are proportional to the -log10(p) value of that SNP.
 The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
@@
 The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
-The same variants identified as being low quality independently in each population are found in other populations. }
+The same variants identified as being low quality independently in each population are found in other populations. \sgcomment{What does this last sentence mean??}}
   \label{OverLap}
@@
 \begin{tabular}{l l l r r r}
-  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
+  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
 28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
@@
 The variants reaching genome wide significance have a star ( * ).
-Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of $ p > -\log_{10}(\alpha)$ are deemed statistically significant.
-The black text colour indicates that this variant may be spurious, grey text colour indicates that these variants are likely correct but may be impacted by a calling bias.}
+The black text colour indicates that this variant may be spurious, grey text colour indicates that these variants are likely correct but may be impacted by a calling bias.
+\sgcomment{This is not what the color indicates!
+Give the criterion first, interpretation afterwards!}}
 \label{gwasTable}
@@
 Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
+\sgcomment{I don't know if you can have "here" hyperlinks in papers?}
 
@@
 \end{tabular}
-\caption{Number of variants included in the analysis by category}
+\caption{Number of variants included in the analysis by category \sgcomment{Include in Table 1?}}
 \label{totTable}
@@
 \section{Supplementary Figures}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
 
@@
 These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
-In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model.}  
+In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model. \sgcomment{label axes}}  
 \label{CompareModel}
 \end{figure}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
 
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
-There are 15,018 SNPs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
+There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
-There are 2,121 INDELs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
+There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
-There are 4,405 SNPs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
+There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the -log10 p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
-There are 642 INDELs that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
+There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
 The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
@@
 This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
-Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017}.
-Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.}  
+Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017} \sgcomment{missing a verb?}.
+Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.
+\sgcomment{This caption could be clarified a lot!}}
 \label{90HanSFS}
@@
 This plot shows  12,086 significant variants identified in the reverse GWAS model including all populations.
- Only 1,637 of the variants are present in the high depth resequenced individuals.}  
+ Only 1,637 of the variants are present in the high depth resequenced individuals. \sgcomment{This caption could be clarified to be more self-contained}}  
 \label{90HanSFS_30}
@@
 \includegraphics[width=15cm,keepaspectratio]{./Figures/Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of suspicious variants identified in publications, for individuals with $Q$ scores above and below 30.
+\caption{Site frequency spectrum plot comparing the frequency of suspicious \sgcomment{Make sure suspicious is defined throughout} variants identified in publications, for individuals with $Q$ scores above and below 30.
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
@@
 This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
-Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.}
+Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
+\sgcomment{This is a bit too much interpretation and redundant with the main text (also, not clear how the figure relates to caption)}}
 \label{Omni}
@@
 \caption{\textbf{A} 
-The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality. 
+The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality. \sgcomment{Also show original?}
 \textbf{B} 

### 2019-04-30 legacy.tex
updated text with 10kb window analysis

@@
 There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
-These three enrichments can be summarized as G**${\rightarrow}$GG*.
-\textbf{B}
-Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset \sgcomment{Citation? That will be discussed in more detail elsewhere?}.
+These three enrichments can be summarized as G**${\rightarrow}$GG*. \textbf{B} 
+Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
 Crosses $(+)$ are variants that associate with $Q$ in the JPT. 
@@
 This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
-The same analysis was performed independently for each of the populations in the 1kGP.  \sgcomment{Label axes!}}
+The same analysis was performed independently for each of the populations in the 1kGP.}
  \label{SFS}
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-While trying to follow up on this observation in a larger and more recent Japanese cohort, we did not find this particular signature.
+While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
 When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
@@
 \caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project. 
-Individuals are ranked by the date of the earliest sequencing data is used for individuals sequenced more than once.
-The x-axis is ranked by the mean sequencing date per population.
-\sgcomment{Not sure that I understand last 2 sentences.
-What it looks like to me: Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates).}.
+ Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
  \textbf{B} The x-axis is sorted by the sequencing date per individual. 
@@
 
-Thus, this mutational signal is heavily enriched in suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
+Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
 The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses where quality of imputation matters, identifying suspicious variants may be preferable.
+For other analyses where quality of imputation matters, identifying $Q$-associated variants may be preferable.
 
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-Lists of suspicious variants and individuals with low $Q$ are provided as supplementary material \ref{rsIDList}.
+Lists of $Q$-associated variants and individuals with low $Q$ are provided as supplementary material \ref{rsIDList}.
+
+With the exception of chromosome 1 these variants appear to be evenly distributed across chromosomes (Figure \ref{CountPerChrom}).
+Clusters of variants reaching significance are apparent in 10kb regions of the genome (Figure \ref{VariantDensity}).
+These regions coincide with regions with known sequencing issues like HLA and other immunoglobulin receptor genes.
 
@@
 \centering
-\begin{tabular}{l  r r}
-                      & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 5,186 & 17,124 \\  
-{INDEL} & 509 & 1,571 \\ \hline
+\begin{tabular}{l  r r r}
+                      & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
+{SNP}  & 5,186 & 17,124 & 22,310\\  
+{INDEL} & 509 & 1,571 & 2,080\\ \hline
+{Total} & 5,695 & 18,695 & 24,390\\
 \end{tabular}
@@
 SNPs and INDELs were also analyzed separately.
-\sgcomment{All four variant categories were subjected to a distinct FDR estimation.?}
-The total number of 24,390 are statistically significantly associated to $Q$. \sgcomment{Have sums in table?}}
+All four variant categories were subjected to a distinct FDR estimation.
+A total of 24,390 are statistically significantly associated to $Q$.}
 \label{sigTable}
@@
 \caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-The size of the crosses ( + ) \sgcomment{I don't see crosses} are proportional to the -log10(p) value of that SNP.
+The size of the vertical bars ( I ) are proportional to the -log10(p) value of that SNP.
 The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
@@
 The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
-The same variants identified as being low quality independently in each population are found in other populations. \sgcomment{What does this last sentence mean??}}
+The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$.}
   \label{OverLap}
@@
 We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 24,390 suspicious variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Supplementary Figure \ref{90HanSFS_full}).
+Of the 24,390 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Supplementary Figure \ref{90HanSFS_full}).
 From this subset, only 1,637 are present in the resequenced data.
@@
 However, three studies reported on SNPs with strong bias.
-One such study reports associations with seven suspicious variants that we find to be highly biased.
-This study considered copy number of \sgcomment{Epstein-Barr Virus} sequence in the 1kGP as a phenotype.
+One such study reports associations with seven $Q$-associated variants that we find to be highly biased.
+This study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype.
 Thus the phenotype in that study is likely confounded by the same technical artefacts.
@@
 \end{tabular}
-\caption{A list of recent publications that reported suspicious variants as close to or above the genome-wide significant threshold.
+\caption{A list of recent publications that reported $Q$-associated variants as close to or above the genome-wide significant threshold.
 The variants reaching genome wide significance have a star ( * ).
-The black text colour indicates that this variant may be spurious, grey text colour indicates that these variants are likely correct but may be impacted by a calling bias.
-\sgcomment{This is not what the color indicates!
-Give the criterion first, interpretation afterwards!}}
+The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure \ref{Pub_30}).}
 \label{gwasTable}
@@
 Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
-\sgcomment{I don't know if you can have "here" hyperlinks in papers?}
 
@@
 In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
+We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al.
+2017.
+\citep{Harris2017a}.
+Scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
 
@@
 \end{tabular}
-\caption{Number of variants included in the analysis by category \sgcomment{Include in Table 1?}}
+\caption{Number of variants included in the analysis by category.}
 \label{totTable}
@@
 
-Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. \sgcomment{Not sure why, but the equation numbers don't line up}
+Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. 
+\sgcomment{Not sure why, but the equation numbers don't line up}
 $\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
@@
 The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model.
-We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs.
-\sgcomment{This is not consistent with what is said in the text.
-Also, population labels?
-what is M?}
-To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip} \sgcomment{OMNI 1M?}
-
-\subsection{Mutation Spectrum}
-\sgcomment{This paragraph should perhaps be in the QC section above?}
-We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al.
-2017.
-\citep{Harris2017a}.
-Scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
+We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs ($M =2,306,130$ from the 1kGP OMNI 2.5). 
+\sgcomment{This is not consistent with what is said in the text. Also, population labels? what is M?}
+To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}
 
@@
 We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
+We would also like to thank the members of the Gravel lab for their help with coding and brainstorming ideas.
 
@@
 \label{CompareModel}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/CountPerChrom.jpg}
+\caption{The proportion of $Q$-associated variants per chromosome.}  
+\label{CountPerChrom}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/VariantDensity.jpg}
+\caption{The number of $Q$-associated variants per 10kb window across the genome. 
+56 regions have more than 10 variants per window. 
+These regions are listed in the \ref{RegionsOver10.bed}{Bed} file.}  
+\label{VariantDensity}
 \end{figure}
@@
 \includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data.
-This plot only shows significant SNPs identified in the single population tests of CHS and CHB populations.
-Only 9 of the 296 significant SNPs identified in the resequenced Han samples \citep{Lan2017} \sgcomment{missing a verb?}.
-Because the resequencing did not include all the individuals from the original 1kGP dataset, we downsampled the data to match the resequenced individuals.
-\sgcomment{This caption could be clarified a lot!}}
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 296 variants associated with $Q$ in the single population tests within the 1kGP CHB and CHS, 9 are present in the resequenced data \citep{Lan2017}.}  
 \label{90HanSFS}
@@
 \includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequenced data. 
-This plot shows  12,086 significant variants identified in the reverse GWAS model including all populations.
- Only 1,637 of the variants are present in the high depth resequenced individuals. \sgcomment{This caption could be clarified to be more self-contained}}  
+\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 12,086 variants associated with $Q$ in the GCAT model including all populations, 1,637 are present in the high depth resequenced individuals.}  
 \label{90HanSFS_30}
@@
 \includegraphics[width=15cm,keepaspectratio]{./Figures/Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of suspicious \sgcomment{Make sure suspicious is defined throughout} variants identified in publications, for individuals with $Q$ scores above and below 30.
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
@@
 \includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of suspicious variants for individuals with $Q$ scores above and below 30.
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants for individuals with $Q$ scores above and below 30.
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.}
@@
 \caption{We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
-This does not mean that these variants are false positives, but rather, that in the 1kGP, individuals with low $Q$ were less likely to be called as having a variant in that position. 
-Moreover, using the 1kGP as a reference panel might impact the population allele frequencies observed for these positions.
-\sgcomment{This is a bit too much interpretation and redundant with the main text (also, not clear how the figure relates to caption)}}
+Variants highly associated to $Q$ tend to have low global allele frequencies.}}
 \label{Omni}
@@
 \caption{\textbf{A} 
-The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality. \sgcomment{Also show original?}
+The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
 \textbf{B} 

### 2019-04-30 legacy.tex
added discussion of weird regions

@@
 
-With the exception of chromosome 1 these variants appear to be evenly distributed across chromosomes (Figure \ref{CountPerChrom}).
-Clusters of variants reaching significance are apparent in 10kb regions of the genome (Figure \ref{VariantDensity}).
-These regions coincide with regions with known sequencing issues like HLA and other immunoglobulin receptor genes.
+$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
+At a 10kb scale, we also see rather uniform distribution with a small number of regions showing a strong enrichment for such variants (Figure \ref{VariantDensity}).
+We identify 56 regions of 10kb each that collectively contain 1132 $Q$- associated (less than 5\% of problematic variants).
+These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary file \sgcomment{XX}.
+Distribution of association statistics in a sample of these regions is provided in Figure \ref{x}.
 
@@
 \caption{We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
-Variants highly associated to $Q$ tend to have low global allele frequencies.}}
+Variants highly associated to $Q$ tend to have low global allele frequencies.}
 \label{Omni}

### 2019-04-30 legacy.tex
nearly done

@@
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing a strong enrichment for such variants (Figure \ref{VariantDensity}).
-We identify 56 regions of 10kb each that collectively contain 1132 $Q$- associated (less than 5\% of problematic variants).
-These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary file \sgcomment{XX}.
-Distribution of association statistics in a sample of these regions is provided in Figure \ref{x}.
+We identify 56 regions of 10kb each that collectively contain 1,132 $Q$- associated (less than 5\% of problematic variants).
+These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary file RegionsOver10\_geneID.bed.
+Distribution of association statistics in a sample of these regions is provided in Figure \ref{Top10}.
 
@@
 
-%While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
-%In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$.
-%We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
-%Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS.
-(The supplementary material contains a list of variants statistically associated to $Q$).
-
 \section{Conclusion}
@@
 The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for removal or re-sequencing.
-
-%We have not identified the precise mechanism through which the technical artifact was introduced.
-Given that modern platforms seem to have resolved this technical issue, but that recent studies continue to be affected by the legacy batch effect, we found it worthwhile to draw attention to the statistical artefact while a sequencing forensics investigation can take place.
 
@@
 By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
-See supplementary file  \ref{rsIDList} for a list of variants.
+See supplementary file rsIDList.txt for a list of variants and adjusted \textit{p}-values.
 
@@
 56 regions have more than 10 variants per window. 
-These regions are listed in the \ref{RegionsOver10.bed}{Bed} file.}  
+These regions are listed in the Supplementary file RegionsOver10.bed.}  
 \label{VariantDensity}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{./Figures/top10.jpg}
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the ten 10kb windows with the most  $Q$-associated variants per 10kb across the genome. 
+These regions are listed in the Supplementary file RegionsOver10.bed.}  
+\label{Top10}
 \end{figure}

### 2019-04-30 legacy.tex
minor tweaks

@@
 To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
-We also considered two alternative approaches to account for confounders, namely using the leading five  principal component analysis, and using population membership as covariates. 
+We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
 These models were broadly consistent (See Figure \ref{CompareModel}).
@@
 The null model is 
-\begin{align} \label{lfa}
+\begin{align} \label{lfa_basic}
 y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big).
@@
 
-\begin{align}\label{gcat}
+\begin{align}\label{gcat_basic}
  y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg).
@@
 
-Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \ref{gcat} reduces to Model \ref{lfa}. 
-\sgcomment{Not sure why, but the equation numbers don't line up}
+Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \eqref{gcat_basic} reduces to Model \eqref{lfa_basic}. 
 $\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
@@
 The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model.
-We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs ($M =2,306,130$ from the 1kGP OMNI 2.5). 
-\sgcomment{This is not consistent with what is said in the text. Also, population labels? what is M?}
+We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
 To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}

### 2019-04-30 legacy.tex
minor update

@@
 These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations. 
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{...}.
-\sgcomment{Give the distribution of large and weak}.
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}.
+\luke{One the one hand,18,930 variants are strongly biased, whereas 3,430 are weakly biased}.
 Most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}).
@@
 These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
-In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model. \sgcomment{label axes}}  
+In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model.}  
 \label{CompareModel}

### 2019-05-01 Misc/QE/main.tex
minor text and major clean up of git repo

--- 2018-11-26 QE/main.tex
@@
 
-Bottlenecks can drastically change the allele frequencies within a population and lead to rapid divergences \citep{Rosenberg2002, Charlesworth2003}.
+Bottlenecks can drastically change the allele frequencies within a population and lead to rapid divergences \citep{Rosenberg2002, Charlesworth2003, Balick2015}.
 Rare alleles in the source population can increase significantly in frequency or even become fixed in the founder population.
@@
 
-\subsection{Founder effect or ancient population structure?}
+\subsection{Test statistics : founder effect or ancient population structure?}
 Using Fst statistics we can infer the relative differences between sub-populations of Quebec (see \citep{Roy-Gagnon2011}).
@@
 
-\subsection{The geography of genetics}
-
-Preliminary analyses of the Genizon dataset are in agreement with previous work.
+\subsection{Visualizing population structure}
 In collaboration with another graduate student from the Gravel lab, Alex Diaz-Papkovich, we performed a UMAP projection of the Genizon genotype data \citep{mcinnes2018umap,diaz2018}.
@@
 UMAP has been shown to outperform dimension reduction techniques like tSNE both computationally and at preserving both local and global population structure \citep{diaz2018, mcinnes2018umap}.
-This projection shows significant population structure in Quebec (Figure \ref{UMAP}).
-
-We colored this plot based on the three dimensional coordinates of the UMAP projection space (X,Y,Z) converted to three dimensional color space (R,G,B) (Figure \ref{UMAP}).
+This visualization of Quebec population structure shows five clusters which is in agreement with previous works (Figure \ref{UMAP}) \citep{Gagnon2001, Bherer2011,Roy-Gagnon2011}. 
+\vfill
+\subsection{The geography of genetics}
+The three dimensional coordinates of the UMAP space (X,Y,Z) were converted to three dimensional color space (R,G,B) (Figure \ref{UMAP}).
 Therefore, Individuals with similar color are closer to each other in the projection and presumably share more recent ancestry.
 To link the genetic data to geographic coordinates, we used the marriage locations of the ancestors of those linked to the BALSAC database.
-The ancestral locations of these individuals in combination with the three dimensional UMAP coloring scheme show geographic clustering of the sub-populations becomes apparent.
-Five genetic clusters are visible in the UMAP projection, these clusters also appear to be geographically linked.
+The ancestral locations of these individuals in combination with the three dimensional UMAP coloring scheme show geographic clustering of the five sub-populations.
 This is in agreement with what Roy-Gagnon observed \citep{Roy-Gagnon2011}.
-These preliminary results are very promising as they appear to capture the geographic and genetic clustering of the Quebec population.
+These preliminary results are very promising as they appear to capture the geographic and genetic clustering of the French-Canadian population of Quebec.
 
@@
 Dr. Christian Dina in the University of Nantes in France has expressed interest in sharing French genomics data for this project.
-\clearpage
 
+\newgeometry{
+	paper=a4paper, % Change to letterpaper for US letter
+	inner=2cm, % Inner margin
+	outer=2cm, % Outer margin
+	bindingoffset=0cm, % Binding offset
+	top=2.6cm, % Top margin
+	bottom=1.5cm, % Bottom margin
+}
 \chapter{Estimating human mutation rates}
+
+\section{Review of current methods and limitations}
 
@@
 In addition, more recent studies have found that mutation rates can even vary among human populations. \citep{Harris2015a, Harris2017a}.
-%Converting the mutation rate per generation to a yearly rate is also met with difficulties when estimating average human generation time.
-%Generation times are known to vary among current populations, and likely varied over longer time spans \citep{Henn2012, Moorjani2016}
 
-\section{Review of current methods and limitations}
-
-%\subsection{Phylogenetic methods}
+\subsection{Phylogenetic methods}
 The first human mutation rates were estimated by Haldane (1935) using the rates of children born with rare diseases from unaffected parents \citep{Haldane1935}.
@@
 
-%\subsection{Direct methods}
+\subsection{Direct methods}
 Instead of using distant ancestry and fossil records, another more direct approach to estimating mutation rates is to sequence the genomes of parents and offspring. 
@@
 
-%\subsection{Population genetics methods}
+\subsection{Population genetics methods}
 Methods for estimating mutation rates using population genetics are increasingly being used.
@@
 These segments shared by multiple individuals are compared and mutations in these segments are then used in combination with the number of generations separating two individuals \citep{Campbell2012}.
-Using autozygous regions from 70 trio sequenced Hutterite families, they obtained an estimated mutation rate of $1.2\times10^{-8}$.
+An estimated mutation rate of $1.2\times10^{-8}$ was obtained by using autozygous regions from 70 trio sequenced Hutterite families.
 Palamara et al.
-(2015) describe an even more generalized method by identifying IBD segments from unrelated individuals and inferring the number of generations using coalescent theory \citep{Palamara2015,Baharian2016}.
+(2015) describe an even more generalized method by identifying IBD segments from unrelated individuals and inferring the number of generations using coalescent theory \citep{Palamara2012,Palamara2015,Baharian2016}.
 This method then regresses the number of mutations observed per base against the time of most recent common ancestor.
 The slope of the regression is the mutation rate per generation and the intercept is the inferred genotyping false-positive rate.
-Applying this method to 498 individuals from 250 families, they measured a mutation rate of ($2.08\times10^{-8}$).
+Applying this method to 498 individuals from 250 families, they measured a mutation rate of ($2.08\times10^{-8}$)\citep{Palamara2015}.
 %The major limitation to this method comes from the coalescence estimations.
@@
 \section{Goals}
-%\subsection{Applying new methods to new data}
 The aim of the third chapter of this thesis will be to explore mutation rate estimations using a combination of sequence data and genealogical data from the French-Canadian population.
@@
 
+\restoregeometry 
 %----------------------------------------------------------------------------------------
@@
 \begin{figure}
-\includegraphics[width=0.85\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
+\begin{center}
+\includegraphics[width=0.75\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
 
@@
 \label{MapQual}
+\end{center}
 \end{figure}

### 2019-05-01 legacy.tex
minor update
minor text and major clean up of git repo

@@
 The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model.
-We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
+We approximated the population structure using $K=5$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
 To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}
@@
 Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2.
-We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni 1M chip} genotype data.
+We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni 2.5M chip} genotype data.
 The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.
@@
         \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatSNPs.jpg}
-        \label{fig:a}
+
     \end{subfigure} %
@@
         \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatSNPs.jpg}
-        \label{fig:b}    
+
     \end{subfigure} 
@@
         \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatIndels.jpg}
-        \label{fig:a}
+
     \end{subfigure} %
@@
         \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatIndels.jpg}
-        \label{fig:b}    
+
     \end{subfigure} 
@@
         \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatSNPs.jpg}
-        \label{fig:a}
+
     \end{subfigure} %
@@
         \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatSNPs.jpg}
-        \label{fig:b}    
+
     \end{subfigure} 
@@
         \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatIndels.jpg}
-        \label{fig:a}
+
     \end{subfigure} %
@@
         \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatIndels.jpg}
-        \label{fig:b}    
+
     \end{subfigure} 

### 2019-05-01 rick/misc/1KGP_v2.Rmd
minor text and major clean up of git repo

--- 2019-03-07 rick/1KGP.Rmd
@@
 
-to_datalist <- function(i, data_genotype, data_iaf){
+to_datalist <- function(i){
 
@@
      return(data)       
-}
-
-get_af_data <- function(){
-
-  data <- read.bed("./data/plink/ALL_Pops")
-  message("plink data loaded")
-  str(data)
-  LF <- readRDS("./data/preprocessed/lf.rds")
-
-  # compute the first d factors including the intercept
-
-  message("running  af")
-  tic()
-  data_iaf <- af(data, LF)
-  toc()
-
-  saveRDS(data_iaf, "./data/preprocessed/iaf.rds")
-
-  loci <- c(1:dim(data)[1])
-  names(loci) <- loci
-
-  message("to datalist")
-
-  tic()
-  data_list <- future_map(loci, 
-                 to_datalist,
-                 data_genotype=data,
-                 data_iaf=data_iaf)
-
-  toc()
-
-  data_list <- 
-  enframe(data_list,
-          name = "locus", 
-          value = "dataset")
-
-  return(data_list)
-
 }
@@
 ```{r}
-run_workflow <- function(loci){
-
-  data_list <- get_af_data()
-
-  saveRDS(data_list, "./data/preprocessed/data_list.rds")
+fit_data <- function(data_list){
 
@@
 ```{r}
-data_list <- run_workflow ()
+  data_genotype <- read.bed("./data/plink/ALL_Pops")
+  message("plink data loaded")
+
+  # compute the first d factors including the intercept
+
+  data_iaf <- readRDS("./data/preprocessed/iaf.rds")
+
+  loci <- c(1:dim(data_genotype)[1])
+  names(loci) <- loci
+
+  message("to datalist")
+
+  tic()
+  data_list <- future_map(loci, 
+                 to_datalist)
+
+  toc()
+
+  data_list <- 
+  enframe(data_list,
+          name = "locus", 
+          value = "dataset")
+
+saveRDS(data_list, "./data/preprocessed/data_list.rds")
+```
+
+```{r}
+data_list <- fit_data (data_list)
 data_list

### 2019-05-01 rick/misc/2_run_lfa.Rmd
minor text and major clean up of git repo

---
title: "HGP analysis: LFA"
author: 
- name: Rick Farouni, 
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

# Preparations

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), fig.width=16)
options(scipen=999)
```

```{r}
library(lfa)
library(tictoc)
```
```{r}
gc() #- do it now
gcinfo(TRUE) #-- in the future, show when R does it
```

```{bash}
free -t -g
```

```{r load}
tic()
 data <- read.bed("./data/plink/ALL_Pops")
toc()
 message("plink data loaded")
```

```{bash}
free -t -g
```

```{r lfa}
tic()
LF <- lfa(data, 4)
toc()
message("lfa run complete")
```

```{r save}
tic()
saveRDS(LF, "./data/preprocessed/lf.rds")
toc()
```
```{r}
gcinfo(verbose = FALSE) #-- don't show it anymore
gc()
```

```{r}
sessionInfo()
```

### 2019-05-01 PlottingScripts/AllSig_RunModel.R

@@ -1,108 +1,0 @@
-library(dplyr)
-library(data.table)
-library(ggplot2)
-library(glm2)
-library(cowplot)
-
-#read colors per pop (used in 1kG Nature)
-colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
-MyColour <- as.character(colors $hex)
-names(MyColour) <- colors $Pop
-
-path = '/Users/luke/genomes/genomes/hg19/Genotypes/'
-
-#read Genotypes
-GT = fread('~/Documents/QualityPaper/sig/SigVar_Pop_0.01_rsID.genotype.txt')
-ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames_rsID.txt', header=F)
-names(GT) = as.character(unlist(ColNames))
-names(GT)[4] = 'rsID'
-GT$ID <- as.numeric(as.character(GT$ID))
-
-samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt')
-samples$Pop = factor(samples$Pop, levels=c('FIN','GBR','CEU','IBS','TSI','CHS','CDX','CHB','JPT','KHV','GIH','STU','PJL','ITU','BEB','PEL','MXL','CLM','PUR','ASW','ACB','GWD','YRI','LWK','ESN','MSL'))
-
-model = apply(GT[,-(1:3)], 1, function(x)
-						glm2(x ~
-							samples$Pop +
-							samples$average_quality_of_mapped_bases))
-#get coef of Qual per site
-coef = lapply(model, function(x) coef(x)[[27]])				
-
-dev = lapply(model, function(x) anova(x, test='Chi')[[3,2]])
-
-plt = cbind(unlist(coef), GT[,c(1:3)])
-plt = cbind(plt, unlist(dev))
-plt$ID = as.numeric(as.character(plt$ID))
-plt$pval = -log10(pchisq(plt$V2, 1, lower.tail=F))
-
-#to deal with duplicate entries, merge based on Chr Pos and AF
-SigPos = fread('/Users/luke/Documents/QualityPaper/sig/SignificantVariants.txt')
-
-plt = merge(plt, SigPos, by = c('CHROM','POS','ID'))
-
[…]

### 2019-05-01 PlottingScripts/Clean_Figure1.R

@@ -1,238 +1,0 @@
-library(ggplot2)
-library(colorspace)
-library(cowplot)
-library(data.table)
-
-if(F){
-#f1kGP='~/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered.freq.frq'
-f1kGP='/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_JPT_1.frq'
-fnag='~/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'
-hweFile='/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT_sig6.hwe'
-
-#Allele Frequency files
-JPT<-fread(f1kGP, fill=T, col.names=c('CHROM','POS','N_ALLELES','N_CHR','JPT_AF','JPT_MAF'), colClasses=c('numeric'))
-NAG<-fread(fnag, fill=T,col.names=c('CHROM','POS','N_ALLELES','N_CHR','NAG_AF','NAG_MAF'), colClasses=c('numeric'))
-
-#merge them together to get joint frequency spectrum
-join<-merge(JPT[which(JPT$JPT_AF <= 1),],
- 			NAG[which(NAG$NAG_AF <= 1),], 
- 			by=c('CHROM','POS'), all=T)
-join[is.na(join)]<-0
-rm(NAG)
-rm(JPT)
-#REMOVE  HWE SITES
-NAG_hwe<-fread('/Users/luke/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered_sig6.freq.hwe')
-JPT_hwe <- fread('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered_sig6.freq.hwe')
-hwe=merge(NAG_hwe[,c('V1','V2')], JPT_hwe[,c('V1','V2')], by=c('V1','V2'), all=T)
-
-join= join[! paste(join$CHROM, join$POS) %in% c(paste(hwe$V1,hwe$V2)), ]
-join=join[,c('CHROM','POS','JPT_MAF','JPT_AF','NAG_MAF','NAG_AF')]
-
-join$CHROM<-as.numeric(as.character(join$CHROM))
-join$POS<-as.numeric(as.character(join$POS))
[…]

### 2019-05-01 PlottingScripts/Clean_Figure3.R

@@ -1,82 +1,0 @@
-#load libraries
-library(ggplot2)
-library(cowplot)
-library(data.table)
-#pop names, used to read/write files
-pops<-read.table('/Users/luke/Documents/MutSpect/WrapUpPlots/NamePop.txt')
-#read colors per pop (used in 1kG Nature)
-colors<-read.table("~/Dropbox/LukeTemp/NamePopBigPop_color.html.txt", comment.char='>',col.names=c('Pop','hex','BigPop'))
-MyColour <- as.character(colors $hex)
-names(MyColour) <- colors $Pop
-
-phase1=c('YRI','LWK','CHB','CHD','JPT','CEU','LWK')
-
-#read the list of sig snps
-#bad=read.table('~/Documents/1FinalGWAS/1kGP_GenomeWide.4bed_filtered_sig6.POS.txt',col.names=c('Chr', 'Pos'))
-#dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
-#fileNames = list.files(path=dir, pattern='dev10.csv', full.names = T)
-#Reg = do.call(rbind, lapply(fileNames, function(x) fread(x)))
-
-Reg<- fread('/Users/luke/Documents/QualityPaper/Misc/AllRegressionsOver10.txt')
-Reg<-Reg[which(Reg$Pos != 'Pos'),]
-Reg$Chr<-as.numeric(as.character(Reg$Chr))
-Reg$Pos<-as.numeric(as.character(Reg$Pos))
-Reg$dev<-as.numeric(as.character(Reg$dev))
-Reg$plog10 <- - pchisq(Reg$dev, 1, lower.tail=F, log.p=T)/log(10)
-Reg<-Reg[which(Reg$plog10 >= 6),]
-
-NumPop<-as.data.frame(table(Reg$Chr,Reg$Pos))
-UnPop<-NumPop[which(NumPop$Freq==1),]
-NumPop<-NumPop[which(NumPop$Freq>1),]
-NumPop$Var1<-as.numeric(as.character(NumPop$Var1))
-NumPop$Var2<-as.numeric(as.character(NumPop$Var2))
-
-
-m=merge(Reg,NumPop, by.y=c('Var1','Var2'), by.x=c('Chr','Pos'))
-n=as.data.frame(table(m$Pop))
-plt=unique(merge(m,n,by.x='Pop',by.y='Var1'))
-plt$title = paste(plt$Pop, ' : ',plt$Freq.y)
[…]

### 2019-05-01 PlottingScripts/Clean_RegressionPlots.R

@@ -1,169 +1,0 @@
-## Deal with Regression output
-
-library(dplyr)
-library(magrittr)
-library(data.table)
-library(ggplot2)
-library(RColorBrewer)
-library(multtest)
-library(cowplot)
-library(reshape2)
-
-Regression <-fread('~/Documents/QualityPaper/Misc/GenomeWide.Regression.txt', colClasses='numeric',col.names=c('CHROM','POS','ID','dev.p'))
-
-Regression $ID<-as.numeric(as.character(Regression $ID))
-Regression $CHROM<-as.numeric(as.character(Regression $CHROM))
-Regression $POS<-as.numeric(as.character(Regression $POS))
-Regression $dev.p<-as.numeric(as.character(Regression $dev.p))
-Regression <- Regression[complete.cases(Regression),]
-
-
-#read in list of global singletons
-singles<-fread(
-	'/Users/luke/genomes/genomes/1kGP_NotFiltered/GenomeWide_doubles.frq',
-	col.names=c('CHROM','POS'))
-singles$drop<-'Yes' #dummy variable for merging
-
-#join the singles to the Reg to get info of what sites to remove
-Regression <-left_join(Regression, singles, by=c('CHROM','POS'))
-rm(singles)
-
-#this removes the singletons
-Regression <-Regression[which(is.na(Regression$drop)),]
-Regression$drop<-NULL
-
-##read Repeats
-Repeats <- fread('~/Documents/QualityPaper/Misc/Reg_strictMask_nestedRepeats.POS',fill=T,col.names=c('CHROM','POS'))
-Repeats$Repeat='yes'
-
-##read Indels
-Indels <- fread('/Users/luke/Documents/QualityPaper/Misc/Reg_indels.POS', skip=1 ,col.names=c('CHROM','POS'))
-Indels$Indel = 'yes'
-
-#read in rsIDs (used for GWAS catalogue search)
-rsID<-fread('/Users/luke/genomes/genomes/hg19/phase3/rsID.txt')
-colnames(rsID)<-c('CHROM','POS','rsID')
-
-Regression <- left_join(Regression, Repeats, by=c('CHROM','POS'))
-
-Regression <- left_join(Regression, Indels, by=c('CHROM','POS'))
-
[…]

### 2019-05-01 PlottingScripts/CompareModels.R

@@ -1,205 +1,0 @@
-library(dplyr)
-library(data.table)
-library(ggplot2)
-library(glm2)
-library(cowplot)
-
-#read Genotypes
-GT = fread('gzcat ~/genomes/genomes/hg19/Genotypes/CHR22.Genotypes.txt.gz',nrows=50000)
-Pos = GT[,c(1:3)]
-names(Pos) <- c('Chr','Pos','AF')
-ColNames = read.table('/Users/luke/genomes/genomes/hg19/Genotypes/ColNames.txt', header=F)
-names(GT) = as.character(unlist(ColNames))
-
-samples = fread('/Users/luke/Documents/PCAperPop/Name_Pop_Qual_PC1_PC2_PC3_PC4_PC5.txt') 
-
-Pop = apply(GT[,-(1:3)], 1, function(x)
-						glm2(x ~
-							samples$Pop +
-							samples$average_quality_of_mapped_bases))
-#save(Pop, file='~/Documents/QualityPaper/Misc/Temp.Pop.RData')
-#rm(Pop)
-
-PC = apply(GT[,-(1:3)], 1, function(x)
-						glm2(x ~
-							samples$PC1 + 
-							samples$PC2 +
-							samples$PC3 +
-							samples$PC4 +
-							samples$PC5 + 
-							samples$average_quality_of_mapped_bases))
-#save(PC, file='~/Documents/QualityPaper/Misc/Temp.PC.RData')
-#rm(PC)	
-					
-PopPC = apply(GT[,-(1:3)], 1, function(x)
-						glm2(x ~
-							samples$Pop +
-							samples$PC1 + 
-							samples$PC2 +
-							samples$PC3 +
-							samples$PC4 +
-							samples$PC5 + 
-							samples$average_quality_of_mapped_bases))						
-#save(PopPC, file='~/Documents/QualityPaper/Misc/Temp.PopPC.RData')
-#load('~/Documents/QualityPaper/Misc/Temp.Pop.RData')
-#load('~/Documents/QualityPaper/Misc/Temp.PC.RData')																							
-#get coef of Qual per site
-coefPop = lapply(Pop, function(x) coef(x)[[27]])
-coefPC = lapply(PC, function(x) coef(x)[[7]])
-coefPopPC = lapply(PopPC, function(x) coef(x)[[32]])
-
[…]

### 2019-05-01 PlottingScripts/Han_1kGP_SFS.R

@@ -1,70 +1,0 @@
-##plot SFS sig Han
-library(data.table)
-library(ggplot2)
-library(reshape2)
-
-awkPos<-fread('/Users/luke/Documents/QualityPaper/sig/Total_Sig_awk_POS.txt')
-
-Qual<-fread('~/Dropbox/LukeTemp/SubMeta.txt')
-Qual$V1<-NULL
-
-Han<- fread('~/genomes/genomes/Han90/83_Han_Sig.frq')
-Han$V1 <- as.numeric(Han$V1)
-Han$V2 <- as.numeric(Han$V2)
-
-					
-KGP <- read.table('~/Documents/QualityPaper/SigVCF/1kGP_90Han_shared.frq',
-					sep="\t",fill=TRUE,col.names=paste("V",1:8,sep=''), skip=1)
-					
-plt<-unique(merge(Han, KGP, by=c('V1','V2'), all=T))
-
-plt[which(is.na(plt$V6.x)),]$V6.x <- 0
-plt[which(is.na(plt$V6.y)),]$V6.y <- 0
-
-ggplot(plt, aes(x=V6.x, y=V6.y))+geom_bin2d(binwidth=c(1/83,1/83))+scale_fill_distiller(palette = "Spectral")+labs(x='high depth',y='1kGP')+theme_classic()
-
-ggsave('~/Documents/QualityPaper/Figures/Han83.jpg',height=5,width=5)
-
-#gt1<-fread('/Users/luke/Documents/QualityPaper/SigVCF/1kGP_all.genotypes.txt')
-gt1<-fread('~/Documents/QualityPaper/SigVCF/1kGP_83Han_genomeWide.genotypes.txt')
-gt1.m<-melt(gt1, id.vars = c('#CHROM','POS'))
-colnames(gt1.m)=c('CHROM','POS','Sample','kGT')
-
-
-#gtHan<-fread('~/genomes/genomes/Han90/83_Han_all.genotypes.txt')
-gtHan <-fread('~/genomes/genomes/Han90/90_Han_Chinese_SigPos.genotypes.txt')
-gtHan$'#CHROM' <- as.integer(as.character(gsub('chr','',gtHan$'#CHROM')))
-gtHan.m<-melt(gtHan, id.vars = c('#CHROM','POS'))
-colnames(gtHan.m)=c('CHROM','POS','Sample','HanGT')
-
[…]

### 2019-05-01 PlottingScripts/Han_1kGP_SFS_clean.R

@@ -1,118 +1,0 @@
-##plot SFS sig Han
-library(data.table)
-library(ggplot2)
-library(reshape2)
-library(cowplot)
-library(dplyr)
-
-#zgrep '^#' $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz     > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf
-
-#for f in `seq 1 22`; do echo $f; bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariantsCHROM_POS.txt $GenoPath/ALL.chr${f}.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz | grep -v '#' >> ~/Documents/QualityPaper/sig/SigVar_Pop_0.01.vcf; done
-#java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar extractFields ~/Documents/QualityPaper/sig/SigVar_0.01_Pop.vcf CHROM POS AF "GEN[*].GT" | sed "s/0|0/0/g ; s/0|[1-9]/1/g ; s/[1-9]|0/1/g ; s/[1-9]|[1-9]/1/g ; /|/d" | tail -n +2 > ~/Documents/QualityPaper/sig/SigVar_Pop_0.01_genotype.txt
-
-#bcftools view -R /Users/luke/Documents/QualityPaper/sig/SignificantVariants_chr_CHROM_POS.txt /Users/luke/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_sorted.vcf.gz  -Oz -o /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.01.vcf.gz
-
-#java -jar ~/bin/snpEff_latest_core/snpEff/SnpSift.jar extractFields /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.01.vcf.gz CHROM POS AF "GEN[*].GT" | sed "s/0\/0/0/g ; s/0\/[1-9]/1/g ; s/[1-9]\/0/1/g ; s/[1-9]\/[1-9]/1/g ; /\//d" | tail -n +2 > /Users/luke/genomes/genomes/Han90/SignificantSNPs_Han90_0.01_genotype.txt
-
-
-#read Genotypes
[…]

### 2019-05-01 PlottingScripts/HapMap+GnomAD.R

@@ -1,40 +1,0 @@
-library(data.table)
-library(ggplot2)
-library(stringr)
-library(cowplot)
-
-#GP<-fread('~/Documents/QualityPaper/sig/Total_Sig_POSfreq.txt',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
-GP<-fread('/Users/luke/Documents/QualityPaper/sig/SigVar_0.001.frq',fill=TRUE,col.names=c('CHROM','POS','rsID', 'AF'))
-GP$CHROM<-as.numeric(as.character(GP$CHROM))
-GP$POS<-as.numeric(as.character(GP$POS))
-
-
-
-HM<-fread('/Users/luke/genomes/genomes/HapMap/hapmap_3.3.b37_sig_rsID.frq', sep='\t',fill=T, col.names=c('CHROM','POS','rsID','AF', 'AN'))
-HM<-HM[which(complete.cases(HM)==T),]
-
-
-plt1<-merge(GP, HM, by=c('CHROM','POS','rsID'))
-
-ggplot(plt1, aes(x=AF.y, y=AF.x))+geom_point(shape=1)+geom_point(data=plt1[which(plt1$rsID=='rs6057648'),], color='blue')+geom_point(data=plt1[which(plt1$rsID=='rs301'),], color='green')+labs(x='Allele Frequency in HapMap', y='Allele Frequency in 1kGP', title='Significant SNPs found in HapMap and 1kGP')+theme_classic()+xlim(c(0,1))+ylim(c(0,1))
-
-ggsave('/Users/luke/Documents/QualityPaper/Figures/Hap.jpg',height=7,width=7)
-
-
-GN<-fread('/Users/luke/genomes/GnomAD/gnomad_AF.frq')
-GN$AF_raw<-as.numeric(as.character(GN$AF_raw))
-GN$AC_raw<-as.numeric(as.character(GN$AC_raw))
-ggplot(GN, aes(x=reorder(POS,CHROM),y=AF_raw))+geom_point(shape=1)+theme_classic()+theme(axis.text.x=element_blank())
-
-GNm<-melt(GN[,-c('AF_raw')], id=c('CHROM','POS','ID'))
-GNm$value<-as.numeric(as.character(GNm$value))
-GNm<-GNm[which(GNm$value!=0),]
-
-plt<-merge(GP, GN, by=c('CHROM','POS'))
-
[…]

### 2019-05-01 PlottingScripts/ManhattanPlot.R

@@ -1,184 +1,0 @@
-##Manhattan
-library(dplyr)
-library(magrittr)
-library(data.table)
-library(ggplot2)
-library(multtest)
-library(cowplot)
-#dir='/Volumes/gravel/luke_projects/1000Genomes/Regression/'
-#out='/Volumes/gravel/luke_projects/1000Genomes/MeanDev/'
-dir='~/Documents/Regression/'
-out='~/Documents/Regression/'
-
-if(F){ #this is the loop that took the mean of the deviances
-for(chrom in seq(4,22)){
-	print(chrom)
-	#read in the regressions for each pop
-	fileNames = list.files(path=dir,pattern=paste("CHR",chrom,"_indels.Regression",sep=''), full.names = T)
-	if(length(fileNames) != 0){
-	Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T, sep=' ')))
-	
-	print(head(Reg))
-	#had duplicate headers, so had to remove them and reset columns to numeric
-	Reg<-Reg[which(Reg$Pos != 'Pos'),]
-	Reg$Chr<-as.numeric(as.character(Reg$Chr))
-	Reg$Pos<-as.numeric(as.character(Reg$Pos))
-	Reg$dev<-as.numeric(as.character(Reg$dev))
-	#chisquared random variable
-	#taking the sum of the deviances (mean is to take into account the DF)
-	Reg <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
-	
-	write.table(Reg,paste(out,chrom,'indels_MeanDev.csv',sep=''))
-	}
-}
-}
-
-#read in rsIDs (used for GWAS catalogue search)
-rsID<-fread('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_Chr.Pos.rsID.txt')
-colnames(rsID)<-c('Chr','Pos','rsID')
-
-#read in list of global singletons
-singles<-fread(
-	'~/Documents/QualityPaper/Misc/Singletons_Chr.Pos.rsID.txt',
-	col.names=c('Chr','Pos','rsID'))
-singles$drop<-'Yes' #dummy variable for merging
-	
-doubles<-fread(
[…]

### 2019-05-01 PlottingScripts/PlotUnfiltered.R

@@ -1,44 +1,0 @@
-##Manhattan
-library(dplyr)
-library(magrittr)
-library(data.table)
-library(ggplot2)
-library(multtest)
-library(cowplot)
-
-dir='~/Documents/Regression/'
-
-fileNames = list.files(path=dir,pattern=paste("GenomeWide_NOT.Regression_",sep=''), full.names = T)
-
-Reg = do.call(rbind, lapply(fileNames, function(x) read.table(x, header=T, sep=' ')))
-	
-print(head(Reg))
-#had duplicate headers, so had to remove them and reset columns to numeric
-Reg<-Reg[which(Reg$Pos != 'Pos'),]
-Reg$Chr<-as.numeric(as.character(Reg$Chr))
-Reg$Pos<-as.numeric(as.character(Reg$Pos))
-Reg$dev<-as.numeric(as.character(Reg$dev))
-#chisquared random variable
-#taking the sum of the deviances (mean is to take into account the DF)
-Reg <- Reg %>% group_by(Chr,Pos) %>% summarize(SumDev = sum(dev), MeanDev = mean(dev), Count = n())
-
-Reg$plog10 <- - pchisq(Reg$SumDev, Reg$Count, lower.tail=F, log.p=T)/log(10)
-Reg$p <- pchisq(Reg$SumDev, Reg$Count, lower.tail=F)
-
-ggplot(Reg, aes(x=Pos, y = plog10, color=as.factor(Chr)))+
-facet_grid(~Chr, scales='free_x', space='free_x', switch='x')+
-scale_fill_manual (values=getPalette(colourCount))+
-scale_y_continuous(expand=c(0,0))+
-geom_point(alpha=0.3, size=1)+theme_classic()+
-labs(y='-log10(p)', x='Chromosome')+
-theme(plot.title = element_text(hjust = 0.5), 
-plot.subtitle = element_text(hjust = 0.5), 
-axis.text.x=element_blank(), 
-axis.ticks.x=element_blank(),
-axis.line.x=element_blank(), 
-strip.background = element_blank(),
-strip.text.x = element_text(size = 6))+
-guides(color=F)+expand_limits(y=c(0,21))
-
-ggsave('~/Documents/QualityPaper/Figures/ManhattanPlot_NOT.jpg', height=5, width=10)
-ggsave('~/Documents/QualityPaper/Figures/ManhattanPlot_NOT.tiff', height=5, width=10)

### 2019-05-01 PlottingScripts/Resequenced_Dev.R

@@ -1,49 +1,0 @@
-library(data.table)
-library(ggplot2)
-library(cowplot)
-
-CHB.kgp<-fread('/Users/luke/Documents/Regression/GenomeWideRegression_CHB.csv')
-CHS.kgp<-fread('/Users/luke/Documents/Regression/GenomeWideRegression_CHS.csv')
-
-CHB.kgp$Pos <- as.numeric(as.character(CHB.kgp$Pos))
-CHS.kgp$Pos <- as.numeric(as.character(CHS.kgp $Pos))
-CHB.kgp$Chr <- as.numeric(as.character(CHB.kgp$Chr))
-CHS.kgp$Chr <- as.numeric(as.character(CHS.kgp $Chr))
-CHB.kgp$dev <- as.numeric(as.character(CHB.kgp$dev))
-CHS.kgp$dev <- as.numeric(as.character(CHS.kgp $dev))
-
-CHB.kgp<-CHB.kgp[complete.cases(CHB.kgp),]
-CHS.kgp<-CHS.kgp[complete.cases(CHS.kgp),]
-
-
-CHB.han<-fread('~/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_CHB_Regression.txt')
-CHS.han<-fread('~/genomes/genomes/Han90/90_Han_Chinese_GenomeWide_CHS_Regression.txt')
-
-CHB.han$Chr <- as.numeric(as.character(gsub('chr','',CHB.han$Chr)))
-CHS.han$Chr <- as.numeric(as.character(gsub('chr','',CHS.han$Chr)))
-CHB.han$Pos <- as.numeric(as.character(CHB.han$Pos))
-CHS.han$Pos <- as.numeric(as.character(CHS.han$Pos))
-CHB.han$dev <- as.numeric(as.character(CHB.han$dev))
-CHS.han$dev <- as.numeric(as.character(CHS.han$dev))
-
-CHB.han<-CHB.han[complete.cases(CHB.han),]
-CHS.han<-CHS.han[complete.cases(CHS.han),]
-
-CHB<-merge(CHB.han, CHB.kgp, by=c('Chr', 'Pos', 'Pop'), all =T)
-CHS<-merge(CHS.han, CHS.kgp, by=c('Chr', 'Pos', 'Pop'), all= T)
-CHB[is.na(CHB)] <- 0
-CHS[is.na(CHS)] <- 0
-
[…]

### 2019-05-01 PlottingScripts/hwe_af.R

@@ -1,125 +1,0 @@
-library(ggplot2)
-library(colorspace)
-library(cowplot)
-library(data.table)
-
-f1kGP='/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide_JPT_1.frq'
-fnag='~/genomes/genomes/NAG_4bedFiltered/NAG_GenomeWide.4bed_filtered.freq.frq'
-hweFile='/Users/luke/Documents/GWAS_Qual/Final_GWAS/1kGP_GenomeWide_JPT_sig6.hwe'
-
-JPT_hwe <- fread('/Users/luke/genomes/genomes/1kGP_4bedFiltered/1kGP_GenomeWide.4bed_filtered_1.freq.hwe')
-colnames(JPT_hwe)[c(1,2,8)]<-c('CHROM','POS','HWE_p')
-#LogisticRegression
-JPT_reg <- fread('~/Documents/QualityPaper/Misc/GenomeWide.Regression_JPT.csv')
-colnames(JPT_reg)[c(1,2)]<-c('CHROM','POS')
-JPT_reg$CHROM<-as.numeric(as.character(JPT_reg$CHROM))
-JPT_reg$POS<-as.numeric(as.character(JPT_reg$POS))
-JPT_reg$dev<-as.numeric(as.character(JPT_reg$dev))
-JPT_reg<-JPT_reg[complete.cases(JPT_reg)]
-
-#Allele Frequency files
-JPT<-fread(f1kGP, fill=T, col.names=c('CHROM','POS','N_ALLELES','N_CHR','JPT_AF','JPT_MAF'), colClasses=c('numeric'))
-NAG<-fread(fnag, fill=T,col.names=c('CHROM','POS','N_ALLELES','N_CHR','NAG_AF','NAG_MAF'), colClasses=c('numeric'))
-
-#merge them together to get joint frequency spectrum
-join<-merge(JPT[which(JPT$JPT_AF <= 1),],
- 			NAG[which(NAG$NAG_AF <= 1),], 
- 			by=c('CHROM','POS'), all=T)
-join[is.na(join)]<-0
-
-join=join[,c('CHROM','POS','JPT_MAF','JPT_AF','NAG_MAF','NAG_AF')]
-
-join$CHROM<-as.numeric(as.character(join$CHROM))
-join$POS<-as.numeric(as.character(join$POS))
[…]

### 2019-05-01 rick/ProcessVCF.R

@@ -1,208 +1,0 @@
-#Taking Rick's Code and putting it in an Rscript
-
-#load libaries
-library(Matrix)
-library(irlba)
-library(threejs)
-library(tidyverse)
-library(lfa)
-library(furrr)
-library(glm2)
-library(cowplot)
-library(drake)
-options(future.globals.maxSize= 10000000000)
-future::plan(multicore)
-
-#For running in terminal
-args = commandArgs(trailingOnly=T)
-Chr <- args[1]
-#Chr <- '22'
-#set directories and paths
-GenoPath='/Users/luke/genomes/genomes/hg19/phase3/'
-VCFname = '.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz'
-Dir = '/Users/luke/Documents/QualityPaper/rick/'
-setwd(Dir)
-
-#Get ID names
-if(!file.exists("ids.txt")){
-ids <- readLines(pipe(paste("gzcat ",GenoPath,"ALL.chr",Chr,".phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.vcf.gz  |
-                      sed -n /^#CHROM/p | 
-                      tr '\t' '\n' | 
-                      tail -n +10",sep='')))
-
-writeLines(ids, "ids.txt")}
-
-if(file.exists("ids.txt")){
-  ids <- unlist(read.table("ids.txt"))
-  head(ids)
-}
-
-#Get Quality Scores
-if(!file.exists("bas_dt")){
-read_basfiles <- function(bas_filepath) {
-  
-  bas_file <- read_tsv(bas_filepath,
-                  col_types = cols_only('sample' = 'c',        
-                                       'average_quality_of_mapped_bases' = col_guess())) 
-  return( bas_file)
-}
-
-bas_files <- list.files(path = "/Users/luke/genomes/bas_file/1000GenomesBAS/",
-                        pattern="*bam.bas", full.names = TRUE)
-
-bas_dt <- 
-  map_dfr(bas_files,read_basfiles ) %>%
-  rename(avg_qual="average_quality_of_mapped_bases") %>%
-  group_by(sample) %>% summarize(avg_qual = mean(avg_qual))
-
-#keep samples in 1kGP VCF files
-bas_dt <- bas_dt[which((bas_dt$sample %in% ids)==T),]
-
-#get populations of each ID
-ped <- 
[…]

### 2019-05-01 rick/final/Luke_Cluster/3_fit_model_script.R

@@ -1,155 +1,0 @@
-## ------------------------------------------------------------------------
-library(tidyverse)
-library(lfa)
-library(furrr)
-library(glm2)
-library(tictoc)
-library(data.table)
-options(future.globals.maxSize= 1000000000000)
-plan(multicore)
-
-## ------------------------------------------------------------------------
-
-args <- commandArgs(trailingOnly=T)
-
-path <- args[1]
-setwd(path)
-
-plinkPath <- args[2] 
-
-chromosome <- args[3]
-
-data_genotype <- read.bed(paste0(plinkPath,'chr', chromosome))
-
-message("plink data loaded")
-
-## ------------------------------------------------------------------------
-data_bim <- fread(paste0(plinkPath,'chr', chromosome,'.bim'), 
-                  col.names = c('Chr','rsID','X',
-                                'Pos','Ref','Alt'))
-
-rownames(data_genotype) <- data_bim$rsID
-
-## ----lfa-----------------------------------------------------------------
-tic()
-#LF <- lfa(data_genotype, 5)
-toc()
-message("lfa run complete")
-
-## ----save----------------------------------------------------------------
-tic()
-#saveRDS(LF, "./data/preprocessed/lf5.rds")
-toc()
-
-## ----read lf-------------------------------------------------------------
-LF <- readRDS( "./data/preprocessed/lf5.rds")
-
-## ------------------------------------------------------------------------
-#this runs the logistic regression
-lreg <- function(y, X){
-  X <- rbind(X, X)
-  y1 <- as.numeric((y==1) | (y==2))
-  y2 <- as.numeric(y==2)
-  y <- c(y1,y2)
-  b <- glm2(cbind(y, 2-y) ~ -1 + X, family="binomial")$coef
-  
-  return(b)
-  
-}
-#modified from gcat
-fit_glm <- function(..., LF, trait){
-  snp <- c(...)
-  ind <- !is.na(snp) #logical index vector
-  snp_no_na <- snp[ind]
-  LF_no_na <- LF[ind,]
-  p0 <- rep(NA, length(snp))
-  p1 <- rep(NA, length(snp))
-  trait_no_na <- trait[ind]
-  b0 <- lreg(snp_no_na, LF_no_na) #coefficients from logreg 
-  b1 <- lreg(snp_no_na, cbind(LF_no_na, trait_no_na))
-  est0 <- .Call("mv", LF_no_na, b0)
-  est1 <- .Call("mv", cbind(LF_no_na, trait_no_na), b1)
[…]

### 2019-05-01 Latex/legacy.tex
more git cleaning

--- 2019-05-01 legacy.tex
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/Figure1.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}
 \caption{
@@
 \begin{figure}
-\includegraphics[width=0.95\hsize,keepaspectratio]{./Figures/MapQualOverTime.jpg}
+\includegraphics[width=0.95\hsize,keepaspectratio]{../Figures/MapQualOverTime.jpg}
 
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/SNPOverlap6.jpg}
 
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/fits_Significant_Positions_CompareModels.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/fits_Significant_Positions_CompareModels.jpg}
 \caption{Comparison of three logistic regression models for testing association to $Q$.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/CountPerChrom.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/CountPerChrom.jpg}
 \caption{The proportion of $Q$-associated variants per chromosome.}  
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/VariantDensity.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
 \caption{The number of $Q$-associated variants per 10kb window across the genome. 
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/top10.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/top10.jpg}
 \caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the ten 10kb windows with the most  $Q$-associated variants per 10kb across the genome. 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatSNPs.jpg}
 
@@
     	\center    
-        \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatSNPs.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_NonRepeatIndels.jpg}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatIndels.jpg}
 
@@
     	\center    
-        \includegraphics[width=\hsize]{./Figures/QQPlot_NonRepeatIndels.jpg}
+        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatIndels.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatSNPs.jpg}
 
@@
     	\center    
-        \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatSNPs.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{./Figures/ManhattanPlot_RepeatIndels.jpg}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatIndels.jpg}
 
@@
     	\center    
-        \includegraphics[width=\hsize]{./Figures/QQPlot_RepeatIndels.jpg}
+        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatIndels.jpg}
 
@@
 \centering
-\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_singlePop.jpg}
+\includegraphics[width=12cm,keepaspectratio]{../Figures/Han_1kGP_SFS_singlePop.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
@@
 \centering
-\includegraphics[width=12cm,keepaspectratio]{./Figures/Han_1kGP_SFS_FullModel.jpg}
+\includegraphics[width=12cm,keepaspectratio]{../Figures/Han_1kGP_SFS_FullModel.jpg}
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/Published_OverUnder30.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/Published_OverUnder30.jpg}
 \caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{./Figures/OverUnder30.jpg}
+\includegraphics[width=15cm,keepaspectratio]{../Figures/OverUnder30.jpg}
 \caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants for individuals with $Q$ scores above and below 30.
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/Omni_AF.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/Omni_AF.jpg}
 \caption{We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/PC1_Correlation.jpg}
 \caption{Plots of data metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{./Figures/MutationSpectrum_cutOff.png}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
 \caption{\textbf{A} 

### 2019-05-01 Misc/rick_misc/DownloadVCF_ConvertToPlink.Rmd
more git cleaning

#  Download data and convert to Plink format
```{bash}
GenoName='phase3_shapeit2_mvncall_integrated_v5a.20130502'
GenoPath='/Users/luke/genomes/genomes/hg19/phase3'
PlinkPath='/Users/luke/genomes/genomes/hg19/plink'
PlinkDir='/Users/luke/bin/plink_mac'
cd $GenoPath

for f in `seq 1 22`
  do echo $f
  if [ ! -f $GenoPath/ALL.chr${f}.${GenoName}.genotypes.vcf.gz ]; then
    echo "Downloading VCF CHR ${f}"
    wget ftp://ftp-trace.ncbi.nih.gov/1000genomes/ftp/release/20130502/ALL.chr${f}.${GenoName}.genotypes.vcf.gz
  fi

  if [ ! -f $PlinkPath/chr${f}.bed ]; then
    echo "Converting VCF to Plink : Chr ${f}"
    $PlinkDir/plink \
    --vcf $GenoPath/ALL.chr${f}.${GenoName}.genotypes.vcf.gz \
    --make-bed \
    --out $PlinkPath/chr${f}.bed
  fi
```

### 2019-05-01 Misc/rick_misc/plot.Rmd
more git cleaning

--- 2019-05-01 rick/misc/1KGP_v2.Rmd
@@
 library(tidyverse)
-library(lfa)
-library(furrr)
-library(broom)
 library(tictoc)
-options(future.globals.maxSize= 1000000000000)
+options(future.globals.maxSize= 10000000000)
 plan(multicore)
-```
-
-```{r}
-gc()
-```
-
-```{r}
-gcinfo(TRUE) 
-```
-
-#  Fit model
-
-```{r}
-
-to_datalist <- function(i){
-
-    data <- tibble(genotype=data_genotype[i,],
-               iaf=data_iaf[i,])
-
-     return(data)       
-}
-
-get_predicted_mean <- function(model_fit) {
- predicted <- 
-   augment(model_fit, type.predict = "response") %>%
-   select("genotype","offset.iaf.",".fitted") %>%
-   rename(iaf="offset.iaf.", mu=".fitted" )
-  return(predicted)
-}
-
-get_stats <- function(model_fit){
-  stats_dt <- 
-    tidy(model_fit)  %>%
-    filter(term=="avg_qual_mean") %>%
-    select(estimate, p.value) %>%
-    mutate(deviance= anova(model_fit)["Deviance"][[1]][2])
-  return(stats_dt)
-}
-
-fit_model <- function(dataset){
-
-   fit <- glm(genotype ~ avg_qual_mean + offset(iaf),
-                           family ="binomial",
-                           data=dataset)
-
-   stats <- get_stats(fit)
-   predicted <- get_predicted_mean(fit)
-
-   return(list(stats, predicted))
-}
-
-get_plot <- function(data){
-
- p <- ggplot(bind_cols(data, 
-                   bas_dt %>% 
-                     select(Population,
-                            avg_qual_mean))) +
-    geom_point(aes(avg_qual_mean, 
-                   genotype),
-               size=.5,
-               alpha= 0.4) +
-    geom_line(aes(avg_qual_mean,
-                  iaf, 
-                  colour="af"),
-              linetype="dashed") +
-    geom_line(aes(avg_qual_mean,
-                  mu,
-                  colour="mu")) +
-      theme_bw() +
-    facet_wrap(~Population)
-
-  return(p)
-}
 ```
@@
 ```{r}
-fit_data <- function(data_list){
-
-  message("fitting model")
-  tic()
-  data_list <- 
-    data_list %>%
-  #slice(1:10) %>%
-  transmute(
-    locus=locus,
-    fit1 = future_map(dataset, fit_model))  %>%
-  transmute(
-    locus=locus,
-    stats = future_map(fit1, c(1)),
-    predicted =future_map(fit1, c(2))) %>%
-    unnest(stats) %>%
-    arrange(-deviance)
-
-  toc()
-
-  return(data_list)
-
-}
-
+LF <- readRDS("./data/preprocessed/lf.rds")
 ```
@@
 ```{r}
-  data_genotype <- read.bed("./data/plink/ALL_Pops")
-  message("plink data loaded")
-
-  # compute the first d factors including the intercept
-
-  data_iaf <- readRDS("./data/preprocessed/iaf.rds")
-
-  loci <- c(1:dim(data_genotype)[1])
-  names(loci) <- loci
-
-  message("to datalist")
-
-  tic()
-  data_list <- future_map(loci, 
-                 to_datalist)
-
-  toc()
-
-  data_list <- 
-  enframe(data_list,
-          name = "locus", 
-          value = "dataset")
-
-saveRDS(data_list, "./data/preprocessed/data_list.rds")
+bas_dt <- bind_cols(bas_dt,
+                    tibble(lf1=LF[,1], lf2=LF[,2],lf3=LF[,3]))
 ```
@@
 ```{r}
-data_list <- fit_data (data_list)
-data_list
-```
-
-```{r save_data}
-saveRDS(data_list, "./data/preprocessed/data_fits.rds")
+ggplot(bas_dt,
+       aes(lf2, lf3, color=Population)) +
+  geom_point(size=0.6, alpha=.6) +
+  theme_bw()
 ```
@@
 ```{r}
-n <-20
-target_loci <- 
-  data_list %>% 
-  slice(1:n) %>% 
-  pull(locus)
-target_loci 
+ggplot(bas_dt,
+       aes(lf1, lf3, color=Population)) +
+  geom_point(size=0.6, alpha=.6) +
+  theme_bw()
 ```
@@
 ```{r}
-data_list_subset <- 
-  data_list %>%
-   filter(locus %in% target_loci ) %>%
-    mutate(plotted= future_map(predicted, get_plot)) 
-data_list_subset
+ggplot(bas_dt,
+       aes(lf1, lf2, color=Population)) +
+  geom_point(size=0.6, alpha=.6) +
+  theme_bw()
 ```
-
-```{r}
-data_list_subset %>%
-  pull(plotted)
-```
-
-```{r}
-sessionInfo()
-```

### 2019-05-01 RunAnalysis/2_run_lfa.Rmd
more git cleaning

--- 2019-05-01 rick/misc/2_run_lfa.Rmd
@@
 tic()
-LF <- lfa(data, 4)
+LF <- lfa(data, 5)
 toc()
@@
 tic()
-saveRDS(LF, "./data/preprocessed/lf.rds")
+saveRDS(LF, "./data/preprocessed/lf5.rds")
 toc()

### 2019-05-01 LatexManuscript/legacy.tex
final edits

--- 2019-05-01 Latex/legacy.tex
@@
 When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Supplementary Figure \ref{PC1_Correlation}).
+To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Figure \ref{PC1_Correlation}).
 The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. 
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-Lists of $Q$-associated variants and individuals with low $Q$ are provided as supplementary material \ref{rsIDList}.
+\luke{We also note that 37.5\% of the variants are found within regions flagged by the 1kGP strict callability mask.}
+Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
 
@@
 We identify 56 regions of 10kb each that collectively contain 1,132 $Q$- associated (less than 5\% of problematic variants). 
-These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary file RegionsOver10\_geneID.bed.
+These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary Data. 
 Distribution of association statistics in a sample of these regions is provided in Figure \ref{Top10}.
@@
 We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 24,390 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Supplementary Figure \ref{90HanSFS_full}).
+Of the 24,390 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
 From this subset, only 1,637 are present in the resequenced data.
@@
 
-Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Supplementary Figure \ref{Omni}). 
+Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
 These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
@@
 These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations. 
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between low- and high-$Q$ individuals is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}.
-\luke{One the one hand,18,930 variants are strongly biased, whereas 3,430 are weakly biased}.
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
+The majority (83.6\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.
 Most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}).
@@
 
-For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the associates SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary material).
+For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the associates SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary Data).
 We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
@@
 \section{Methods}
+\subsection{Code and data availability}
+Our pipelines are fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
+This repository has all the scripts used for data download, processing, analysis and plotting.
+Supplementary data as well as preprocessed data for plotting are also available on GitHub.
+
 \subsection{Metadata}
@@
 If a sample was sequenced more than once, we took the average of each $Q$ score from each sequencing instance. 
-The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files \ref{ProcessPlotData.Rmd}.
-
-\subsection{Code and data availability}
-For more details regarding data processing or figure plotting, our pipelines and R Notebooks are publicly available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
-
-Index of BAS files \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000_genomes_project/1000genomes.low_coverage.GRCh38DH.alignment.index}{available here}.
-
-Phase3 analysis sequence index file  \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/20130502.phase3.analysis.sequence.index}{available here}
+The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.
 
@@
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
-The  1000 Genomes \href{http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/accessible_genome_masks/20141020.strict_mask.whole_genome.bed}{strict mask} was used to remove low quality regions of the genome, highly conserved regions were removed using the \href{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/database/phastConsElements100way.txt.gz}{phastCons100way} mask file and highly repetitive regions were also removed using the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file from RepeatMasker.
+The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were also removed using the NestedRepeats mask file from RepeatMasker.
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
 In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
-We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al.
-2017.
-\citep{Harris2017a}.
-Scripts are available \href{https://github.com/LukeAndersonTrocme/QualityPaper}{here}.
+We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}.
 
 For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test (see Table \ref{totTable}).
-We also used the \href{http://hgdownload.cse.ucsc.edu/goldenpath/hg19/database/nestedRepeats.txt.gz}{NestedRepeats} mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
+We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
 
@@
 By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
-See supplementary file rsIDList.txt for a list of variants and adjusted \textit{p}-values.
+See Supplementary Data for a list of variants and adjusted \textit{p}-values.
 
@@
 We approximated the population structure using $K=5$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
-To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{OMNI 2.5M chip}
+To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the OMNI 2.5M chip.
 
@@
 Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2.
-We used the genotyped data from the 1kGP \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni 2.5M chip} genotype data.
+We used the genotyped data from the 1kGP Omni 2.5M chip genotype data.
 The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.
@@
 \clearpage
-\section{Supplementary Figures}
+\section{Figures}
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 56 regions have more than 10 variants per window. 
-These regions are listed in the Supplementary file RegionsOver10.bed.}  
+These regions are listed in the Supplementary Data.}  
 \label{VariantDensity}
@@
 \caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the ten 10kb windows with the most  $Q$-associated variants per 10kb across the genome. 
-These regions are listed in the Supplementary file RegionsOver10.bed.}  
+These regions are listed in the Supplementary Data.}  
 \label{Top10}
@@
 
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatSNPs.jpg}
-
-    \end{subfigure} %
-
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatSNPs.jpg}
-
-    \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
-There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.}
-  \label{NRS_Manhattan}
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{../Figures/OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Two clusters of are visible, where the majority (83.6\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.}
+\label{90HanSFS_full}
 \end{figure}
 
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatIndels.jpg}
-
-    \end{subfigure} %
-
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatIndels.jpg}
-
-    \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
-There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.}
-  \label{NRI_Manhattan}
-  \end{figure}
-
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatSNPs.jpg}
-
-    \end{subfigure} %
-
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatSNPs.jpg}
-
-    \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
-There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.}
-  \label{RS_Manhattan}
-  \end{figure}
-
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatIndels.jpg}
-
-    \end{subfigure} %
-
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatIndels.jpg}
-
-    \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
-There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
-The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.}
-  \label{RI_Manhattan}
+\begin{figure}[h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{../Figures/Published_OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Each of the rsIDs of the variants are labelled for clarity.}
+\label{Pub_30}
 \end{figure}
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Each of the rsIDs of the variants are labelled for clarity.}
-\label{Pub_30}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatSNPs.jpg}
+    \end{subfigure} 
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatSNPs.jpg}
+
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
+There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.}
+  \label{NRS_Manhattan}
   \end{figure}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatIndels.jpg}
+    \end{subfigure} %
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatIndels.jpg}
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.}
-\label{90HanSFS_full}
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
+There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.}
+  \label{NRI_Manhattan}
+  \end{figure}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatSNPs.jpg}
+    \end{subfigure} %
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatSNPs.jpg}
+
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
+There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.}
+  \label{RS_Manhattan}
+  \end{figure}
+\begin{figure} \centering
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatIndels.jpg}
+    \end{subfigure} %
+    \begin{subfigure}[b]{\linewidth}
+    	\center    
+        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatIndels.jpg}
+    \end{subfigure} 
+    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
+There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
+The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20.
+  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.}
+  \label{RI_Manhattan}
   \end{figure}

### 2019-05-01 LatexManuscript/legacy.tex
tweaks and clarifications

@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
+Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts. \sgcomment{Something missing}
 
@@
 To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
-The majority (83.6\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.
-Most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}).
-However, three studies reported on SNPs with strong bias. 
-One such study reports associations with seven $Q$-associated variants that we find to be highly biased. 
+The majority (83.6\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
+By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
+One study reports associations with seven $Q$-associated variants that we find to be highly biased \cite{}. 
 This study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype. 
-Thus the phenotype in that study is likely confounded by the same technical artefacts.
+Thus the phenotype in that study is likely confounded by the same technical artefacts that lead to biased SNP calling.
 

### 2019-05-01 PlottingScripts/ProcessedData/FilterStrictMask.py
update strict script

@@ -1,0 +1,27 @@
+import multiprocessing as mp
+import sys, os
+import subprocess
+
+def Bash_cmd(i):
+    #Set Path and vcfName
+    path = '/Users/luke/genomes/genomes/hg19/phase3/'
+    outPath = '/Users/luke/Documents/Legacy_Misc/Strict/'
+    vcfName = '.phase3_shapeit2_mvncall_integrated_v5a.20130502.genotypes.'
+    Bed = '/Users/luke/genomes/BED_MASKS/20141020.strict_mask.whole_genome.bed'
+
+    In = "{0}ALL.chr{1}{2}vcf.gz".format(path,i,vcfName)
+    Out = "{0}Chr{1}_strict".format(outPath,i)
+
+    subprocess.call("/usr/local/bin/bcftools-1.6/bcftools filter \
+            --regions-file {0} {1} \
+            | /usr/local/bin/bcftools-1.6/bcftools view \
+            --drop-genotypes --output-type v \
+            --output-file {2}".format(Bed, In, Out), shell=True)
+
+if __name__ == '__main__':
+    ListOfChrom=list(range(1,23))
+    pool_size=7 #mp.cpu_count()
+    pool=mp.Pool(processes=pool_size)
+    pool_outputs= pool.map(Bash_cmd, ListOfChrom)
+    pool.close()
+    pool.join()

### 2019-05-02 LatexManuscript/legacy.tex
updated strict mask to scripts and text

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ELIFE ARTICLE TEMPLATE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%% PREAMBLE 
-\documentclass[9pt,lineno]{elife}
+\documentclass[9pt,lineno]{template}
 
@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-Despite considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts. \sgcomment{Something missing}
+\luke{These mismatches were maintained despite} considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
 
@@
 
-This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 15,270 variants associated to $Q$ distributed across the genome (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-\luke{We also note that 37.5\% of the variants are found within regions flagged by the 1kGP strict callability mask.}
+We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
 Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
@@
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing a strong enrichment for such variants (Figure \ref{VariantDensity}). 
-We identify 56 regions of 10kb each that collectively contain 1,132 $Q$- associated (less than 5\% of problematic variants). 
 These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary Data. 
 Distribution of association statistics in a sample of these regions is provided in Figure \ref{Top10}.
+
+Repeat Indels : 181 Repeat SNPs : 3369 NonRepeat Indels : 657 NonRepeat SNPs : 11059
 
@@
                       & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
-{SNP}  & 5,186 & 17,124 & 22,310\\  
-{INDEL} & 509 & 1,571 & 2,080\\ \hline
-{Total} & 5,695 & 18,695 & 24,390\\
+{SNP}  & 3,369 & 11,059 & 14,428\\  
+{INDEL} & 181 & 657 & 838\\ \hline
+{Total} & 3,550 & 11,716 & 15,270\\
 \end{tabular}
@@
 All four variant categories were subjected to a distinct FDR estimation.
-A total of 24,390 are statistically significantly associated to $Q$.}
+A total of 15,270 are statistically significantly associated to $Q$. 
+The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
 \label{sigTable}
@@
 We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 24,390 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
+Of the 15,270 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
 From this subset, only 1,637 are present in the resequenced data.
@@
 
-Among the 24,390 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
+Among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
 These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
@@
 To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
-The majority (83.6\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
+The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
 By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
@@
 
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test (see Table \ref{totTable}).
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
-
-\begin{table}[h]
-\begin{tabular}{l  r r}
-                      & {Repeat}  & {Non-Repeat}       \\ \hline
-{SNP}  & 6,312,620 & 19,846,786   \\  
-{INDEL} &  586,342  & 1,770,315 \\ \hline
-\end{tabular}
-\caption{Number of variants included in the analysis by category.}
-\label{totTable}
-\end{table}
+\luke{We filtered out variants surpassing the significance threshold  flagged by the 1kGP strict mask after the FDR adjustment. }
 
@@
 Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. 
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significance. 
 See Supplementary Data for a list of variants and adjusted \textit{p}-values.
@@
 These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
-In these plots we are comparing the deviance from the null model in the 24,390 variants identified using the LFA model.}  
+In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
 \label{CompareModel}
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/top10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the ten 10kb windows with the most  $Q$-associated variants per 10kb across the genome. 
-These regions are listed in the Supplementary Data.}  
+\includegraphics[width=15cm,keepaspectratio]{../Figures/RegionOver10.jpg}
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
 \label{Top10}
@@
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (83.6\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.}
+Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.}
 \label{90HanSFS_full}

### 2019-05-02 PlottingScripts/ProcessPlotData.Rmd
updated strict mask to scripts and text

@@
 #Directories
-DataPath <- '~/Documents/QualityPaper/ProcessedData/'
-figurePath <- '~/Documents/QualityPaper/Figures/'
+DataPath <- '~/Documents/LegacyData/PlottingScripts/ProcessedData/'
+figurePath <- '~/Documents/LegacyData/Figures/'
 ```
@@
 #see 1_make_quality_dt.Rmd for making this file
-load('/Users/luke/Documents/QualityPaper/rick/bas_dt')
+load(paste0(DataPath,'bas_dt'))
 
@@
                             'character', 'numeric'))
+
+strictMask <- fread(paste0(DataPath,
+                           '1kGP_StrictMask.bed'),
+            col.names = c('Chr','Pos','Pos.1','rsID'))
+#remove positions that don't pass strict mask
+Reg <- anti_join(Reg,
+                 strictMask,
+                 by = c('Chr','Pos'))
+
 #Get P-Values and significant sites
@@
 ```{bash}
-DataPath='/Users/luke/Documents/QualityPaper/ProcessedData'
+DataPath='/Users/luke/Documents/LegacyData/PlottingScripts/ProcessedData'
 RawDataPath='/Users/luke/genomes/genomes/hg19/phase3'
@@
 --out $DataPath/Significant_Positions
+
+#bcftools doesn't take rsID
+#this deals with indels and other mulitallelic variants
+vcftools \
+--vcf $DataPath/Significant_Positions.vcf \
+--snps $DataPath/Significant_rsID.txt \
+--recode --recode-INFO-all \
+--out $DataPath/Significant_Positions_allPop
+
+/Users/luke/bin/plink_mac/plink \
+--vcf $DataPath/Significant_Positions_allPop.recode.vcf \
+--make-bed \
+--out $DataPath/Significant_Positions_allPop
+
+#Get Allele Frequencies
+/Users/luke/bin/plink_mac/plink \
+--bfile $DataPath/Significant_Positions_allPop \
+--freq \
+--out $DataPath/Significant_Positions_allPop
 ```
@@
 ```{bash}
-DataPath='/Users/luke/Documents/QualityPaper/ProcessedData'
+DataPath='/Users/luke/Documents/LegacyData/PlottingScripts/ProcessedData'
 RawDataPath='/Users/luke/genomes/genomes/Han90/90_Han_Chinese_SNP_gatkvqsr_20170424_chr'
@@
 data_bim <- fread(paste0(DataPath,
-                         'Significant_Positions.bim'), 
+                         'Significant_Positions_allPop.bim'), 
                   col.names = c('Chr','rsID','X',
@@
 data_fam <- fread(paste0(DataPath,
-                         'Significant_Positions.fam'), 
+                         'Significant_Positions_allPop.fam'), 
                   col.names = c('familyID','sample',
@@
 
-data_genotype <- read.bed(paste0(DataPath,'Significant_Positions'))
+data_genotype <- read.bed(paste0(DataPath,'Significant_Positions_allPop'))
 rownames(data_genotype) <- data_bim$rsID

### 2019-05-02 LatexManuscript/legacy.tex
updated numbers in text

@@
 
-Among the 296 variants associated with $Q$ within the 1kGP CHB and CHS, and polymorphic the 83 resequenced individuals, only 9 are present in the resequenced data (Figure \ref{90HanSFS}). 
+Among the 296 variants associated with $Q$ within the 1kGP CHB and CHS, and polymorphic the 83 resequenced individuals, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
 This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes. 
@@
 We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 15,270 $Q$-associated variants identified globally, 12,086 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
-From this subset, only 1,637 are present in the resequenced data.
+Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
+From this subset, only 1,139 are present in the resequenced data.
 Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in both datasets.
@@
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1kGP CHB and CHS, 9 are present in the resequenced data \citep{Lan2017}.}  
+Among the 296 variants associated with $Q$ in the single population tests within the 1kGP CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
 \label{90HanSFS}
@@
 \caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 12,086 variants associated with $Q$ in the GCAT model including all populations, 1,637 are present in the high depth resequenced individuals.}  
+Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
 \label{90HanSFS_30}

### 2019-05-02 LatexManuscript/legacy.tex
added one sentence

@@
 These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary Data. 
-Distribution of association statistics in a sample of these regions is provided in Figure \ref{Top10}.
-
-Repeat Indels : 181 Repeat SNPs : 3369 NonRepeat Indels : 657 NonRepeat SNPs : 11059
+Distribution of association statistics in the region with the most $Q$-associated variants is provided in Figure \ref{Top10}.
+\luke{There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask.}
 
@@
 \end{tabular}
-\caption{Number of statistically significant variants per category. 
+\caption{Number of statistically significant variants  passing the 1kGP strict mask per category. 
 Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately.
@@
 \caption{The number of $Q$-associated variants per 10kb window across the genome. 
-56 regions have more than 10 variants per window. 
-These regions are listed in the Supplementary Data.}  
+One region in chromosome 17 has more than 10 variants per window.}  
 \label{VariantDensity}

### 2019-05-02 LatexManuscript/legacy.tex
final edits

@@
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-\luke{These mismatches were maintained despite} considering only sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
+These mismatches were maintained despite only considering sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
 
@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a ``reverse GWAS'' or Genotype-Conditional Association Test, in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
+Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
@@
 
-This method identifies a total of 15,270 variants associated to $Q$ distributed across the genome (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
+This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
@@
 We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
+There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask.
 Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
@@
 $Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
-At a 10kb scale, we also see rather uniform distribution with a small number of regions showing a strong enrichment for such variants (Figure \ref{VariantDensity}). 
-These regions, which include known challenging regions such as HLA and other immunoglobulin receptor genes, are provided in Supplementary Data. 
-Distribution of association statistics in the region with the most $Q$-associated variants is provided in Figure \ref{Top10}.
-\luke{There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask.}
+At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
+We identify a 10kb region in chromosome 17 has 35 $Q$-associated variants.
+Distribution of association statistics in this region is provided in Figure \ref{Top10}.
 
@@
 This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
-However, the vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
 
@@
 Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
-From this subset, only 1,139 are present in the resequenced data.
-Figure \ref{90HanSFS_full} shows a relatively weak correlation to allele frequency for the variants present in both datasets.
-Thus roughly 10\% of the variants identified are associated with $Q$ are likely present in the cell lines but exhibit biased genotypes.
+From this subset, only 1,139 (or 18\%) are present in the resequenced data.
+The allele frequencies of these variants are nearly identical between datasets suggesting that for these 83 individuals, these variants are properly genotyped in the 1kGP.
+There are 5 alleles that show differing frequencies between both datasets that are likely explained by biased genotypes.
+The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
 
@@
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
-\luke{We filtered out variants surpassing the significance threshold  flagged by the 1kGP strict mask after the FDR adjustment. }
+Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
+These variants are only removed after the FDR and excluded from downstream analyses.
 

### 2019-05-02 LatexManuscript/legacy.tex
minor edits
final edits - update 10kb plot

@@
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
-We identify a 10kb region in chromosome 17 has 35 $Q$-associated variants.
+While regions flagged by the 1kGP strict mask appear enriched in $Q$-associated variants, we also identify a 10kb region in chromosome 17 has 35 $Q$-associated variants.
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
@@
 \caption{The number of $Q$-associated variants per 10kb window across the genome.
-One region in chromosome 17 has more than 10 variants per window.}  
+Red bars indicate regions within and black bars indicate regions outside the 1kGP strict mask.
+One region not flagged by the 1kGP strict mask in chromosome 17 has more than 10 variants per window.}  
 \label{VariantDensity}

### 2019-05-02 LatexManuscript/legacy.tex
edits u pto line 163

--- 2019-05-02 LatexManuscript/legacy.tex
@@
 Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus contaminates modern studies in hidden ways, it may be time to remove or upgrade such legacy sequencing data from reference databases.
+Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
+It may be time to retire or upgrade such legacy sequencing data.
 \end{abstract}
@@
 This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
 To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
-There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask.
+
+%We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
+There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask. \sgcomment{Inconsistent?}
 Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
@@
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
-We identify a 10kb region in chromosome 17 has 35 $Q$-associated variants.
+An outlying 10kb region in chromosome 17 \sgcomment{give coordinates} has 35 $Q$-associated variants.
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
@@
 \caption{Number of statistically significant variants  passing the 1kGP strict mask per category. 
-Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately.
-SNPs and INDELs were also analyzed separately.
-All four variant categories were subjected to a distinct FDR estimation.
+Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
 A total of 15,270 are statistically significantly associated to $Q$. 
@@
 
-In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-Here we consider whether variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-
-Among the 296 variants associated with $Q$ within the 1kGP CHB and CHS, and polymorphic the 83 resequenced individuals, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
+In 2017, Lan et al.
+resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
+To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
+Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
 This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.

### 2019-05-02 LatexManuscript/legacy.tex
tweaks throughout,a nd some comments

@@
 An outlying 10kb region in chromosome 17 \sgcomment{give coordinates} has 35 $Q$-associated variants.
+
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
@@
 From this subset, only 1,139 (or 18\%) are present in the resequenced data.
-The allele frequencies of these variants are nearly identical between datasets suggesting that for these 83 individuals, these variants are properly genotyped in the 1kGP.
+The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
 There are 5 alleles that show differing frequencies between both datasets that are likely explained by biased genotypes.
@@
 By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
-One study reports associations with seven $Q$-associated variants that we find to be highly biased \cite{}. 
-This study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype. 
+One study reports associations with seven $Q$-associated variants that we find to be highly biased \sgcomment{cite!}\cite{}. 
+That study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype. 
 Thus the phenotype in that study is likely confounded by the same technical artefacts that lead to biased SNP calling.
@@
 However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes.
-The significant association of a variant with a quality metric is not in itself an indication that the variant is spurious.
-As we mentioned above, it may simply be subject to a weak but significant calling bias.
-It may nevertheless be prudent to carefully examine GWAS associations for such variants, e.g.
+Even though significant association of a variant with a quality metric is not in itself an indication that the variant is spurious, we would recommend to carefully examine GWAS associations for such variants, e.g.
 by repeating the analysis without the 1kGP as part of the imputation panel.
 
-For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the associates SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary Data).
+For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the $Q$-associated SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary Data).
 We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
-Given the value of freely accessible data, resequencing individuals with low Q would  also likely be worth it for the community.
+Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.
 
@@
 The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies.
-The logistic factor analysis is more computationally demanding and produces more robust results.
+The logistic factor analysis is more computationally demanding but produces produces more robust results \sgcomment{We never explained this!}.
 Both approaches produce comparable results.
@@
 However, cleaning up existing databases is also important to avoid generating spurious results. 
-The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for removal or re-sequencing.
+The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for retirement or re-sequencing.
 
@@
 \subsection{Code and data availability}
-Our pipelines are fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
-This repository has all the scripts used for data download, processing, analysis and plotting.
-Supplementary data as well as preprocessed data for plotting are also available on GitHub.
+Since this data is primarily performed using publicly available data, we provide fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
+This repository includes scripts used for data download, processing, analysis and plotting.
+%Supplementary data as well as preprocessed data for plotting are also available on GitHub.
 
@@
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
-The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were also removed using the NestedRepeats mask file from RepeatMasker.
+The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were removed using the NestedRepeats mask file from RepeatMasker.
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
 In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
-We calculated the mutation spectrum of triplets for the list of significant variants for the JPT population using a similar method as described in Harris et al. 2017. \citep{Harris2017a}.
+We calculated the mutation spectrum of base pair triplets for the list of significant variants for the JPT population using a similar method as described in  \citep{Harris2017a}.
 
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test \sgcomment{Explain why is this larger than the previous number?}.
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
 Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
-These variants are only removed after the FDR and excluded from downstream analyses.
+These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
 
@@
 We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
-We would also like to thank the members of the Gravel lab for their help with coding and brainstorming ideas.
+We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.
 

### 2019-05-02 LatexManuscript/legacy.tex
reordering of a few lines

@@
 %We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
-There are 9,124 $Q$-associated variants that are within the flagged regions of the 1kGP strict mask. \sgcomment{Inconsistent?}
 Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
@@
 An outlying 10kb region in chromosome 17 \sgcomment{give coordinates} has 35 $Q$-associated variants.
-
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
+By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome(Figure \ref{VariantDensity}).
 

### 2019-05-03 LatexManuscript/legacy.tex
adressed comments

@@
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
-An outlying 10kb region in chromosome 17 \sgcomment{give coordinates} has 35 $Q$-associated variants.
+An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
@@
 By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
-One study reports associations with seven $Q$-associated variants that we find to be highly biased \sgcomment{cite!}\cite{}. 
+One study reports associations with seven $Q$-associated variants that we find to be highly biased \citep{Mandage2017}. 
 That study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype. 
@@
 The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies.
-The logistic factor analysis is more computationally demanding but produces produces more robust results \sgcomment{We never explained this!}.
+The logistic factor analysis is more computationally demanding but produces more robust results \sgcomment{We never explained this!}.
 Both approaches produce comparable results.
@@
 Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
-In total, 57,838,849 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
+In total, 7,786,023 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
 We calculated the mutation spectrum of base pair triplets for the list of significant variants for the JPT population using a similar method as described in  \citep{Harris2017a}.
 
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test \sgcomment{Explain why is this larger than the previous number?}.
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.

### 2019-05-03 LatexManuscript/legacy.tex
added caption titles

--- 2019-05-02 LatexManuscript/legacy.tex
@@
 \caption{
-\textbf{A} 
+Suspicious mutations carried by individuals with low mapping quality have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. \textbf{A} 
 Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
@@
 
-\caption{\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project. 
+\caption{Sampling and sequencing technologies over time in the 1kGP.
+\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project. 
  Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
@@
 \end{tabular}
-\caption{A list of recent publications that reported $Q$-associated variants as close to or above the genome-wide significant threshold.
+\caption{Recent publications that reported $Q$-associated variants as close to or above the genome-wide significant threshold.
 The variants reaching genome wide significance have a star ( * ).
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
+    \caption{Association of SNPS in non-repetitive regions with $Q$.
+    \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.
 There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
+    \caption{Association of indels in non-repetitive regions with $Q$.
+\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
 There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
+    \caption{Association of SNPS in repetitive regions with $Q$.
+\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
 There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
@@
     \end{subfigure} 
-    \caption{\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
+    \caption{Association of indels in repetitive regions with $Q$.
+\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
 There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.
@@
 \includegraphics[width=\hsize,keepaspectratio]{../Figures/Omni_AF.jpg}
-\caption{We found that 613 of the variants we identified as being associated to $Q$ were present on Illumina's Omni 2.5 chip.
-Variants highly associated to $Q$ tend to have low global allele frequencies.}
+\caption{Estimated frequency and association strength of $Q$- associated variants present on Illumina's Omni 2.5 chip. Variants highly associated to $Q$ tend to have low global allele frequencies.}
 \label{Omni}
@@
 \includegraphics[width=\hsize,keepaspectratio]{../Figures/PC1_Correlation.jpg}
-\caption{Plots of data metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
+\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
 The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.
@@
 \includegraphics[width=\hsize,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
-\caption{\textbf{A} 
+
+\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  \textbf{A} 
 The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.

### 2019-05-03 LatexManuscript/legacy.tex
minor updates

@@
 
-In particular, the latter two studies  identified a heterogeneous mutational signature within 1000 Genomes Japanese individuals.
+In particular, the latter two studies  identified a heterogeneous mutational signature within 1kGP Japanese individuals.
 This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
@@
 \caption{
-Suspicious mutations carried by individuals with low mapping quality have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. \textbf{A} 
+Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. \textbf{A} 
 Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
@@
 These three enrichments can be summarized as G**${\rightarrow}$GG*. \textbf{B} 
-Joint frequency spectrum plot of the Japanese from the 1kGP and a more recent Nagahama dataset.
+Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
 Crosses $(+)$ are variants that associate with $Q$ in the JPT. 
@@
 \textbf{C} 
-Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1kGP.
+Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
 This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
-The same analysis was performed independently for each of the populations in the 1kGP.}
+The same analysis was performed independently for each of the populations in the 1000 Genomes Project.}
  \label{SFS}
@@
 
-\caption{Sampling and sequencing technologies over time in the 1kGP.
-\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes sequencing project. 
+\caption{Sampling and sequencing technologies over time in the 1000 Genomes Project.
+\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes Project. 
  Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
@@
 
-\subsection{Identifying suspicious variants in the 1kGP}
+\subsection{Identifying suspicious variants in the 1000 Genomes Project}
 The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 3,826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
+Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
@@
 We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-
-%We apply the 1kGP strict mask after FDR-adjustments to allow the method to identify problematic regions already flagged as such.
 Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
@@
 \end{tabular}
-\caption{Number of statistically significant variants  passing the 1kGP strict mask per category. 
-Variants that are flagged by the 1kGP nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
+\caption{Number of statistically significant variants  passing the 1000 Genomes Project strict mask per category. 
+Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
 A total of 15,270 are statistically significantly associated to $Q$. 
@@
 \subsection{Metadata}
-The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
+The metadata used in this analysis was compiled from each of the index files from the 1kGP file system. 
 Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file.
@@
 \clearpage
-\section{Figures}
+\section{Supplementary Figures}
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/fits_Significant_Positions_CompareModels.jpg}
+\includegraphics[width=13cm,keepaspectratio]{../Figures/fits_Significant_Positions_CompareModels.jpg}
 \caption{Comparison of three logistic regression models for testing association to $Q$.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/CountPerChrom.jpg}
+\includegraphics[width=11cm,keepaspectratio]{../Figures/CountPerChrom.jpg}
 \caption{The proportion of $Q$-associated variants per chromosome.}  
@@
 \caption{The number of $Q$-associated variants per 10kb window across the genome.
-Red bars indicate regions within and black bars indicate regions outside the 1kGP strict mask.
-One region not flagged by the 1kGP strict mask in chromosome 17 has more than 10 variants per window.}  
+Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
+One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.}  
 \label{VariantDensity}
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/RegionOver10.jpg}
+\includegraphics[width=10cm,keepaspectratio]{../Figures/RegionOver10.jpg}
 \caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
@@
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$ data.}
+Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
 \label{90HanSFS_full}
@@
 \centering
-\includegraphics[width=12cm,keepaspectratio]{../Figures/Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1kGP CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
+\includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_singlePop.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
 \label{90HanSFS}
@@
 \centering
-\includegraphics[width=12cm,keepaspectratio]{../Figures/Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1kGP data to the high depth resequence data for variants that, in the 1kGP, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+\includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_FullModel.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
 Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/Omni_AF.jpg}
+\includegraphics[width=12cm,keepaspectratio]{../Figures/Omni_AF.jpg}
 \caption{Estimated frequency and association strength of $Q$- associated variants present on Illumina's Omni 2.5 chip. Variants highly associated to $Q$ tend to have low global allele frequencies.}
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/PC1_Correlation.jpg}
-\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP.
+\includegraphics[width=12cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
+\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project.
 The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.
@@
 \begin{figure}[h]
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
+\includegraphics[width=12cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
 

### 2019-05-03 LatexManuscript/legacy.tex
minot update
minor update

@@
 It may be time to retire or upgrade such legacy sequencing data.
+\vspace{5mm}
+
+\textbf{Key words} : Batch Effect, Mutational Signature, Statistical Genetics, Population Genetics, Reference Cohorts, Imputation
 \end{abstract}
@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-We therefore decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals.
+We therefore decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
 
@@
 
-We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
 Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
@@
 
-\begin{table}
+\begin{table}[h!]
 \centering
@@
 
-\begin{table}[h]
+\begin{table}[h!]
 \begin{tabular}{l l l r r r}

### 2019-06-05 PlottingScripts/ProcessPlotData.Rmd
update

@@
 data$RevAlt<-ReverseComp(as.character(data$Alt))
-
 #only get one half (to fold over)
@@
 data_fam <- fread(paste0(DataPath,
-                         'Significant_Positions_allPop.fam'), 
+                         'Significant_Positions_allPop.fam'), `
                   col.names = c('familyID','sample',

### 2019-06-05 PlottingScripts/ProcessPlotData.Rmd
update

@@
                       substr(binCount$Var1,5,5))
+binCount$Mut <- gsub('->', ' →',binCount$Mut)
 return(binCount)

### 2019-06-12 LatexManuscript/legacy.tex
final update
many updates for reviewer comments

@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
+Here we perform a \luke{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
@@
                       & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
-{SNP}  & 3,369 & 11,059 & 14,428\\  
-{INDEL} & 181 & 657 & 838\\ \hline
+{SNP}  & 3,369 \luke{(0.53\textperthousand)} & 11,059 (0.56\textperthousand) & 14,428\\  
+{INDEL} & 181 (0.3\textperthousand) & 657 (0.66\textperthousand) & 838\\ \hline
 {Total} & 3,550 & 11,716 & 15,270\\
@@
 
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/SNPOverlap6.jpg}
-
-\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-The size of the vertical bars ( I ) are proportional to the -log10(p) value of that SNP.
-The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
-Phase 1 populations are marked by a star ( * ).
-The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$.}
-  \label{OverLap}
-\end{figure}
-
 \subsection{Cell line or technical artifact}
@@
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
-This is more than our nominal false positive rate of 1\% of the sites. Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
+This is more than our nominal false discovery rate of \luke{$\alpha = 0.01$} of the sites. 
+Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
 
@@
 The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
-There are 5 alleles that show differing frequencies between both datasets that are likely explained by biased genotypes.
+There are 5 \luke{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
@@
 \begin{tabular}{l l l r r r}
-  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
-28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
-	& 	&  	rs201130852 & 5.05 & 72.28\\
+  {Pubmed ID}  & {Disease/Trait} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
+28654678 & EBV copy number in & rs201761909 & 5.7 & 78.11\\
+	&   lymphoblastoid cell lines	&  	rs201130852 & 5.05 & 72.28\\
 	& 	&  	rs201255786 & 5.7 & 68.97\\
@@
 	& 	& 	rs200699422 & 5.3 & 7.43\\
-23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
-28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
-26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
-\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
-\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
-\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
-\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
-\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
-\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
-\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
-\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
-\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
+23527680 & \multicolumn{1}{p{3cm}}{\raggedright Attention deficit\\hyperactivity disorder} & rs6057648 & 5.4 & 20.5\\
+28928442 & \multicolumn{1}{p{3cm}}{\raggedright Cold sores} & rs201471471 & 6.52 & 7.87\\
+26053186 & \multicolumn{1}{p{3cm}}{\raggedright HMPMA levels in smokers} & rs60136336 & 5.7 & 2.25\\ \hline
+\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
+\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Prostate cancer}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
+\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
+\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Blood protein levels}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
+\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{High light scatter\\reticulocyte count}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
+\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Response to\\hepatitis B vaccine}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
+\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Metabolic syndrome}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
+\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Alzheimer disease\\and age of onset}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
+\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Intraocular pressure}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
+\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Immunoglobulin A vasculitis}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
+\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Chronic inflammatory diseases}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
+\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Post bronchodilator FEV1/FVC ratio}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
  	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
 	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
-\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\
+\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diisocyanate-induced asthma}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\
 
@@
 The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies.
-The logistic factor analysis is more computationally demanding but produces more robust results \sgcomment{We never explained this!}.
+The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}.
 Both approaches produce comparable results.
@@
 \label{CompareModel}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\hsize,keepaspectratio]{../Figures/SNPOverlap6.jpg}
+
+\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
+The size of the vertical bars ( I ) are proportional to the -log10(p) value of that SNP.
+The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
+Phase 1 populations are marked by a star ( * ).
+The line plot underneath shows the number of populations for which a variant has reached significance.
+The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$.}
+  \label{OverLap}
 \end{figure}

### 2019-06-12 PlottingScripts/ProcessPlotData.Rmd
many updates for reviewer comments

--- 2019-06-05 PlottingScripts/ProcessPlotData.Rmd
@@
                       substr(binCount$Var1,5,5))
+binCount$Mut <- gsub('->', ' →',binCount$Mut)
+
+binCount$Var1 <- NULL
+
 return(binCount)
@@
 rSNP <- processContext(rSNP)
+
+#Chi-square test of independence
+chi_test <- merge(SNP, rSNP, 
+                  by = c('Start','End', 'Mut'), 
+                  all =T)
+
+chi_test$chi <- 
+  (chi_test$Freq.x - chi_test$Freq.y)^2 / chi_test$Freq.y
+
+chi_test$p_val <- pchisq(chi_test$chi, 
+                         df=96, 
+                         lower.tail=FALSE)
+
+SNP <- merge(SNP, 
+             chi_test %>% 
+               select(c('Start','End', 'Mut',
+                        'chi', 'p_val')), 
+             by = c('Start','End', 'Mut'))
 

### 2019-06-13 LatexManuscript/legacy.tex
updated figures and text

@@
 \caption{
-Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. \textbf{A} 
+Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. 
+\textbf{A} 
+Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
+The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
+\textbf{B} 
+Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
+The size of blue dots are proportional to the number of variants that associate with $Q$ in the JPT. 
+\textbf{C} 
 Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
@@
 There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
-These three enrichments can be summarized as G**${\rightarrow}$GG*. \textbf{B} 
-Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
-Crosses $(+)$ are variants that associate with $Q$ in the JPT. 
-The histogram on the left of the plot is the distribution of significant variants. 
-\textbf{C} 
-Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
-This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project.}
+These three enrichments can be summarized as G**${\rightarrow}$GG*. 
+Stars ( * ) indicate significant difference from the mutational spectrum of a set of control SNPs.
+}
  \label{SFS}
@@
  Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
- \textbf{B} The x-axis is sorted by the sequencing date per individual. 
-The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. 
-The bottom plot indicates the sequencing technologies used over time.}
+The shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. 
+\luke{\textbf{B} The sequencing technologies used over time. A change in technology coincides with an increase in overall quality of data produced.}}
 \label{MapQual}
@@
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
+The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
+\luke{Statistical significance of these enrichment is computed using a chi-squared test as described by Harris 2017 \citep{Harris2017a}. }
 
@@
 	& 	& 	rs200699422 & 5.3 & 7.43\\
-23527680 & \multicolumn{1}{p{3cm}}{\raggedright Attention deficit\\hyperactivity disorder} & rs6057648 & 5.4 & 20.5\\
+23527680 & \multicolumn{1}{p{3cm}}{\raggedright ADHD\textsuperscript{\dag}} & rs6057648 & 5.4 & 20.5\\
 28928442 & \multicolumn{1}{p{3cm}}{\raggedright Cold sores} & rs201471471 & 6.52 & 7.87\\
-26053186 & \multicolumn{1}{p{3cm}}{\raggedright HMPMA levels in smokers} & rs60136336 & 5.7 & 2.25\\ \hline
+26053186 & \multicolumn{1}{p{3cm}}{\raggedright HMPMA\textsuperscript{\ddag} levels in smokers} & rs60136336 & 5.7 & 2.25\\ \hline
 \textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
@@
 \textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Blood protein levels}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{High light scatter\\reticulocyte count}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
-\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Response to\\hepatitis B vaccine}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
-\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Metabolic syndrome}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
+\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{High light scatter\\reticulocyte count}} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
+\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Response to\\hepatitis B vaccine}} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
+\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Metabolic syndrome}}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
 \textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Alzheimer disease\\and age of onset}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
@@
 \textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Chronic inflammatory diseases}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
-\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Post bronchodilator FEV1/FVC ratio}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
- 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
+\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Post bronchodilator}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
+ 	& \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{FEV1/FVC ratio}} 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
 	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
@@
 The variants reaching genome wide significance have a star ( * ).
-The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure \ref{Pub_30}).}
+The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure \ref{Pub_30}).
+\textsuperscript{\dag} Attention deficit hyperactivity disorder. 
+\textsuperscript{\ddag} 3-hydroxy-1-methylpropylmercapturic acid.
+}
 \label{gwasTable}

### 2019-06-14 PlottingScripts/ProcessPlotData.Rmd

@@
 ```
+
+extract context for GCAT test snps of interest
+```{bash}
+#Using GenomeWide.AncestralContext.txt
+#To get mutation context, use $path/AncestralContext.py
+#compare first two columns in both files, extract matches
+path='/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/'
+
+if [ ! -f $path/gcat.sig.pos.context.txt ]; then
+  echo "sig pos context"
+  awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
+  $path/Significant_Positions.txt \
+  $path/GenomeWide.AncestralContext.txt \
+  > $path/gcat.sig.pos.context.txtt
+fi 
+```
+
 Load mutation context and plot enrichment
@@
 
-#Chi-square test of independence
+#chi-squared contingency table tests
 chi_test <- merge(SNP, rSNP, 
@@
 
-chi_test$chi <- 
-  (chi_test$Freq.x - chi_test$Freq.y)^2 / chi_test$Freq.y
+chi_test[is.na(chi_test)] <- 0
 
-chi_test$p_val <- pchisq(chi_test$chi, 
-                         df=96, 
-                         lower.tail=FALSE)
+chi_test$chi_p <- 
+  lapply(split(chi_test %>% 
+                 select(Freq.x,Freq.y), 
+               (1:nrow(chi_test) - 1)), 
+         function(data) 
+           chisq.test(data)$p.value)
 
@@
              chi_test %>% 
-               select(c('Start','End', 'Mut',
-                        'chi', 'p_val')), 
+               select(c('Start','End', 'Mut','chi_p')), 
              by = c('Start','End', 'Mut'))

### 2019-06-14 LatexManuscript/legacy.tex
edit

@@
                       & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
-{SNP}  & 3,369 \luke{(0.53\textperthousand)} & 11,059 (0.56\textperthousand) & 14,428\\  
-{INDEL} & 181 (0.3\textperthousand) & 657 (0.66\textperthousand) & 838\\ \hline
+{SNP}  & 3,369 \textcolor{gray}{0.53\textperthousand} & 11,059 \textcolor{gray}{0.56\textperthousand} & 14,428\\  
+{INDEL} & 181 \textcolor{gray}{0.3\textperthousand} & 657 \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
 {Total} & 3,550 & 11,716 & 15,270\\

### 2019-06-14 LatexManuscript/legacy.tex
edits

@@
 \usepackage{hyperref}
+\usepackage{ulem}
 \hypersetup{
@@
 While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and suspicious GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
-It may be time to retire or upgrade such legacy sequencing data.
+Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and \luke{some questionable \sout{suspicious}} GWAS associations.
+Lower-quality data from the early phases of the 1kGP thus continues to \luke{impact \sout{contaminate}} modern studies in hidden ways.
+\luke{As technologies and analytical methods improve, it is important that we renew and update these data resources accordingly}\sout{It may be time to retire or upgrade such legacy sequencing data.}
 \vspace{5mm}

### 2019-06-14 PlottingScripts/ProcessPlotData.Rmd
edits

@@
                           levels = unique(metaData$sample))
+rownames(metaData) <- NULL
+metaData$order <- rownames(metaData)
 

### 2019-06-17 LatexManuscript/legacy.tex
updated reviewer suggestions

@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-We therefore decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
+\luke{Given this signature had not been reproduced in another dataset}, we \sout{therefore} decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
+\luke{Our initial goal was to explore the population structure that might be responsible for such an intriguing mutational signature.}
 
@@
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
+The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
@@
 
-Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques (Figure \ref{MutSpect}).
+Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques \luke{due to lack of power (Supplementary Figure 000}).
 The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
@@
 The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
+\luke{In the GCAT test, a variant may be significantly associated to $Q$ in some populations, but not necessarily all populations. 
+Thus, the 1,139 variants present in the resequenced Han might have stronger associations to $Q$ in in individuals from other populations.}
 There are 5 \luke{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
+\luke{These  $Q$-associated variants could also compromise the imputation of real variants that happen to be  close by, causing not only improper imputation, but also false negatives for real variants.}
 
-We searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
+We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
+\luke{Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
+Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
+However, the fact that some of these variants are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.}
 
@@
 One study reports associations with seven $Q$-associated variants that we find to be highly biased \citep{Mandage2017}. 
-That study considered copy number of Epstein-Barr virus sequence in the 1kGP as a phenotype. 
+That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
+\luke{It is unlikely that EBV copy number associations to these seven $Q$-associated variants could be explained by population structure as the data quality of an individual is also correlated with sequencing date.
+This suggests that there may be issues with how EBV is mapped to the genomes of individuals with low quality sequence data.}
 Thus the phenotype in that study is likely confounded by the same technical artefacts that lead to biased SNP calling.

### 2019-06-18 LatexManuscript/legacy.tex
updated supplementary plots, and legends.

@@
 While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and \luke{some questionable \sout{suspicious}} GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus continues to \luke{impact \sout{contaminate}} modern studies in hidden ways.
-\luke{As technologies and analytical methods improve, it is important that we renew and update these data resources accordingly}\sout{It may be time to retire or upgrade such legacy sequencing data.}
+Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and \todo{some questionable \sout{suspicious}} GWAS associations.
+Lower-quality data from the early phases of the 1kGP thus continues to \todo{impact \sout{contaminate}} modern studies in hidden ways.
+\todo{As technologies and analytical methods improve, it is important that we renew and update these data resources accordingly}\sout{It may be time to retire or upgrade such legacy sequencing data.}
 \vspace{5mm}
@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\luke{Given this signature had not been reproduced in another dataset}, we \sout{therefore} decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
-\luke{Our initial goal was to explore the population structure that might be responsible for such an intriguing mutational signature.}
+\todo{Given this signature had not been reproduced in another dataset}, we \sout{therefore} decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
+\todo{Our initial goal was to explore the population structure that might be responsible for such an intriguing mutational signature.}
 
@@
 The shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. 
-\luke{\textbf{B} The sequencing technologies used over time. A change in technology coincides with an increase in overall quality of data produced.}}
+\todo{\textbf{B} The sequencing technologies used over time. A change in technology coincides with an increase in overall quality of data produced.}}
 \label{MapQual}
@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a \luke{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
+Here we perform a \todo{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
@@
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-\luke{Statistical significance of these enrichment is computed using a chi-squared test as described by Harris 2017 \citep{Harris2017a}. }
+\todo{Statistical significance of these enrichment is computed using a chi-squared test as described by Harris 2017 \citep{Harris2017a}. }
 
-Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques \luke{due to lack of power (Supplementary Figure 000}).
+Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques \todo{due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps})}.
 The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
@@
 \subsection{Identifying suspicious variants in the 1000 Genomes Project}
-The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
+The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual} and Figure \ref{quality_by_center}).
 This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
@@
 By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome(Figure \ref{VariantDensity}).
+\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
+However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
+These three enrichments can be summarized as T*T${\rightarrow}$TTT.}
 
@@
 A total of 15,270 are statistically significantly associated to $Q$. 
+The grey text is the proportion of significant associations per category.
 The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
@@
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
-This is more than our nominal false discovery rate of \luke{$\alpha = 0.01$} of the sites. 
+This is more than our nominal false discovery rate of \todo{$\alpha = 0.01$} of the sites. 
 Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
@@
 The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
-\luke{In the GCAT test, a variant may be significantly associated to $Q$ in some populations, but not necessarily all populations. 
+\todo{In the GCAT test, a variant may be significantly associated to $Q$ in some populations, but not necessarily all populations. 
 Thus, the 1,139 variants present in the resequenced Han might have stronger associations to $Q$ in in individuals from other populations.}
-There are 5 \luke{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
+There are 5 \todo{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-\luke{These  $Q$-associated variants could also compromise the imputation of real variants that happen to be  close by, causing not only improper imputation, but also false negatives for real variants.}
+\todo{These  $Q$-associated variants could also compromise the imputation of real variants that happen to be  close by, causing not only improper imputation, but also false negatives for real variants.}
 
@@
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-\luke{Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
+\todo{Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
 Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
@@
 That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
-\luke{It is unlikely that EBV copy number associations to these seven $Q$-associated variants could be explained by population structure as the data quality of an individual is also correlated with sequencing date.
+\todo{It is unlikely that EBV copy number associations to these seven $Q$-associated variants could be explained by population structure as the data quality of an individual is also correlated with sequencing date.
 This suggests that there may be issues with how EBV is mapped to the genomes of individuals with low quality sequence data.}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering
@@
 \label{CompareModel}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=13cm,keepaspectratio]{../Figures/histogram_of_sig_snps.jpg}
+\caption{Frequency distribution of the 1034 $Q$ associated SNPs from the single population test in the JPT in blue and all SNPs in black.
+The association test lacks power to identify variants at low frequencies.}
+\label{histogram_of_sig_snps}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=13cm,keepaspectratio]{../Figures/quality_by_center.jpg}
+\caption{Sequencing date of samples included in the 1kGP coloured by sequencing centre. Improvements in sequencing platforms coincides with improvement in sequencing quality.}  
+\label{quality_by_center}
 \end{figure}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
+\centering
+\includegraphics[width=13cm,keepaspectratio]{../Figures/strict_gcat_mutational_enrichment.jpg}
+\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
+A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
+There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
+These three enrichments can be summarized as T*T${\rightarrow}$TTT.
+Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
+\label{strict_gcat_mutational_enrichment}
+\end{figure}
+
+\begin{figure}[ht]
 \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/Omni_AF.jpg}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}

### 2019-06-18 PlottingScripts/ProcessPlotData.Rmd
updated supplementary plots, and legends.

@@
 ```{r}
+########### JPT 1kGP ###########
 #get positions of significant SNPs
@@
 rm(NotSig) #save memory
+
+########### GCAT 1kGP ###########
+#see RegressionResults.rmd for Regression
+
+#get positions of random SNPs as control
+set.seed(123)
+
+number_of_sig <- 
+  nrow(Regression[which(
+    Regression$Adjusted_Log10p 
+    > -log10(0.01)),])
+
+gcat.random.pos <- 
+  Regression[sample(nrow(Regression), 
+                    number_of_sig),
+             c('Chr','Pos')]
+
+write.table(gcat.random.pos, 
+            file = paste0(DataPath,
+                        'gcat.random.pos.txt'), 
+            col.names = FALSE, 
+            row.names = FALSE, 
+            quote = FALSE, 
+            sep = '\t')
+
+strict_sig <- nrow(RegressionMasked[which(
+    RegressionMasked$Adjusted_Log10p 
+    > -log10(0.01)),])
+
+strict.gcat.random.pos <- 
+  RegressionMasked[sample(nrow(RegressionMasked), 
+                    strict_sig),
+             c('Chr','Pos')]
+
+write.table(strict.gcat.random.pos, 
+            file = paste0(DataPath,
+                        'strict.gcat.random.pos.txt'), 
+            col.names = FALSE, 
+            row.names = FALSE, 
+            quote = FALSE, 
+            sep = '\t')
+
+strict.gcat.pos <- 
+  RegressionMasked[which(
+    RegressionMasked$Adjusted_Log10p 
+    > -log10(0.01)),
+    c('Chr','Pos')]
+
+write.table(strict.gcat.pos, 
+            file = paste0(DataPath,
+                        'strict.gcat.pos.txt'), 
+            col.names = FALSE, 
+            row.names = FALSE, 
+            quote = FALSE, 
+            sep = '\t')
+
 ```
@@
 #compare first two columns in both files, extract matches
-path='/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/'
+out_path='/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/'
+path='/Users/luke/Documents/LegacyData/PlottingScripts/ProcessedData/'
 
-if [ ! -f $path/gcat.sig.pos.context.txt ]; then
+if [ ! -f $path/strict.gcat.random.pos.context.txt ]; then
+  echo "gcat random pos context"
+  awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
+  $path/strict.gcat.random.pos.txt \
+  $path/GenomeWide.AncestralContext.txt \
+  > $path/strict.gcat.random.pos.context.txt
+fi
+
+if [ ! -f $path/strict.gcat.sig.pos.context.txt ]; then
   echo "sig pos context"
   awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
-  $path/Significant_Positions.txt \
+  $path/strict.gcat.pos.txt \
   $path/GenomeWide.AncestralContext.txt \
-  > $path/gcat.sig.pos.context.txtt
+  > $path/strict.gcat.sig.pos.context.txt
 fi 
@@
 
+#toggle for all snps or just JPT 1kGP snps
+name <- 'strict.gcat.' #'gcat.' # ''
+
 #load SNP context
-SNP <- read.table(paste0(DataPath,
-                         'sig.pos.context.txt'), 
+SNP <- read.table(paste0(DataPath, name,'sig.pos.context.txt'), 
                 sep = '\t', 
@@
                 col.names = contextHeader)
-rSNP <- read.table(paste0(DataPath,
+
+rSNP <- read.table(paste0(DataPath, name,
                           'random.pos.context.txt'), 
@@
 saveRDS(SNP,
-            paste0(DataPath,
+            paste0(DataPath, name,
                    'SNP_Context.rds'))
@@
 saveRDS(rSNP,
-            paste0(DataPath,
+            paste0(DataPath, name,
                    'rSNP_Context.rds'))

### 2019-06-20 LatexManuscript/legacy.tex
update

@@
 Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
-This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual
+This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual.
 The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
@@
 A total of 15,270 are statistically significantly associated to $Q$. 
-The grey text is the proportion of significant associations per category.
+\todo{The grey text is the proportion of significant associations per category.}
 The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}

### 2019-06-20 LatexManuscript/legacy.tex
some small edits

@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}
+\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
 \caption{
@@
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-\todo{Statistical significance of these enrichment is computed using a chi-squared test as described by Harris 2017 \citep{Harris2017a}. }
+\todo{Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}. }
 
@@
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
-By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome(Figure \ref{VariantDensity}).
+By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
 \todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.

### 2019-06-25 PlottingScripts/ProcessPlotData.Rmd
Updated figure 1 with pluralization and finalized significance test.

@@
 #toggle for all snps or just JPT 1kGP snps
-name <- 'strict.gcat.' #'gcat.' # ''
+name <-''#'strict.gcat.' #'gcat.' # '' #
 
@@
 
-rSNP <- read.table(paste0(DataPath, name,
-                          'random.pos.context.txt'), 
+rSNP <- read.table(paste0(DataPath,'GenomeWide.AncestralContext.txt'), 
                 sep = '\t',
@@
 SNP <- processContext(SNP)
+
 rSNP <- processContext(rSNP)
+rSNP <- 
+  rSNP %>%
+  filter(Start != "N" &
+           End != "N")
+
+#get proportions
+total_count <- sum(rSNP$Freq)
+total_sample <- sum(SNP$Freq)
+rSNP$Freq <- rSNP$Freq / total_count * total_sample
 

### 2019-06-25 LatexManuscript/legacy.tex
text update

@@
 While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and \todo{some questionable \sout{suspicious}} GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus continues to \todo{impact \sout{contaminate}} modern studies in hidden ways.
-\todo{As technologies and analytical methods improve, it is important that we renew and update these data resources accordingly}\sout{It may be time to retire or upgrade such legacy sequencing data.}
+Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
+Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
+It may be time to retire or upgrade such legacy sequencing data.
 \vspace{5mm}
@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\todo{Given this signature had not been reproduced in another dataset}, we \sout{therefore} decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
-\todo{Our initial goal was to explore the population structure that might be responsible for such an intriguing mutational signature.}
+Given that this signature had not yet been reproduced in another dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
 
@@
 \textbf{C} 
-Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to a random set of SNPs.
+Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the proportional distribution of all SNPs.
 The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
-There is also a slight enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
+There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
 These three enrichments can be summarized as G**${\rightarrow}$GG*. 
-Stars ( * ) indicate significant difference from the mutational spectrum of a set of control SNPs.
+Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
 }
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
+\todo{Our initial goal was to determine if this signature could be explained by population structure or technological error.}
 While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
@@
 The shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. 
-\todo{\textbf{B} The sequencing technologies used over time. A change in technology coincides with an increase in overall quality of data produced.}}
+\textbf{B} Sequencing technologies used over the course of the 1000 Genomes Project.}
 \label{MapQual}
@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a \todo{genotype conditional association test (GCAT), where} genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
+Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
@@
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-\todo{Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}. }
+Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}.
 
-Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques \todo{due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps})}.
+Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
 The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).

### 2019-06-25 PlottingScripts/ProcessedData/ScoopCheck.Rmd
new figure

---
title: "Comparing Results with Mafessoni et al."
author: 
- name: Luke Anderson-Trocmé
  affiliation:
  - &cruk Department of Human Genetics, McGill University,  Montreal, Canada
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

load libraries
```{r, message = FALSE}
library(ggplot2)
library(data.table)
library(tidyverse)

#Directories
DataPath <- '~/Documents/LegacyData/PlottingScripts/ProcessedData/'
figurePath <- '~/Documents/LegacyData/Figures/'
```

define functions
```{r, include = FALSE}
ReverseComp <- function(x)
        chartr("ATGC","TACG",
        sapply(
          lapply(
          strsplit(x, NULL), 
          rev), 
          paste, 
          collapse = ""))

processContext <- function(data){

#get reverse complement of context, ref and alt
data$Rev<-ReverseComp(as.character(data$Context))
data$RevRef<-ReverseComp(as.character(data$Ref))
data$RevAlt<-ReverseComp(as.character(data$Alt))
#only get one half (to fold over)
AC<-data[which(data$Ref 
               %in% 
                 c('A','C')),]
TG<-data[which(data$Ref 
               %in% 
                 c('T','G')),]
#fold over
TG$Context <- TG$Rev
TG$Ref <- TG$RevRef
TG$Alt <- TG$RevAlt
#bind them together
data <- rbind(AC,TG)
#combine it all together to get a context
data$Mut <- paste0(data$Ref,'->', 
                substr(data$Context,1,1),
                data$Alt,
                substr(data$Context,3,3))
#get the counts of each bin
binCount <- as.data.frame(table(data$Mut))
#process for plotting
binCount$Start <- substr(binCount$Var1,4,4)
binCount$End <- substr(binCount$Var1,6,6)
binCount$Mut <- paste(substr(binCount$Var1,1,3),
                      substr(binCount$Var1,5,5))
binCount$Mut <- gsub('->', ' →',binCount$Mut)

binCount$Var1 <- NULL

return(binCount)
}

plotContext <- function(data, name){

plt <- ggplot(data, 
              aes(x = End, 
                  y = Start, 
                  fill = Freq)) +
    geom_tile() +
    facet_grid(.~Mut) +
    labs(y = name) +
    theme_classic() +
    theme(axis.title.x = element_blank(),
          axis.line = element_blank(),
          axis.text = element_text(size = 20),
          axis.title.y = element_text(size = 20),
          strip.placement = "outside",
          strip.background = element_blank(),
          strip.text = element_text(size = 20),
          legend.text = element_text(size = 20),
          legend.title = element_text(size = 20),
          panel.border = element_rect(colour = "black", 
                                      fill = NA, 
                                      size = 1),
          plot.margin = unit(c(0,0,0,1.5), "lines")) +
    scale_fill_gradient(limits = c(0, maxFreq),
                        low = 'white', 
                        high = 'royalblue1',
                        guide = guide_legend(title = 'Variant\ncount',
                                             direction = 'vertical'))
return(plt)
}
```

load data
```{r}
t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)
t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)

us <- fread('/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/Significant_Positions_and_Pvals.txt')

totalT<- unique(rbind(t4,t5,t6,t7,t8))
totalTus <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
totalTusX <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
totalTusX[is.na(totalTusX)] <- 1

```

```{r}
p_comp <- 
  ggplot(totalTusX, 
       aes(x=Adjusted_Log10p, 
           y= -log10(V5))) +
  geom_point(shape = 1) +
  labs(x = 'Q-association',
       y = 'LD-association')

p_comp
```

```{r}
mafessoni_snps <- 
  totalT[,c('V1','V3')]

write.table(mafessoni_snps, 
            file = paste0(DataPath,
                        'mafessoni_snps.pos.txt'), 
            col.names = FALSE, 
            row.names = FALSE, 
            quote = FALSE, 
            sep = '\t')

```

extract context for Mafessoni SNPs
```{bash}
#Using GenomeWide.AncestralContext.txt
#To get mutation context, use $path/AncestralContext.py
#compare first two columns in both files, extract matches
out_path='/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/'
path='/Users/luke/Documents/LegacyData/PlottingScripts/ProcessedData/'

if [ ! -f $path/mafessoni_snps.pos.context.txt ]; then
  echo "pos context"
  awk 'NR==FNR{a[$1,$2];next} ($1,$2) in a' \
  $path/mafessoni_snps.pos.txt \
  $path/GenomeWide.AncestralContext.txt \
  > $path/mafessoni_snps.pos.context.txt
fi  
```

```{r}
contextHeader = c('Chr','Pos','Ref','Alt','Flip','Context')

#load SNP context
mSNP <- read.table(paste0(DataPath, 
                          'mafessoni_snps.pos.context.txt'), 
                sep = '\t', 
                header  = FALSE,
                col.names = contextHeader)

mSNP <- processContext(mSNP)

maxFreq <- max(mSNP$Freq)
mSNPplt <- plotContext(mSNP,'Mafessoni')

m_plot <-
  plot_grid(p_comp,
          mSNPplt,
          ncol = 1,
          rel_heights = c(2, 1))

ggsave(paste0(figurePath,
              'Mafessoni_MutSpect.jpg'),
       m_plot, 
       height = 10,
       width = 18)

m_plot
```

```{r}

```

```{r}

```

### 2019-06-25 LatexManuscript/legacy.tex
updated text and bib

@@
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Here we point out how large and previously unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
+\todo{Mafessoni et al.
+recently identified a batch effect in the 1kGP using an approach that identifies individuals and loci with unusual linkage disequelibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}
+Here we point out how large and largely unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
 
@@
 
+\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a different method.
+We find that only 10\% of the variants they identified are in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
+Moreover, the variants they identified are not enriched in the mutational spectra described above \ref{Mafessoni_MutSpect}. }
+
 \begin{table}[h!]
@@
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
-This is more than our nominal false discovery rate of \todo{$\alpha = 0.01$} of the sites. 
+\todo{This is more than the number of false positives predicted by our nominal false discovery rate of $\alpha = 0.01$ ($0.01 \times 296 = 2.96$)}. 
 Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
@@
 The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
-\todo{In the GCAT test, a variant may be significantly associated to $Q$ in some populations, but not necessarily all populations. 
-Thus, the 1,139 variants present in the resequenced Han might have stronger associations to $Q$ in in individuals from other populations.}
-There are 5 \todo{variants} that show differing frequencies between both datasets that are likely explained by biased genotypes.
+\todo{Some mutational signatures are heterogenous where not all suspicious variants are completely spurious in all individuals or populations. 
+Some mutational signatures reproduce across cohorts, whereas others do not and may be due to the different sequencing platforms used.}
+There are 5 variants that show differing frequencies between both datasets that are likely explained by biased genotypes.
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-\todo{These  $Q$-associated variants could also compromise the imputation of real variants that happen to be  close by, causing not only improper imputation, but also false negatives for real variants.}
+These $Q$-associated variants could also compromise the imputation of nearby real-variants causing not only improper imputation, but also false negatives for real variants.
 
@@
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-\todo{Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
+Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
 Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
-However, the fact that some of these variants are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.}
+However, the fact that some of these variants are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
 
@@
 That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
-\todo{It is unlikely that EBV copy number associations to these seven $Q$-associated variants could be explained by population structure as the data quality of an individual is also correlated with sequencing date.
-This suggests that there may be issues with how EBV is mapped to the genomes of individuals with low quality sequence data.}
-Thus the phenotype in that study is likely confounded by the same technical artefacts that lead to biased SNP calling.
+\todo{It is plausible that the EBV copy number phenotype is sensitive to mapping and in this study, may be confounded by the same technical artefacts that lead to biased SNP calling.}
 
@@
 \centering
+\includegraphics[width=13cm,keepaspectratio]{../Figures/Mafessoni_MutSpect.jpg}
+\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
+\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
+\label{Mafessoni_MutSpect}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
 \includegraphics[width=11cm,keepaspectratio]{../Figures/CountPerChrom.jpg}

### 2019-07-03 LatexManuscript/legacy.tex
updated text

@@
 \todo{Mafessoni et al.
-recently identified a batch effect in the 1kGP using an approach that identifies individuals and loci with unusual linkage disequelibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}
-Here we point out how large and largely unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
+recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}8
+Here we point out how these and additional unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
 
@@
 \subsection{Cell line or technical artifact}
+\todo{
+To assess whether or not the variants we've identified are the result of cell line or technical artifact, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
+Variants that do not reproduce across these cohorts are likely the result of technical artifacts, however variants that do reproduce across cohorts are either the result of a cell line artifact or biased genotypes.
+ }
 
@@
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
-\todo{This is more than the number of false positives predicted by our nominal false discovery rate of $\alpha = 0.01$ ($0.01 \times 296 = 2.96$)}. 
+\todo{This is more than the 2.96 predicted false positives according to our nominal false discovery rate of $\alpha = 0.01$}. 
 Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
@@
 From this subset, only 1,139 (or 18\%) are present in the resequenced data.
+\todo{The higher rate of variants that reproduce across cohorts might be due to the fact that the average effect sizes are smaller in the GCAT model.}
 The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
-\todo{Some mutational signatures are heterogenous where not all suspicious variants are completely spurious in all individuals or populations. 
-Some mutational signatures reproduce across cohorts, whereas others do not and may be due to the different sequencing platforms used.}
 There are 5 variants that show differing frequencies between both datasets that are likely explained by biased genotypes.
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
+%Some mutational signatures are heterogenous where not all suspicious variants are completely spurious in all individuals or populations. 
+%Some mutational signatures reproduce across cohorts, whereas others do not and may be due to the different sequencing platforms used.
 
@@
 \section{Discussion}
+
+Recommendations : doing their approach as well as ours. We recommend removing the variants we ID and the ones they find. Regression with QC metrics (shoutout to them)
 
@@
 \section{Supplementary Figures}
+
 \renewcommand{\thefigure}{S\arabic{figure}}
@@
 
-\begin{figure}[ht]
+\begin{figure}[tbp]
 \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}[tbp]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/Omni_AF.jpg}
@@
 
-\begin{figure}[ht]
+\begin{figure}[tbp]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
@@
 
-\begin{figure}[ht]
+\begin{figure}[h!]
 \includegraphics[width=12cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
-
 \caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  \textbf{A} 

### 2019-07-04 LatexManuscript/legacy.tex
comments throughout

@@
 \todo{Mafessoni et al.
-recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}8
-Here we point out how these and additional unreported batch effects in the early phases of the 1kGP still lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
+recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}\sgcomment{You are talking about things that have not been introduced yet...
+Maybe: Mafessoni et al.
+recently identified batch effects in the 1kGP by looking for individuals with excess linkage disequilibrium among distant variants.
+}
+Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
 
@@
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
-In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in a number of recent publications.
+In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications and online resources.
 
@@
 \caption{
-Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome. 
+Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies \sgcomment{, and are distributed across the genome--this is outdated}.
 \textbf{A} 
 Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
-This GWAS identified $587\ \  p < 10^{-8}$ and $1034\ \ p < 10^{-6}$ SNPs that were associated to the average $Q$ of SNPs mapped for an individual.
-The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
+This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
+($587$ SNPs had  $p < 10^{-8}.$)
+\sgcomment{Why do we say the following?} The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
 \textbf{B} 
 Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
-The size of blue dots are proportional to the number of variants that associate with $Q$ in the JPT. 
+The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
 \textbf{C} 
-Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the proportional distribution of all SNPs.
+Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
 The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-\todo{Our initial goal was to determine if this signature could be explained by population structure or technological error.}
+\todo{Our initial goal was to determine if \sgcomment{whether?} this signature could be explained by population structure or technological error.}
 While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
@@
  Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
-The shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. 
+The shape indicates whether the individual \sgcomment{belongs to -- could we clarify this?} Phase 1 or Phase 3 of the 1000 Genomes project. 
 \textbf{B} Sequencing technologies used over the course of the 1000 Genomes Project.}
@@
 By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
-\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
+\sgcomment{There seems to be a paragraph break needed, a logical link missing, or both?}\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
 However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
@@
 \todo{
-To assess whether or not the variants we've identified are the result of cell line or technical artifact, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
-Variants that do not reproduce across these cohorts are likely the result of technical artifacts, however variants that do reproduce across cohorts are either the result of a cell line artifact or biased genotypes.
+To assess whether or not $Q$-associated variants results from mutations occurring in the creation or maintenance of the cell lines, or from technical artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
+Variants that do not reproduce across these cohorts are likely the result of technical artifacts.
+By contrast, if the same mutations are observed at identical frequencies in independently sequenced cohorts, we can conclude that recurrent cell line mutations are likely to blame.
+Finally, to complicate matters, we may have a third situation where a real variant has a genotyping error probability that correlates with $Q$.
+In this case, we expect the variant may reproduce, but at different frequency.
  }
@@
 resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
-To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
+%To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}). 
-\todo{This is more than the 2.96 predicted false positives according to our nominal false discovery rate of $\alpha = 0.01$}. 
-Thus a small number of variants associated to $Q$ are present in the population but with somewhat biased genotypes.
+This is more than the 2.96 predicted false positives according to our nominal false discovery rate of $\alpha = 0.01$. 
+Thus a small number of variants associated to $Q$ in the JPT are present in the population but with biased genotypes.
 
@@
 Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
-From this subset, only 1,139 (or 18\%) are present in the resequenced data.
-\todo{The higher rate of variants that reproduce across cohorts might be due to the fact that the average effect sizes are smaller in the GCAT model.}
-The allele frequencies of these variants are nearly identical between datasets suggesting that among these 83 individuals, these variants are properly genotyped in the 1kGP.
-There are 5 variants that show differing frequencies between both datasets that are likely explained by biased genotypes.
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
+Five variants show differing frequencies between both datasets and are likely explained by biased genotypes.
+Finally, 1,139 (or 18\%) are present in the resequenced data.
+The allele frequencies of these variants are nearly identical between datasets suggesting that variant calls are accurate for these 83 individuals.
+This is likely because these variants have biased genotypes in populations other than the CHB/CHD.
+
 %Some mutational signatures are heterogenous where not all suspicious variants are completely spurious in all individuals or populations. 
@@
 
-Among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
+Finally, among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
 These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
@@
 They also removed rare alleles and alleles with high degrees of missingness.
-Despite using state-of-the-art quality controls, these variants managed not only to be imputed onto real genotype data, but they also reached genome wide significance for association with biological traits.
+Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
 

### 2019-07-04 PlottingScripts/ProcessPlotData.Rmd
updated text and supplementary figures.

@@
 ```
+
+##Imputation Scores
+```{r}
+library(data.table)
+library(spatstat)
+library(rdist)
+library(Rfast)
+#Imputation scores for Japanese genotyped individuals
+suspicious <- fread('/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/Significant_Positions_and_Pvals.txt')
+#set path and array name
+jpt_imp_path = "/Users/luke/Documents/Legacy_Misc/Misc/ImputationScores/"
+array = "610k" #2500k_2500o #ASA #CoreExome24
+#list of filenames
+jpt_imp_files = list.files(path=jpt_imp_path, 
+                           pattern=paste0(array,".imputed.1.info.gz"),
+                           full.names = T) 
+# Read the files
+jpt_imp <- lapply(jpt_imp_files, 
+                  function(x) {fread(paste('gzcat', x), 
+                                          header = T, 
+                                          sep ="\t")})
+# Combine them
+jpt_imp <- do.call("rbind", 
+                   lapply(jpt_imp, as.data.frame))
+
+jpt_imp <- separate(data = jpt_imp, 
+                    col = SNP, 
+                    into = c("Chr", "Pos"), 
+                    sep = ":")
+
+jpt_imp <- merge(jpt_imp, suspicious, 
+                 by = c('Chr', 'Pos'), all.x = T)
+
+jpt_imp$Pos <- as.numeric(as.character(jpt_imp$Pos))
+
+jpt_imp
+```
+
+```{r}
+
+imputed_position <- jpt_imp$Pos
+suspicious_position <- jpt_imp[which(jpt_imp$Adjusted_Log10p > 2),]$Pos
+
+jpt_imp$closest <- unlist(lapply(imputed_position, function(x) suspicious_position[which.min(abs(suspicious_position - x))]))
+jpt_imp$dist <- abs(jpt_imp$closest - jpt_imp$Pos)
+
+jpt_imp$dist_2 <- round(jpt_imp$dist, -1)
+
+jpt_imp_sub <- jpt_imp %>%
+  select(AvgCall, Rsq, dist_2) %>%
+  group_by(dist_2) %>%
+  summarise(AvgCall_mean = mean(AvgCall),
+            Rsq_mean = mean(Rsq))
+
+#plot distance from nearest suspicious variant 
+Rsq <- 
+  ggplot(jpt_imp_sub, 
+       aes(x= dist_2, 
+           y = Rsq_mean)) +
+  geom_point() +
+  labs(x = "Distance from nearest suspicious variant",
+       y = "Rsq") +
+  xlim(c(0,10000))
+
+AvgCall <- 
+  ggplot(jpt_imp_sub, 
+       aes(x= dist_2, 
+           y = AvgCall_mean)) +
+  geom_point() +
+  labs(x = "Distance from nearest suspicious variant",
+       y = "Average Call") +
+  xlim(c(0,10000))
+
+plot_grid(AvgCall, Rsq, ncol=1)
+
+ggsave(paste0(figurePath,
+              "Imputation_Distance_10s.jpg"),
+       height = 9,
+       width = 9)
+```
+
+```{r}
+
+```

### 2019-07-04 PlottingScripts/ProcessedData/ScoopCheck.Rmd
updated text and supplementary figures.

@@
 ```{r}
+#load regression results
+Regression <- readRDS('/Users/luke/Documents/Legacy_Misc/Misc/Regression.Rmd')
+#load supplementary files from Mafessoni
 t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
-t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
+
+#t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
+
 t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)
-t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
-t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)
 
-us <- fread('/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/Significant_Positions_and_Pvals.txt')
+#t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
 
-totalT<- unique(rbind(t4,t5,t6,t7,t8))
-totalTus <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
-totalTusX <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
+#t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)
+
+#get total unique positions from most conservative tables (i.e cutoff of 1% MAF)
+totalT<- unique(rbind(t4,t6))
+
+#totalTus <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
+totalTusX <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
 totalTusX[is.na(totalTusX)] <- 1
 
+totalTusX
 ```
@@
 ```{r}
+#select rows of interest
+#either present in Mafessoni OR significant in our data
+library(ggExtra)
+
+to_plot <- totalTusX %>%
+           filter(V5 < 1 | Adjusted_Log10p > 2)
+
 p_comp <- 
-  ggplot(totalTusX, 
+  ggplot(to_plot, 
        aes(x=Adjusted_Log10p, 
            y= -log10(V5))) +
-  geom_point(shape = 1) +
+  geom_point(shape = 1,
+             size = 1,
+             alpha = 0.3) +
   labs(x = 'Q-association',
-       y = 'LD-association')
+       y = 'LD-association') +
+  theme_classic() +
+  theme(axis.title = element_text(size = 20),
+        axis.text = element_text(size = 20))
+
+p_comp <- ggExtra::ggMarginal(p_comp, type = "histogram")
 
@@
 maxFreq <- max(mSNP$Freq)
-mSNPplt <- plotContext(mSNP,'Mafessoni')
+mSNPplt <- plotContext(mSNP,'LD-associated')
 
@@
           ncol = 1,
-          rel_heights = c(2, 1))
+          rel_heights = c(4, 1))
 
@@
        m_plot, 
-       height = 10,
-       width = 18)
+       height = 12,
+       width = 10)
 
@@
 ```{r}
+#missing in Mafessoni 
+m_maf <- to_plot %>%
+           filter(V5 == 1 & Adjusted_Log10p > 2)
 
+#missing in ours
+m_us <- to_plot %>%
+           filter(V5 < 1 & Adjusted_Log10p < 2)
+
+#present in both
+both <- to_plot %>%
+           filter(V5 < 1 & Adjusted_Log10p > 2)
+
+print(paste("missing in Mafessoni : ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(to_plot)*100) )
+print(paste("missing in ours : ",nrow(m_us), "or : ",nrow(m_us)/nrow(to_plot)*100))
+print(paste("present in both : ",nrow(both), "or : ",nrow(both)/nrow(to_plot)*100))
+print(paste("total : ", nrow(to_plot)))
 ```

### 2019-07-04 LatexManuscript/legacy.tex
updated text and supplementary figures.
minor updates

@@
 \todo{Mafessoni et al.
-recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}\sgcomment{You are talking about things that have not been introduced yet...
+recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium (LD) patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}\sgcomment{You are talking about things that have not been introduced yet...
 Maybe: Mafessoni et al.
-recently identified batch effects in the 1kGP by looking for individuals with excess linkage disequilibrium among distant variants.
+recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants.
 }
@@
 Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies \sgcomment{, and are distributed across the genome--this is outdated}.
+\luke{not sure what this means?}
 \textbf{A} 
@@
 ($587$ SNPs had  $p < 10^{-8}.$)
-\sgcomment{Why do we say the following?} The same analysis was performed independently for each of the populations in the 1000 Genomes Project.
 \textbf{B} 
@@
  Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
-The shape indicates whether the individual \sgcomment{belongs to -- could we clarify this?} Phase 1 or Phase 3 of the 1000 Genomes project. 
+The shape indicates whether the individual was first released in Phase 1 or Phase 3 of the 1000 Genomes project. 
 \textbf{B} Sequencing technologies used over the course of the 1000 Genomes Project.}
@@
 By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
-\sgcomment{There seems to be a paragraph break needed, a logical link missing, or both?}\todo{The mutational 3-mer substitution patterns of this list of variants is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
+
+\todo{The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
 However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
@@
 
-\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a different method.
-We find that only 10\% of the variants they identified are in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
-Moreover, the variants they identified are not enriched in the mutational spectra described above \ref{Mafessoni_MutSpect}. }
+\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics.
+They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
+We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
+We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. 
+Interetingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above \ref{Mafessoni_MutSpect}.
+These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suit of association tests.}
 
@@
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-Some of these publications noted that these variants were not flanked by linkage disequilibrium peaks \citep{Mandage2017}.
+Some of these publications noted that these variants were not flanked by LD peaks \citep{Mandage2017}.
 Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
@@
 \caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
+There are a total of 40,266 variants in this plot with only 1328 ( 3\% ) are statistically significant in both test.
+There are 23172  variants ( 58\% ) that are $Q$-associated but not LD-associated and 15766 ( 39\% ) that are LD-associated but not $Q$-associated.
 \textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  

### 2019-07-04 PlottingScripts/ProcessPlotData.Rmd
minor updates

@@
 jpt_imp_files = list.files(path=jpt_imp_path, 
-                           pattern=paste0(array,".imputed.1.info.gz"),
+                           pattern=paste0(array,".imputed.*.info.gz"),
                            full.names = T) 

### 2019-07-04 LatexManuscript/legacy.tex
minor updates

@@
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-\todo{Mafessoni et al.
-recently identified a batch effect in the 1kGP without using any quality control metrics, but instead, using an approach that identifies individuals and loci with unusual linkage disequilibrium (LD) patterns, these result partially overlap with the variants we identified using a more statistically powerful method \citep{mafessoni2018turning}.}\sgcomment{You are talking about things that have not been introduced yet...
-Maybe: Mafessoni et al.
-recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants.
-}
+\todo{Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.}
 Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-\todo{Our initial goal was to determine if \sgcomment{whether?} this signature could be explained by population structure or technological error.}
+Our initial goal was to determine whether this signature could be explained by population structure or technological error.
 While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
@@
 
-\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics.
-They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
+\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics \citep{mafessoni2018turning}.
+They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ (Figure \ref{Mafessoni_MutSpect}).
 We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
 We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interetingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above \ref{Mafessoni_MutSpect}.
+Interetingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
 These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suit of association tests.}
@@
 \todo{
-To assess whether or not $Q$-associated variants results from mutations occurring in the creation or maintenance of the cell lines, or from technical artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
+To assess whether or not $Q$-associated variants result from mutations occurring in the creation or maintenance of the cell lines, or from technical artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
 Variants that do not reproduce across these cohorts are likely the result of technical artifacts. 
@@
 Finally, to complicate matters, we may have a third situation where a real variant has a genotyping error probability that correlates with $Q$. 
-In this case, we expect the variant may reproduce, but at different frequency.
+In this case, we expect the variant may reproduce, but at different frequency in some individuals.
  }

### 2019-07-05 PlottingScripts/ProcessPlotData.Rmd
updated figures

@@
 
-jpt_imp <- merge(jpt_imp, suspicious, 
-                 by = c('Chr', 'Pos'), all.x = T)
+#jpt_imp <- merge(jpt_imp, suspicious, 
+#                 by = c('Chr', 'Pos'), all.x = T)
 
 jpt_imp$Pos <- as.numeric(as.character(jpt_imp$Pos))
+jpt_imp$Chr <- as.numeric(as.character(jpt_imp$Chr))
 
@@
 
-imputed_position <- jpt_imp$Pos
-suspicious_position <- jpt_imp[which(jpt_imp$Adjusted_Log10p > 2),]$Pos
+close_list <- data.frame()
+for(chrom in seq(1,22)){
+  print(chrom)
 
-jpt_imp$closest <- unlist(lapply(imputed_position, function(x) suspicious_position[which.min(abs(suspicious_position - x))]))
-jpt_imp$dist <- abs(jpt_imp$closest - jpt_imp$Pos)
+  imputed_position <- 
+    jpt_imp %>% 
+    filter(Chr == chrom) %>%
+    select(Pos) %>%
+    unlist()
+
+  suspicious_position <- 
+    suspicious %>% 
+    filter(Chr == chrom) %>%
+    select(Pos) %>%
+    unlist()
+
+  close <- 
+  unlist(lapply(imputed_position, 
+                function(x) suspicious_position[which.min(abs(suspicious_position - x))]))
+
+  out <- data.frame("Pos" = imputed_position, 
+                       "closest" = close)
+  rownames(out) <- NULL
+
+  close_list <- rbind(close_list, out)
+}
+
+close_list
+```
+
+```{r}
+
+jpt_imp$dist <- abs(close_list$closest - close_list$Pos)
 
@@
        y = "Rsq") +
-  xlim(c(0,10000))
+  xlim(c(0,5000))
 
@@
        y = "Average Call") +
-  xlim(c(0,10000))
+  xlim(c(0,5000))
 
@@
 ggsave(paste0(figurePath,
-              "Imputation_Distance_10s.jpg"),
+              "Imputation_Distance_100s_gw.jpg"),
        height = 9,

### 2019-07-05 LatexManuscript/legacy.tex
updated figure 1

@@
 \centering
-\includegraphics[width=13cm,keepaspectratio]{../Figures/histogram_of_sig_snps.jpg}
+\includegraphics[width=8cm,keepaspectratio]{../Figures/histogram_of_sig_snps.jpg}
 \caption{Frequency distribution of the 1034 $Q$ associated SNPs from the single population test in the JPT in blue and all SNPs in black.
@@
 \centering
-\includegraphics[width=13cm,keepaspectratio]{../Figures/strict_gcat_mutational_enrichment.jpg}
+\includegraphics[width=10cm,keepaspectratio]{../Figures/strict_gcat_mutational_enrichment.jpg}
 \caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
@@
 
+\clearpage
+
 \begin{figure}[tbp]
-\includegraphics[width=12cm,keepaspectratio]{../Figures/Omni_AF.jpg}
+\includegraphics[width=11cm,keepaspectratio]{../Figures/Omni_AF.jpg}
 \caption{Estimated frequency and association strength of $Q$- associated variants present on Illumina's Omni 2.5 chip. Variants highly associated to $Q$ tend to have low global allele frequencies.}
@@
 \begin{figure}[tbp]
-\includegraphics[width=12cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
+\includegraphics[width=11cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
 \caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project. 
-The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.
-}
+The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.  }
 \label{PC1_Correlation}
@@
 
-\begin{figure}[h!]
-\includegraphics[width=12cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
-\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  \textbf{A} 
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
+\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+\textbf{A} 
 The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.

### 2019-07-08 LatexManuscript/legacy.tex
a few edits and comments

@@
 \caption{
-Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies \sgcomment{, and are distributed across the genome--this is outdated}.
-\luke{not sure what this means?}
+Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
 \textbf{A} 
@@
 \todo{The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
-However, another enrichment profile is significantly enriched : TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
-These three enrichments can be summarized as T*T${\rightarrow}$TTT.}
+There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT. }\sgcomment{Point to figure}
 
 \todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics \citep{mafessoni2018turning}.
-They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ (Figure \ref{Mafessoni_MutSpect}).
+They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ 
+\sgcomment{the previous phrase is confusing to me. E.g., does "that" stand for patterns or variants?
+Maybe replace by :  
+A recent publication by Mafessoni et al.
+also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify over 15,000 suspicious variants \citep{mafessoni2018turning}.
+  } (Figure \ref{Mafessoni_MutSpect}).
 We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
-We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interetingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
-These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suit of association tests.}
+We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. \sgcomment{Present proportions for both, or counts for both? Did they use only strict mask? }
+Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
+These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.}
 
@@
 \centering
-\begin{tabular}{l  r r r}
-                      & {Repeat}  & {Non-Repeat}   & {Total}    \\ \hline
-{SNP}  & 3,369 \textcolor{gray}{0.53\textperthousand} & 11,059 \textcolor{gray}{0.56\textperthousand} & 14,428\\  
-{INDEL} & 181 \textcolor{gray}{0.3\textperthousand} & 657 \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
-{Total} & 3,550 & 11,716 & 15,270\\
+\begin{tabular}{l  r r r  r r}
+                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
+{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
+{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
+{Total} & 3,550& & 11,716 & & 15,270\\
 \end{tabular}
-\caption{Number of statistically significant variants  passing the 1000 Genomes Project strict mask per category. 
+\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
 Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
 A total of 15,270 are statistically significantly associated to $Q$. 
-\todo{The grey text is the proportion of significant associations per category.}
+\todo{The grey text is the proportion of $Q$-associated variants per category.}
 The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
@@
 \todo{
-To assess whether or not $Q$-associated variants result from mutations occurring in the creation or maintenance of the cell lines, or from technical artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
-Variants that do not reproduce across these cohorts are likely the result of technical artifacts. 
-By contrast, if the same mutations are observed at identical frequencies in independently sequenced cohorts, we can conclude that recurrent cell line mutations are likely to blame. 
-Finally, to complicate matters, we may have a third situation where a real variant has a genotyping error probability that correlates with $Q$. 
-In this case, we expect the variant may reproduce, but at different frequency in some individuals.
+To assess whether $Q$-associated variants result from sequencing artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
+Variants that do not reproduce across cohorts are likely the result of sequencing artifacts. 
+Variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.   
  }
@@
 A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
-There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and ACA${\rightarrow}$AAA (reverse complement : TGT${\rightarrow}$TTT).
+There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
 These three enrichments can be summarized as T*T${\rightarrow}$TTT.
@@
 \includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
-\caption{The number of $Q$-associated variants per 10kb window across the genome.
+\caption{The number of $Q$-associated variants per 10kb window across the genome. \sgcomment{Any reason why this is plotted until 35?}
 Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.

### 2019-07-08 LatexManuscript/legacy.tex
more edits and comments

@@
 That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
-\todo{It is plausible that the EBV copy number phenotype is sensitive to mapping and in this study, may be confounded by the same technical artefacts that lead to biased SNP calling.}
+\todo{It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.} \sgcomment{Don't forget to explain that the authors dealt with this using another QC metric!}
 

### 2019-07-08 LatexManuscript/legacy.tex
pushing changes

--- 2019-07-08 LatexManuscript/legacy.tex
@@
 
-\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics \citep{mafessoni2018turning}.
-They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ 
-\sgcomment{the previous phrase is confusing to me. E.g., does "that" stand for patterns or variants?
-Maybe replace by :  
 A recent publication by Mafessoni et al.
-also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify over 15,000 suspicious variants \citep{mafessoni2018turning}.
-  } (Figure \ref{Mafessoni_MutSpect}).
+also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify nearly 20,000 suspicious variants \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}).
+They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
 We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
@@
 Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
-These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.}
+These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
 
@@
 
-In 2017, Lan et al.
-resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
+In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
 %To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
@@
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-Some of these publications noted that these variants were not flanked by LD peaks \citep{Mandage2017}.
-Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
-However, the fact that some of these variants are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
 
@@
 
-These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+\todo{All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
+Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
 Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
+However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
 
@@
 By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
-One study reports associations with seven $Q$-associated variants that we find to be highly biased \citep{Mandage2017}. 
-That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
-\todo{It is plausible that the EBV copy number phenotype is sensitive to mapping and in this study, may be confounded by the same technical artefacts that lead to biased SNP calling.}
+One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
+The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.}
 
@@
 
-Recommendations : doing their approach as well as ours. We recommend removing the variants we ID and the ones they find. Regression with QC metrics (shoutout to them)
-
 The variants identified in this study are likely to be technical artifacts from legacy technologies.
@@
 
-For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the $Q$-associated SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary Data).
+For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the individuals with $Q<30$ and to filter out the LD-associated variants identified by \cite{mafessoni2018turning} and the $Q$-associated variants we identified (lists of such variants and sample IDs are provided in the Supplementary Data).
 We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
@@
 The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. 
-Both approaches produce comparable results.
+All three approaches accounting for population structure we tested produce comparable results.
 
@@
 The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for retirement or re-sequencing.
+
+\section{Author's Note}
+\todo{While reviewing our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}
 
@@
 \caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
-There are a total of 40,266 variants in this plot with only 1328 ( 3\% ) are statistically significant in both test.
-There are 23172  variants ( 58\% ) that are $Q$-associated but not LD-associated and 15766 ( 39\% ) that are LD-associated but not $Q$-associated.
+There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both test.
+There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
 \textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  

### 2019-07-08 PlottingScripts/ProcessedData/ScoopCheck.Rmd
pushing changes

@@
 #load supplementary files from Mafessoni
+#t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
+#t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
+#t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)
+
 t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
-
-#t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
-
 t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)
-
-#t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
-
-#t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)
 
@@
 
-#totalTus <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
-totalTusX <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
-totalTusX[is.na(totalTusX)] <- 1
+Merged_Regression <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
+Merged_Regression[is.na(Merged_Regression)] <- 1
 
-totalTusX
+Merged_Regression
 ```
@@
 
-to_plot <- totalTusX %>%
+to_plot <- Merged_Regression %>%
            filter(V5 < 1 | Adjusted_Log10p > 2)
@@
         axis.text = element_text(size = 20))
-
-p_comp <- ggExtra::ggMarginal(p_comp, type = "histogram")
 

### 2019-07-09 LatexManuscript/legacy.tex
updates

@@
 
-\todo{All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
@@
 The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
-It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.}
+It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
 

### 2019-07-09 PlottingScripts/ProcessedData/ScoopCheck.Rmd
updates

--- 2019-06-25 PlottingScripts/ProcessedData/ScoopCheck.Rmd
@@
                         low = 'white', 
-                        high = 'royalblue1',
+                        high = 'black',
                         guide = guide_legend(title = 'Variant\ncount',
@@
 ```{r}
-t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
-t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
-t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)
-t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
-t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)
+#load regression results
+Regression <- readRDS('/Users/luke/Documents/Legacy_Misc/Misc/Regression.Rmd')
+#load supplementary files from Mafessoni
+#coding
+t4 <- read.table('/Users/luke/Downloads/evy199_supp/TableS4.bed', comment.char = "@", header = T)
+t4$type <- "rare_coding"
+t5 <- read.table('/Users/luke/Downloads/evy199_supp/TableS5.bed', comment.char = "@", header = T)
+t5$type <- "common_coding"
+#intergenic
+t6 <- read.table('/Users/luke/Downloads/evy199_supp/TableS6.bed', comment.char = "@", header = T)
+t6$type <- "common_intergenic"
+t7 <- read.table('/Users/luke/Downloads/evy199_supp/TableS7.bed', comment.char = "@", header = T)
+t7$type <- "rare_intergenic"
 
-us <- fread('/Users/luke/Documents/LegacyData/LatexManuscript/SupplementaryData/Significant_Positions_and_Pvals.txt')
+#get total unique positions from most conservative tables
+totalT<- 
+  rbind(t4,t5,t6,t7) %>%
+  group_by(X.chr, end) %>% 
+  top_n(1) #keep unique positions only
 
-totalT<- unique(rbind(t4,t5,t6,t7,t8))
-totalTus <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
-totalTusX <- merge(us, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
-totalTusX[is.na(totalTusX)] <- 1
+merged_regression <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('X.chr','end'), all.x=T)
+merged_regression[is.na(merged_regression)] <- 1
 
+#remove positions that don't pass strict mask
+merged_regression_masked <- 
+  anti_join(merged_regression, 
+            strictMask,
+            by = c('Chr','Pos','rsID'))
+
+#remove positions that DO pass strict mask
+merged_regression_unmasked <- 
+  anti_join(merged_regression, 
+            merged_regression_masked)
+
+nrow(merged_regression)
+
+nrow(merged_regression_unmasked)
+
+nrow(merged_regression_masked)
+
+merged_regression_unmasked$strict_mask <- "Pass strict mask"
+merged_regression_masked$strict_mask <- "Fail strict mask"
+
+merged_regression <- rbind(merged_regression_masked, merged_regression_unmasked)
 ```
@@
 ```{r}
+#select rows of interest
+#either present in Mafessoni OR significant in our data
+
+to_plot <- merged_regression %>%
+           filter(pvalue_glm_filt < 1 | Adjusted_Log10p > 2)
+
 p_comp <- 
-  ggplot(totalTusX, 
+  ggplot(to_plot, 
          aes(x=Adjusted_Log10p, 
-           y= -log10(V5))) +
-  geom_point(shape = 1) +
+             y= -log10(pvalue_glm_filt), 
+             color = strict_mask)) +
+  geom_point(shape = 1,
+             size = 1,
+             alpha = 0.3) +
   labs(x = 'Q-association',
-       y = 'LD-association')
+       y = 'LD-association') +
+  scale_color_grey(start = 0.1, end = 0.6) +
+  guides(color = F) +
+  facet_wrap(.~strict_mask)+
+  theme_classic() +
+  theme(axis.title = element_text(size = 20),
+        axis.text = element_text(size = 20))
 
 p_comp
+```
+```{r}
+#missing in Mafessoni
+p <- to_plot %>%
+  filter(strict_mask == 'Pass')
+
+m_maf <- p %>%
+           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+
+#missing in ours
+m_us <- p %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+
+#present in both
+both <- p %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+
+print(paste("missing in Mafessoni (Pass): ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(p)*100) )
+print(paste("missing in ours (Pass): ",nrow(m_us), "or : ",nrow(m_us)/nrow(p)*100))
+print(paste("present in both (Pass): ",nrow(both), "or : ",nrow(both)/nrow(p)*100))
+print(paste("total (Pass): ", nrow(p)))
+
+#missing in Mafessoni
+f <- to_plot %>%
+  filter(strict_mask == 'Fail')
+
+m_maf <- f %>%
+           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+
+#missing in ours
+m_us <- f %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+
+#present in both
+both <- f %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+
+print(paste("missing in Mafessoni (Fail): ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(f)*100) )
+print(paste("missing in ours (Fail): ",nrow(m_us), "or : ",nrow(m_us)/nrow(f)*100))
+print(paste("present in both (Fail): ",nrow(both), "or : ",nrow(both)/nrow(f)*100))
+print(paste("total (Fail): ", nrow(f)))
 ```
@@
 mafessoni_snps <- 
-  totalT[,c('V1','V3')]
+  totalT[,c('X.chr','end')]
 
@@
 ```
-
-```{r}
-
-```

### 2019-07-09 LatexManuscript/legacy.tex
removed some blue text, fixed S8

@@
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-\todo{Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.}
+Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.
 Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
@@
 
-\todo{The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
-There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT. }\sgcomment{Point to figure}
+The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
+There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT (Figure \ref{strict_gcat_mutational_enrichment}).
 
@@
 A total of 15,270 are statistically significantly associated to $Q$. 
-\todo{The grey text is the proportion of $Q$-associated variants per category.}
+The grey text is the proportion of $Q$-associated variants per category.
 The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
@@
 \subsection{Cell line or technical artifact}
-\todo{
 To assess whether $Q$-associated variants result from sequencing artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
@@
 Variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.
-}
 
@@
 \includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
-\caption{The number of $Q$-associated variants per 10kb window across the genome. \sgcomment{Any reason why this is plotted until 35?}
+\caption{The number of $Q$-associated variants per 10kb window across the genome.
 Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.

### 2019-07-09 LatexManuscript/legacy.tex
update fig text
minor text, pre meeting push

@@
 Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
+The plot is zoomed in on frequencies below 0.5 for clarity.
 The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
@@
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
 It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
@@
 \section{Author's Note}
-\todo{While reviewing our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}
+\todo{While revising our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}
 
@@
 We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
+\todo{We would like to thank Rajendra Mandage, Gabriel Santpere and Arcadi Navarro for sharing more details relating to the results from their publication.}
 We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.

### 2019-07-09 LatexManuscript/legacy.tex
update to acknowledgments

@@
 
-\section{Author's Note}
-\todo{While revising our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}
-
 \section{Methods}
@@
 \section{Acknowledgments}
-We would like to thank Kelly Harris for sharing her mutation spectrum scripts.
-\todo{We would like to thank Rajendra Mandage, Gabriel Santpere and Arcadi Navarro for sharing more details relating to the results from their publication.}
+We would like to thank K. Harris for sharing her mutation spectrum scripts, F. Mafessoni, G. Santpere and A. Navarro for feedback on earlier versions of the manuscript.
 We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.

### 2019-07-10 PlottingScripts/ProcessPlotData.Rmd
updated figure 1
squared fig1

@@
 #toggle for all snps or just JPT 1kGP snps
-name <-''#'strict.gcat.' #'gcat.' # '' #
+name <-'strict.gcat.' #'gcat.' # '' #
 
@@
 
-jpt_imp$dist_2 <- round(jpt_imp$dist, -1)
+jpt_imp$dist_2 <- round(jpt_imp$dist, -3)
 
@@
 ggsave(paste0(figurePath,
-              "Imputation_Distance_100s_gw.jpg"),
+              "Imputation_Distance_1000s_gw.jpg"),
        height = 9,

### 2019-07-10 PlottingScripts/ProcessedData/ScoopCheck.Rmd
squared fig1

@@
 p <- to_plot %>%
-  filter(strict_mask == 'Pass')
+  filter(strict_mask == 'Pass strict mask')
 
@@
 f <- to_plot %>%
-  filter(strict_mask == 'Fail')
+  filter(strict_mask == 'Fail strict mask')
 

### 2019-07-10 LatexManuscript/legacy.tex
cleaned up supplementary figures

@@
 
-\caption{Variants associated with average quality of mapped bases $Q$ in more than one population.
-The size of the vertical bars ( I ) are proportional to the -log10(p) value of that SNP.
-The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
-Phase 1 populations are marked by a star ( * ).
-The line plot underneath shows the number of populations for which a variant has reached significance.
-The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$.}
+\caption{Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
+\textbf{A} For each population, we plot the number of variants associated to at least one other population.
+\textbf{B} The number of populations for which a variant has reached significance.}
   \label{OverLap}
@@
 \includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
-\caption{The number of $Q$-associated variants per 10kb window across the genome.
+\caption{\textbf{A} The number of $Q$-associated variants per 10kb window across the genome.
 Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.}  
+One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
+\textbf{B} Distribution of variant counts per 10kb window. Where variants that fail the strict mask are in grey and those that pass are in black. }  
 \label{VariantDensity}

### 2019-07-10 PlottingScripts/ProcessedData/ScoopCheck.Rmd
saving r scripts

@@
              alpha = 0.3) +
-  labs(x = 'Q-association',
-       y = 'LD-association') +
+  labs(x = '-log10(p) Q-association',
+       y = '-log10(p) LD-association') +
   scale_color_grey(start = 0.1, end = 0.6) +
@@
        m_plot, 
-       height = 12,
+       height = 10,
        width = 10)

### 2019-07-10 LatexManuscript/legacy.tex
updated plot with variant count

@@
 A recent publication by Mafessoni et al.
-also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify nearly 20,000 suspicious variants \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}).
+also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}).
 They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
-We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
-We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. \sgcomment{Present proportions for both, or counts for both? Did they use only strict mask? }
+We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
+We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
 Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).

### 2019-07-10 PlottingScripts/ProcessedData/ScoopCheck.Rmd
updated plot with variant count

@@
 
-merged_regression_unmasked$strict_mask <- "Pass strict mask"
-merged_regression_masked$strict_mask <- "Fail strict mask"
+merged_regression_unmasked$strict_mask <- paste0("Pass strict mask (",nrow(merged_regression_unmasked),")")
+merged_regression_masked$strict_mask <- paste0("Fail strict mask (",nrow(merged_regression_masked),")")
 
@@
 print(paste("total (Fail): ", nrow(f)))
+
+m_maf <- to_plot %>%
+           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+
+#missing in ours
+m_us <- to_plot %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+
+#present in both
+both <- to_plot %>%
+           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+
+print(paste("missing in Mafessoni (total): ",nrow(m_maf), "or : ", nrow(m_maf)/nrow(to_plot)*100) )
+print(paste("missing in ours (total): ",nrow(m_us), "or : ",nrow(m_us)/nrow(to_plot)*100))
+print(paste("present in both (total): ",nrow(both), "or : ",nrow(both)/nrow(to_plot)*100))
+print(paste("total (total): ", nrow(to_plot)))
 ```

### 2019-07-10 LatexManuscript/legacy.tex
updated supplementary figure text

@@
 \caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
-There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both test.
+There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
 There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
+For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
+For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
 \textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/VariantDensity.jpg}
+\includegraphics[width=14cm,keepaspectratio]{../Figures/VariantDensity.jpg}
 \caption{\textbf{A} The number of $Q$-associated variants per 10kb window across the genome.

### 2019-07-10 LatexManuscript/legacy.tex
minor

@@
 Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
+Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
 We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 

### 2019-07-11 LatexManuscript/legacy.tex
figure reference

@@
 A recent publication by Mafessoni et al.
-also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}).
-They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
-We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
+also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}A).
+They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$.
+We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method.
 We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
+Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
 These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.

### 2019-07-12 LatexManuscript/legacy.tex
figure text
edited cell line artefact

@@
 Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
-A total of 15,270 are statistically significantly associated to $Q$. 
+A total of 15,270 are statistically significantly associated to $Q$ and pass the 1kGP strict mask. 
 The grey text is the proportion of $Q$-associated variants per category.
@@
 
-\subsection{Cell line or technical artifact}
-To assess whether $Q$-associated variants result from sequencing artifacts, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
-Variants that do not reproduce across cohorts are likely the result of sequencing artifacts. 
-Variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.
+\subsection{Spurious variants, biased genotypes, or cell line artefacts?}
+To assess whether $Q$-associated variants are spurious variants resulting from sequencing artifacts or real variants exhibiting biased genotyping, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
+$Q$-associated variants that do not reproduce across sequencing experiments are likely the result of sequencing artifacts in the 1kGP. By contrast, 
+$Q$-associated variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.
+Finally, given our nominal false discovery rate of $\alpha = 0.01$, we expect approximately $1\%$ of $Q$-associated variants to be false discoveries, i.e., variants that are not associated with $Q$ and that should therefore reproduce across experiments.
 
@@
 Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}).   
-This is more than the 2.96 predicted false positives according to our nominal false discovery rate of $\alpha = 0.01$. 
-Thus a small number of variants associated to $Q$ in the JPT are present in the population but with biased genotypes.
+This is slightly more but consistent with the 3 false positives predicted according to the $\alpha = 0.01$ nominal false discovery rate. 
+Thus the majority of $Q$-associated variants in the CHB or CHS  appear to be spurious variants.
 
@@
 Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
-The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, supporting sequencing rather than cell line artifacts.
+The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, suggesting that they are spurious variants.
 Five variants show differing frequencies between both datasets and are likely explained by biased genotypes.
-Finally, 1,139 (or 18\%) are present in the resequenced data.
-The allele frequencies of these variants are nearly identical between datasets suggesting that variant calls are accurate for these 83 individuals.
-This is likely because these variants have biased genotypes in populations other than the CHB/CHD.
+Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
+Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies not only spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
 

### 2019-07-12 LatexManuscript/legacy.tex
a word

--- 2019-07-11 LatexManuscript/legacy.tex
@@
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
 It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

### 2019-07-12 PlottingScripts/ProcessedData/ScoopCheck.Rmd
updated figure

@@
                         high = 'black',
+                        breaks = c(100,250,500),
                         guide = guide_legend(title = 'Variant\ncount',
@@
 
-merged_regression_unmasked$strict_mask <- paste0("Pass strict mask (",nrow(merged_regression_unmasked),")")
-merged_regression_masked$strict_mask <- paste0("Fail strict mask (",nrow(merged_regression_masked),")")
+merged_regression_unmasked$strict_mask <- "Pass strict mask"
+merged_regression_masked$strict_mask <- "Fail strict mask"
 
@@
 to_plot <- merged_regression %>%
-           filter(pvalue_glm_filt < 1 | Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt < 1 | Adjusted_Log10p > 2)
 
@@
 p <- to_plot %>%
-  filter(strict_mask == 'Pass strict mask')
+  dplyr::filter(strict_mask == paste0("Pass strict mask (",nrow(merged_regression_unmasked),")"))
 
 m_maf <- p %>%
-           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
 
@@
 m_us <- p %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
 
@@
 both <- p %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
 
@@
 f <- to_plot %>%
-  filter(strict_mask == 'Fail strict mask')
+  dplyr::filter(strict_mask == paste0("Fail strict mask (",nrow(merged_regression_masked),")"))
 
 m_maf <- f %>%
-           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
 
@@
 m_us <- f %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
 
@@
 both <- f %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
 
@@
 m_maf <- to_plot %>%
-           filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt == 1 & Adjusted_Log10p > 2)
 
@@
 m_us <- to_plot %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p < 2)
 
@@
 both <- to_plot %>%
-           filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
+           dplyr::filter(pvalue_glm_filt < 1 & Adjusted_Log10p > 2)
 

### 2019-07-12 LatexManuscript/legacy.tex
update to imputation section

@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-These $Q$-associated variants could also compromise the imputation of nearby real-variants causing not only improper imputation, but also false negatives for real variants.
+These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
 
@@
 \centering
+\includegraphics[width=10cm,keepaspectratio]{../Figures/Imputation_Distance_100s_gw.jpg}
+\caption{Imputation scores are not impacted by nearby suspicious variants.
+By comparing the imputation scores from Japanese Nagahama genotype data to the distance from the nearest suspicious variant, we find no impact on nearby variants.
+The only variants that do appear to have lower imputation scores are the suspicious variants themselves with a distance of zero on the plot.
+}
+\label{imputation_score}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
 \includegraphics[width=15cm,keepaspectratio]{../Figures/OverUnder30.jpg}

### 2019-07-12 LatexManuscript/legacy.tex
small edits

--- 2019-07-12 LatexManuscript/legacy.tex
@@
 %%% PREAMBLE 
-\documentclass[9pt,lineno]{template}
+\documentclass[9pt,article]{template}
 
@@
 $Q$-associated variants that do not reproduce across sequencing experiments are likely the result of sequencing artifacts in the 1kGP. By contrast, 
-$Q$-associated variants that do reproduce across studies could result from (recurrent) cell line mutations or, more likely, from existing variants whose accurate genotyping depends on $Q$.
+$Q$-associated variants that do reproduce across experiments could result from (recurrent) cell line mutations or, more likely, from existing variants whose genotyping depends on $Q$.
 Finally, given our nominal false discovery rate of $\alpha = 0.01$, we expect approximately $1\%$ of $Q$-associated variants to be false discoveries, i.e., variants that are not associated with $Q$ and that should therefore reproduce across experiments.
@@
 In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-%To assess consistency between the two datasets, we consider consistency of genotype calls for $Q$-associated variants that are predicted to be polymorphic in these 83 individuals according to the 1kGP.
-Among the 296 such variants that were $Q$-associated in the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}).   
+Among the 296 such variants that were $Q$-associated in single-population tests for the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}).   
 This is slightly more but consistent with the 3 false positives predicted according to the $\alpha = 0.01$ nominal false discovery rate. 
@@
 Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
-Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies not only spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
-
-%Some mutational signatures are heterogenous where not all suspicious variants are completely spurious in all individuals or populations. 
-%Some mutational signatures reproduce across cohorts, whereas others do not and may be due to the different sequencing platforms used.
+Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
 
@@
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
 It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.

### 2019-07-12 PlottingScripts/ProcessPlotData.Rmd

@@
 ```{r}
+strictMask <- fread(paste0(DataPath,
+                           '1kGP_StrictMask.bed'),
+            col.names = c('Chr','Pos','Pos.1','rsID'))
+
 phase1 <- c('YRI','LWK','CHB','CHD','JPT','CEU','LWK')
@@
 
-strictMask <- fread(paste0(DataPath,
-                           '1kGP_StrictMask.bed'),
-            col.names = c('Chr','Pos','Pos.1','rsID'))
+# positions that don't pass strict mask
+Reg_not_strict <- inner_join(Reg,
+                 strictMask,
+                 by = c('Chr','Pos')) %>%
+  select('Chr','Pos','Population','deviance') %>%
+  mutate(strict = "fail")
+
 #remove positions that don't pass strict mask
-Reg <- anti_join(Reg,
+Reg_strict <- anti_join(Reg,
                  strictMask,
-                 by = c('Chr','Pos'))
+                 by = c('Chr','Pos')) %>%
+  mutate(strict = "pass")
+
+Reg_mask <- rbind(Reg_strict, Reg_not_strict)
 
 #Get P-Values and significant sites
-Reg$plog10 <- -log10(pchisq(Reg$deviance, 1, 
+Reg_mask$plog10 <- -log10(pchisq(Reg_mask$deviance, 1, 
                             lower.tail = FALSE))
-Reg<-Reg[which(Reg$plog10 >= 6),]
+Reg_mask<-Reg_mask[which(Reg_mask$plog10 >= 6),]
 #Count number of populations where a variant is significant
-NumPop <- as.data.frame(table(Reg$Chr,Reg$Pos))
-NumPop <- NumPop[which(NumPop$Freq>1),]
+NumPop <- as.data.frame(table(Reg_mask$Chr,Reg_mask$Pos))
 names(NumPop) <- c('Chr','Pos', 'SigHits')
@@
 
-m <- merge(Reg,
-           NumPop, 
-           by = c('Chr','Pos'))
-#count number of significant sites per population
-n <- as.data.frame(table(m$Pop))
-#put it all together for plotting
-overLap <- unique(merge(m,n,
-                        by.x = 'Population',
-                        by.y = 'Var1'))
-#include number of significant sites in titles
-overLap$title = paste0(overLap$Pop,
-                       ' : ', 
-                       overLap$Freq)
-#add asterix for Phase 1 pops
-Phase1Title <- paste('*',
-                     overLap[which(overLap$Pop 
-                                   %in% phase1),]$title)
-overLap[which(overLap$Pop
-              %in% phase1),]$title <- Phase1Title
-#set order by SigHits, Chromosome and Position
-overLap <- overLap[with(overLap, 
-                        order(-SigHits,
-                              Chr,
-                              Pos)),]
-overLap$Pos <- factor(overLap$Pos, 
-                      levels = unique(overLap$Pos))
-#this is for keeping track of the number of variants
-overLap$nrow <- seq(length = nrow(overLap))
-#this is for discrete point size
-overLap$Pcat <- 4
-overLap[which(overLap$plog10 < 8),]$Pcat <- 2
-overLap[which((overLap$plog10 > 8) & 
-              (overLap$plog10 < 10)),]$Pcat <- 3
+SinglePop <- NumPop[which(NumPop$SigHits==1),]
+SinglePop <- merge(Reg_mask,
+                   SinglePop, 
+                   by = c('Chr','Pos')) %>%
+  group_by(Population, strict) %>%
+  summarize(single_count = n())
 
-overLap
+DoublePop <- NumPop[which(NumPop$SigHits>1),]
+DoublePop <- merge(Reg_mask,
+                   DoublePop, 
+                   by = c('Chr','Pos')) %>%
+  group_by(Population, strict) %>%
+  summarize(double_count = n())
+
+Combined_count <- merge(DoublePop, SinglePop, by = c('Population', 'strict'))
+
+Combined_count[which(Combined_count$Pop%in% phase1),]$Population <- paste(Combined_count[which(Combined_count$Pop%in% phase1),]$Population, '*')
 
 #save to file
-saveRDS(overLap,
+saveRDS(Combined_count,
             paste0(DataPath,
                    'overLap.rds'))
+
+pos_count <- Reg_mask %>%
+  group_by(Chr,Pos,strict) %>%
+  summarize(count = n()) %>%
+  group_by(count, strict) %>%
+  summarize(freq = n())
+
+#save to file
+saveRDS(pos_count,
+            paste0(DataPath,
+                   'pos_count.rds'))
 ```

### 2019-07-12 LatexManuscript/legacy.tex
update figure S4

@@
 
-\caption{Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
+\caption{Comparing results from the single population tests, we find 812 variants that pass the strict mask that are independently associated with average quality of mapped bases $Q$ in more than one population.
 \textbf{A} For each population, we plot the number of variants associated to at least one other population.
-\textbf{B} The number of populations for which a variant has reached significance.}
+Phase 1 populations are indicated with a star (*).
+\textbf{B} The number of populations for which a variant has reached significance. Many of the variants associated in only one population do not pass the strict mask. }
   \label{OverLap}

### 2019-07-12 LatexManuscript/legacy.tex
figure S4

--- 2019-07-12 LatexManuscript/legacy.tex
@@
 
-\caption{Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
-\textbf{A} For each population, we plot the number of variants associated to at least one other population.
+\caption{Results from the single population test.
+We find a total of 1,451 variants associated with the average quality of mapped bases $Q$ where 812 of them are found in at least one other population.
+Variants that pass the strict mask are in light grey while those that do not are in dark grey.
+%Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
+\textbf{A} For each population, we plot the number of variants associated to $Q$.
 \textbf{B} The number of populations for which a variant has reached significance.}

### 2019-07-12 LatexManuscript/legacy.tex
final push

@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
+Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
@@
 
-\caption{Results from the single population test.
-We find a total of 1,451 variants associated with the average quality of mapped bases $Q$ where 812 of them are found in at least one other population.
-Variants that pass the strict mask are in light grey while those that do not are in dark grey.
-%Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
+\caption{Results from the single population test. We find a total of 1,165 variants associated with the average quality of mapped bases $Q$.
+Phase 1 populations are indicated by a star ( * ).
 \textbf{A} For each population, we plot the number of variants associated to $Q$.

### 2019-07-12 LatexManuscript/legacy.tex
clarifications, typos, etc

@@
 While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
-While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
+While investigating an intriguing putative population stratification within the Japanese population, we identified a batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
 Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
@@
 
-Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
+Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants to estimate allele frequencies and to answer a wide range of medical and evolutionary questions.
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-Our initial goal was to determine whether this signature could be explained by population structure or technological error.
+Our initial goal was to determine whether this signature could be explained by population structure or by technological bias.
 While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
-When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}B).
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
@@
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}.
+Statistical significance of these enrichments is computed using a chi-squared test following \cite{Harris2017a}.
 
@@
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses where quality of imputation matters, identifying $Q$-associated variants may be preferable.
+For other analyses where quality of imputation matters, identifying and discarding $Q$-associated variants may be preferable.
 
@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
+Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.
 
@@
 
-\caption{Results from the single population test. We find a total of 1,165 variants associated with the average quality of mapped bases $Q$.
-Phase 1 populations are indicated by a star ( * ).
-\textbf{A} For each population, we plot the number of variants associated to $Q$.
+\caption{Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
+\textbf{A} For each population, we plot the number of variants associated to at least one other population.
 \textbf{B} The number of populations for which a variant has reached significance.}

### 2019-07-12 LatexManuscript/legacy.tex

@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.
+
+\sgcomment{initially was: Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.}
+
+Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
@@
 
-\caption{Comparing results from the single population tests, we find 812 variants independently associated with average quality of mapped bases $Q$ in more than one population.
-\textbf{A} For each population, we plot the number of variants associated to at least one other population.
+\caption{Results from the single population test. We find a total of 1,165 variants associated with the average quality of mapped bases $Q$.
+Phase 1 populations are indicated by a star ( * ).
+\textbf{A} For each population, we plot the number of variants associated to $Q$.
 \textbf{B} The number of populations for which a variant has reached significance.}

### 2019-07-12 LatexManuscript/legacy.tex
reverted overlap statement

@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.
 
-\sgcomment{initially was: Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.}
-
-Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
-
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or \sgcomment{"or" in this context is ambiguous.
+Do you mean "this approach is called ..." or do you mean something else?} Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
 We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 

### 2019-07-12 LatexManuscript/legacy.tex
two comments

@@
 
-This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+This method \sgcomment{GCAT?} identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.

### 2019-07-12 LatexManuscript/legacy.tex
low-quality individuals

@@
 We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
-These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
+Interestingly, the variants identified by Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
+These results indicate that there may be a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
 
@@
 Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
-Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
+Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes.
+The similarity in allele frequencies for these variants suggests that the biases in allele frequencies are present in populations other than the CHB and CHD.
 
@@
 We found that all of the variants associated with $Q$ were imputed back in the samples.
-This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
+This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the individuals with low $Q$.
 These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).

### 2019-07-12 LatexManuscript/legacy.tex
some comments and edits

@@
 %%% PREAMBLE 
-\documentclass[9pt,article]{template}
+\documentclass[9pt,lineno]{template}
 
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the individuals with low $Q$.
-These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
+These $Q$-associated variants could also compromise the imputation of nearby real variants because haplotype matching would be compromised.
+However when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
 
-We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
+We searched the literature for GWAS that have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
@@
 
-All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
-Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
-Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
+%Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
+Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
 However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
+\sgcomment{I'm not sure what the previous sentence is saying}
 

### 2019-07-13 LatexManuscript/legacy.tex
typos and edits

@@
 The plot is zoomed in on frequencies below 0.5 for clarity.
-The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
+The size of blue dots are proportional to the number of suspicious SNPs in a given frequency bin (out of the 1034 identified by GWAS).  
 \textbf{C} 
-Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
+Mutation spectrum of these 1034 variants compared to the expectation from the distribution of all SNPs.
 The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
@@
                       & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
-{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
-{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
+{SNP}  & 3,369 &\textcolor{gray}{  $5.3\times 10^{-4}$}& 11,059 &\textcolor{gray}{  $5.6\times 10^{-4}$}& 14,428\\  
+{INDEL} & 181& \textcolor{gray}{  $3.0\times 10^{-4}$}& 657& \textcolor{gray}{  $6.6\times 10^{-4}$} & 838\\ \hline
 {Total} & 3,550& & 11,716 & & 15,270\\
@@
 
-These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
+These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct association.
 To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
@@
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
-It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
+Even though these seven variants are reported in the GWAS catalogue, they were not flanked by LD peaks and were correctly removed from downstream analyses by \citep{Mandage2017}. 
+It is plausible that the EBV copy number phenotype estimation is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
 
@@
 To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
+Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer actively introducing bias in recent sequence data.
 
@@
 
-On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
+On a technical front, we were surprised that such strong associations between variants and technical covariates in the 1kGP project had not been identified before. 
 The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
 The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}.
-All three approaches accounting for population structure we tested produce comparable results.
 
@@
 \subsection{Code and data availability}
-Since this data is primarily performed using publicly available data, we provide fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
+Since this analysis is primarily performed using publicly available data, we provide fully reproducible and publicly available pipelines on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
 This repository includes scripts used for data download, processing, analysis and plotting.
@@
 For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al. 
-as they applied the current state of the art quality thresholds to remove questionable sequences for detecting population level differences.
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
@@
 
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
+For the GCAT analysis, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
 Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
-These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
+These variants are only removed after the FDR \sgcomment{I thought we used distinct FDR For all four categories?} and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
 
@@
 To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
-The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.
+The deviance test statistic under the null model is approximately chi-square distributed with one degree of freedom.
 
@@
 Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significance. 
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significant. 
 See Supplementary Data for a list of variants and adjusted \textit{p}-values.

### 2019-07-13 LatexManuscript/legacy.tex
reverse GWAS and stuff

@@
 
-We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
+We therefore performed the same association approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
 Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.
@@
 \caption{Comparison of three logistic regression models for testing association to $Q$.
-These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
+These methods model each individual-specific allele frequency as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
 In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
@@
 These three enrichments can be summarized as T*T${\rightarrow}$TTT.
-Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
+Stars on the figure indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
 \label{strict_gcat_mutational_enrichment}
@@
 \includegraphics[width=13cm,keepaspectratio]{../Figures/Mafessoni_MutSpect.jpg}
-\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
+\caption{ \textbf{A} Comparison of the p-values from the LD-based association from \cite{mafessoni2018turning} to p-values from the $Q$ association performed here.
 There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
@@
 For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
-For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
+For the 14,616 variants that do not pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
 \textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
@@
 \includegraphics[width=10cm,keepaspectratio]{../Figures/RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the GCAT logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
 \label{Top10}
@@
 \caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
+The black dashed lines indicates equal allele frequencies while the red dotted line indicates variants twice as frequent in individuals with $Q$ scores below 30.
+The majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
 \label{90HanSFS_full}
@@
 \includegraphics[width=15cm,keepaspectratio]{../Figures/Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in GWAS publications, for individuals with $Q$ scores above and below 30.
 The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
@@
 \includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
+\caption{Joint site frequency spectrum plot of 1000 Genomes Project data and the high depth resequence data, restricted to variants that are both associated with $Q$ in single population tests and polymorphic in the 83 individuals that were resequenced.
+Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequence data \citep{Lan2017}.}  
 \label{90HanSFS}
@@
 \includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+\caption{Joint frequency spectrum for the original 1000 Genomes Project data and the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
 Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
@@
     \caption{Association of SNPS in non-repetitive regions with $Q$.
-    \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions. 
+    \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the GCAT logistic regression analysis for SNPs in non repetitive regions. 
     There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
@@
     \caption{Association of indels in non-repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
+\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the GCAT logistic regression analysis for INDELs in non repetitive regions.
     There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
-    The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
+    The circles ( o ) are variants that reached values greater than 20, for clarity we set a hard ceiling at 20. 
   \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.}
@@
     \end{subfigure} 
-    \caption{Association of SNPS in repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
+    \caption{Association of SNPS in repetitive regions with $Q$. \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the GCAT logistic regression analysis for SNPs in repetitive regions. 
     There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
     The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.}
+  \textbf{B} QQ plot of the unadjusted p values for the GCAT logistic regression analysis for SNPs in repetitive regions.}
   \label{RS_Manhattan}
@@
     \end{subfigure} 
-    \caption{Association of indels in repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
+    \caption{Association of indels in repetitive regions with $Q$.  \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the GCAT logistic regression analysis for INDELs in repetitive regions. 
     There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
     The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.}
+  \textbf{B} QQ plot of the unadjusted p values for the GCAT logistic regression analysis for INDELs in repetitive regions.}
   \label{RI_Manhattan}

### 2019-07-13 LatexManuscript/legacy.tex
addressed some changes

--- 2019-07-13 LatexManuscript/legacy.tex
@@
 
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or \sgcomment{"or" in this context is ambiguous.
-Do you mean "this approach is called ..." or do you mean something else?} Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure.
+This approach is called Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
 We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
@@
 
-This method \sgcomment{GCAT?} identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+The GCAT method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
@@
 Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
-However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
-\sgcomment{I'm not sure what the previous sentence is saying}
+These results indicate that there is a need to remove these variants to avoid spurious GWAS associations.
 
@@
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
-Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
-These variants are only removed after the FDR \sgcomment{I thought we used distinct FDR For all four categories?} and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
+Variants flagged by the 1kGP strict mask are included in the association test and included in the four separate FDR adjustments for the four categories of variants (INDELs in Repeat and Non-Repeat regions and SNPs in Repeat and Non-Repeat regions).
+These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
 

### 2019-08-08 LatexManuscript/MBE/legacy.tex
moved things to one folder for mbe upload

--- 2019-07-13 LatexManuscript/legacy.tex
@@
 While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
-While investigating an intriguing putative population stratification within the Japanese population, we identified a batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
+While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
 Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
@@
 The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
+Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Genome Diversity project (SGDP)\citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
 Many more genomes are available indirectly through servers providing imputation services \citep{mccarthy2016reference} or summary statistics for variant frequency estimation \citep{Lek2016}.
@@
 
-Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants to estimate allele frequencies and to answer a wide range of medical and evolutionary questions.
+Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
 This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-Given that this signature had not yet been reproduced in another dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.
-
+\todo{While this signature had been partially reproduced in the SGDP, it had yet to be reproduced in a Japanese dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
@@
 \begin{figure}
-\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{../Figures/Figure1.jpg}}
+\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}}
 \caption{
@@
 The plot is zoomed in on frequencies below 0.5 for clarity.
-The size of blue dots are proportional to the number of suspicious SNPs in a given frequency bin (out of the 1034 identified by GWAS).  
+The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
 \textbf{C} 
-Mutation spectrum of these 1034 variants compared to the expectation from the distribution of all SNPs.
+Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
 The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
@@
 Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-Our initial goal was to determine whether this signature could be explained by population structure or by technological bias.
+Our initial goal was to determine whether this signature could be explained by population structure or technological error.
 While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
-When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}B).
+When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
 Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
@@
 \begin{figure}
-\includegraphics[width=0.95\hsize,keepaspectratio]{../Figures/MapQualOverTime.jpg}
+\includegraphics[width=0.95\hsize,keepaspectratio]{MapQualOverTime.jpg}
 
@@
 These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-Statistical significance of these enrichments is computed using a chi-squared test following \cite{Harris2017a}.
+Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}.
 
@@
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses where quality of imputation matters, identifying and discarding $Q$-associated variants may be preferable.
+For other analyses where quality of imputation matters, identifying $Q$-associated variants may be preferable.
 
@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 812 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each.
+Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
 
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure.
-This approach is called Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
+To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
 We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
@@
 
-The GCAT method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
 Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
@@
 We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interestingly, the variants identified by Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
-These results indicate that there may be a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
+Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
+These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
 
@@
                       & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
-{SNP}  & 3,369 &\textcolor{gray}{  $5.3\times 10^{-4}$}& 11,059 &\textcolor{gray}{  $5.6\times 10^{-4}$}& 14,428\\  
-{INDEL} & 181& \textcolor{gray}{  $3.0\times 10^{-4}$}& 657& \textcolor{gray}{  $6.6\times 10^{-4}$} & 838\\ \hline
+{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
+{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
 {Total} & 3,550& & 11,716 & & 15,270\\
@@
 Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
-Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes.
-The similarity in allele frequencies for these variants suggests that the biases in allele frequencies are present in populations other than the CHB and CHD.
+Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
 
@@
 We found that all of the variants associated with $Q$ were imputed back in the samples.
-This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the individuals with low $Q$.
-These $Q$-associated variants could also compromise the imputation of nearby real variants because haplotype matching would be compromised.
-However when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
+This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
+These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
 
-We searched the literature for GWAS that have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
+We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
 The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
@@
 
-These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
 They also removed rare alleles and alleles with high degrees of missingness.
-%Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
-Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
-These results indicate that there is a need to remove these variants to avoid spurious GWAS associations.
+Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
+Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
+However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
 
-These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct association.
+These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
 To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
@@
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-Even though these seven variants are reported in the GWAS catalogue, they were not flanked by LD peaks and were correctly removed from downstream analyses by \citep{Mandage2017}. 
-It is plausible that the EBV copy number phenotype estimation is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
+These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
 
@@
 To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer actively introducing bias in recent sequence data.
+Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
 
@@
 
-On a technical front, we were surprised that such strong associations between variants and technical covariates in the 1kGP project had not been identified before. 
+On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
 The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
 The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. 
+All three approaches accounting for population structure we tested produce comparable results.
 
@@
 \subsection{Code and data availability}
-Since this analysis is primarily performed using publicly available data, we provide fully reproducible and publicly available pipelines on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
+Since this data is primarily performed using publicly available data, we provide fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
 This repository includes scripts used for data download, processing, analysis and plotting.
@@
 For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al.
+as they applied the current state of the art quality thresholds to remove questionable sequences for detecting population level differences.
 Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
@@
 
-For the GCAT analysis, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
+For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
 We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
-Variants flagged by the 1kGP strict mask are included in the association test and included in the four separate FDR adjustments for the four categories of variants (INDELs in Repeat and Non-Repeat regions and SNPs in Repeat and Non-Repeat regions).
+Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
 These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
@@
 To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
-The deviance test statistic under the null model is approximately chi-square distributed with one degree of freedom.
+The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.
 
@@
 Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significant. 
+By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significance. 
 See Supplementary Data for a list of variants and adjusted \textit{p}-values.
@@
 \centering
-\includegraphics[width=13cm,keepaspectratio]{../Figures/fits_Significant_Positions_CompareModels.jpg}
+\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
 \caption{Comparison of three logistic regression models for testing association to $Q$.
@@
 \centering
-\includegraphics[width=8cm,keepaspectratio]{../Figures/histogram_of_sig_snps.jpg}
+\includegraphics[width=8cm,keepaspectratio]{histogram_of_sig_snps.jpg}
 \caption{Frequency distribution of the 1034 $Q$ associated SNPs from the single population test in the JPT in blue and all SNPs in black.
@@
 \centering
-\includegraphics[width=13cm,keepaspectratio]{../Figures/quality_by_center.jpg}
+\includegraphics[width=13cm,keepaspectratio]{quality_by_center.jpg}
 \caption{Sequencing date of samples included in the 1kGP coloured by sequencing centre. Improvements in sequencing platforms coincides with improvement in sequencing quality.}  
@@
 \begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{../Figures/SNPOverlap6.jpg}
+\includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
 
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{../Figures/strict_gcat_mutational_enrichment.jpg}
+\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
 \caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
@@
 \centering
-\includegraphics[width=13cm,keepaspectratio]{../Figures/Mafessoni_MutSpect.jpg}
+\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
 \caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
@@
 \centering
-\includegraphics[width=11cm,keepaspectratio]{../Figures/CountPerChrom.jpg}
+\includegraphics[width=11cm,keepaspectratio]{CountPerChrom.jpg}
 \caption{The proportion of $Q$-associated variants per chromosome.}  
@@
 \centering
-\includegraphics[width=14cm,keepaspectratio]{../Figures/VariantDensity.jpg}
+\includegraphics[width=14cm,keepaspectratio]{VariantDensity.jpg}
 \caption{\textbf{A} The number of $Q$-associated variants per 10kb window across the genome.
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{../Figures/RegionOver10.jpg}
+\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
 \caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{../Figures/Imputation_Distance_100s_gw.jpg}
+\includegraphics[width=10cm,keepaspectratio]{Imputation_Distance_100s_gw.jpg}
 \caption{Imputation scores are not impacted by nearby suspicious variants.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/OverUnder30.jpg}
+\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
 \caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
@@
 \centering
-\includegraphics[width=15cm,keepaspectratio]{../Figures/Published_OverUnder30.jpg}
+\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
 \caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
@@
 \centering
-\includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_singlePop.jpg}
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
 \caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
@@
 \centering
-\includegraphics[width=9cm,keepaspectratio]{../Figures/Han_1kGP_SFS_FullModel.jpg}
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
 \caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{ManhattanPlot_NonRepeatSNPs.jpg}
     \end{subfigure} 
@@
     	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{QQPlot_NonRepeatSNPs.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_NonRepeatIndels.jpg}
+        \includegraphics[width=\hsize]{ManhattanPlot_NonRepeatIndels.jpg}
     \end{subfigure} %
@@
     	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_NonRepeatIndels.jpg}
+        \includegraphics[width=\hsize]{QQPlot_NonRepeatIndels.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{ManhattanPlot_RepeatSNPs.jpg}
     \end{subfigure} %
@@
     	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatSNPs.jpg}
+        \includegraphics[width=\hsize]{QQPlot_RepeatSNPs.jpg}
 
@@
     \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{../Figures/ManhattanPlot_RepeatIndels.jpg}
+        \includegraphics[width=\hsize]{ManhattanPlot_RepeatIndels.jpg}
     \end{subfigure} %
@@
     	\center    
-        \includegraphics[width=\hsize]{../Figures/QQPlot_RepeatIndels.jpg}
+        \includegraphics[width=\hsize]{QQPlot_RepeatIndels.jpg}
     \end{subfigure} 
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{../Figures/Omni_AF.jpg}
+\includegraphics[width=11cm,keepaspectratio]{Omni_AF.jpg}
 \caption{Estimated frequency and association strength of $Q$- associated variants present on Illumina's Omni 2.5 chip. Variants highly associated to $Q$ tend to have low global allele frequencies.}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{../Figures/PC1_Correlation.jpg}
+\includegraphics[width=11cm,keepaspectratio]{PC1_Correlation.jpg}
 \caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project. 
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{../Figures/MutationSpectrum_cutOff.png}
+\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
 \caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  

### 2019-08-08 LatexManuscript/MBE/legacy_mbe.tex
moved things to one folder for mbe upload

--- 2019-08-08 LatexManuscript/MBE/legacy.tex
@@
-%%% PREAMBLE 
-\documentclass[9pt,article]{template}
+\documentclass[nogrid]{MBE}%
 
+\usepackage{url}
+%\usepackage[T1]{fontenc}
 \usepackage[export]{adjustbox}
@@
 }
+\usepackage{xcolor}
+
+\jshort{mst}
+
+\volname{}
+
+\jvolume{0}
+
+\jvol{}
+
+\jissue{0}
+
+\pubyear{2019}
+
+\mstype{Article}
+
+\artid{012}
 
@@
 \newcommand{\logit}{\operatorname{logit}}
-%\newcommand{\qq}{m}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Legacy Data Confounds Genomics Studies}
+\title{ {{L}}egacy Data Confounds Genomics Studies}
 
-\author[1,2]{Luke Anderson-Trocm\'e}
-\author[1,2]{Rick Farouni}
-\author[1,2]{Mathieu Bourgey}
-\author[3]{Yoichiro Kamatani}
-\author[3]{Koichiro Higasa}
-\author[4,5]{Jeong-Sun Seo}
-\author[4]{Changhoon Kim}
-\author[3]{Fumihiko Matsuda}
-\author[1,2]{Simon Gravel}
+\author[Anderson-Trocm\'e et al.]{
+{{L}}uke \surname{Anderson-Trocm\'e},$^{\ast,1,2}$ 
+Rick Farouni,$^{1,2}$ 
+Mathieu Bourgey,$^{1,2}$ 
+Yoichiro Kamatani,$^{3}$ 
+Koichiro Higasa,$^{3}$ 
+Jeong-Sun Seo,$^{4,5}$ 
+Changhoon Kim$^{4}$ 
+Fumihiko Matsuda$^{3}$ 
+and Simon Gravel$^{1,2}$}
 
-\affil[1]{Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada}
-\affil[2]{McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada}
-\affil[3]{Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
-\affil[4]{Bioinformatics Institute, Macrogen Inc., Seoul, 08511, Republic of Korea}
-\affil[5]{Precision Medicine Center, Seoul National University Bundang Hospital, Seongnam, 13605, Republic of Korea}
-\corr{simon.gravel@mcgill.ca}{SG}
+\address{$^{1}$Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada\\
+$^{2}$McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada\\ 
+$^{3}$Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan\\ 
+$^{4}$Bioinformatics Institute, Macrogen Inc., Seoul, 08511, Republic of Korea\\ 
+$^{5}$Precision Medicine Center, Seoul National University Bundang Hospital, Seongnam, 13605, Republic of Korea}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE START
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\history{Received 3 May 2019; reviews returned 27 May 2019}
+
+\coresp{E-mail: simon.gravel@mcgill.ca}
+
+\editor{Sara Tinsley}
+
+\abstract{Recent reports have identified differences in the mutational spectra across human populations.
+While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
+While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
+Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
+Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
+It may be time to retire or upgrade such legacy sequencing data.}
+
+\keyword{Batch Effect, Mutational Signature, Statistical Genetics, Population Genetics, Reference Cohorts, Imputation}
 
@@
 \maketitle
-\begin{abstract}
-Recent reports have identified differences in the mutational spectra across human populations.
-While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
-While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
-It may be time to retire or upgrade such legacy sequencing data.
-\vspace{5mm}
 
-\textbf{Key words} : Batch Effect, Mutational Signature, Statistical Genetics, Population Genetics, Reference Cohorts, Imputation
-\end{abstract}
-
-\section{Introduction}
+\section{{Introduction}\label{sec:Intro}}
 
@@
 
-\bibliography{Legacy}
+\bibliographystyle{natbib}%%%%natbib.sty
+\bibliography{legacy_mbe}%%%Legacy.bib
 

### 2019-08-08 PlottingScripts/ProcessPlotData.Rmd
moved things to one folder for mbe upload

@@
 NumPop$Pos<-as.numeric(as.character(NumPop$Pos))
+all_sites <- merge(Reg_mask,
+                   NumPop, 
+                   by = c('Chr','Pos')) %>%
+  group_by(Population, strict) %>%
+  summarize(total_count = n())
+all_sites[which(all_sites$Population%in% phase1),]$Population <- 
+  paste0('*',all_sites[which(all_sites$Population%in% phase1),]$Population)
 

### 2019-08-08 LatexManuscript/MBE/legacy.tex
modified sentence

@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\todo{While this signature had been partially reproduced in the SGDP, it had yet to be reproduced in a Japanese dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
+\todo{This signature had been partially reproduced at continental resolution in the SGDP. 
+However, due to smaller sample sizes, statistical power lacked to confirm the signature.
+Moreover, the result had yet to be reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.

### 2019-08-08 LatexManuscript/MBE/legacy.tex
chromosomal enrichment

@@
 Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
+\todo{The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal enrichment in the Japanese, however other signals identified by Harris and Pritchard appear unchanged including the continental enrichment of *AC${\rightarrow}$*CC signal in East Asians compared to Africans as reported by Harris and Pritchard and replicated in the SGDP (Figure \ref{MutSpect} and \ref{MutSpect_continental}).}
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
@@
 $Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
+\todo{Chromosome 1 shows enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite accounting for chromosomal length (Figure \ref{MutEnrichment_chrom}).} 
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
@@
 \centering
+\includegraphics[width=10cm,keepaspectratio]{mutational_enrichment_per_chrom.jpg}
+\caption{Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
+Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.}  
+\label{MutEnrichment_chrom}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
 \includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
@@
 
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
+\caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+The  *AC${\rightarrow}$*CC mutational signature in East Asian (EAS) Population remains despite removing individuals with average quality per mapped bases $Q$ below a threshold of 30.
+Points indicate significant mutation spectrum contrasts with $p <10^{-5}$}
+\label{MutSpect_continental}
+\end{figure}
+
 \end{document}

### 2019-08-08 LatexManuscript/old/legacy.tex
chromosomal enrichment

--- 2019-07-13 LatexManuscript/legacy.tex
@@
 Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
-These results indicate that there is a need to remove these variants to avoid spurious GWAS associations.
+Removing these variants from imputation servers could help avoid spurious associations.
 

### 2019-08-08 PlottingScripts/ProcessPlotData.Rmd
chromosomal enrichment

@@
 }
+```
+
+##per chromosome
+```{r}
+ChromLength <- fread(paste0(DataPath,
+                            'GRCh37_ChromLength.txt'),
+                     sep='\t',
+                     col.names = c('Chr','Length'))
+
+snp_list <- list()
+snp_combined <- data.frame()
+maxFreq <- 0
+for(chrom in 1:22) {
+  snp_list[[chrom]] <- 
+    processContext(SNP %>% filter(Chr == chrom))
+  if(max(snp_list[[chrom]]$Freq) > maxFreq){
+    maxFreq <- max(snp_list[[chrom]]$Freq)
+  }
+  snp_list[[chrom]]$Chr <- chrom
+  snp_combined <- rbind(snp_combined,snp_list[[chrom]])
+}
+
+plt_list <-list()
+for(chrom in 1:22) {
+  plt_list[[chrom]] <- 
+    plotContext(snp_list[[chrom]],paste('Chrom',chrom)) + 
+                theme(legend.position = "none")
+}
+
+plot_grid(plt_list[[1]] , plt_list[[2]], plt_list[[3]], plt_list[[4]], plt_list[[5]], plt_list[[6]], plt_list[[7]], plt_list[[8]], plt_list[[9]], plt_list[[10]], plt_list[[11]], plt_list[[12]], plt_list[[13]], plt_list[[14]], plt_list[[15]], plt_list[[16]], plt_list[[17]], plt_list[[18]], plt_list[[19]], plt_list[[20]],plt_list[[21]], plt_list[[22]], ncol=2)
+
+ggsave(paste0(figurePath,
+              'mutational_enrichment_per_chrom.jpg'),
+              height=24, 
+              width=12)
+
+snp_combined$title <- paste(snp_combined$Start,"*",snp_combined$End)
+
+snp_combined <- merge(snp_combined, ChromLength, by = 'Chr')
+
+snp_combined$norm_freq <- snp_combined$Freq / snp_combined$Length
+
+ggplot(snp_combined, 
+       aes(x=Mut, 
+           y = norm_freq, 
+           color = as.factor(Chr)))+
+  geom_point(shape =1) +
+  geom_text_repel(data = snp_combined %>% 
+              filter(Freq > 200),
+              show.legend = FALSE,
+    aes(label = title,
+        hjust = -1))+
+  guides(color=guide_legend(title="Chromosome"))
+
+ggplot(snp_combined, 
+       aes(x=Start, 
+           y = Freq, 
+           color = as.factor(Chr)))+
+  geom_point()+
+  facet_grid(End~Mut) +
+  guides(color=guide_legend(title="Chromosome"))+
+  theme(strip.placement = "outside",
+        strip.background = element_blank(),
+        strip.text = element_text(size = 20),
+        legend.text = element_text(size = 20),
+        legend.title = element_text(size = 20),
+        axis.text = element_text(size = 20), 
+        panel.border = element_rect(colour = "black", 
+                                    fill = NA, 
+                                    size = 1))
+ggsave(paste0(figurePath,
+              'mutational_spectrum_per_chrom.jpg'),
+              height=10, 
+              width=12)
+
+  geom_text_repel(data = snp_combined %>% 
+                    filter(Freq > 200),
+                  aes(x=Mut, 
+                  y = Freq,
+                  label = Chr))+
+  guides(color = F) +
+```
+
+```{r}
 

### 2019-08-12 LatexManuscript/MBE/legacy.tex
commit while modifying things

@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\todo{This signature had been partially reproduced at continental resolution in the SGDP. 
-However, due to smaller sample sizes, statistical power lacked to confirm the signature.
-Moreover, the result had yet to be reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
+
+\todo{
+An excess of this signature has been observed in East Asians compared to Afriacas in the SGDP. 
+At a continental resolution, this signature has been partially reproduced in the SGDP -- an independent dataset with smaller sample size from the 1kGP \citep{Aikens2018, Harris2017a}. 
+Nevertheless, the heterogeneous signature reported by \cite{Harris2017a} had yet to be reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.}
+The herertogeneous signatrue reported by harris et al had not , so we decided.
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
@@
 
-$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
-\todo{Chromosome 1 shows enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite accounting for chromosomal length (Figure \ref{MutEnrichment_chrom}).} 
+$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}A).
+\todo{Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes (Figure \ref{CountPerChrom}B).} 
 At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
@@
 \centering
-\includegraphics[width=10cm,keepaspectratio]{mutational_enrichment_per_chrom.jpg}
-\caption{Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
-Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.}  
-\label{MutEnrichment_chrom}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
 \includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
@@
 \includegraphics[width=11cm,keepaspectratio]{CountPerChrom.jpg}
-\caption{The proportion of $Q$-associated variants per chromosome.}  
+\caption{
+\textbf{A}
+The proportion of $Q$-associated variants per chromosome.
+\textbf{B}
+Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
+Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
+}  
 \label{CountPerChrom}

### 2019-08-12 LatexManuscript/MBE/legacy.tex
new supplementary plots and modified text

@@
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-
-\todo{
-An excess of this signature has been observed in East Asians compared to Afriacas in the SGDP. 
-At a continental resolution, this signature has been partially reproduced in the SGDP -- an independent dataset with smaller sample size from the 1kGP \citep{Aikens2018, Harris2017a}. 
-Nevertheless, the heterogeneous signature reported by \cite{Harris2017a} had yet to be reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.}
-The herertogeneous signatrue reported by harris et al had not , so we decided.
+\todo{An excess of this signature has been observed in East Asians compared to Africans in the 1kGP and reproduced in the SGDP  \citep{Aikens2018, Harris2017a}.
+The heterogeneous signature reported by \cite{Harris2017a} had not yet been reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.}
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
@@
 $Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}A).
-\todo{Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes (Figure \ref{CountPerChrom}B).} 
-At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
-An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
+\todo{Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite normalizing for the number of variants tested per chromosome (Figure \ref{CountPerChrom}B).}
+At a 1Mb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{CountPerChrom}C).
+%An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
 Distribution of association statistics in this region is provided in Figure \ref{Top10}.
-By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
+By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{CountPerChrom}D).
 
@@
 \centering
-\includegraphics[width=11cm,keepaspectratio]{CountPerChrom.jpg}
+\includegraphics[width=11cm,keepaspectratio]{PerChromosome.jpg}
 \caption{
@@
 Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
+\textbf{C} The number of $Q$-associated variants per 10kb window across the genome.
+Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
+One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
+\textbf{D} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
 }  

### 2019-08-12 LatexManuscript/MBE/legacy.tex
updated top regions plots

@@
 At a 1Mb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{CountPerChrom}C).
-%An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
-Distribution of association statistics in this region is provided in Figure \ref{Top10}.
+Three outlying 1Mb regions in chromosomes 1, 2 and 17 have over 30 $Q$-associated variants.
+Distribution of association statistics in these regions are provided in Figure \ref{Top10}.
 By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{CountPerChrom}D).
@@
 Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
+%One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
 \textbf{D} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
@@
 \centering
-\includegraphics[width=14cm,keepaspectratio]{VariantDensity.jpg}
-\caption{\textbf{A} The number of $Q$-associated variants per 10kb window across the genome.
-Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
-\textbf{B} Distribution of variant counts per 10kb window. Where variants that fail the strict mask are in grey and those that pass are in black. }  
-\label{VariantDensity}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
 \includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome.}  
 \label{Top10}

### 2019-08-13 LatexManuscript/MBE/legacy.tex
minor
reordered figures

@@
 
-To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
+To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}A).
 Traditionally, genome wide association studies use genotypes as the independent variable. 
 Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
-We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
+We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$ . 
 While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
@@
 To assess whether $Q$-associated variants are spurious variants resulting from sequencing artifacts or real variants exhibiting biased genotyping, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
-$Q$-associated variants that do not reproduce across sequencing experiments are likely the result of sequencing artifacts in the 1kGP. By contrast, 
-$Q$-associated variants that do reproduce across experiments could result from (recurrent) cell line mutations or, more likely, from existing variants whose genotyping depends on $Q$.
+$Q$-associated variants that do not reproduce across sequencing experiments are likely the result of sequencing artifacts in the 1kGP. 
+By contrast, $Q$-associated variants that do reproduce across experiments could result from (recurrent) cell line mutations or, more likely, from existing variants whose genotyping depends on $Q$. 
 Finally, given our nominal false discovery rate of $\alpha = 0.01$, we expect approximately $1\%$ of $Q$-associated variants to be false discoveries, i.e., variants that are not associated with $Q$ and that should therefore reproduce across experiments.
@@
 These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{OverUnder30}).
 The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
@@
 
-The variants identified in this study are likely to be technical artifacts from legacy technologies.
+The variants identified in this study \todo{could be explained by} technical artifacts from legacy technologies. %are likely to be
 Different sequencing technologies will have different error profiles. 
@@
 
-To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
+\todo{An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain biologically or technologically as neither are expected to produce chromosome-specific differences in error rates.
+Bioinformatics data processing errors are one possibility that would be consistent with this as chromosomes are commonly treated and analyzed separately.
+However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
+Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.}
 
@@
 For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the individuals with $Q<30$ and to filter out the LD-associated variants identified by \cite{mafessoni2018turning} and the $Q$-associated variants we identified (lists of such variants and sample IDs are provided in the Supplementary Data).
-We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
+We also recommend that imputation servers discard individuals with low $Q$ (or at least provide the option of performing the imputation without).
 Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.
@@
 \section{Supplementary Figures}
+\extrafloats{100}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
-\caption{Comparison of three logistic regression models for testing association to $Q$.
-These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
-In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
-\label{CompareModel}
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{PC1_Correlation.jpg}
+\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project. 
+The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.  }
+\label{PC1_Correlation}
 \end{figure}
@@
 \label{histogram_of_sig_snps}
+\end{figure}
+
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
+\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+\textbf{A} 
+The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
+\textbf{B} 
+Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
+\label{MutSpect}
+\end{figure}
+
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
+\caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+The  *AC${\rightarrow}$*CC mutational signature in East Asian (EAS) Population remains despite removing individuals with average quality per mapped bases $Q$ below a threshold of 30.
+Points indicate significant mutation spectrum contrasts with $p <10^{-5}$}
+\label{MutSpect_continental}
 \end{figure}
@@
 
-\begin{figure}[tbp]
+\begin{figure}
 \centering
-\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
-\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
-A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
-There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
-These three enrichments can be summarized as T*T${\rightarrow}$TTT.
-Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
-\label{strict_gcat_mutational_enrichment}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
-\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
-There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
-There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
-For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
-For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
-\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
-\label{Mafessoni_MutSpect}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=11cm,keepaspectratio]{PerChromosome.jpg}
-\caption{
-\textbf{A}
-The proportion of $Q$-associated variants per chromosome.
-\textbf{B}
-Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
-Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
-\textbf{C} The number of $Q$-associated variants per 10kb window across the genome.
-Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-%One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
-\textbf{D} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
-}  
-\label{CountPerChrom}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome.}  
-\label{Top10}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{Imputation_Distance_100s_gw.jpg}
-\caption{Imputation scores are not impacted by nearby suspicious variants.
-By comparing the imputation scores from Japanese Nagahama genotype data to the distance from the nearest suspicious variant, we find no impact on nearby variants.
-The only variants that do appear to have lower imputation scores are the suspicious variants themselves with a distance of zero on the plot.
-}
-\label{imputation_score}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
-\label{90HanSFS_full}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Each of the rsIDs of the variants are labelled for clarity.}
-\label{Pub_30}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
-\label{90HanSFS}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
-\label{90HanSFS_30}
+\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
+\caption{Comparison of three logistic regression models for testing association to $Q$.
+These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
+In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
+\label{CompareModel}
 \end{figure}
@@
 \begin{figure}[tbp]
+\centering
+\includegraphics[width=11cm,keepaspectratio]{PerChromosome.jpg}
+\caption{
+\textbf{A}
+The proportion of $Q$-associated variants per chromosome.
+\textbf{B}
+Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
+Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
+\textbf{C} The number of $Q$-associated variants per 1Mb window across the genome.
+Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
+Three regions not flagged by the 1000 Genomes Project strict mask in chromosomes 1, 2 and 17 have more than 30 variants per window.
+\textbf{D} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
+}  
+\label{CountPerChrom}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome.}  
+\label{Top10}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
+\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
+A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
+There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
+These three enrichments can be summarized as T*T${\rightarrow}$TTT.
+Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
+\label{strict_gcat_mutational_enrichment}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
+\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
+There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
+There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
+For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
+For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
+\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
+\label{Mafessoni_MutSpect}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
+\label{90HanSFS}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
+\label{90HanSFS_full}
+\end{figure}
+
+\begin{figure}[tbp]
 \includegraphics[width=11cm,keepaspectratio]{Omni_AF.jpg}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{PC1_Correlation.jpg}
-\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project. 
-The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.  }
-\label{PC1_Correlation}
+\centering
+\includegraphics[width=10cm,keepaspectratio]{Imputation_Distance_100s_gw.jpg}
+\caption{Imputation scores are not impacted by nearby suspicious variants.
+By comparing the imputation scores from Japanese Nagahama genotype data to the distance from the nearest suspicious variant, we find no impact on nearby variants.
+The only variants that do appear to have lower imputation scores are the suspicious variants themselves with a distance of zero on the plot.
+}
+\label{imputation_score}
 \end{figure}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
-\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
-\textbf{A} 
-The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
-\textbf{B} 
-Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
-\label{MutSpect}
+\centering
+\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
+\label{OverUnder30}
 \end{figure}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
-\caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
-The  *AC${\rightarrow}$*CC mutational signature in East Asian (EAS) Population remains despite removing individuals with average quality per mapped bases $Q$ below a threshold of 30.
-Points indicate significant mutation spectrum contrasts with $p <10^{-5}$}
-\label{MutSpect_continental}
+\centering
+\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Each of the rsIDs of the variants are labelled for clarity.}
+\label{Pub_30}
 \end{figure}

### 2019-08-13 LatexManuscript/MBE/legacy.tex
updated font size
couple of edits to edits

@@
 
-\usepackage[export]{adjustbox}
-\usepackage{lscape}
-\usepackage{afterpage}
+%\usepackage[export]{adjustbox}
+%\usepackage{lscape}
+%\usepackage{afterpage}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
-\usepackage{ulem}
+%\usepackage{ulem}
 \hypersetup{
@@
 
-In particular, the latter two studies  identified a heterogeneous mutational signature within 1kGP Japanese individuals.
+\todo{Another signature, *AC $\rightarrow$ *CC, has been observed at higher frequency in East Asians compared to Africans in the 1kGP and the SGDP  \citep{Aikens2018, Harris2017a}}. 
+ These two studies also found heterogeneous frequency of this signature among 1kGP Japanese individuals.
 This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
 A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\todo{An excess of this signature has been observed in East Asians compared to Africans in the 1kGP and reproduced in the SGDP  \citep{Aikens2018, Harris2017a}.
-The heterogeneous signature reported by \cite{Harris2017a} had not yet been reproduced in a Japanese dataset, so we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama.}
+This observation could not be reproduced in SGDP due to the small number of Japanese samples \citep{Aikens2018}.
+\todo{We therefore decided to follow up on this observation by using a newly sequenced dataset of Japanese individuals from Nagahama.}
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
@@
 \clearpage
+
 \section{Supplementary Figures}
+\clearpage
+
 \extrafloats{100}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
+\includegraphics[width=7cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
 \caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  

### 2019-08-14 LatexManuscript/MBE/legacy.tex
updated plots and text

@@
 Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
+\textbf{D} Blue bars indicate the density of variants passing the strict mask with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.
 Three regions not flagged by the 1000 Genomes Project strict mask in chromosomes 1, 2 and 17 have more than 30 variants per window.
-\textbf{D} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
+\textbf{E} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
 }  
@@
 \includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome.}  
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome. 
+In these plots, blue points are variants with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.}  
 \label{Top10}

### 2019-08-15 LatexManuscript/MBE/legacy.tex
removed blue text and edited .cls file for MBE upload

@@
 
-\todo{Another signature, *AC $\rightarrow$ *CC, has been observed at higher frequency in East Asians compared to Africans in the 1kGP and the SGDP  \citep{Aikens2018, Harris2017a}}. 
+Another signature, *AC $\rightarrow$ *CC, has been observed at higher frequency in East Asians compared to Africans in the 1kGP and the SGDP  \citep{Aikens2018, Harris2017a}. 
  These two studies also found heterogeneous frequency of this signature among 1kGP Japanese individuals.
@@
 This observation could not be reproduced in SGDP due to the small number of Japanese samples \citep{Aikens2018}.
-\todo{We therefore decided to follow up on this observation by using a newly sequenced dataset of Japanese individuals from Nagahama.}
+We therefore decided to follow up on this observation by using a newly sequenced dataset of Japanese individuals from Nagahama.
 While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
@@
 Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
-\todo{The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal enrichment in the Japanese, however other signals identified by Harris and Pritchard appear unchanged including the continental enrichment of *AC${\rightarrow}$*CC signal in East Asians compared to Africans as reported by Harris and Pritchard and replicated in the SGDP (Figure \ref{MutSpect} and \ref{MutSpect_continental}).}
+The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal enrichment in the Japanese, however other signals identified by Harris and Pritchard appear unchanged including the continental enrichment of *AC${\rightarrow}$*CC signal in East Asians compared to Africans as reported by Harris and Pritchard and replicated in the SGDP (Figure \ref{MutSpect} and \ref{MutSpect_continental}).
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
@@
 $Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}A).
-\todo{Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite normalizing for the number of variants tested per chromosome (Figure \ref{CountPerChrom}B).}
+Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite normalizing for the number of variants tested per chromosome (Figure \ref{CountPerChrom}B).
 At a 1Mb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{CountPerChrom}C).
@@
 
-The variants identified in this study \todo{could be explained by} technical artifacts from legacy technologies. %are likely to be
+The variants identified in this study could be explained by technical artifacts from legacy technologies. %are likely to be
 Different sequencing technologies will have different error profiles. 
@@
 
-\todo{An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain biologically or technologically as neither are expected to produce chromosome-specific differences in error rates.
+An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain biologically or technologically as neither are expected to produce chromosome-specific differences in error rates.
 Bioinformatics data processing errors are one possibility that would be consistent with this as chromosomes are commonly treated and analyzed separately.
 However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.}
+Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
 

### 2019-08-15 LatexManuscript/MBE/legacy.tex
edited the discussion

@@
 
-\subsection{Suspicious variants impact modern genomics analyses}
-
-State of the art imputation servers use a combination of many databases including some that are not freely available.
-From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
-
-To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
-We found that all of the variants associated with $Q$ were imputed back in the samples.
-This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
-
-We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-
-Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used genotyping chip data \citep{yucesoy2015genome, ellinghaus2016analysis}.
-
-All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
-They also removed rare alleles and alleles with high degrees of missingness.
-Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
-Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
-However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
-
-These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{OverUnder30}).
-The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
-By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
-One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
-It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
-
 \begin{table}[h!]
@@
 
+\subsection{Suspicious variants impact modern genomics analyses}
+
+State of the art imputation servers use a combination of many databases including some that are not freely available.
+From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
+
+To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
+We found that all of the variants associated with $Q$ were imputed back in the samples.
+This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
+These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
+
+We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
+The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
+
+Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
+One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used genotyping chip data \citep{yucesoy2015genome, ellinghaus2016analysis}.
+
+All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
+They also removed rare alleles and alleles with high degrees of missingness.
+Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
+Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
+However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
+
+These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{OverUnder30}).
+The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
+By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
+One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
+These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
+It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
+
 \section{Discussion}
@@
 
-An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain biologically or technologically as neither are expected to produce chromosome-specific differences in error rates.
-Bioinformatics data processing errors are one possibility that would be consistent with this as chromosomes are commonly treated and analyzed separately.
-However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
+An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic differences across chromosomes.
+Issues arising from subtle differences in how these data were integrated or processed are one explanation that would be consistent with this as chromosomes are commonly treated and analyzed separately.
+However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation, combination of sequencing technologies and data processing pipelines used throughout the 1kGP.
 Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
@@
 
-\bibliography{Legacy}
+%\bibliography{Legacy}
+\begin{thebibliography}{}
+
+\bibitem[{1000 Genomes Project Consortium},
+  2010]{1000GenomesProjectConsortium2010}
+{1000 Genomes Project Consortium} (2010).
+\newblock {A map of human genome variation from population-scale sequencing.}
+\newblock {\em Nature}, 467(7319):1061--73.
+
+\bibitem[{1000 Genomes Project Consortium},
+  2012]{The1000GenomesProjectConsortium2012}
+{1000 Genomes Project Consortium} (2012).
+\newblock {An integrated map of genetic variation}.
+\newblock {\em Nature}, 135:0--9.
+
+\bibitem[Aikens et~al., 2019]{Aikens2018}
+Aikens, R.~C., Johnson, K.~E., and Voight, B.~F. (2019).
+\newblock {Signals of Variation in Human Mutation Rate at Multiple Levels of
+  Sequence Context}.
+\newblock {\em Molecular Biology and Evolution}.
+
+\bibitem[Alexandrov et~al., 2013]{Alexandrov2013}
+Alexandrov, L.~B., Nik-Zainal, S., Wedge, D.~C., Aparicio, S.~A., Behjati, S.,
+  Biankin, A.~V., Bignell, G.~R., Bolli, N., Borg, A., B{\o}rresen-Dale, A.~L.,
+  Boyault, S., Burkhardt, B., Butler, A.~P., Caldas, C., Davies, H.~R.,
+  Desmedt, C., Eils, R., Eyfj{\"{o}}rd, J.~E., Foekens, J.~A., Greaves, M.,
+  Hosoda, F., Hutter, B., Ilicic, T., Imbeaud, S., Imielinsk, M., J{\"{a}}ger,
+  N., Jones, D.~T., Jonas, D., Knappskog, S., Koo, M., Lakhani, S.~R.,
+  L{\'{o}}pez-Ot{\'{i}}n, C., Martin, S., Munshi, N.~C., Nakamura, H.,
+  Northcott, P.~A., Pajic, M., Papaemmanuil, E., Paradiso, A., Pearson, J.~V.,
+  Puente, X.~S., Raine, K., Ramakrishna, M., Richardson, A.~L., Richter, J.,
+  Rosenstiel, P., Schlesner, M., Schumacher, T.~N., Span, P.~N., Teague, J.~W.,
+  Totoki, Y., Tutt, A.~N., Vald{\'{e}}s-Mas, R., {Van Buuren}, M.~M., {Van 'T
+  Veer}, L., Vincent-Salomon, A., Waddell, N., Yates, L.~R., Zucman-Rossi, J.,
+  {Andrew Futreal}, P., McDermott, U., Lichter, P., Meyerson, M., Grimmond,
+  S.~M., Siebert, R., Campo, E., Shibata, T., Pfister, S.~M., Campbell, P.~J.,
+  and Stratton, M.~R. (2013).
+\newblock {Signatures of mutational processes in human cancer}.
+\newblock {\em Nature}, 500(7463):415--421.
+
+\bibitem[Astle et~al., 2016]{astle2016allelic}
+Astle, W.~J., Elding, H., Jiang, T., Allen, D., Ruklisa, D., Mann, A.~L., Mead,
+  D., Bouman, H., Riveros-Mckay, F., Kostadima, M.~A., et~al. (2016).
+\newblock The allelic landscape of human blood cell trait variation and links
+  to common complex disease.
+\newblock {\em Cell}, 167(5):1415--1429.
+
+\bibitem[Balding and Nichols, 1995]{balding1995method}
+Balding, D.~J. and Nichols, R.~A. (1995).
+\newblock A method for quantifying differentiation between populations at
+  multi-allelic loci and its implications for investigating identity and
+  paternity.
+\newblock {\em Genetica}, 96(1-2):3--12.
+
+\bibitem[Benjamini et~al., 2006]{Benjamini2006}
+Benjamini, Y., Krieger, A.~M., and Yekutieli, D. (2006).
+\newblock {Adaptive linear step-up procedures that control the false discovery
+  rate}.
+\newblock {\em Biometrika}.
+
+\bibitem[Consortium et~al., 2015]{phase3}
+Consortium, . G.~P. et~al. (2015).
+\newblock A global reference for human genetic variation.
+\newblock {\em Nature}, 526(7571):68.
+
+\bibitem[Ebejer et~al., 2013]{Ebejer2013}
+Ebejer, J.~L., Duffy, D.~L., {Van Der Werf}, J., Wright, M.~J., Montgomery, G.,
+  Gillespie, N.~A., Hickie, I.~B., Martin, N.~G., and Medland, S.~E. (2013).
+\newblock {Genome-wide association study of inattention and
+  hyperactivity-impulsivity measured as quantitative traits}.
+\newblock {\em Twin Research and Human Genetics}.
+
+\bibitem[Ellinghaus et~al., 2016]{ellinghaus2016analysis}
+Ellinghaus, D., Jostins, L., Spain, S.~L., Cortes, A., Bethune, J., Han, B.,
+  Park, Y.~R., Raychaudhuri, S., Pouget, J.~G., H{\"u}benthal, M., et~al.
+  (2016).
+\newblock Analysis of five chronic inflammatory diseases identifies 27 new
+  associations and highlights disease-specific patterns at shared loci.
+\newblock {\em Nature genetics}, 48(5):510.
+
+\bibitem[Gao et~al., 2018]{gao2018genome}
+Gao, X.~R., Huang, H., Nannini, D.~R., Fan, F., and Kim, H. (2018).
+\newblock Genome-wide association analyses identify new loci influencing
+  intraocular pressure.
+\newblock {\em Human molecular genetics}, 27(12):2205--2213.
+
+\bibitem[Harris, 2015]{Harris2015a}
+Harris, K. (2015).
+\newblock {Evidence for recent, population-specific evolution of the human
+  mutation rate}.
+\newblock {\em Proceedings of the National Academy of Sciences},
+  112(11):3439--3444.
+
+\bibitem[Harris and Pritchard, 2017]{Harris2017a}
+Harris, K. and Pritchard, J.~K. (2017).
+\newblock {Rapid evolution of the human mutation spectrum}.
+\newblock {\em eLife}, 6.
+
+\bibitem[Herold et~al., 2016]{herold2016family}
+Herold, C., Hooli, B.~V., Mullin, K., Liu, T., Roehr, J.~T., Mattheisen, M.,
+  Parrado, A.~R., Bertram, L., Lange, C., and Tanzi, R.~E. (2016).
+\newblock Family-based association analyses of imputed genotypes reveal
+  genome-wide significant association of alzheimer?s disease with osbpl6,
+  ptprg, and pdcl3.
+\newblock {\em Molecular psychiatry}, 21(11):1608.
+
+\bibitem[{International HapMap Consortium}, 2005]{HapMap2005}
+{International HapMap Consortium} (2005).
+\newblock {A haplotype map of the human genome.}
+\newblock {\em Nature}, 437(7063):1299--320.
+
+\bibitem[Kraja et~al., 2011]{Kraja2011}
+Kraja, A.~T., Vaidya, D., Pankow, J.~S., Goodarzi, M.~O., Assimes, T.~L.,
+  Kullo, I.~J., Sovio, U., Mathias, R.~A., Sun, Y.~V., Franceschini, N.,
+  Absher, D., Li, G., Zhang, Q., Feitosa, M.~F., Glazer, N.~L., Haritunians,
+  T., Hartikainen, A.~L., Knowles, J.~W., North, K.~E., Iribarren, C., Kral,
+  B., Yanek, L., O'Reilly, P.~F., McCarthy, M.~I., Jaquish, C., Couper, D.~J.,
+  Chakravarti, A., Psaty, B.~M., Becker, L.~C., Province, M.~A., Boerwinkle,
+  E., Quertermous, T., Palotie, L., Jarvelin, M.~R., Becker, D.~M., Kardia,
+  S.~L., Rotter, J.~I., Chen, Y. D.~I., and Borecki, I.~B. (2011).
+\newblock {A bivariate genome-wide approach to metabolic syndrome: STAMPEED
+  Consortium}.
+\newblock {\em Diabetes}.
+
+\bibitem[Lan et~al., 2017]{Lan2017}
+Lan, T., Lin, H., Zhu, W., Laurent, T. C. A.~M., Yang, M., Liu, X., Wang, J.,
+  Wang, J., Yang, H., Xu, X., and Guo, X. (2017).
+\newblock Deep whole-genome sequencing of 90 han chinese genomes.
+\newblock {\em GigaScience}, 6(9):gix067.
+
+\bibitem[Lek et~al., 2016]{Lek2016}
+Lek, M., Karczewski, K.~J., Minikel, E.~V., Samocha, K.~E., Banks, E., Fennell,
+  T., O'Donnell-Luria, A.~H., Ware, J.~S., Hill, A.~J., Cummings, B.~B.,
+  Tukiainen, T., Birnbaum, D.~P., Kosmicki, J.~A., Duncan, L.~E., Estrada, K.,
+  Zhao, F., Zou, J., Pierce-Hoffman, E., Berghout, J., Cooper, D.~N., Deflaux,
+  N., DePristo, M., Do, R., Flannick, J., Fromer, M., Gauthier, L., Goldstein,
+  J., Gupta, N., Howrigan, D., Kiezun, A., Kurki, M.~I., Moonshine, A.~L.,
+  Natarajan, P., Orozco, L., Peloso, G.~M., Poplin, R., Rivas, M.~A.,
+  Ruano-Rubio, V., Rose, S.~A., Ruderfer, D.~M., Shakir, K., Stenson, P.~D.,
+  Stevens, C., Thomas, B.~P., Tiao, G., Tusie-Luna, M.~T., Weisburd, B., Won,
+  H.~H., Yu, D., Altshuler, D.~M., Ardissino, D., Boehnke, M., Danesh, J.,
+  Donnelly, S., Elosua, R., Florez, J.~C., Gabriel, S.~B., Getz, G., Glatt,
+  S.~J., Hultman, C.~M., Kathiresan, S., Laakso, M., McCarroll, S., McCarthy,
+  M.~I., McGovern, D., McPherson, R., Neale, B.~M., Palotie, A., Purcell,
+  S.~M., Saleheen, D., Scharf, J.~M., Sklar, P., Sullivan, P.~F., Tuomilehto,
+  J., Tsuang, M.~T., Watkins, H.~C., Wilson, J.~G., Daly, M.~J., and MacArthur,
+  D.~G. (2016).
+\newblock {Analysis of protein-coding genetic variation in 60,706 humans}.
+\newblock {\em Nature}, 536(7616):285--291.
+
+\bibitem[L{\'o}pez-Mej{\'\i}as et~al., 2017]{lopez2017genome}
+L{\'o}pez-Mej{\'\i}as, R., Carmona, F.~D., Casta{\~n}eda, S., Genre, F.,
+  Remuzgo-Mart{\'\i}nez, S., Sevilla-Perez, B., Ortego-Centeno, N., Llorca, J.,
+  Ubilla, B., Mijares, V., et~al. (2017).
+\newblock A genome-wide association study suggests the hla class ii region as
+  the major susceptibility locus for iga vasculitis.
+\newblock {\em Scientific reports}, 7(1):5088.
+
+\bibitem[Lutz et~al., 2015]{lutz2015genome}
+Lutz, S.~M., Cho, M.~H., Young, K., Hersh, C.~P., Castaldi, P.~J., McDonald,
+  M.-L., Regan, E., Mattheisen, M., DeMeo, D.~L., Parker, M., et~al. (2015).
+\newblock A genome-wide association study identifies risk loci for spirometric
+  measures among smokers of european and african ancestry.
+\newblock {\em BMC genetics}, 16(1):138.
+
+\bibitem[Mafessoni et~al., 2018]{mafessoni2018turning}
+Mafessoni, F., Prasad, R.~B., Groop, L., Hansson, O., and Pr{\"u}fer, K.
+  (2018).
+\newblock Turning vice into virtue: Using batch-effects to detect errors in
+  large genomic data sets.
+\newblock {\em Genome biology and evolution}, 10(10):2697--2708.
+
+\bibitem[Mallick et~al., 2016]{Mallick2016}
+Mallick, S., Li, H., Lipson, M., Mathieson, I., Gymrek, M., Racimo, F., Zhao,
+  M., Chennagiri, N., Nordenfelt, S., Tandon, A., Skoglund, P., Lazaridis, I.,
+  Sankararaman, S., Fu, Q., Rohland, N., Renaud, G., Erlich, Y., Willems, T.,
+  Gallo, C., Spence, J.~P., Song, Y.~S., Poletti, G., Balloux, F., {Van Driem},
+  G., {De Knijff}, P., Romero, I.~G., Jha, A.~R., Behar, D.~M., Bravi, C.~M.,
+  Capelli, C., Hervig, T., Moreno-Estrada, A., Posukh, O.~L., Balanovska, E.,
+  Balanovsky, O., Karachanak-Yankova, S., Sahakyan, H., Toncheva, D.,
+  Yepiskoposyan, L., Tyler-Smith, C., Xue, Y., Abdullah, M.~S., Ruiz-Linares,
+  A., Beall, C.~M., {Di Rienzo}, A., Jeong, C., Starikovskaya, E.~B., Metspalu,
+  E., Parik, J., Villems, R., Henn, B.~M., Hodoglugil, U., Mahley, R.,
+  Sajantila, A., Stamatoyannopoulos, G., Wee, J.~T., Khusainova, R.,
+  Khusnutdinova, E., Litvinov, S., Ayodo, G., Comas, D., Hammer, M.~F.,
+  Kivisild, T., Klitz, W., Winkler, C.~A., Labuda, D., Bamshad, M., Jorde,
+  L.~B., Tishkoff, S.~A., Watkins, W.~S., Metspalu, M., Dryomov, S., Sukernik,
+  R., Singh, L., Thangaraj, K., Pa{\"{a}}bo, S., Kelso, J., Patterson, N., and
+  Reich, D. (2016).
+\newblock {The Simons Genome Diversity Project: 300 genomes from 142 diverse
+  populations}.
+\newblock {\em Nature}, 538(7624):201--206.
+
+\bibitem[Mandage et~al., 2017]{Mandage2017}
+Mandage, R., Telford, M., Rodr{\'{i}}guez, J.~A., Farr{\'{e}}, X., Layouni, H.,
+  Marigorta, U.~M., Cundiff, C., Heredia-Genestar, J.~M., Navarro, A., and
+  Santpere, G. (2017).
+\newblock {Genetic factors affecting EBV copy number in lymphoblastoid cell
+  lines derived from the 1000 Genome Project samples}.
+\newblock {\em PLoS ONE}.
+
+\bibitem[Mathieson and Reich, 2017]{Mathieson2017a}
+Mathieson, I. and Reich, D. (2017).
+\newblock {Differences in the rare variant spectrum among human populations}.
+\newblock {\em PLoS Genetics}, 13(2).
+
+\bibitem[McCarthy et~al., 2016]{mccarthy2016reference}
+McCarthy, S., Das, S., Kretzschmar, W., Delaneau, O., Wood, A.~R., Teumer, A.,
+  Kang, H.~M., Fuchsberger, C., Danecek, P., Sharp, K., et~al. (2016).
+\newblock A reference panel of 64,976 haplotypes for genotype imputation.
+\newblock {\em Nature genetics}, 48(10):1279.
+
+\bibitem[Minoche et~al., 2011]{Minoche2011}
+Minoche, A.~E., Dohm, J.~C., and Himmelbauer, H. (2011).
+\newblock {Evaluation of genomic high-throughput sequencing data generated on
+  Illumina HiSeq and Genome Analyzer systems}.
+\newblock {\em Genome Biology}, 12(11).
+
+\bibitem[Nagy et~al., 2017]{nagy2017exploration}
+Nagy, R., Boutin, T.~S., Marten, J., Huffman, J.~E., Kerr, S.~M., Campbell, A.,
+  Evenden, L., Gibson, J., Amador, C., Howard, D.~M., et~al. (2017).
+\newblock Exploration of haplotype research consortium imputation for
+  genome-wide association studies in 20,032 generation scotland participants.
+\newblock {\em Genome medicine}, 9(1):23.
+
+\bibitem[Nishida et~al., 2018]{nishida2018key}
+Nishida, N., Sugiyama, M., Sawai, H., Nishina, S., Sakai, A., Ohashi, J., Khor,
+  S.-S., Kakisaka, K., Tsuchiura, T., Hino, K., et~al. (2018).
+\newblock Key hla-drb1-dqb1 haplotypes and role of the btnl2 gene for response
+  to a hepatitis b vaccine.
+\newblock {\em Hepatology}, 68(3):848--858.
+
+\bibitem[Park et~al., 2015]{park2015mercapturic}
+Park, S.~L., Carmella, S.~G., Chen, M., Patel, Y., Stram, D.~O., Haiman, C.~A.,
+  Le~Marchand, L., and Hecht, S.~S. (2015).
+\newblock Mercapturic acids derived from the toxicants acrolein and
+  crotonaldehyde in the urine of cigarette smokers from five ethnic groups with
+  differing risks for lung cancer.
+\newblock {\em PLoS One}, 10(6):e0124841.
+
+\bibitem[Pfeifer et~al., 2002]{Pfeifer2002}
+Pfeifer, G.~P., Denissenko, M.~F., Olivier, M., Tretyakova, N., Hecht, S.~S.,
+  and Hainaut, P. (2002).
+\newblock {Tobacco smoke carcinogens, DNA damage and p53 mutations in
+  smoking-associated cancers}.
+\newblock {\em Oncogene}, 21-48(6):7435--7451.
+
+\bibitem[Pleasance et~al., 2010]{Pleasance2010}
+Pleasance, E.~D., Stephens, P.~J., O'Meara, S., McBride, D.~J., Meynert, A.,
+  Jones, D., Lin, M.~L., Beare, D., Lau, K.~W., Greenman, C., Varela, I.,
+  Nik-Zainal, S., Davies, H.~R., Ord{\~{o}}ez, G.~R., Mudie, L.~J., Latimer,
+  C., Edkins, S., Stebbings, L., Chen, L., Jia, M., Leroy, C., Marshall, J.,
+  Menzies, A., Butler, A., Teague, J.~W., Mangion, J., Sun, Y.~A., McLaughlin,
+  S.~F., Peckham, H.~E., Tsung, E.~F., Costa, G.~L., Lee, C.~C., Minna, J.~D.,
+  Gazdar, A., Birney, E., Rhodes, M.~D., McKernan, K.~J., Stratton, M.~R.,
+  Futreal, P.~A., and Campbell, P.~J. (2010).
+\newblock {A small-cell lung cancer genome with complex signatures of tobacco
+  exposure}.
+\newblock {\em Nature}, 463(7278):184--190.
+
+\bibitem[Pritchard et~al., 2000]{pritchard2000inference}
+Pritchard, J.~K., Stephens, M., and Donnelly, P. (2000).
+\newblock Inference of population structure using multilocus genotype data.
+\newblock {\em Genetics}, 155(2):945--959.
+
+\bibitem[Shiraishi et~al., 2015]{Shiraishi2015a}
+Shiraishi, Y., Tremmel, G., Miyano, S., and Stephens, M. (2015).
+\newblock {A Simple Model-Based Approach to Inferring and Visualizing Cancer
+  Mutation Signatures}.
+\newblock {\em PLoS Genetics}, 11(12).
+
+\bibitem[Song et~al., 2015]{song2015testing}
+Song, M., Hao, W., and Storey, J.~D. (2015).
+\newblock Testing for genetic associations in arbitrarily structured
+  populations.
+\newblock {\em Nature genetics}, 47(5):550.
+
+\bibitem[Spracklen et~al., 2017]{spracklen2017association}
+Spracklen, C.~N., Chen, P., Kim, Y.~J., Wang, X., Cai, H., Li, S., Long, J.,
+  Wu, Y., Wang, Y.~X., Takeuchi, F., et~al. (2017).
+\newblock Association analyses of east asian individuals and trans-ancestry
+  analyses with european individuals reveal new loci associated with
+  cholesterol and triglyceride levels.
+\newblock {\em Human molecular genetics}, 26(9):1770--1784.
+
+\bibitem[Suhre et~al., 2017]{suhre2017connecting}
+Suhre, K., Arnold, M., Bhagwat, A.~M., Cotton, R.~J., Engelke, R., Raffler, J.,
+  Sarwath, H., Thareja, G., Wahl, A., DeLisle, R.~K., et~al. (2017).
+\newblock Connecting genetic risk to disease end points through the human blood
+  plasma proteome.
+\newblock {\em Nature communications}, 8:14357.
+
+\bibitem[Tian et~al., 2017]{tian2017genome}
+Tian, C., Hromatka, B.~S., Kiefer, A.~K., Eriksson, N., Noble, S.~M., Tung,
+  J.~Y., and Hinds, D.~A. (2017).
+\newblock Genome-wide association and hla region fine-mapping studies identify
+  susceptibility loci for multiple common infections.
+\newblock {\em Nature communications}, 8(1):599.
+
+\bibitem[van Dijk et~al., 2014]{VanDijk2014}
+van Dijk, E.~L., Auger, H., Jaszczyszyn, Y., and Thermes, C. (2014).
+\newblock {Ten years of next-generation sequencing technology}.
+\newblock {\em Trends in Genetics}, 30(9):418--426.
+
+\bibitem[Xu et~al., 2012]{xu2012genome}
+Xu, J., Mo, Z., Ye, D., Wang, M., Liu, F., Jin, G., Xu, C., Wang, X., Shao, Q.,
+  Chen, Z., et~al. (2012).
+\newblock Genome-wide association study in chinese men identifies two new
+  prostate cancer risk loci at 9q31. 2 and 19q13. 4.
+\newblock {\em Nature genetics}, 44(11):1231.
+
+\bibitem[Yucesoy et~al., 2015]{yucesoy2015genome}
+Yucesoy, B., Kaufman, K.~M., Lummus, Z.~L., Weirauch, M.~T., Zhang, G.,
+  Cartier, A., Boulet, L.-P., Sastre, J., Quirce, S., Tarlo, S.~M., et~al.
+  (2015).
+\newblock Genome-wide association study identifies novel loci associated with
+  diisocyanate-induced occupational asthma.
+\newblock {\em Toxicological Sciences}, 146(1):192--201.
+
+\end{thebibliography}
 
@@
 
-\extrafloats{100}
+%\extrafloats{100}
 

### 2019-08-15 LatexManuscript/MBE/legacy.tex
modified one sentence

@@
 An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic differences across chromosomes.
-Issues arising from subtle differences in how these data were integrated or processed are one explanation that would be consistent with this as chromosomes are commonly treated and analyzed separately.
+Subtle differences in how these data were integrated or processed could explain such differences, as chromosomes are commonly treated and analyzed separately. 
 However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation, combination of sequencing technologies and data processing pipelines used throughout the 1kGP.

### 2019-08-21 LatexManuscript/MBE/legacy.tex
final push

@@
 %\usepackage{ulem}
+\usepackage{textcomp}
 \hypersetup{
@@
 
-An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic differences across chromosomes.
-Subtle differences in how these data were integrated or processed could explain such differences, as chromosomes are commonly treated and analyzed separately. 
+An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic biases across chromosomes.
+Subtle differences in how these data were integrated or processed could explain such biases, as chromosomes are commonly treated and analyzed separately. 
 However, to pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation, combination of sequencing technologies and data processing pipelines used throughout the 1kGP.
@@
 We also recommend that imputation servers discard individuals with low $Q$ (or at least provide the option of performing the imputation without).
-Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.
+Given the value of freely accessible data, resequencing individuals with low $Q$ would  also likely be a worthwhile investment for the community.
 

### 2019-08-22 LatexManuscript/MBE/legacy.tex
supp figures

@@
 }
+% Cross ref with supplement
+\usepackage{xr}
+\externaldocument{supplementary_figures}
+\AtBeginDocument{\let\textlabel\label}
 
@@
 
-\clearpage
-
-\section{Supplementary Figures}
-\clearpage
-
-%\extrafloats{100}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{PC1_Correlation.jpg}
-\caption{Sequencing metrics against the prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1000 Genomes Project. 
-The average quality per mapped bases $Q$ per individual shows some clustering with individuals with low-quality data showing elevated rates of the signature.  }
-\label{PC1_Correlation}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,keepaspectratio]{histogram_of_sig_snps.jpg}
-\caption{Frequency distribution of the 1034 $Q$ associated SNPs from the single population test in the JPT in blue and all SNPs in black.
-The association test lacks power to identify variants at low frequencies.}
-\label{histogram_of_sig_snps}
-\end{figure}
-
-\begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
-\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
-\textbf{A} 
-The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
-\textbf{B} 
-Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
-\label{MutSpect}
-\end{figure}
-
-\begin{figure}[tbp]
-\includegraphics[width=7cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
-\caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
-The  *AC${\rightarrow}$*CC mutational signature in East Asian (EAS) Population remains despite removing individuals with average quality per mapped bases $Q$ below a threshold of 30.
-Points indicate significant mutation spectrum contrasts with $p <10^{-5}$}
-\label{MutSpect_continental}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=13cm,keepaspectratio]{quality_by_center.jpg}
-\caption{Sequencing date of samples included in the 1kGP coloured by sequencing centre. Improvements in sequencing platforms coincides with improvement in sequencing quality.}  
-\label{quality_by_center}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}
-
-\caption{Results from the single population test. We find a total of 1,165 variants associated with the average quality of mapped bases $Q$.
-Phase 1 populations are indicated by a star ( * ).
-\textbf{A} For each population, we plot the number of variants associated to $Q$.
-\textbf{B} The number of populations for which a variant has reached significance.}
-  \label{OverLap}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
-\caption{Comparison of three logistic regression models for testing association to $Q$.
-These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
-In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
-\label{CompareModel}
-\end{figure}
-
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{ManhattanPlot_NonRepeatSNPs.jpg}
-    \end{subfigure} 
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{QQPlot_NonRepeatSNPs.jpg}
-
-    \end{subfigure} 
-    \caption{Association of SNPS in non-repetitive regions with $Q$.
-    \textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions. 
-    There are 15,018 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
-    The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in non repetitive regions.}
-  \label{NRS_Manhattan}
-  \end{figure}
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{ManhattanPlot_NonRepeatIndels.jpg}
-    \end{subfigure} %
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{QQPlot_NonRepeatIndels.jpg}
-
-    \end{subfigure} 
-    \caption{Association of indels in non-repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.
-    There are 2,121 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
-    The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in non repetitive regions.}
-  \label{NRI_Manhattan}
-  \end{figure}
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{ManhattanPlot_RepeatSNPs.jpg}
-    \end{subfigure} %
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{QQPlot_RepeatSNPs.jpg}
-
-    \end{subfigure} 
-    \caption{Association of SNPS in repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.
-    There are 4,405 SNPs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
-    The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for SNPs in repetitive regions.}
-  \label{RS_Manhattan}
-  \end{figure}
-\begin{figure} \centering
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\hsize]{ManhattanPlot_RepeatIndels.jpg}
-    \end{subfigure} %
-    \begin{subfigure}[b]{\linewidth}
-    	\center    
-        \includegraphics[width=\hsize]{QQPlot_RepeatIndels.jpg}
-    \end{subfigure} 
-    \caption{Association of indels in repetitive regions with $Q$.
-\textbf{A} Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.
-    There are 642 INDELs that reach $p$ values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment.  
-    The circles ( o ) are variants that reached values greater than 20, for clarity we implemented hard ceiling at 20. 
-  \textbf{B} QQ plot of the unadjusted p values for the reverse GWAS logistic regression analysis for INDELs in repetitive regions.}
-  \label{RI_Manhattan}
-  \end{figure}
-
-\clearpage
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=11cm,keepaspectratio]{PerChromosome.jpg}
-\caption{
-\textbf{A}
-The proportion of $Q$-associated variants per chromosome.
-\textbf{B}
-Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
-Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
-\textbf{C} The number of $Q$-associated variants per 1Mb window across the genome.
-Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-\textbf{D} Blue bars indicate the density of variants passing the strict mask with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.
-Three regions not flagged by the 1000 Genomes Project strict mask in chromosomes 1, 2 and 17 have more than 30 variants per window.
-\textbf{E} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
-}  
-\label{CountPerChrom}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome. 
-In these plots, blue points are variants with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.}  
-\label{Top10}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
-\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
-A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
-There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
-These three enrichments can be summarized as T*T${\rightarrow}$TTT.
-Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
-\label{strict_gcat_mutational_enrichment}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
-\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
-There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
-There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
-For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
-For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
-\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
-\label{Mafessoni_MutSpect}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
-\label{90HanSFS}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
-\label{90HanSFS_full}
-\end{figure}
-
-\begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{Omni_AF.jpg}
-\caption{Estimated frequency and association strength of $Q$- associated variants present on Illumina's Omni 2.5 chip. Variants highly associated to $Q$ tend to have low global allele frequencies.}
-\label{Omni}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{Imputation_Distance_100s_gw.jpg}
-\caption{Imputation scores are not impacted by nearby suspicious variants.
-By comparing the imputation scores from Japanese Nagahama genotype data to the distance from the nearest suspicious variant, we find no impact on nearby variants.
-The only variants that do appear to have lower imputation scores are the suspicious variants themselves with a distance of zero on the plot.
-}
-\label{imputation_score}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
-\label{OverUnder30}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Each of the rsIDs of the variants are labelled for clarity.}
-\label{Pub_30}
-\end{figure}
-
 \end{document}

### 2019-08-22 LatexManuscript/mbe/supplementary_figures.tex
supp figures

--- 2019-08-08 LatexManuscript/MBE/legacy.tex
@@
 \usepackage{hyperref}
-\usepackage{ulem}
+%\usepackage{ulem}
+\usepackage{textcomp}
 \hypersetup{
@@
 }
-
-\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
-\newcommand{\luke}[1]{\textcolor{blue}{Luke: #1}}
-\newcommand{\todo}[1]{\textcolor{blue}{*#1*}}
-\newcommand{\alex}[1]{\textcolor{red}{Alex: #1}}
-\newcommand{\logit}{\operatorname{logit}}
-%\newcommand{\qq}{m}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Legacy Data Confounds Genomics Studies}
-
-\author[1,2]{Luke Anderson-Trocm\'e}
-\author[1,2]{Rick Farouni}
-\author[1,2]{Mathieu Bourgey}
-\author[3]{Yoichiro Kamatani}
-\author[3]{Koichiro Higasa}
-\author[4,5]{Jeong-Sun Seo}
-\author[4]{Changhoon Kim}
-\author[3]{Fumihiko Matsuda}
-\author[1,2]{Simon Gravel}
-
-\affil[1]{Department of Human Genetics, McGill University, Montreal, QC H3A 0G1, Canada}
-\affil[2]{McGill University and Genome Quebec Innovation Centre, Montreal, QC H3A 0G1, Canada}
-\affil[3]{Center for Genomic Medicine, Graduate School of Medicine, Kyoto University, Kyoto 606-8501, Japan}
-\affil[4]{Bioinformatics Institute, Macrogen Inc., Seoul, 08511, Republic of Korea}
-\affil[5]{Precision Medicine Center, Seoul National University Bundang Hospital, Seongnam, 13605, Republic of Korea}
-\corr{simon.gravel@mcgill.ca}{SG}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE START
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Cross ref with main text
+\usepackage{xr}
+\externaldocument{legacy}
 
@@
 
-\maketitle
-\begin{abstract}
-Recent reports have identified differences in the mutational spectra across human populations.
-While some of these reports have been replicated in other cohorts, most have been reported only in the 1000 Genomes Project (1kGP) data.
-While investigating an intriguing putative population stratification within the Japanese population, we identified a previously unreported batch effect leading to spurious mutation calls in the 1kGP data and to the apparent population stratification.
-Because the 1kGP data is used extensively, we find that the batch effects also lead to incorrect imputation by leading imputation servers and a small number of suspicious GWAS associations.
-Lower-quality data from the early phases of the 1kGP thus continues to contaminate modern studies in hidden ways.
-It may be time to retire or upgrade such legacy sequencing data.
-\vspace{5mm}
+\section{Supplementary Figures}
+\clearpage
 
-\textbf{Key words} : Batch Effect, Mutational Signature, Statistical Genetics, Population Genetics, Reference Cohorts, Imputation
-\end{abstract}
-
-\section{Introduction}
-
-\subsection{Batch Effects in Aging Reference Cohort Data}
-
-The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
-Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project (1kGP)\citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012,phase3}, and the Simons Genome Diversity project (SGDP)\citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses.
-Many more genomes are available indirectly through servers providing imputation services \citep{mccarthy2016reference} or summary statistics for variant frequency estimation \citep{Lek2016}.
-
-The first genomes in the 1kGP were sequenced 10 years ago \citep{VanDijk2014}. Since then, sequencing platforms have rapidly improved.  
-The second phase of the 1kGP implemented multiple technological and analytical improvements over its earlier phases \citep{The1000GenomesProjectConsortium2012,phase3}, leading to heterogeneous sample preparations and data quality over the course of the project.
-
-Yet, because of the extraordinary value of freely available data, early data from the 1kGP is still widely used to impute untyped variants, to estimate allele frequencies, and to answer a wide range of medical and evolutionary questions.
-This raises the question of whether and how such legacy data should be included in contemporary analyses alongside more recent cohorts.
-Mafessoni et al. recently identified batch effects in the 1kGP by looking for individuals with excess LD among distant variants. \citep{mafessoni2018turning}.
-Here we point out how these and additional unreported batch effects in the early phases of the 1kGP lead to incorrect genetic conclusions through population genetic analyses and spurious GWAS associations as a result of imputation using the 1kGP as a reference.
-
-\subsection{Mutational Signatures}
-
-Different mutagenic processes may preferentially affect different DNA motifs. 
-Certain mutagens in tobacco smoke, for example, have been shown to preferentially bind to certain genomic motifs leading to an excess of G to T transversions \citep{Pfeifer2002,Pleasance2010}. 
-Thus, exposure of populations to different mutational processes can be inferred by considering the DNA context of polymorphism in search of \textit{signatures} of different mutational processes \citep{Alexandrov2013,Shiraishi2015a}.
-Such genome-wide mutational signatures have been used as diagnostic tools for cancers (e.g., \cite{Alexandrov2013,Shiraishi2015a}).
-
-In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures which can be revealed in large sequencing panels. 
-In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}.
-Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
-Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.
-
-In particular, the latter two studies  identified a heterogeneous mutational signature within 1kGP Japanese individuals.
-This heterogeneity is intriguing because differences in germline signatures accumulate over many generations.
-A systematic difference within the Japanese population would suggest sustained environmental or genetic differences across sub-populations within Japan with little to no gene flow.
-\todo{While this signature had been partially reproduced in the SGDP, it had yet to be reproduced in a Japanese dataset, we decided to follow up on this observation, by using a newly sequenced dataset of Japanese individuals from Nagahama \citep{Harris2017a,Aikens2018}.}
-While we were unable to reproduce the mutational heterogeneity within the Japanese population, we could trace back the source of the discrepancy to a technical artefact in the 1kGP data.
-In addition to creating biases in mutational signatures, this artefact leads to spurious imputation results which have found their way in recent publications and online resources.
-
-The results section is organized as follows.
-We first attempt to reproduce the original signal and identify problematic variants in the JPT cohort from the 1kGP. 
-Next, we expand our analysis to the other populations in the 1kGP and identify lists of variants that show evidence for technical bias.
-Finally, we investigate how these variants have impacted modern genomics analyses.
-
-\begin{figure}
-\scalebox{0.85}{\includegraphics[width=\hsize,keepaspectratio]{Figure1.jpg}}
-\caption{
-Suspicious mutations carried by individuals with low quality data have distinct mutational profiles, reproduce poorly across studies, and are distributed across the genome.
-\textbf{A} 
-Genome wide association of the average quality of mapped bases $Q$ for the 104 Japanese individuals included in the 1000 Genomes Project. 
-This GWAS identified $1034$ SNPs associated to the average $Q$ of SNPs mapped for an individual, with  $p < 10^{-6}$
-($587$ SNPs had  $p < 10^{-8}.$)
-\textbf{B} 
-Joint frequency spectrum plot of the Japanese from the 1000 Genomes Project and a more recent Japanese dataset from Nagahama.
-The plot is zoomed in on frequencies below 0.5 for clarity.
-The size of blue dots are proportional to the number of variants in a given frequency bin that associate with $Q$ in the JPT. 
-\textbf{C} 
-Mutation spectrum of the 1034 variants that associated with $Q$ in the JPT($p < 10^{-6}$), compared to the expectation from the distribution of all SNPs.
-The majority of the variants with significant associations to $Q$ have the *AC${\rightarrow}$*CC mutational pattern. 
-There is also an enrichment in GA*${\rightarrow}$GG* and GC*${\rightarrow}$GG* mutations. 
-These three enrichments can be summarized as G**${\rightarrow}$GG*. 
-Stars ( * ) indicate a significant deviation from the expected mutational spectrum from all SNPs.
-}
- \label{SFS}
-\end{figure}
-
-\section{Results}
-
-\subsection{A peculiar mutational signature in Japan}
-
-Harris and Pritchard reported an excess of a 3-mer substitution patterns *AC${\rightarrow}$*CC in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
-Our initial goal was to determine whether this signature could be explained by population structure or technological error.
-While trying to follow up on this observation in a larger and more recent Japanese cohort from Nagahama, we did not find this particular signature.
-When comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed a number of single nucleotide polymorphisms (SNPs) private to one of the two groups (Figure \ref{SFS}).
-Given the similarity of the two populations, this strongly suggests a technical difference rather than a population structure effect.
-These mismatches were maintained despite only considering sites that satisfied strict quality masks and Hardy-Weinberg equilibrium in both cohorts.
-
-When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Figure \ref{PC1_Correlation}).
-The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. 
-Thus, sequences called from low-$Q$ data contain variants that reproduce poorly across studies and exhibit a particular mutational signature.
-
-\begin{figure}
-\includegraphics[width=0.95\hsize,keepaspectratio]{MapQualOverTime.jpg}
-
-\caption{Sampling and sequencing technologies over time in the 1000 Genomes Project.
-\textbf{A} The average quality of mapped bases $Q$ for each individual per population included in the 1000 Genomes Project. 
- Populations are ranked by mean sequencing date (the earliest sequencing date was used for individuals with multiple dates). 
-The shape indicates whether the individual was first released in Phase 1 or Phase 3 of the 1000 Genomes project. 
-\textbf{B} Sequencing technologies used over the course of the 1000 Genomes Project.}
-\label{MapQual}
-\end{figure}
-
-To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
-Traditionally, genome wide association studies use genotypes as the independent variable. 
-Here we perform a genotype conditional association test (GCAT), where genotypes are now the dependent variable that we predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
-We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
-While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$).
-The variants that are associated to $Q$ have a significant enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}C).
-These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.
-Statistical significance of these enrichment is computed using a chi-squared test following \cite{Harris2017a}.
-
-Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal, however other signals identified by Harris and Pritchard appear unchanged (Figure \ref{MutSpect}).
-For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
-For other analyses where quality of imputation matters, identifying $Q$-associated variants may be preferable.
-
-\subsection{Identifying suspicious variants in the 1000 Genomes Project}
-The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual} and Figure \ref{quality_by_center}).
-This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
-By 2011, older sequencing technologies were phased out, and methods became more consistent, resulting in higher and more uniform quality.
-
-We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
-
-To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}.
-We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
-These models were broadly consistent (See Figure \ref{CompareModel}).
-
-This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
-Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
-To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
-We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
-Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.
-
-$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}).
-At a 10kb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{VariantDensity}).
-An outlying 10kb region in chromosome 17 (bases : 22,020,000 to  22,030,000) has 35 $Q$-associated variants.
-Distribution of association statistics in this region is provided in Figure \ref{Top10}.
-By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{VariantDensity}).
-
-The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
-There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT (Figure \ref{strict_gcat_mutational_enrichment}).
-
-A recent publication by Mafessoni et al.
-also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}A).
-They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$.
-We find that 1,279 ( 3\% ) of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method.
-We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
-These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{l  r r r  r r}
-                      & {Repeat} & \textcolor{gray}{Proportion} & {Non-Repeat} & \textcolor{gray}{Proportion}   & {Total}    \\ \hline
-{SNP}  & 3,369 &\textcolor{gray}{0.53\textperthousand} & 11,059 & \textcolor{gray}{0.56\textperthousand} & 14,428\\  
-{INDEL} & 181& \textcolor{gray}{0.3\textperthousand} & 657& \textcolor{gray}{0.66\textperthousand} & 838\\ \hline
-{Total} & 3,550& & 11,716 & & 15,270\\
-\end{tabular}
-\caption{Number and proportion of $Q$-associated variants passing the 1000 Genomes Project strict mask per category. 
-Variants that are flagged by the 1000 Genomes Project nested repeat mask file were analyzed separately for FDR calculation. SNPs and INDELs were also analyzed separately.
-A total of 15,270 are statistically significantly associated to $Q$ and pass the 1kGP strict mask. 
-The grey text is the proportion of $Q$-associated variants per category.
-The number of variants included in the analysis for SNPs, SNPs in repeat regions, INDELs and INDELs in repeat regions are  19,846,786, 6,312,620, 1,770,315 and 586,342 respectively.}
-\label{sigTable}
-\end{table}
-
-\subsection{Spurious variants, biased genotypes, or cell line artefacts?}
-To assess whether $Q$-associated variants are spurious variants resulting from sequencing artifacts or real variants exhibiting biased genotyping, we compare the original 1kGP sequence data to more recent sequencing efforts using the same cell lines.
-$Q$-associated variants that do not reproduce across sequencing experiments are likely the result of sequencing artifacts in the 1kGP. By contrast, 
-$Q$-associated variants that do reproduce across experiments could result from (recurrent) cell line mutations or, more likely, from existing variants whose genotyping depends on $Q$.
-Finally, given our nominal false discovery rate of $\alpha = 0.01$, we expect approximately $1\%$ of $Q$-associated variants to be false discoveries, i.e., variants that are not associated with $Q$ and that should therefore reproduce across experiments.
-
-In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-Among the 296 such variants that were $Q$-associated in single-population tests for the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}).   
-This is slightly more but consistent with the 3 false positives predicted according to the $\alpha = 0.01$ nominal false discovery rate. 
-Thus the majority of $Q$-associated variants in the CHB or CHS  appear to be spurious variants.
-
-We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
-The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, suggesting that they are spurious variants.
-Five variants show differing frequencies between both datasets and are likely explained by biased genotypes.
-Finally, 1,139 (or 18\%) are present in the resequenced data at comparable frequencies.
-Because this is higher than the false discovery rate, we conclude that the cohort-wide GCAT test identifies predominantly spurious variants, but also variants that have biased genotypes in populations other than the CHB/CHD.
-
-Finally, among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
-These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
-
-\subsection{Suspicious variants impact modern genomics analyses}
-
-State of the art imputation servers use a combination of many databases including some that are not freely available.
-From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.
-
-To investigate whether suspicious calls from the 1kGP are imputed into genotyping studies, we submitted genotype data for the first two chromosomes of the 1kGP genotype data to the Michigan Imputation Server.
-We found that all of the variants associated with $Q$ were imputed back in the samples.
-This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
-
-We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
-The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
-
-Eleven of these studies included the 1kGP in their reference panel for imputation \citep{xu2012genome, lutz2015genome, park2015mercapturic, astle2016allelic, herold2016family,  suhre2017connecting, lopez2017genome, tian2017genome,  spracklen2017association,  nagy2017exploration, gao2018genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
-One study used an in-house reference panel for imputation \citep{nishida2018key}, two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013} and two studies used genotyping chip data \citep{yucesoy2015genome, ellinghaus2016analysis}.
-
-All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
-They also removed rare alleles and alleles with high degrees of missingness.
-Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
-Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
-However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.
-
-These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
-The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
-By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
-One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
-These seven variants were not flanked by LD peaks and were correctly removed from downstream analyses however, they were still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
-It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.
-
-\begin{table}[h!]
-\begin{tabular}{l l l r r r}
-  {Pubmed ID}  & {Disease/Trait} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$\\-$\log_{10} p$\\(adjusted)}      \\ \hline
-28654678 & EBV copy number in & rs201761909 & 5.7 & 78.11\\
-	&   lymphoblastoid cell lines	&  	rs201130852 & 5.05 & 72.28\\
-	& 	&  	rs201255786 & 5.7 & 68.97\\
-	& 	&  	rs200655768 & 6.52 & 66.67\\
-	& 	&  	rs184202621 & 5.52 & 60.45\\
-	& 	& 	rs80274284 & 6 & 56.15\\
-	& 	& 	rs200699422 & 5.3 & 7.43\\
-23527680 & \multicolumn{1}{p{3cm}}{\raggedright ADHD\textsuperscript{\dag}} & rs6057648 & 5.4 & 20.5\\
-28928442 & \multicolumn{1}{p{3cm}}{\raggedright Cold sores} & rs201471471 & 6.52 & 7.87\\
-26053186 & \multicolumn{1}{p{3cm}}{\raggedright HMPMA\textsuperscript{\ddag} levels in smokers} & rs60136336 & 5.7 & 2.25\\ \hline
-\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
-\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Prostate cancer}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{HDL cholesterol}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Blood protein levels}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
-\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{High light scatter\\reticulocyte count}} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
-\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Response to\\hepatitis B vaccine}} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
-\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Metabolic syndrome}}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
-\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Alzheimer disease\\and age of onset}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
-\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Intraocular pressure}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
-\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Immunoglobulin A vasculitis}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
-\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Chronic inflammatory diseases}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
-\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Post bronchodilator}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
- 	& \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{FEV1/FVC ratio}} 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
-	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
-\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diisocyanate-induced asthma}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\
-
- \hline
-\end{tabular}
-\caption{Recent publications that reported $Q$-associated variants as close to or above the genome-wide significant threshold.
-The variants reaching genome wide significance have a star ( * ).
-The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure \ref{Pub_30}).
-\textsuperscript{\dag} Attention deficit hyperactivity disorder. 
-\textsuperscript{\ddag} 3-hydroxy-1-methylpropylmercapturic acid.
-}
-\label{gwasTable}
-\end{table}
-
-\section{Discussion}
-
-The variants identified in this study are likely to be technical artifacts from legacy technologies.
-Different sequencing technologies will have different error profiles. 
-A report comparing the Genome Analyzer II (GAII) to the Illumina HiSeq found that the GAII had much higher rates of reads below a quality score of 30 \citep{Minoche2011} with, for instance, different patterns of quality decrease along reads.
-Differences in read quality and error profiles in turn require different calling pipelines.
-
-To pinpoint the precise technical source of the discrepancy would require further forensic inquiries into the details of the heterogeneous sample preparation and data processing pipelines used throughout the 1kGP.
-Given the progress in sequencing and calling that occurred since the early phases of the 1kGP (Figure \ref{MapQual}), it is likely that the source of these biases is not longer being actively introduced in recent sequence data.
-
-However, because the 1kGP data is widely used as a reference database, these variants are still being imputed onto new genotype data and can then impact association studies for a variety of phenotypes.
-Even though significant association of a variant with a quality metric is not in itself an indication that the variant is spurious, we would recommend to carefully examine GWAS associations for such variants, e.g.
-by repeating the analysis without the 1kGP as part of the imputation panel.
-
-For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the individuals with $Q<30$ and to filter out the LD-associated variants identified by \cite{mafessoni2018turning} and the $Q$-associated variants we identified (lists of such variants and sample IDs are provided in the Supplementary Data).
-We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without).
-Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.
-
-\section{Conclusion}
-
-On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
-The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
-The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. 
-All three approaches accounting for population structure we tested produce comparable results.
-
-More generally, to improve the quality of genomic reference datasets, we can proceed by addition of new and better data and by better curation of existing data.
-Given rapid technological progress, the focus of genomic research is naturally on the data generation side. 
-However, cleaning up existing databases is also important to avoid generating spurious results. 
-The present findings suggest that a substantial fraction of data from the final release of the 1kGP project is overdue for retirement or re-sequencing.
-
-\section{Methods}
-\subsection{Code and data availability}
-Since this data is primarily performed using publicly available data, we provide fully reproducible and publicly available on \href{https://github.com/LukeAndersonTrocme/LegacyData}{GitHub}.
-This repository includes scripts used for data download, processing, analysis and plotting.
-%Supplementary data as well as preprocessed data for plotting are also available on GitHub.
-
-\subsection{Metadata}
-The metadata used in this analysis was compiled from each of the index files from the 1kGP file system. 
-Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file.
-Each BAS file has metadata regarding each sequencing event for each sample. 
-If a sample was sequenced more than once, we took the average of each $Q$ score from each sequencing instance. 
-The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.
-
-\subsection{Quality Controls}
-For the mutation spectrum analysis, we reproduced the quality control and data filtering pipelines used by Harris et al.
-as they applied the current state of the art quality thresholds to remove questionable sequences for detecting population level differences.
-Several mask files were applied to remove regions of the genome that might be lower quality, or might have very different mutation rates or base pair complexity compared to the rest of the genome. 
-The 1kGP strict mask was used to remove low quality regions of the genome, highly conserved regions were removed using the phastCons100way mask file and highly repetitive regions were removed using the NestedRepeats mask file from RepeatMasker.
-Furthermore, only sites with missingness below 0.01, MAF less than 0.1, and MAF greater than 0.9 were considered.
-In total, 7,786,023 diallelic autosomal variants passed our quality controls for the mutation spectrum analysis.
-We calculated the mutation spectrum of base pair triplets for the list of significant variants for the JPT population using a similar method as described in  \citep{Harris2017a}.
-
-For the reverse GWAS, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test.
-We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
-Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
-These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
-
-\subsection{Testing the association of quality to genotype}
-When conducting a statistical analysis of population genetics data, we must account for population structure.
-In a typical GWAS, we are interested in modelling the phenotype as a function of the genotype.
-Here we have the opposite situation, where the quantitative variable ($Q$) is used as an explanatory variable. 
-So we consider models where the genotype $y$ is a function of an expected frequency $\pi_{si},$ based on population structure, and $Q$. 
-The null model is 
-\begin{align} \label{lfa_basic}
-y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big).
-\end{align} 
-The expected frequency for a SNP $s$ and individual $i$ can be estimated using principal component analysis, categorical population labels, or logistic factor analysis \citep{song2015testing}.
-The alternative model then takes in $Q$ as a covariate:
-
-\begin{align}\label{gcat_basic}
- y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, \logit^{-1}\Big(\logit(\pi_{si}) + \beta_s q_i\Big) \bigg).
-\end{align}
-
-Under the null hypothesis the slope coefficient $\beta_s$ is zero and Model \eqref{gcat_basic} reduces to Model \eqref{lfa_basic}. 
-$\beta_s$ denotes the association to average quality of mapped bases $Q$ to genotype $y_{s}$. 
-To test the null hypothesis, we use the generalized likelihood ratio test statistic, whose deviance is a measure of the marginal importance of adding $Q$ in the model. 
-The deviance test statistic under the null model is approximately chi-square distributed with one degrees of freedom.
-
-We run a total of $S$ regressions, where $S$ is the total number of genomic loci.
-Given the large number of tests, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
-By using a nominal Type-I error rate $\alpha = 0.01$, a total of 15,270 variants were found to be statistically significance. 
-See Supplementary Data for a list of variants and adjusted \textit{p}-values.
-
-\subsection{Individual-specific allele frequency}
-Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}.
-These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization. 
-\begin{align}
-\mathbf{L }= \mathbf{AH}
-\end{align} 
-where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i^\text{th}$ individual and the $s^\text{th}$ row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$.
-When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
-\begin{align} \label{lfa}
-y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
-\end{align} 
-for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
-
-To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing}.
-The GCAT is a regression approach that assumes the following model.
-\begin{align}\label{gcat}
- y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
-\end{align} 
-for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
-The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and are therefore used directly in the model.
-We approximated the population structure using $K=5$ latent components from a subsampled genotype matrix consisting of $M = 2,306,130$ SNPs (we picked SNPs from the 1kGP OMNI 2.5). 
-To avoid possible biases in computing PCA from the biased variants, we considered the genotype matrix $L$ obtained by downsampling 1kGP variants the positions from the OMNI 2.5M chip.
-
-\subsection{Imputation}
-Using the Michigan Imputation Server, we imputed the genotype data from 1kGP for chromosomes 1 and 2.
-We used the genotyped data from the 1kGP Omni 2.5M chip genotype data.
-The VCF file returned from the server was then downloaded and used to search for the number of significant variants successfully imputed.
-
-\section{Acknowledgments}
-We would like to thank K. Harris for sharing her mutation spectrum scripts, F. Mafessoni, G. Santpere and A. Navarro for feedback on earlier versions of the manuscript.
-We would also like to thank the members of the Gravel lab for their help with coding and useful discussions.
-
-\bibliography{Legacy}
-
-\clearpage
-\section{Supplementary Figures}
+%\extrafloats{100}
 
@@
 \label{histogram_of_sig_snps}
+\end{figure}
+
+\begin{figure}[tbp]
+\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
+\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+\textbf{A} 
+The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
+\textbf{B} 
+Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
+\label{MutSpect}
+\end{figure}
+
+\begin{figure}[tbp]
+\includegraphics[width=7cm,keepaspectratio]{CutOffID_QUAL30_heatmap_v_AFR_nosingle.jpg}
+\caption{Comparing continental mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
+The  *AC${\rightarrow}$*CC mutational signature in East Asian (EAS) Population remains despite removing individuals with average quality per mapped bases $Q$ below a threshold of 30.
+Points indicate significant mutation spectrum contrasts with $p <10^{-5}$}
+\label{MutSpect_continental}
 \end{figure}
@@
 
-\begin{figure}[tbp]
+\begin{figure}
 \centering
-\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
-\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
-A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
-There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
-These three enrichments can be summarized as T*T${\rightarrow}$TTT.
-Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
-\label{strict_gcat_mutational_enrichment}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
-\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
-There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
-There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
-For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
-For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
-\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
-\label{Mafessoni_MutSpect}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=11cm,keepaspectratio]{CountPerChrom.jpg}
-\caption{The proportion of $Q$-associated variants per chromosome.}  
-\label{CountPerChrom}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=14cm,keepaspectratio]{VariantDensity.jpg}
-\caption{\textbf{A} The number of $Q$-associated variants per 10kb window across the genome.
-Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
-One region not flagged by the 1000 Genomes Project strict mask in chromosome 17 has more than 10 variants per window.
-\textbf{B} Distribution of variant counts per 10kb window. Where variants that fail the strict mask are in grey and those that pass are in black. }  
-\label{VariantDensity}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
-\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the 10kb window with the most  $Q$-associated variants per 10kb across the genome.}  
-\label{Top10}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=10cm,keepaspectratio]{Imputation_Distance_100s_gw.jpg}
-\caption{Imputation scores are not impacted by nearby suspicious variants.
-By comparing the imputation scores from Japanese Nagahama genotype data to the distance from the nearest suspicious variant, we find no impact on nearby variants.
-The only variants that do appear to have lower imputation scores are the suspicious variants themselves with a distance of zero on the plot.
-}
-\label{imputation_score}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
-\label{90HanSFS_full}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
-\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
-The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
-Each of the rsIDs of the variants are labelled for clarity.}
-\label{Pub_30}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
-\label{90HanSFS}
-\end{figure}
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
-\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
-Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
-\label{90HanSFS_30}
+\includegraphics[width=13cm,keepaspectratio]{fits_Significant_Positions_CompareModels.jpg}
+\caption{Comparison of three logistic regression models for testing association to $Q$.
+These methods model each genotype as a logistic function using principal components (PC), Population membership (Pop) or LFA as an offset.
+In these plots we are comparing the deviance from the null model in the 15,270 variants identified using the LFA model.}  
+\label{CompareModel}
 \end{figure}
@@
 \begin{figure}[tbp]
+\centering
+\includegraphics[width=11cm,keepaspectratio]{PerChromosome.jpg}
+\caption{
+\textbf{A}
+The proportion of $Q$-associated variants per chromosome.
+\textbf{B}
+Mutation spectrum enrichment of  $Q$-associated variants across each chromosome.
+Chromosome 1 has a strong enrichment in *AC${\rightarrow}$*CC mutational pattern.
+\textbf{C} The number of $Q$-associated variants per 1Mb window across the genome.
+Grey bars indicate regions within and black bars indicate regions outside the 1000 Genomes Project strict mask.
+\textbf{D} Blue bars indicate the density of variants passing the strict mask with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.
+Three regions not flagged by the 1000 Genomes Project strict mask in chromosomes 1, 2 and 17 have more than 30 variants per window.
+\textbf{E} Distribution of variant counts per 1Mb window. Where variants that fail the strict mask are in grey and those that pass are in black. 
+}  
+\label{CountPerChrom}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=10cm,keepaspectratio]{RegionOver10.jpg}
+\caption{Manhattan plot of the $-\log_{10}(p)$ values for the reverse GWAS logistic regression analysis for the three 1Mb windows with the most  $Q$-associated variants per 1Mb across the genome. 
+In these plots, blue points are variants with the *AC${\rightarrow}$*CC mutational pattern whereas red points have the T*T${\rightarrow}$TTT mutational pattern.}  
+\label{Top10}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=10cm,keepaspectratio]{strict_gcat_mutational_enrichment.jpg}
+\caption{Mutation spectrum of the 15,270 $Q$-associated variants that pass the 1kGP strict mask filter, compared to a random set of non-associated variants.
+A large fraction of the  $Q$-associated variants have the *AC${\rightarrow}$*CC mutational pattern.
+There is also an enrichment in TAT${\rightarrow}$TTT, TCT${\rightarrow}$TTT and  TGT${\rightarrow}$TTT (shown as reverse complement  ACA${\rightarrow}$AAA).
+These three enrichments can be summarized as T*T${\rightarrow}$TTT.
+Stars ( * ) indicate significant difference from the mutational spectrum of a set of control variants of equal size.}
+\label{strict_gcat_mutational_enrichment}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=13cm,keepaspectratio]{Mafessoni_MutSpect.jpg}
+\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
+There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both tests.
+There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
+For the 21,416 variants that pass the strict mask, 629 ( 3\% ) variants are both $Q$-associated and LD-associated; 8,495 variants ( 40\% ) are $Q$-associated but not LD-associated and 12,292 ( 57\% ) are LD-associated but not $Q$-associated.
+For the 14,616 variants that don't pass the strict mask, 650 ( 3\% ) variants are both $Q$-associated and LD-associated; 14,616 variants ( 70\% ) are $Q$-associated but not LD-associated and 5,625 ( 27\% ) are LD-associated but not $Q$-associated.
+\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
+\label{Mafessoni_MutSpect}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_singlePop.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 296 variants associated with $Q$ in the single population tests within the 1000 Genomes Project CHB and CHS, 6 are present in the resequenced data \citep{Lan2017}.}  
+\label{90HanSFS}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=9cm,keepaspectratio]{Han_1kGP_SFS_FullModel.jpg}
+\caption{Site frequency spectrum plot comparing the original 1000 Genomes Project data to the high depth resequence data for variants that, in the 1000 Genomes Project, are both associated with $Q$ and polymorphic in the 83 individuals that were resequenced.
+Among the 6,307 variants associated with $Q$ in the GCAT model including all populations, 1,139 are present in the high depth resequenced individuals.}  
+\label{90HanSFS_full}
+\end{figure}
+
+\begin{figure}[tbp]
 \includegraphics[width=11cm,keepaspectratio]{Omni_AF.jpg}
@@
 \begin{figure}[tbp]
-\includegraphics[width=11cm,keepaspectratio]{MutationSpectrum_cutOff.png}
-\caption{Comparing mutational signatures after removing $Q$-associated variants and after removing individuals with low $Q$.  
-\textbf{A} 
-The  *AC${\rightarrow}$*CC mutational signature in JPT remains despite removing variants associated to quality.
-\textbf{B} 
-Removing individuals with average quality per mapped bases $Q$ below a threshold of 30 removes the mutational signature completely. }
-\label{MutSpect}
+\centering
+\includegraphics[width=15cm,keepaspectratio]{OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the allele frequency difference between individuals with low- and high-$Q$.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Two clusters of are visible, where the majority (92.7\%) of the $Q$-associated variants are more than twice as frequent in individuals with low-$Q$.}
+\label{OverUnder30}
+\end{figure}
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{Published_OverUnder30.jpg}
+\caption{Site frequency spectrum plot comparing the frequency of $Q$-associated variants identified in publications, for individuals with $Q$ scores above and below 30.
+The black dashed lines indicates equal allele frequencies while the red dotted line for variants twice as frequent in individuals with $Q$ scores below 30.
+Each of the rsIDs of the variants are labelled for clarity.}
+\label{Pub_30}
 \end{figure}

### 2019-08-22 LatexManuscript/MBE/legacy.tex
hard coded supp ref

@@
 When mismatch sites are removed from the 1kGP data, the  *AC${\rightarrow}$*CC signal disappears (Figure \ref{SFS}).
-To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Figure \ref{PC1_Correlation}).
+To identify possible technical reasons for the difference, we performed regressions of the prevalence of the  *AC${\rightarrow}$*CC mutational signature against different individual-level quality metrics provided by the 1kGP (see Figure S1).
+%\ref{PC1_Correlation}
 The average quality of mapped bases  $Q$ per individual stood out as a strong correlate : Individuals with low $Q$ show elevated rates of the signature. 
@@
 
-Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Supplementary Figure \ref{histogram_of_sig_snps}).
-The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal enrichment in the Japanese, however other signals identified by Harris and Pritchard appear unchanged including the continental enrichment of *AC${\rightarrow}$*CC signal in East Asians compared to Africans as reported by Harris and Pritchard and replicated in the SGDP (Figure \ref{MutSpect} and \ref{MutSpect_continental}).
+Thus, this mutational signal is heavily enriched in $Q$-associated SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low $Q$ remain unidentifiable using association techniques due to lack of power (Figure S2).
+%\ref{histogram_of_sig_snps}
+The removal of individuals with $Q$ below 30 successfully removes the *AC${\rightarrow}$*CC signal enrichment in the Japanese, however other signals identified by Harris and Pritchard appear unchanged including the continental enrichment of *AC${\rightarrow}$*CC signal in East Asians compared to Africans as reported by Harris and Pritchard and replicated in the SGDP (Figure S3 and S4).
+%\ref{MutSpect} and \ref{MutSpect_continental}
 For population genetic analyses sensitive to the accumulation of rare variants, the removal of individuals with low $Q$ appears preferable to filtering specific low-quality SNPs. 
@@
 \subsection{Identifying suspicious variants in the 1000 Genomes Project}
-The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual} and Figure \ref{quality_by_center}).
+The distribution of $Q$ across 1kGP populations shows that many populations have distributions of $Q$ scores comparable to that of the JPT, especially populations sequenced in the phase 1 of the project: sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual} and Figure S5).
+%\ref{quality_by_center}
 This variability could result from evolving sequence platform and protocols or variation between sequencing centres. 
@@
 We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 23 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
-Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).
+Over 1,165 variants were independently associated to low $Q$ with $ p < 10^{-6}$ in each (Figure S6). %\ref{OverLap}
 
@@
 We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
-These models were broadly consistent (See Figure \ref{CompareModel}).
+These models were broadly consistent (See Figure S7). %\ref{CompareModel}
 
-This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}).
+This method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures S8, S9, S10 and S11). 
+%\ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}
 Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
@@
 
-$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure \ref{CountPerChrom}A).
-Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite normalizing for the number of variants tested per chromosome (Figure \ref{CountPerChrom}B).
-At a 1Mb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure \ref{CountPerChrom}C).
+$Q$-associated variants are distributed across the genome, with chromosome 1 showing an excess of such variants, and other chromosomes being relatively uniform (Figure S12A). %\ref{CountPerChrom}
+Chromosome 1 shows strong enrichments in the  *AC${\rightarrow}$*CC signal compared to other chromosomes despite normalizing for the number of variants tested per chromosome (Figure S12B).
+%\ref{CountPerChrom}
+At a 1Mb scale, we also see rather uniform distribution with a small number of regions showing an enrichment for such variants (Figure S12C). %\ref{CountPerChrom}
 Three outlying 1Mb regions in chromosomes 1, 2 and 17 have over 30 $Q$-associated variants.
-Distribution of association statistics in these regions are provided in Figure \ref{Top10}.
-By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure \ref{CountPerChrom}D).
+Distribution of association statistics in these regions are provided in Figure S13. %\ref{Top10}.
+By contrast, variants that do not pass the 1kGP strict mask are more unevenly distributed across the genome (Figure S12D). %\ref{CountPerChrom}
 
 The mutational 3-mer substitution patterns of this list of variants from the GCAT model is similar to the signature identified in the single population test of the 1kGP JPT in that there is an enrichment in *AC${\rightarrow}$*CC.
-There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT (Figure \ref{strict_gcat_mutational_enrichment}).
+There is also an enrichment of mutations from TAC, TCT, and TGT to the homonucleotide TTT (Figure S14). %\ref{strict_gcat_mutational_enrichment}
 
 A recent publication by Mafessoni et al.
-also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}A).
+also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify 19,196 suspicious variants with 67\% of them passing the 1kGP strict mask \citep{mafessoni2018turning} (Figure S15A).
+%\ref{Mafessoni_MutSpect}
 They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$.
@@
 We also find 23,111 $Q$-associated variants that are not associated to abnormal LD patterns. 
-Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}B).
+Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure S15B).
 These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.
@@
 In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}. 
-Among the 296 such variants that were $Q$-associated in single-population tests for the CHB or CHS, only 6 are present in the resequenced data (Figure \ref{90HanSFS}).   
+Among the 296 such variants that were $Q$-associated in single-population tests for the CHB or CHS, only 6 are present in the resequenced data (Figure S16). %\ref{90HanSFS}   
 This is slightly more but consistent with the 3 false positives predicted according to the $\alpha = 0.01$ nominal false discovery rate. 
@@
 We did a similar analysis using all variants identified in the GCAT model (rather than only variants significantly associated to $Q$ within the CHB and CHS). 
-Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure \ref{90HanSFS_full}).
+Of the 15,270 $Q$-associated variants identified globally, 6,307 are polymorphic in the 1kGP for the 83 resequenced individuals  (See Figure S17). %\ref{90HanSFS_full}
 The vast majority of polymorphisms associated with $Q$ are not present at all in the resequencing dataset, suggesting that they are spurious variants.
@@
 
-Finally, among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure \ref{Omni}). 
+Finally, among the 15,270 $Q$-associated variants, 613 are present on Illumina's Omni 2.5 chip (See Figure S18). % \ref{Omni} 
 These are likely among the small number of variants that are present in the data but exhibit biased genotyping in 1kGP.
@@
 The variants reaching genome wide significance have a star ( * ).
-The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure \ref{Pub_30}).
+The black text colour indicates that this variant is twice as frequent in individuals with $Q$ < 30, grey text colour indicates that these variants are less than twice as frequent in individuals with $Q$ < 30 (See Figure S21).
+%\ref{Pub_30}
 \textsuperscript{\dag} Attention deficit hyperactivity disorder. 
@@
 This suggests that the imputation reference panel still includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual.
-These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure \ref{imputation_score}).
+These $Q$-associated variants could also compromise the imputation of nearby real-variants, however when considering the imputation scores of genotype data from Japanese individuals from Nagahama, there does not appear to be any impact on nearby variant imputation scores (Figure S19).
+% \ref{imputation_score}
 
@@
 These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
-To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{OverUnder30}).
+To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure S20).
+%\ref{OverUnder30}
 The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
-By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
+By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure S21), with three exceptions. %\ref{Pub_30}
 One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
@@
 
-An enrichment in mutational profiles in one chromosome (Figure \ref{CountPerChrom}B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic biases across chromosomes.
+An enrichment in mutational profiles in one chromosome (Figure 12B) is difficult to explain through biological mechanisms or sequencing technology as neither are expected to produce systematic biases across chromosomes.
+%\ref{CountPerChrom}
 Subtle differences in how these data were integrated or processed could explain such biases, as chromosomes are commonly treated and analyzed separately. 
